\chapter*{Acknowledgements}
A number of marvelous people have helped me make this revision possible, and many clever minds produced ideas that I have incorporated into the new system. I cannot hope to properly credit everyone who contributed, knowingly or otherwise, but I can at least make an attempt here. The following individuals have my gratitude:

Connor Haines, for being a great brainstorming partner and excellent roommate.

Zontan Ferrah, for frequently telling me when my ideas were stupid. Sometimes he was even right.

Desmond Henderson, for his remarkably thorough and insightful feedback, and for being the first person brave enough to run a game with Rise.

Kyle McCauley, Linchaun Zhang, and Scott Kottkamp for helping to correct foolish errors that once littered this document and for providing ideas about how to improve the system.

Tarkisflux, for his excellent \href{http://dnd-wiki.org/wiki/Tome_of_Prowess_(3.5e_Sourcebook)}{Tome of Prowess}, feedback on spell design, and more.

Kholai, for keeping me on my toes in our monstrously long discussions of rules minutiae.

Douglas Milewski, for his insightful essay on the nature of fighters. It helped inspire me to think that I could make D\&D right.

Rich Burlew, for providing a forum for house rule enthusiasts like myself to gather and for his insightful articles that guided my thought.

Shannon Carty, for the donation of her vocabulary.

Dave Rosenberg, for his insight into the nature of druids.

Ryan Beck, for his helpful commentary and criticism.

Jordan MacAvoy, for his endless patience and creativity that repeatedly revealed flaws in the wording and design of this book.

Wizards of the Coast, for making a great game and releasing it under the OGL license, which makes all of this tinkering possible.

In addition, the art for this book came from \href{https://huggingface.co/spaces/stabilityai/stable-diffusion}{Stable Diffusion} and \href{https://openai.com/dall-e-2/}{Dall-E 2}, which is very exciting! We live in the future!

The emoji images used in this book come from Twemoji. They are used and redistributed under the CC-BY-4.0 license terms offered by the \href{https://twemoji.twitter.com/}{Twemoji} project.
