\chapter{Items and Equipment}\label{Items and Equipment}

Items are divided into five broad categories:
\begin{raggeditemize}
  \item Weapons are typically used to make \glossterm{strikes}.
    Weapons come in many forms, and can have very different uses in combat.
  \item Armor is worn or wielded to protect yourself.
    There are two types of armor: body armor and shields.
  \item Apparel items are worn on the body, but unlike armor they have no intrinsic defensive properties.
    Rings, cloaks, and all manner of other worn items are considered apparel.
    The only mechanically significant apparel items are \magical.
  \item Implements are typically held while casting spells or using other magical abilities.
    The only mechanically significant implements are \magical.
  \item Tools are used to achieve an effect instead of worn on the body.
    Potions, traps, and magic bags are all tools.
\end{raggeditemize}

\section{Wealth and Item Ranks}\label{Wealth}\label{Wealth and Item Ranks}
  The worth of an item can be measured with money, or with the more abstract concept of an item's rank.
  Money is typically measured with gold pieces, or gp.
  An item with a given rank is typically worth a certain number of gold pieces, as shown in the table below.
  Consumable items are worth a fifth as much as a permanent item of the same rank.

  \begin{columntable}
    \begin{dtabularx}{\columnwidth}{l l X X}
      \tb{Rank} & \tb{Power} & \tb{Typical Permanent Item Price} & \tb{Typical Consumable Price} \tableheaderrule
      0         & 0          & 10 gp or less                     & 2 gp or less \\
      1         & 2          & 40 gp                             & 10 gp        \\
      2         & 4          & 200 gp                            & 40 gp        \\
      3         & 6          & 1,000 gp                          & 200 gp       \\
      4         & 8          & 5,000 gp                          & 1,000 gp     \\
      5         & 10         & 25,000 gp                         & 5,000 gp     \\
      6         & 12         & 125,000 gp                        & 25,000 gp    \\
      7         & 14         & 625,000 gp                        & 125,000 gp   \\
    \end{dtabularx}
  \end{columntable}

\section{Weapons}
  Weapons are used to make \glossterm{strikes} in combat.
  Your class determines which weapons you are proficient with.

  Each weapon deals a specific amount of damage if you hit with it.
  Weapons can also have other special properties to change how you use them.
  For example, many weapons grant a small accuracy bonus, and some weapons are ranged or can be used to attack multiple creatures at the same time.
  There are too many unique weapons to even summarize them here.
  Your character sheet should indicate what weapons you have, and how they modify your attacks.

\section{Armor}
  Armor protects you from incoming attacks.
  Body armor has two main effects: it increases your Armor defense, and it increases your \glossterm{durability}.
  Shields only increase your armor defense, and they require a hand to use.
  Your class determines which armor you are proficient with.

  There are three main types of armor.
  \begin{raggeditemize}
    \item Light: Not very protective, but has few penalties. Typically useful for characters with a high Dexterity, since they can dodge more easily while wearing it.
    \item Medium: Reasonable for most characters to use, with solid protection and only minor penalties.
    \item Heavy: Extremely protective, but heavy body armor reduces your speed by 10 feet. Proficiency with heavy armor is relatively rare.
  \end{raggeditemize}

\section{Magic Items}
  There are a vast array of magical items that exist in Rise.
  If you find a magic item, the GM should explain what it can do, assuming that you can identify it.
  In order to use magic equipment, you have to \glossterm{attune} to it, which costs one of your \glossterm{attunement points}.
  Magic tools, like permanently glowing rocks or bags of holding, don't require attunement.

\section{Other Items}
  There are many other items that you might find or use on your journey.
  Rise doesn't define specific rules for many regular items that exist in the world, like backpacks or ropes.
  Talk to your GM about what makes sense for you to have in your game.
