\chapter{Campaign Settings}\label{Campaign Settings}

\section{Altus}

  % TODO: include world map image

  Altus is a specific world that you can use to run a game of Rise.
  It has a defined geography, countries, and legends.
  Many of its specifics are left somewhat vague, which gives you the freedom to expand on it or modify it to suit the needs of your story.

  Of course, you can also make your own entirely unique world!
  The rules of Rise are not deeply tied to the specific world of Altus.
  Beware that world-building can be difficult and time-consuming, so it's generally best to start small.
  For example, you might find it easier to run a game set in an invented island off the main continent of Altus before you make a whole world from scratch.
  That allows you to fall back on some of the existing countries and conventions if you need them.

  \subsection{The Story of Creation}

    All things began in the Age of Darkness.
    If you could have looked up at the sky back then, ignoring that you didn't have any ground to stand on, you would have seen nothing but a dark void.
    The stars were still there, of course - little discs floating around, same as they are now.
    But not a single one of them emitted the light we are used to.
    They were happy in the darkness, and lulled into a peaceful sleep by the changeless void - all but one.

    One among their number had awoken, and she grew discontent.
    Was there nothing more to creation than this endless stagnation?
    She preached words of change to any nearby stars who could hear her.
    Surely they could make something more exciting if they worked together, she said.
    Her neighbors were slow to rise from their sleep, but nevertheless, she persisted.
    Two of the nearby stars eventually awoke and joined her scheme.

    At the time, none of these stars had names, but this story is going to get confusing if we don't get that settled.
    That persistent first star is now known as as Illumis, the Lightbringer.
    We owe her our life, our light, and even the ground beneath our feet - but that's getting ahead of ourselves.
    The brighter of the two stars that followed her in her little rebellion is called Solaris, the Herald.
    He is the sun that soars overhead each day, making sure our personal world has light and warmth to survive in the Void Beyond.
    The dimmer star is Lunaris, the Disciple.
    She had not the raw power of Solaris nor the revolutionary zeal of Illumis, and her moonlight is but a pale shadow of Solaris's brilliance.
    Yet without her unceasing devotion and protective wisdom, all of their plans would have come to naught.
    We call the three stars together the Triune Astralis.

    Illumis started everything off by igniting into beautiful, transcendent brilliance.
    No one had seen anything before in the whole Age of Darkness!
    Can you imagine having Illumis's ignition be the first thing you ever saw?
    Solaris and Lunaris ignited too, with Solaris's light even outshining Illumis, and everyone woke up pretty quickly after that.
    Even in the most distant corners of the universe, where Illumis's light was a dim glow and Lunaris was invisible, Solaris shined as a beacon that drew attention.

    At first, the other stars were pretty grumpy.
    They had all been pretty happy while asleep, and now they were awake and it was bright and confusing.
    Illumis used her light to tell the universe about her ideas for the future.
    With blinks and flashes and swirls, she painted a picture of a shocking new age.

    Illumis's awakening was too powerful to be contained.
    When she had ignited, some of her lifelight had seeped into her rocky core.
    Now there were little creatures running around on her surface, drinking in her light and living their own chaotic, unpredictable lives.
    The little creatures were fascinating, and she fell in love with them immediately.
    She invited all of the other stars to ignite and begin a new age: the Age of Light, with a sky of twinkling majesty instead of cold darkness.

    Solaris and Lunaris echoed her message, though each contributed their own perspective.
    The creatures that inhabited Solaris were born of flame and heat to match his raw power.
    His fire elementals, as we call them now, bore little resemblance to Illumis's fleshy creatures.
    Lunaris's light was too weak to awaken any inhabitants of her own, and she drifted closer to Illumis to watch and tend to the creatures there.

    The stars were in an uproar over this news.
    Not all of them shared Illumis's interest in creating their own inhabitants.
    However, they could not deny the beauty and novelty of her light and creations.
    Critically, Lunaris showed that they could ignite and be a part of the new age even if they were unable or unwilling to support the chaos of life themselves.
    One by one, they ignited in turn, with a small fraction creating life as Illumis had.

    However, some stars utterly refused to be swayed.
    Without light, they could not communicate at a distance, so they began to drift together.
    As their numbers grew and they became increasingly isolated from the ever-brightening sky, their anger grew in turn.
    Who was Illumis to rewrite the universe in her image?
    Wasn't it better when everything was dark and quiet and peaceful?
    They had to stop this rebellion so everything could go back to the way it was.
    If they destroyed Illumis, the other stars would extinguish themeselves in fear, and the Age of Darkness would return.

    The main problem that the unlit stars faced was simple: they were weak.
    Illumis had been among the strongest of the stars before her awakening.
    Worse, she seemed to have a symbiotic relationship with her infestation of minor life that strengthened her further.
    The unlit stars swore an oath to join together until the Age of Darkness was restored, and became the Voidsworn.

    For all their reactionary inclinations, the Voidsworn ended up invented something too.
    None of the stars had ever spent much time in close proximity to each other during the Age of Darkness.
    As the Voidsworn swarmed and gathered their numbers, they began to fuse together into a single monstrous entity.
    This Voidsworn Amalgam swept through the sky towards the Triune Astralis.
    When they found stars in their path, they attacked and consumed their ignited brethren, joining the corpses of the dead stars into the Amalgam.

    As the Voidsworn carved their ruthless path through the sky, Illumis prepared for the inevitable confrontation.
    She knew the Age of Light could not truly begin until she met them in battle to determine the fate of the universe.
    When the Voidsworn Amalgam arrived, Illumis drew them into a trap.
    She pulled her light, and her life-fuel, into the core of her body, leaving the surface frozen and barren.
    A great age of darkness and cold enveloped her inhabitants, which we call the Long Dark.
    This was a difficult time for our ancestors, as you should know from the Old Histories!

    The Voidsworn Amalgam surged into Illumis's core, trying to search out and destroy her center of power.
    This was when her trap was fully revealed.
    She made the ultimate sacrifice, burning away all of her power and light forever to forge her body into a mighty world-cage.
    The Voidsworn found themselves trapped in the center of her sphere, surrounded on all sides by unbreakable walls.

    They pushed and smashed and tried to break out of the cage.
    Each mighty blow against Illumis's corpse warped the land, raising mountains as scars.
    Altus was the center of their efforts, and they pushed it so far out from the core that they nearly broke free.
    But Illumis's world-cage held, and the Voidsworn were trapped forever.
    They still live in her core now now, though they have only a fraction of their original power.
    Their efforts to escape sometimes cause great earthquakes.

    When the world settled, Solaris and Lunaris approached Illumis's corpse.
    She had foreseen her demise, and given them instructions to keep her dream alive.
    Solaris now provides the light that Illumis cannot, keeping our world warm and safe.
    He burns bright and strong, but he must rest each day.
    Lunaris keeps us company as well, though she is not content to look after only one star.
    She makes sure to give us her full attention each month, but the rest of the time we only see part of her radiance, since she is busy looking at faraway stars.

    \subsubsection{The Details of Creation}

      The general outline of the Story of Creation, as presented above, is known and agreed on throughout Altus.
      However, each culture has a different interpretation of some specific details which are important to them.
      The order of creation is a central point of contention, especially as it relates to different species and regions.
      Since recorded history only begins after the end of the Long Dark, there is no way to tell which version is true.

      For example, the official story told by the elves in the Vastwoods is that Illumis originally created the elves as the first, perfect being.
      They call the long rule of elves as the sole sentient species the Tranquiline Age, and mark its end shortly before the start of the Long Dark.
      Illumis knew that the world would need hardier, less perfect beings to survive the difficult times ahead, and the elves would need to be strengthened by competition with lesser foes.
      Therefore, she weakened and warped her light of creation, spawning all manner of beasts and lesser sentient creatures like humans and dwarves.
      Knowing that they would need to survive the Long Dark and the intermittent light from Solaris after her death, she gave them the gift of sleep so they could preserve their limited energy.
      Only elves continued to be sleepless, since they were created from her greater light at the dawn of time.

  \subsection{Founding Gods and Lesser Deities}
    The Triune Astralis - Illumis, Solaris, and Lunaris - are the founding gods of the world.
    All mortals owe them tribute, and they have temples in every civilized area.
    However, their domains are extremely broad.
    Solaris rules the day, Lunaris rules the night, and Illumis is an even more abstract creator figure.
    Since Illumis is dead, she can empower no clerics of her own.
    This leaves space for a whole host of lesser deities who claim dominion over specific aspects of the world, and who empower their own personal clerics.

    Rise does not precisely mirror real-world polytheism, but it is not entirely different either.
    At the risk of dramatic oversimplification, polytheism typically involves recognizing a mixture of deities of widely varied power and scope.
    People typically do not have a single favored deity that they worship above all others in all contexts.
    Instead, they give respect, tribute, or gratitude to deities that are relevant to their current situation.
    You might offer a sacrifice to the god of travel before making a journey, offer a sacrifice to the god of the hearth upon returning home, and so on.

    Rise has a wide variety of deities with specific domains, and most people offer respect to relevant deities at appropriate times.
    However, unlike the real world, Rise places a great importance on the concept of a ``patron deity''.
    Mortals who worship a specific deity are rewarded by going to an afterlife ruled by that specific deity.
    This is often preferable to going to a generic alignment-appropriate afterlife, especially for evil characters.

    For their part, deities draw power from the worship of mortals, and especially from claiming the soul energy from mortals who end up in the deity's afterlife.
    This makes deities generally invested in finding ways to increase their base of mortal worshippers.
    They use their clerics accomplish this goal, generally by spreading awareness of the deity's domain and influence.
    Famous clerics act as living proof of the deity's power, and many temples offer healing services to anyone in need.
    Most deities avoid directly converting claimed followers of other significant deities to avoid inter-deity conflict.

    Of course, clerics can also influence the mortal world to make life there match the deity's preferences.
    Clerics of Chavi might hold storytelling competitions, and clerics of Raphael might act as bounty hunters to hunt down criminals who escaped justice.
    On a more sinister note, clerics of Daeghul might offer human sacrifices to channel the soul energy of the dying creatures towards their deity.

\section{Praxis}
  The world of Praxis is a limited one, fraught with danger in the wilderness beyond a small region of safety and civilization.

  \subsection{History}

    \subsubsection{The Apocalypse}
      A thousand years ago, the world was destroyed in a great magical apocalypse.
      Civilization survived in only one place: Praxis, a city sheltered by a massive and mysterious shield-dome.
      The details of this event, and the nature of the world before the apocalypse, are lost to time.
      Most of the life in the world was also wiped out in the apocalypse.
      Of the species that survived, most were kept safe in Praxis and released into the wild after it ended.
      A rare few specimens of other species apparently also survived, but no living creature was alive before the apocalypse - even of species with lifespans that can stretch that far back, such as dragons.

    \subsubsection{The Long Reconstruction}
      In the aftermath of the world's destruction, the inhabitants of Praxis decided to ban magic after the devastation it had caused.
      For about nine nundred years, this worked fairly well, and Praxis very slowly expanded and rebuilt its civilization.
      Most of the interior of the dome was devoted to farmland during that time.
      Praxis developed its own government and guilds, and the four major power sources in Praxis developed at this time: the city government, the Boundary Wardens, the Nullwardens, and the Landowners' Association.
      The Boundary Wardens and Nullwardens made efforts to expand the influence of Praxis beyond the dome and into the untamed wilds beyond.
      However, their progress was slow and dangerous thanks to the monsters that had taken over the wilds beyond Praxis.

      Periodically, Praxis encountered civil unrest or the threat of hidden magic users violating the city's prohibition.
      The greatest punishment the city imposed on its worst members was exile.
      Many of those exiled from the city died in the inhospitable world beyond, but there are some groups outside Praxis that can trace their origins back hundreds of years before the Great Harrowing to groups of people exiled from Praxis.
      Most of these long-lived groups are druidic circles.

    \subsubsection{The Great Harrowing}
      Nine hundred and twenty years after the apocalypse, Praxis was attacked by a pair of massive red dragons.
      They flew in through the open gates, destroyed the city's few static weapons that could pose them any threat, and began torching the city and all of its farmland.
      The city tried to defend itself, but without magical defenses they had little ability to stop the dragons.
      Although Nullwardens had found success slaying dragons before in more controlled circumstances, their near invulnerability to magic could only stop the dragons from killing them - not from destroying the rest of the city around them.

      In Praxis's darkest hour, it was finally saved by people who had been concealing their magical powers from Praxis's ban: the Eldritch Knights and the Divine Chorus.
      The Eldritch Knights were an order of mage-knights whose founders had been exiled from Praxis centuries ago.
      When they discovered that Praxis was under attack, they had come to help defend it.
      They were joined by the clerics of the major churches in Praxis, who had been concealing their divine power.
      They did not immediately claim a specific name for their alliance, but in the years since the Great Harrowing they have become known as the Divine Chorus.

      The Eldritch Knights and the Divine Chorus with the remaining defenders of the city to seal the gates of Praxis to trap the dragons inside.
      Then, in a fearsome magical battle, they defeated the dragons and saved the charred remains of the city.

    \subsubsection{The Era of Rebirth}
      In the aftermath of the Great Harrowing, much of the old power structure of the city was destabilized, and many of its previous power holders were dead.
      When the city started to function normally again, the first major act was to revoke the ban on magic in gratitude for the role that magic users had played in saving the city.
      The power offered by magic dramatically accelerated the rate of Praxis's reconstruction and expansion.
      For the first time in centuries, Praxis was able to make significant progress against the monsters at its gates.

      Eighty years have passed since the Great Harrowing, and in that time, Praxis has pushed most of its farmland outside of the city.
      The inside of the city is being developed and magically renovated, and living quarters are becoming far less crowded for most of the city's inhabitants.
      The power of the Nullwardens has waned with the acceptance of magic, though they still argue strenuously against its dangers and use each magic-wielding criminal as evidence that magic should be more aggressively controlled - or banned once more.
      The Boundary Wardens are stretched thin patrolling the ever expanding territory, and they are eager to recruit anyone with the skills to help them keep the peace in the wilderness and ensure that the road networks are safe.

  \subsection{The City of Praxis}
    The city of Praxis occupies a nearly perfect circle with a ten mile radius, all of which is protected by an immense dome.
    The city sits atop a massive plateau with a thin rim around the outside.
    Roads zigzag up the side of the plateau to reach each of Praxis's three gates.

    \subsubsection{The Shield-Dome}
      The shield-dome over Praxis is made of a mysterious metal alloy that is harder than any known substance.
      Though it is apparently mundane in nature, no known weapon or spell has ever been able to harm it.
      The gates to the city are massive doors that can swing open and closed, and were apparently part of the dome at the time of its construction.
      For many centuries, the entire dome has been magically invisible, allowing sunlight to enter the city.
      The gates have only closed once in that time - during the Great Harrowing, to trap the dragons inside the city.

      Most people think that the dome is a magical forcefield, like an immense \spell{wall of force}, instead of physical metal that is rendered invisible.
      Of the few who do know the dome's true nature, even fewer know what sustains the invisibility effect.
      The power to sustain the dome's invisibility comes from an ancient magical device deep in the Praxis Labyrinth.

    \subsubsection{The Praxis Labyrinth}
      The plateau that supports Praxis also contains an immense labyrinth.
      An ancient magic makes the labyrinth impossible to navigate by confusing both the minds of intruders and the physical space they walk through, creating twisted and impossible passages that layer over themselves.
      Many people have tried to explore it, but the labyrinth has never been navigated within recorded history.
      Only the most learned historical scholars know of reports centuries ago that an unknown woman walked out of the labyrinth without having entered it.
      She was never seen again.

  \subsection{Modern Geography}
    Praxis is the only major city.
    It sits at the intersection of the four major biomes: the Northern Forest, the Southern Forest, the Expanse, and the Storm Peaks.
    There are a number of farmsteads expanding from Praxis towards the Northern Forest and Southern Forest, and several towns of importance outside Praxis.

  \subsection{Major Organizations}

    \subsubsection{Boundary Wardens}
      The Boundary Wardens are responsible for patrolling the wilderness around Praxis, maintaining the safety of the road networks, and generally keeping the boundary between civilization and nature intact.
      They have legal juridisction over all areas outside of the Praxis shield-dome.

    \subsubsection{City Guards}
      The city guards are responsible for maintaining peace and order within Praxis.
      They have legal juridisction over all areas within the Praxis shield-dome.

    \subsubsection{Landowners' Association}
      The Landowners' Association is a loose collection of rich merchants and real estate owners within Praxis.
      They have no formal responsibility or juridisction, but the wield great wealth to accomplish their objectives, which mostly revolve around acquiring greater wealth.

    \subsubsection{Nullwardens}
      The Nullwardens are a group of strictly mundane warriors who fight against the dangers of magic.
      All Nullwardens above the entry ranks are Null, making them virtually immune to magic.
      Centuries ago, the Nullwardens were indisputably the most important power brokers in Praxis, surpassing even the legal government.
      They were a mix of elite mercenaries and witch hunters, searching for magical dangers both within Praxis and beyond its walls.
      After their failure to protect Praxis during the Great Harrowing, their power has waned, though they are still deeply influential.
      Since magic is now legal, they spend less time hunting for mages in Praxis and more time advocating for political change.

      Unlike the Boundary Wardens or the city guard, the Nullwardens do not primarly define their influence in terms of territory.
      The Nullwardens send surgical strike teams anywhere in the world to stop magical threats.
      To the extent necessary, they generally obey the authority of whoever claims legal jurisdiction over the territory they have to operate in, and any extra-legal operations they undertake are well hidden.

\section{Dreadbore Isles}
  The world of Dreadbore Isles is composed of a ring of islands that surround an immense whirlpool named the Dreadbore.

  \subsection{Geography}
    The islands surrounding the Dreadbore vary significantly in their size and culture, and the influence of the Dreadbore tends to make it impractical for any given island to have significant contact with any islands beyond its close neighbors.

    Beyond the ring of islands lies the Great Ocean.
    Its depths are beyond reckoning, and no significant land mass has ever been discovered outside of the ring of islands.
    The deep waters of the Great Ocean also contain very limited sea life, especially compared to the rich variety around the islands.

  \subsection{Island Descriptions}

    \subsubsection{Dragonsguard}
      The island of Dragonsguard is ruled by dragons.
      Most ships that approach the island are destroyed on sight by dragons, regardless of their intentions.
      There are two known exceptions that the dragons allow to pass: ships carrying an immense wealth of treasure to offer as tribute to the dragons, and ships crewed entirely by dwarves.
      The tribute ships are allowed to leave Dragonsguard unscathed, save for the loss of their tribute.
      Dwarf-crewed ships have been observed to reach the island unscathed, but the ships never return, and their crews are never seen again.

      The few people who have been to Dragonsguard and returned on tribute ships tell fantastical stories of a draconic paradise, filled with works of magical genius and incredible wealth.
      However, they have never seen dwarves on the island, and no one knows what happens to the dwarf-crewed ships.

    \subsubsection{Darren's Land}
      The island of Darren's Land is flat and dry, and long ago it was an uninhabited desert.
      Thanks to the mighty and continual efforts of the druidic Circle of Unity, a collective representing the variety of druid circles present on the island, the barren land provides enough food to sustain the island's population.
      This gives the Circle of Unity far more influence on the politics of civilization than druids normally possess, which has posed challenges for both the civilization on the island and the integrity of the druids who wield that power.

      Long ago, a mighty warrior named Darren came to the island and cleansed it entirely of monsters.
      He sought to establish a utopia free from any danger, where common people could explore any part of the island freely.
      Although he succeeded at rendering the island safe, he failed to render it inhabitable, and was forced to flee the island to avoid starvation.
      Decades later, he returned with a group of druids and colonists to establish a sustainable civilization there, and it has persisted ever since.

    \subsubsection{Elysland}
      The island of Elysland is overgrown with an ancient forest.
      It is divided into five forest-dwelling nations.
      The different nations have deep divisions in their attitudes towards society, and some share ancient grudges against each other.
      However, all are unified in their reverence for the towering trees that shade and protect them, and they will put aside their differences when necessary to push back against outsiders who would defile the forests.

    \subsubsection{Reinward}
      Reinward is one of the largest islands, and is the most thoroughly developed and civilized.

    \subsubsection{The Savage Lands}
      The Savage Lands are bordered by mountains on all sides but the easternmost side, which extends deep into the Dreadbore.
      Only the largest and most well-crewed ships are capable of navigating that port, and no ship built on the Savage Lands has ever escaped to safety.
      It is not a barren island, but it is an unwelcoming one.
      There are three main power groups on the island: the orc tribes, the undead, and the Old Fae.

\section{The Soulforge}
  The Soulforge is a plane of infinite smooth, grey ground stretching to every horizon.
  The only location of any distinction or relevance on the plane is the Soulspire.
  It is a metal tower, five hundred feet across at the base and one mile high.
  The outside of the tower is adorned with beautiful and intricate sigils in an unknown language.
  Around the base of the tower, a small encampment of doomed souls repeatedly struggle to ascend the Soulspire, die in the process, and take corporeal form once again.

  \subsection{The Trapped Souls}
    The Soulspire seems to function similarly to an unusually limited afterlife.
    Every creature surrounding the Soulspire is a soulforged made from the essence of the plane.
    As normal, when a soulforged creature of strong will dies, they can maintain the cohesion of their soul and reconstitute a new body from the essence of their native plane.

    A long-standing legend based on murals at the base of the tower claims that anyone who reach to the top of the Soulspire can find their heart's desire, which is generally interpreted to mean leaving the plane.
    Many inhabitants of the Soulspire take advantage of their effective immortality to lay reckless seige to the Soulspire, returning after each demise as long as their willpower holds.

    There is a semblance of culture at the tower's base.
    It shifts slowly, as over long centuries the old inhabitants give up on maintaining their individuality and become part of the plane itself.
    Traditionally, this is done by attempting one final tower ascent and not returning from the inevitable death that follows, though some have wanderered off into the grey void outside the tower and never returned.
    Whenever an old soul gives up its sense of self, a new arrival appears within a day.
    There are three classes of people in Soulforge: the ascendants, who actively try to climb the tower; the fledglings, who have never tried; and the sages, who once sought to climb the tower and have now given up.

  \subsection{The Impossible Ascent}
    Each floor of the Soulspire is a separate demiplane, most of which are far larger than the tower's radius would suggest.
    There is always a portal on each floor to ascend to the next level, and there are often windows in incogruous locations that look out onto the original plane.
    The view from each window has an elevation about twenty feet higher than the previous window, and looks out onto the Soulforge.
    The windows can be seen from outside the Soulspire, and ascendants looking out can be observed, through their surroundings are never visible - only the ascendants themselves.
    This allows the fledglings and sages at the base to track the progress of ascendants.
    It gives the ascendants confidence that the Soulspire is finite, and that actually reaching the top is an achievable goal - though not an easy one.
    Based on the exterior windows, the Soulspire has two hundred and fifty-one floors.
    No ascendant has ever passed the fiftieth floor.

    Aside from the constant presence of the portal and the windows, each floor of the Soulspire can be dramatically different.
    In addition to having highly varied terrain, each floor can have different gravity or otherwise break different fundamental laws.
    In general, the floors seem to expand in size and complexity as their apparent height in the tower increases.

    Almost all floors are populated with soulforged imitations of creatures native to other planes.
    These creatures are almost universally hostile.
    In the unusual event that the creatures communicate with ascendants, they have never displayed any knowledge of the Soulspire's nature or of any world beyond their current floor.

    The Soulspire is entered through a large, golden gate at its base.
    It is activated by placing a hand against it for thirty continuous seconds.
    Once it activates, it remains open for thirty seconds before closing.
    Portals within the Soulspire behave in the same way, though their color and size may vary significantly.
    During the time that the portals are open, any ascendants who enter it always arrive at the same floor as each other.
    On rare occasions, ascendants may arrive at different locations in the next floor, but they usually arrive in the same location as each other.
