% TODO: Condition has a specific meaning now; move this content to Glossary
\chapter{Conditions}\label{Conditions}

\condition{blinded} A blinded creature cannot see. It automatically fails at actions which depend on vision, including simply seeing the locations of other objects and creatures (but see \pcref{Awareness}). It has a 50\% miss chance with \glossterm{strikes} and vision-related checks, even if it knows the location of its target. Finally, it is \defenseless.

%\condition{blown Away} Depending on its size, a creature can be blown away by winds of high velocity. A creature on the ground that is blown away is knocked down and rolls 1d4 \x 10 feet, taking 1d4 points of nonlethal damage per 10 feet. A flying creature that is blown away is blown back 2d6 \x 10 feet and takes 2d6 points of nonlethal damage due to battering and buffering.

\condition{bloodied} At or below half hit points. Bloodied creatures take a \minus4 penalty to Fortitude and Mental defense.

\condition{charmed} A charmed creature is mentally influenced to like another creature.
It always sees the words and actions of the creature that charmed it in the most favorable way, as a close friend or trusted ally.
A charmed creature cannot be controlled like an automaton, but can be persuaded to take particular actions with the Persuasion skill (see \pcref{Persuasion}).
It treats the creature that charmed it as a friend (a \plus10 relationship modifier) for the purpose of Persuasion checks.

%\condition{checked} Prevented from achieving forward motion by an applied force, such as wind. Checked creatures on the ground merely stop. Checked flying creatures move back a distance specified in the description of the effect.

\condition{confused} A confused creature is unable to independently control its actions. \confusionexplanation

\condition{crouching} A crouching creature gains a \plus1 bonus to physical defenses against ranged attacks.
However, it takes a \minus1 penalty to physical \glossterm{accuracy} with melee attacks and physical defenses against melee attacks, and moves at half speed.

\condition{dazed} A dazed creature acts during the \glossterm{delayed action phase} instead of the \glossterm{action phase}.

\condition{dazzled} A dazzled creature has difficulty seeing.
It loses any special vision abilities it has, such as \glossterm{darkvision} or \glossterm{low-light vision}.
In addition, it takes a \minus2 penalty to \glossterm{accuracy} with \glossterm{strikes} and vision-related checks.

\condition{dead} A dead creature's soul leaves its body. Dead creatures cannot benefit from normal or magical healing, but they can be restored to life via magic (see \pcref{Resurrecting the Dead}). A dead body decays normally unless magically preserved.

\condition{deafened} A deafened creature cannot hear. It automatically fails at actions which depend on hearing. In addition, it has a 20\% failure chance when casting any spell with verbal components.

\condition{defenseless} A defenseless creature is unable to defend itself in melee combat. It takes a \minus2 penalty to physical defenses against melee attacks. Any creature not capable of using a weapon or shield to defend itself, such as most unarmed creatures, is defenseless.

\condition{disabled} A disabled creature has no hit points remaining.
It takes a \minus4 penalty to \glossterm{accuracy} and checks.

\condition{disoriented} During each movement phase, a disoriented creature is compelled to move its full speed in a random direction.
It moves as far as it can, but will not sprint or take similar strenuous actions to increase its speed.

\condition{dominated} A charmed creature is mentally compelled to obey another creature.
It obeys the commands of the creature of the dominated it unquestioningly, as an automaton.
If it does not understand the language of the creature that dominated it, it still attempts to obey as much as possible, and simple commmands (such as ``attack'' or ``follow'') can usually be communicated successfully.

\condition{dying} A dying creature is unconscious and near death. See \pcref{Dying}.

\condition{encumbered} An encumbered creature has its motion restricted by armor or weight. It may be unable to use certain class ability and abilities which require free motion. See \pcref{Encumbrance} for details.

\condition{exhausted} An exhausted creature moves at half speed, cannot sprint (see \pcref{Sprint}), and takes a \minus2 penalty to \glossterm{accuracy}, checks, and defenses.

\condition{fascinated} A fascinated creature can take no actions. It remains in place, giving its total attention to some object, creature, or effect. It takes a \minus4 penalty to skill checks made as reactions, such as Awareness checks. If the creature notices any threat against it, such as an approaching enemy, it is no longer fascinated.

\condition{fatigued} A fatigued creature moves at half speed and cannot sprint (see \pcref{Sprint}).

\condition{frightened} A frightened creature takes a \minus4 penalty to \glossterm{accuracy} and checks as long as it is within \rngmed range of the source of its fear.
These penalties do not stack with the penalties for being \shaken or \panicked.

If the source of a frightened creature's fear is a creature and is \glossterm{defeated}, this effect is broken.

\condition{goaded} A goaded creature wants to attack the creature that it is goaded by.
If it is within \rngmed range of the taunting creature, it is takes a \minus2 to all attacks that do not directly affect that creature.
If that creature is rendered \helpless, surrenders, or is otherwise unable to fight, this effect immediately ends.
The \conditionlink{taunted} condition is a more severe version of this effect.

\condition{grappled} A grappled creature is in wrestling or some other form of hand-to-hand struggle with one or more attackers. While grappled, you suffer certain penalties and restrictions, as described below.

\begin{itemize}
    \item You must use a free hand (or equivalent limbs) to grapple, preventing you from taking any actions which would require having two free hands. If you cannot free a hand, you suffer a \minus10 penalty to accuracy with all physical attacks, including grapple attacks, until you have a free hand.
    \item You take a \minus4 penalty to physical defenses against creatures you are not grappling with.
    \item You take a \minus4 penalty to accuracy with weapons that are not light, since they are too large and cumbersome to be used effectively in a grapple.
    \item Spellcasting is extremely difficult. You cannot cast spells with somatic components. Casting a spell without somatic components requires a Concentration check with a DR equal to 20 \add double spell level (see \pcref{Concentration}).
    \item You cannot move normally (but see Move the Grapple, below).
\end{itemize}

Other than the restrictions listed above, you can act normally. You can also try to move the grapple, escape the grapple, or pin your opponent. See \pref{Grapple} for more information.

\condition{helpless} A helpless creature is completely at an opponent's mercy. Its physical defenses are equal to 10 \add its size modifier. Paralyzed, bound, and unconscious creatures are helpless. Helpless creatures can be killed instantly by a coup de grace (see \pcref{Coup de Grace}).

\condition{ignited} An ignited creature has been set on fire.
It takes 1d6 fire damage at the end of each round, and takes a \minus2 penalty to \glossterm{accuracy}, checks, and defenses.
As a move action, an ignited creature can make a DR 10 Dexterity check to put out the flames.
This action requires a free hand.
Dropping prone as part of the action gives a \plus5 bonus on this check.

\condition{immobilized} An immobilized creature can't move out of the space it was in when it became immobilized. Immobilized flying creatures that have the ability to hover can maintain their initial altitude. All other flying creatures subjected to this condition descend at a rate of 20 feet per round until they reach the ground, taking no falling damage.

\condition{invisible} An invisible creature or object cannot be seen. Creatures unable to see an invisible creature are \defenseless against its attacks. Attackers suffer a 50\% miss chance even if they know the location of the invisible creature. See \pcref{Awareness}, and \pcref{Stealth}, for how to identify invisible creatures.

%\condition{knocked Down} Depending on their size, creatures can be knocked down by winds of high velocity. Creatures on the ground are knocked prone by the force of the wind. Flying creatures are instead blown back 1d6 \x 10 feet.

\condition{nauseated} A nauseated creature takes a \minus4 penalty to \glossterm{accuracy}, checks, and defenses.
These penalties do not stack with the penalties for being \sickened.

\condition{panicked} A panicked creature takes a \minus4 penalty to \glossterm{accuracy} and checks as long as it is within \rngmed range of the source of its fear.
In addition, it must flee from the source of its fear by any means necessary if it is within that range.
If unable to flee, it must do nothing other than take the \glossterm{total defense} action every round (see \pcref{Total Defense}).
These penalties do not stack with the penalties for being \shaken or \panicked.

If the source of a panicked creature's fear is a creature and is \glossterm{defeated}, this effect is broken.

\condition{paralyzed} A paralyzed creature is unable to take physical actions. It has effective Dexterity and Strength scores of \minus10 and is \helpless, but can take purely mental actions. This can cause flying creatures to crash, swimming creatures to drown, and so on. A creature can move through a space occupied by a paralyzed creature -- ally or otherwise. Each square occupied by a paralyzed creature, however, counts as 2 squares.

\condition{petrified} A petrified creature has been turned to stone. It is neither alive nor dead, but is unconscious and unable to take actions, and its body is an inanimate statue. If the statue is broken or damaged before the creature is restored to its original state, the creature has equivalent damage or deformities.

\condition{pinned} A pinned creature is held completely immobile in a grapple.
The only physical actions it can make are to escape the grapple (see \pcref{Grappling}).
Like a \glossterm{helpless} creature, its physical defenses are equal to 10 \add its size modifier.
Unlike a helpless creature, a pinned creature cannot be killed instantly by a coup de grace.

\condition{prone} A prone creature is lying on the ground, rather than standing normally.
It takes a \minus2 penalty to \glossterm{accuracy} with physical melee attacks and physical defenses.
It gains a \minus2 bonus to physical defenses against ranged attacks.
A creature can stand up from being prone instead of moving during the movement phase.
This generally requires one free hand.

\condition{shaken} A shaken creature takes a \minus2 penalty to \glossterm{accuracy}, checks, and defenses as long as it is within 100 feet of the source of its fear.
These penalties do not stack with the penalties for being \frightened or \panicked.

If the source of a shaken creature's fear is a creature and is \glossterm{defeated}, this effect is broken.

\condition{sickened} A sickened creature takes a \minus2 penalty to \glossterm{accuracy}, checks, and defenses.
These penalties do not stack with the penalties for being \nauseated.

\condition{slowed} A slowed creature cannot act during the movement phase, and moves at half speed.

\condition{squeezing} A squeezing creature is trying to move though an area too small for it to fight in normally.
While squeezing, a creature moves at half speed and takes a \minus2 penalty to physical accuracy, physical checks, and physical defenses.
For details, see \pcref{Squeezing}.

\parhead{Stable} A creature who was dying but who has stopped losing hit points and still has vital damage is stable. The creature is no longer dying, but is still unconscious. See \pcref{Stable}.

\condition{staggered} A staggered creature is temporarily overwhelmed by physical trauma.
It takes a \minus4 penalty to attack, checks, and defenses.
The first time you become \glossterm{bloodied} in an encounter, you become staggered until the end of the next round.

\condition{stunned} A stunned creature cannot take any actions during the \glossterm{action phase} or \glossterm{delayed action phase} except the \textit{recover} and \textit{desperate recovery} actions (see \pcref{Recover}, and \pcref{Desperate Recovery}).

\condition{taunted} A taunted creature is compelled to attack the creature that it is taunted by.
If it is within \rngmed range of the taunting creature, it takes a \minus4 penalty to \glossterm{accuracy} with all attacks that do not directly affect that creature.
If that creature is rendered \helpless, surrenders, or is otherwise unable to fight, this effect immediately ends.
The \conditionlink{goaded} condition is a less severe version of this effect.

\condition{unaware} An unaware creature does not know that it is being attacked. Successful physical attacks against an unaware creature automatically threaten critical hits. After being attacked, an unaware creature typically stops being unaware of future attacks, even if cannot see or identify its attacker.

\condition{unencumbered} An unencumbered creature does not have its motion restricted by armor or weight. See \pcref{Encumbrance} for details.
