\chapter{Magic Items}

Magic items are objects that have been imbued with magical energy. They can take almost any form, and their potential uses are only as limited as the magic that created them.

\section{Magic Item Types}
    Magic items are divided into three broad categories:
    \begin{itemize}
        \item Apparel items provide access to their abilities while worn.
            A \mitem{flaming burst full plate} and a \mitem{ring of protection} are apparel items.
        \item Weapons are used to make physical attacks. They provide access to their abilities when wielded.
            A \mitem{\plus1 flaming longsword} and a \mitem{\plus3 vampiric scythe} are weapons.
        \item Implements are used to cast spells. They provide access to their abilities when wielded and attuned to.
            A \mitem{\plus2 staff of fire} and a \mitem{\plus5 staff of time} are implements.
        \item Tools provide access to their abilities when used in some way.
            A \mitem{bag of holding} is a tool.
    \end{itemize}

\section{Using Magic Items}

    To use a magic item, it must be activated, although sometimes activation simply means wearing or holding the item.

    \subsection{Daily Item Activations}
        In general, you can activate any combination of magic items you possess a number of times per day equal to half your level (minimum 1).
        Once you have used up your activations for the day, you can't activate any more magic items until the next day, though you can continue to use items that don't require activation (such as most magic weapons) normally.

    \subsection{Activation Methods}
        Magic items can be activated in one of four ways:

        \begin{itemize}
            \item Command word
            \item Specific action
            \item Spell completion
            \item Triggered
        \end{itemize}

        These methods are described below.

        \parhead{Command Word} A character must speak a special word defined by the item to activate it.
        Unless otherwise stated, activating a command word magic item is a standard action.

        \parhead{Specific Action} A character must perform a specific action defined by the item to activate it.
        For example, a creature might need to drink the item, wrap its cloak around itself, or perform some other task.
        The activation time for such items can be varied.
        Unless otherwise stated, activating a specific action magic item is a standard action.

        \parhead{Spell Completion} A character must complete a spell defined by the item to activate it.
        Unless otherwise stated, activating a spell completion magic item is a standard action.
        Both verbal and somatic components may be required, as appropriate to the spell to be completed.

        In order to activate a spell completion item, a character must be able to cast spells of that level and she must have that spell on his spell list.

        \parhead{Triggered} A creature must fulfill some triggering condition defined by the item to activate it.
        For example, a triggered magic item might activate when a character strikes a foe, is damaged, or is affected by a particular kind of magic.
        Unless otherwise stated, activating a triggered item is an immediate action.
        It is done completely mentally, requiring no physical action, so it can be done even while paralyzed.
        Some triggered items activate automatically, without requiring an action of any kind.

\section{Magic Item Effects}
    Most magic items either provide numerical bonuses or emulate the effect of a spell in some way.

    \subsection{Item Power}\label{Item Power}
        The strength of an item's effects depends on its item power.
        If the item is not being used or worn by a creature, its item power is equal to its base item power.
        If it is being used or worn by a creature, its item power is equal its base item power or the level of the creature, whichever is higher.

        In addition to modifying an item's special effects, item power also affects an item's defenses.
        An item's Fortitude and Mental defenses are equal to 10 \add its item power.
        Its Armor defense and Reflex defense are not affected by item power, and are solely determined by the item's size and shape.

    \subsection{Removing Magic Items}
        Unless otherwise noted, magic items that have effects on the creature using the item must continue to be worn or held as long as the effect lasts.
        If a magic item is removed before its effect's duration is up, the effect ends.
        Items which are consumed when used or which do not affect their user are unaffected by this rule.

    \subsection{Special Attacks}
        If a magic item requires a successful special attack to have its full effect, the accuracy is listed in the item's description.
        Typically, the accuracy is equal to the item's power.

\section{Item Description Format}

    Each general type of magic item gets an overall description, followed by descriptions of specific items.

    General descriptions include notes on activation, random generation, and other material.
    The AD, hardness, hit points, and break DR are given for typical examples of some magic items.
    The AD assumes that the item is unattended and includes a \minus10 penalty for the item's effective Dexterity of \minus10.
    If a creature holds the item, use the creature's Dexterity in place of the \minus10 penalty.

    Some individual items, notably those that simply store spells and nothing else, don't get full-blown descriptions.
    Reference the spell's description for details, modified by the form of the item (potion, scroll, wand, and so on).
    Assume that the spell is cast at the minimum level required to cast it.

    Items with full descriptions have their abilities detailed, and each of the following topics is covered in notational form at the end of the description.

    \parhead{Minimum Power} The next item in a notational entry gives the minimum power of the item.
    An item's minumum power can affect its item power when used (see \pcref{Item Power}).

    For potions and scrolls, the item's minimum power must be at least twice the level of the spell contained.
    Generally, an item's minimum power is the same as the minimum spellpower required to create the item.

    \parhead{Aura} Most of the time, a Spellcraft check will reveal the school of magic associated with a magic item and the strength of the aura an item emits.
    This information (when applicable) is given at the beginning of the item's notational entry.
    See the Spellcraft skill for details.

    \subparhead{Ability Tags} Unless otherwise noted, an item's abilities have all tags included in the item's aura.

    \parhead{Requirements} The qualifications that must be met to create the item, %described in \pcref{Creating Magic Items}.

    \parhead{Market Price} This gold piece value, given following the word ``Price,'' represents the price someone should expect to pay to buy the item.
    The market price for an item that can be constructed with an item creation feat is usually equal to the base price plus the price for any components.

    \parhead{Cost to Create} The next part of a notational entry is the cost in gp to create the item, given following the word ``Cost.'' This information appears only for items with components which make their market prices higher than their base prices.
    The cost to create includes the costs derived from the base cost plus the costs of the components.

    Items without components do not have a ``Cost'' entry.
    For them, the market price and the base price are the same.
    The cost in gp is 1/2 the market price.

    \parhead{Weight} The notational entry for many wondrous items ends with a value for the item's weight.
    When a weight figure is not given, the item has no weight worth noting (for purposes of determining how much of a load a character can carry).

\section{Apparel}

    \subsection{Apparel Item Limitations}

        There are a wide variety of magic items, but there is a limit to how many apparel items a character can wear at any given time.
        For humanoid-shaped creatures, there are four core areas on the body that can hold magic items: the arms, the head, the torso, and the legs.
        A creature can wear only one item in the hands, arms, and legs, but it can use two torso items, provided that they are not identical (such as two belts).
        In addition, it can wear a suit of armor and two magic rings (one on each hand), for a total of eight item slots.
        Any additional items worn do not function.

        Creatures with non-humanoid body structures may have different arrangement of items.
        For example, a wolf would be able to wear items on its head, its torso, and on two different sets of legs.
        Regardless of body shape or size, most creatures may not have more than eight item slots, and many have fewer.
        A list of common items and their placement on the body is given below.

        \begin{itemize}
            \item Arm items:
                \begin{itemize}
                    \item Bracers, bracelets, gauntlets, and gloves.
                \end{itemize}
            \item Armor:
                \begin{itemize}
                    \item Body armor, shields
                \end{itemize}
            \item Head items:
                \begin{itemize}
                    \item Hats, headbands, and helmets
                \end{itemize}
            \item Leg items:
                \begin{itemize}
                    \item Boots and shoes.
                \end{itemize}
            \item Rings
            \item Torso items:
                \begin{itemize}
                    \item Amulets, belts, cloaks, mantles, necklaces, robes, shirts, and vests.
                \end{itemize}
        \end{itemize}

        Of course, a character may carry or possess as many items of the same type as he wishes.
        However, additional equipped items beyond those listed above have no effect.

        A rare few apparel items can be ``worn'' without taking up space on a character's body.
        The description of an item indicates when it has this property.

        \onecolumn

    \subsection{Apparel Table}

        \begin{longtabuwrapper}
            \begin{longtabu}{l X l l l l}
                \lcaption{Apparel Properties}\\
                \tb{Armor and Shields} & \tb{Description} & \tb{Cost} & \tb{Item Level} & \tb{Location} & \tb{Page} \\
                \mitemref{Feather} \minus1 & \minus1 encumbrance penalty & 100 gp & 2nd & Body, Shield & \mitempref{Feather} \\
                \mitemref{Arrow Deflection} & Can instantly deflect a ranged attack & \mitempricetable{Active}{1} & Shield & \mitempref{Arrow Deflection} \\
                \mitemref{Resilient} & Can grant damage reduction & \mitempricetable{Active}{1} & Body & \mitempref{Resilient} \\
                \mitemref*{Feather} \minus2 & \minus2 encumbrance penalty & 500 gp & 4th & Body, Shield & \mitempref{Feather} \\
                \mitemref{Arrow Catching} & Redirects arrows to hit you & \mitempricetable{Active}{2} & Shield & \mitempref{Arrow Catching} \\
                \mitemref{Bashing} & Deals more damage when used to bash & \mitempricetable{Passive}{1} & Shield & \mitempref{Bashing} \\
                \mitemref{Magic Resistant} & Can grant magic resistance & \mitempricetable{Active}{2} & Body & \mitempref{Magic Resistant} \\
                \mitemref{Flaming Burst} & Can deal fire damage when hit or missed & \mitempricetable{Active}{3} & Body, Shield & \mitempref{Flaming Burst} \\
                \mitemref{Freezing Burst} & Can deal cold damage when hit or missed & \mitempricetable{Active}{3} & Body, Shield & \mitempref{Freezing Burst} \\
                \mitemref{Shocking Burst} & Can deal electricity damage when hit or missed & \mitempricetable{Active}{3} & Body, Shield & \mitempref{Shocking Burst} \\
                \mitemref*{Feather} \minus3 & \minus3 encumbrance penalty & 2,500 gp & 8th & Body, Shield & \mitempref{Feather} \\
                \mitemref{Fortified} & Grants immunity to critical hits & \mitempricetable{Passive}{3} & Body & \mitempref{Fortified} \\
                \mitemref*{Feather} \minus4 & \minus4 encumbrance penalty & 12,500 gp & 12th & Body, Shield & \mitempref{Feather} \\
                \mitemref{Energy Resistant} & Reduces energy damage & \mitempricetable{Passive}{4} & Body & \mitempref{Energy Resistant} \\
                \mitemref*{Feather} \minus5 & \minus5 encumbrance penalty & 62,500 gp & 16th & Body, Shield & \mitempref{Feather} \\
                \mitemref{Flameforged} & Deals fire damage when hit or missed & \mitempricetable{Passive}{5} & Body, Shield & \mitempref{Flameforged} \\
                \mitemref{Iceforged} & Deals cold damage when hit or missed & \mitempricetable{Passive}{5} & Body, Shield & \mitempref{Iceforged} \\
                \mitemref{Invulnerable} & Reduces physical damage & \mitempricetable{Passive}{5} & Body & \mitempref{Invulnerable} \\
                \mitemref{Sparkforged} & Deals electricity damage when hit or missed & \mitempricetable{Passive}{5} & Body, Shield & \mitempref{Sparkforged} \\
                \mitemref{Spell Reflecting} & Can reflect spells & \mitempricetable{Active}{7} & Shield & \mitempref{Spell Reflecting} \\

                \midrule
                \tb{Arms} & \tb{Description} & \tb{Cost} & \tb{Item Level} & \tb{Location} & \tb{Page} \\
                \mitemref{Greatreach Bracers} & Can grant increased reach & \mitempricetable{Active}{0} & Arms & \mitempref{Greatreach Bracers} \\
                \mitemref{Burning Gloves} & Can throw ball of fire & \mitempricetable{Active}{1} & Arms & \mitempref{Burning Gloves} \\
                \mitemref{Gauntlets of Improvisation} \plus1 & Grants \plus1 with improvised weapons & 200 gp & 3rd & Arms & \mitempref{Gauntlets of Improvisation} \\
                \mitemref{Bracers of Archery} & Grants proficiency with bows & \mitempricetable{Passive}{1} & Arms & \mitempref{Bracers of Archery} \\
                \mitemref{Bracers of Armor} & Grants invisible body armor & \mitempricetable{Passive}{1} & Arms & \mitempref{Bracers of Armor} \\
                \mitemref{Bracers of Repulsion} & Can shove foe when hit & \mitempricetable{Active}{2} & Arms & \mitempref{Bracers of Repulsion} \\
                \mitemref{Gauntlet of the Ram} & Can shove foe after unarmed attack & \mitempricetable{Active}{2} & Arms & \mitempref{Gauntlet of the Ram} \\
                \mitemref{Puppeteer's Glove} & Can concentrate on Figments more easily & \mitempricetable{Active}{2} & Arms & \mitempref{Puppeteer's Glove} \\
                \mitemref*{Gauntlets of Improvisation} \plus2 & Grants \plus2 effect & 1,000 gp & 6th & Arms & \mitempref{Gauntlets of Improvisation} \\
                \mitemref{Throwing Gloves} & Grants ability to throw anything accurately & \mitempricetable{Passive}{2} & Arms & \mitempref{Throwing Gloves} \\
                \mitemref*{Gauntlets of Improvisation} \plus3 & Grants \plus3 effect & 5,000 gp & 10th & Arms & \mitempref{Gauntlets of Improvisation} \\
                \mitemref*{Gauntlets of Improvisation} \plus4 & Grants \plus4 effect & 25,000 gp & 13th & Arms & \mitempref{Gauntlets of Improvisation} \\
                \mitemref*{Gauntlets of Improvisation} \plus5 & Grants \plus5 effect & 125,000 gp & 17th & Arms & \mitempref{Gauntlets of Improvisation} \\

                \midrule
                \tb{Head} & \tb{Description} & \tb{Cost} & \tb{Item Level} & \tb{Location} & \tb{Page} \\
                \mitemref{Hat of Disguise} & Can disguise your appearance & \mitempricetable{Active}{1} & Head & \mitempref{Hat of Disguise} \\
                \mitemref{Mask of Water Breathing} & Can grant ability to breathe water & \mitempricetable{Active}{2} & Head & \mitempref{Mask of Water Breathing} \\

                \midrule
                \tb{Legs} & \tb{Description} & \tb{Cost} & \tb{Item Level} & \tb{Location} & \tb{Page} \\
                \mitemref{Boots of Water Walking} & Can walk on water & \mitempricetable{Active}{0} & Legs & \mitempref{Boots of Water Walking} \\
                \mitemref{Boots of Earth's Embrace} & Can ignore forced movement & \mitempricetable{Active}{1} & Legs & \mitempref{Boots of Earth's Embrace} \\
                \mitemref{Boots of Speed} & Can double movement speed & \mitempricetable{Active}{1} & Legs & \mitempref{Boots of Speed} \\
                \mitemref{Boots of Freedom} & Can ignore movement-impairing effects & \mitempricetable{Active}{1} & Legs & \mitempref{Boots of Speed} \\
                \mitemref{Shrinking Horseshoes} & Can shrink you & \mitempricetable{Active}{1} & Legs & \mitempref{Shrinking Horseshoes} \\
                \mitemref{Boots of Elvenkind} & Grants \plus4 to Stealth & 500 gp & 4th & Legs & \mitempref{Boots of Elvenkind} \\
                \mitemref{Sandals of Sprinting} & Grants \plus4 to Sprint & 500 gp & 4th & Legs & \mitempref{Sandals of Sprinting} \\
                \mitemref{Boots of Swift Passage} & Can teleport you short distances & \mitempricetable{Active}{2} & Legs & \mitempref{Boots of Swift Passage} \\
                \mitemref{Boots of the Winterlands} & Grants benefits in cold environments & \mitempricetable{Passive}{1} & Legs & \mitempref{Boots of the Winterlands} \\
                \mitemref{Boots of Levitation} & Can levitate you & \mitempricetable{Active}{2} & Legs & \mitempref{Boots of Levitation} \\
                \mitemref{Boots of Striding and Springing} & Grants \plus10 foot speed & \mitempricetable{Passive}{3} & Legs & \mitempref{Boots of Striding and Springing} \\
                \mitemref{Boots of Teleportation} & Can teleport you long distances & \mitempricetable{Active}{4} & Legs & \mitempref{Boots of Teleportation} \\
                \mitemref{Seven League Boots} & Can teleport you exactly 25 miles & \mitempricetable{Active}{6} & Legs & \mitempref{Seven League Boots} \\

                \midrule
                \tb{Rings} & \tb{Description} & \tb{Cost} & \tb{Item Level} & \tb{Location} & \tb{Page} \\
                \mitemref{Protection} \plus1 & Grants \plus1 armor enhancement & 100 gp & 2nd & Ring & \mitempref{Protection} \\
                \mitemref{Energy Resistance} & Can briefly reduce energy damage & \mitempricetable{Active}{1} & Ring & \mitempref{Energy Resistance} \\
                \mitemref{Animal Friendship} & Grants \plus4 to Creature Handling & 500 gp & 4th & Ring & \mitempref{Animal Friendship} \\
                \mitemref*{Protection} \plus2 & Grants \plus2 armor enhancement & 500 gp & 4th & Ring & \mitempref{Protection} \\
                \mitemref{Nourishment} & Grants food and drink automatically & \mitempricetable{Passive}{1} & Ring & \mitempref{Nourishment} \\
                \mitemref{Invisibility} & Can grant invisibility & \mitempricetable{Active}{2} & Ring & \mitempref{Invisibility} \\
                \mitemref{Counterspells} & Can counter spell cast into ring & \mitempricetable{Active}{3} & Ring & \mitempref{Counterspells} \\
                \mitemref{Heroic Vengeance} & Can deal damage when foe rolls a 20 & \mitempricetable{Active}{3} & Ring & \mitempref{Heroic Vengeance} \\
                \mitemref*{Protection} \plus3 & Grants \plus3 armor enhancement & 2,500 gp & 8th & Ring & \mitempref{Protection} \\
                \mitemref{Sustenance} & Grants food, drink, and sleep automatically & \mitempricetable{Passive}{3} & Ring & \mitempref{Sustenance} \\
                \mitemref*{Protection} \plus4 & Grants \plus4 armor enhancement & 12,500 gp & 12th & Ring & \mitempref{Protection} \\
                \mitemref{Regeneration} & Grants healing each round & \mitempricetable{Passive}{6} & Ring & \mitempref{Regeneration} \\
                \mitemref*{Protection} \plus5 & Grants \plus5 armor enhancement & 62,500 gp & 16th & Ring & \mitempref{Protection} \\

                \midrule
                \tb{Torso} & \tb{Description} & \tb{Cost} & \tb{Item Level} & \tb{Location} & \tb{Page} \\
                \mitemref{Amulet of Mighty Fists} \plus1 & Grants \plus1 unarmed weapon enhancement & 200 gp & 3rd & Torso & \mitempref{Amulet of Mighty Fists} \\
                \mitemref{Belt of Heroic Recovery} & Can heal when you roll a 20 & \mitempricetable{Active}{1} & Torso & \mitempref{Belt of Heroic Recovery} \\
                \mitemref{Healing Belt} & Can heal touched creature & \mitempricetable{Active}{1} & Torso & \mitempref{Healing Belt} \\
                \mitemref{Obscuring Cloak} & Can summon fog around you & \mitempricetable{Active}{1} & Torso & \mitempref{Obscuring Cloak} \\
                \mitemref{Hunter's Cloak} & Can grant invisibility while immobile & \mitempricetable{Active}{2} & Torso & \mitempref{Hunter's Cloak} \\
                \mitemref*{Amulet of Mighty Fists} \plus2 & Grants \plus2 unarmed weapon enhancement & 1,000 gp & 6th & Torso & \mitempref{Amulet of Mighty Fists} \\
                \mitemref{Belt of Dwarvenkind} & Grants \plus2 Fortitude, dwarven characteristics & \mitempricetable{Passive}{2} & Torso & \mitempref{Belt of Dwarvenkind} \\
                \mitemref*{Amulet of Mighty Fists} \plus3 & Grants \plus3 unarmed weapon enhancement & 5,000 gp & 10th & Torso & \mitempref{Amulet of Mighty Fists} \\
                \mitemref{Amulet of Nondetection} & Grants resistance to divination spells & \mitempricetable{Passive}{3} & Torso & \mitempref{Amulet of Nondetection} \\
                \mitemref{Vanishing Cloak} & Can teleport and summon fog around you & \mitempricetable{Active}{4} & Torso & \mitempref{Vanishing Cloak} \\
                \mitemref{Assassin's Cloak} & Can grant invisibility while immobile, and shortly afterward  & \mitempricetable{Active}{5} & Torso & \mitempref{Assassin's Cloak} \\
                \mitemref*{Amulet of Mighty Fists} \plus4 & Grants \plus4 unarmed weapon enhancement & 25,000 gp & 13th & Torso & \mitempref{Amulet of Mighty Fists} \\
                \mitemref{Amulet of the Planes} & Can transport you to other planes & \mitempricetable{Active}{6} & Torso & \mitempref{Amulet of the Planes} \\
                \mitemref*{Amulet of Mighty Fists} \plus5 & Grants \plus5 unarmed weapon enhancement & 125,000 gp & 17th & Torso & \mitempref{Amulet of Mighty Fists} \\
            \end{longtabu}
        \end{longtabuwrapper}
        \twocolumn

    \subsection{Armor Overview}

        Magic body armor and shields protect the wearer to a greater extent than their nonmagical equivalents.
        All magic armor has an enhancement bonus to improve your hit points and ability to resist attacks.
        In addition to an enhancement bonus, magic armor may have special properties or be made of an unusual material.

        Body armor is always created so that even if the type of armor comes with boots or gauntlets, these pieces can be switched for other magic boots or gauntlets.

        \subsubsection{Armor Enhancement Bonuses}\label{Armor Enhancement Bonuses}

            Magic armor can have enhancement bonuses ranging from \plus1 to \plus5.
            Each \plus1 of enhancement bonus increases your maximum hit points by an amount equal to the item's power, and grants you an additional defensive legend point each day.
            If you stop using the armor, you lose the legend points and your maximum hit points returns to its normal value.

            These legend points can only be gained once per day, regardless of the number of items you use.
            If you use both magic body armor and a magic shield, or change between different pieces of armor, use only the highest bonus that applies.
            If you change from a weaker magical item to a stronger magical item, you gain bonuses equal to the difference between the two items.

        \subsubsection{Prices}\label{Armor Prices}
            The prices of enhancement bonuses to armor are listed in \trefnp{Magic Armor Prices}, and the prices of magical properties are listed on \trefnp{Apparel Properties}.
            If armor has a magical property, the price of the property is added to the price of the armor.
            The number of properties on the armor cannot exceed the enhancement bonus of the armor.
            Additionally, the price of all properties cannot exceed twice the price of the enhancement bonus on the armor.

            \begin{dtable}
                \lcaption{Magic Armor Prices}
                \begin{dtabularx}{\columnwidth} {>{\ccol}X c c}
                    \tb{Enhancement Bonus} & \tb{Base Price} & \tb{Item Level} \\
                    \bottomrule
                    \plus1 & 100 gp    & 2nd  \\
                    \plus2 & 500 gp    & 4th  \\
                    \plus3 & 2,500 gp  & 8th  \\
                    \plus4 & 12,500 gp & 12th \\
                    \plus5 & 62,500 gp & 16th \\
                \end{dtabularx}
            \end{dtable}

            \parhead{Base Power for Armor and Shields} The base power of a magic shield or magic armor with a magical property is given in the item description.
            For an item with only an enhancement bonus, the base power is equal to three times the enhancement bonus.
            If an item has both an enhancement bonus and a magical property, the higher of the two base powers must be met.

            \parhead{Shields} Shield enhancement bonuses do not act as accuracy or damage bonuses when the shield is used in a bash.
            However, a shield can be enhanced as a weapon.

            \parhead{Hardness and Hit Points} Each \plus1 of enhancement bonus adds 2 to an armor or shield's hardness and \plus10 to its hit points.

            \parhead{Activation} A character benefits from magic armor and shields in exactly the way a character benefits from nonmagical armor and shields -- by wearing them.
            Magical properties on body armor are usually activated if the character is hit or damaged, while magical properties on shields are usually activated if the character avoids an attack.

    \subsection{Armor Properties}\label{Armor Properties}

        \begin{magicitemdef}{Arrow Catching}[Active]{2}[Shield]{Evocation [Telekinesis]}{as shield}
            \magicitemability{Active}[Standard action] If you trace a circle in the air with this shield, it pulls in nearby ranged weapons.
            Whenever a creature within a \areasmall emanation of you would be attacked by a ranged weapon, the attack is redirected to target you instead.
            Resolve the attack as if it had initially targeted you, except that the attack is not affected by cover or concealment.
            Projectile and thrown objects that are Medium size or larger are not affected by this redirection.
        \end{magicitemdef}

        \begin{magicitemdef}{Arrow Deflection}[Active]{1}[Shield]{Transmutation [Imbuement]}{as shield}
            \magicitemability{Triggered}[Immediate action] When you are attacked by a ranged weapon, you can activate this item.
            If you do, the shield moves itself to deflect the attack without harm.
            You must be aware of the attack to deflect it in this way.
            Projectile and thrown objects that are Medium size or larger are not affected by this deflection.
        \end{magicitemdef}

        \begin{magicitemdef}{Bashing}[Passive]{1}[Shield]{Transmutation [Enhancement]}{as shield}
            \magicitemability{Passive} This shield deals damage with shield bash attacks as if it was two size categories larger than normal (so a Medium-sized light shield deals 1d6 damage, and a Medium-sized heavy shield deals 1d8 damage).
        \end{magicitemdef}

        \begin{magicitemdef}{Energy Resistant}[Passive]{4}[Body]{Abjuration [Shielding]}{as armor}
            \magicitemability{Passive} You have damage reduction against \glossterm{energy damage} equal to the item's power.
            Whenever you resist energy with this item, it sheds light as a torch for 5 rounds.
            The color of the light depends on the energy damage resisted: blue for cold, yellow for electricity, red for fire, brown for sonic, and white for none.
        \end{magicitemdef}

        \begin{magicitemdef}{Feather}{1}[Body, Shield]{Transmutation [Enhancement]}{as armor}
            \magicitemability{Passive} This armor has a reduced \glossterm{encumbrance penalty}.
            The price depends on the penalty reduction, as shown in the table below.

            Its base power is equal to three times the item's penalty reduction.
            To craft the item, you must have a number of ranks in the relevant Craft skill equal to the item's base power \add 4.
        \end{magicitemdef}

        \begin{dtable}
            \lcaption{Feather Armor}
            \begin{dtabularx}{\columnwidth} {l X X}
                \tb{Bonus} & \tb{Base Price} & \tb{Item Level} \\
                \bottomrule
                \minus1    & 100 gp          & 2nd             \\
                \minus2    & 500 gp          & 4th             \\
                \minus3    & 2,500 gp        & 8th             \\
                \minus4    & 12,500 gp       & 12th            \\
                \minus5    & 62,500 gp       & 16th            \\
            \end{dtabularx}
        \end{dtable}

        %\mitemreq{Transmutation}{Imbuement}{1}{varies}{Craft (as armor) varies}

        \begin{magicitemdef}{Flaming Burst}[Active]{3}[Body, Shield]{Evocation [Fire]}{as armor}
            \magicitemability{Triggered}[Immediate action] When you are hit or missed by a melee attack, you can activate this item.
            Body armor triggers if the attack hits, and shields trigger if the attack misses.
            If you activate the item, the attacking creature takes 1d6 fire damage per two item power.

            After you activate this ability, the item is wreathed in flame, causing it to shed light as a torch.
            This effect lasts for 5 rounds, during which time you cannot activate this ability again.
        \end{magicitemdef}

        \begin{magicitemdef}{Flameforged}[Active]{5}[Body, Shield]{Evocation [Fire]}{as armor}
            \magicitemability{Passive} When a creature hits or misses you with a melee attack, the attacking creature takes 1d6 fire damage per two item power.
            Body armor triggers if the attack hits, and shields trigger if the attack misses.
            A creature can only be dealt damage by this item once per round.
        \end{magicitemdef}

        \begin{magicitemdef}{Fortified}[Passive]{3}[Body]{Transmutation [Enhancement]}{as armor}
            \magicitemability{Passive} You are immune to critical hits.
        \end{magicitemdef}

        \begin{magicitemdef}{Freezing Burst}[Active]{3}[Body, Shield]{Evocation [Cold]}{as armor}
            \magicitemability{Triggered}[Immediate action] When you are hit or missed by a melee attack, you can activate this item.
            Body armor triggers if the attack hits, and shields trigger if the attack misses.
            If you activate the item, the attacking creature takes 1d6 cold damage per two item power.

            After you activate this ability, the item radiates frigid cold, causing it to snuff out torches and other small fires within a 5 foot radius.
            This effect lasts for 5 rounds, during which time you cannot activate this ability again.
        \end{magicitemdef}

        \begin{magicitemdef}{Glamered}[Active]{2}[Body]{Illusion [Glamer]}{as armor}
            \magicitemability{Active}[Standard action] If you trace the symbol of a mask on your chest, this armor appears to change shape and form to assume the appearance of a normal set of clothing.
            You may choose the design of the clothing.
            The armor retains all its properties (including weight and sound) when glamered.

            The armor remains disguised until you trace the symbol of the mask in the reverse direction, at which point it regains its normal appearance.
        \end{magicitemdef}

        \begin{magicitemdef}{Iceforged}[Active]{5}[Body, Shield]{Evocation [Cold]}{as armor}
            \magicitemability{Passive} When a creature hits or misses you with a melee attack, the attacking creature takes 1d6 cold damage per two item power.
            Body armor triggers if the attack hits, and shields trigger if the attack misses.
            A creature can only be dealt damage by this item once per round.
        \end{magicitemdef}

        \begin{magicitemdef}{Invulnerable}[Passive]{5}[Body]{Abjuration [Shielding]}{as armor}
            \magicitemability{Passive} You have physical damage reduction equal to the item's power.
            This allows you to ignore the first points of physical damage you take each round.
            Adamantine weapons ignore this damage reduction and negate it for 1 round.
        \end{magicitemdef}

        \begin{magicitemdef}{Magic Resistant}[Active]{2}[Body]{Abjuration [Shielding]}{as armor}
            \magicitemability{Active}[Standard action] If you trace an inverted arcane sigil on your chest, you gain \glossterm{magic resistance} equal to 10 \add the item's power.
            The magic resistance lasts for 5 rounds.
        \end{magicitemdef}

        \begin{magicitemdef}{Resilient}[Active]{1}[Body]{Abjuration [Shielding]}{as armor}
            \magicitemability{Active}[Standard action] If you strike your chest with a weapon or other hard object, you gain physical damage reduction equal to the item's power for 5 rounds.
            This allows you to ignore the first points of physical damage you take each round.
            Adamantine weapons ignore this damage reduction and negate it for 1 round.
        \end{magicitemdef}

        \begin{magicitemdef}{Shocking Burst}[Active]{3}[Body, Shield]{Evocation [Electricity]}{as armor}
            \magicitemability{Triggered}[Immediate action] When you are hit or missed by a melee attack, you can activate this item.
            Body armor triggers if the attack hits, and shields trigger if the attack misses.
            If you activate the item, the attacking creature takes 1d6 electricity damage per two item power.

            After you activate this ability, the item crackles with electrical energy, causing you to radiate flashing light as a torch.
            This effect lasts for 5 rounds, during which time you cannot activate this ability again.
        \end{magicitemdef}

        \begin{magicitemdef}{Sparkforged}[Active]{5}[Body, Shield]{Evocation [Electricity]}{as armor}
            \magicitemability{Passive} When a creature hits or misses you with a melee attack, the attacking creature takes 1d6 electricity damage per two item power.
            Body armor triggers if the attack hits, and shields trigger if the attack misses.
            A creature can only be dealt damage by this item once per round.
        \end{magicitemdef}

        \begin{magicitemdef}{Spell Reflecting}[Active]{7}[Shield]{Abjuration [Shielding]}{as armor}[This shield's surface is completely reflective, allowing it to act as a mirror.]
            \magicitemability{Triggered}[Immediate action] When you are targeted by a \glossterm{targeted spell}, you can activate this shield to reflect the spell back at its caster.
            The spell has its normal effects, except that the caster is the target instead of you.
            Any other targets of the spell are affected normally.

            After you activate this ability, the shield's surface becomes dully metallic instead of reflective.
            This effect lasts for 5 rounds, during which time you cannot activate this ability again.
        \end{magicitemdef}


    \subsection{Arms}

        \begin{magicitemdef}{Bracers of Archery}[Passive]{1}[Arms]{Transmutation [Enhancement]}{leather or metal}
            \magicitemability{Passive} You are proficient with bows.
        \end{magicitemdef}

        \begin{magicitemdef}{Bracers of Armor}[Passive]{1}[Arms]{Abjuration [Force]}{leather or metal}
            \magicitemability{Passive} You gain a \plus2 bonus to Armor defense.
            The protection from these bracers is treated as body armor, and does not stack with any other body armor you wear.
            Since this armor is made of magical force, incorporeal creatures can't bypass it the way they do normal armor.
        \end{magicitemdef}

        \begin{magicitemdef}{Bracers of Repulsion}[Active]{2}[Arms]{Evocation [Telekinesis]}{leather or metal}
            \magicitemability{Triggered}[Immediate action] When a foe damages you with a melee attack, you can activate this item.
            If you do, you can make a shove attack against the attacking creature at the end of the round.
            Your accuracy is equal to the item's power \add the damage its attack dealt to you.

            After you activate this item, barely visible fields of telekinetic energy surround the bracers.
            This effect lasts for 5 rounds, during which time you cannot activate this ability again.
        \end{magicitemdef}

        \begin{magicitemdef}{Burning Gloves}[Active]{1}[Arms]{Evocation [Fire]}{leather or textiles}
            \magicitemability{Active}[Standard action] By flicking your fingers, you can create fire in your hand.
            You can immediately throw this fire at a creature or object within \rngmed range.
            If you do, you make Item power vs. Reflex attack against the target.
            A successful attack means the target takes 1d6 fire damage per item power.
            A failed attack deals half damage.
        \end{magicitemdef}

        \begin{magicitemdef}{Gauntlets of Improvisation}{1}[Arms]{Transmutation [Enhancement]}{bone or metal}
            \magicitemability{Passive} These gauntlets have an enhancement bonus to improve your improvised weapons.
            This functions like an enhancement bonus with a weapon (see \pcref{Weapon Enhancement Bonuses}), except that the damage bonus applies to all improvised weapons you wield.

            \spellspecial The price of the gauntlets depend on their enhancement bonus, as shown in \trefnp{Gauntlets of Improvisation}.
            Its base power is equal to three times its enhancement bonus.
            To craft the item, you must have a number of ranks in Craft (bone or metal) equal to the item's base power \add 4.
        \end{magicitemdef}

        \begin{dtable}
            \lcaption{Gauntlets of Improvisation}
            \begin{dtabularx}{\columnwidth}{l X X}
                \tb{Bonus} & \tb{Base Price} & \tb{Item Level} \\
                \bottomrule
                \plus1 & 200 gp & 3rd \\
                \plus2 & 1000 gp & 6th \\
                \plus3 & 5,000 gp & 10th \\
                \plus4 & 25,000 gp & 13th \\
                \plus5 & 125,000 gp & 17th \\
            \end{dtabularx}
        \end{dtable}

        \begin{magicitemdef}{Gauntlet of the Ram}[Active]{2}[Arms]{Evocation [Telekinesis]}{bone or metal}
            \magicitemability{Triggered}[Immediate action] When you make a successful unarmed attack with this gauntlet against a foe, you can activate it to immediately make a shove attack against the struck foe.
            You gain a bonus on the shove attack equal to the damage you dealt.
            In addition, you do not need to move with the foe to push it backwards.

            After you activate this ability, the gauntlet grows small ram horns.
            This effect lasts for 5 rounds, during which time you cannot activate this ability again.
        \end{magicitemdef}

        \begin{magicitemdef}{Greatreach Bracers}[Active]{0}[Arms]{Transmutation [Imbuement]}{bone or metal}
            \magicitemability{Active}[Swift action] You can activate this item to increase your \glossterm{reach} by 5 feet until the end of the round.

            After you activate this ability, the bracers visually grow in size, though not in weight or encumbrance.
            This effect lasts for 5 rounds, during which time you cannot activate this ability again.
        \end{magicitemdef}

        \begin{magicitemdef}{Puppeteer's Glove}[Active]{2}[Arms]{Illusion [Figment]}{leather or textiles}
            \magicitemability{Triggered}[Immediate action] When you cast a \glossterm{Figment} spell, you can activate this glove.
            If you do, you can concentrate on the spell as a \glossterm{free action} by controlling the figment with your glove.
            If you are unable to control the figment with the glove, such as if you use the gloved hand for any other purpose, you lose your concentration on the figment.
            You must retain line of sight and line of effect to the figment to control it.

            After you activate this ability, the tips of the glove grow faintly visible strings.
            The strings extend four inches from the glove before vanishing into nothingness.
            This effect lasts for five rounds, during which time you cannot activate this ability again.
        \end{magicitemdef}

        \begin{magicitemdef}{Throwing Gloves}[Passive]{2}[Arms]{Transmutation [Enhancement]}{leather or textiles}
            \magicitemability{Passive} You can throw any item as if you were proficient with throwing it, granting you a \plus4 bonus to accuracy.
            This does not improve your ability to throw items you are already proficient with throwing, such as darts.
        \end{magicitemdef}

    \subsection{Head}

        \begin{magicitemdef}{Hat of Disguise}[Active]{2}[Head]{Illusion [Glamer]}{textiles}
            \magicitemability{Active}[Standard action] If you spin this hat on your head once, you can change your appearance for 1 hour, as the effect of the \spell{illusory guise} ritual.
            Your effective spellpower is equal to your item power.
        \end{magicitemdef}

        \begin{magicitemdef}{Mask of Water Breathing}[Active]{2}[Head]{Transmutation [Imbuement]}{textiles}
            \magicitemability{Active}[Standard action] If you speak a command word while wearing this mask, you gain the ability to breathe water for 1 hour.
            This does not prevent you from breathing air, and does not grant you the ability to breathe other liquids.
        \end{magicitemdef}

    \subsection{Legs}

        \begin{magicitemdef}{Boots of Earth's Embrace}[Active]{1}[legs]{Transmutation [Imbuement]}{leather or textiles}
            \magicitemability{Triggered}[Immediate action] When you are hit by an attack that would deal physical damage or force you to move (such as a shove attack), you can activate these boots.
            If you do, until the end of the round, you are immune to forced movement effects and you gain damage reduction against physical damage equal to your item power.

            After you activate this ability, the boots appear to be made from solid rock, though this does not impede your movement.
            This effect lasts for 5 rounds, during which time you cannot activate this ability again.
        \end{magicitemdef}

        \begin{magicitemdef}{Boots of Elvenkind}[500]{1}[Legs]{Transmutation [Enhancement]}{leather or textiles}
            \magicitemability{Passive} You gain a \plus4 bonus to Stealth checks.
        \end{magicitemdef}

        \begin{magicitemdef}{Boots of Freedom}[Active]{1}[Legs]{Transmutation [Imbuement]}{leather or textiles}
            \magicitemability{Active}[Standard action] If you wiggle your toes, you gain the effects of the \spell{freedom} spell.
            This benefit lasts as long as you move continuously without taking any other action, and for 5 rounds thereafter (maximum 5 minutes).
        \end{magicitemdef}

        \begin{magicitemdef}{Boots of Levitation}[Active]{2}[Legs]{Evocation [Telekinesis]}{leather or textiles}
            \magicitemability{Active}[Standard action] By lifting and planting one leg in mid-air, as if climbing an invisible stair, you can gain the benefit of the \spell{levitate} spell on yourself for 5 rounds.
        \end{magicitemdef}

        \begin{magicitemdef}{Boots of Speed}[Active]{1}[Legs]{Transmutation [Temporal]}{leather or textiles}
            \magicitemability{Active}[Standard action] If you stomp your foot on the ground three times, you gain the effects of the \spell{haste} spell.
            This effect lasts as long as you move continuously without taking any other action, and for 5 rounds thereafter (maximum 5 minutes).
        \end{magicitemdef}

        \begin{magicitemdef}{Boots of Striding and Springing}[Passive]{3}[Legs]{Transmutation [Enhancement]}{leather or textiles}
            \magicitemability{Passive} You gain a \plus10 foot bonus to your land speed.
            A high land speed increases your ability to jump (see \pcref{Jump}).
        \end{magicitemdef}

        \begin{magicitemdef}{Boots of Swift Passage}[Active]{2}[Legs]{Conjuration [Teleportation]}{leather or textiles}
            \magicitemability{Active}[Standard action] If you take three rapid steps in the shape of a triangle, you can teleport to an unoccupied location within \rngmed range of your original location.
            If that destination is invalid, this ability has no effect, though this does not prevent you from using the boots again.
            You must be able to move in order to activate this item.
            If you are \grappled, \immobilized, or otherwise unable to move normally, you cannot activate it.

            After you activate this ability successfully, the boots shimmer and seem faintly translucent.
            This effect lasts for 5 rounds, during which time you cannot activate this ability again.
        \end{magicitemdef}

        \begin{magicitemdef}{Boots of Teleportation}[Active]{4}[Legs]{Conjuration [Teleportation]}{leather or textiles}
            \magicitemability{Active}[Standard action] If you click your heels together three times, you can teleport up to 1,000 feet to a location you can specify, as the \spell{dimension door} spell.
        \end{magicitemdef}

        \begin{magicitemdef}{Boots of Water Walking}[Active]{0}[Legs]{Transmutation [Imbuement, Water]}{leather or textiles}
            \magicitemability{Active}[Standard action] If you step into water with your left foot, you can walk on water, as the \spell{water walk} spell.
            This effect lasts as long as you move continuously on water without taking any other action, and for 5 rounds thereafter (maximum 1 hour).
        \end{magicitemdef}

        \begin{magicitemdef}{Boots of the Winterlands}[Passive]{1}[Legs]{Evocation/Transmutation [Enhancement]}{leather or textiles}
            \magicitemability{Passive} You can travel across snow and ice without slipping or suffering movement penalties for the terrain.
            In addition, the boots keep you warm, protecting you in conditions as cold as \minus50 Fahrenheit.
        \end{magicitemdef}

        \begin{magicitemdef}{Sandals of Sprinting}[500]{1}[Legs]{Transmutation [Enhancement]}{leather or textiles}
            \magicitemability{Passive} You gain a \plus4 bonus to Sprint checks.
        \end{magicitemdef}

        \begin{magicitemdef}{Seven League Boots}[Active]{6}[Legs]{Conjuration [Teleportation]}{leather or textiles}
            \magicitemability{Active}[Move action] If you take an extraordinarily long step, you can teleport exactly 25 miles in a direction you specify.
            If this would place you within a solid object or otherwise impossible space, the boots will shunt you up to 1,000 feet in any direction to the closest available space, and you take 1d10 physical damage.
            If there is no available space within 1,000 feet of your intended destination, the effect fails and you take 2d10 physical damage.
        \end{magicitemdef}

        \begin{magicitemdef}{Shrinking Horseshoes}[Active]{1}[Legs]{Transmutation [Shaping, Sizing]}{metal}[These horseshoes are intended to be worn by horses and similar animals. Similar items may exist with the same magical property that can be worn by other creatures.]
            \magicitemability{Active}[Standard action] If tap a short pattern with one of your feet, you shrink to become one size category smaller, as the effect of the \spell{shrink} spell.
            This pattern is simple, and an animal can be taught the pattern as a trick with a DR 15 Creature Handling check (see \pcref{Training Creatures}).
            If you are riding a mount wearing these horseshoes, it can use one of your item uses in place of its own to activate this item.
        \end{magicitemdef}

        \begin{magicitemdef}{Winged Boots}[Active]{4}[Legs]{Transmutation [Imbuement]}{leather or textiles}
            \magicitemability{Active}[Standard action] If you tap your boots together in midair, you gain a 30 foot fly speed for 5 rounds.
            While \unencumbered, you can fly (see \pcref{Flying}).

            After you activate this ability, small wings sprout from the sides of the boots.
            This effect lasts for 5 minutes, during which time you cannot activate this ability again.
        \end{magicitemdef}

    \subsection{Rings}

        \parhead{Physical Description} Rings have no appreciable weight.
        Although exceptions exist that are crafted from glass or bone, the vast majority of rings are forged from metal -- usually precious metals such as gold, silver, and platinum.
        A typical ring has AD 13, 10 hit points, hardness 10, and a break DR of 25.

        \begin{magicitemdef}{Animal Friendship}[500]{1}[Ring]{Transmutation [Enhancement]}{bone, metal, or jewelry}
            \magicitemability{Passive} You gain a \plus4 bonus to Creature Handling checks.
        \end{magicitemdef}

        \begin{magicitemdef}{Counterspells}[Active]{3}[Ring]{Abjuration [Thaumaturgy]}{crafting}
            \magicitemability{Active}[Immediate action] If you cast a \glossterm{targeted spell} while wearing this ring, you can activate it.
            If you do, the spell becomes stored in the ring instead of having its normal effect.
            The next time a creature wearing the ring becomes targeted by the stored spell, that spell is automatically counterspelled.
            This discharges the energy stored in the ring, and it becomes unable to counter spells until another spell is stored in it.

            The ring may only store one spell at a time.
            If a new spell is stored into the ring while another spell is already inside it, the new spell is stored and the old spell is discharged without effect.
            Storing a spell in the ring requires an item use, but countering a spell does not.
        \end{magicitemdef}

        \begin{magicitemdef}{Energy Resistance}[Active]{1}[Ring]{Abjuration [Shielding]}{bone, metal, or jewelry}
            \magicitemability{Triggered}[Immediate action] When you would take \glossterm{energy damage}, you can activate this ring.
            If you do, you gain damage reduction against \glossterm{energy damage} equal to the item's power until the end of the round.

            After you activate this ability, the ring sheds light as a torch.
            The color of the light depends on the energy damage resisted: blue for cold, yellow for electricity, red for fire, brown for sonic, and white for none.
            This effect lasts for 5 rounds, during which time you cannot activate this ability again.
        \end{magicitemdef}

        \begin{magicitemdef}{Heroic Vengeance}[Active]{3}[Ring]{Abjuration}{bone, metal, or jewelry}
            \magicitemability{Triggered}[Immediate action] When a foe within \rngmed range rolls a natural 20 on an attack against you, you can activate this item.
            If you do, the attacking creature takes 1d10 divine damage per item power.
        \end{magicitemdef}

        \begin{magicitemdef}{Invisibility}[Active]{2}[Ring]{Illusion [Glamer]}{bone, metal, or jewelry}
            \magicitemability{Active}[Standard action] If you put the hand wearing this ring into a pocket or otherwise conceal it from sight, you gain the effects of the \spell{invisibility} spell for 5 rounds.
        \end{magicitemdef}

        \begin{magicitemdef}{Nourishment}[Passive]{1}[Ring]{Conjuration/Transmutation [Creation, Temporal]}{bone, metal, or jewelry}
            \magicitemability{Passive} You continuously gain nourishment, and no longer need to eat or drink.
            The ring must be worn for 24 hours before it begins to work.
        \end{magicitemdef}

        \begin{magicitemdef}{Protection}{1}[Ring]{Abjuration [Shielding]}{bone, metal, or jewelry}
            \magicitemability{Passive} This ring has an enhancement bonus to improve your defenses.
            Each \plus1 of enhancement bonus grants temporary hit points equal the item's power, and grants you an additional defensive legend point each day.
            If you stop using the ring, you lose the temporary hit points and the legend points.
            The benefits of this ring function in the same way as enhancement bonuses on magic armor, and they not stack with those bonuses.

            These bonuses can only be gained once per day, regardless of the number of items you use.
            If you wear two rings, or change between different rings, use only the highest bonus that applies.
            If you change from a weaker magical item to a stronger magical item, you gain bonuses equal to the difference between the two items.

            \spellspecial The price of the ring depends on its enhancement bonus, as shown in the table below.
            Its base power is equal to three times its enhancement bonus.
            To craft the item, you must have a number of ranks in Craft (bone, metal, or jewelry) equal to the item's base power \add 4.
        \end{magicitemdef}

        \begin{dtable}
            \lcaption{Ring of Protection}
            \begin{dtabularx}{\columnwidth} {l X X}
                \tb{Bonus} & \tb{Base Price} & \tb{Item Level} \\
                \bottomrule
                \plus1 & 100 gp    & 2nd  \\
                \plus2 & 500 gp    & 4th  \\
                \plus3 & 2,500 gp  & 8th  \\
                \plus4 & 12,500 gp & 12th \\
                \plus5 & 62,500 gp & 16th \\
            \end{dtabularx}
        \end{dtable}

        \begin{magicitemdef}{Regeneration}[Passive]{6}[Ring]{Transmutation [Imbuement]}{bone, metal, or jewelry}
            \magicitemability{Passive} At the end of each round, you heal hit points equal your item power.
            Only damage taken while wearing the ring can be healed in this way.
        \end{magicitemdef}

        \begin{magicitemdef}{Sustenance}[Passive]{3}[Ring]{Conjuration/Transmutation [Creation, Temporal]}{bone, metal, or jewelry}
            \magicitemability{Passive} You continuously gain nourishment, and no longer need to eat or drink.
            In addition, you need only one-quarter your normal amount of sleep (or similar activity, such as elven trance) each day.

            The ring must be worn for 24 hours before it begins to work.
        \end{magicitemdef}

    \subsection{Torso}

        \begin{magicitemdef}{Amulet of Mighty Fists}{1}[Torso]{Transmutation [Enhancement]}{bone or jewelry}
            \magicitemability{Passive} This amulet has an enhancement bonus to improve your natural attacks and unarmed attacks.
            This functions like an enhancement bonus with a weapon (see \pcref{Weapon Enhancement Bonuses}), except that the damage bonus applies to your natural attacks and unarmed attacks.

            \spellspecial The price of the amulet depends on its enhancement bonus, as shown in \trefnp{Amulet of Mighty Fists}.
            Its base power is equal to three times its enhancement bonus.
            To craft the item, you must have a number of ranks in Craft (jewelry) equal to the item's base power \add 4.
        \end{magicitemdef}

        \begin{dtable}
            \lcaption{Amulet of Mighty Fists}
            \begin{dtabularx}{\columnwidth}{l X X}
                \tb{Bonus} & \tb{Base Price} & \tb{Item Level} \\
                \bottomrule
                \plus1 & 200 gp & 3rd \\
                \plus2 & 1000 gp & 6th \\
                \plus3 & 5,000 gp & 10th \\
                \plus4 & 25,000 gp & 13th \\
                \plus5 & 125,000 gp & 17th \\
            \end{dtabularx}
        \end{dtable}

        \begin{magicitemdef}{Amulet of the Planes}[Active]{6}[Torso]{Conjuration [Teleportation, Planar]}{bone, jewelry}
            By holding this amulet in one hand and concentrating on a specific plane for 1 minute, you can create the effects of a \ritual{plane shift} ritual.
            Activating the item successfully requires a DR 15 Knowledge (planes) check.
            If you fail, your activation of the amulet has no effect (but still consumes an item use).
            If you roll a 1 and fail, you and any creatures with you are transported to a random plane.
            Each time you successfully activate this item in the same day, the DR of the check increases by 5.
        \end{magicitemdef}

        \begin{magicitemdef}{Amulet of Nondetection}[Passive]{3}[Torso]{Abjuration [Shielding]}{bone, jewelry}
            \magicitemability{Passive} You gain the benefits of the \spell{nondetection} ritual.
            If a divination is attempted against you, the caster must make a spellpower check against a DR equal to 15 \add the item's power.
        \end{magicitemdef}

        \begin{magicitemdef}{Assassin's Cloak}[Active]{5}[Torso]{Illusion [Glamer]}{textiles}
            \magicitemability{Active}[Standard action] If you wrap this cloak around yourself so it covers your whole body, you can activate this item.
            If you do, you become \glossterm{invisible}.
            This effect lasts until you move, and for 5 rounds thereafter.

            If you attack, you become visible immediately, but the effect does not end.
            At the end of each round, if you did not attack a creature that round, you become invisible again.
        \end{magicitemdef}

        \begin{magicitemdef}{Belt of Dwarvenkind}[Passive]{2}[Torso]{Divination/Transmutation [Enhancement, Imbuement]}{leather or textiles}
            \magicitemability{Passive} You gain a \plus2 bonus to Fortitude defense.
            In addition, you gain dwarven characteristics.
            You gain a \plus4 bonus to social checks when dealing with dwarves, but take a \minus2 penalty with all other creatures.
            You also gain the benefits of the Stonecunning feat, regardless of whether you meet the prerequisites.
        \end{magicitemdef}

        \begin{magicitemdef}{Healing Belt}[Active]{1}[Torso]{Vivimancy [Life]}{leather or textiles}
            \magicitemability{Active}[Standard action] If you grab this belt in one hand and touch a willing creature with the other, the touched creature is healed for 1d6 damage per item power.
            If you heal yourself, you only need one hand free to grab the belt.
        \end{magicitemdef}

        \begin{magicitemdef}{Belt of Heroic Recovery}[Active]{1}[Torso]{Abjuration/Vivimancy [Positive]}{leather or textiles}
            \magicitemability{Triggered}[Immediate action] When you roll a natural 20 on an attack roll, you can activate this item. If you do, you heal 1d6 damage per item power.
        \end{magicitemdef}

        \begin{magicitemdef}{Hunter's Cloak}[Active]{2}[Torso]{Illusion [Glamer]}{textiles}
            \magicitemability{Active}[Standard action] If you wrap this cloak around yourself so it covers your whole body, you can activate this item.
            If you do, you become \glossterm{invisible}.
            This effect lasts until you move or attack, at which point it immediately ends.
        \end{magicitemdef}

        % +1 level for extra thick fog in the first round
        \begin{magicitemdef}{Obscuring Cloak}[Active]{1}[Torso]{Conjuration [Creation, Fog]}{textiles}
            \magicitemability{Active}[Standard action] If you wrap this cloak around yourself with one hand, you can create a zone of fog in a \areamed radius cylinder centered on you.
            All creatures within or looking through the area treat everything they see as if it had \concealment.
            The effect lasts for 5 rounds.
            For the first round of the effect, the fog is unusually thick, blocking all sight beyond 10 feet.
            This can allow you to hide unobserved.
        \end{magicitemdef}

        \begin{magicitemdef}{Vanishing Cloak}[Active]{4}[Torso]{Conjuration [Creation, Fog, Teleportation]}{textiles}
            \magicitemability{Active}[Standard action] If you wrap this cloak around yourself with one hand, you can teleport up to an unoccupied location within \rngmed range of your original location.
            In addition, you create a cloud of fog centered on you, as the \spell{fog cloud} spell.
            You can choose whether the cloud appears at your original location, or at your location after teleporting.
            For the first round of the effect, the fog is unusually thick, blocking all sight beyond 10 feet.
            This can allow you to hide unobserved (see \pcref{Stealth}).

            If your intended destination is invalid, you do not teleport, though the fog cloud is still created.
        \end{magicitemdef}

\section{Weapons}
    Magic weapons improve a character's combat abilities.
    They must be wielded to gain their effects.
    All magic weapons have an enhancement bonus to improve your damage and ability to hit.
    In addition to an enhancement bonus, magic weapons may have special properties or be made of an unusual material.

    \subsection{Weapon Table}

        \onecolumn
        \begin{longtabuwrapper}
            \begin{longtabu}{l X l l l l}
                \lcaption{Weapon Properties}\\
                \tb{Name} & \tb{Description} & \tb{Cost} & \tb{Item Level} & \tb{Weapon Type} & \tb{Page} \\
                \mitemref{Bane} & Property only works against some creatures & Varies & Varies & Varies & \mitempref{Bane} \\
                \mitemref{Sweeping} & Can make extra attack against another foe & \mitempricetable{Active}{0} & Melee & \mitempref{Sweeping} \\
                \mitemref{Rebounding} & Returns to you slowly when thrown & \mitempricetable{Passive}{0} & Melee & \mitempref{Rebounding} \\
                \mitemref{Forceful} & Can knock struck creature flying & \mitempricetable{Active}{1} & Any & \mitempref{Forceful} \\
                \mitemref{Morphing} & Can transform into similar weapon & \mitempricetable{Active}{1} & Any & \mitempref{Morphing} \\
                \mitemref{Surestrike} & Can reroll missed attack & \mitempricetable{Active}{1} & Any & \mitempref{Surestrike} \\
                \mitemref{Merciful} & Deals nonlethal damage & \mitempricetable{Passive}{1} & Any & \mitempref{Merciful} \\
                \mitemref{Transforming} & Can transform into any weapon & \mitempricetable{Active}{2} & Any & \mitempref{Transforming} \\
                \mitemref{Vampiric} & Can heal based on damage dealt & \mitempricetable{Active}{2} & Melee & \mitempref{Vampiric} \\
                \mitemref{Defending} & Grants defensive legend points instead of offensive & \mitempricetable{Passive}{2} & Melee & \mitempref{Defending} \\
                \mitemref{Flaming} & Can deal fire damage on hit & \mitempricetable{Active}{3} & Any & \mitempref{Flaming} \\
                \mitemref{Freezing} & Can deal cold damage on hit & \mitempricetable{Active}{3} & Any & \mitempref{Flaming} \\
                \mitemref{Longshot} & Has doubled range increment & \mitempricetable{Passive}{2} & Ranged & \mitempref{Longshot} \\
                \mitemref{Poisoning} & Can instantly duplicate minor poison & \mitempricetable{Active}{3} & Melee & \mitempref{Poisoning} \\
                \mitemref{Returning} & Returns to you instantly when thrown & \mitempricetable{Passive}{2} & Melee & \mitempref{Returning} \\
                \mitemref{Shocking} & Can deal electricity damage on hit & \mitempricetable{Active}{3} & Any & \mitempref{Shocking} \\
                \mitemref{Thieving} & Can absorb objects & \mitempricetable{Active}{3} & Melee & \mitempref{Thieving} \\
                \mitemref{Thundering} & Can deafen on hit & \mitempricetable{Active}{3} & Any & \mitempref{Thundering} \\
                \mitemref{Fixating} & Can automatically critical after striking foe repeatedly & \mitempricetable{Active}{4} & Melee & \mitempref{Fixating} \\
                \mitemref{Seeking} & Reduce or ignore miss chances & \mitempricetable{Passive}{4} & Any & \mitempref{Seeking} \\
                \mitemref{Disorienting} & Can disorient foe & \mitempricetable{Active}{6} & Any & \mitempref{Disorienting} \\
                \mitemref{Speed} & Grants extra strikes & \mitempricetable{Passive}{5} & Any & \mitempref{Speed} \\
                \mitemref{Toxic} & Can instantly coat itself with major poison & \mitempricetable{Active}{6} & Melee & \mitempref{Poisoning} \\
                \mitemref{Soulreaving} & Deals delayed damage instead of immediate damage & \mitempricetable{Passive}{7} & Any & \mitempref{Soulreaving} \\
                \mitemref{Vorpal} & Deals massive damage on a natural 20 & \mitempricetable{Passive}{8} & Melee & \mitempref{Vorpal} \\
            \end{longtabu}
        \end{longtabuwrapper}
        \twocolumn

    \subsection{Weapon Enhancement Bonuses}\label{Weapon Enhancement Bonuses}

        Magic weapons can have enhancement bonuses ranging from \plus1 to \plus5.
        You gain a bonus to damage on physical attacks using a magic weapon equal to the weapon's enhancement bonus.
        In addition, each \plus1 of enhancement bonus grants you an additional offensive legend point each day.

        These legend points can only be gained once per day, regardless of the number of weapons you use.
        If you use multiple weapons at once, or change between different weapons, use only the highest number of legend points that applies.
        If you change from a weaker magical weapon to a stronger magical weapon, you gain legend points equal to the difference between the two enhancement bonuses.

        Offensive legend points gained from weapon enhancement bonuses do not stack with offensive legend points gained from implement enhancement bonuses.
        Use the higher of the two.

    \subsection{Weapon Prices}\label{Weapon Prices}
        The prices of enhancement bonuses to weapons are listed in \trefnp{Magic Weapon Prices}, and the prices of magical properties are listed on \trefnp{Weapon Properties}.
        If a weapon has a magical property, the price of the property is added to the price of the weapon.
        The number of properties on the weapon cannot exceed the enhancement bonus of the weapon.
        Additionally, the price of all properties cannot exceed twice the price of the enhancement bonus on the weapon.

        \begin{dtable}
            \lcaption{Magic Weapon Prices}
            \begin{dtabularx}{\columnwidth} {>{\ccol}X c c}
                \tb{Enhancement Bonus} & \tb{Base Price} & \tb{Item Level} \\
                \bottomrule
                \plus1 & 200 gp     & 3rd  \\
                \plus2 & 1,000 gp   & 6th  \\
                \plus3 & 5,000 gp   & 10th \\
                \plus4 & 25,000 gp  & 13th \\
                \plus5 & 125,000 gp & 17th \\
            \end{dtabularx}
        \end{dtable}

        \parhead{Base Power for Weapons} The base power of a magic weapon with a magical property is given in the item description.
        For an item with only an enhancement bonus, the base power is three times the enhancement bonus.
        If an item has both an enhancement bonus and a magical property, the higher of the two base powers is used.

        \parhead{Hardness and Hit Points} Each \plus1 of enhancement bonus adds 2 to a weapon's hardness and \plus10 to its hit points.

        \parhead{Ranged Weapons and Ammunition} The enhancement bonus from a ranged weapon does not stack with the enhancement bonus from ammunition.
        Only the higher of the two enhancement bonuses applies.
        Magical properties are applied from both sources, as long as they are not identical.
        If conflicting properties exist, the property on the ammunition takes precedence.

        Magic ammunition loses its magic after being fired, whether it hits or misses.

        \parhead{Light Generation} Some magic weapons shed light equivalent to a light spell (bright light in a 20-foot radius, shadowy light in a 40-foot radius).
        The light on such weapons cannot normally be shut off.
        Some of the specific weapons detailed below always or never glow, as defined in their descriptions.

        \parhead{Activation} Usually, a character benefits from a magic weapon in the same way a character benefits from a mundane weapon -- by attacking with it.
        Magical properties on weapons are usually activated if the character strikes a foe with the weapon.

        \parhead{Magic Weapons and Critical Hits} Some weapon qualities and some specific weapons have an extra effect on a critical hit.
        These special effects function against creatures not subject to critical hits.
        When fighting against such creatures, roll for critical hits as you would against any other creature subject to critical hits.
        On a successful critical roll, apply the special effect, but do not multiply the weapon's regular damage.

    \subsection{Weapon Properties}\label{Weapon Properties}

        \begin{magicitemdef}{Bane}{2}{Transmutation [Imbuement]}{as weapon}
            \spellspecial A bane weapon excels at attacking a specific type of creature.
            Any magical weapon property that targets a creature with its effect can be designated as a ``bane'' ability, causing it to only function against a specific kind of creature.
            In exchange, the ability costs half the normal price in raw materials to add to the weapon.
            A list of possible foes is described on the following table.
        \end{magicitemdef}

        \begin{dtable}
            \lcaption{Bane Creature Types}
            \begin{dtabularx}{\columnwidth}{>{\lcol}X >{\lcol}X}
                \tb{Designated Foe}  & \tb{Designated Foe} \\
                \bottomrule
                Aberrations          & Animals             \\
                Animates             & Constructs          \\
                Humanoids, civilized & Humanoids, savage   \\
                Magical beast        & Monstrous humanoid  \\
                Outsiders, aligned   & Undead              \\
            \end{dtabularx}
        \end{dtable}

        \begin{magicitemdef}{Defending}[Passive]{2}[Melee]{Abjuration [Shielding]}{as weapon}
            \magicitemability{Passive} You can use the legend points granted by this weapon's enhancement bonus as defensive legend points, in addition to using them as offensive legend points.
            This stacks with any defensive legend points granted by armor, but not with other offensive legend points granted by weapons.
        \end{magicitemdef}

        \begin{magicitemdef}{Disorienting}[Active]{6}[Melee, Ranged]{Enchantment [Compulsion, Mind]}{as weapon}
            \magicitemability{Triggered}[Immediate action] When you hit a foe with this weapon, you can activate it to make the struck creature \disoriented for 2 rounds.

            After you activate this ability, the weapon cackles gleefully.
            It cackles again every time you hit a foe with it.
            This effect lasts for 5 rounds, during which time you cannot activate this ability again.
        \end{magicitemdef}

        \begin{magicitemdef}{Longshot}[Passive]{2}[Ranged]{Transmutation [Enhancement]}{as weapon}
            \magicitemability{Passive} Ranged attacks with this weapon have twice the normal \glossterm{range increment}.
            This can affect both thrown weapons and projectile weapons.
        \end{magicitemdef}

        \begin{magicitemdef}{Flaming}[Active]{3}[Melee, Ranged]{Evocation [Fire]}{as weapon}
            \magicitemability{Active}[Immediate action] When you hit with a physical attack with this weapon, you can activate it.
            If you do, the struck creature takes 1d6 fire damage per two item power.

            After you activate this ability, the item is wreathed in flame, causing it to shed light as a torch.
            This effect lasts for 5 rounds, during which time you cannot activate this ability again.
        \end{magicitemdef}

        \begin{magicitemdef}{Freezing}[Active]{3}[Melee, Ranged]{Evocation [Cold]}{as weapon}
            \magicitemability{Active}[Immediate action] When you hit with a physical attack with this weapon, you can activate it.
            If you do, the struck creature takes 1d6 cold damage per two item power.

            After you activate this ability, the item radiates frigid cold, causing it to snuff out torches and other small fires within a 5 foot radius.
            This effect lasts for 5 rounds, during which time you cannot activate this ability again.
        \end{magicitemdef}

        \begin{magicitemdef}{Forceful}[Active]{1}[Melee, Ranged]{Evocation [Telekinesis]}{as weapon}
            \magicitemability{Triggered}[Immediate action] When you damage a foe with this weapon, you can activate it.
            If you do, you make a shove attack against your target.
            You gain a bonus equal to the damage you dealt on the attack, and do not have to move with your foe to knock it back the full distance.

            After you activate this ability, the weapon is visibly sheathed in telekinetic force.
            This effect lasts for 5 rounds, during which time you cannot activate this ability again.
        \end{magicitemdef}

        \begin{magicitemdef}{Fixating}[Active]{4}[Melee]{Divination}{as weapon}
            \magicitemability{Active}[Swift action] When you activate this weapon, your attacks fixate on targets you have hit previously.
            Until the end of the round, all of your successful \glossterm{strikes} against creature that you hit with strikes in both of the last two rounds are automatically critical hits.

            After you activate this ability, the weapon twists slightly in your hand to point towards the target, if it is nearby and visible.
            This effect lasts for 5 rounds, during which time you cannot activate this ability again.
        \end{magicitemdef}

        \begin{magicitemdef}{Merciful}[Passive]{1}[Melee, Ranged]{Transmutation [Imbuement]}{as weapon}
            \magicitemability{Passive} This weapon deals \glossterm{nonlethal} damage.
            As a standard action, you may run your hand along the \glossterm{striking surface} of the weapon.
            This changes the weapon from dealing nonlethal damage to lethal damage, or vice versa.
        \end{magicitemdef}

        \begin{magicitemdef}{Morphing}[Active]{1}[Melee, Ranged]{Transmutation [Shaping]}{as weapon}
            \magicitemability{Active}[Standard action] If you grab this weapon with both hands and strike it against your knee or another hard object while visualizing a different weapon type, this weapon transforms into the visualized weapon type.
            The new weapon type must be from the same weapon group as the weapon's original type.
            This weapon remains transformed until you transform it again.
        \end{magicitemdef}

        \begin{magicitemdef}{Poisoning}[Active]{3}[Melee]{Conjuration/Transmutation [Creation, Temporal]}{as weapon}
            \magicitemability{Passive} This weapon has a small slot in its hilt (or similar gripping surface). Up to one ounce of liquid can be placed in the slot or removed as a standard action. A nonmagical liquid placed in this slot is kept fresh, decaying at a rate of one minute per day. In addition, a non-magical liquid worth less than 100gp can be quickly duplicated to coat the weapon's surface.
            \magicitemability{Active}[Swift action] If you press a small button on the hilt (or similar gripping surface) of this weapon, the weapon's striking surface becomes coated in the liquid stored in the weapon's slot.

            After you activate this ability, it takes 5 rounds for the weapon to create more liquid.
            During this time, you cannot activate this ability again.
        \end{magicitemdef}

        %price as 6th level spell? similar to 20 uses of a one-shot item worth 1000 gp
        \begin{magicitemdef}{Toxic}[Active]{6}[Melee]{Conjuration/Transmutation [Creation, Temporal]}{as weapon}
            \magicitemability{Passive} This weapon functions like a \magicitem{poisoning} weapon, except that it can duplicate nonmagical liquids worth up to 1,000 gp.
            In addition, up to five different poisons can be stored within the weapon.
            When you coat the weapon with poison, you may choose which poison to use.
        \end{magicitemdef}

        \begin{magicitemdef}{Rebounding}[Passive]{0}[Melee]{Conjuration [Teleportation]}{as weapon}
            \magicitemability{Passive} After being thrown or fired, a returning weapon teleports back to the creature that threw or fired it.
            It returns to the thrower at the end of the round (and is therefore ready to use again in the next round).

            Catching a rebounding weapon when it comes back is a free action.
            If you can't catch it, the weapon drops to the ground in the square from which it was thrown.
        \end{magicitemdef}

        \begin{magicitemdef}{Returning}[Passive]{2}[Melee]{Conjuration [Teleportation]}{as weapon}
            \magicitemability{Passive} This weapon functions like a \magicitem{returning} weapon, except that the weapon teleports back to the creature that threw or fired it immediately after the attack is resolved, allowing the creature to make multiple attacks in the same round with the same thrown weapon.
        \end{magicitemdef}

        \begin{magicitemdef}{Seeking}[Passive]{4}[Ranged]{Divination [Knowledge]}{as weapon}
            \magicitemability{Passive} This weapon automatically veers towards its intended target if misaimed.
            Attacks with this weapon that would suffer a 50\% miss chance instead suffer a 20\% miss chance.
            In additions, such attacks ignore all other effects that would apply a 20\% miss chance.
        \end{magicitemdef}

        \begin{magicitemdef}{Shocking}[Active]{3}[Melee, Ranged]{Evocation [Electricity]}{as weapon}
            \magicitemability{Active}[Immediate action] When you hit with a physical attack with this weapon, you can activate it.
            If you do, the struck creature takes 1d6 electricity damage per two item power.

            After you activate this ability, the item crackles with electrical energy, causing you to radiate flashing light as a torch.
            This effect lasts for 5 rounds, during which time you cannot activate this ability again.
        \end{magicitemdef}

        % As 7th level spell? I have no idea.
        % There aren't actually any spells this can be based on.
        \begin{magicitemdef}{Soulreaving}[Passive]{7}[Melee, Ranged]{Necromancy [Soul]}{as weapon}
            \magicitemability{Passive} Attacks with this weapon ignore all damage reduction and hardness, but do not immediately deal hit point damage.
            In fact, a creature struck by the weapon does not feel the attack at all.
            Damage that would be dealt by the weapon can be delayed indefinitely.
            While the damage is delayed, it cannot be removed by any means short of the creature's death and resurrection.

            A soulreaving weapon has no effect on objects or constructs.
            While wielded, it has physical form only for its wielder, making it impossible to disarm.
            While not in use, it can be picked up and touched normally.

            \magicitemability{Triggered}[Immediate] If you bury the weapon in your heart as a \glossterm{strike}, you can activate it.
            If you do, the delayed damage is converted into real damage.
            This deals no damage to you, but any other creatures that have been dealt damage by the weapon immediately the delayed damage the weapon has stored up for them.
            Any such damage dealt in excess of the creature's hit points is converted directly into vital damage.
            This removes all delayed damage the weapon has dealt.
        \end{magicitemdef}

        \begin{magicitemdef}{Speed}[Passive]{5}[Melee, Ranged]{Transmutation [Temporal]}{as weapon}
            \magicitemability{Passive} Whenever you make a \glossterm{standard attack} with this weapon, you can make an additional \glossterm{strike} with a \minus5 penalty to accuracy.
            This does not stack with any other effects which grant extra strikes.
        \end{magicitemdef}

        \begin{magicitemdef}{Surestrike}[Active]{1}[Melee, Ranged]{Divination [Knowledge]}{as weapon}
            \magicitemability{Triggered}[Immediate action] When you miss a \glossterm{strike} with this weapon, you can activate it to reroll the attack roll.
            You must take the second result.

            After you activate this ability, you see shadowy glimpses of alternate futures superimposed over objects and creatures you see.
            This effect lasts for 5 rounds, during which time you cannot activate this ability again.
        \end{magicitemdef}

        \begin{magicitemdef}{Sweeping}[Active]{0}[Melee]{Transmutation [Enhancement]}{as weapon}
            \magicitemability{Triggered}[Immediate action] When you hit a foe with a melee attack with this weapon, you can activate it.
            If you do, you can make an extra \glossterm{strike} at another creature you threaten.

            After you activate this ability, the weapon always feels slightly loose in your hands, no matter how tightly you grip it.
            This effect lasts for 5 rounds, during which time you cannot activate this ability again.
        \end{magicitemdef}

        \begin{magicitemdef}{Thieving}[Active]{3}[Melee]{Transmutation [Shaping, Sizing]}{as weapon}
            \magicitemability{Triggered}[Immediate action] When you strike an unattended object with this weapon, if the object is at least one size category smaller than the weapon, you may activate the weapon.
            If you do, the object is absorbed into the weapon, leaving no trace that it ever existed.
            If you use this weapon to make a successful disarm attempt, the disarmed object is considered unattended, allowing you to activate the weapon to absorb the object.

            \magicitemability{Passive} Items absorbed by the weapon remain stored in the weapon indefinitely.
            As a standard action, you can run your hand along the length of the striking surface of the weapon.
            If you do, the last item absorbed by the weapon appears in your hand.

            The weapon can hold no more than three objects at once.
            If you attempt to absorb an object while the weapon is full, the attempt fails.
        \end{magicitemdef}

        \begin{magicitemdef}{Thundering}[Active]{3}[Melee, Ranged]{Evocation [Sonic]}{as weapon}
            \magicitemability{Active}[Immediate action] When you hit with a physical attack with this weapon, you activate it.
            If you do, it emits a deafening thundering noise, causing the struck creature to be \deafened for 2 rounds.

            After you activate this ability, the item continues to rumble threateningly.
            This effect lasts for 5 rounds, during which time you cannot activate this ability again.
        \end{magicitemdef}

        \begin{magicitemdef}{Transforming}[Active]{2}[Melee, Ranged]{Transmutation}{as weapon}
            \magicitemability{Active}[Standard action] This weapon functions like a \magicitem{morphing} weapon, except that the weapon's new type may be from any weapon group.
        \end{magicitemdef}

        \begin{magicitemdef}{Vampiric}[Active]{2}[Melee]{Necromancy [Life]}{as weapon}
            \magicitemability{Active}[Immediate action] After dealing damage with a \glossterm{strike} with this weapon, you can activate it.
            If you do, the weapon heals you for an amount of damage equal to the physical damage dealt with that attack.

            After you activate this ability, the weapon glows with a dark light.
            This effect lasts for 5 rounds, during which time you cannot activate this ability again.
        \end{magicitemdef}

        \begin{magicitemdef}{Vorpal}[Passive]{8}[Melee]{Transmutation [Enhancement]}{as weapon}
            \magicitemability{Passive} When you roll a natural 20 with this weapon, you deliver a vorpal blow.
            You immediately make any number of additional strikes against the struck creature until you miss a strike or hit with 10 successful strikes.
            All damage dealt in this way is considered to come from a single attack.
            This effect replaces the normal benefits of delivering a critical hit.
        \end{magicitemdef}

\section{Implements}
    Implements can take many forms: staffs, wands, holy symbols, and more.
    Like weapons, implements must be wielded to gain their effects.
    However, while weapons are used to deal physical damage to enemies, implements are used to cast spells.
    All magical implements have an enhancement bonus to improve your spell damage.
    In addition to an enhancement bonus, magical implements may have a single magical property.

    \onecolumn
    \subsection{Implement Table}

        \begin{longtabuwrapper}
            \begin{longtabu}{l X l l l l}
                \lcaption{Implement Properties}\\
                \tb{Name} & \tb{Description} & \tb{Cost} & \tb{Item Level} & \tb{Magic Source} & \tb{Page} \\
                \mitemref{Electricity, Lesser} & Know \spell{shocking grasp} spell & \mitempricetable{Active}{1} & Arcane, Nature & \mitempref{Electricity, Lesser} \\
                \mitemref{Fire, Lesser} & Know \spell{burning hands} spell & \mitempricetable{Active}{1} & Arcane, Nature & \mitempref{Fire, Lesser} \\
                \mitemref{Missiles, Lesser} & Know \spell{magic missile} spell & \mitempricetable{Active}{1} & Arcane & \mitempref{Missiles, Lesser} \\
                \mitemref{Time, Lesser} & Know \spell{haste} spell & \mitempricetable{Active}{1} & Arcane & \mitempref{Time, Lesser} \\
                \mitemref{Cold, Lesser} & Know \spell{cone of cold} spell & \mitempricetable{Active}{2} & Arcane, Nature & \mitempref{Cold, Lesser} \\
                \mitemref{Curses, Lesser} & Know \spell{curse of blood and bone} spell & \mitempricetable{Active}{2} & Arcane, Divine & \mitempref{Curses, Lesser} \\
                \mitemref{Acid} & Know \spell{acid arrow} spell & \mitempricetable{Active}{3} & Arcane & \mitempref{Acid} \\
                \mitemref{Electricity} & Know \spell{lightning bolt}, \spell{shocking grasp} spells & \mitempricetable{Active}{3} & Arcane, Nature & \mitempref{Electricity} \\
                \mitemref{Fire} & Know \spell{fireball}, \spell{burning hands} spells & \mitempricetable{Active}{3} & Arcane, Nature & \mitempref{Fire} \\
                \mitemref{Sympathetic} & Heal when you deal damage & \mitempricetable{Active}{3} & Divine & \mitempref{Sympathetic} \\
                \mitemref{Time} & Know \spell{slow}, \spell{haste} spells & \mitempricetable{Active}{3} & Arcane & \mitempref{Time} \\
                \mitemref{Missiles} & Know \spell{missile storm}, \spell{magic missile} spells & \mitempricetable{Active}{4} & Arcane & \mitempref{Missiles} \\
                \mitemref{Cold} & Know \spell{ice spike}, \spell{cone of cold} spells & \mitempricetable{Active}{5} & Arcane, Nature & \mitempref{Cold} \\
                \mitemref{Curses} & Know \spell{curse of the wayward mind}, \spell{curse of blood and bone} spells & \mitempricetable{Active}{5} & Arcane, Divine & \mitempref{Curses} \\
                \mitemref{Electricity, Greater} & Know \spell{chain lightning}, \spell{lightning bolt}, \spell{shocking grasp} spells & \mitempricetable{Active}{6} & Arcane, Nature & \mitempref{Electricity, Greater} \\
                \mitemref{Fire, Greater} & Know \spell{immolation}, \spell{fireball}, \spell{burning hands} spells & \mitempricetable{Active}{6} & Arcane, Nature & \mitempref{Fire, Greater} \\
                \mitemref{Cold, Greater} & Know \spell{ice storm}, \spell{ice spike}, \spell{cone of cold} spells & \mitempricetable{Active}{7} & Arcane, Nature & \mitempref{Cold, Greater} \\
                \mitemref{Time, Greater} & Know \spell{time stop}, \spell{slow}, \spell{haste} spells & \mitempricetable{Active}{9} & Arcane & \mitempref{Time, Greater} \\
            \end{longtabu}
        \end{longtabuwrapper}
        \twocolumn

    \subsection{Implement Enhancement Bonuses}\label{Implement Enhancement Bonuses}

        Magical implements can have enhancement bonuses ranging from \plus1 to \plus5.
        Each \plus1 of enhancement bonus grants the ability to cast a spell without expending a spell slot once per day.
        You can use this ability to replace a spell slot of any level of spells you can cast.
        This does not take an action, besides the action normally required to cast the spell, or an item use.

        In addition, each \plus1 of enhancement bonus grants you an additional offensive legend point each day.
        These legend points can only be gained once per day, regardless of the number of implements you use.
        If you use multiple implements at once, or change between different implements, use only the highest number of legend points that applies.
        If you change from a weaker magical implement to a stronger magical implement, you gain legend points equal to the difference between the two enhancement bonuses.

        Offensive legend points gained from weapon enhancement bonuses do not stack with offensive legend points gained from implement enhancement bonuses.
        Use the higher of the two.

        \parhead{Implement Prices} The prices of enhancement bonuses to implements are listed in \trefnp{Implement Prices}, and the prices of magical properties are listed on \trefnp{Implement Properties}.
        If an implement has a magical property, the price of the property is added to the price of the staff.
        An implement can only have one property, and the price of that property cannot exceed twice the price of the enhancement bonus on the implement.

        \begin{dtable}
            \lcaption{Implement Prices}
            \begin{dtabularx}{\columnwidth} {>{\ccol}X c c}
                \tb{Enhancement Bonus} & \tb{Base Price} & \tb{Item Level}\\
                \bottomrule
                \plus1 & 150 gp    & 3rd  \\
                \plus2 & 800 gp    & 5th  \\
                \plus3 & 4,000 gp  & 9th  \\
                \plus4 & 20,000 gp & 13th \\
                \plus5 & 95,000 gp & 17th \\
            \end{dtabularx}
        \end{dtable}

        \parhead{Somatic Components} While wielding an implement, you may gesture with it and channel magic through it.
        These qualify as somatic components for the purpose of casting spells.

        \parhead{Spells Known and Attunement} Magical implements can provide additional spells known.
        Gaining access to the spells known from an implement requires attuning to it, which takes 10 minutes of concentration and an item use.
        Attunement lasts until you attune to a different implement.
        While you are attuned to an implement you wield, you gain access to additional spells known, as given in the description of the item.

        You must have be able to cast and learn spells of a given level to learn and cast spells of that level from an implement.

    \subsection{Implement Types}

        \subsubsection{Holy Symbols}

            \parhead{Physical Description} A typical holy symbol is a no larger than 4 inches in each dimension and can be easily held in the palm of a hand.
            Most holy symbols are metal, but they can be made from wood, bone, or even more exotic materials, depending on the deity they symbolize.

            \parhead{Special Rules} All holy symbols are divine implements.
            Most holy symbols are designed to be worn as an amulet in addition to being held in the hand.
            When worn in this way, the holy symbol occupies a torso body slot.
            A holy symbol worn on the body cannot be used to perform somatic components for spellcasting.
            However, it still grants enhancement bonuses and spells known as if it was being actively wielded.

        \subsubsection{Staffs}

            \parhead{Physical Description} A typical staff is 4 feet to 7 feet long and 2 inches to 3 inches thick, weighing about 5 pounds.
            Most staffs are wood, but a rare few are bone, metal, or even glass.
            (These are extremely exotic.)
            Staffs often have a gem or some device at their tip or are shod in metal at one or both ends.

            Staffs are often decorated with carvings or runes.
            A typical staff is like a walking stick, quarterstaff, or cudgel.
            It has AD 7, 10 hit points, hardness 5, and a break DR of 24.

        \subsubsection{Wands}

            \parhead{Physical Description} A typical wand is 6 inches to 12 inches long and about 1/4 inch thick, and often weighs no more than 1 ounce.
            Most wands are wood, but some are bone.
            A rare few are metal, glass, or even ceramic, but these are quite exotic.
            Occasionally, a wand has a gem or some device at its tip, and most are decorated with carvings or runes.
            A typical wand has AD 7, 5 hit points, hardness 5, and a break DR of 16.

            \parhead{Special Rules}
            Unlike most spellcasting implements, enhancement bonuses from wands wands only affect spells from a particular school of magic (see \pcref{Schools of Magic}), or spells with a particular tag (see \pcref{Ability Tags}).
            The prices for enhancement bonuses on wands are correspondingly lower, as given on \trefnp{Wand Prices}.

            \subparhead{Multiple Schools} Some rare wands provide bonuses to two schools.
            The enhancement bonus on a wand costs twice as much if it provides bonuses to two schools.
            A wand cannot provide an enhancement bonus to more than two schools.

            \begin{dtable}
                \lcaption{Wand Prices}
                \begin{dtabularx}{\columnwidth} {>{\ccol}X c c}
                    \tb{Enhancement Bonus} & \tb{Base Price} & \tb{Item Level}\\
                    \bottomrule
                    \plus1 wand & 50 gp     & 1st  \\
                    \plus2 wand & 300 gp    & 3rd  \\
                    \plus3 wand & 1,500 gp  & 7th  \\
                    \plus4 wand & 7,500 gp  & 10th \\
                    \plus5 wand & 35,000 gp & 14th \\
                \end{dtabularx}
            \end{dtable}

            % 1st-3rd should have 1 spell on the staff
            % 4th-6th should have 2 spells on the staff
            % 7th-9th should have 3 spells on the staff

    \subsection{Implement Properties}\label{Implement Properties}

        \begin{magicitemdef}[Arcane]{Acid}[Active]{3}{Conjuration [Acid, Creation, Physical]}{as implement}
            \magicitemability{Attuned} While wielding this item, you gain \spell{acid arrow} as a spell known.
            In addition, you can use item uses in place of spell slots to cast spells known from this item.
        \end{magicitemdef}

        \begin{magicitemdef}[Arcane, Nature]{Cold, Lesser}[Active]{2}{Evocation [Cold]}{as implement}
            \magicitemability{Attuned} While wielding this item, you gain \spell{cone of cold} as a spell known.
            In addition, you can use item uses in place of spell slots to cast spells known from this item.
        \end{magicitemdef}

        \begin{magicitemdef}[Arcane, Nature]{Cold}[Active]{5}{Conjuration/Evocation [Cold, Creation]}{as implement}
            \magicitemability{Attuned} While wielding this item, you gain \spell{ice spike} and \spell{cone of cold} as spells known.
            In addition, you can use item uses in place of spell slots to cast spells known from this item.
        \end{magicitemdef}

        \begin{magicitemdef}[Arcane, Nature]{Cold, Greater}[Active]{7}{Conjuration/Evocation [Cold, Creation]}{as implement}
            \magicitemability{Attuned} While wielding this item, you gain \spell{ice storm}, \spell{ice spike}, and \spell{cone of cold} as spells known.
            In addition, you can use item uses in place of spell slots to cast spells known from this item.
        \end{magicitemdef}

        \begin{magicitemdef}[Arcane, Divine]{Curses, Lesser}[Active]{2}{Vivimancy [Curse]}{as implement}
            \magicitemability{Attuned} You gain \spell{curse of blood and bone} as a spell known.
            In addition, you can use item uses in place of spell slots to cast spells known from this item.
        \end{magicitemdef}

        \begin{magicitemdef}[Arcane, Divine]{Curses}[Active]{5}{Vivimancy [Curse]}{as implement}
            \magicitemability{Attuned} You gain \spell{curse of the wayward mind} and \spell{curse of blood and bone} as spells known.
            In addition, you can use item uses in place of spell slots to cast spells known from this item.
        \end{magicitemdef}

        \begin{magicitemdef}[Arcane, Nature]{Electricity, Lesser}[Active]{1}{Evocation [Electricity]}{as implement}
            \magicitemability{Attuned} While wielding this item, you gain \spell{shocking grasp} as a spell known.
            In addition, you can use item uses in place of spell slots to cast spells known from this item.
        \end{magicitemdef}

        \begin{magicitemdef}[Arcane, Nature]{Electricity}[Active]{3}{Evocation [Electricity]}{as implement}
            \magicitemability{Attuned} While wielding this item, you gain \spell{lightning bolt} and \spell{shocking grasp} as spells known.
            In addition, you can use item uses in place of spell slots to cast spells known from this item.
        \end{magicitemdef}

        \begin{magicitemdef}[Arcane, Nature]{Electricity, Greater}[Active]{6}{Evocation [Electricity]}{as implement}
            \magicitemability{Attuned} While wielding this item, you gain \spell{chain lightning}, \spell{lightning bolt}, and \spell{shocking grasp} as spells known.
            In addition, you can use item uses in place of spell slots to cast spells known from this item.
        \end{magicitemdef}

        \begin{magicitemdef}[Arcane, Nature]{Fire, Lesser}[Active]{1}{Evocation [Fire]}{as implement}
            \magicitemability{Attuned} While wielding this item, you gain \spell{burning hands} as a spell known.
            In addition, you can use item uses in place of spell slots to cast spells known from this item.
        \end{magicitemdef}

        \begin{magicitemdef}[Arcane, Nature]{Fire}[Active]{3}{Evocation [Fire]}{as implement}
            \magicitemability{Attuned} While wielding this item, you gain \spell{fireball} and \spell{burning hands} as spells known.
            In addition, you can use item uses in place of spell slots to cast spells known from this item.
        \end{magicitemdef}

        \begin{magicitemdef}[Arcane, Nature]{Fire, Greater}[Active]{7}{Evocation [Fire]}{as implement}
            \magicitemability{Attuned} While wielding this item, you gain \spell{immolation}, \spell{fireball}, and \spell{burning hands} as spell known.
            In addition, you can use item uses in place of spell slots to cast spells known from this item.
        \end{magicitemdef}

        \begin{magicitemdef}[Divine, Nature]{Healing, Lesser}[Active]{1}{Vivimancy [Life]}{as implement}
            \magicitemability{Attuned} You gain \spell{cure wounds} as a spell known.
            In addition, you can use item uses in place of spell slots to cast spells known from this item.
        \end{magicitemdef}

        \begin{magicitemdef}[Arcane, Nature]{Healing}[Active]{6}{Vivimancy [Life]}{as implement}
            \magicitemability{Attuned} You gain \spell{heal} and \spell{cure wounds} as spells known.
            In addition, you can use item uses in place of spell slots to cast spells known from this item.
        \end{magicitemdef}

        \begin{magicitemdef}[Arcane]{Missiles, Lesser}[Active]{1}{Evocation}{as implement}
            \magicitemability{Attuned} While wielding this item, you gain \spell{magic missile} as a spell known.
            In addition, you can use item uses in place of spell slots to cast spells known from this item.
        \end{magicitemdef}

        \begin{magicitemdef}[Arcane]{Missiles}[Active]{4}{Evocation}{as implement}
            \magicitemability{Attuned} While wielding this item, you gain \spell{missile storm} and \spell{magic missile} as spells known.
            In addition, you can use item uses in place of spell slots to cast spells known from this item.
        \end{magicitemdef}

        \begin{magicitemdef}[Arcane, Nature]{Shielding, Lesser}[Active]{2}{Abjuration [Shielding]}{as implement}
            \magicitemability{Attuned} You gain \spell{inertial shield} as a spell known.
            In addition, you can use item uses in place of spell slots to cast spells known from this item.
        \end{magicitemdef}

        \begin{magicitemdef}[Arcane, Nature]{Shielding}[Active]{4}{Abjuration/Evocation [Shielding]}{as implement}
            \magicitemability{Attuned} You gain \spell{fire shield} as spells known.
            In addition, you can use item uses in place of spell slots to cast spells known from this item.
        \end{magicitemdef}

        \begin{magicitemdef}[Divine]{Sympathetic}[Active]{3}{Vivimancy [Life]}{as implement}
            \magicitemability{Active}[Immediate action] When you deal damage to a creature with a spell you cast, you can activate this item.
            If you do, you heal an ally within \rngclose range for 1d6 points of damage per two item power.

            After you activate this ability, the holy symbol radiates a gentle warmth, and a faint heartbeat can be felt from it.
            This effect lasts for 5 rounds, during which time you cannot activate this ability again.
        \end{magicitemdef}

        \begin{magicitemdef}[Arcane]{Time, Lesser}[Active]{1}{Transmutation [Temporal]}{as implement}
            \magicitemability{Attuned} While wielding this item, you gain \spell{haste} as a spell known.
            In addition, you can use item uses in place of spell slots to cast spells known from this item.
        \end{magicitemdef}

        \begin{magicitemdef}[Arcane]{Time}[Active]{3}{Transmutation [Temporal]}{as implement}
            \magicitemability{Attuned} While wielding this item, you gain \spell{slow} and \spell{haste} as spells known.
            In addition, you can use item uses in place of spell slots to cast spells known from this item.
        \end{magicitemdef}

        \begin{magicitemdef}[Arcane]{Time, Greater}[Active]{9}{Transmutation [Temporal]}{as implement}
            \magicitemability{Attuned} While wielding this item, you gain \spell{time stop}, \spell{haste}, and \spell{slow} as spells known.
            In addition, you can use item uses in place of spell slots to cast spells known from this item.
        \end{magicitemdef}

\section{Tools}

    \onecolumn
    \subsection{Tool Table}

        \begin{longtabuwrapper}
            \begin{longtabu}{l X l l l l}
                \tb{Wondrous Items} & \tb{Description} & \tb{Cost} & \tb{Item Level} & \tb{Page} \\
                \mitemref{Answerstone} & Answers yes or no questions & \mitempricetable{Active}{1} & \mitempref{Answerstone} \\
                \mitemref{Bag of Holding}, Small & Objects within are weightless & \mitempricetable{Passive}{0} & \mitempref{Bag of Holding} \\
                \mitemref*{Bag of Holding}, Medium & Objects within are weightless & \mitempricetable{Passive}{1} & \mitempref{Bag of Holding} \\
                \mitemref{Candle of Truth} & Can create a zone of truth & \mitempricetable{Active}{2} & \mitempref{Candle of Truth} \\
                \mitemref{Friendstone} & Can summon an attuned creature once & 1,500 gp & 7th & \mitempref{Friendstone} \\
                \mitemref*{Bag of Holding}, Large & Objects within are weightless & \mitempricetable{Passive}{2} & \mitempref{Bag of Holding} \\
                \mitemref{Truthstone} & Answers yes or no questions accurately & \mitempricetable{Active}{3} & \mitempref{Answerstone} \\
                \mitemref{Witch's Broom} & Flying broomstick & \mitempricetable{Passive}{7} & \mitempref{Witch's Broom} \\

                \tb{Scrolls and Potions} & \tb{Description} & \tb{Cost} & \tb{Item Level} & \tb{Page} \\
                1st-level & Contains 1st level spell & 10 gp & 1/2 & \tdash \\
                2nd-level & Contains 2nd level spell & 40 gp & 1st & \tdash \\
                3rd-level & Contains 3rd level spell & 100 gp & 2nd & \tdash \\
                4th-level & Contains 4th level spell & 250 gp & 3rd & \tdash \\
                5th-level & Contains 5th level spell & 600 gp & 5th & \tdash \\
                6th-level & Contains 6th level spell & 1,500 gp & 7th & \tdash \\
                7th-level & Contains 7th level spell & 3,000 gp & 9th & \tdash \\
                8th-level & Contains 8th level spell & 7,000 gp & 11th & \tdash \\
                9th-level & Contains 9th level spell & 15,000 gp & 12th & \tdash \\
            \end{longtabu}
        \end{longtabuwrapper}
        \twocolumn

    \subsection{Scrolls}
        A scroll is a spell that has been stored in written form.
        You can use a scroll to cast the spell on the scroll, just like casting a spell you know.
        Once a scroll is used, the magic is drained from it, and it cannot be used again.

        \parhead{Physical Description} A typical scroll is a heavy sheet of fine vellum or high-quality paper.
        An area about 8 1/2 inches wide and 11 inches long is sufficient to hold one spell.
        The sheet is reinforced at the top and bottom with strips of leather slightly longer than the sheet is wide.
        A scroll holding more than one spell has the same width (about 8 1/2 inches) but is an extra foot or so long for each extra spell.
        A scroll has AD 9, 1 hit point, hardness 0, and a break DR of 8.

        To protect it from wrinkling or tearing, a scroll is rolled up from both ends to form a double cylinder.
        (This also helps the user unroll the scroll quickly.) The scroll is usually placed in a tube of ivory, jade, leather, metal, or wood.
        Most scroll cases are inscribed with magic symbols which often identify the owner or the spells stored on the scrolls inside.
        The symbols can also be used to hide magic traps.

        \parhead{Activation} Activating a scroll simply means casting the spell written on the scroll.
        This functions exactly like casting the spell normally, except that you must spend an item use and the scroll itself rather than a spell slot, and you do not need to know the spell.
        You use your own spellpower with that spell source to determine the effects of the spell.
        You must holding the scroll and able to see and read the writing on it.

        \parhead{Activation Restrictions} To successfully activate a scroll, you must be able to cast spells of the same level with the same \glossterm{spell source} as the scroll.
        In addition, the spell on the scroll must be on your \glossterm{spell list} for that source.

        \parhead{Deciphering Scrolls} The writing on a scroll must be deciphered before a character can use it or know exactly what spell it contains.
        This requires a \ritual{read magic} ritual or a successful Spellcraft check (DR 20 \add double spell level).
        Deciphering a scroll to determine its contents does not activate the scroll.
        However, a scroll can be cursed or magically trapped to harm a creature trying to decipher it.

    \subsection{Potions and Oils}

        Potions and oils are magical liquids that can be activated to produce spell effects when drunk or used to coat a surface.
        Potions have their effects when drunk, and always target the creature drinking the potion.
        Oils have their effects when applied externally, and always target the creature or object the oil is applied to.
        Once activated, the magic in a potion or oil is expended and useless.

        Both potions and oils can only mimic the effects of \glossterm{targeted spells} with a casting time of a standard action or less.
        Some spells are useless to create as a potion or oil.
        For example, the \spell{suggestion} spell requires speaking a suggestion as part of casting the spell.
        Since the drinker of a \spell{suggestion} potion would not hear the suggestion used to cast the spell, such a potion would be useless.

        Once activated, the effects of a potion or oil function as if the target had cast the spell on itself.
        Neither the character activating the effect nor the target of the effect can make any decisions about the spell's effects, including concentrating on the effect.
        All decisions must be made by the creator of the potion when it is made.

        \parhead{Physical Description} A typical potion or oil consists of 3 ounces of liquid held in a ceramic or glass vial fitted with a tight stopper.
        The stoppered container is usually no more than 2 inches wide and 4 inches high.
        The vial has AD 13, 1 hit point, hardness 1, and a break DR of 12.
        Vials hold 3 ounces of liquid.

        \parhead{Identifying Potions and Oils}
        The magical aura on a potion or oil can be identified with a Spellcraft check (see \pcref{Spellcraft}), revealing the nature of the spell contained in the potion.
        If you sample a taste from the potion as you use the Spellcraft skill, you gain a \plus5 bonus to the Spellcraft check.
        You cannot accidentally activate the spell effect while sampling the potion, but you risk being poisoned or otherwise negatively effected if the potion is not as it seems.
        The \spell{identify} ritual can also be used to identify potions.

        Alternately, you can simply learn to identify potions by memory.
        Potions with the same spell effect from the same \glossterm{spell source} almost always have a similar appearance and taste.
        This can be dangerous if you encounter a liquid designed to look and taste like a potion it is not.

        \parhead{Activation} As a standard action, you can drink a potion, apply an oil to a creature or object within your \glossterm{reach}, or feed a potion to another willing creature within your reach.
        If you spend a magic item use, the spell contained takes effect at the same time that spells cast that round resolve.
        If you drink multiple potions or apply multiple oils at the same time, you can still only activate one of them.

        If you are physically unable to drink a potion or smear on oil, you cannot activate the item's effects.
        You can attempt to apply an oil to an unwilling target.
        To do so, you must make a touch attack against your target (see \pcref{Touch Attacks}).

        \parhead{Spellpower} The spellpower for a standard potion or oil is normally equal to twice the spell level of the spell in the potion.
        It is possible to create potions and oils with a higher spellpower than that minimum.
        For every 2 points by which the spellpower of a potion or oil increases above the minimum for its spell level, the price doubles additively.
        For example, a potion of a 1st level spell with a spellpower of 6 would cost 30 gp.

        \parhead{Mishaps} Extraordinarily powerful potions and oils can be dangerous to activate.
        Whenever you activate a potion or oil, if the \glossterm{power} of the potion exceeds your level, roll 1d10 and subtract the difference between your level and the potion's power.
        Compare the result to \trefnp{Potion Mishaps}.

        \begin{dtable}
            \lcaption{Potion Mishaps}
            \begin{dtabularx}{\columnwidth}{l X}
                \tb{Mishap Result} & \tb{Mishap Effect}                                       \\
                \bottomrule
                6\plus             & No additional effects                                    \\
                3--5               & You also take damage equal to the potion's spellpower    \\
                1--2               & As above, and you are \nauseated for 1 round             \\
                0 or below         & As above, and the potion does not have its normal effect \\
            \end{dtabularx}
        \end{dtable}

    \subsection{Rituals}

        \begin{dtable}
            \caption{Ritual Costs}
            \begin{dtabularx}{\columnwidth}{X l l l }
                \tb{Ritual Level} & \tb{Cost to Perform} & \tb{Cost to Scribe} & \tb{Item Level} \\
                \bottomrule
                1st-Level & 5 gp    & 50 gp    & 1st  \\
                2nd-Level & 20 gp   & 200 gp   & 3rd  \\
                3rd-Level & 50 gp   & 500 gp   & 4th  \\
                4th-Level & 125 gp  & 1250 gp  & 7th  \\
                5th-Level & 300 gp  & 3000 gp  & 9th  \\
                6th-Level & 750 gp  & 7500 gp  & 11th \\
                7th-Level & 1500 gp & 15000 gp & 12th \\
                8th-Level & 3500 gp & 35000 gp & 14th \\
                9th-Level & 7500 gp & 75000 gp & 16th \\
            \end{dtabularx}
        \end{dtable}

    \subsection{Wondrous Items}

        Wondrous items are items which are inherently magical in some way.

        \begin{magicitemdef}{Answerstone}[Active]{1}{Divination [Knowledge]}{bone, metal, or stone}
            \magicitemability{Active}[Standard action] If you shake this stone and ask a simple yes or no question, you receive an answer of either ``yes'', ``no'', or ``unclear''.
            The answer has a 75\% chance of being correct.
            If the answerstone is incorrect, it will answer ``yes'' or ``no'' randomly.
            After being asked a question once, it will always give the same answer to that question (and any sufficiently similar question) for 24 hours.

            The answerstone will always answer ``unclear'' if asked questions about events more than thirty minutes into the future or past, or farther than ten miles from its current location.
            It can only answer questions about observable quantities (including spells).

            Although answerstones are usually made of stone, they can also be made of other sturdy materials.
        \end{magicitemdef}

        \begin{magicitemdef}{Truthstone}[Active]{3}{Divination [Knowledge]}{bone, metal, or stone}
            \magicitemability{Active}[Standard action] This item functions like an answerstone, except that it does not have a random chance of failure.
        \end{magicitemdef}

        \begin{dtable*}
            \lcaption{Bags of Holding}
            \begin{dtabularx}{\textwidth}{l l X X X}
                \tb{Size} & \tb{Bag Weight} & \tb{Weight Limit} & \tb{Base Price} & \tb{Item Level} \\
                \bottomrule
                Small   & 1/2 lb. & 500 lb. & 200 gp    & 3rd  \\
                Medium  & 4 lb.   & 2,000 lb. & 800 gp  & 5th  \\
                Large   & 32 lb.  & 8,000 lb. & 2,000 gp  & 8th \\
            \end{dtabularx}
        \end{dtable*}
        \begin{magicitemdef}{Bag of Holding}{1}{Conjuration [Planar]}{textiles}
            \magicitemability{Passive} A bag of holding appears to be an unusually thick cloth sack.
            Objects completely inside the bag are nearly weightless, and do not change the bag's weight.
            There are three sizes of bags: Small, Medium, and Large.
            Each size of bag can hold an object or creature of its size, and has a limit on the weight it can contain.
            Any weight beyond the bag's weight limit is not negated, causing the bag to weigh more.
            The weight limits for each bag are shown in \trefnp{Bags of Holding}.
        \end{magicitemdef}

        \begin{magicitemdef}{Bottle of Air}[Active]{1}{Conjuration [Creation]}{ceramics}
            This item appears to be a normal glass bottle with a cork.
            \magicitemability{Active}[Standard action] If you uncork the bottle, it continually fills itself with clean air for 1 hour.
            It will remain filled with clean air even if taken into an airless environment or environment with dangerous gases, allowing you to breathe from the bottle.
            Liquid can still fill the bottle, and the bottle will continuously bubble as air is created until the liquid is removed.
            If the bottle is corked (a standard action), it stops filling itself with air until activated again.
        \end{magicitemdef}

        \begin{magicitemdef}{Candle of Truth}[Active]{2}{Enchantment [Compulsion, Mind]}{alchemy}
            This item appears to be a simple white candle.
            \magicitemability{Active}[Standard action] If you light this candle, it mimics the effect of the \spell{zone of truth} ritual, with a spellpower equal to its item power.
            The zone lasts as long as the candle remains lit, which is normally 1 hour.
            If the candle is snuffed, the zone immediately ends.
            While lit, the candle does not burn down, and provides servicable candle-light.
        \end{magicitemdef}

        \begin{magicitemdef}{Friendstone}[1500]{6}{Conjuration/Divination [Scrying, Teleportation]}{ceramics or stone}
            This item appears to be a glossy, smooth stone or ball of glass about three inches in diameter.
            \magicitemability{Active}[Standard action] If you press the stone to your forehead and speak your full name, you can attune to it.
            If you do, you can be summoned by a creature using the stone.
            This replaces the stone's attunement to any other creature.
            \magicitemability{Active}[Standard action] If you grasp the stone in your hand and throw it to the ground, you can activate it.
            if you do, you instantly teleport the creature attuned to the stone to the stone's location.
            The creature is granted a brief glimpse of the stone's surroundings, as if looking out through the stone, and may refuse the summoning.
            If the creature refuses, or if it is physically impossible for the creature to appear, your activation of the stone has no effect.
            If the creature accepts, the stone is destroyed and the creature is unerringly teleported into the stone's location.
            The creature must be on the same plane as the stone, but the teleportation works across any distance.
        \end{magicitemdef}

        \begin{magicitemdef}{Witch's Broom}[Passive]{7}{Divination/Transmutation [Communication, Imbuement]}{wood}
            \magicitemability{Passive} This broom has a 30 foot fly speed with good maneuverability.
            You can ride the broom to fly.
            Riding the broom is like riding a mount, except that you control it perfectly and some actions (such as taking cover behind the mount) are infeasible due to the shape of the broom.
            It can carry up to 1,000 pounds before it becomes unable to fly.
            \magicitemability{Active}[Standard action] As a standard action, you can speak a command word to summon the broom.
            If you do, and the broom is within \rngext range, it flies next to you and hovers at a convenient height to ride.
            The broom must be able to hear the commmand word.
        \end{magicitemdef}

\section{Special Materials}

    Some materials are inherently magical.
    Items made from such materials gain special properties automatically.

    \subsection{Material Types}
        There are three main types of materials used to create items: metal, wood, and leather.
        In order to create an item with a special material, the item must normally be made with a material of the appropriate material type.
        In the case of items that are made from multiple materials, such as polearms, the primary functional part of the item determines which type of special material can be used to create it.
        For example, the material type for a weapon is generally determined by its striking surface.

    \subsection{Special Material Descriptions}

        \begin{dtable*}
            \lcaption{Special Materials}
            \begin{dtabularx}{\columnwidth}{l l X l l}
                \tb{Material Name} & \tb{Material Type} & \tb{Effect} & \tb{Weapon Price (Level)} & \tb{Armor Price (Level)} \\
                \bottomrule
                Adamantine  & Metal   & Grant or overcome physical damage reduction & 12,000 gp (12th) & 30,000 gp (14th) \\
                Cold Iron   & Metal   & Bonuses against enchantments and mages      & 2,000 gp (8th) & 5,000 gp (10th) \\
                Darkwood    & Wood    & Extraordinarily light                       & 2,000 gp (8th) & 5,000 gp (10th) \\
                Dragonbone  & Metal   & Inherently magical                          & 30,000 gp (14th) & 60,000 gp (16th) \\
                Dragonhide  & Leather & Resist energy                               & \tdash & 2,000 gp (8th) \\
                Dragonscale & Metal   & Resist energy                               & \tdash & 2,000 gp (8th) \\
                Ironwood    & Metal   & Metallic wood                               & \tdash & 100 gp (5th) \\
                Mithral     & Metal   & Extraordinarily light                       & 2,000 gp (8th) & 5,000 gp (10th) \\
                Silvered    & Metal   & Effective against supernatural creatures    & 100 gp (3rd) & 200 gp (3rd) \\
            \end{dtabularx}
        \end{dtable*}

        \parhead{Adamantine}
        Adamantine is a rare metal that is among the hardest substances known.

        Adamantine weapons ignore the \glossterm{hardness} of creatures and objects.

        Adamantine body armor grants its wearer damage reduction against physical damage equal to your item power.
        This damage reduction is overcome by adamantine weapons.
        Adamantine shields have no special effects.

        Adamantine items of any kind have double the hit points and hardness of an equivalent item of the same type.

        \parhead{Cold Iron}

        Cold iron is iron that has been magically smelted without the use of heat.

        Cold iron items are especially effective against some supernatural creatures, as indicated in the creature descriptions.
        In addition, cold iron items have inherent \glossterm{magic resistance} equal to 10 \add the item's power.
        This magic resistance is unique to the item, and is not granted to its user.
        However, the maximum number and price of magical properties that can be imbued into the armor is halved.

        When you damage a creature with a cold iron weapon, it is \impaired with casting arcane spells for 2 rounds.

        Cold iron body armor grants its wearer immunity to \glossterm{Compulsion} and \glossterm{Delusion} effects.
        Cold iron shields have no special effects.

        \parhead{Darkwood}
        Darkwood is a rare magic wood that is very light.

        Darkwood weapons have the Finesse weapon property, allowing you to apply your Dexterity instead of your Strength when determining your accuracy with physical attacks using the weapon, even if it isn't a light weapon for you.

        Darkwood shields have their \glossterm{encumbrance penalty} reduced by 2, and their arcane spell failure reduced by 10\%.

        Darkwood items of any kind weigh half as much as an equivalent item of the same type.

        \parhead{Dragonbone}

        Dragon bones can be used in place of metal when creating weapons and armor.

        Dragonbone weapons grant twice the normal number of offensive legend points from their enhancement bonus.
        This does not affect the damage dealt by the weapon.

        Dragonbone armor grants its wearer \glossterm{magic resistance} equal to 10 \add item power.

        \parhead{Dragonhide}

        Dragon hide can be used as leather when creating armor.

        Dragonhide body armor grants damage reduction equal to your level against the energy type used by the dragon the hide is taken from.

        \parhead{Dragonscale}

        Dragon scales can be used in place of metal when creating armor.

        Dragonscale armor grants damage reduction equal to your level against the energy type used by the dragon the hide is taken from.

        \parhead{Ironwood}
        Ironwood is a magic wood created by the \spell{ironwood} ritual.
        It has been magically hardened, giving it the strength of iron.
        Ironwood can be used in place of metal when creating weapons and armor.

        Ironwood items have no special properties except that they are made from wood instead of metal.

        \parhead{Mithral}
        Mithral is a rare metal that is very light, but has the strength of iron.

        Mithral weapons have the Finesse weapon property, allowing you to apply your Dexterity instead of your Strength when determining your accuracy with physical attacks using the weapon, even if it isn't a light weapon for you.

        Mithral armor has its \glossterm{encumbrance penalty} reduced by 2, and its arcane spell failure reduced by 10.

        Mithral items of any kind weigh half as much of an equivalent item of the same type.

        \parhead{Silvered}

        Silvered items have silver infused into them in the process of their creation.
        Only metal items made of iron or steel can be silvered.

        Silvered items are especially effective against some supernatural creatures, as indicated in the creature descriptions.
        However, silvered weapons deal damage as if they were one size category smaller, and mithral armor has its defense bonus reduced by 1.

\section{Magic Item Rules}

    \subsection{Magic Item Auras}

        Magic items radiate magical auras which can be detected with the Spellcraft skill (see \pcref{Spellcraft}).
        Each item describes the auras that can be detected on it, including the strength, school, and descriptors, as appropriate.

    \subsection{Damaging Magic Items}

        A magic item is normally unharmed by attacks unless it is unattended or is specifically targeted by an effect.
        A magic item's non-physical defenses are all equal to 10 \add the spellpower of the item.
        The only exceptions to this are intelligent magic items, which apply their Willpower to their Mental defense.

        Magic items, unless otherwise noted, take damage as nonmagical items of the same sort.
        A damaged magic item continues to function, but if it is \glossterm{broken}, its magic ceases to function until it is repaired.
        If it is destroyed, all its magical power is lost.

    \subsection{Repairing Magic Items}

        A magic item which is \glossterm{broken} (but not destroyed) can be repaired for 10\% of the value of the item.

    \subsection{Intelligent Items}

        Some magic items, particularly weapons, have an intelligence all their own.
        Only permanent magic items (as opposed to those with a single use or those with charges) can be intelligent.
        (This means that potions, scrolls, and wands, among other items, are never intelligent.)

        In general, fewer than 1\% of magic items have intelligence.

    \subsection{Cursed Items}

        Some items are cursed -- incorrectly made, or corrupted by outside forces.
        Cursed items might be particularly dangerous to the user, or they might be normal items with a minor flaw, an inconvenient requirement, or an unpredictable nature.
        Many cursed items are difficult to identify and remove, requiring the use of rituals such as \spell{remove curse}.
