\chapter{Monster Mechanics}

Monsters are all of the various non-humanoid creatures that exist in the world of Rise.
Many of them are dangerous, and adventurers may need to fight them.
This chapter describes the rules for monsters, and the combat statistics for a variety of monsters.

\section{Monster Statistics}
  Like player characters, monsters have levels, attributes, and abilities.
  However, they do not have classes, insight points, or many other elements of characters.
  This section defines how monsters function.

  \subsection{Monster Roles}
    Each monster has a role.
    A monster's role influences its statistics, and generally indicates the way it fights.
    There are six roles: brute, skirmisher, warrior, sniper, mystic, and leader.
    The effects of each role are described below.

    \subsubsection{Brute}
      A typical brute rushes into melee and smashes its enemies with highly damaging attacks.
      Brutes have high hit points and reasonable defenses, but also become \glossterm{injured} easily.
      Most brutes charge directly at the enemy and fight in melee, though some prefer to use a few ranged attacks before entering the fray.

      \parhead{Defenses} \plus4 Armor, \plus5 Brawn, \plus4 Fortitude, \plus4 Reflex, \plus3 Mental
      \parhead{Durability} \plus6.
      \parhead{Injury Point} Three-quarters of total hit points.

    \subsubsection{Leader}
      A typical leader commands their allies from behind the front lines, but is willing to enter the fray to protect more vulnerable allies.
      Leaders have well-rounded statistics, but they do not generally fight well alone.
      Instead, they tend to be more effective when aiding or commanding their allies.

      \parhead{Defenses} \plus4 Armor, \plus4 Brawn, \plus4 Fortitude, \plus4 Reflex, \plus4 Mental
      \parhead{Durability} \plus4.
      \parhead{Injury Point} Half of total hit points.

    \subsubsection{Skirmisher}
      A typical skirmisher uses its high mobility to fight only when it is advantageous to do so.
      Skirmishers can be hard to hit, though they are vulnerable to attacks that can't be dodged.
      They are likely to avoid intimidating front-line fighters.
      Instead, they ambush the vulnerable back lines if possible.

      \parhead{Defenses} \plus5 Armor, \plus4 Brawn, \plus4 Fortitude, \plus5 Reflex, \plus4 Mental
      \parhead{Durability} \plus2.
      \parhead{Injury Point} Half of total hit points.

    \subsubsection{Sniper}
      A typical sniper uses ranged attacks while staying as far away from combat as possible.
      Snipers tend to be fragile, so they can be quickly dispatched by creatures that are able to close the distance.

      \parhead{Defenses} \plus3 Armor, \plus3 Brawn, \plus3 Fortitude, \plus4 Reflex, \plus4 Mental
      \parhead{Durability} \plus2.
      \parhead{Injury Point} Three-quarters of total hit points.

    \subsubsection{Warrior}
      A typical warrior forms a defensive front line to protect any vulnerable allies.
      Warriors have high defenses, and can be very difficult to kill.
      However, they tend to lack offensive power and mobility, so they risk being outflanked and irrelevant.

      \parhead{Defenses} \plus5 Armor, \plus4 Brawn, \plus5 Fortitude, \plus3 Reflex, \plus4 Mental
      \parhead{Durability} \plus4.
      \parhead{Injury Point} One third of total hit points.

  \subsection{Level Scaling}
    Each monster has a level that indicates its approximate strength.
    Monsters have some specific level scaling rules that differ from the rules for player characters.
    This scaling is summarized in \tref{Monster Advancement}.
    In general, their automatic level scaling is designed to keep them on par with the power that player characters gain at higher levels, but without all of the complexity of high-level PCs.
    High level monsters should have slightly higher base values for their attributes, though this is not quantified, since monster attributes can be highly varied.
    They have the following specific changes.
    \begin{raggeditemize}
      \item Monsters calculate their \glossterm{injury point} based on their role rather than their level and Constitution.
        Their role defines their injury point as some multiplier of the monster's total hit points.
      \item Monsters only add half their attributes to their defenses instead of their full attribute.
      \item Monsters gain a \plus1 bonus to two attributes at 3rd level and every 3 levels thereafter.
        This replaces the attribute bonus player characters get.
      % Enhancement accuracy bonuses come online at rank 3 and rank 7, and we want to compensate for that.
      % We want to make sure this kind of universal bonus lands on an odd level to avoid two-point jumps with a single level.
      \item Monsters gain a \plus1 bonus to all defenses at 7th level and 19th level.
      % Our ideal HP by rank calcs assume we upgrade to a 80% hit rate at rank 5.
      % This is also nicely between the defense bonus scaling.
      % The 21st level bonus is mainly just because monsters otherwise don't get much of anything from 21st level,
      % and we want to make sure that extremely high level monsters are appropriately dangerous.
      \item Monsters gain a \plus1 accuracy bonus at 13th level and 21st level.
      % Monsters need to deal enough damage, and they don't get the extra damage effects that players get.
      \item Monsters gain a \plus1 power bonus at 5th level, 11th level, and 17th level.
    \end{raggeditemize}

    % NOTE: Keep in sync with Ship Statistics in Ship Combat
    \begin{dtable}
      \lcaption{Monster Advancement}
      \begin{compresseddtabularx}{\columnwidth}{l l l l >{\lcol}X}
        \tb{Level} & \tb{Rank} & \tb{Durability} & \tb{Bonus}\fn{1} & \tb{Special}             \tableheaderrule
        1st        & 1         & \plus0          & \tdash           & \tdash                                     \\
        2nd        & 1         & \plus1          & \plus1           & \tdash                                     \\
        3rd        & 1         & \plus2          & \plus1           & \plus1 to two attributes                   \\
        4th        & 2         & \plus2          & \plus2           & HP: 2x \glossterm{durability}              \\
        5th        & 2         & \plus3          & \plus2           & \plus1 power                               \\
        6th        & 2         & \plus4          & \plus3           & \plus1 to two attributes                   \\
        7th        & 3         & \plus4          & \plus3           & HP: 3x durability, \plus1 defenses  \\
        8th        & 3         & \plus5          & \plus4           & \tdash                                     \\
        9th        & 3         & \plus6          & \plus4           & \plus1 to two attributes                   \\
        10th       & 4         & \plus6          & \plus5           & HP: 4x durability                          \\
        11th       & 4         & \plus7          & \plus5           & \plus1 power                               \\
        12th       & 4         & \plus8          & \plus6           & \plus1 to two attributes                   \\
        13th       & 5         & \plus8          & \plus6           & HP: 6x durability, \plus1 accuracy         \\
        14th       & 5         & \plus9          & \plus7           & \tdash                                     \\
        15th       & 5         & \plus10         & \plus7           & \plus1 to two attributes                   \\
        16th       & 6         & \plus10         & \plus8           & HP: 8x durability                          \\
        17th       & 6         & \plus11         & \plus8           & \plus1 power                               \\
        18th       & 6         & \plus12         & \plus9           & \plus1 to two attributes                   \\
        19th       & 7         & \plus12         & \plus9           & HP: 10x durability, \plus1 defenses        \\
        20th       & 7         & \plus13         & \plus10          & \tdash                                     \\
        21st       & 7         & \plus14         & \plus10          & \plus1 accuracy, \plus1 to two attributes  \\
      \end{compresseddtabularx}
      1. This bonus applies to the monster's \glossterm{accuracy}, \magical power, mundane power, trained skills, and defenses. \\
    \end{dtable}

  \subsection{Elite Monsters}\label{Elite Monsters}
    Elite monsters are approximately four times as strong as an ordinary monster.
    They have a number of benefits and modifiers to make them more appropriate as a solo encounter than simply using a very high level ordinary monster:
    % Elite monsters deal about 1.3*1.2*2 = 3.1x damage of a baseline monster, with higher defenses.
    % Their higher attributes and access to weird abilities also increase offensive output in ways that are harder to calculate.
    % They also don't reduce their damage from losing 1/4 of their HP, so they are overall similar in power to 4x normal monsters.
    % NOTE: Keep in sync with elite ships
    \begin{raggeditemize}
      \item Elite monsters gain a \plus2 bonus to their \glossterm{power} and all \glossterm{defenses}.
      \item Elite monsters have three times the \glossterm{hit points} of standard monsters.
      \item Elite monsters can have a maximum starting attribute of 6 (see \pcref{Monster Attributes}).
        % elite actions should be about as strong as regular actions, so 2x more damage
      \item Elite monsters can take an additional \glossterm{elite action} each round (see \pcref{Elite Actions}).
      \item Elite monsters automatically remove \glossterm{conditions} (see \pcref{Monster Conditions}).
    \end{raggeditemize}

  \subsection{Monster Vital Wounds}
    Monsters do not normally make vital rolls like player characters do.
    Unless otherwise specified on the monster's description, vital wounds have no negative effects on monsters.
    Instead, once a monster gains a vital wound, it is \glossterm{defeated}.
    Defeated monsters may be unconscious, dead, or merely flee the fight at the discretion of the Game Master.

  \subsection{Monster Conditions}\label{Monster Conditions}
    Monsters normally can't use the \textit{recover} ability.
    Standard monsters can only remove \glossterm{conditions} by taking a \glossterm{short rest}.
    However, \glossterm{elite} monsters can remove conditions more easily with the \ability{elite cleanse} ability.
    Intelligence monsters can generally choose which condition they remove, while unintelligent monsters may choose instinctually or randomly.

    % The timing here is important.
    % If the monster could remove conditions as an action on its turn, then offensive debuffs would have a chance to be removed before they ever affected the monster's actions.
    % Conditions should never be worse than brief effects, so this has to be an end of round trigger.
    \begin{triggeredability}{Elite Cleanse}{Triggered}
      \rankline
      At the end of each round, roll 1d10.
      On an 7 or higher, the monster can remove one \glossterm{condition}.
      On a 10, it can also remove a different condition.
      Conditions applied during the current round cannot be removed in this way.
    \end{triggeredability}

    Whenever an elite monster gains a unique condition, it is added to a ``condition track''.
    The first condition is to the first slot in the track, the second condition is added to the second slot, and so on.
    If a monster gains two instances of the same condition, they are put in the same location on the track.

    At the end of each round, each elite monster rolls 1d6.
    If the result matches an existing condition on its condition track, that condition is removed unless all instances of that condition were gained during the current round.
    When a monster has multiple copies of the same condition, only one instance of that condition is removed, so it may have to roll the same number multiple times to fully remove the condition.
    If a location on the condition track becomes empty, all of the conditions to the right of that location shift down by 1.
    This means the highest occupied number on the condition track should always match the number of unique conditions the monster has.
    The oldest condition should always be number 1, and the most recent condition should have the highest occupied number on the track.

  \subsection{Monster Resources}
    Monsters have no resources, and their fatigue tolerance is 0.
    They are normally unable to use universal abilities that would cause them to increase their fatigue level, such as the \textit{desperate exertion} ability.
    They can use the \ability{sprint} ability, but only during the \glossterm{action phase}, since that does not increase their fatigue.

    If a monster has an explicitly listed special ability that increases its fatigue level, it can use that ability.
    Similarly, monsters that are allied with player characters may choose to increase their fatigue level to receive healing from the players, such as from the \spell{restoration} spell.
    This would immediately cause them to suffer fatigue penalties, since their fatigue tolerance is 0.

    At the GM's discretion, individual monsters may technically be attuned to their abilities or be attuned to magic items they find, but this is not an explicit part of monster definitions.
    This is typically only relevant if one of the players has the Null feat (see \pcref{Feats}).

  \subsection{Monster Ability Access}\label{Monster Ability Access}
    Although monsters do not gain archetypes, their special abilities should still be balanced appropriately for their maximum accessible rank.
    Like a player character, a monster's maximum accessible rank is based on its level, as shown in \tref{Monster Advancement}.

    It's not always meaningful to give every monster multiple active abilities.
    However, maneuvers provide significant power at high levels, so high level monsters should generally use special abilities like maneuvers instead of simple strikes.
    This also ensures that high level monsters feel meaningfully different from low-level monsters, rather than simply having higher statistics.

  \subsection{Monster Attributes}\label{Monster Attributes}
    Each of a standard monster's attributes can range from \minus9 to 4.
    Elite monsters can have attributes up to 6 at level 1, and they generally have higher attributes overall.
    A few specific monsters of legendary power can exceed this limit.
    In general, a monster with higher attributes will be stronger, but not all monsters need to start with the same attribute total.

  \subsection{Monster Natural Weapons}\label{Monster Natural Weapon}
    Most monsters that do not have hands, or equivalently agile limbs capable of holding weapons and shields, have more powerful natural weapons.
    They treat all non-\weapontag{Light} natural weapons as if they had the \weapontag{Heavy} weapon tag and were being held in two hands.
    This gives them a \plus1 damage bonus per 3 \glossterm{power}.

    Some specific monsters, like horses, do not gain this benefit because they are unable to effectively use their whole body to support their attacks.

  \subsection{Monster Dual Strikes}\label{Monster Dual Strikes}
    Monsters with a sufficiently high Dexterity are able to make dual strikes (see \pcref{Dual Strikes}).
    However, most monsters can only make \glossterm{dual strikes} with two copies of the same \weapontag{Light} natural weapon.
    Monsters that use \glossterm{manufactured weapons} can make dual strikes with them in the same way as player characters.

\section{Monster Combat Mechanics}

  \subsection{Monster Actions}\label{Monster Actions}
    All monsters are able to make movements and take free actions, minor actions, and standard actions in the same way as player characters.

    \subsubsection{Elite Actions}\label{Elite Actions}
      Elite monsters can take an additional \glossterm{elite action} each round.
      Every elite monster has at least one special ability which requires an elite action to use.
      Elite actions cannot be converted into standard or minor actions, and elite monsters cannot use a standard action to use abilities that require an elite action.

      In general, elite monsters should be designed to attack multiple different PCs in every round.
      If the full damage output of an elite monster is brought to bear on a single PC, it will likely kill that PC quickly, taking a player out of the fight too early.
      Alternately, if that PC is able to avoid or resist the damage, it will trivialize the fight and remove the sense of danger.
      Instead, damage should be spread throughout the party to ensure that multiple people feel threatened and have to adjust their actions based on their personal danger tolerance.

      Although splitting attacks between multiple defenders is usually a suboptimal battle strategy, it can be narratively justified in a variety of ways.
      It may be physically impractical for a monster to use all of its natural weapons on the same creature.
      Monsters may want to inflict debuffs on multiple attackers to reduce the damage they take, which may require dealing enough damage to inflict HP-based debuffs.

  \subsection{Encounter Balancing}\label{Encounter Balancing}
    In general, a group of PCs of a given level will have an appropriate challenge from fighting an equal number of monsters of their level.
    Alternately, they can fight a single elite monster of their level.
    You can generally replace one monster of a given level with two monsters that are two or three levels lower.
    Similarly, you can replace two monsters of a given level with one monster that is two or three levels higher.
    To make encounters easier or harder, you can change the number of monsters or the level of the monsters.

    It is generally not a good idea for PCs to fight monsters five or more levels higher or lower than their own level.
    They may find that their attacks always miss, or never miss, and other aspects of the encounter may be similarly imbalanced in ways that are not easily remedied by simply changing the number of enemies.

    There are many ways that you can make encounters easier or harder in ways that don't simply involve changing the level and number of monsters.
    If monsters enter the battle gradually over time, rather than all at once, you can use a larger number of monsters without overwhelming the players.
    Similarly, having multiple separate fights without letting the players finish a \glossterm{short rest} can dramatically increase the danger and difficulty of the later fights.

    The terrain and physical location of a fight can also dramatically change its difficulty.
    Narrow corridors and tight spaces favor smaller groups, highly defensive characters, and mid-range spellcasters who can attack safely from the back lines.
    Wide-open terrain favors large groups, highly mobile creatures, and long-range combatants like archers.
    A party with good tactics can easily hold a five-foot tunnel against dozens or even hundreds of melee-only enemy combatants, even if they would be swarmed and killed easily by those same enemies in an open field.

  \subsection{Monster Knowledge}
    You can remember relevant knowledge about monsters by making Knowledge skill checks.
    Each monster has a set of associated information that you can learn with a knowledge check of the listed difficulty value.
    Most monsters have multiple tiers of information that you can learn depending on how good your Knowledge check is.
    All information recorded in these descriptions is accurate, though lower check results may not provide full context for a monster's behavior or nature.

% \section{Quick Monster Creation}

%   This table describes the statistics for a non-elite monster with 0 in all attributes.
%   It can be a useful reference to when creating custom monsters at a particular role and level.
%   The format is extremely terse to save space: ``HP/IP; Armor/Fort/Ref/Ment''.

%   \input{generated/monster_reference_table.tex}

\chapter{Monster Descriptions}

  This chapter contains detailed descriptions for a wide variety of monsters.

  \input{generated/monster_descriptions.tex}
