\chapter{Mystic Spheres}\label{Mystic Spheres}

\section{Categories of Magic}
    % TODO: description

    \subsection{Magic Sources}
        There are four \glossterm{magic sources} that characters can use to cast spells and perform rituals: arcane (cast by mages), divine (cast by clerics and paladins), nature (cast by druids), and pact (cast by warlocks).
        Each magic source has a set of associated \glossterm{mystic spheres} (see Mystic Spheres, below).
        % TODO: more description

        \subsubsection{Characters with Multiple Magic Sources}
            A character can have access to multiple sources of magic through the use of abilities like the Hybrid Training ability (see \featpref{Half-Elves}).
            The \glossterm{mystic spheres}, spells, and rituals that character knows are tracked separately for each source of magic that character has access to.
            If you have access to the same spell or ritual from multiple sources, the two versions of the ability are generally considered to be the same ability.
            When you cast the spell or perform the ritual, you choose which source you are using for the ability.

    \subsection{Mystic Spheres}
        A \glossterm{mystic sphere} is a collection of thematically related magical effects that includes both \glossterm{spells} and \glossterm{rituals}.
        Each \glossterm{mystic sphere} is associated with a single school of magic and any number of \glossterm{magic sources}.
        The mystic spheres are listed at \pcref{Mystic Sphere Lists}.

    \subsection{Magic Schools}
        There are nine schools of magic.
        Each school of magic has a set of thematically related effects.
        Every spell and ritual belongs to at least one school of magic.
        % TODO: wording.
        When you cast a spell or perform a ritual, that ability is considered to be from the source you used to gain access to the ability.

        The schools of magic are described below.

        \subsubsection{Abjuration}
            Abjuration spells and rituals reduce or negate damage, magic, and other effects.
            They can be used to protect allies and remove harmful magic.

        \subsubsection{Channeling}
            Channeling spells and rituals call upon the power of deities or other supernatural entities.
            They can be used to do anything those entities could do.
            Only divine spellcasters have access to Channeling spells.

        \subsubsection{Conjuration}
            Conjuration spells and rituals create and transport objects and creatures.
            They can be used to summon allies, transport creatures, and create objects from thin air.

        \subsubsection{Divination}
            Divination spells and rituals grant knowledge.
            They can be used to reveal hidden truths, predict the future, or communicate at great distances.

        \subsubsection{Enchantment}
            Enchantment spells and rituals alter the minds of creatures.
            They can be used to influence, control, or debilitate creatures.
            Almost all enchantment spells and rituals have the \glossterm{Compulsion} or \glossterm{Emotion} tags,
            and some have the \glossterm{Subtle} tag as well (see \pcref{Ability Tags}).

        \subsubsection{Evocation}
            Evocation spells and rituals create and manipulate energy.
            They can be used to inflict damage with energy blasts or manipulate the environment.

        \subsubsection{Illusion}
            Illusion spells and rituals create or manipulate sensory impressions.
            They can be used to create or remove light, conceal things that exist, or cause creatures to perceive things that do not exist.

        \subsubsection{Transmutation}
            Transmutation spells and rituals change the properties of creatures and objects.
            They can be used to grant new abilities, enhance existing abilities, change a target's form, or even alter the flow of time itself.

        \subsubsection{Vivimancy}
            Vivimancy spells and rituals manipulate the power of life and death, as well as souls.
            They can be used to heal or inflict wounds, resurrect the dead, create undead monsters, and cripple the bodies of creatures.

\section{Spell and Ritual Mechanics}\label{Spell and Ritual Mechanics}

    Spells and rituals share many common properties, defined here.

    \subsection{Casting Components}\label{Casting Components}
        Unless otherwise noted, all spells and rituals require both \glossterm{verbal components} to cast or perform.
        In addition, arcane spells, pact spells, and all rituals require \glossterm{somatic components}.
        You cannot start casting a spell or performing a ritual without all required components.
        If you lose those components before the ability resolves, it is \glossterm{miscast}.

        To provide the verbal component for a spell or ritual, you must speak in a strong voice with a volume at least as loud as ordinary conversation.
        To provide the somatic component for a spell or ritual, you must make a measured and precise movement of at least one free hand.

        \subsubsection{Somatic Component Failure}\label{Somatic Component Failure}
            Encumbrance from armor interferes with the \glossterm{somatic components} required to perform arcane spells, pact spells, and all rituals.
            When you cast a spell or perform a ritual that requires \glossterm{somatic components} while you have an \glossterm{encumbrance}, you must roll 1d10.
            If your result is less than or equal to your \glossterm{encumbrance}, you \glossterm{miscast} the ability (see \pcref{Miscasting}).
            When you perform a ritual, this roll must be repeated at the end of each action phase during the ritual.

    \subsection{Concentration}\label{Concentration}
        Some abilities, such as spells and rituals, require focused thought to use successfully.
        If your focus is tested, such as by casting a spell defensively (see \pcref{Defensive Casting}), you may need to make a \glossterm{concentration} check.
        Your bonus with a concentration check is equal to your level or your Willpower, whichever is higher.
        You apply any \glossterm{overwhelm penalties} you suffer as penalties to concentration checks.

        Success on a concentration check generally means you successfully use the ability that triggered the check.
        Failure generally means that you do not use the ability successfully.

        If your concentration is tested in multiple ways simultaneously, you only make one \glossterm{concentration} check.
        You take a \minus2 penalty to the check for each effect that tests your concentration simultaneously.

    \subsection{Miscasting}\label{Miscasting}

        If you start using a spell or ritual and fail to complete it successfully for any reason, you miscast it.
        A miscast spell or ritual does not have its normal effect.
        Instead, a wave of magical energy causes a \glossterm{miscast backlash}.
        When a mystic backlash occurs, make an attack against the Mental defense of yourself and all creatures in a 5 foot radius from you.
        Your \glossterm{power} with this ability is equal to your \glossterm{power} with the spell or ritual you tried to use.
        On a hit, each target takes thaumatic \glossterm{standard damage} \minus1d.

        \parhead{Impossible Spells and Rituals} When you try to use a spell or ritual in an impossible way, the ability is \glossterm{miscast} instead.
        This most commonly happens if you attempt to declare an invalid target for a spell.
        For example, if you try to cast a spell that only affects living creatures on a creature that is undead, the spell would be miscast.

        % TODO: This may be more clear if minor + no miscast becomes a tag.
        \parhead{Minor Action Spells} Spells that require a \glossterm{minor action} to cast are not miscast if your concentration is broken while casting the spell.
        Instead, the spell simply fizzles without any effect.

    \subsection{Focused and Defensive Casting}\label{Focused and Defensive Casting}
        The concentration required to use spells and rituals can impair your ability to defend yourself.
        When you use a spell or ritual, you choose whether to use \glossterm{focused casting} or \glossterm{defensive casting}.
        For rituals, you can change your choice during each phase that you spend performing the ritual.
        % This is a hack to prevent playing games involving delaying casting until the delayed action phase
        % to avoid being disruptible without taking defense penalties.
        You cannot use \glossterm{focused casting} if you are distracted, as your concentration is divided.
        You are distracted if you are experiencing violent motion, such as if you are on a galloping horse, or if you already took damage during the current round.

        \subsubsection{Focused Casting}\label{Focused Casting}
            If you use a spell or ritual with \glossterm{focused casting}, you suffer a \minus4 penalty to all defenses during each \glossterm{phase} where you concentrate on the ability.

        \subsubsection{Defensive Casting}\label{Defensive Casting}
            If you use a spell or ritual using \glossterm{defensive casting}, you suffer no penalty to defenses.
            In exchange, you have to make a \glossterm{concentration} check to use the ability successfully (see \pcref{Concentration}).
            The \glossterm{difficulty rating} of the check is equal to 5 \add three times the level of the spell you are casting, as shown on \tref{Concentration DRs}.
            Success means that you use the ability successfully.
            Failure means that you \glossterm{miscast} the spell or ritual (see Miscasting, below).

            When you use defensive casting to perform a ritual, you must make a new check during each phase.

            \begin{dtable}
                \lcaption{Concentration DRs}
                \begin{dtabularx}{\columnwidth}{l X}
                    \tb{Spell Level} & \tb{DR} \tableheaderrule
                    Cantrip & 5 \\
                    1st level & 8 \\
                    2nd level & 11 \\
                    3rd level & 15 \\
                    4th level & 18 \\
                    5th level & 21 \\
                    6th level & 24 \\
                    7th level & 27 \\
                \end{dtabularx}
            \end{dtable}


    \subsection{Augments}\label{Augments}
        There are a number of \glossterm{augments} that can be applied to spells and rituals to increase their power.
        Each augment has a name, a level, and an effect.
        When you cast a spell or perform a ritual, can choose to apply any number of augments you know to the spell or ritual.
        For each augment you apply, you increase the spell or ritual's level by an amount equal to the augment's level.
        In exchange, the ability gains the effects of that augment.
        If an augment would increase the spell or ritual's level beyond the maximum level you can cast, you cannot apply the augment to that ability.

        \parhead{Augments and Cantrips}
        You can apply augments to \glossterm{cantrips}.
        Cantrips are considered to be 0th level spells for this purpose.

        \subsubsection{Augment Descriptions}\label{Augment Descriptions}

            \augment{1}{Cryptic} The spell's visual effects and magical aura changes to mimic a different spell of your choice.
            You may choose any combination of spells you know, along with any other augments, that result in a spell of the same level or lower as the spell you are casting.
            This affects inspection of the spell itself by any means, such as with the Spellcraft skill (see \pcref{Spellcraft}).
            However, it does not alter the mechanical effects of the spell in any way.
            If the spell's effects depend on visual components, the spell may fail to work if you alter the spell's visuals too much. 

            \augment{1}{Selective} You may freely exclude any areas from the spell's effect.
            However, all squares in the final area of the spell must be contiguous.
            You cannot create split a spell's area into multiple completely separate areas.
            \par This augment can be applied to any spell or ritual that affects an area.

            \augment{1}{Silent} You do not need to use \glossterm{verbal components} to cast the spell.
            \par This augment can be applied to any spell.

            \augment{1}{Stilled} You do not need to use \glossterm{somatic components} to cast the spell.
            \par This augment can be applied to any spell.

            \augment{2}{Accelerated} The ritual takes half the normal amount of time to perform.
            \par This augment can be applied to any ritual.

            \augment{2}{Empowered} The ability's \glossterm{power} increases by 2.
            This augment can be applied multiple times.
            Its effects stack.
            \par This augment can be applied to any spell or ritual.

            % TODO: Too weak at 2?
            \augment{2}{Extended} The ability's range increases by one step, to a maximum of \rngext.
            The steps are, in order: \rngclose, \rngmed, \rnglong, and \rngext.
            This augment can be applied multiple times.
            Each time, the ability's range increases by an additional step.
            \par This augment can be applied to any spell or ritual with a range that is one of the above ranges.

            % Too weak at 2, weird narrative at 1?
            \augment{2}{Giant} The ability can affect a target one size category larger.
            This augment can be applied multiple times.
            Its effects stack.
            \par This augment can be applied to any spell or ritual that that has a maximum size category of targets that it can affect.

            \augment{2}{Legion} If the spell would normally affect five or more specific targets, it instead affects five times that many targets.

            \augment{2}{Precise} You gain a \plus1 bonus to accuracy with the ability.
            This augment can be applied multiple times.
            Its effects stack.
            \par This augment can be applied to any spell or ritual that has an attack roll.

            \augment{2}{Quickened} You can cast the spell as a \glossterm{minor action}.
            In exchange, you cannot take any actions during the \glossterm{action phase} or \glossterm{delayed action phase} of the next round.
            \par This augment can be applied to any spell.

            \augment{2}{Reach} When you cast the spell, choose a location within \rngclose range.
            The spell takes effect as if you were in the chosen location.
            This affects your \glossterm{line of effect} for the spell, but not your \glossterm{line of sight} (since you still see from your normal location).
            % Wording?
            Since a spell's range is measured from your location, this augment can allow you to affect targets outside your normal range.
            For example, a cone that normally bursts out from you would instead originate from your chosen location, potentially avoiding an obstacle between you and your target.
            % Are there other edge cases that could cause problems?
            \par This augment can be applied to any spell that does not create an \glossterm{emanation}.

            % too weak at 3?
            \augment{3}{Phasing} When determining whether you have \glossterm{line of sight} and \glossterm{line of effect} to a particular location with the spell, you can ignore a single solid obstacle up to five feet thick.
            This can allow you to cast spells through solid walls, though it does not grant you the ability to see through the wall.
            \par This augment can be applied to any spell with a range.

            % too weak at 3?
            \augment{3}{Widened} The ability's area increases by one step, to a maximum of \areahuge.
            The steps are, in order: \areasmall, \areamed, \arealarge, and \areahuge.
            If you widen a line-shaped area, the line also increases in width by 5 feet.
            A line used to define a wall does not have a width.
            This augment can be applied multiple times.
            Each time, the ability's area increases by an additional step.
            \par This augment can be applied to any spell or ritual with an area that is one of the above areas.

            % \augment{4}{Dual} The spell targets an additional creature within range.
            % This augment can be applied multiple times.
            % Its effects stack.
            % \par This augment can be applied to any spell that has a range and affects a single target of the caster's choice.
            % It cannot be applied to spells that affect a single specific target, such as the caster.
            % In addition, it cannot be applied to \glossterm{Attune} spells.

            % too weak at 4?
            \augment{4}{Echoing} During the \glossterm{delayed action phase} of the next round, the spell's effect occurs again.
            All choices you made for the original casting of the spell are made identically for the repeat casting.
            It affects the same area, targets, and so on.

            If the spell is now invalid, such as if all of its targets are out of range, the additional casting has no effect.
            \par This augment can be applied to any spell that does not have the \glossterm{Attune} tag.

            % Are there any exploitable spells?
            \augment{4}{Innate} The spell loses the \glossterm{AP} tag, allowing you to cast it without spending an \glossterm{action point}.
            \par This augment can be applied to any spell with the \glossterm{AP} tag that requires a standard action to cast and which does not have the \glossterm{Sustain} tag.

    \subsection{Dismissal}
        As a \glossterm{minor action}, you can dismiss any spells or rituals you used that have lasting effects.
        This requires the same casting components (verbal and somatic) as casting the spell or performing the ritual normally.
        Spells and rituals can also be dismissed in other ways, such as after their effects have finished.
        When a spell or ability is dismissed, all of its lingering effects immediately end.
        % Should a spell be dismissed for all targets simultaneously, or per target?

    \subsection{Spell Power}

        The \glossterm{power} of your spells and rituals is the same as your \glossterm{power} with other magical abilities, which is equal to the higher of your level and Willpower.
        You can voluntarily reduce the \glossterm{power} of the spells you cast.
        % Keeping the ``twice spell level'' progression is fine despite that no longer matching spell acquisition
        The minimum power you can use to cast a spell is equal to twice the spell's level.

    \subsection{Resurrecting the Dead}\label{Resurrecting the Dead}
        Several rituals have the power to restore dead characters to life.

        When a living creature dies, its soul departs its body, travels through the Astral Plane, and goes to abide on the plane where the creature's deity resides.
        If the creature did not worship a deity, its soul departs to the plane corresponding to its alignment.
        Bringing a creature back from the dead means retrieving their soul and returning it to their body.

        \subparhead{Death and Old Age} While a creature is dead, it still tracks that time towards its maximum age.
        A creature's maximum age is largely determined by the strength of its soul, not the condition of its body.
        No magic can return a creature to life when it has passed its maximum age.

        \subparhead{Preventing Revivification} Enemies can take steps to make it more difficult for a character to be returned from the dead.
        Except for \spell{true resurrection}, every ritual to raise the dead requires a body, so keeping or destroying the body is an effective deterrent.
        The \spell{soul bind} ritual prevents any sort of revivification unless the soul is first released.

        \subparhead{Revivification against One's Will} A soul cannot be returned to life if it does not wish to be.
        A soul infallibly knows the name, alignment, and patron deity (if any) of the character attempting to revive it and may refuse to return on that basis.

\section{Spells}\label{Spells}
    % TODO: better description
    A \glossterm{spell} is a discrete magical effect with a name, a level, and an effect.
    Each \glossterm{mystic sphere} has a number of spells associated with it.
    An ability that gives you access to \glossterm{mystic spheres} will define how many spells you know from that sphere.
    It will also define the maximum level of spell you can cast.

    When you cast a spell, you can apply any number of \glossterm{augments} to the spell (see \pcref{Augments}).
    You cannot learn or cast spells whose spell level exceeds your maximum spell level.

    \subsection{Action Points}
        Most spells have the \glossterm{AP} tag, which means they cost an \glossterm{action point} to cast.
        However, some spells have the \glossterm{Attune} tag instead, and do not cost action points to cast.
        Instead, they require a creature to attune to the spell's effect, which costs an action point.
        For details, see \pcref{Attunement}.

    \subsection{Cantrips}\label{Cantrips}
        Some \glossterm{mystic spheres} have minor spells called cantrips.
        Anyone who has access to a mystic sphere knows all cantrips from that sphere.
        Cantrips are considered to be 0th level spells for the purpose of abilities that care about spell level, such as \glossterm{concentration} (see \pcref{Concentration}).

\section{Rituals}\label{Rituals}
    Each \glossterm{mystic sphere} has a number of \glossterm{rituals}.
    Some spellcasting characters can learn and perform rituals.
    Rituals are ceremonies that create magical effects.
    Like spells, each ritual has a name, a level, and an effect.
    Although rituals are similar to spells, abilities that affect spells do not affect rituals unless they say they do in their descriptions.
    % Oddly located
    An ability that gives you access to rituals will define the maximum level of ritual that you can perform.

    You don't memorize a ritual as you would a normal spell.
    Rituals are too complex for all but the most knowledgeable sages to commit to memory.
    To perform a ritual, you need to read from a book or a scroll containing it.
    You must have access to the \glossterm{mystic sphere} a ritual is from in order to perform the ritual.

    \subsection{Ritual Descriptions}
        % TODO: proper chapter references
        Rituals are described in the body of the \glossterm{mystic sphere} they are associated with, following the description of spells from that mystic sphere.

    \subsection{Ritual Books}
        A ritual book contains one or more rituals that you can use as frequently as you want, as long as you can spend the time and \glossterm{action points} to perform the ritual.
        Scribing a ritual in a ritual book costs an amount of precious inks.

    \subsection{Ritual Costs}\label{Ritual Costs}
        The costs to scribe rituals are described on \trefnp{Ritual Costs}.
        \begin{dtable}
            \lcaption{Ritual Costs}
            \begin{dtabularx}{\columnwidth}{X l l}
                \tb{Ritual Level} & \tb{Cost to Scribe} & \tb{Item Level} \tableheaderrule
                1st-Level & 125 gp     & 2nd  \\
                2nd-Level & 800 gp     & 5th  \\
                3rd-Level & 2,750 gp   & 8th  \\
                4th-Level & 10,000 gp  & 11th \\
                5th-Level & 37,000 gp  & 14th \\
                6th-Level & 125,000 gp & 17th \\
                7th-Level & 400,000 gp & 20th \\
            \end{dtabularx}
        \end{dtable}

    \subsection{Performing Rituals}
        To perform a ritual, you must have a ritual book containing the ritual and the material components required for the ritual.
        Unless otherwise specified, performing a ritual requires spending a single \glossterm{action point}.
        Some rituals require multiple action points to complete.
        Other creatures can supply action points to help you perform rituals; see Ritual Participants, below.

        % The action point cost for 24 hour rituals is equal to (ritual level ^ 2) * 2.
        % Should this be specified explicitly?

        Rituals with the \glossterm{Attune} (ritual) tag do not cost action points to perform, but require a ritual participant to attune to the ritual's effect, which costs an action point.
        For details, see \pcref{Attunement}.

        \subsubsection{Ritual Participants}
            Creatures can assist in the performance of rituals even if they are unable to perform rituals themselves.
            A creature that helps perform a ritual is called a ritual participant, and the creature performing the ritual is called the ritual leader.
            A ritual participant may spend an action point in place of or in addition to the action point spent by the creature performing the ritual.
            It may also \glossterm{attune} to the effect of the ritual in place of the creature performing the ritual.
            Only one creature may attune to the ritual's effect in this way.
            If multiple creatures are willing to spend action points or attune to effects, the ritual leader decides which creatures spend action points or attune to the ritual's effects.

            The steps required to participate in rituals can be complex.
            Ritual participants must be given specific instructions for the actions they must perform during a ritual by a creature who knows how to perform the ritual.
            This instruction generally takes half the time required to perform the ritual.
            A creature cannot participate in rituals unless it has an Intelligence of at least 0, can speak at least one language, and has the fine motor control required to perform the \glossterm{somatic components} of rituals.

            Normally, a ritual participant can only contribute one action point.
            If the participant has access to the same \glossterm{magic source} as the ritual, they can contribute any number of action points.

            \parhead{Changing Ritual Participation}
            Rituals are deeply complex magic, and they cannot be abandoned or paused partway through.
            If the number of ritual participants in a ritual decreases below its initial value, the ritual fails at the end of the next round if the number of participants is not restored.
            However, ritual participants can transfer their participation to other creatures without disrupting the ritual.

            In order to transfer ritual participation, the new creature must be able to participate in the ritual, and must immediately spend the same number of action points as the creature that it is taking over from.
            Similarly, the ritual leader can transfer their leadership to another creature.
            In addition to the requirements for transferring ritual participation, the new leader must know the ritual and be able to perform it themselves.

            Changing ritual participation and leadership is usually done when performing extraordinarily long or demanding rituals.

    \subsection{Magical Writings}
        To record a spell in written form, a character uses complex notation that describes the magical forces involved in the spell.
        The notation constitutes a universal language that spellcasters have discovered, not invented.
        Each writer uses this universal system regardless of their native language or culture.
        However, each character uses the system in their own way.
        Another person's magical writing remains incomprehensible to even the most powerful spellcaster until they take the time to study and decipher it.

        You can identify magical writing using the Spellcraft skill (see \pcref{Identify Magical Writing}).


\section{Mystic Sphere Lists}\label{Mystic Sphere Lists}

    \input{mystic_sphere_lists.tex}

\section{Mystic Sphere Descriptions}\label{Mystic Sphere Descriptions}

    \input{mystic_sphere_descriptions.tex}
