\tooldescription{Bag of Tricks}{Varies}{see text}{Varied Conjuration (Summoning)}{}{\tdash} This small sack appears normal and empty.
However, anyone reaching into the bag feels a small, fuzzy ball.
If the ball is removed and tossed up to 20 feet away, it turns into an animal.
The animal serves the character who drew it from the bag for 1 minute (or until slain or ordered back into the bag), at which point it disappears.
It can follow any of the commands described in the Creature Handling skill.
Each of the three kinds of a bag of tricks produces a different set of animals.
Use the following tables to determine what animals can be drawn out of each.

Animals produced are always random, and only one may exist at a time.
If a new animal is drawn from the bag, the previous animal is immediately unsummoned.

\begin{dtable}
    \lcaption{Bag of Tricks Types}
    \begin{dtabularx}{\columnwidth}{l X X}
        \tb{Bag Type} & \tb{Base Price} & \tb{Item Level} \\
        \hline
        Gray          & 200 gp          & 3rd             \\
        Rust          & 2,000 gp        & 8th             \\
        Tan           & 12,000 gp       & 12th            \\
    \end{dtabularx}
\end{dtable}

\begin{dtable}
    \lcaption{Bag of Tricks Animals}
    \begin{dtabularx}{\columnwidth}{l X X X}
        \tb{Roll} & \tb{Gray Bag} & \tb{Rust Bag} & \tb{Tan Bag} \\
        \hline
        1-20   & Dire rat   & Dire weasel   & Snake, giant constrictor \\
        21-40  & Cat        & Lion          & Dire lion \\
        41-60  & Badger     & Dire badger   & Dire wolverine \\
        61-80  & Owl        & Giant owl     & Griffon \\
        81-100 & Wolf       & Dire wolf    & Rhinoceros \\
    \end{dtabularx}
\end{dtable}

\mitemreq{Conjuration}{Summoning}{1 (gray), 3 (rust), or 5 (tan)}{2 (gray), 6 (rust), or 10 (tan)}{Craft (textiles) 6 (gray), 10 (rust), or 14 (tan)}



\tooldescription{Bead of Force}{600}{5th}{Moderate Evocation (Control) [Force]}{10th}{Standard (specific action)} This small black sphere appears to be a lusterless pearl.
By holding it in your hand and throwing it at an object or creature, you can surround the struck target with a \spell{resilient sphere}, as the spell.
The bead is treated as a thrown weapon with a range increment of 20 feet.
Your Reflex attack bonus for the \spell{resilient sphere} effect is equal to your level \add 5.

If you miss your target, the bead will still activate if it strikes a solid object, potentially wasting its effect.
After being activated, the bead is destroyed.

\mitemreqdesc{Evocation}{Control}{Force}{5}{10th}{Craft (jewelry) 14}



                \mitemref{Circlet of Perception} & Grants \plus2 Perception & \mitempricetable{Passive}{2} & Head & \mitempref{Circlet of Perception} \\
                \mitemref{Headband of Intellect} & Grants \plus2 Intelligence & \mitempricetable{Passive}{2} & Head & \mitempref{Headband of Intellect} \\
                \mitemref{Circlet of Perception, Greater} & Grants \plus4 Perception & \mitempricetable{Passive}{4} & Head & \mitempref{Circlet of Perception, Greater} \\
                \mitemref{Headband of Intellect, Greater} & Grants \plus4 Intelligence & \mitempricetable{Passive}{4} & Head & \mitempref{Headband of Intellect, Greater} \\
                \mitemref{Belt of Constitution} & Grants \plus2 Constitution & \mitempricetable{Passive}{2} & Torso & \mitempref{Belt of Constitution} \\
                \mitemref{Torc of Willpower} & Grants \plus2 Willpower & \mitempricetable{Passive}{2} & Torso & \mitempref{Torc of Willpower} \\
                \mitemref{Belt of Constitution, Greater} & Grants \plus4 Constitution & \mitempricetable{Passive}{4} & Torso & \mitempref{Belt of Constitution, Greater} \\
                \mitemref{Torc of Willpower, Greater} & Grants \plus4 Willpower & \mitempricetable{Passive}{4} & Torso & \mitempref{Torc of Willpower, Greater} \\
                \mitemref{Gloves of Dexterity} & Grants \plus2 Dexterity & \mitempricetable{Passive}{2} & Arms & \mitempref{Gloves of Dexterity} \\
                \mitemref{Gloves of Dexterity, Greater} & Grants \plus4 Dexterity & \mitempricetable{Passive}{4} & Arms & \mitempref{Gloves of Dexterity, Greater} \\
                \mitemref{Gauntlets of Strength} & Grants \plus2 Strength & \mitempricetable{Passive}{2} & Arms & \mitempref{Gauntlets of Strength} \\
                \mitemref{Gauntlets of Strength, Greater} & Grants \plus4 Strength & \mitempricetable{Passive}{4} & Arms & \mitempref{Gauntlets of Strength, Greater} \\


        \begin{magicitemdef}{Gloves of Dexterity}[Passive]{2}[Arms]{Transmutation [Augment]}{leather or textiles}
            \magicitemability{Passive} You gain a \plus2 bonus to Dexterity, up to a maximum Dexterity equal to the item's power.
        \end{magicitemdef}

        \begin{magicitemdef}{Gloves of Dexterity, Greater}[Passive]{4}[Arms]{Transmutation [Augment]}{leather or textiles}
            \magicitemability{Passive} You gain a \plus4 bonus to Dexterity, up to a maximum Dexterity equal to the item's power.
        \end{magicitemdef}


        \begin{magicitemdef}{Circlet of Perception}[Passive]{2}[Head]{Transmutation [Augment]}{metal}
            \magicitemability{Passive} You gain a \plus2 bonus to Perception, up to a maximum Perception equal to the item's power.
        \end{magicitemdef}

        \begin{magicitemdef}{Circlet of Perception, Greater}[Passive]{4}[Head]{Transmutation [Augment]}{metal}
            \magicitemability{Passive} You gain a \plus4 bonus to Perception, up to a maximum Perception equal to the item's power.
        \end{magicitemdef}

        \begin{magicitemdef}{Headband of Intellect}[Passive]{2}[Head]{Transmutation [Augment]}{metal}
            \magicitemability{Passive} You gain a \plus2 bonus to Intelligence, up to a maximum Intelligence equal to the item's power.
        \end{magicitemdef}

        \begin{magicitemdef}{Headband of Intellect, Greater}[Passive]{4}[Head]{Transmutation [Augment]}{metal}
            \magicitemability{Passive} You gain a \plus4 bonus to Intelligence, up to a maximum Intelligence equal to the item's power.
        \end{magicitemdef}

        \begin{magicitemdef}{Belt of Constitution}[Passive]{2}[Torso]{Transmutation [Augment]}{leather or textiles}
            \magicitemability{Passive} You gain a \plus2 bonus to Constitution, up to a maximum Constitution equal to the item's power.
        \end{magicitemdef}

        \begin{magicitemdef}{Belt of Constitution, Greater}[Passive]{4}[Torso]{Transmutation [Augment]}{leather or textiles}
            \magicitemability{Passive} You gain a \plus4 bonus to Constitution, up to a maximum Constitution equal to the item's power.
        \end{magicitemdef}

        \begin{magicitemdef}{Torc of Willpower}[Passive]{2}[Torso]{Transmutation [Augment]}{metal or jewelry}
            \magicitemability{Passive} You gain a \plus2 bonus to Willpower, up to a maximum Willpower equal to the item's power.
        \end{magicitemdef}

        \begin{magicitemdef}{Torc of Willpower, Greater}[Passive]{4}[Torso]{Transmutation [Augment]}{metal or jewelry}
            \magicitemability{Passive} You gain a \plus4 bonus to Willpower, up to a maximum Willpower equal to the item's power.
        \end{magicitemdef}

        \begin{magicitemdef}{Gauntlets of Strength}[Passive]{2}[Arms]{Transmutation [Augment]}{bone or metal}
            \magicitemability{Passive} You gain a \plus2 bonus to Strength, up to a maximum Strength equal to the item's power.
        \end{magicitemdef}

        \begin{magicitemdef}{Gauntlets of Strength, Greater}[Passive]{4}[Arms]{Transmutation [Augment]}{bone or metal}
            \magicitemability{Passive} You gain a \plus4 bonus to Strength, up to a maximum Strength equal to the item's power.
        \end{magicitemdef}

        \mitemref{Boots of Mobility} & Grants \plus4 to Jump, Sprint & 1,000 gp & 6th & Legs & \mitempref{Boots of Mobility} \\

        \begin{magicitemdef}{Boots of Mobility}[1000]{1}[Legs]{Transmutation [Enhancement]}{leather or textiles}
            \magicitemability{Passive} You gain a \plus4 bonus to Jump and Sprint checks.
        \end{magicitemdef}
