
\feat{Elemental Body}{Bloodline, Elemental}
\featpre Any three elemental bloodline feats.
\featben You have a 50\% chance to ignore critical hits on you, treating them as regular hits instead.

\feat{Elemental Force}{Bloodline, Elemental}
\featpre Elemental Heritage.
\featben Once per day per elemental bloodline feat you possess, you may unleash the power of your element on your foe as an attack action.
Air allows you to make a shove attack with a \plus4 bonus, and you use your Constitution in place of your Strength to attack.
Earth allows you to make a trip attack with a \plus4 bonus, and you use your Constitution in place of your Strength to attack.
Fire allows you to make a touch attack that causes your foe to be \ignited for 5 rounds if you hit.
Water allows you to make a touch attack that dehydrates your foe for 5 rounds if you hit, making it \fatigued.

\feat{Elemental Heritage}{Bloodline, Elemental}
\featben You have the essence of an elemental in your body.
When you take this feat, choose a type of elemental to be your elemental ancestor: air, earth, fire, or water.
Air and fire elemental heritage grants a \plus2 bonus to your Reflex defense, while earth and water elemental heritage grant a \plus2 bonus to your Fortitude defense.

\feat{Elemental Mastery}{Bloodline, Elemental}
\featpre Elemental Heritage.
\featben In circumstances that depend on your elemental ancestor, you gain a \plus1 bonus to physical attacks.
Air grants a bonus when you are airborne or fighting airborne creatures.
Earth grants a bonus when both you and your foe are standing on unworked earth or stone.
Fire grants a bonus when either you or your foe is ignited, or when you are making attacks that deal fire damage.
Water grants a bonus when both you and your foe are touching water.
If you have four or more elemental bloodline feats, this bonus increases to \plus2.

\feat{Elemental Movement}{Bloodline, Elemental}
\featpre Elemental Heritage.
\featben You gain a movement ability based on you choice of elemental ancestor.
Air halves the damage you take from falling and improves the maneuverability of any flight abilities you possess by one category.
Earth gives you a \plus2 bonus to maneuver defense against attacks that would force you to move, such as shove and trip attacks.
Fire increases your land speed by 5 feet.
Water gives you a swim speed equal to your base land speed.

\feat{Alacritous Surge}{Surge}
\featpres 16th level, Dex 7.
\featben You may gain the benefits of the \spell{haste}
spell for up to 5 rounds each day.
Activating this ability is a swift action, and it lasts until you will it to stop (a free action) or until the duration expires.

\feat{Deathless Surge}{Surge}
\featpres 12th level, Wil 5.
\featben You may gain the benefits of the \spell{death ward}
spell for up to 5 rounds each day.
Activating this ability is a swift action, and it lasts until you will it to stop (a free action) or until the duration expires.

\feat{Freeing Surge}{Surge}
\featpres 16th level, Dexterity 7.
\featben You may gain the benefits of the \spell{freedom}
spell for up to 5 rounds each day.
Activating this ability is a swift action, and it lasts until you will it to stop (a free action) or until the duration expires.

\feat{Healing Surge}{Surge}
\featpres 8th level, Constitution 3.
\featben Three times per day, you may heal 4d6 points of damage as a swift action.
This healing increases by 1d6 at 12th level and every 4 levels thereafter.
After using this ability, you cannot use it again for 5 rounds.

\feat{Prescient Surge}{Awareness, Surge}
\featpres 12th level, Wisdom 5.
\featben Three times per day, you may roll twice on any single physical attack or opposed check.
You must decide to use this ability before making the roll.
Activating this ability is an immediate action.

\feat{Shattering Surge}{Power, Surge}
\featpres 8th level, Strength 3.
\featben Three times per day, when you successfully attempt to disarm an object, you may cause it to be affected as if by the targeted version of the \spell{shatter} spell with a spellpower equal to your level.
Activating this ability is an immediate action.

\feat{Surge of Speed}{Surge}
\featpres 8th level, Dexterity 3.
\featben You may gain the benefits of the \spell{expeditious retreat}
spell for up to 5 rounds each day.
Activating this ability is a swift action, and it lasts until you will it to stop (a free action) or until the duration expires.


\feat{Bane of the Weak Minded}{Bane, Magic}
\featpre Ability to cast an Enchantment spell of 2nd level or higher.
\featben You gain a \plus1 bonus to attacks and defenses against \disoriented creatures.
\featspecial \featbanenotes

\feat{Dragonbane}{Bane}
\featpre Fought a dragon.
\featben You gain a \plus2 bonus to attacks and defenses against dragons.
\featspecial \featbanenotes

\feat{Fearless Rage}{Class}
\featpre Rage ability.
\featben You are immune to fear effects while raging.

\feat{Extra Channeling}{Class}
\featpre Ability to channel energy.
\featben You can channel energy three more times per day.

\feat{Extra Invocation}{Class}
\featpre Ability to use an arcane invocation.
\featben You learn a new arcane invocation.
You cannot learn invocations from your prohibited schools, if any, with this feat.

\feat{Extra Rage}{Class}
\featpre Ability to rage.
\featben You can rage one more time per day.

\feat{Extra Smiting}{Class}
\featpre Smite ability.
\featben You can smite three more times per day.
If you have more than one smite ability, you choose which ability this applies to.
\parhead{Special}
You can take more this once if you have more than one smite ability.
Its effects do not stack.
Each time you take the feat, you choose a different smite ability you have and apply the effects to that ability.

\feat{Extra Wild Aspect}{Class}
\featpre Wild aspect ability.
\featben You gain three extra uses of your wild aspect ability.

\feat{Mighty Summons}{Magic}
\featpre Ability to cast a Conjuration (Summoning) spell of at least 2nd level.
\featben Each creature you conjure with a summoning spell gains a \plus2 bonus to Strength and Constitution for the duration of the spell that summoned it.

\feat{Residual Awareness}{Magic}
\featpre Ability to cast a Divination (Awareness) spell.
\featben Whenever you cast a Divination (Awareness) spell, you gain an enhancement bonus to Awareness checks equal to the level of the spell for 5 rounds.

\feat{Residual Beguilement}{Magic}
\featpre Ability to cast an Enchantment (Beguilement) spell of at least 3rd level.
\featben Whenever you cast a Enchantment (Beguilement) spell, you gain an enhancement bonus to Persuasion checks equal to the level of the spell for 5 minutes.

\feat{Residual Knowledge}{Magic}
\featpre Ability to cast a Divination (Knowledge) spell.
\featben Whenever you cast a Divination (Knowledge) spell, you gain an enhancement bonus to Knowledge checks equal to the level of the spell for 5 rounds.

\feat{Residual Shield}{Magic}
\featpre Ability to cast an Abjuration (Shielding) spell.
\featben Whenever you cast an Abjuration (Shielding) spell with a duration of \durshort or longer on another creature, you gain the benefits on yourself for 1 round.

\feat{Dodge}{Combat, Defense, Reaction}
\featpre Dexterity 5, combat prowess 4.
\featben You gain a \plus1 bonus to physical defenses (Armor, Manuever, Reflex).
This bonus increases to \plus2 at Dexterity 10, and to \plus3 at Dexterity 20.

\feat{Focused Mind}{Racial}
\featpre Elf, Int 3.
\featben You gain a \plus2 bonus to Mental defense.
In addition, you can use your Intelligence instead of your Constitution on Concentration checks, such as when you cast spells.

\feat{Patient Spell}{Metamagic}
\featpre 2nd level spell known.
\featben A patient spell requires a full-round action to cast.
In exchange, you gain a \plus2 bonus to spellpower with the spell.
A patient spell uses up a spell slot one level higher than the spell's actual level.

\feat{Destructive}{General}
\featpre Strength 3.
\featben You gain a \plus5 bonus to Strength checks made to break, bend, or crush objects.


\feat{Creative Magic}{Magic, Style}
\featpre 4th level spell known.
\featben You gain a \plus2 bonus to spellpower with all spells that you have not cast within the past 5 rounds.
This bonus increases by \plus1 when you gain access to 6th and 8th level spells.

\feat{Devastating Magic}{Magic, Style}
\featpre 8th level spells.
\featben Your spells always inflict their critical success effects if they succeed by any amount, rather than if they succeed by 10.
This has no effect on creatures whose level exceeds your spellpower.

\feat{Flameborne Magic}{Magic, Style}
\featpre 4th level or higher \glossterm{Fire} spell known.
\featben As long as you are within \rngmed range of an active fire of at least Medium size, you gain a \plus2 bonus to spellpower with Fire spells.
If the source is at least Large size, this bonus applies to all spells you cast.
This bonus increases by \plus1 when you gain access to 6th and 8th level spells.

\feat{Peaceful Magic}{Magic, Style}
\featpre 4th level or higher \glossterm{Abjuration} spell known.
\featben If you have not taken damage within the past 5 rounds, you gain a \plus2 bonus to spellpower.
This bonus increases by \plus1 when you gain access to 6th and 8th level spells.

\feat{Blood Magic}{Magic, Style}
\featpre 4th level Vivimancy spell known.
\featben At the end of each round, if you took at least as much lethal damage as your spellpower during that round, you gain a \plus2 bonus to spellpower during the next round.
In addition, as a swift action, you can deal yourself life damage equal to your spellpower.
This bonus increases by \plus1 when you gain access to 6th and 8th level spells.

\feat{Unliving Conduit}{Magic, Style}
\featpre 4th level or higher \glossterm{Vivimancy} spell known.
\featben As long as you are within \rngclose range of an undead creature (other than yourself), you gain a \plus2 bonus to spellpower with Vivimancy spells.
If the undead creature is under your control, this bonus applies to all spells you cast.
This bonus increases by \plus1 when you gain access to 6th and 8th level spells.

\feat{Brutal Magic}{Magic, Style}
\featpres 4th level spells.
\featben At the end of each round, if you dealt damage with a \glossterm{strike} that round, you gain a \plus2 bonus to spellpower during the next round.
This bonus increases by \plus1 when you gain access to 6th and 8th level spells.


\feat{Supreme Combat Insight}{Awareness, Combat}
\featpres Perception 15, combat prowess 16.
\featben You automatically know the exact status of all creatures you can see within \rngmed range of you.
This includes the following properties:
\begin{itemize}
    \item Current hit points
    \item Any special types of damage the creature has taken (such as vital damage)
    \item The exact nature and remaining duration of all physical conditions and extraordinary abilities
    \item The remaining duration, but not exact nature, of all spells and supernatural abilities
\end{itemize}



    At 7th level, if you have Perception 9, you can attempt to disable a foe's wings.
    As a standard action, you can make a single physical attack against a creature's Armor and Fortitude defenses.
    Success against Armor defense means the target takes damage from your weapon.
    This damage is doubled for each strike you can make beyond the first.
    If you also beat the target's Fortitude defense, it loses the ability to fly for 2 rounds.
    You must be able to hit the target's wings, which is usually only possible with melee attacks if it is no more than one size category larger than you.
    This only affects creatures who use wings or other physical means to fly, and has no effect on creatures with magical flight.

    At 11th level, if you have Perception 12, you can deliver a killing blow.
    As a standard action, you can make a single physical attack against a creature.
    Success against Armor defense means the target takes damage from your weapon.
    This damage is doubled for each strike you can make beyond the first.
    Any damage in excess of the target's remaining hit points is dealt as \glossterm{vital damage}.
