\chapter{Optional Rules}

\section{Attributes}

    \subsection{Other Methods of Attribute Generation}
        Point buy offers the fairest and most customizable system for determining attribute scores, ensuring that players can be almost any character they want to be. However, some groups may wish to determine attribute scores differently. Other options are provided below.

        \subsubsection{Simple Random Point Buy}
            With this method, you have only a small degree of control over your attribute scores, but all characters generated in this way are equally powerful.
            As with the point buy method, all your attribute scores start at 0, and you get 15 points to distribute among your attribute scores.
            However, you do not have full control over how to distribute those points.

            For each attribute, starting with the attributes you care about most, roll 1d8.
            You spend that many points on that attribute, ignoring any extra points that can't be spent
            For example, if you roll a 4, you spend 3 points on the attribute, causing you to start with a 2.
            If you do not have enough points remaining to spend the amount indicated by the die roll, spend as many as you can and move on to the next attribute.

            If you have points remaining after rolling all of your attribute scores, you may distribute the points freely among your abilities, using the normal point buy rules.
            You cannot increase the starting value of any individual attribute by more than 1 during this stage.
            If any of your attributes start as a 0, you may choose to lower them to gain the normal benefits from having low attributes (see \pcref{Attribute Penalties}).

            To further limit your character creation options, you may choose to randomize the order in which you roll your attributes instead of rolling them in an order of your choice.

        \subsubsection{Smoothed Random Point Buy}
            This method functions like the Simple Random Point Buy method, except that the resulting attribute values have a smoother distribution, and you can randomly end up with attribute penalties.

            For each attribute, starting with the attributes you care about most, roll 4d6.
            Then, remove any one of the rolls after seeing the results.
            Sum the results of the remaining three dice and spend the appropriate number of attribute points as indicated in \trefnp{Smoothed Random Point Buy Results}.
            If you do not have enough points remaining to spend the amount indicated by the die roll, spend as many as you can and move on to the next ability.

            If you have points remaining after rolling all of your attribute scores, you may distribute the points freely among your abilities, using the normal point buy rules.
            You cannot increase the starting value of any individual attribute by more than 1 during this stage.

            To further limit your character creation options, you may choose to randomize the order in which you roll your attributes instead of rolling them in an order of your choice.

            \begin{dtable}
                \lcaption{Smoothed Random Point Buy Results}
                \begin{dtabularx}{\columnwidth}{X X X}
                    \tb{Roll} & \tb{Base Attribute} & \tb{Point Cost} \tableheaderrule
                    3-4       & \minus2             & 0\fn{1} \\
                    5-6       & \minus1             & 0\fn{2} \\
                    7-8       & 0                   & 0       \\
                    9-10      & 1                   & 1       \\
                    11-12     & 2                   & 3       \\
                    13-15     & 3                   & 5       \\
                    16-18     & 4                   & 8       \\
                \end{dtabularx}
                1 You gain one \glossterm{insight point}. \\
                2 You gain an additional \glossterm{trained skill}. \\
            \end{dtable}

        \subsubsection{Classic Hardcore}

            This method is completely random and can generate very overpowered or underpowered characters.
            It represents the unfairness of the world, where some people are just better or worse than others.
            For each attribute, roll 2d6, take the average (rounded down), and subtract 2.
            If you roll a 1 on both dice, treat the average as a 0.
            The result is your base value for that attribute.

\section{Epic Fate}
    After 21st level, characters no longer gain levels normally.
    However, they can still increase their personal power as they make progress towards their ultimate fate.

    When you reach 21st level, you may choose an epic fate that you qualify for, or you may delay choosing until you meet the prerequisites for your desired fate.
    You do not start with any ranks in you chosen epic fate.
    Each epic fate specifies ways that you can make progress towards that epic fate.
    Whenever you make dramatic progress towards your epic fate, your rank in that epic fate may increase, at the discretion of the Game Master.

    None of the epic fate abilities have a tag to indicate that they are \glossterm{magical} abilities.
    Many of them are not fundamentally \glossterm{mundane} in nature, but they are beyond normal magic, and effects like an \spell{antimagic field} cannot interact with or suppress them.

    \subsection{Artificial Immortality}
        You have sought out strange magical power in search of a way to artificially prolong your life.
        As your power grows, you become increasingly able to resist death and return from it.
        Eventually, you will transcend death entirely.

        \parhead{Prerequisites} You must perform a series of rituals to prepare yourself for immortality, at least one of which must be rank 7 or higher. There are many kinds of immortality that you can pursue with this epic fate, and the exact nature of the rituals will change depending on the type of immortality you pursue. For example, you can have a phylactery regenerate a new body for you like a lich, you can create clones of yourself that you inhabit if your first body dies, or you can modify your body to regenerate after death from mortal wounds. This immortality may change your base species, such as if you become a lich or move your body into a flesh golem. If it does, you retain all benefits and modifiers from your original species other than size and gain the effects of the new species in addition.

        \parhead{Progression} You must discover powerful new magic rituals that support your particular form of immortality. This generally requires exploring sites of ancient magic, gaining favor with powerful creatures who have relevant knowledge or abilities, and independent experimentation based on your findings.

        \subsubsection{Artifical Immortality Ranks}

            \parhead{Rank 1 -- Life After Death} If you die from any cause other than old age, you resurrect according to nature of your chosen immortality. Your specific form of immortality determines where you return, such as at the site of your death, at the current location of your corpse, or in an entirely separate location. The timing of your resurrection may also differ based on your immortality, but you cannot complete your resurrection sooner than one day after the time of your death. After you resurrect in this way, this ability does not function for one week, allowing you to be killed normally.

            \parhead{Rank 2 -- Death Familiarity} You become so familiar with the trauma of injury and death that your body adapts to it.
            You gain a \plus2 bonus to vital rolls.
            In addition, you cannot gain more than two vital wounds from a single damage roll, regardless of the amount of damage you take.

            \parhead{Rank 3 -- Artificial Life} Whenever you resurrect with your \textit{life after death} ability, your new body gains a \plus1 bonus to Strength, Dexterity, and Constitution. This bonus does not stack if you resurrect multiple times. In addition, that resurrection functions even if the cause of your death was old age, and you can control the physical age of your new body.

            \parhead{Rank 4 -- Deathcaller} You are deeply familiar with death, and know how to most effectively inflict it on others.
            Whenever you inflict a \glossterm{vital wound} on a living creature, you may kill that creature outright.

            \parhead{Rank 5 -- True Immortality} You become fully immortal. There is no time limit after the resurrection from your \textit{life after death} ability where you become vulnerable to a true death. In addition, the resurrection can complete as quickly as the end of the next round after your death. If a physical component limits your immortality, such as a phylactery, it can no longer be damaged or destroyed without the direct intervention of a rank 5 Slayer.

    \subsection{Deity}
        People have begun to worship you, putting you on the path to become a deity.
        As your followers grow, you become capable of ever greater miraculous acts, and you can grant your followers some of your power.
        Eventually, you ascend into the pantheon of gods.

        \parhead{Prerequisites} You must have at least a hundred sentient worshippers to choose this epic fate.
        In addition, you must not have any cleric archetypes.

        \parhead{Progression} To progress towards this epic fate, you must gain a significant number of additional worshippers.
        In general, you must at least double your worshippers to progress towards each new rank of this fate, though this can vary widely.
        Having worshippers among many different places is more valuable than converting an isolated group to worship you, though both are helpful.

        \subsubsection{Deity Ranks}
            \parhead{Rank 1 -- Domain Influence} Choose a cleric domain.
            You gain all abilities from that domain except for its mastery ability.
            In addition, your worshippers become eligible to gain cleric archetypes, though they cannot exceed a maximum rank in those archetypes of twice your rank in this epic fate (to a maximum of 8).
            This does not grant additional archetypes to worshippers who have already chosen their three archetypes, and is usually only relevant to NPC worshippers.

            \parhead{Rank 2 -- Prayers} You hear all prayers directed to you.
            Once per week, you can teleport yourself and up to ten \glossterm{allies} any distance within the same plane as a \glossterm{standard action}.
            Your destination must either be a worshipper actively praying to you or a holy place dedicated to you.
            In addition, choose a second cleric domain.
            You gain all abilities from that domain except for its mastery ability.

            \parhead{Rank 3 -- Domain Mastery} Choose a third cleric domain.
            You gain all abilities from that domain.
            In addition, you gain the mastery ability from the domains you chose with your \textit{domain influence} and \textit{prayers} abilities.

            \parhead{Rank 4 -- Demigod} You become a demigod.
            You no longer age normally, and you cannot die from old age.
            You become a \glossterm{planeforged} native to an Aligned Plane matching your alignment.
            For details about the aligned planes, see the Tome of Guidance.
            While you are on that plane, you can teleport to any plane with your \textit{prayers} ability from this epic fate.
            In addition, you can use that teleportation ability once per hour instead of once per week.

            \parhead{Rank 5 -- Deification} You become a deity.
            You are transported to an Aligned Plane matching your alignment, and you gain divine dominion over an amount of territory in that plane.
            While you are in your territory, you can can freely reshape your territory with a thought to match your desires, and you are immune to all damage and \glossterm{conditions}.

            Regardless of which plane you are on, you can teleport to anywhere within your home plane as a \glossterm{standard action}.
            In addition, there is no limit on the number of times you can teleport with your \textit{prayers} ability from this epic fate.

    \subsection{Hero of Legend}
        You are widely known as a hero, rescuing those in need.
        As your deeds of heroism spread, you gain abilities to help you protect others.
        Although you will eventually die, your legend will live on, inspiring others to save people as you did.

        \parhead{Prerequisites} You must be publicly known to be involved with saving at least one major country or similarly large group of people from some sort of disaster to choose this epic fate.
        In addition, you must have a Willpower of at least 1.

        \parhead{Progression} To progress towards this epic fate, you must publicly contribute to saving large numbers of people from death or other major disasters in a way that builds your reputation.

        \subsubsection{Hero of Legend Ranks}
            \parhead{Rank 1 -- Heroic Intervention} As a \glossterm{free action}, you may choose an \glossterm{ally} adjacent to you.
            Whenever that creature would be the target of an attack that phase, you are targeted by that attack instead.
            If the attack would have targeted both you and that ally, the attack only targets you once, not twice.
            This ability has the \abilitytag{Swift} tag.

            \parhead{Rank 2 -- Unstoppable Hero} You gain a \plus4 bonus to all defenses, and you gain a \plus50 bonus to your maximum \glossterm{hit points}.
            In addition, you gain a \plus20 foot bonus to your speed with all of your \glossterm{movement modes}.

            \parhead{Rank 3 -- Sheltering Aura} Your \textit{heroic intervention} ability from this epic fate affects any number of \glossterm{allies} within a \areamed radius \glossterm{emanation} from you.
            Whenever an affected ally is attacked, you teleport into an empty space next to that creature, or into its space if no empty space is available.
            If multiple allies are attacked simultaneously, you can choose where you end up at the end of the series of teleportations.

            \parhead{Rank 4 -- Inspiring Hero} The area of your \textit{heroic intervention} ability increases to a \areagarg radius \glossterm{emanation}.
            In addition, each creature with a mind affected by that ability is so inspired by your example that it gains a \plus1 bonus to its Willpower permanently.
            As normal, this bonus does not stack with itself.

            \parhead{Rank 5 -- Answer the Call} You gain an intuitive sense for when people need your aid.
            Whenever someone on the same plane as you is in danger, you are aware of the existence of that danger.
            You can sense the general category of danger (fire, combat, drowning, etc.) and a very approximate direction and distance.
            This generally allows you to sense if a large number of people are in danger from the same thing.
            As a \glossterm{standard action}, you can teleport any distance within that plane to reach a person in danger.

    \subsection{Slayer}
        You are widely known as a killer of legendary skill.
        As your body count increases, you gain abilities to help you track down and kill increasingly powerful foes.
        Eventually, your powers threaten the gods themselves, allowing you a unique ability to transcend death.

        \parhead{Prerequisites} You must be publicly known to be involved with slaying at least one \glossterm{elite} creature with a level of at least 21.

        \parhead{Progression} To progress towards this epic fate, you must publicly contribute to slaying increasingly dangerous and fearsome foes in a way that builds your reputation.

        \subsubsection{Slayer Ranks}
            \parhead{Rank 1 -- Lethality} Whenever you get a \glossterm{critical hit}, you double all of your damage bonuses along with your damage dice.
            This bonus stacks with other abilities with the same effect, such as the \ability{sneak attack} ability.

            \parhead{Rank 2 -- Precision Killer} You gain a \plus4 bonus to \glossterm{accuracy}.
            In addition, you can inflict \glossterm{critical hits} on creatures that would otherwise be immune to critical hits from you due to their size or body structure.

            \parhead{Rank 3 -- Mark of the Slayer} As a \glossterm{standard action}, you can choose to mark any creature you can unambiguously identify.
            This includes any creature you can see, as well as any creature you know the name of and can differentiate from other similar creatures.
            You can only mark one creature at a time, and applying a new mark replaces any previous mark.
            You cannot use this ability to replace a mark that is less than a week old if the recipient of the previous mark still lives.

            This mark is visible on the creature's body with a design that is recognizably yours.
            It appears on top of any clothing or other attempt to conceal it, even if the creature is invisible.
            Anyone can recognize the significance of the mark with a \glossterm{difficulty value} 15 Knowledge (arcana or local) check, and creatures that understand the significance of the mark may refuse to give your target aid of any kind to avoid risking your wrath.

            You know the exact distance and direction to any creature you have marked with this ability that is on the same plane as you.
            As a \glossterm{standard action}, you can create a \glossterm{scrying sensor} adjacent to them that you can see and hear through.
            The sensor lasts as long as you \glossterm{sustain} it as a \glossterm{free action}.
            It moves to stay adjacent to the target, regardless of its speed.

            \parhead{Rank 4 -- Slayer's Journey} As a \glossterm{standard action}, you can \glossterm{teleport} yourself and up to ten \glossterm{allies} any distance within the same plane to the location of a creature affected by your \textit{mark of the slayer} ability from this epic fate.
            You cannot precisely choose the destination of this ability, and it does not leave you immediately adjacent to the marked creature.
            Generally, it leaves you just outside any sort of fortress or defenses the marked creature has constructed.
            After you use this ability, you cannot use it to travel to the same creature for a day.
            This does not limit your ability to travel to a different creature if you mark a different creature.

            \parhead{Rank 5 -- Godslayer}
            Your attacks ignore most forms of general immunity.
            This does not help you ignore specific immunities, such as fire elemental's immunity to fire damage.
            However, you can destroy artifacts and even inflict damage and conditions on deities in their divine dominion.
            As a result, even deities fear to interfere with you directly.
            If you ever die, you can generally threaten or fight your way past any planar guardians to leave your afterlife whenever you want.
            After you do this once, you become a \glossterm{planeforged} native to your afterlife plane, since your new body is formed from the raw material of that plane.

\section{Uncommon Species}\label{Uncommon Species}

    \subsection{Animal Hybrid}
        Animal hybrids are humanoid creatures that are a combination of humans and animals.
        The abilities of an animal hybrid depend on the type of animal it is based on.

        \parhead{Size} Medium.
        \parhead{Attributes} The attributes of an awakened animal depend on its size.
        \parhead{Special Abilities} As the original animal.
        \parhead{Automatic Languages} Common and any one \glossterm{common language} (see \tref{Common Languages}).

        \subsubsection{Sample Animal Hybrids}

            \parhead{Hybrid Bee}

            \subparhead{Special Abilities}
            \parhead{Attribute} \plus1 Dexterity, \minus1 Constitution.
            \begin{itemize}
                \itemhead{Low-light Vision} A hybrid bee has \trait{low-light vision}, allowing it to see clearly in \glossterm{shadowy illumination} (see \pcref{Low-light Vision}).
                \itemhead{Stinger} A hybrid bee has a stinger natural weapon (see \pcref{Natural Weapons}).
                    Whenever it causes a creature to lose \glossterm{hit points} with that natural weapon, the struck creature is poisoned by giant wasp venom (see \pcref{Poison}).
                    Its stage 1 effect makes the target \slowed while the poison lasts.
                    Its stage 3 effect makes the target \immobilized while the poison lasts.
                \itemhead{Winged Agility} A hybrid bee has wings that are not strong enough to help it fly.
                    However, the wings still help it stabilize its movements.
                    It gains a \plus3 bonus to the Balance and Jump skills.
            \end{itemize}

            \parhead{Hybrid Shark}

            \subparhead{Special Abilities}
            \begin{itemize}
                \itemhead{Bloodscent} A hybrid shark has the scent ability (see \pcref{Scent}).
                    In addition, it gains a \plus10 bonus to Awareness checks to detect blood.
                \itemhead{Bite} A hybrid shark's mouth is elongated, which it can use as a bite attack (see \pcref{Natural Weapons}).
                    A hybrid shark's bite deals 1d6 damage.
                \itemhead{Gills} You can breathe water as easily as a human breathes air, preventing you from drowning or suffocating underwater.
                \itemhead{Swim Speed} A hybrid shark has a swim speed equal to the base speed for its size.
            \end{itemize}

            \parhead{Hybrid Wolf}

            \subparhead{Special Abilities}
            \begin{itemize}
                \itemhead{Scent} A hybrid wolf has the scent ability (see \pcref{Scent}).
                \itemhead{Bite} A hybrid wolf's mouth is elongated, which it can use as a bite attack (see \pcref{Natural Weapons}).
                    A hybrid wolf's bite deals 1d6 damage.
                \itemhead{Low-light Vision} A hybrid wolf has \trait{low-light vision}, allowing it to see clearly in \glossterm{shadowy illumination} (see \pcref{Low-light Vision}).
            \end{itemize}

    \subsection{Awakened Animal}

        Awakened animals are animals that have been granted sentience by the \spell{awaken} ritual.
        The abilities of an awakened animal depend on the type of animal it is.

        \parhead{Size} Small or Medium, as original animal.
        \parhead{Attributes} The attributes of an awakened animal depend on its size.
        \subparhead{Medium} No change.
        \subparhead{Small} \minus2 Strength, \plus1 Dexterity.
        \parhead{Special Abilities} As the original animal.
        \parhead{Automatic Languages} Common.

        \subsubsection{Sample Awakened Animals}

            \parhead{Cat}

            \subparhead{Size} Small. This gives a cat a 20 foot \glossterm{base speed} and a \plus5 bonus to the Stealth skill, among other effects (see \pcref{Size Categories}).
            \subparhead{Attributes} \minus2 Strength, \plus1 Dexterity
            \subparhead{Special Abilities}
            \begin{itemize}
                \itemhead{Claws} A cat's paws end in claws, which it can use to attack (see \pcref{Natural Weapons}). A cat's claws do 1d4 damage.
                \itemhead{Low-light Vision} A cat has \trait{low-light vision}, allowing it to see clearly in \glossterm{shadowy illumination} (see \pcref{Low-light Vision}).
                \itemhead{Scent} A cat has the scent ability (see \pcref{Scent}).
            \end{itemize}

    \subsection{Changeling}

        \parhead{Size} Medium.
        \parhead{Attributes} No change.
        \parhead{Special Abilities}
        \begin{itemize}
            \itemhead{Alter Shape} A changeling can alter its physical form in minor ways. As a standard action, a changeling can make a Disguise check with a \plus10 bonus to alter its body. This ability does not alter the changeling's equipment, which may give away its identity unless disguised normally.

            This is a \glossterm{magical} ability.
        \end{itemize}
        \parhead{Bonus Languages} Any.
        \parhead{Automatic Languages} Common, any two \glossterm{common languages}.

    \subsection{Dragon}
        Ancient dragons are magical creatures of immense power and wisdom, and are far more powerful than any ordinary character of the same level.
        However, young dragons can be played as characters, though their unique abilities do pose unique challenges.

        \parhead{Creature Type} Unlike most other playable species, dragons are considered magical beasts instead of humanoids.
        \parhead{Size} Small. This gives a dragon a 20 foot \glossterm{base speed} and a \plus5 bonus to the Stealth skill, among other effects (see \pcref{Size Categories}).
        \parhead{Attributes} \minus2 Strength, \plus1 Dexterity.
        \parhead{Special Abilities}
        \begin{itemize}
            \itemhead{Dragon Archetype} Dragons only gain two class archetypes instead of three.
                Instead, they treat the Dragon archetype as one of their archetypes, and they gain ranks in it just like they gain ranks in class archetypes.
            \itemhead{Draconic Senses} Dragons have \trait{darkvision} with a 60 foot range, allowing them to see in complete darkness (see \pcref{Darkvision}).
                In addition, dragons gain \trait{low-light vision}, allowing them to see clearly in \glossterm{shadowy illumination} (see \pcref{Low-light Vision}).
            \itemhead{Draconic Weapons} Dragons have a bite natural weapon and two claw natural weapons.
                For details, see \pcref{Natural Weapons}.
            \itemhead{Draconic Wings} Dragons have scaly wings that sprout from their backs.
                These wings grant them a glide speed equal to the \glossterm{base speed} for their size (see \pcref{Gliding}).
                The wings themselves are \glossterm{mundane}, but the ability to fly and glide with them is \glossterm{magical}.
            \itemhead{Dragon Type} Each dragon has a single type from among the dragon types on \trefnp{Dragon Types}.
                They are immune to the damage type dealt by their type's breath weapon.
            \itemhead{Limited Equipment} A dragon's claws are not able to effectively wield shields or manufactured weapons.
                They can wear armor, but it is treated as barding instead of normal armor, increasing its cost.
                In general, dragon-fitted barding is rare or nonexistent even in large cities, so a dragon's armor must usually be created specifically for them.
        \end{itemize}
        \parhead{Automatic Languages} Common, Draconic, any one \glossterm{common language}.

        \subsubsection{Dragon Archetype}

            \cf{Dgn}[1]{Draconic Breath} You can use the \textit{breath weapon} ability as a \glossterm{standard action}.
            % +1d to compensate for cooldown
            \begin{activeability}{Breath Weapon}
                \rankline
                Make an attack vs. Reflex against everything in the area defined by your dragon type (see \trefnp{Dragon Types}).
                After you use this ability, you \glossterm{briefly} cannot use it again.
                \hit Each target takes damage equal to 1d8 plus half your \glossterm{power}.
                The damage type is defined by your dragon type.

                \rankline
                \rank{2} The damage increases to 1d10.
                    In addition, the area affected by your breath weapon increases.
                    A line breath weapon becomes a \arealarge, 5 ft.\ wide line.
                    A cone breath weapon becomes a \areamed cone.
                \rank{3} The damage increases to 2d6.
                \rank{4} The damage increases to 2d8.
                    In addition, the area affected by your breath weapon increases.
                    A line breath weapon becomes a \areahuge, 10 ft.\ wide line.
                    A cone breath weapon becomes a \arealarge cone.
                \rank{5} The damage increases to 2d10.
                \rank{6} The damage increases to 4d6.
                    In addition, the area affected by your breath weapon increases.
                    A line breath weapon becomes a \areagarg, 15 ft.\ wide line.
                    A cone breath weapon becomes a \areahuge cone.
                \rank{7} The damage increases to 4d8.
            \end{activeability}

            \cf{Dgn}[2]{Draconic Body} You gain a \plus1 bonus to Armor defense.
            In addition, you gain a \plus1d damage bonus with all \glossterm{natural weapons}.

            \cf{Dgn}[3]{Draconic Flight}[Magical] Your wings grow larger, granting you a limited ability to fly.
            You gain a \glossterm{fly speed} equal to the \glossterm{base speed} for your size with a maximum height of 15 feet (see \pcref{Flying}).
            As a \glossterm{free action}, you can increase your \glossterm{fatigue level} by one to ignore this height limit until the end of the round.

            \cf{Dgn}[4]{Draconic Bulk} Your size category increases to Medium.
            This increases the \glossterm{base speed} for your size.
            You reduce your Dexterity by 1 and increase your Strength by 2.
            In addition, you gain a \plus2 bonus to your \glossterm{power} with all abilities.

            \cf{Dgn}[5]{Greater Draconic Body} The defense bonus from your \textit{draconic body} ability increases to \plus2.
            In addition, the damage bonus increases to \plus2d.

            \cf{Dgn}[6]{Greater Draconic Flight} The maximum height from your \textit{draconic flight} ability increases to 60 feet.
            In addition, you gain a \plus10 foot bonus to your fly speed with that ability.

            \cf{Dgn}[7]{Greater Draconic Bulk} Your size category increases to Large.
            The speed bonus from your \textit{draconic bulk} ability increases to \plus20 feet, the attribute modifiers to Dexterity and Strength increase to \minus2 and \plus3 respectively, and the power bonus increases to \plus6.
            You gain a slam natural weapon, except that it also has the Sweeping (1) \glossterm{weapon tag} (see \pcref{Natural Weapons}).
            % TODO: too fast?
            In addition, you gain a \plus30 foot bonus to your fly speed with your \textit{draconic flight} ability, but your maneuverability drops to poor maneuverability (see \pcref{Flying Maneuverability}).

        \subsubsection{Basic Class Abilities}
            If you choose dragon as your base class, you gain the following abilities.

            \cf{Drg}{Power} Your automatic bonus to power based on your rank in this archetype is the same as a fighter (see \pcref{Class-Based Power Bonuses}).

            \cf{Drg}{Defenses}
            You gain the following bonuses to your \glossterm{defenses}: \plus7 Fortitude, \plus3 Reflex, \plus5 Mental.
            In addition, you gain a \plus1 bonus to your \glossterm{vital rolls}.

            \cf{Drg}{Resources} You have the following \glossterm{resources}:
            \begin{itemize}
                \item Three \glossterm{attunement points}, which you can use to attune to items and abilities that affect you (see \pcref{Attunement Points}).
                \item A \glossterm{fatigue tolerance} equal to 5 \add your Constitution \add half your Willpower.
                    Your fatigue tolerance makes it easier for you to use powerful abilities that fatigue you (see \pcref{Fatigue}).
                \item A number of \glossterm{insight points} equal to 2 \add your Intelligence.
                    You can spend insight points to gain additional abilities (see \pcref{Insight Points}).
                \item Four \glossterm{trained skills} from among your \glossterm{class skills}, plus additional trained skills equal to your Intelligence (see \pcref{Skills}).
            \end{itemize}

            \cf{Drg}{Weapon Proficiencies} 
            You are not proficient with any weapon groups, even simple weapons.
            You are still proficient with your natural weapons.

            \cf{Drg}{Armor Proficiencies} 
            You are proficient with light and medium armor.
            Armor shaped appropriately for dragons can be hard to find, and may need to be crafted individually for the dragon.

            \cf{Drg}{Skills}
            You have the following \glossterm{class skills}:
            \begin{itemize}
                \item \subparhead{Strength} Climb, Jump, Swim.
                \item \subparhead{Dexterity} Balance, Stealth.
                \item \subparhead{Constitution} Endurance.
                \item \subparhead{Intelligence} Craft, Deduction, Knowledge (arcana), Medicine.
                \item \subparhead{Perception} Awareness, Creature Handling, Social Insight, Survival.
                \item \subparhead{Other} Deception, Intimidate, Persuasion.
            \end{itemize}

    \subsection{Drakkenfel}

        A drakkenfel is created when a dragon's scales are removed while the dragon still lives. The scales retain much of the dragon's power, and without them, the dragon is cursed to continue its life as a diminished, mortal creature.
        Drakkenfel are extremely rare, as there are few who are powerful enough to subdue a dragon to extract its scales -- and fewer still who would dare to do so, knowing that they would earn the eternal enmity of dragonkind.

        Physically, a drakkenfel resembles a wyrmling dragon, except that it is completely scaleless.
        Its skin is leathery and rough, and some drakkenfel bear scars from the ritual that removed their scales.
        Most drakkenfel retain tatters of their wings, but they are always nonfunctional.

        \parhead{Size} Small. This gives a drakkenfel a 20 foot \glossterm{base speed} and a \plus5 bonus to the Stealth skill, among other effects (see \pcref{Size Categories}).
        \parhead{Attributes} \minus2 Strength, either \plus1 Dexterity or \plus1 Constitution.
        \parhead{Special Abilities}
        \begin{itemize}
            \itemhead{Bite} A drakkenfel's mouth can be used to bite (see \pcref{Natural Weapons}). A drakkenfel's bite attack deals 1d6 damage.
            \itemhead{Draconic Essence} Each drakkenfel was once a type of dragon.
                When creating a drakkenfel, choose which type of dragon it used to be.
                This is inherent to the drakkenfel, and cannot be changed.
                A list of dragons and their associated energy type is given on \tref{Dragon Types}.
            \itemhead{Damage Tolerance} A drakkenfel is \trait{impervious} to damage of the type associated with its \textit{draconic essence}.
            \itemhead{Low-light Vision} Drakkenfel have \trait{low-light vision}, allowing them to see clearly in \glossterm{shadowy illumination} (see \pcref{Low-light Vision}).
            \itemhead{Sleeping Dragon} If a drakkenfel recovers its stolen scales, it immediately becomes a dragon again.
                Its statistics become identical to its statistics before losing its scales, including level.
        \end{itemize}
        \parhead{Automatic Languages} Common, Draconic, any one \glossterm{common language}.

    \subsection{Drow}

        Drow are an offshoot group of elves that live deep underground.
        The deep caves are a far harsher environment than the surface world.
        Resources are scarce, and dangerous monsters are far more common.
        In order to survive, drow were forced to adopt a variety of practices condemned by surface civilizations.
        The most notorious are their frequent use of poison, their refusal to take prisoners, their willingness to eat any non-drow creatures they kill, even sentient creatures.
        In addition, drow society tends to reward selfishness and ambition more explicitly than surface civilizations, and the vast majority of drow are evil.

        When drow find opportunities to reach the surface world, they seek to conquer territory for themselves, usually with great violence.
        They have always been defeated and banished back to their caves, but surface civilizations still remember the danger that drow pose.
        Even more so than tieflings or orcs, who are already viewed with suspicion, drow are anathema in almost any civilized society.
        Drow who escape the deep caves are more likely to find a peaceful existence on other planes that do not fear an underground invasion.

        \parhead{Size} Medium.
        \parhead{Attributes} \minus1 Constitution, \plus1 Dexterity
        \parhead{Special Abilities}
        \begin{itemize}
            \itemhead{Darkvision} Drow have \trait{darkvision} with a 120 foot range, allowing them to see in complete darkness (see \pcref{Darkvision}).
            \itemhead{Deep Darkness}[Magical] A drow can use the \textit{deep darkness} ability as a \glossterm{standard action}.
                \begin{sustainability}{Deep Darkness}{\abilitytag{Sustain} (minor)}
                    \rankline
                    \target{One \glossterm{zone} within \rngmed range}
                    You can choose this ability's radius, up to a maximum of a \areamed radius.
                    Light within or passing through the area is dimmed to be no brighter than \glossterm{shadowy illumination}
                    Any object or effect which blocks light also blocks this spell's effect.
                \end{sustainability}
            \itemhead{Drow Prejudice} Almost all surface-dwellers have negative associations with drow.
                Drow have an Opposition relationship with most people that they meet, which influences people's behavior and makes Persuasion checks harder (see \pcref{Persuasion}).
                People in some locations, such as deep underground, do not have this attitude.
            \itemhead{Keen Senses} Drow gain a \plus2 bonus to the Awareness skill (see \pcref{Awareness}).
            \itemhead{Poison Tolerance} Drow are \trait{impervious} to poison.
            \itemhead{Sensitive Eyes} Drow take a \minus2 penalty to \glossterm{accuracy} while they are in \glossterm{bright illumination}.
                This penalty is doubled while they are in \glossterm{brilliant illumination}.
            \itemhead{Trance} Drow do not sleep, and are immune to \glossterm{magical} effects that would cause them to sleep.
                Instead of sleeping, drow can trance for 4 hours.
                An elf in trance may make Perception-based checks at a \minus5 penalty.
                Drow must still avoid strenuous activity for 8 hours to heal and gain other benefits of taking a \glossterm{long rest}.
        \end{itemize}
        \parhead{Automatic Languages} Common, Elven, Undercommon

    \subsection{Dryaidi}

        Dryaidi are humanoid creatures that resemble plants. They are descended from dryads.

        \parhead{Size} Medium.
        \parhead{Attributes} \plus1 Constitution, \minus1 Dexterity.
        \parhead{Special Abilities}
        \begin{itemize}
            \itemhead{Ingrain} A dryaidi use the \textit{ingrain} ability as a standard action.
                \begin{activeability}{Ingrain}
                    \rankline
                    The dryaidi's land speed becomes 5 feet, regardless of any modifiers that normally apply.
                    It gains a \plus4 bonus to Fortitude defense and a \plus1 bonus to Armor defense.
                    When moving, it may ignore \glossterm{difficult terrain} of any kind.
                    If the dryaidi takes a \glossterm{long rest} while this ability is active, it acquires nutrients sufficient to replace a day's worth of food and water.

                    This ability lasts until the dryaidi ends it as a standard action.
                \end{activeability}
            \itemhead{Plant Nature} A dryaidi is considered both a creature and a plant.
            \itemhead{Unhurried and Unfaltering} Dryaidi have a \minus10 penalty to their \glossterm{land speed}.
                However, wearing medium or heavy \glossterm{body armor} does not reduce a dryaidi's movement speed (see \pcref{Armor Usage Classes}).
        \end{itemize}
        \parhead{Automatic Languages} Common, Sylvan.

    \subsection{Eladrin}

        \parhead{Size} Medium.
        \parhead{Attributes} \minus1 Constitution, either \plus1 Dexterity or \plus1 Willpower
        \parhead{Special Abilities}
        \begin{raggeditemize}
            \itemhead{Fae Step} As a standard action, you can use the \ability{fae step} ability.
            \begin{activeability}{Fae Step}
                \rankline
                You \glossterm{teleport} into an unoccupied location within \shortrange on a stable surface that can support your weight.
                If the destination is invalid, this ability fails with no effect.

                \rankline
                This ability improves based on your rank in your highest-rank archetype.
                \rank{3} The range increases to \medrange.
                \rank{5} The range increases to \longrange.
                \rank{7} The range increases to \distrange.
            \end{activeability}
            \itemhead{Fae Season} Eladrin respond strongly to their emotions, and change their abilities based on the season they currently represent.
                An eladrin must choose one of the following seasons when it finishes a \glossterm{short rest}.
                The chosen season lasts until it changes to a different season.
                \subcf{Spring} \plus1 bonus to Mental defense, \minus1 penalty to Fortitude defense.
                Eladrin expressing the spring season are filled with the joy of a new year.
                However, they are also visibly thinner and more frail, as if recovering from a long winter.
                \subcf{Summer} \plus1 bonus to Fortitude defense, \minus1 penalty to Reflex defense.
                Eladrin expressing the summer season are visibly hearty and a little more plump.
                However, they also move with all the alacrity of a long summer day.
                \subcf{Autumn} \plus1 bonus to all checks, \minus1 penalty to \glossterm{accuracy}.
                Eladrin expressing the autumn season embody the spirit of the harvest.
                They are filled with goodwill towards all creatures, and prefer finding peaceful solutions to problems.
                Their bodies tend to be firm and toned, reflecting the hard work required to prepare for the winter.
                \subcf{Winter} \plus1 bonus to \glossterm{vital rolls}, \minus1 penalty to Mental defense.
                Eladrin expressing the winter season are prepared for the worst.
                They tend to be dour and pessimistic, but they press on despite the certainty of doom.
            \itemhead{Low-light Vision} Eladrin have \trait{low-light vision}, allowing them to see clearly in \glossterm{shadowy illumination} (see \pcref{Low-light Vision}).
            \itemhead{Trance} Eladrin do not sleep, and are immune to \glossterm{magical} effects that would cause them to sleep.
                Instead of sleeping, eladrin can trance for 4 hours.
                An eladrin in trance may make Perception-based checks at a \minus5 penalty.
                Eladrin must still avoid strenuous activity for 8 hours to heal and gain other benefits of taking a \glossterm{long rest}.
        \end{raggeditemize}
        \parhead{Species Feat Options} 
        \parhead{Automatic Languages} Common, Sylvan, and any one \glossterm{common language} (see \tref{Common Languages}).

    \subsection{Harpy}
        Harpies are winged creatures with the upper body of a humanoid and the lower body of a bird.
        The vast majority of harpies are female, but male harpies do exist.

        \parhead{Creature Type} Unlike most other playable species, harpies are considered monstrous humanoids instead of humanoids.
        \parhead{Size} Medium.
        \parhead{Attributes} \minus1 Intelligence, \plus1 Dexterity.
        \parhead{Special Abilities}
        \begin{itemize}
            \itemhead{Prehensile Talons} Harpies have a talon natural weapon on each foot (see \pcref{Natural Weapons}).
                In addition, they can use their feet as \glossterm{free hands}.
                However, their talons are not able to effectively wield shields or manufactured weapons.
            \itemhead{Wings} Harpies have a feathered wings that sprout from their shoulders in place of arms.
                These wings grant them a glide speed equal to the \glossterm{base speed} for their size (see \pcref{Gliding}).
            \itemhead{Armless} Harpies do not have arms or hands.
            \itemhead{Harpy Archetype} Harpies only gain two class archetypes instead of three.
                Instead, they treat the Harpy archetype as one of their archetypes, and they gain ranks in it just like they gain ranks in class archetypes.
        \end{itemize}
        \parhead{Automatic Languages} Common.
        
        \subsubsection{Harpy Archetype}

            \cf{Hrp}[1]{Winged Agility} You gain a \plus4 bonus to the Balance and Jump skills as long as you can use your wings to help you.

            \cf{Hrp}[1]{Winged Combat} As a \glossterm{free action}, you can make a short hop off of solid ground to hover in your space using your wings.
            This allows you to use both of your feet without standing on them during that phase, so you can attack with both talons at once.
            You must land at the end of the phase, and you can only use this ability once per round.
            This brief hovering does not cause you to suffer penalties for flying.

            \cf{Hrp}[2]{Luring Song} You can use the \textit{luring song} ability as a standard action.
            \begin{activeability}{Luring Song}[\abilitytag{Auditory}, \abilitytag{Compulsion}]
                    \rankline
                    Make an attack vs. Mental against a creature within \longrange.
                    \hit As a \glossterm{condition}, the target must move towards you as best it can during each \glossterm{movement phase}.
                    In addition, it cannot move farther away from you at any time, except as necessary to get closer to you (such as to avoid an intervening obstacle).
                    It can otherwise act freely, and is still able to attack you and your allies.

                    If you attack the target with any ability other than this one, this effect is automatically broken.
                    When this effect ends, the target becomes immune to this effect until it takes a \glossterm{short rest}.
                    \crit The condition must be removed twice before the effect ends.

                    \rankline
                    \rank{4} You can target an additional creature within range.
                    \rank{6} The maximum number of targets increases to 5.
                \end{activeability}

            \cf{Hrp}[2]{Raking Talons} You gain a \plus1d damage bonus with all \glossterm{natural weapons}.

            \cf{Hrp}[3]{Flight} You gain a \glossterm{fly speed} equal to the \glossterm{base speed} for your size with a maximum height of 15 feet (see \pcref{Flying}).
            As a \glossterm{free action}, you can increase your \glossterm{fatigue level} by one to ignore this height limit until the end of the round.

            \cf{Hrp}[4]{Greater Winged Agility} You can use your wings to help you maneuver on the ground more effectively, such as by briefly hovering or gliding over obstacles.
            The Balance and Jump bonuses from your \textit{winged agility} ability increase to \plus8.
            In addition, you ignore \glossterm{difficult terrain}.

            \cf{Hrp}[4]{Greater Winged Combat} You can use your \textit{winged combat} any number of times in the same round.
            You must still land at the end of each phase.

            \cf{Hrp}[5]{Greater Raking Talons} The damage bonus from your \textit{raking talons} ability improves to \plus2d.

            % dazed enemies in medarea radius is a rank 5 spell. charm + not a condition make this better.
            \cf{Hrp}[5]{Siren Song} You can use the \textit{siren song} ability as a standard action.
            \begin{sustainability}{Siren Song}{\abilitytag{Auditory}, \abilitytag{Emotion}, \abilitytag{Sustain} (minor)}
                \rankline
                Make an attack vs. Mental against any number of creatures within a \medarea radius from you.
                \hit Each target is both \charmed by you and \dazed as long as it can still hear you sing.
                It remains dazed even if it stops being charmed, such as if you or your allies attack it.
                This ability does not have the \abilitytag{Subtle} tag, so an observant target may notice that it is being influenced.
                \rankline
                \rank{7} The area increases to a \largearea radius.
            \end{sustainability}

            \cf{Hrp}[6]{Agile Flight} Your flight improves to have good maneuverability (see \pcref{Flying Maneuverability}).
            In addition, your maximum height increases to 30 feet.

            \cf{Hrp}[7]{Mythic Siren} You gain a \plus5 \glossterm{accuracy} bonus with your \ability{luring song} and \ability{siren song} abilities.

        \subsubsection{Basic Class Abilities}
            If you choose harpy as your base class, you gain the following abilities.

            \cf{Hrp}{Power} Your automatic bonus to power based on your rank in this archetype is the same as a fighter (see \pcref{Class-Based Power Bonuses}).

            \cf{Hrp}{Defenses}
            You gain the following bonuses to your \glossterm{defenses}: \plus1 Armor, \plus3 Fortitude, \plus5 Reflex, \plus7 Mental.
            In addition, you gain a \plus1 bonus to your \glossterm{vital rolls}.

            \cf{Hrp}{Resources} You have the following \glossterm{resources}:
            \begin{itemize}
                \item Three \glossterm{attunement points}, which you can use to attune to items and abilities that affect you (see \pcref{Attunement Points}).
                \item A \glossterm{fatigue tolerance} equal to 4 \add your Constitution \add half your Willpower.
                    Your fatigue tolerance makes it easier for you to use powerful abilities that fatigue you (see \pcref{Fatigue}).
                \item A number of \glossterm{insight points} equal to 2 \add your Intelligence.
                    You can spend insight points to gain additional abilities (see \pcref{Insight Points}).
                \item Six \glossterm{trained skills} from among your \glossterm{class skills}, plus additional trained skills equal to your Intelligence (see \pcref{Skills}).
            \end{itemize}

            \cf{Hrp}{Weapon Proficiencies} 
            You are not proficient with any weapon groups, even simple weapons.
            You are still proficient with your natural weapons.

            \cf{Hrp}{Armor Proficiencies} 
            You are proficient with light armor.

            \cf{Hrp}{Skills}
            You have the following \glossterm{class skills}:
            \begin{itemize}
                \item \subparhead{Strength} Climb, Jump.
                \item \subparhead{Dexterity} Balance, Flexibility, Perform, Stealth.
                \item \subparhead{Perception} Awareness, Creature Handling, Deception, Persuasion, Survival.
                \item \subparhead{Other} Intimidate.
            \end{itemize}

    \subsection{Kit}

        Kit are humanoid creatures that have noticeable foxlike characteristics.
        They are descended from natural fox spirits.
        All kit have at least one tail, and some have multiple tails.
        Their tails are distinctly fluffy and fox-like, and most kit put effort into concealing their tails to avoid revealing their true nature.

        \parhead{Size} Medium.
        \parhead{Attributes} No change.
        \parhead{Special Abilities}
        \begin{itemize}
            \itemhead{Foxlike Agility} Kit gain a \plus2 bonus to the Balance and Stealth skills.
            \itemhead{Illusory Guise} As a standard action, a kit can magically disguise its physical appearance in minor ways.
                This functions like the \textit{disguise creature} ability with a \plus4 bonus, except that a kit cannot change the appearance of its equipment, creature type, or number of limbs, including any tails it may have (see \pcref{Disguise Creature}).
                This is a \glossterm{magical} ability.
                It lasts until the kit \glossterm{dismisses} it as a free action or uses this ability again.
            \itemhead{Instictive Trickster} Kit gain a \plus2 bonus to the Deception and Social Insight skills.
            \itemhead{Low-light Vision} Kit have \trait{low-light vision}, allowing them to see clearly in \glossterm{shadowy illumination} (see \pcref{Low-light Vision}).
        \end{itemize}
        \parhead{Automatic Languages} Common, any one \glossterm{common language}.

    \subsection{Naiadi}
        Naiadi are humanoid creatures descended from naiads.
        Most naiadi are unusually physically appealing, but show no other outward signs of their heritage.

        \parhead{Size} Medium.
        \parhead{Attributes} No change.
        \parhead{Special Abilities}
        \begin{itemize}
            \itemhead{Create Water} A naiadi can cast the \spell{create water} cantrip.
                When they do so, they do not require verbal or somatic \glossterm{casting components}, and their spellcasting rank is considered to be equal to their rank in their highest rank archetype.
                If they would already know that cantrip through the Aquamancy sphere, the volume of water created with the cantrip doubles.
            \itemhead{Enchanting Appearance} A naiadi gains a \plus2 bonus to the Creature Handling, Perform, and Persuasion skills.
            \itemhead{Low-light Vision} Naiadi have \trait{low-light vision}, allowing them to see clearly in \glossterm{shadowy illumination} (see \pcref{Low-light Vision}).
            \itemhead{Naiadi Archetype} You may choose three class archetypes, as normal.
                However, you may choose the Naiadi archetype in place of one of your class archetypes.
                If you do, you gain ranks in it just like you gain ranks in class archetypes.
                You cannot choose naiadi as your base class.
            \itemhead{Water Affinity} A naiadi has a \glossterm{swim speed} equal to the \glossterm{base speed} for their size.
                In addition, they can breathe clean water like a human breathes air.
        \end{itemize}
        \parhead{Automatic Languages} Common, Sylvan, any one \glossterm{common language}.

        \subsubsection{Naiadi Archetype}

            \cf{Nai}[1]{Water Bond}[Magical] You can form a bond with a fresh stream, lake, or other Gargantuan or larger body of fresh water (not salt water).
            Forming a bond or severing a bond takes one week of meditation and ritual, periodically interrupted by rest.
            Forming a bond also requires asking permission from the water.
            Any individual body of water can only be bonded to one naiad or naiadi in this way.

            As long as your bonded water remains clean, pure, and large enough to be a valid subject of bonding, you gain a \plus1 bonus to Mental defense, a \plus1 bonus to your \glossterm{power}, and a bonus equal to twice your rank in this archetype to your \glossterm{hit points}.
            If your bonded water becomes contaminated or shrinks below the minimum size, these bonuses are inverted into penalties until you sever the bond.
            You can passively observe the general health and status of water you are bonded to, including knowing when significant pollutants enter the water and when the water grows or shrinks significantly.

            \cf{Nai}[2]{Fluidseeker} You gain a \plus1 bonus to \glossterm{accuracy} against creatures significantly composed of water or watery fluids.
            This is true of almost all living creatures.

            \cf{Nai}[2]{Freshwater Fountain} The volume of water you can create with the \spell{create water} cantrip increases by five times.

            \cf{Nai}[3]{Aqueous Form} You can cast the \spell{aqueous form} spell.
            When you do, you do not require verbal or somatic \glossterm{casting components}, and you use your rank in this archetype as your your spellcasting rank.
            In addition, it does not reduce your \glossterm{damage resistance}.

            \cf{Nai}[4]{Greater Water Bond} The bonuses from your \textit{water bond} ability increase to \plus2 Mental defense, \plus2 power, and three times your rank in this archetype to your hit points.

            \cf{Nai}[5]{Greater Fluidseeker} The accuracy bonus from your \textit{fluidseeker} ability increases to \plus2.

            \cf{Nai}[5]{Greater Freshwater Fountain} The multiplier from your \textit{freshwater fountain} ability increases to twenty times the normal volume of water.

            \cf{Nai}[6]{Greater Aqueous Form} When you cast the \spell{aqueous form} spell, it does not have the \abilitytag{Attune} tag.
            Instead, it lasts until you \glossterm{dismiss} it as a \glossterm{free action}.

            \cf{Nai}[7]{Supreme Water Bond} The bonuses from your \textit{water bond} ability increase to \plus3 Mental defense, \plus4 power, and three times your rank in this archetype to your hit points.

    \subsection{Oozeborn}
        Oozeborn are sentient ooze creatures that have gained true sentience through a strange quirk of their birth.
        They are very rare to see in civilized lands, as most oozeborn lack the opportunity to discover more than the dark caves in which they were spawned.
        Since they often grow up without mentorship from any civilized creature, oozeborn tend to have odd mannerisms and a poor ability to mask their emotions, even after spending years in civilization.
        Old oozeborn may eventually adapt to societal norms and act perfectly natural, or they may abandon civilized company entirely.

        The body of an oozeborn is amorphous, and they lack any identifiable internal organs.
        Their natural color depends on the nature of the ooze that spawned them, so green and gray are the most common colors.
        Adventuring oozeborn typically assume a bipedal shape for both practical and social convenience, but their natural shape is a loosely spherical blob.
        Unconscious oozeborn revert to their default state automatically, though some learn to maintain a semblance of cohesion while asleep.

        \parhead{Creature Type} Unlike most other playable species, oozeborn are considered animates instead of humanoids.
        \parhead{Size} Medium.
        \parhead{Attributes} \minus1 Intelligence, \plus1 Constitution.
        \parhead{Special Abilities}
        \begin{itemize}
            \itemhead{Acid Tolerance} Ooozeborn are \trait{impervious} to acid damage.
            \itemhead{Amorphous Form} An oozeborn's natural form is a loosely spherical blob.
                They have a \minus5 foot penalty to their \glossterm{land speed}, but they gain a \plus5 bonus to the Flexibility skill (see \pcref{Flexibility}).
                They can use the \ability{mold body} ability as a standard action to adopt a particular shape.
                \begin{sustainability}{Mold Body}{\abilitytag{Sustain} (free)}
                    \rankline
                    You make a Disguise check to alter your appearance (see \pcref{Disguise Creature}).
                    This physically changes your body to match the results of your disguise.
                    You gain a \plus4 bonus on the check, and you ignore penalties for changing your gender, species, subtype, age, and number of limbs (up to 4).
                    However, this effect is unable to alter your equipment in any way.

                    You can sustain this ability for any length of time without mental strain, ignoring the normal 5 minute limit.
                \end{sustainability}
            \itemhead{Compressible Body} Oozeborn can compress their head and shoulders down to a minimum of a one inch radius, allowing them to squeeze through very small areas.
                Their clothing or armor is not compressed, so they may limit their ability to move through extremely narrow spaces.
            \itemhead{Darkvision} Oozeborn have \trait{darkvision} with a 60 foot range, allowing them to see in complete darkness (see \pcref{Darkvision}).
            \itemhead{Oozeborn Archetype} Oozeborn only gain two class archetypes instead of three.
                Instead, they treat the Oozeborn archetype as one of their archetypes, and they gain ranks in it just like they gain ranks in class archetypes.
        \end{itemize}
        \parhead{Automatic Languages} Common.
        
        \subsubsection{Oozeborn Archetype}

            \cf{Ooz}[1]{Acidic Body} You gain a slam \glossterm{natural weapon}, which deals 1d10 damage and requires a free hand to use, except that it also has the Long \glossterm{weapon tag} (see \pcref{Natural Weapons}).
            In addition, all damage you deal with natural weapons is acid damage in addition to its other types.
            This does not affect damage you deal with manufactured weapons.

            \cf{Ooz}[1]{Poison Dissolution} You are \trait{impervious} to poisons.

            \cf{Ooz}[2]{Darkborn Senses} You gain \trait{blindsense} with a 60 foot range, allowing you to sense your surroundings without light (see \pcref{Blindsense}).
            If you already have the blindsense ability, you increase its range by 60 feet.
            In addition, you gain \trait{blindsight} with a 15 foot range, allowing you to see without light (see \pcref{Blindsight}).
            If you already have the blindsight ability, you increase its range by 15 feet.

            \cf{Ooz}[2]{Ingest Object} You can use the \textit{ingest object} ability as a standard action.
            This functions like the \spell{absorb object} spell, except that you do not gain the ability to absorb objects that are larger than Medium size.
            Anything you absorb in this way takes a single point of \glossterm{environmental} acid damage during each \glossterm{action phase} while it remains absorbed.
            This damage is insufficient to hurt most objects made from wood, stone, or metal, but it can destroy more fragile objects like paper or complex mechanical traps.

            \cf{Ooz}[3]{Greater Amorphous Form} You gain a \plus4 bonus to your defenses when determining whether a \glossterm{strike} gets a \glossterm{critical hit} against you instead of a normal hit.

            \cf{Ooz}[3]{Greater Compressible Body} You reduce your penalties for \squeezing by 1.

            \cf{Ooz}[4]{Greater Acidic Body} You gain a \plus1d damage bonus with natural weapons.

            \cf{Ooz}[4]{Acidic Apotheosis} You are \trait{immune} to acid damage and poisons.

            \cf{Ooz}[5]{Greater Darkborn Senses} The range of your \trait{blindsense} increases by 60 feet.
            In addition, the range of your \trait{blindsight} increases by 15 feet.

            \cf{Ooz}[5]{Greater Ingest Object} The maximum number of objects you can absorb with your \textit{ingest object} ability increases to 2.
            In addition, you may absorb \glossterm{allies} with that ability in addition to unattended objects.

            \cf{Ooz}[6]{Supreme Amorphous Form} The bonus from your \textit{greater amorphous form} ability increases to \plus8.
            In addition, your \ability{mold body} ability loses the \abilitytag{Sustain} (free) tag.
            Instead, it lasts until you choose to \glossterm{dismiss} it as a \glossterm{free action}.
            This allows you to maintain your shape while unconscious.

            \cf{Ooz}[6]{Supreme Compressible Body} You reduce your penalties for squeezing by 2, which means you take no penalties for squeezing unless you use the \ability{tight squeeze} ability (see \pcref{Flexibility}).

            \cf{Ooz}[7]{Supreme Acidic Body} The bonus from your \textit{greater acidic body} ability increases to \plus2d.

            \cf{Ooz}[7]{Third Arm} When you use your \ability{mold body} ability, you can create three arms instead of two.
            You can use all three hands as free hands.
            For example, this can allow you to use a heavy weapon and a shield simultaneously.

        \subsubsection{Basic Class Abilities}
            If you choose oozeborn as your base class, you gain the following abilities.

            \cf{Ooz}{Power} Your automatic bonus to power based on your rank in this archetype is the same as a fighter (see \pcref{Class-Based Power Bonuses}).

            \cf{Ooz}{Defenses}
            You gain the following bonuses to your \glossterm{defenses}: \plus7 Fortitude, \plus3 Reflex, \plus5 Mental.
            In addition, you gain a \plus2 bonus to your \glossterm{vital rolls}.

            \cf{Ooz}{Resources} You have the following \glossterm{resources}:
            \begin{itemize}
                \item Three \glossterm{attunement points}, which you can use to attune to items and abilities that affect you (see \pcref{Attunement Points}).
                \item A \glossterm{fatigue tolerance} equal to 6 \add your Constitution \add half your Willpower.
                    Your fatigue tolerance makes it easier for you to use powerful abilities that fatigue you (see \pcref{Fatigue}).
                \item A number of \glossterm{insight points} equal to 1 \add your Intelligence.
                    You can spend insight points to gain additional abilities (see \pcref{Insight Points}).
                \item Three \glossterm{trained skills} from among your \glossterm{class skills}, plus additional trained skills equal to your Intelligence (see \pcref{Skills}).
            \end{itemize}

            \cf{Ooz}{Weapon Proficiencies} 
            You are proficient with simple weapons.

            \cf{Ooz}{Armor Proficiencies} 
            You are proficient with light armor.
            However, you cannot wear armor unless you sustain your \ability{mold body} ability, so you cannot wear armor while sleeping.

            \cf{Ooz}{Skills}
            You have the following \glossterm{class skills}:
            \begin{itemize}
                \item \subparhead{Strength} Climb, Jump, Swim.
                \item \subparhead{Dexterity} Balance, Flexibility, Sleight of Hand, Stealth.
                \item \subparhead{Constitution} Endurance.
                \item \subparhead{Intelligence} Craft, Knowledge (dungeoneering).
                \item \subparhead{Perception} Awareness, Survival.
                \item \subparhead{Other} Intimidate.
            \end{itemize}

    \subsection{Orc}
        Orcs are green-skinned humanoid creatures known for their strength and brutality.

        \parhead{Size} Medium.
        \parhead{Attributes} \plus1 starting Strength, \minus1 starting Intelligence.
        \parhead{Special Abilities}
        \begin{itemize}
            \itemhead{Darkvision} Orcs have \trait{darkvision} with a 60 foot range, allowing them to see in complete darkness (see \pcref{Darkvision}).
            \itemhead{Intimidating} Orcs gain a \plus3 bonus to the Intimidate skill (see \pcref{Intimidate}).
            \itemhead{Mighty} You gain a \plus1 bonus to your Strength for the purpose of determining your \glossterm{weight limits} (see \pcref{Weight Limits}).
            \itemhead{Orcish Prejudice} Most people have negative associations with orcs thanks to their violent and dangerous behavior.
                Orcs have an Acquaintance (negative) relationship with most people that they meet, which influences people's behavior and makes Persuasion checks harder (see \pcref{Persuasion}).
                People in some locations, such as orcish tribes, do not have this attitude.
            \itemhead{Powerful} You gain a bonus equal to a quarter of your level (minimum 1) to your \glossterm{power}.
        \end{itemize}
        \parhead{Automatic Languages} Common, Orc.

    \subsection{Tieflings}

        Tieflings are humanoid creatures descended from fiends.
        They inherit a tendency towards evil from their ancestors, and are therefore viewed with great suspicion by most civilized societies.
        Good-aligned tieflings exist, but they may have difficulty using their natural talents for subterfuge and deceit for noble ends, and they often struggle with hidden vices.

        \parhead{Size} Medium.
        \parhead{Attributes} No change.
        \parhead{Special Abilities}
        \begin{itemize}
            \itemhead{Darkvision} Tieflings have \trait{darkvision} with a 60 foot range, allowing them to see in complete darkness (see \pcref{Darkvision}).
            \itemhead{Demonic Prejudice} Most people have negative associations with tieflings thanks to the malign influence that demons have on the world.
                Tieflings have an Opposition relationship with most people that they meet, which influences people's behavior and makes Persuasion checks harder (see \pcref{Persuasion}).
                People in some locations, such as the Abyss, do not have this attitude.
            \itemhead{Hellfire Tolerance} Tieflings are \trait{impervious} to fire damage.
            \itemhead{Infernal Presence} Tieflings gain a \plus2 bonus to the Deception and Intimidate skills.
            \itemhead{Tiefling Archetype} You may choose three class archetypes, as normal.
                However, you may choose the Tiefling archetype in place of one of your class archetypes.
                If you do, you gain ranks in it just like you gain ranks in class archetypes.
                You cannot choose tiefling as your base class.
        \end{itemize}
        \parhead{Automatic Languages} Abyssal, Common, any one \glossterm{common language}.

        \subsubsection{Tiefling Archetype}
            \cf{Tif}[1]{Abyssal Hop} You can use the \ability{abyssal hop} ability as a standard action.
            \begin{activeability}{Abyssal Hop}[\abilitytag{Magical}]
                \rankline
                You teleport horizontally into an unoccupied location within \shortrange on a stable surface that can support your weight.
                If the destination is invalid, this spell fails with no effect.
                In addition, make an attack vs. Reflex against each \glossterm{enemy} adjacent to your location after you arrive.
                \hit Each target takes 1d4 \add half power fire damage.
                \rankline
                \rank{2} The damage increases to 1d6.
                \rank{3} The damage increases to 1d10.
                \rank{4} The damage increases to 2d6.
                \rank{5} The damage increases to 2d10.
                \rank{6} The damage increases to 4d6.
                \rank{7} The damage increases to 4d10.
            \end{activeability}

            \cf{Tif}[1]{Infernal Resilience} You gain a bonus equal to twice your rank in this archetype to your \glossterm{damage resistance}.

            \cf{Tif}[2]{Infernal Ancestry} You deepen your connection to a particular aspect of your demonic ancestry.
            Choose one of the following infernal ancestries: hellfire conduit, tempting allure, or unholy might.
            You gain a benefit based on your chosen ancestry.
            \begin{itemize}
                \item Hellfire Conduit: You can use the \textit{infernal rebuke} ability as a standard action.
                    \begin{activeability}{Infernal Rebuke}[\abilitytag{Magical}]
                        \rankline
                        Make an attack vs. Fortitude against one creature within \shortrange.
                        You gain a \plus2 bonus to \glossterm{accuracy} with this attack if the target attacked you during the previous round.
                        \hit The target takes 1d10 \add \glossterm{power} fire damage.

                        \rankline
                        \rank{3} The damage increases to 2d6.
                        \rank{4} The damage increases to 2d10.
                        \rank{5} The damage increases to 4d6.
                        \rank{6} The damage increases to 4d10.
                        \rank{7} The damage increases to 5d10.
                    \end{activeability}
                \item Tempting Allure: You gain a \plus2 bonus to the Deception, Disguise, and Persuasion skills.
                    In addition, you can use the \ability{charming temptation} ability as a standard action.
                    \begin{sustainability}{Charming Temptation}{\abilitytag{Emotion}, \abilitytag{Magical}, \abilitytag{Subtle}, \abilitytag{Sustain} (minor)}
                        \rankline
                        \noindent

                        Make an attack vs. Mental against one creature within \medrange.
                        You take a \minus10 penalty to \glossterm{accuracy} with this attack against creatures who have made an attack or been attacked since the start of the last round.
                        \vspace{0.25em}
                        \hit The target is \charmed by you.
                        Any act by you or by creatures that appear to be your allies that threatens or harms the charmed person breaks the effect.
                        Harming the target is not limited to dealing it damage, but also includes causing it significant subjective discomfort.
                        An observant target may interpret overt threats to its allies as a threat to itself.
                        \crit Actions which threaten the charmed person without harming them do not break the effect.

                        \rankline

                        \noindent The attack's \glossterm{accuracy} increases by \plus1 for each rank beyond 2.
                        \vspace{0.1em}
                    \end{sustainability}
                \item Unholy Might: You gain two claw natural weapons and one bite natural weapon (see \pcref{Natural Weapons}).
                    In addition, you gain a \plus1d bonus to your damage with all weapons.
            \end{itemize}

            \cf{Tif}[3]{Abysswalker} You can use your \ability{abyssal hop} ability to teleport as a \glossterm{movement} instead of as a standard action.
            When you do, you do not deal fire damage at your destination, and you \glossterm{briefly} cannot use that ability as a movement again.

            \cf{Tif}[4]{Greater Infernal Ancestry} The benefits of your \textit{infernal ancestry} ability improve.
            \begin{itemize}
                \item Hellfire Conduit: You gain a \plus3 bonus to your \glossterm{power}.
                    In addition, the area affected by your \textit{abyssal hop} ability increases to a \smallarea radius from your destination.
                \item Tempting Allure: The skill bonuses from your \textit{infernal ancestry} ability increase to \plus3.
                    In addition, you can use the \ability{dominating temptation} ability as a standard action.
                    \begin{activeability}{Dominating Temptation}[\abilitytag{Magical}]
                        \rankline
                        \noindent
                        Make an attack vs. Mental against one creature within \shortrange.%
                        \vspace{0.25em}
                        \hit The target is \stunned as a \glossterm{condition}.
                        \crit The target is \confused instead of stunned.
                        In addition, if the target is humanoid and was already confused from a previous casting of this spell, you may \glossterm{attune} to this ability.
                        If you do, it becomes \dominated by you for the duration of that attunement.

                        \rankline
                        \noindent The attack's \glossterm{accuracy} increases by \plus1 for each rank beyond 4.
                        \vspace{0.1em}
                    \end{activeability}
                \item Unholy Might: You can use the \ability{unholy strength} ability as a standard action.
                    \begin{attuneability}{Unholy Surge}{\abilitytag{Attune}, \abilitytag{Magical}}
                        \rankline
                        You gain a \plus1 \glossterm{magic bonus} to your Strength.
                    \end{attuneability}
            \end{itemize}

            \cf{Tif}[5]{Greater Hellfire Tolerance} You become \trait{immune} to fire damage.

            \cf{Tif}[5]{Greater Infernal Resilience}  The bonus from your \ability{infernal resilience} ability increases to three times your rank in this archetype.

            \cf{Tif}[6]{Supreme Infernal Ancestry} The benefits of your \textit{infernal ancestry} ability reach their peak.
            \begin{itemize}
                \item Hellfire Conduit: The power bonus from your \textit{greater infernal ancestry} ability increases to \plus6.
                    In addition, the area affected by your \textit{abyssal hop} ability increases to a \medarea radius from your destination.
                \item Tempting Allure: The skill bonuses from your \textit{infernal ancestry} ability increase to \plus4.
                    In addition, your \textit{tempting domination} ability can dominate non-humanoid creatures.
                \item Unholy Might: The damage bonus from your \textit{infernal ancestry} ability increases to \plus2d.
                    In addition, your \textit{unholy surge} ability loses the \abilitytag{Attune} tag.
                    Instead, it lasts until you \glossterm{dismiss} it as a \glossterm{free action}.
            \end{itemize}

            \cf{Tif}[7]{Greater Abyssal Hop} When you use your \ability{abyssal hop} ability, you no longer require \glossterm{line of sight} or \glossterm{line of effect} to your destination.
            In addition, when you use it to teleport as a standard action, the range increases to \distrange.

        % \subsubsection{Basic Class Abilities}
        %     If you choose tiefling as your base class, you gain the following abilities.

        %     \cf{Tif}{Defenses}
        %     You gain the following bonuses to your \glossterm{defenses}: \plus3 Fortitude, \plus5 Reflex, \plus7 Mental.

        %     \cf{Tif}{Resources} You have the following \glossterm{resources}:
        %     \begin{itemize}
        %         \item Four \glossterm{attunement points}, which you can use to attune to items and abilities that affect you (see \pcref{Attunement Points}).
        %         \item A \plus2 bonus to your \glossterm{fatigue tolerance}, which makes it easier for you to use powerful abilities that fatigue you (see \pcref{Fatigue}).
        %         \item Two \glossterm{insight points}, which you can spend to gain additional abilities or proficiencies (see \pcref{Insight Points}).
        %         \item Five \glossterm{trained skills}, which you can spend to learn skills (see \pcref{Trained Skills}).
        %     \end{itemize}

        %     \cf{Tif}{Weapon Proficiencies} 
        %     You are proficient with simple weapons.

        %     \cf{Tif}{Armor Proficiencies} 
        %     You are proficient with light and medium armor.

        %     \cf{Tif}{Skills}
        %     You have the following \glossterm{class skills}:
        %     \begin{itemize}
        %         \item \subparhead{Strength} Climb, Jump.
        %         \item \subparhead{Dexterity} Balance, Sleight of Hand, Stealth.
        %         \item \subparhead{Intelligence} Craft, Disguise, Knowledge (arcana, planes), Linguistics
        %         \item \subparhead{Perception} Awareness, Social Insight.
        %         \item \subparhead{Other} Deception, Intimidate, Perform, Persuasion.
        %     \end{itemize}

\section{Classes}
    \subsection{Bard}
        A bard is a rogue with the ability to perform magical feats through music.
        It is unclear whether bards actually draw power from music in the same way that druids draw power from nature, or whether they simply channel their innate magical talent through music.
        The bard class functions like the rogue class, with the following exceptions:
        \begin{itemize}
            \item A bard cannot choose the \textit{assassin} archetype. However, the \textit{arcane magic} sorcerer archetype is considered to be part of their class, and they may choose that archetype without spending insight points to multiclass.
            \item A bard casts spells without \glossterm{somatic components}.
            \item A bard can only cast spells while sustaining a performance with the Perform skill. This performance can be either a mundane performance or a \textit{bardic performance} ability.
        \end{itemize}

    \subsection{Blighter}
        Blighter practice a strange inversion of druidic traditions.
        While druids venerate nature in all its forms, blighters dedicate their lives to the destruction of nature for its own sake.
        They rip power directly from the death of natural beings, using it to fuel their own warped version of nature magic.
        The blighter class functions like the druid class, with the following exceptions:
        \begin{itemize}
            \item Whenever a blighter rests, they automatically destroy nature and kill anything living around them.
                Plants wither and die, insects fall dead in the air, and so on.
                A ten minute rest destroys life in a radius equal to five feet times the blighter's highest rank in the blighter class (minimum 5 feet total).
                In general, Diminuitive or larger creatures and Medium or larger plants suffer no ill effects, though creatures may feel subtle pains.
                An eight hour rest destroys life in ten times that radius, and kills life one size category larger.
                Resting beyond that point does not increase the radius or severity of the effect.
                This destruction spreads out gradually throughout the resting period, and even a partially completed rest destroys some natural life.
            \item A blighter cannot choose the \textit{wildspeaker} archetype.
                However, the \textit{domain influence} cleric archetype is considered to be part of their class, and they may choose that archetype without spending insight points to multiclass.
                A blighter can only choose the Death, Destruction, and Evil domains.
            \item A blighter cannot gain access to the \sphere{verdamancy} mystic sphere by any means.
        \end{itemize}

    \subsection{Faebonder}
        A faebonder is a warlock who made their pact with a fae creature instead of a demon.
        The faebonder class functions like the warlock class, with the following exceptions:
        \begin{itemize}
            \item The magic source for the faebonder class is nature magic instead of pact magic.
                This changes the \glossterm{mystic spheres} a faebonder has access to and all other effects based on their source of magic.
                However, they still require both \glossterm{verbal components} and \glossterm{somatic components} to cast spells from the faebonder class (see \pcref{Casting Components}).
            \item A faebonder cannot choose the \textit{blessings of the abyss} archetype. However, the \textit{elementalist} druid archetype is considered to be part of their class, and they may choose that archetype without spending insight points to multiclass.
            \item Faebonders add Knowledge (nature) to their class skill list and remove Knowledge (planes).
        \end{itemize}

    \subsection{Favored Soul}
        A favored soul is a warlock who made their pact with a deity instead of a demon.
        This is an unusual arrangement, as deities would normally influence their clerics to achieve their aims.
        However, in special circumstances, a deity may want to empower a non-worshipper to influence mortal affairs.
        The favored soul class functions like the warlock class, with the following exceptions:
        \begin{itemize}
            \item The magic source for the favored soul class is divine magic instead of pact magic.
                This changes the \glossterm{mystic spheres} a favored soul has access to and all other effects based on their source of magic.
                However, they still require both \glossterm{verbal components} and \glossterm{somatic components} to cast spells from the favored soul class (see \pcref{Casting Components}).
            \item A favored soul cannot choose the \textit{blessings of the abyss} archetype. However, the \textit{domain influence} cleric archetype is considered to be part of their class, and they may choose that archetype without spending insight points to multiclass.
            \item Favored souls add Knowledge (religion) to their class skill list and remove Knowledge (planes).
        \end{itemize}

    \subsection{Shaman}
        A shaman, like a cleric, is a divine worshipper.
        However, while clerics worship powerful, well-established deities, shamans worship more primitive deities of lesser power.
        As a result, their divine powers are more limited and take different forms.
        Shamans are common among less civilized humanoid societies like bugbears.
        The shaman class functions like the cleric class, with the following exceptions:
        \begin{itemize}
            \item The magic source for the shaman class is nature magic instead of divine magic.
                This changes the \glossterm{mystic spheres} a shaman has access to and all other effects based on their source of magic.
            \item A shaman cannot choose the \textit{divine spell mastery} archetype. However, the \textit{elementalist} druid archetype is considered to be part of their class, and they may choose that archetype without spending insight points to multiclass.
            \item A shaman cannot gain access to more than two \textit{mystic spheres} from the magic source granted by the shaman class by any means.
            \item Shamans add Knowledge (nature) to their class skill list and remove Knowledge (planes).
        \end{itemize}

\section{Alternate Play Styles}

    \subsection{Being Surrounded}\label{Being Surrounded}
        Normally, exact positioning doesn't matter that much in combat.
        This makes it easier to play without a grid, or to just spend less time worrying about the details of everyone's positions on a grid.
        With this optional rule, you can make positioning more important in combat, increasing tactical depth for melee characters.
        This generally has the downside of making movement more complicated, however, as combatants try to surround others and avoid being surrounded themselves.

        If you play with this alternate rule, when you are being attacked by multiple foes at once, you are less able to defend yourself.
        If every space adjacent to you either contains an \glossterm{enemy} or is adjacent to an \glossterm{enemy}, you are surrounded.
        A creature that is surrounded takes a \minus2 penalty to its Armor and Reflex defenses.
        When determining whether you are surrounded, ignore any enemies that are sharing space with you, and ignore any enemies that are at least two size categories smaller than you.

        Any effect that makes a creature immune to being \partiallyunaware, such as the \spell{foresight} spell, also makes that creature immune to being surrounded.

    \subsection{Expanded Insight Points}
        Normally, \glossterm{insight points} can only be used to learn new special abilities from your class, or from a small number of feats.
        This alternate rule you to spend insight points to gain a wide variety of other proficiencies and benefits.
        This makes character creation more complicated, but it also allows you to personalize your character much more precisely.

        If you play with this alternate rule, increase the number of insight points that each class grants as part of its base class abilities by 1.
        In addition, you can spend insight points in any of the following ways.
        \begin{raggeditemize}
            \item You can spend an \glossterm{insight point} to gain an additional \glossterm{trained skill}.
            \item You can spend an \glossterm{insight point} to gain proficiency in an additional \glossterm{usage class} of armor (light, medium, or heavy).
                You must be proficient with light armor to become proficient with medium armor, and you must be proficient with medium armor to become proficient with heavy armor.
            \item You can spend an \glossterm{insight point} to gain proficiency in an additional \glossterm{weapon group}.
            \item You can spend two \glossterm{insight points} to gain proficiency with \glossterm{exotic weapons} from a single \glossterm{weapon group} you are already proficient with.
            \item You can spend an \glossterm{insight point} to learn two \glossterm{common languages} or one \glossterm{rare language} (see \pcref{Communication and Languages}).
        \end{raggeditemize}

    \subsection{Obscure Magic Items}\label{Obscure Magic Items}
        The base rules of Rise make it fairly easy to identify magic items.
        This keeps the pace of the game up when players find magic items frequently.
        However, you may choose to treat magic items as being more rare and mysterious.
        If you do, make the following changes:
        \begin{itemize}
            \item The \ability{identify item} ability from the Craft and Knowledge skills provides no information about how to use a magic item's properties or what they might be.
                It can still be used to identify whether or not an item is magical.
            \item The Knowledge (items) Knowledge skill is removed entirely.
            \item Magic items are more rare, and therefore more valuable.
                Calculate the prices for all magic items as if they were one rank higher than they actually are.
                Rank 7 magic items cannot be bought for any price - they are simply too rare.
            \item All spells with the \abilitytag{Attune} tag require an additional \glossterm{attunement point} to attune to.
                If magic items are hard to find and use, spellcasters gain a powerful benefit, since their personal attunement spells are still reliably available.
                This change ensures that spellcasters still gain a benefit from their personal access to magic, but they are not drastically more powerful than characters who depend on finding useful magic items.
        \end{itemize}

        You may also want to add complex or unintuitive activation conditions to magic items.
        For example, \mitem{boots of speed} may only function while hopping on one foot, or while you are not wearing socks.
        This can encourage players to experiment more with magic items to figure out how to use them.

    \subsubsection{Restricted Archetype Order}
        Normally, when a character in Rise levels up, they can freely choose which of their class archetypes they want to rank up (as long as they don't exceed their maximum rank).
        However, this means that most levels require making a choice that may be confusing for newer players.
        The process of leveling up can be simplified if each player chooses an order for their archetypes.

        With this variant, each character has a primary archetype, a secondary archetype, and a tertiary archetype.
        This choice is made at character creation.
        Whenever they increase their maximum rank, they increase their rank in their primary archetype.
        In their next level up, they increase their rank in their secondary archetype, and then finally their tertiary archetype.

    \subsection{Sleeping While Encumbered}
        Normally, characters can sleep in their armor without any penalty.
        This is unrealistic, but it can be time-consuming to make everyone track how their sleeping statistics differ from their waking statistics.
        Being ambushed while sleeping is very rare in most games, so it's generally not worth the hassle.
        However, if you want a more realistic game with more punishing night ambushes, you can use this alternate rule.

        If you play with this alternate rule, resting in armor is difficult.
        If you take a \glossterm{long rest} while you have any encumbrance, you finish your rest with a \glossterm{fatigue level} equal to the value of your encumbrance.
        In addition, only half the time you spend sleeping while you have encumbrance counts as sleep for the purpose of determining your fatigue (see \pcref{Sleep and Fatigue}).

    \subsection{Tap Out}
        With this optional rule, whenever you gain a vital wound, you can ``tap out'' to guarantee that you survive while taking your character out of the fight.
        If you tap out, you treat the result of the vital roll for that vital wound as a 10, regardless of any bonuses or penalties you would normally have to the vital roll.
        However, you fall unconscious immediately, and you cannot regain consciousness by any means until you take a \glossterm{short rest}.

        This optional rule significantly reduces the likelihood of character death, and makes fights less likely to impose long-term consequences on characters.
        However, it also makes vital wounds more likely to entirely knock characters out of a fight, which can increase the risk that the entire party is defeated.
