\chapter{Reference}\label{Reference}

\section{Ability Tags}\label{Ability Tags}

  \abilitytagdef{Acid} Acid abilities use physical acid to cause their effects.
  They are often more effective against objects.

  \abilitytagdef{Air} Air abilities tap into the elemental force of air.

  \abilitytagdef{Attune} Attune abilities require an \glossterm{attunement point} to maintain.
  For details, see \pcref{Attuned Abilities}.

  \abilitytagdef{Auditory} Auditory abilities use sound to cause their effects.
  Covering or plugging your ears makes you \impervious to Auditory attacks.
  Objects and \deafened creatures are immune to Auditory abilities.

  \abilitytagdef{Barrier} Barrier abilities create obstacles that block or punish passage through them.
  They create a powerful magical resonance around themselves that inhibits other barriers from being formed.
  % TODO: awkward wording
  Any part of a Barrier ability that would extend within 5 feet of an already active Barrier ability is not created, unless the new barrier simply extends an already existing wall along its length.

  \abilitytagdef{Brawling} Brawling abililities rely on brute force in hand to hand combat.
  Your accuracy with Brawling abilities is equal to half the sum of your level and Strength (see \pcref{Brawling Accuracy}).
  Many Brawling abilities are described in \pcref{Universal Combat Abilities}.

  \abilitytagdef{Clinch} You gain a \plus2 accuracy bonus with this ability against creatures that are \grappled by you or who are grappling you.

  \abilitytagdef{Cold} Cold abilities draw energy out of the environment to cause their effects.

  \abilitytagdef{Compulsion} Compulsion abilities forcibly alter a creature's actions, but do not necessarily affect its opinions or personality.
  They have no effect on objects or creatures without minds.

  \abilitytagdef{Creation} Creation abilities create permanent physical objects.
  Objects created with Creation abilities are identical to objects created through more mundane means.

  \abilitytagdef{Curse} Curse abilities lay supernatural curses on their targets.
  They cannot be \glossterm{dismissed}, but can be removed with the \spell{dispel curse} spell.

  \abilitytagdef{Detection}\label{Detection} Detection abilities reveal magical auras or information within an area.
  They can penetrate up to 1 foot of stone, 1 inch of common metal, a thin sheet of lead, or 3 feet of wood or dirt.
  For its ability to penetrate other materials, use the most similar substance from the list above.

  \abilitytagdef{Earth} Earth abilities tap into the elemental force of earth.

  \abilitytagdef{Electricity} Electricity abilities use electrical energy to cause their effects.
  Their effects typically create \glossterm{bright illumination} in a \smallarea radius that immediately ends when they are done.

  \abilitytagdef{Emotion} Emotion abilities alter a creature's opinons or personality, but do not necessarily affect their actions.
  They have no effect on objects or creatures without minds.

  \abilitytagdef{Fire} Fire abilities tap into the elemental force of fire.
  Their effects typically create \glossterm{bright illumination} in a \smallarea radius that lingers until the end of the current round.
  Damage from Fire abilities can set flammable objects on fire.

  \abilitytagdef{Impact} You get a \glossterm{glancing blow} with this ability when you would miss by 5 or less (see \pcref{Glancing Blows}).
  This does not allow you to get glancing blows if that would normally be impossible for you.

  \abilitytagdef{Keen} You gain a \plus2 \glossterm{accuracy} bonus with this ability for the purpose of determining whether you get a \glossterm{critical hit}.

  \abilitytagdef{Manifestation} Manifestation abilities create temporary constructs formed from raw magical energy.
  Objects and creatures created with Manifestation abilities seem real on the surface, but they have no internal structure.
  A \glossterm{difficulty value} 5 Knowledge (items) or relevant Craft check reveals that a manifested item is a magical fabrication rather than an ordinary item.
  When an object or creature created by a Manifestation ability is destroyed or killed, or when the duration of the ability that created it ends, it disappears without a trace.

  \abilitytagdef{Ritual} Ritual abilities have a number of shared properties.
  For details, see \pcref{Spell and Ritual Mechanics}.

  \abilitytagdef{Scrying} Scrying abilities create one or more invisible magical sensors that send you information.
  Unless otherwise noted, the sensor created has the same powers of sensory acuity that you possess.
  This includes the effect of any abilities which target you personally, such as spells to increase your visual acuity, but not abilities which affect an area around you.
  However, the sensor is treated as a separate, independent sensory organ, and it functions normally even if you have been blinded, deafened, or otherwise suffered sensory impairment.
  \par The sensor can be dismissed as if it were an active spell.
  You cannot create a sensor in a location with lead sheeting between you and the location, and you sense that the effect is blocked in this way.

  \abilitytagdef{Size-Based} Size-Based abilities are limited based on your own size.
  They have no effect on creatures or objects that are two or more size categories larger than you.

  \abilitytagdef{Speech} Speech abilities use words to achieve their ends.
  You must specify a language when using a Speech effect, and the language must be one you know (or have memorized the correct words to say). They have no effect on objects or creatures that do not understand the chosen language.

  \abilitytagdef{Spell} Spell abilities have a number of shared properties.
  For details, see \pcref{Spell and Ritual Mechanics}.

  \abilitytagdef{Subdual} All damage dealt by this ability is \glossterm{subdual damage} (see \pcref{Subdual Damage}).

  \abilitytagdef{Subtle} Subtle abilities have no visual or otherwise perceivable manifestation.
  Creatures successfully affected by Subtle abilities do not generally know that they are being influenced.
  However, a creature that successfully resists a Subtle ability can generally notice that it resisted a special effect of some kind, just like a non-Subtle ability.
  You can notice the effects of a Subtle ability on yourself with the Awareness skill (see \pcref{Awareness}).

  \abilitytagdef{Sustain} Sustain abilities require an action to maintain.
  The tag includes an action type, such as (minor), which indicates the type of action required to sustain the ability.
  If it also includes ``attuneable'', you can choose to \glossterm{attune} to the effect instead of sustaining it every round.
  For details, see \pcref{Sustained Abilities}.

  \abilitytagdef{Swift} Swift abilities take effect before non-Swift abilities used during the same phase.
  For details, see \pcref{Swift Abilities}.

  \abilitytagdef{Trap} Trap abilities create triggered effects that punish trespassing.
  They create a powerful magical resonance around themselves that inhibits other barriers from being formed.
  Any part of a Trap ability that would extend within 15 feet of an already active Trap ability is not created.

  \abilitytagdef{Visual} Visual abilities use visible objects or forces to cause their effects.
  Visual abilities do not determin their \glossterm{line of effect} like other abilities.
  Instead, they use their \glossterm{line of sight} to also determine their \glossterm{line of effect}.
  For example, a Visual effect is not blocked by glass, but it is blocked by thick smoke.

  Creatures and objects that cannot see a Visual effect are immune to it.
  Simply closing your eyes only makes you \impervious to Visual effects.
  Completely covering your eyes, such as with your hands or with a blindfold, makes you immune.
  Special vision abilities that replace normal vision entirely, such as \trait{blindsight} and \trait{tremorsense}, are unable to perceive Visual effects.

  \abilitytagdef{Water} Water abilities tap into the elemental force of water.

  \newpage
\section{Circumstances and Debuffs}\label{Circumstances and Debuffs}

  \debuffdef{blinded} A blinded creature cannot use its eyes.
  Assuming it does not have other methods of sight, it automatically fails at actions which depend on vision, including simply seeing the locations of other objects and creatures.
  This makes it at least \partiallyunaware of all attacks, and it may be fully \unaware as normal.

  \buffdef{braced} An braced creature gains a \plus2 bonus to all defenses.

  \debuffdef{charmed} A charmed creature is mentally influenced to like another creature.
  It always sees the words and actions of the creature that charmed it in the most favorable way, as a close friend or trusted ally.
  A charmed creature cannot be controlled like an automaton, but can be persuaded to take particular actions with the Persuasion skill (see \pcref{Persuasion}).
  It treats the creature that charmed it as a friend (a \plus10 relationship modifier) for the purpose of Persuasion checks.

  Any act by the charming creature or by creatures that appear to be its allies that threatens or harms the charmed person breaks the effect.
  Harming a charmed creature is not limited to dealing it damage, but also includes causing it significant subjective discomfort.
  An observant creature may interpret overt threats to its allies as a threat to itself.

  \debuffdef{confused} A confused creature takes a \minus2 penalty to all defenses and is unable to independently control its actions.
  This penalty does not stack with the \stunned effect.
  When a creature becomes confused, and at the beginning of each round, it randomly decides to have one of two behaviors that round: attack its \glossterm{enemies}, or protect itself and its \glossterm{allies} without attacking.
  Within those constraints, it can freely choose its actions.
  If it can't carry out the indicated action, it does nothing but babble incoherently.

  % Kind of an awkward hack
  A confused creature automatically stops being confused after ten minutes, even if it was unable to rest due to its confusion.

  \debuffdef{dazzled} A dazzled creature has difficulty seeing.
  It loses the benefits of special vision traits that require eyes, such as \trait{darkvision} and \trait{low-light vision}.
  In addition, it treats everything as if it had \glossterm{concealment}.
  Among other effects, this gives its \glossterm{targeted} attacks a 20\% \glossterm{miss chance}.

  \debuffdef{deafened} A deafened creature cannot hear.
  It automatically fails at actions which depend on hearing, and is immune to Auditory abilities.
  In addition, it has a 20\% failure chance when casting any spell with verbal components.

  \debuffdef{dominated}[dominate] A dominated creature is mentally compelled to obey another creature.
  It obeys the commands of the creature of the dominated it unquestioningly, as an automaton.
  If it does not understand the language of the creature that dominated it, it still attempts to obey as much as possible, and simple commmands (such as ``attack'' or ``follow'') can usually be communicated successfully.
  A creature that is both dominated and \confused obeys its dominated orders, ignoring the confusion.

  \buffdef{empowered} An empowered creature deals \glossterm{extra damage} with all of its damaging abilities equal to its \glossterm{character rank}.
  Being empowered does not improve your \glossterm{reactive attacks}.

  \debuffdef{enraged} An enraged creature must spend a \glossterm{standard action} to make an attack during each round.
  It can still take other actions normally.
  The creature's attacks do not have to specifically target other creatures, so it can attack inanimate objects.
  If it is unable to take standard actions, such as if it is unconscious, it suffers no extra penalty.

  % Kind of an awkward hack
  An enraged creature automatically stops being enraged after ten minutes, even if it was unable to rest due to its rage.

  \buffdef{focused} A focused creature can \glossterm{reroll} attack rolls once and keep the higher result.
  This only affects attack rolls, not \glossterm{miss chances} or \glossterm{checks}.
  Being focused does not improve your \glossterm{reactive attacks}.

  \buffdef{fortified} A fortified creature gains a \plus2 bonus to its Brawn, Fortitude, and Mental defenses.

  \debuffdef{frightened} A frightened creature takes a \minus2 penalty to its Mental defense.
  In addition, it takes a \minus2 penalty to \glossterm{accuracy} against the source of its fear.
  This does not stack with the \panicked effect.
  If the source of a frightened creature's fear is \glossterm{defeated} or otherwise stops being relevant to the creature, this effect is broken.
  % TODO: why is this necessary? can this be removed? it seems surprising
  Being frightened is always an \abilitytag{Emotion} effect, even if it is caused by an ability that does not have that tag.

  \debuffdef{goaded} A goaded creature takes a \minus2 penalty to \glossterm{accuracy} against creatures other than the creature that goaded it it as long as it is within \rngmed range of of that creature.
  If the goading creature is \glossterm{defeated}, this effect is broken.
  If a creature is goaded by multiple different creatures simultaneously, it suffers the accuracy penalty on all of its attacks.
  Being goaded is always an \abilitytag{Emotion} effect, even if it is caused by an ability that does not have that tag.

  \debuffdef{grappled} A grappled creature is wrestling or in some other form of hand-to-hand struggle with at least one other creature.
  While grappled, you suffer certain penalties and restrictions, as described below.
  \begin{raggeditemize}
    \item One of your hands cannot be used for any purposes.
      This prevents humanoid creatures from taking any actions which would require having two free hands, such as attacking with \weapontag{Heavy} weapons.
      This does not affect creatures without hands.
    \item You take a \minus2 penalty to Armor and Reflex defenses.
    \item You cannot move unless you \glossterm{push} all creatures grappling you, such as with the \ability{shove} ability (see \pcref{Shove}).
      In addition, you cannot \glossterm{push} a creature grappling with you so it stops being adjacent to you.
      However, you can use the \ability{shove} ability to affect creatures you are grappling with as a \glossterm{move action} instead of as a standard action.
    \item You can use the \ability{escape grapple} and \ability{maintain grapple} abilities to stop or continue grappling (see \pcref{Universal Combat Abilities}).
  \end{raggeditemize}

  \debuffdef{helpless} A helpless creature is unable to move or defend itself.
  It takes a \minus8 penalty to its Armor, Brawn, and Reflex defenses.
  These penalties do not stack with the penalties for being unaware (partially or fully) or unsteady.
  Paralyzed, bound, and unconscious creatures are helpless.

  \buffdef{honed} A honed creature gains a \plus4 accuracy bonus for the purpose of determining whether it gets a \glossterm{critical hit}.
  As normal, a miss or glancing blow cannot also get a critical hit.

  \buffdef{maximized} A maximized creature deals maximum damage whenever it deals damage.
  This only affects damage caused directly by the creature, just like \glossterm{extra damage}.
  For example, a maximized creature that \glossterm{flings} a creature does not maximize the impact damage from the fling, if any.
  Being maximized does not improve your \glossterm{reactive attacks}.

  \debuffdef{panicked} A panicked creature takes a \minus4 penalty to its Mental defense.
  In addition, it is unable to make any attacks that include the source of its fear as a target.
  The penalty from this effect does not stack with the \frightened or \panicked effects.
  If the source of a panicked creature's fear is \glossterm{defeated}, this effect is broken.
  Being panicked is always an \abilitytag{Emotion} effect, even if it is caused by an ability that does not have that tag.

  \debuffdef{paralyzed} A paralyzed creature is unable to take physical actions. It is \helpless, but can take purely mental actions. This can cause flying creatures to crash, swimming creatures to drown, and so on. Any creature can move through a space occupied by a paralyzed creature without slowing down, and creatures can stand in a square with a paralyzed creature without \squeezing.

  \debuffdef{partially unaware} A creature that is partially unaware knows that something is nearby, but is missing information about the exact location or nature of the creature, object, or attack it is partially unaware of.
  Creatures take a \minus2 penalty to all defenses against attacks that they are partially unaware of.
  They have a 50\% miss chance with \glossterm{targeted} atacks against creatures and objects that they are partially unaware of, and they can only attempt to target creatures and objects that they know the location of.

  These penalties do not stack with the penalties for being \unaware or \unsteady.
  For details, see \pcref{Awareness and Surprise}.

  \buffdef{primed} A primed creature automatically \glossterm{explodes} on the first die when making attack rolls (see \pcref{Exploding Dice}).
  This only affects attack rolls, not \glossterm{miss chances} or \glossterm{checks}.
  Since a 10 normally explodes, this has no effect if you roll a 10.
  Being primed does not improve your \glossterm{reactive attacks}.

  \debuffdef{prone} A prone creature is lying on the ground, rather than standing normally.
  It takes a \minus2 penalty to Armor and Reflex defenses.
  However, it gains a \plus4 bonus to all defenses against ranged \glossterm{strikes} as long as the attacker is not adjacent to it.
  It moves at half of its normal speed.
  Creatures that are not on the ground, such as flying or gliding creatures, are immune to being knocked prone.

  If a creature becomes prone while in a precarious situation, such as on a narrow ledge, it may fall.
  Mounted creatures that are knocked prone fall off their mounts.
  Creatures cannot glide or fly while prone.

  A creature can stand up from being prone as part of a \glossterm{movement} using one of their move speeds.
  This generally requires one \glossterm{free hand}.
  Standing up from a prone position costs half of the creature's speed during that movement.

  \buffdef{shielded} A shielded creature gains a \plus2 bonus to its Armor and Reflex defenses.

  \debuffdef{slowed} A slowed creature has a \minus10 foot penalty to its \glossterm{movement speed} and takes a \minus2 penalty to its Armor and Reflex defenses.

  \debuffdef{squeezing} A squeezing creature is trying to move though an area too small for it to fight in normally.
  While squeezing, a creature moves at half speed and takes a \minus2 penalty to its Armor and Reflex defenses.
  For details, see \pcref{Squeezing}.

  \buffdef{steeled} A steeled creature cannot receive critical hits.
  If an attack against it would be a critical hit, the attack is a normal hit instead.

  \debuffdef{stunned} A stunned creature takes a \minus2 penalty to all defenses.
  This does not stack with the \confused effect.

  \debuffdef{unaware} A creature that is unaware makes no attempt to defend itself.
  Creatures take a \minus4 penalty to all defenses against attacks that they are unaware of.
  They are completely unable to use \glossterm{targeted} abilities against creatures and objects that they are unaware of.

  These penalties do not stack with the penalties for being \partiallyunaware or \unsteady.
  For details, see \pcref{Awareness and Surprise}.

  \debuffdef{unconscious} While you are unconscious, you are \helpless, \unaware, and completely unable to take any actions.
  Some sensory abilities, such as the Awareness skill, can be used reactively while you are asleep, but not while you are forcibly knocked unconscious.

  \debuffdef{unsteady} A creature that is unsteady takes a \minus2 penalty to its \glossterm{accuracy} and Armor, Brawn, and Reflex defenses.
  These penalties do not stack with the penalties for being \partiallyunaware or \unaware.
  Creatures that are \glossterm{midair} and not native to the air, climbing without a \glossterm{climb speed}, or in liquid without a \glossterm{swim speed} are unsteady.

  \newpage
\section{Traits}\label{Traits}
  % TODO: should there be subgroups within this section, or should it be strictly alphabetical?
  % A possible subgroup would be "sensory traits".

  \traitdef{Amphibious}{amphibious}
  An amphibious creature can hold its breath ten times longer than normal (see \pcref{Endurance}).

  \traitdef{Blindsense}{blindsense}
  A creature with blindsense can sense the location of everything in its surroundings.
  It does not need to use its eyes to gain this benefit.
  This ability works regardless of concealment, invisibility, or light levels.
  Blindsense always has a specific range limit, and provides no benefit beyond that range.
  However, it only grants knowledge of location, not actual sight, so does not mitigate any \glossterm{miss chances} that would apply.
  It also does not mitigate \glossterm{cover} or otherwise allow sensing through objects that block \glossterm{line of effect}.
  Blindsense does not require using eyes.

  \traitdef{Blindsight}{blindsight}
  A creature with blindsight can perceive its surroundings perfectly regardless of concealment, invisibility, or light levels.
  It does not need to use its eyes to gain this benefit.
  This allows the creature to ignore all \glossterm{miss chances} caused by those effects.
  Blindsight always has a specific range limit, and provides no benefit beyond that range.
  It also does not mitigate \glossterm{cover} or otherwise allow sensing through objects that block \glossterm{line of effect}.
  Blindsight does not require using eyes.

  \traitdef{Darkvision}{darkvision}
  A creature with darkvision can see perfectly in both complete darkness and \glossterm{dim illumination} just like a human does in \glossterm{bright illumination}.
  Darkvision always has a specific range limit, and provides no benefit beyond that range.
  As long as a creature with darkvision is in \glossterm{bright illumination} or \glossterm{brilliant illumination}, their darkvision stops working.
  The darkvision \glossterm{briefly} stays disabled even after they leave the lit area.
  Darkvision requires using eyes, and is disabled while you are \dazzled.

  \traitdef{Floating}{floating}
  A floating creature is only affected weakly by gravity.
  It does not fall while it is able to use a \glossterm{fly speed}, and it is unable to \glossterm{glide}.
  This does not increase its \glossterm{height limit} for any \glossterm{fly speed} it may have.
  A floating creature above the height limit for its fly speed falls down until its fly speed becomes functional again, which does not cause it to take \glossterm{falling damage}.
  Some creatures can freely choose whether they float, potentially allowing them to glide.

  \traitdef{Impervious}{impervious}
  % Are there any other things you can be impervious to?
  A creature can be impervious to a particular ability tag.
  Creatures gain a \plus4 bonus to all defenses against attacks that they are impervious to.
  In addition, they take no damage if the attack misses or gets a \glossterm{glancing blow}, even if the attack would normally deal half damage on a miss.
  If an attack has multiple tags, a creature is impervious to the attack if it is impervious to any of the attack's tags.
  % There shouldn't be any conditional tags like this
  % For attacks with random effects, such as the \spell{chromatic orb} spell, determine the random effect before determining if the creature is impervious.

  \traitdef{Immune}{immune}
  A creature that is immune to an attack is completely unaffected by it.
  Creatures and objects can be immune to specific ability tag or debuffs.
  It is also possible to be immune to more specific effects, like being \grappled or gaining \glossterm{conditions}.

  If an attack has multiple tags, a creature is immune to the attack if it is immune to any of the attack's tags.
  However, being immune to a particular effect of an attack does not make you immune to the rest of the attack's effects.
  For example, if you are immune to being \stunned, you still take full damage from an attack that deals damage and stuns you.

  \traitdef{Incorporeal}{incorporeal}
  An incorporeal creature does not have a physical body.
  It is both \trait{intangible} and \trait{floating}.
  In addition, it can enter or pass through solid objects.
  It must remain adjacent to the object's exterior at all times.

  While an incorporeal creature is inside of an object, it can choose whether it is completely enveloped or partially contained.
  If it is completely enveloped, it does not have \glossterm{line of sight} or \glossterm{line of effect} outside of the object, limiting its ability to attack.
  If it is partially contained, it has \glossterm{cover}, but can otherwise attack and be attacked normally.

  \traitdef{Intangible}{intangible}
  An intangible creature has a physical body that is gaseous or otherwise amorphous.
  It cannot be \glossterm{touched}, and is never considered to be \squeezing.
  % Should this reference Armor defense? What abilities target Armor defense that should affect a ghost?
  % Answer: Smite, spellblade, ki monk, and so on. But it would be useful to find a way to exclude Polymorph.
  It is \trait{immune} to \glossterm{mundane}, \atCreation, and \atManifestation abilities.
  It gains a \plus5 bonus to Stealth and ignores the effects of abilities that only work if it has a corporeal body.
  This includes all \abilitytag{Brawling} abilities, \glossterm{difficult terrain}, being detected by \trait{tremorsense}, setting off pressure plates, and so on.
  It can move freely through spaces occupied by enemies, but cannot pass through solid objects.

  Many intangible creatures have no Strength attribute.
  If an intangible creature has a Strength attribute, it has some ability to manipulate the physical world despite being intangible.
  Unless otherwise noted, an intangible creature with a Strength attribute may selectively choose whether it wants to interact with physical objects.

  \traitdef{Invisible}{invisible}
  An invisible creature or object cannot be seen with light.
  Creatures unable to see an invisible creature are at least \partiallyunaware of its attacks, and they can be fully \unaware as normal depending on their level of awareness.
  Attackers suffer a 50\% \glossterm{miss chance} with \glossterm{targeted} attacks even if they know the location of the invisible creature.
  See \pcref{Awareness}, and \pcref{Stealth}, for how to identify invisible creatures.

  % This name is awkward, so creatures aren't specifically called out as being 
  \traitdef{Legless}{legless}
  A legless creature has no legs.
  Legless creatures cannot jump, but they are immune to being \prone.

  \traitdef{Lifesense}{lifesense}
  Lifesense functions like \trait{blindsense}, except that it only grants knowledge of the location of living things.
  This includes both creatures and non-creatures, such as plants.

  \traitdef{Lifesight}{lifesight}
  Lifesight functions like \trait{blindsight}, except that it can only see living things.
  This includes both creatures and non-creatures, such as plants.

  \traitdef{Low-light Vision}{low-light vision}
  A creature with low-light vision can see perfectly in \glossterm{dim illumination}, just like a human does in \glossterm{bright illumination}.
  This provides no benefit in areas of complete darkness.
  Low-light vision requires using eyes, and is disabled while you are \dazzled.

  \traitdef{Mindless}{mindless}
  A mindless creature lacks a normal mind.
  It does not have Intelligence or Willpower attributes, and has no Mental defense.
  Any attacks against it that would normally use its Mental defense use its Fortitude defense instead.
  It uses its Strength to determine its \glossterm{magical power} instead of its Willpower.

  Unlike a \trait{simple-minded} creature, a mindless creature cannot normally be given instructions or communicated with in any meaningful way.
  They still act autonomously, but in a way that is entirely instinctual, without the ability to form thoughts or plan ahead.
  Some magic can influence or control mindless creatures, but that magic works on different principles than ordinary mind-influencing magic.
  A mindless creature has no soul, so if it dies, it cannot be resurrected.
  Mindless creatures are immune to \abilitytag{Compulsion} and \abilitytag{Emotion} abilities.

  \traitdef{Multipedal}{multipedal}
  A multipedal creature typically uses more than four legs to move.
  Multipedal creatures gain a \plus10 foot bonus to their \glossterm{movement speed} and a \plus5 bonus to the Balance skill.

  \traitdef{Quadrupedal}{quadrupedal}
  A quadrupedal creature typically uses four legs to move.
  Quadrupedal creatures gain a \plus10 foot bonus to their \glossterm{movement speed}.

  \traitdef{Scent}{scent}
  A creature with the scent ability has an unusually good sense of smell.
  It gains a \plus10 bonus to scent-based Awareness checks.
  In addition, it is able to follow scent-based tracks (see \pcref{Tracking}).

  \traitdef{Simple-Minded}{simple-minded}
  A simple-minded creature has an incompletely functioning mind.
  It can follow simple instructions, but is not fully \glossterm{sentient} or capable of complex reasoning.
  It has no soul, so if it dies, it cannot be resurrected.
  Simple-minded creatures are immune to \abilitytag{Emotion} abilities.
  However, they are \vulnerable to \abilitytag{Compulsion} attacks.

  \traitdef{Telepathy}{telepathy}
  A creature with telepathy has the ability to mentally communicate with other nearby creatures.
  All telepathy abilities have a defined \glossterm{range}.
  Unless otherwise specified, a telepathic creature can only communicate with one creature at a time.

  As a \glossterm{free action}, a telepathic creature can open a telepathic communication channel with one creature it sees within the range of its telepathy ability.
  That channel remains open until the telepath \glossterm{dismisses} it.
  The target does not have to be willing to receive telepathic communication in this way.
  While this channel is open, the telepathic creature can cause the target to ``hear'' the telepathic creature's voice inside the target's head.
  If the target attempts to mentally reply while the channel is open, the telepathic creature can similarly ``hear'' the reply in its head as if the target was speaking.
  This does not generally grant the ability to detect any other thoughts, though exceptionally stupid targets may accidentally broadcast their private thoughts.

  Telepathic communication uses words, so it still requires a shared language to be intelligible, even though the words are only imagined.
  A telepathic creature may attempt to telepathically communicate with creatures without a language, though this is generally unproductive.
  A skilled telepath can customize the mental ``voice'' it projects in the same way that a creature can attempt to disguise or alter its voice when speaking.

  \traitdef{Tremorsense}{tremorsense}
  Tremorsense functions like \trait{blindsense}, except that it requires an uninterrupted path through solid objects instead of \glossterm{line of effect}.
  This makes it incapable of sensing flying creatures, but it ignores \glossterm{cover} and can even sense through solid obstacles that are no more than half a foot thick.

  \traitdef{Tremorsight}{tremorsight}
  Tremorsense functions like \trait{blindsight}, except that it requires an uninterrupted path through solid objects instead of \glossterm{line of effect}.
  This makes it incapable of seeing flying creatures, but it ignores \glossterm{cover} and can even see through solid obstacles that are no more than half a foot thick.

  \traitdef{Vulnerable}{vulnerable}
  % Are there any other kinds of vulnerabilities besides tags and silver / cold iron?
  A creature can be vulnerable to a particular ability tag or weapon special material.
  It takes a \minus4 penalty to all defenses against attacks with that tag, or attacks using weapons that it is vulnerable to.
  If a creature would be both vulnerable and \impervious or immune to an ability, it is treated as vulnerable and not impervious or immune.
  % For attacks with random effects, such as the \ability{sudden entropy} ability from the Entropist feat, determine the random effect before determining if the creature is vulnerable.

\newpage
\section{Creature Tags}\label{Creature Tags}

  Creatures can have tags.
  Unlike traits, which are defined above, creature tags are intrinsic properties of a creature that cannot be granted or removed temporarily by any means.
  Each creature tag has a defined narrative implication, and many creature tags also affect a creature's abilities.
  If a creature tag has combat-relevant effects, those effects will be included in its main description.
  However, less common narrative effects of creature tags are only defined fully here.

  If one tag always comes with other tags, only the primary tag is listed in a creature's description.
  For example, all creatures with the \creaturetag{Dragon} tag also have the \creaturetag{Immortal} tag, but a monster's description would only list the \creaturetag{Dragon} tag to save space.

  By default, a creature with no tags is assumed to be a \glossterm{living} creature with a body, mind, and soul.
  It needs to eat, drink, and sleep, and it dies of old age eventually.

  \creaturetagdef{Aberration}{aberration}
  Aberrations have souls that are owned by the Precursors in the Eternal Void.
  When an aberration dies, its soul does not splinter, and neither Nature nor ordinary deities can claim any reward.
  Instead, the soul slowly drifts through the Astral Plane for centuries until it finally returns to the Eternal Void.

  \creaturetagdef{Animal}{animal}
  Animals are simple, \glossterm{mundane} creatures.
  They cannot have \magical abilities, and cannot have an Intelligence above \minus8.
  The Creature Handling skill is more effective on animals, as indicated in that skill's description.

  % TODO: should constructs regain hit points when they rest?
  \creaturetagdef{Construct}{construct} A construct is a creature that is made of nonsapient matter.
  Its inanimate body is given a semblance of life and sentience by some form of magic.
  Like other creatures, they can move and follow instructions.
  However, they lack agency and cannot make the independent decisions.
  They are \trait{mindless}, making them immune to \abilitytag{Compulsion} and \abilitytag{Emotion} effects.
  Constructs lack a soul and cannot be resurrected by any means if they are destroyed.

  Constructs are considered to be both creatures and objects, and are affected by abilities which affect either.
  They are not alive, and they do not need to eat, drink, or sleep.
  They are always considered to be \glossterm{attended} by themselves, so they are never affected by abilities that only affect unattended objects, even while unconscious.

  Constructs are not affected by the Medicine skill, and do not normally remove \glossterm{vital wounds} when they take a \glossterm{long rest}.
  Instead, their vital wounds must be repaired manually.
  This functions like like \ability{accelerate recovery} ability from the Medicine skill, except that it uses an appropriate Craft skill and raw materials appropriate to the construct's construction.

  \creaturetagdef{Dragon}{dragon} A dragon is a powerful magical creature with four legs and two wings.
  Dragons are \creaturetag{immortal} creatures that can feed on magic and valuable metals, including gold. 
  They are known for their vast hoards of wealth.
  Since dragons never stop growing in size or strength, elder dragons are some of the most powerful creatures in existence.

  % TODO: What makes fey unique? Short lifespan?
  \creaturetagdef{Fey}{fey}
  Fey are \magical creatures that are vulnerable to cold iron weapons.
  They have a natural connection to Discord, the plane of chaos, but are not planeforged.

  \creaturetagdef{Humanoid}{humanoid}
  A humanoid is a Medium creature with two arms and two legs.
  Humanoids can have magical abilities, but are usually not intrinsically magical.
  Most civilized creatures and societies are predominantly humanoid.

  \creaturetagdef{Immortal}{immortal}
  An immortal creature can never die of old age.
  Most immortal creatures can still be killed and need to take actions to sustain their life, such as eating and drinking.
  Some immortal creatures completely cease aging once they reach maturity, while others continue to grow or change over time without limit.

  \creaturetagdef{Indwelt}{indwelt}
  An indwelt is a creature that is made of nonsapient matter.
  Its inanimate body was awakened to life by connection to an external soul.
  Unlike constructs, indwelt have agency and true intelligence.

  The soul of an indwelt has no connection to the matter that composes its body.
  This contrasts with undead, which always maintain a connection between a body and its original soul.
  As a result, an indwelt's connection to its physical body is weak.
  If an indwelt is killed, it can be resurrected, but its previous body is not considered its corpse in the same way that a human's dead body is.
  This means it cannot be resurrected by magic that uses the corpse of the deceased creature.

  Indwelt are considered to be both creatures and objects, and are affected by abilities which affect either.
  They are alive if their base matter is alive, but not if their base matter is dead or inorganic.
  If they are alive, they need to eat, drink, and sleep.
  They are always considered to be \glossterm{attended} by themselves, so they are never affected by abilities that only affect unattended objects, even while unconscious.

  \creaturetagdef{Planebound}{planebound}
  A planebound creature cannot leave a specific plane by any means.
  Effects which would cause it to leave that plane have no effect on that creature.
  Most planebound creatures are \creaturetag{planeforged} creatures from that plane.

  \creaturetagdef{Planeforged}{planeforged}
  A planeforged is a creature that is made from the concentrated planar essence of only one or two planes.
  It is not \glossterm{living}.
  When a planeforged dies, its soul generally does not splinter.
  Instead, the soul returns intact to its home plane, where it may reform a new body if it still has the necessary willpower.
  Some planeforged are also \creaturetag{planebound}, including the bodies of dead souls in their afterlife.

  \creaturetagdef{Undead}{undead}
  Undead are creatures that are made from the corpse or spirit of a dead creature.
  They are animated by some part of the soul of the original creature.

  Although undead are not \glossterm{living}, they are affected in unusual ways by effects that directly manipulate life energy.
  They can be targeted as if they were living \glossterm{allies} by \magical effects that would cause living creatures to regain \glossterm{hit points}.
  Whenever they would regain \glossterm{hit points} from an ability that normally only affects living creatures, they instead take damage equal to the hit points that they would have regained, ignoring any hit point maximum the ability would normally have.

  Any other effects beyond simple hit point recovery are ignored.
  For example, if a cleric uses their \ability{divine aid} ability to heal an undead creature, the undead would take damage, but it would not gain any bonus to its defenses.

  % TODO: should undead heal from vivimancy effects? Is it worth introducing a new damage type just for that?
  % Similarly, undead are always considered living creatures for the purpose of effects that would normally deal damage to living creatures.
  % Whenever they would take damage from an ability that normally only affects living creatures, they instead regain hit points equal to the damage that they would have taken. 
  % This ability does not have the \abilitytag{Swift} tag, so it resolves after incoming attacks during the current phase.
  % Any other effects beyond simple damage are ignored.
