\newpage
\section{Relics}\label{Relics}

  Most items can be bought, sold, and crafted.
  Relics are rare items that cannot be acquired normally.
  They are intended to be interesting items that can make an individual campaign more fun when included, but which would cause problems if they functioned like ordinary magic items.
  Unless you specifically say that a relic item exists, it does not exist in the world at all.
  Normally, they cannot be crafted, purchased in towns, or chosen with any special character abilities that can normally choose any magic item property.

  In general, relics can provide more narrative power than a magic item can typically provide.
  If they are too common, they can undercut what makes individual characters feel unique.
  Mid to high level characters can become extremely wealthy.
  If they could easily buy all of the low rank relics, they would have a vast amount of magical control over small aspects of the universe.
  This would be too complicated to manage easily.
  In addition, characters with their own unique abilities that grant them narrative power would feel less special.

  Categorizing items as relics can also solve world-building issues.
  Sending stones are a useful item, and it can be helpful for the story for players to have the ability to talk to distant friendly NPCs.
  However, sending stones would also radically reshape the way long-distance communication would work in the world.
  It would be reasonable for rulers of major countries to have an array of sending stones that would allow them to communicate with neighboring kingdoms directly, or to directly command their generals in the field.
  This type of instant communication is not really genre-appropriate for a classic fantasy setting, and pushes the setting in a much more modern direction.
  If sending stones are so rare and valuable that this becomes impractical, players would be strongly incentivized to sell them in exchange for vast wealth, since the value they provide to players is more of a small convenience than a necessity.
  All of this can be solved by making sending stones relics.
  That means they can be safely given to players without disrupting the broader fantasy world.

  Relics are also a good way to categorize any new items you want to create in the universe.
  If you want to give the players a custom item that you created, but you don't want to deal with the long-term implications of making that a normal item that exists in the universe, you can call it a relic.

  \begin{longcolumn}
    \input{generated/relics_table.tex}
  \end{longcolumn}

  \input{generated/relics.tex}
