\chapter{Attributes}
A character has six attributes: Strength (Str), Dexterity (Dex), Constitution (Con), Intelligence (Int), Wisdom (Wis), and Charisma (Cha).

\section{Attribute Scores}
Each attribute, after changes made because of race, has an attribute score. This represents your character's raw talent in that area. A 0 in an attribute represents average human capacity. That doesn't mean that every commoner has a 0 in every attribute; not everyone is average, after all.

When your character makes a d20 roll to do something based on an attribute, you usually add the character's score in the attribute to the roll. You also apply your character's attribute to some other values which are related to d20 rolls. For example, you add your Dexterity to your armor class. A positive attribute is called a bonus, and a negative attribute is called a penalty.

Your attribute score can never exceed 10. Magical and extraordinary creatures can exceed this limitation, but ordinary mortals are limited by their physical form.

\subsection{Attribute Values}
Each attribute also has an associated attribute value. The value is the number you use when your character does something related to the attribute which isn't related to a d20 roll. For example, your Intelligence value determines the number of languages your character knows.

Your attribute value is equal to your attribute score divided by two (rounded down).

\subsection{Attributes and Spellcasters}
The attribute that governs spellcasting depends on what type of spellcaster your character is: Intelligence for wizards; Wisdom for druids; or Charisma for paladins, sorcerers, bards, and most clerics. That attribute is called your casting attribute. In order to cast spells of a given level, a spellcaster must have a casting attribute score that is at least equal to the spell's level. For example, a sorcerer with a 4 Charisma can cast up to 4th level spells. In addition to having a high attribute score, a spellcaster must be of high enough class level to be able to cast spells of a given spell level. (See the class descriptions for details.)

\section{The Attributes}
Each attribute partially describes your character and affects some of his or her actions.

\subsection{Strength (Str)}
Strength measures your character's muscle and physical power. This attribute is especially important for fighters, barbarians, paladins, rangers, and monks because it helps them prevail in combat. Strength also limits the amount of equipment your character can carry.

You apply your character's Strength modifier to:
\begin{itemize*}
\item Attack rolls with medium and heavy weapons, including medium and heavy thrown weapons.
\item Climb, Jump, and Swim checks. These are the skills that have Strength as their key attribute.
\item Strength checks (for breaking down doors and the like).
\item Half your Strength modifier (rounded down) is applied to Fortitude saving throws to resist poisons and similar effects.
\end{itemize*}
You apply your character's Strength value to:
\begin{itemize*}
\item Damage rolls when using weapons. (\emph{Exceptions:} Light weapons receive only one half the character's Strength value, while two-handed attacks receive one and a half times the Strength value. Some weapons, such as crossbows, do not apply Strength to damage. A Strength penalty is never halved, even when using light weapons.)
\item Skill points for skills which have Strength as their key attribute.
\end{itemize*}
\par A character needs a 13 Strength to wield medium weapons in one hand, or to wield heavy weapons effectively at all.

\subsection{Dexterity (Dex)}
Dexterity measures hand-eye coordination, agility, and reflexes. This attribute is the most important attribute for rogues, but it's also high on the list for characters who typically wear light or medium armor (rangers and barbarians) or no armor at all (monks, wizards, and sorcerers), and for anyone who wants to be a skilled archer.

You apply your character's Dexterity modifier to:
\begin{itemize*}
\item Ranged attack rolls, except for those made with medium or heavy thrown weapons.
\item Attack rolls with light weapons
\item Armor Class (AC), provided that the character can react to the attack.
\item Reflex saving throws, for avoiding fireballs and other attacks that you can escape by moving quickly.
\item Initiative checks to react first in combat (along with Wisdom).
 \item Balance, Escape Artist, Hide, Move Silently, Ride, Sleight of Hand, and Tumble checks.  These are the skills that have Dexterity as their key attribute.
\end{itemize*}
You apply your character's Dexterity value to:
\begin{itemize*}
\item Skill points for skills which have Dexterity as their key attribute.
\end{itemize*}

\subsection{Constitution (Con)}
Constitution represents your character's health and stamina. A Constitution bonus increases a character's hit points, so the attribute is important for all classes.

You apply your character's Constitution modifier to:
\begin{itemize*}
\item Fortitude saving throws, for resisting poison and similar threats.
\item Concentration checks. Concentration is an attribute important to spellcasters.
\end{itemize*}

\par You apply your character's Constitution value to:
\begin{itemize*}
\item The hit points gained when a character gains a level (though a penalty can never drop a result below 1 -- that is, a character always gains at least 1 hit point each time he or she advances in level). If a character's Constitution score changes enough to alter his or her Constitution value, the character's hit points also increase or decrease accordingly.
\item Skill points for skills which have Constitution as their key attribute.
\end{itemize*}

\subsection{Intelligence (Int)}
Intelligence determines how well your character learns and reasons. This attribute is important for wizards because it affects how many spells they can cast, how hard their spells are to resist, and how powerful their spells can be.

You apply your character's Intelligence modifier to:
\begin{itemize*}
\item Half your Intelligence modifier (rounded down) is applied to Will saving throws to resist mental influence.
\item Craft, Disable Device, Forgery, Knowledge, and Linguistics checks. These are the skills that have Intelligence as their key attribute.
\end{itemize*}

\par You apply your Intelligence value to:
\begin{itemize*}
\item The number of languages your character knows at the start of the game.
\item Skill points for skills which have Intelligence as their key attribute.
\end{itemize*}

\par An animal has an Intelligence score of \minus5 or lower. A creature of humanlike intelligence has a score of at least \minus4.

\subsection{Wisdom (Wis)}
Wisdom describes a character's common sense, perception, and intuition. While Intelligence represents one's attribute to analyze information, Wisdom represents being in tune with and aware of one's surroundings. Wisdom is the most important attribute druids, and it is also important for rangers. If you want your character to be perceptive, put a high score in Wisdom. Every creature has a Wisdom score.

You apply your character's Wisdom modifier to:
\begin{itemize*}
\item Half your Wisdom modifier is applied to Reflex saving throws to recognize and avoid dangerous effects.
\item Half your Wisdom modifier is applied to initiative checks to react first in combat (along with Dexterity).
\item Heal, Listen, Profession, Sense Motive, Spellcraft, Spot, and Survival checks. These are the skills that have Wisdom as their key attribute.
\end{itemize*}
You apply your character's Wisdom value to:
\begin{itemize*}
\item Skill points for skills which have Wisdom as their key attribute.
\end{itemize*}

\subsection{Charisma (Cha)}
Charisma measures a character's force of personality, willpower, persuasiveness, personal magnetism, and attribute to lead. This attribute represents actual strength of personality, not merely how one is perceived by others in a social setting. Charisma is most important for bards, most clerics, paladins, and sorcerers. Every creature has a Charisma score.

You apply your character's Charisma modifier to:
\begin{itemize*}
\item Will saving throws to resist mental influence.
\item Bluff, Diplomacy, Disguise, Gather Information, Handle Animal, Intimidate, Perform, and Use Magic Device checks. These are the skills that have Charisma as their key attribute.
\item Checks that represent attempts to influence others.
\item Turning or rebuking attempts for clerics and paladins attempting to influence zombies, vampires, and other undead.
\end{itemize*}
You apply your character's Charisma value to:
\begin{itemize*}
\item Skill points for skills which have Charisma as their key attribute.
\end{itemize*}

When an attribute score changes, all attributes associated with that score change accordingly, including skill points.

For information on how to determine your character's attribute scores, see Determining Attribute Scores, below.

%Better placed elsewhere. But where?
\section{Determining Attribute Scores}
There are several options for how to determine attribute scores. Under most circumstances, point buy is the best option.

\subsection{Point Buy}
With this method, you can fully control your character's attribute scores to match what you want your character to be. All your character's attribute scores start at 0. You get 15 points to distribute among your character's attribute scores. Attribute scores can be bought according to the costs on \trefnp{Attribute Score Point Costs}.

\begin{dtable}
\lcaption{Attribute Score Point Costs}
\begin{tabularx}{\columnwidth}{X X X X}
\thead{Attribute Score} & \thead{Point Cost} & \thead{Attribute Score} & \thead{Point Cost} \\
\minus2 & \minus2\fn{1} & 3 & 3 \\
\minus2 & \minus1\fn{1} & 4 & 5 \\
0 & 0 & 5 & 7 \\
1 & 1 & 6 & 10 \\
2 & 2 & 7 & 13 \\
\end{tabularx}
1 No more than two attribute scores can be reduced below 10 in this way.
\end{dtable}

\subsection{Other Methods}
Point buy offers the fairest and most customizable system for determining attribute scores, ensuring that players can be almost any character they want to be. However, some groups may wish to determine attribute scores differently. Other options are provided below.

\subsubsection{Predefined Attribute Scores}
This is the simplest method. Simply take the following set of attribute scores and distribute them as you choose among your character's abilities: 5, 4, 3, 2, 0, \minus2. This set of attribute scores is called the ``elite array''. Alternately, you may use the ``savant array'': 6, 4, 2, 0, 0, \minus2.

\subsubsection{Semi-Randomized Point Buy}
With this method, you have only a small degree of control over your character's attribute scores, but all characters generated in this way are equally powerful. As with the point buy method, all your character's attribute scores start at 10, and you get 15 points to distribute among your character's attribute scores. However, you do not have full control over how to distribute those points.

Roll 4d6 for each attribute score, dropping the lowest die with each roll. First roll for the attribute scores that you care about most, and save the least important attribute scores for last. After rolling for an attribute score, sum results on the three highest dice and consult \trefnp{Semi-Randomized Point Buy Results} and spend the appropriate number of points to yield an attribute score, as indicated by \trefnp{Attribute Score Point Costs}. If you do not have enough points remaining to spend the amount indicated by the die roll, spend as many as you can and move on to the next ability.

If you have points remaining after rolling all of your attribute scores, you may distribute the points freely among your abilities, using the normal point buy rules. You cannot buy any stat above 15 during this stage.

After all of your points have been spent, you may swap any two of your attribute scores.

\begin{dtable}
\lcaption{Semi-Randomized Point Buy Results}
\begin{tabularx}{\columnwidth}{X X X}
\thead{Roll} & \thead{Points Spent} & \thead{Attribute Score} \\
3-7   & (\plus2 points)\fn{1} & \minus2 \\
8-9   & (\plus1 point)\fn{1} & \minus9 \\
10    & 0  & 0 \\
11    & 1  & 1 \\
12    & 2  & 2 \\
13    & 3  & 3 \\
14    & 5  & 4 \\
15    & 7  & 5 \\
16    & 10 & 6 \\
17    & 10 (\plus1 point)\fn{2}  & 6 \\
18    & 13 (\plus2 points)\fn{2} & 7 \\
\end{tabularx}
1 You gain extra points for having low stats. You can gain these points any number of times per character. \\
2 You gain extra points for rolling high. You can only gain the points for any individual roll once per character.
\end{dtable}

For characters with more extreme attribute scores, roll d20 for each attribute score and consult \trefnp{Alternate Semi-Randomized Point Buy Results}.

\begin{dtable}
\lcaption{Alternate Semi-Randomized Point Buy Results}
\begin{tabularx}{\columnwidth}{X X X}
\thead{Roll} & \thead{Points Spent} & \thead{Attribute Score} \\
1     & \minus2\fn{1} & \minus2 \\
2-3   & \minus1\fn{1} & \minus2 \\
4-5   & 0 & 0 \\
6-7   & 1 & 1 \\
8-10  & 2 & 2 \\
11-13 & 3 & 3 \\
14-16 & 5 & 4 \\
17-18 & 7 & 5 \\
19 & 10 & 6 \\
20    & 13 (\plus1)\fn{2} & 7 \\
\end{tabularx}
1 You gain extra points for having low stats. You can gain these points any number of times per character. \\
2 You gain extra points for rolling high. You can only gain the points for any individual roll once per character.
\end{dtable}

\subsubsection{Random Point Buy}
This method gives you no control over the character whatsoever, while still ensuring that all characters generated are equally powerful. It functions as the semi-randomized point buy method, except that you also randomize the order in which the attribute scores are rolled.

\subsubsection{Weighted Semi-Randomized Point Buy}
This method gives you more control over your attribute scores, while still requiring you to deal with attribute score advantages or disadvantages that you might not have chosen. It functions like the semi-randomized point buy method, except that you do not roll 3d6 for every attribute score. You have a pool of 18 dice. You can distribute those dice as you choose between each attribute score, except the last one, which is automatically assigned your remaining points. If you roll more than three dice for an attribute score, keep only the highest 3 dice. If you assign fewer than three dice to roll for an attribute score, add 1 to the result for every die less than 3 you are rolling. You can roll as many as six dice for a single attribute score, and you cannot roll less than one die for an attribute score.

For example, player hoping to play a barbarian might assign six dice to Strength, five dice to Constitution, three dice to Dexterity, and two dice to each of Intelligence and Wisdom, leaving the rest of the points to Charisma.

\subsubsection{Classic Hardcore}

This method is completely random and can generate very overpowered or underpowered chracters. It represents the unfairness of the world, where some people are just better or worse than others. Roll 3d6 for each attribute score and subtract 10 from each result. The result is the attribute score.