\chapter{Advanced Feats}

\begin{dtable!*}
\lcaption{General Feats}
\begin{tabularx}{\textwidth}{>{\lcol}p{15em} >{\lcol}p{15em} >{\lcol}X}
\thead{General Feats} & \thead{Prerequisites} & \thead{Benefit} \\
\featref{Mental Fortress} & Cha 9 or base Will save \plus18 & Immune to hostile mind-affecting effects \\
\end{tabularx}
\end{dtable!*}

\begin{dtable!*}
\lcaption{Class Feats}
\begin{tabularx}{\textwidth}{>{\lcol}p{15em} >{\lcol}p{15em} >{\lcol}X}
\thead{Class Feats} & \thead{Prerequisites} & \thead{Benefit} \\
\featref{Improved Channeling} & Channel energy 3d6 & \plus2 level when channeling energy \\
\end{tabularx}
\end{dtable!*}

\begin{dtable!*}
\lcaption{Magic Feats}
\begin{tabularx}{\textwidth}{>{\lcol}p{10em} >{\lcol}p{10em} >{\lcol}X >{\lcol}p{10em}}
\thead{Magic Feats} & \thead{Prerequisites} & \thead{Benefit} & \thead{Feat Type} \\
\featref{Empathic Casting} & 3rd level Enchantment (Emotion) spell & Trade Will saves for caster level with mind-affecting spells & Style \\
\featref{Mass Transporter} & 5th level Conjuration (Translocation) spell & Transport many creatures at once & \x \\ 
\featref{Residual Beguilement} & 3rd level Enchantment (Beguilement) spell & Gain Persuasion bonus after casting a spell & \x \\
\featref{Retributive Counterspell} &  4th level spells & Countered spells rebound on original caster & \x \\
\featref{Spell Specialization} & Magic level 8th, Spell Focus &  \plus4 caster level with specific type of magic, \minus2 penalty with other types. \\
\featref{Wardwalker} & 3rd level Abjuration (Warding) spell & Wards you encounter trigger one round late \\
\end{tabularx}
\end{dtable!*}

\begin{dtable!*}
\lcaption{Skill Feats}
\begin{tabularx}{\textwidth}{>{\lcol}p{15em} >{\lcol}p{15em} >{\lcol}X}
\thead{Skill Feats} & \thead{Prerequisites} & \thead{Benefit} \\
\featref{Legendary Balance} & Acrobatics 13 ranks & Balance on impossible surfaces \\
\featref{Legendary Climber} & Climb 13 ranks & Gain climb speed, climb on impossible surfaces \\
\featref{Legendary Craftsman} & Craft 13 ranks & Craft items with fewer material components \\
\featref{Legendary Devicesmith} & Devices 13 ranks & Disable active spell effects \\
\featref{Legendary Disguise} & Disguise 13 ranks & Alter magical auras with disguise \\
\featref{Legendary Escapist} & Escape Artist 13 ranks & Escape from magical effects \\
\featref{Legendary Liar} & Bluff 13 ranks & Lies become undetectable by magic \\
\featref{Legendary Tumbler} & Acrobatics 13 ranks & Tumble through enemies without provoking \\
\featref{Skill Mastery}\fn{1} & Any skill 10 ranks & Always take 10 on skill checks \\
\featref{Veteran's Knowledge} & Base attack bonus \plus8 & Identify monsters without Knowledge \\
\end{tabularx}
1 You can gain this feat multiple times. Its effects do not stack. Each time you take the feat, it has a different effect. \\
\end{dtable!*}

\begin{dtable!*}
\lcaption{Performance Feats}
\begin{tabularx}{\textwidth}{>{\lcol}p{15em} >{\lcol}p{15em} >{\lcol}X}
\thead{Performance Feats} & \thead{Prerequisites} & \thead{Benefit} \\
\featref{Dual Performance} & Perform 14 ranks, any three Performance feats & Sustain two performances at once \\
\tind \featref{Suggestive Performance} & Perform 10 ranks, Fascinating Performance & Suggest action to fascinated creatures \\
\featref{Freeing Performance} & Perform 12 ranks  & Dispel magical effects on ally \\
\featref{Inspire Greatness} & Perform 12 ranks  & Grant allies bonuses to saving throws and AC \\
\featref{Inspire Heroics} & Perform 18 ranks  & Grant allies temporary hit points and extra attack \\
\featref{Rapid Performance} & Perform 14 ranks, any three Performance feats & Perform as a swift action \\
\end{tabularx}
\end{dtable!*}

\begin{dtable!*}
\lcaption{Combat Feats}
\begin{tabularx}{\textwidth}{>{\lcol}p{10em} >{\lcol}p{10em} >{\lcol}X >{\lcol}p{10em}}
    \thead{Combat Feats} & \thead{Prerequisites} & \thead{Benefit} & \thead{Feat Type} \\
\featref{Bewildering Strike} & Base attack bonus \plus8 & Attack bewilders foe & Strike \\
\featref{Contingent Counter} & Int 3, base attack bonus \plus8 & Prepare to attack when attack misses you & Reaction \\
\featref{Driving Assault} & Str 5, base attack bonus \plus8 & Bull rushed foe provokes attack from you & Power, Maneuver \\
\featref{Heartseeker} & Base attack bonus \plus8 & Trade attack bonus to double critical threat range & Style \\
\featref{Immobilizing Strike} & Base attack bonus \plus8 & Melee attack prevents foe from moving & Strike \\
\featref{Inescapable Bulwark of Defense} & Base attack bonus \plus8, Bulwark of Defense & Enemies cannot avoid provoking attacks of opportunity & Reaction, Style \\
\featref{Overwhelming Force} & Str 5, base attack bonus \plus8 & Apply full Strength to damage when using two hands & Power \\
\featref{Perfect Shot} & Dex 5, base attack bonus \plus8 & Ignore less than total cover/concealment on ranged attacks & Precision, Style \\
\featref{Pierce Wings} & Base attack bonus \plus8 & Ranged attack removes foe's ability to fly & Strike \\ 
\featref{Sickening Strike} & Base attack bonus \plus8 & Attack sickens foe & Strike \\
\featref{Tactical Prediction} & Int 3, Base attack bonus \plus8 & Predict foe's next action & Awareness \\
\featref{Wall Slam} & Str 5, base attack bonus \plus8 & Slam opponent into wall to deal bonus damage & Maneuver, Power \\
\end{tabularx}
\end{dtable!*}

\begin{dtable!*}
\lcaption{Combat Feats}
\begin{tabularx}{\textwidth}{>{\lcol}p{10em} >{\lcol}p{10em} >{\lcol}X >{\lcol}p{10em}}
    \thead{Combat Feats} & \thead{Prerequisites} & \thead{Benefit} & \thead{Feat Type} \\
\featref{Confusing Strike} & Base attack bonus \plus12 & Attack confuses bloodied foe & Strike \\
\featref{Contingent Attack} & Int 5, base attack bonus \plus12 & Prepare to attack as immediate action & Reaction \\
\featref{Legendary Awareness} & Base attack bonus \plus12, any three Awareness feats & Immune to overwhelm & Awareness \\
\featref{Legendary Finesse} & Base attack bonus \plus12, any three Finesse feats & Add half Dexterity to damage & Finesse \\
\featref{Legendary Maneuver Master} & Base attack bonus \plus12, any three Maneuver feats & Never provoke, deal damage with very successful maneuvers & Maneuver \\ 
\featref{Legendary Mobility} & Base attack bonus \plus12, any three Mobility feats & Movement does not provoke & Mobility\\
\featref{Legendary Mounted Warrior} & Base attack bonus \plus12, any three Mounted feats & Share damage with mount & Mounted\\
\featref{Legendary Power} & Base attack bonus \plus12, any three Power feats & Wield weapons as if they were less encumbering & Power\\ 
\featref{Legendary Precision} & Base attack bonus \plus12, any three Precision feats & Very accurate hits deal maximum damage & Precision\\ 
\featref{Legendary Style} & Base attack bonus \plus12, any three Style feats & Use two styles at once & \x \\
\featref{Manyshot} & Dex 7, base attack bonus \plus12 & Shoot two projectiles simultaneously & Precision, Style \\
\featref{Master Tactician} & Int 3, base attack bonus \plus12 & Ready full-round actions with allies & Reaction \\
\featref{Penetrating Strike} & Base attack bonus \plus12 & Attack removes foe's damage reduction \\
\featref{Whirlwind Attack} & Dex 5, base attack bonus \plus12 & One melee attack against each opponent within reach & Style \\
\featref{Counterstorm} & Base attack bonus \plus16 & Foes that miss you provoke & Reaction, Style\\
\featref{Demolishing Strike} & Base attack bonus \plus16 & Attack deals double damage & Strike \\
\featref{Executioner} & Base attack bonus \plus16, any five Reaction feats & Gain free attacks against foes near death & Reaction \\
\end{tabularx}
\end{dtable!*}

\begin{dtable!*}
\lcaption{Metamagic Feats}
\begin{tabularx}{\textwidth}{>{\lcol}p{15em} >{\lcol}p{7.5em} >{\lcol}p{7.5em} >{\lcol}X}
\thead{Metamagic Feats} & \thead{Prerequisites} & \thead{Spell Level Increase} & \thead{Benefit} \\
\featref{Quicken Spell} & Caster level 6th & \plus2 & Cast spell as swift action, but lose next action \\
\featref{Reach Spell} & Caster level 6th & \plus2 & Cast touch spell at Close range \\
\featref{Shape Spell} & Caster level 6th & \plus2 & Exclude areas within spell's area \\
\featref{Widen Spell} & Caster level 8th & \plus3 & Double spell's area
\end{tabularx}
\end{dtable!*}

\begin{dtable!*}
\lcaption{Item Creation Feats}
\begin{tabularx}{\textwidth}{>{\lcol}p{15em} >{\lcol}p{15em} >{\lcol}X}
\thead{Item Creation Feats} & \thead{Prerequisites} & \thead{Benefit} \\
\featref{Imbuement Admixture} & Caster level 6th & Combine multiple spells to create items \\
\end{tabularx}
\end{dtable!*}

\begin{dtable!*}
\lcaption{Surge Feats}
\begin{tabularx}{\textwidth}{>{\lcol}p{15em} >{\lcol}p{15em} >{\lcol}X}
\thead{Surge Feats} & \thead{Prerequisites} & \thead{Benefit} \\
\featref{Alacritous Surge} & 16th level, Dex 7 & Briefly gain effects of \spell{haste} \\
\featref{Deathless Surge} & 12th level, Cha 5 & Briefly gain effects of \spell{death ward} \\
\featref{Freeing Surge} & 16th level, Dex 7 & Briefly gain effects of \spell{freedom} \\
\featref{Healing Surge} & 8th level, Con 3 & Heal damage three times per day \\
\featref{Prescient Surge} & 12th level, Wis 5 & Roll twice three times per day \\
\featref{Restorative Surge} & 8th level, Con 3 & Heal ability damage three times per day \\
\featref{Shattering Surge} & 8th level, Str 3 & \spell{Shatter} struck object three times per day \\
\featref{Surge of Speed} & 8th level, Dex 3 & Briefly gain effects of \spell{expeditious retreat} \\
\end{tabularx}
\end{dtable!*}

\subsection{General Feats}

\feat{Mental Fortress}{General}
\featpre Charisma 9 or base Will save \plus18.
\featben You become immune to hostile mind-affecting effects.

\subsection{Class Feats}

\feat{Improved Channeling}{Class}
\parhead{Prerequisite} Channel energy 3d6.
\parhead{Benefit} You gain a \plus2 competence bonus to your effective cleric level when channeling energy.

\subsection{Magic Feats}

\feat{Empathic Casting}{Magic, Style}
\featpre Ability to cast an Enchantment (Emotion) spell of at least 3rd level.
\featben While using this style, you gain a \plus2 competence bonus to caster level with mind-affecting spells. In exchange, you take a \minus2 penalty to Will saves. At magic level 10th, and every 5 magic levels after that, you increase the bonus and penalty by 1.

\feat{Retributive Counterspell}{Magic}
\parhead{Prerequisites} Ability to cast 4th level spells.
\parhead{Benefit} As part of the action to counter a spell, you may expend an Abjuration (Negation) spell of 4th level or higher. If you do, the counterspelled spell is turned back on the caster as if it were affected by the \spell{spell turning} spell. If it cannot be affected by \spell{spell turning}, such as if it is a spell that only affects the caster, it is simply countered as normal.

\feat{Spell Specialization}{Magic}
Choose a school of magic or a spell descriptor to which you already have applied the Spell Focus feat.
\featpre Magic level 8th, Spell Focus.
\featben You gain a \plus4 competence bonus to your caster level when casting spells and using spell-like abilities from the school of magic or with the spell descriptor you select. In exchange, you take a \minus2 penalty to caster level with all other spells and spell-like abilities.

\feat{Wardwalker}{Magic}
\featpre Ability to cast an Abjuration (Warding) spell of at least 3rd level.
\featben When you trigger an Abjuration (Warding) spell, it does not take effect until one round after you triggered it, giving you time to escape.

\subsection{Skill Feats}

\feat{Legendary Balance}{Skill}
\featpre Acrobatics 13 ranks
\featben You can balance on surfaces that cannot support your weight. The DC is 30 for liquids such as water, 35 for dense gases and raw energy, and 40 for ordinary air. While balancing in this way, you must take a move action each round to continue moving; you cannot remain in the same place in consecutive rounds, or you will fall. The DC increases by 2 for each consecutive round that you spend balancing in this way.

\feat{Legendary Climber}{Skill}
\featpre Climb 13 ranks.
\featben You gain a climb speed equal to your land speed. This grants several benefits. 
\begin{itemize*}
  \item You can always take 10 on Climb checks, even when rushed, endangered, or distracted. 
  \item You gain a \plus5 competence bonus on Climb checks.
  \item A successful Climb check allows you to move a distance equal to your climb speed.
\end{itemize*}

In addition, you can now climb surfaces that are perfectly smooth. The DC is 30 for perfectly smooth vertical surfaces, and 40 to climb on a perfectly smooth ceiling.

\feat{Legendary Craftsman}{Skill}
\featpre Craft (any) 13 ranks.
\featben You can craft items with whatever materials you have on hand. When making an item, if your Craft check is 30 or higher, you use half the costly material components to make the item, replacing them with other, less expensive components. For every 10 points by which the check exceeds 30, you halve the required material components again. This only applies once per item; use the highest result achieved while making that item.

This feat does not affect the material components required to craft magical items.

\feat{Legendary Devicesmith}{Skill}
\featpre Devices 13 ranks.
\featben You can disable spell effects on objects or areas as if they were merely complex devices. You can make a Devices check against an active spell effect within 5 feet of you. If your check result is 30 or higher, it may dispel the effect. The DC to dispel the effect is equal to 20 \add the spell's caster level. Failure indicates that the spell is not dispelled. 

You must be aware of a spell to disable it, either through the Spellcraft skill or because the effect is noticeable. You cannot disable spell effects on creatures.

\feat{Legendary Disguise}{Skill}
\featpre Disguise 13 ranks.
\featben Your disguises can change the magic on a creature. When you make a Disguise check, if the result is 30 or higher, you can decide how that creature and any items on the creature appear when examined by divination spells. For example, you could cause all of its equipment to appear nonmagical, or you could cause it to have a strong aura of good when examined with \spell{detect good}. You cannot create an aura of overwhelming strength with this skill.

Anyone using divination magic on the creature must make a caster level check with a DC equal to your Disguise check in order to perceive the truth. Regardless of the result of the check, the caster is not aware that the check was made.

\feat{Legendary Escapist}{Skill}
\featpre Escape Artist 13 ranks.
\featben You can attempt to escape from magic itself, slipping hostile spells off of your body so they dissipate harmlessly. As a standard action, you can make an Escape Artist check to throw off magical effects on you. If the result is 30 or higher, it may dispel the effect. The DC to dispel the effect is equal to 20 \add the effect's caster level. Failure indicates that the spell is not dispelled.

You must be aware of a spell to disable it, either through the Spellcraft skill or because the effect is noticeable. You can only dispel spell effects which target you directly, not area effects which include you. If a spell targets multiple creatures, you can only remove its effects on you.

\feat{Legendary Handler}{Skill}
\featpre Creature Handling 13 ranks.
\featben You gain two abilities to help you handle creatures. When pacifying a creature, if your check result is 30 or higher, you may convince it to perform a trick of your choice instead of simply pacifying it, whether or not it knows the trick.

In addition, when training a creature, if your check result is 30 or higher, you can train it in only a hour. The creature remembers the trick just as if you had spent the full amount of time training it.

\feat{Legendary Healer}{Skill}
\featpre Heal 13 ranks.
\featben A number of times per day equal to 3 \add half your Wisdom, you may take a standard action to make a Heal check on a creature you can touch. The creature heals hit points equal to twice your check result. For every twenty hit points you would restore, you can instead cure one point of critical damage.

\feat{Legendary Intimidation}{Skill}
\featpre Intimidate 13 ranks.
\featben When you demoralize an opponent, if your check result exceeds 30, the creature is frightened instead of being shaken.

\feat{Legendary Leaper}{Skill}
\featpre Athletics 13 ranks.
\featben You are always treated as having a running start when jumping. When jumping, if your check result is 30 or higher, you can take a standard action at the midpoint of the jump.

\feat{Legendary Liar}{Skill}
\featpre Bluff 13 ranks.
\featben Your lies can fool even magic. When you make a Bluff check, if the result is 30 or higher, anyone using magic to reveal or prevent lies must make a caster level check with a DC equal to your Bluff check in order to affect your lie. Regardless of the result of the check, the caster is not aware that the check was made.

\feat{Legendary Swimmer}{Skill}
\featpre Swim 13 ranks.
\featben You gain a swim speed equal to your land speed. This grants several benefits.
\begin{itemize*}
  \item You can always take 10 on Swim checks, even when rushed, endangered, or distracted. 
  \item You gain a \plus5 competence bonus on Swim checks.
  \item A successful Swim check allows you to move a distance equal to your swim speed.
\end{itemize*}

When swimming, if your check result is 30 or higher, you do not suffer any penalties to melee attack rolls, checks, or armor class for being underwater. You can make a Swim check as a swift action to gain this benefit in a round when you do not make a Swim check to move.

\feat{Legendary Tumbler}{Skill}
\featpre Acrobatics 13 ranks.
\featben When tumbling, if your check result is 30 or higher, you gain two benefits. First, you do not provoke attacks of opportunity for that movement. Second, you can tumble through areas occupied by enemies as if they were unoccupied.

\feat{Skill Mastery}{Skill}
Choose a skill.
\featpre 10 ranks in the chosen skill.
\featben When making a skill check with your chosen skill, you may take 10 even if stress and distraction would normally prevent you from doing so. When you take 10, you treat any roll of less than 10 as if it were a 10.
\parhead{Special} You can gain this feat multiple times. Its effects do not stack. Each time you take the feat, it applies to a new skill.

\feat{Veteran's Experience}{Skill}
\featpre Base attack bonus \plus8
\featben You can use your battlefield experience in place of learned knowledge to identify monsters. When attempting to identify a monster, you may roll your base attack bonus \add your Intelligence. A successful check gives you the same information as a Knowledge check would.

\subsection{Performance Feats}

\feat{Dual Performance}{Performance, Skill}
\featpres Perform 14 ranks, any three Performance feats.
\featben You can sustain two performances as part of the same swift action. You take a \minus10 penalty to other Perform checks while sustaining two performances at once.

\feat{Freeing Performance}{Performance, Skill}
\featpre Perform 12 ranks.
\featben By spending a minute performing continuously, you can make a DC 28 Perform check to free an ally within \rngmed range from hostile magic. This creates an effect similar to the \spell{break enchantment} spell, with a caster level equal to your Perform ranks.

You can use any combination of performance feats you possess a number of times per day equal to the number of performance feats you have \add half your Charisma. If you fail the Perform check, the use of the ability is wasted.

\feat{Inspire Greatness}{Performance, Skill}
\featpre Perform 12 ranks.
\featben As a standard action, you can make a DC 24 Perform check to inspire greatness in allies within \rngmed range of you, including yourself. You can affect one creature per two Perform ranks you have. The affected creatures gain a \plus4 enhancement bonus to saving throws and dodge modifier. This bonus increases to \plus5 at 20 Perform ranks.

You can use any combination of performance feats you possess a number of times per day equal to the number of performance feats you have \add half your Charisma. If you fail the Perform check, the use of the ability is wasted. This is a mind-affecting ability.

\feat{Inspire Heroics}{Performance, Skill}
\featpre Perform 18 ranks.
\featben As a standard action, you can make a DC 32 Perform check to inspire heroics in allies within \rngmed range of you, including yourself. You can affect one creature per two Perform ranks you have. All affected creatures gain temporary hit points equal to your check result and can take an extra attack at a \minus5 penalty when making a full attack. This extra attack is not cumulative with similar effects. You can sustain the performance as a swift action.

\feat{Rapid Performance}{Performance, Skill}
\featpre Perform 14 ranks, any three Performance feats.
\featben You can use performance feats as a swift action by taking a \minus5 penalty on the Perform check to activate the feat.

\feat{Suggestive Performance}{Performance, Skill}
\featpre Perform 10 ranks, Fascinating Performance.
\featben As a swift action, you can make a DC 20 Perform check to suggest an action to a single creature you are currently fascinating with the Fascinating Performance feat. This functions as the \spell{suggestion} spell, with a saving throw DC equal to your Perform check. You can affect an additional creature at 12 Perform ranks and every 2 Perform ranks thereafter.

You can use any combination of performance feats you possess a number of times per day equal to the number of performance feats you have \add half your Charisma. If you fail the Perform check, the use of the ability is wasted.

\subsection{Surge Feats}
%Design guidelines: effective CL = 1/2 HD, min ability score = effective CL-1
%Uses/day: 3/day for single use abilities, normal duration for buffs (splitting
%duration is already good?)
%Personal range penalizes spell level, but shouldn't make these cheaper.
%Treat personal range spells as one higher level.

\feat{Alacritous Surge}{Surge}
\featpre 16th level, Dex 7
\featben You may gain the benefits of the \spell{haste} spell for up to 5 rounds each day. Activating this ability is a swift action, and it lasts until you will it to stop (a free action) or until the duration expires.

\feat{Deathless Surge}{Surge}
\featpre 12th level, Cha 5
\featben You may gain the benefits of the \spell{death ward} spell for up to 5 rounds each day. Activating this ability is a swift action, and it lasts until you will it to stop (a free action) or until the duration expires.

\feat{Freeing Surge}{Surge}
\featpre 16th level, Dex 7
\featben You may gain the benefits of the \spell{freedom} spell for up to 5 rounds each day. Activating this ability is a swift action, and it lasts until you will it to stop (a free action) or until the duration expires.

\feat{Healing Surge}{Surge}
\featpre 8th level, Con 3
\featben Three times per day, you may heal 4d6 points of damage as a standard action. This healing increases by 1d6 at 10th level and at every 2 levels thereafter.

\feat{Prescient Surge}{Awareness, Surge}
\featpre 12th level, Wis 5
\featben Three times per day, you may roll twice on any single attack roll, opposed check, or saving throw. You must decide to use this ability before making the roll. Activating this ability is an immediate action.

\feat{Restorative Surge}{Surge}
\featpre 8th level, Con 3
\featben Three times per day, you may heal two points of ability damage as a swift action. This healing increases by 1 at 12th level and every 4 levels thereafter.

\feat{Shattering Surge}{Power, Surge}
\featpre 8th level, Str 3
\featben Three times per day, when you successfully attempt to sunder an object, you may cause it to be affected as if by the targeted version of the \spell{shatter} spell with a caster level equal to your level. Activating this ability is an immediate action.

\feat{Surge of Speed}{Surge}
\featpre 8th level, Dex 3
\featben You may gain the benefits of the \spell{expeditious retreat} spell for up to 5 rounds each day. Activating this ability is a swift action, and it lasts until you will it to stop (a free action) or until the duration expires.
