\chapter{Advanced Classes}
Characters above 6th level can increase substantially in power and versatility. Rules for these advanced characters are described below.

\section{Class Descriptions}

\subsection{Barbarian}
\begin{dtable*}
\lcaption{The Barbarian}
\begin{tabularx}{\textwidth}{>{\ccol}p{\levelcol} >{\ccol}p{\babcolgood} *{3}{>{\ccol}p{\babcolgood}} X}
\thead{Level} & \thead{Base Attack Bonus} & \thead{Fort Save} & \thead{Ref Save} & \thead{Will Save} & \thead{Special} \\
1st & \plus1         & \plus3 & \plus1 & \plus0 & Damage reduction, rage \plus2 \\
2nd & \plus2         & \plus4 & \plus2 & \plus1 & Fast movement, uncanny dodge \\
3rd & \plus3         & \plus5 & \plus3 & \plus1 & Endurance, channeled rage \\
4th & \plus4         & \plus6 & \plus4 & \plus2 & Grit \\
5th & \plus5         & \plus7 & \plus4 & \plus2 & \\
6th & \plus6/\plus1  & \plus8 & \plus5 & \plus3 & Improved uncanny dodge, channeled rage \\
7th & \plus7/\plus2  & \plus9 & \plus6 & \plus3 & Larger than life \\
8th & \plus8/\plus3  & \plus10& \plus7 & \plus4 & Rage \plus3 \\
9th & \plus9/\plus4  & \plus11& \plus8 & \plus4 & Channeled rage \\
10th& \plus10/\plus5 & \plus12& \plus8 & \plus5 & Greater uncanny dodge \\
11th& \plus11/\plus6/\plus1  & \plus13 & \plus9 & \plus5 & Tireless rage \\
12th& \plus12/\plus7/\plus2  & \plus14 & \plus10& \plus6 & Channeled rage, chaotic rage \\
13th& \plus13/\plus8/\plus3  & \plus15 & \plus10& \plus6 & Indomitable will \\
14th& \plus14/\plus9/\plus4  & \plus16 & \plus11& \plus7 & Rage \plus4 \\
15th& \plus15/\plus10/\plus5 & \plus17 & \plus12& \plus7 & Channeled rage \\
16th& \plus16/\plus11/\plus6/\plus1 & \plus18 & \plus13& \plus8 & Improved grit \\
17th& \plus17/\plus12/\plus7/\plus2 & \plus19 & \plus13& \plus8 & Larger than belief \\
18th& \plus18/\plus13/\plus8/\plus3 & \plus20 & \plus14& \plus9 & Channeled rage \\
19th& \plus19/\plus14/\plus9/\plus4 & \plus21 & \plus15& \plus9 & Deathless rage \\
20th& \plus20/\plus15/\plus10/\plus5& \plus22 & \plus16 & \plus10 & Endless rage, rage \plus5
\end{tabularx}
\end{dtable*}

\cf{Bbn}{Channeled Rage} Some channeled rages require higher levels in barbarian.

\subcf{Mindless Rage} The barbarian becomes immune to mind-affecting spells and effects for the duration of his rage.

\textit{Prerequisites:} Barbarian level 15th.

\subcf{Spellbreaker Rage (Su)} The barbarian gains spell resistance while raging. A creature with spell resistance may always make a saving throw when a spell is cast on it. The saving throw type is indicated by the spell. If it succeeds, the spell has no effect on it.

\textit{Prerequisites:} Barbarian level 9th.

\subcf{Terrifying Rage (Su)} Any enemy beginning its turn within the barbarian's threatened area must make a Will save or be shaken for 5 rounds. The save DC is equal to 10 \add the barbarian's level \add his Charisma modifier. This can only affect any individual creature once per 24 hours.

\textit{Prerequisites:} Barbarian level 6th.

\subcf{Unstoppable Rage} Each round, the barbarian can bull rush an opponent as a swift action that does not provoke attacks of opportunity.

\textit{Prerequisites:} Barbarian level 6th.

\cf{Bbn}{Improved Uncanny Dodge (Ex)} At 6th level and higher, a barbarian is always treated as being threatened by two fewer creatures than he actually is for the purpose of determining overwhelm penalties. This defense can deny a rogue the ability to sneak attack the barbarian.
\par If a character already has uncanny dodge (see above) from a second class and gains improved uncanny dodge, the character stacks those levels to determine if he should gain greater uncanny dodge, and to determine the minimum level a rogue must be to flank the character.

\cf{Bbn}{Larger than Life (Ex)} A barbarian of 7th level or higher holds the strength of a giant in the body of a man (or woman). The barbarian is treated as being one size category larger than he actually is for the purpose of combat maneuvers he performs or is the target of, checks that are affected by size (such as Strength checks to break down doors), and whether a creature's special attacks based on size can affect him if doing so is advantageous to him. In addition, though he uses weapons of the same size, his weapons deal damage as if they were one size category larger, including natural weapons and unarmed strikes. The barbarian's space and reach remain those of a creature of his actual size. The benefits of this class feature stack with the effects of spells and abilities that increase the barbarian's size category.

\cf{Bbn}{Rage} At 8th level, the barbarian's rage bonus increases to \plus3 (and 3 temporary hit points per barbarian level). It increases to \plus4 at 14th level, and to \plus5 at 20th level. The penalty to Armor Class remains the same.

\cf{Bbn}{Greater Uncanny Dodge (Ex)} At 10th level and higher, a barbarian no longer suffers overwhelm penalties, regardless of the number of foes surrounding him.

\cf{Bbn}{Tireless Rage (Ex)} At 11th level and higher, a barbarian no longer becomes fatigued at the end of his rage.

\cf{Bbn}{Chaotic Rage (Ex)} At 12th level, the barbarian gains the ability to change channeled rage abilities at will, without consuming an additional use of his rage ability. He may not change channeled rages in this way more than once per round.

\cf{Bbn}{Indomitable Will (Ex)} At 13th level, a barbarian becomes immune to compulsion and domination spells and effects.

\cf{Bbn}{Improved Grit (Ex)} At 16th level, a barbarian's fortitude knows no bounds. If he fails a Fortitude save against an effect that deals half damage on a successful save, he takes only half damage.

\cf{Bbn}{Larger than Belief (Ex)} At 17th level, the barbarian's larger than life ability improves. He is treated as being two size categories larger than he actually is.

\cf{Bbn}{Deathless Rage (Ex)} At 19th level and higher, a raging barbarian can scorn death and unconsciousness. As long as his rage continues, he is not staggered at 0 hit points, and cannot take critical damage. However, every 50 points of damage he takes in excess of his hit points reduces the duration of his rage by one round, and the Endless Rage channeled rage ability does not extend the duration of his rage if he is at 0 hit points. Once his rage ends, the effects of the barbarian's wounds apply normally if they have not been healed. This ability does not prevent death from spell effects such as \spell{finger of death} or \spell{disintegrate}.

\cf{Bbn}{Limitless Rage (Ex)} At 20th level, the barbarian may rage at will. He no longer has any limitation on the number of times he can rage each day. He may still rage no more than once per encounter.

\subsection{Cleric}
\begin{dtable*}
\lcaption{The Cleric}
\begin{tabularx}{\textwidth}{>{\ccol}p{2em} >{\ccol}p{7em} *{3}{>{\ccol}p{\savecol}} >{\lcol}X}
\thead{Level} & \thead{Base Attack Bonus} & \thead{Fort Save} & \thead{Ref Save} & \thead{Will Save} & \thead{Special} \\
1st & \plus0 & \plus1 & \plus0 & \plus3 & Matters of faith, lesser domain aspect, spontaneous casting \\
2nd & \plus1 & \plus2 & \plus1 & \plus4         & Channel energy, lesser domain aspect \\
3rd & \plus2 & \plus3 & \plus1 & \plus5         & Domain power \\
4th & \plus3 & \plus4 & \plus2 & \plus6         & Domain power \\
5th & \plus3 & \plus4 & \plus2 & \plus7         & Channelled domain power \\
6th & \plus4 & \plus5 & \plus3 & \plus8         & \x  \\
7th & \plus5 & \plus6 & \plus3 & \plus9         & Domain aspect  \\
8th & \plus6/\plus1 & \plus7 & \plus4 & \plus10    & \x  \\
9th & \plus6/\plus1 & \plus7 & \plus4 & \plus11    & Channelled domain power  \\
10th & \plus7/\plus2 & \plus8 & \plus5 & \plus12    & \x  \\
11th & \plus8/\plus3 & \plus9 & \plus5 & \plus13   & Domain aspect  \\
12th & \plus9/\plus4 & \plus10& \plus6 & \plus14    & \x  \\
13th & \plus9/\plus4 & \plus10& \plus6 & \plus15    & Greater channeled domain power  \\
14th & \plus10/\plus5 & \plus11& \plus7 & \plus16    & \x  \\
15th & \plus11/\plus6/\plus1 & \plus12& \plus7 & \plus17 & Greater channeled domain power  \\
16th & \plus12/\plus7/\plus2 & \plus13& \plus8 & \plus18 & \x  \\
17th & \plus12/\plus7/\plus2 & \plus13& \plus8 & \plus19 & Domain mastery  \\
18th & \plus13/\plus8/\plus3 & \plus14& \plus9 & \plus20 & \x  \\
19th & \plus14/\plus9/\plus4 & \plus15& \plus9 & \plus21 & Domain mastery  \\
20th & \plus15/\plus10/\plus5 & \plus16 & \plus10 & \plus22 & \x  \\
\end{tabularx}
\end{dtable*}

\cf{Clr}{Spells} A cleric gains additional spells per day and spells known as he increases in level.

\begin{dtable}
    \lcaption{Spells per Day} 
    \centering
    \begin{tabularx}{\columnwidth}{>{\ccol}X *{10}{>{\ccol}p{\spellcol}}}
        \thead{Level} & \thead{1st} & \thead{2nd} & \thead{3rd} & \thead{4th} & \thead{5th} & \thead{6th} & \thead{7th} & \thead{8th} & \thead{9th} \\
        1st & 3 & \x & \x & \x & \x & \x & \x & \x & \x \\
        2nd & 4 & \x & \x & \x & \x & \x & \x & \x & \x \\
        3rd & 5 & \x & \x & \x & \x & \x & \x & \x & \x \\
        4th & 6 & 3 & \x & \x & \x & \x & \x & \x & \x \\
        5th & 6 & 4 & \x & \x & \x & \x & \x & \x & \x \\
        6th & 6 & 5 & 3 & \x & \x & \x & \x & \x & \x \\
        7th & 6 & 6 & 4 & \x & \x & \x & \x & \x & \x \\
        8th & 6 & 6 & 5 & 3 & \x & \x & \x & \x & \x \\
        9th & 6 & 6 & 6 & 4 & \x & \x & \x & \x & \x \\
        10th & 6 & 6 & 6 & 5 & 3 & \x & \x & \x & \x \\
        11th & 6 & 6 & 6 & 6 & 4 & \x & \x & \x & \x \\
        12th & 6 & 6 & 6 & 6 & 5 & 3 & \x & \x & \x \\
        13th & 6 & 6 & 6 & 6 & 6 & 4 & \x & \x & \x \\
        14th & 6 & 6 & 6 & 6 & 6 & 5 & 3 & \x & \x \\
        15th & 6 & 6 & 6 & 6 & 6 & 6 & 4 & \x & \x \\
        16th & 6 & 6 & 6 & 6 & 6 & 6 & 5 & 3 & \x \\
        17th & 6 & 6 & 6 & 6 & 6 & 6 & 6 & 4 & \x \\
        18th & 6 & 6 & 6 & 6 & 6 & 6 & 6 & 5 & 3 \\
        19th & 6 & 6 & 6 & 6 & 6 & 6 & 6 & 6 & 4 \\
        20th & 6 & 6 & 6 & 6 & 6 & 6 & 6 & 6 & 6 \\
    \end{tabularx}
\end{dtable}

\begin{dtable}
\lcaption{Cleric Spells Known}
\centering
\begin{tabularx}{\columnwidth}{>{\ccol}X *{10}{>{\ccol}p{\spellcol}}}
& \multicolumn{10}{c}{\thead{---{}---{}---{}---{}---{}---{}---{}---Spells Known---{}---{}---{}---{}---{}---{}---{}---}} \\
\thead{Level} & \thead{1st} & \thead{2nd} & \thead{3rd} & \thead{4th} & \thead{5th} & \thead{6th} & \thead{7th} & \thead{8th} & \thead{9th} \\
1st  & 0\plus 2 & \x & \x & \x & \x & \x & \x & \x & \x \\
2nd  & 1\plus 2 & \x & \x & \x & \x & \x & \x & \x & \x \\
3rd  & 2\plus 2 & \x & \x & \x & \x & \x & \x & \x & \x \\
4th  & 2\plus 2 & 0\plus 2 & \x & \x & \x & \x & \x & \x & \x \\
5th  & 3\plus 2 & 1\plus 2 & \x & \x & \x & \x & \x & \x & \x \\
6th  & 3\plus 2 & 1\plus 2 & 0\plus 2 & \x & \x & \x & \x & \x & \x \\
7th  & 3\plus 2 & 2\plus 2 & 1\plus 2 & \x & \x & \x & \x & \x & \x \\
8th  & 3\plus 2 & 2\plus 2 & 1\plus 2 & 0\plus 2 & \x & \x & \x & \x & \x \\
9th  & 3\plus 2 & 2\plus 2 & 2\plus 2 & 1\plus 2 & \x & \x & \x & \x & \x \\
10th & 3\plus 2 & 2\plus 2 & 2\plus 2 & 1\plus 2 & 0\plus 2 & \x & \x & \x & \x \\
11th & 3\plus 2 & 2\plus 2 & 2\plus 2 & 2\plus 2 & 1\plus 2 & \x & \x & \x & \x \\
12th & 3\plus 2 & 2\plus 2 & 2\plus 2 & 2\plus 2 & 1\plus 2 & 0\plus 2 & \x & \x & \x \\
13th & 3\plus 2 & 2\plus 2 & 2\plus 2 & 2\plus 2 & 2\plus 2 & 1\plus 2 & \x & \x & \x \\
14th & 3\plus 2 & 2\plus 2 & 2\plus 2 & 2\plus 2 & 2\plus 2 & 1\plus 2 & 0\plus 2 & \x & \x \\
15th & 3\plus 2 & 2\plus 2 & 2\plus 2 & 2\plus 2 & 2\plus 2 & 2\plus 2 & 1\plus 2 & \x & \x \\
16th & 3\plus 2 & 2\plus 2 & 2\plus 2 & 2\plus 2 & 2\plus 2 & 2\plus 2 & 1\plus 2 & 0\plus 2 & \x \\
17th & 3\plus 2 & 2\plus 2 & 2\plus 2 & 2\plus 2 & 2\plus 2 & 2\plus 2 & 1\plus 2 & 1\plus 2 & \x \\
18th & 3\plus 2 & 2\plus 2 & 2\plus 2 & 2\plus 2 & 2\plus 2 & 2\plus 2 & 1\plus 2 & 1\plus 2 & 0\plus 2 \\
19th & 3\plus 2 & 2\plus 2 & 2\plus 2 & 2\plus 2 & 2\plus 2 & 2\plus 2 & 1\plus 2 & 1\plus 2 & 1\plus 2 \\
20th & 3\plus 2 & 2\plus 2 & 2\plus 2 & 2\plus 2 & 2\plus 2 & 2\plus 2 & 1\plus 2 & 1\plus 2 & 1\plus 2
\end{tabularx}
\end{dtable}

\cf{Clr}{Domain Aspect (Su)} At 7th level, a cleric gains a domain aspect from one of his domains. Domain aspects do not require an action to activate. Options for domain aspects are listed below.

\subcf{Air -- Stormwalker} The cleric suffers no penalties for inclement weather or severe winds and takes half damage from falling damage.
\subcf{Chaos -- Fortune's Friend} Whenever the cleric rolls randomly for an effect, such as when he is affected by the \spell{confusion} spell, he may roll twice and take whichever result he prefers.
\subcf{Death -- Lifedrinker} Whenever the cleric kills a creature with a death effect other than \spell{death knell}, he automatically gains the benefits of a \spell{death knell} spell as if it was cast on the creature he killed.
\subcf{Destruction -- Swordcleaver} Whenever the cleric breaks or destroys an object with a melee attack, he may take a free melee attack on a creature adjacent to him at the same attack bonus.
\subcf{Earth -- Anchored} The cleric gains a \plus4 enhancement bonus to CMD against bull rush, overrun, and trip attempts while standing on solid ground.
\subcf{Evil -- Malevolent Magic} Good creatures take a \minus2 penalty to saving throws against the cleric's spells.
\subcf{Fire -- Flamebearer} The cleric gains Spell Focus (Fire) as a bonus feat.
\subcf{Good -- Purifying Magic} Evil creatures take a \minus2 penalty to saving throws against the cleric's spells.
\subcf{Knowledge -- Knowledge Mastery} The cleric may choose a number of Knowledge skills equal half to his Intelligence (minimum 1). He may take 10 with those skills if he is not in danger or rushed.
\subcf{Law -- Certain Triumph} Whenever the cleric would take 10, he may instead take 12, treating any roll lower than a 12 as if it had been a 12.
\subcf{Magic -- Metamagic Feat} The cleric gains a bonus metamagic feat.
\subcf{Nature -- Favored Terrain} The cleric gains a favored terrain, as the ranger class feature.
\subcf{Protection -- Faithful Shield} The cleric may maintain concentration on Abjuration (Shielding) effects as a swift action.
\subcf{Strength -- Mighty Magic} The cleric can add half his Strength to his casting attribute to meet the minimum attribute requirements to cast spells.
\subcf{Travel -- Rapid Traveller} The cleric gains a \plus10 foot competence bonus to his base land speed.
\subcf{Trickery -- }
\subcf{Vitality -- }
\subcf{War -- Weapon Specialization} The cleric gains Weapon Specialization in his deity's favored weapon group as a bonus feat. If he does not have a deity, he gains it in a weapon group related to his alignment or ideals.
\subcf{Water -- Water Breathing} The cleric may breathe and speak normally while underwater, as the \spell{water breathing} ritual. He may also pass through boggy or wet areas with no penalty to his movement speed.

\par The cleric gains an additional domain aspect from one of his domains at 11th level.

\cf{Clr}{Channeled Domain Power (Su)} The cleric gains an additional channelled domain power at 9th level.

\cf{Clr}{Greater Channelled Domain Power (Su)} At 13th level, a cleric gains a greater channelled domain power from one of his domains. Using a greater channelled domain power consumes two uses of the cleric's channel energy ability. Instead of channeling positive or negative energy, the cleric instead gains the effect of the greater channelled domain power. Options for greater channeled domain powers are listed below.
\subcf{Air -- Mantle of Air} As a swift action, the cleric can surround himself in a mantle of air for 5 rounds. Thrown and projectile weapons have a 50\% chance to miss him while this effect is active. Unusually large weapons, such as a giant's boulders, may suffer a decreased miss chance as appropriate to their size.
\subcf{Chaos -- Anarchic Weapon} As a swift action, the cleric can imbue a touched weapon with the anarchic weapon property for 5 rounds.
\subcf{Death -- Channel Death} The cleric channels negative energy as the Death channelled domain power, except that any creature brought to 0 hit points by this effect immediately dies. This is a death effect.
\subcf{Destruction -- Tide of Destruction} The cleric channels destructive energy as the Destruction channelled domain power, except that any creature damaged by the effect is also filled with a destructive resonance for 5 rounds. the first time each round that each affected creature takes damage, that damage is increased by half the cleric's level. This is considered a circumstance bonus to damage.
\subcf{Earth -- Mantle of Earth} As a swift action, the cleric can surround himself in a mantle of earth for 5 rounds. He gains physical damage reduction equal to half his cleric level that is only overcome by adamantine weapons.
\subcf{Evil -- Unholy Weapon} As a swift action, the cleric can imbue a touched weapon with the unholy weapon property for 5 rounds.
\subcf{Fire -- Mantle of Fire} As a swift action, the cleric can surround himself in a mantle of fire for 5 rounds. He gains the effect of a \spell{fire shield} spell, with a caster level equal to his cleric level.
\subcf{Good -- Holy Weapon} As a swift action, the cleric can imbue a touched weapon with the holy weapon property for 5 rounds.
\subcf{Knowledge -- See the Truth} As a swift action, the cleric can gain the benefit of the \spell{true seeing} spell for 1 round.
\subcf{Law -- Axiomatic Weapon} As a swift action, the cleric can imbue a touched weapon with the axiomatic weapon property for 5 rounds.
\subcf{Life -- Persistent Life} The cleric can restore life, as the \spell{raise dead} spell, to a touched corpse that died no more than 5 rounds previously.
\subcf{Magic -- Enhance Metamagic} The cleric can use this power as part of casting a spell affected by a metamagic feat. If he does, the spell costs a spell slot of one level lower than normal, and applying the metamagic does not increase the casting time of the spell.
\subcf{Nature -- Wild Aspect} When the cleric gains this ability, he chooses one wild aspect ability, as if he were a were a druid of a level equal to his cleric level. When he uses this ability, he may assume that wild aspect. This effect lasts as long as that wild aspect would normally last.
\subcf{Protection -- Mass Sanctuary} The cleric channels protective energy as the Protection channelled domain power, except that instead of being healed, each affected creature separately gains the benefit of a \spell{sanctuary} spell for 5 rounds.  If a member of the group attacks, the effect is broken for that creature, but not for the whole group.
\subcf{Strength -- Surge of Strength} As a swift action, the cleric can add his cleric level as an enhancement bonus to his Strength for a single round.
\subcf{Travel -- Uninhibited Movement} As a swift action, the cleric can gain the ability to move without provoking any attacks of opportunity for a single round.
\subcf{Trickery -- Swift Invisibility} As a swift action, the cleric can gain the benefit of the \spell{invisibility} spell for a single round.
\subcf{War -- Warmaster's Boon} The cleric can use this power as part of casting a spell that targets himself with a duration of \durshort or longer and a range greater than personal. If he uses the power, the spell affects all of his allies within an \areamed radius. However, the spell lasts for no longer than half his Charisma score in rounds.
\subcf{Water -- Aquatic Globe} The cleric creates water out of thin air in an immobile \areamed radius emanation from his location for 5 rounds. Everything within the area is treated as if it were underwater. At the end of the duration, the water evaporates, leaving no trace that it was ever there.

\par The cleric gains an additional greater channelled domain power at 15th level.

\cf{Clr}{Domain Mastery (Su)} At 17th level, a cleric gains a domain mastery from one of his domains. Options for domain masteries are listed below.

\subcf{Air -- Flight} The cleric gains a fly speed (good maneuverability) equal to his land speed. He may remain flying for up to 5 rounds at a time. After that, he must land for 1 round before he can fly again.
\subcf{Chaos -- Avatar of Luck} Once per round, the cleric can add d6 as a circumstance bonus to any attack roll or check. He may declare the use of the ability after failing the roll, but before any additional effects are resolved, potentially making it succeed where it would have failed.
\subcf{Death -- Deathfeeder} The cleric constantly radiates a \areamed radius emanation of death. Whenever a creature dies within the area, he gains the benefits of the \spell{death knell} spell as if it had been cast on the creature.
\subcf{Destruction -- Ruinbringer} The cleric's attacks and spells ignore all damage reduction and hardness (but not damage immunity).
\subcf{Earth -- Earth Glide} The cleric gains the earth glide ability, as an earth elemental.
\subcf{Evil -- Avatar of Evil} The cleric continuously gains the benefits of the \spell{protection from good} spell, with a caster level equal to his cleric level. If the effect is dispelled or suppressed, he can resume it as a swift action.
\subcf{Fire -- Flame Incarnate} The cleric gains the fire subtype, making him immune to fire but giving him a 50\% vulnerability to cold damage. When ever he uses fire spells or effects, he may freely exclude areas or creatures within the area of effect.
\subcf{Good -- Avatar of Good} The cleric continuously gains the benefits of the \spell{protection from evil} spell, with a caster level equal to his cleric level. If the effect is dispelled or suppressed, he can resume it as a swift action.
\subcf{Knowledge -- Combat Insight} The cleric gains a \plus2 circumstance bonus to attack rolls, checks, and saving throws against creatures he has identified with a successful Knowledge check.
\subcf{Law -- Avatar of Order} Once per round, the cleric can take 5 on an attack roll or check, even while stressed or distracted. He may declare the use of this ability after rolling below a 5, but before any additional effects are resolved, potentially causing it to succeed where it would have failed.
\subcf{Life -- Fountain of Life} The cleric gains fast healing 1. All of the cleric's healing spells and abilities restore critical damage as easily as if it were hit point damage.
\subcf{Magic -- Spellbreaker} The cleric gains spell resistance. A creature with spell resistance may always make a saving throw when a spell is cast on it. The saving throw type is indicated by the spell. If it succeeds, the spell has no effect on it.
\subcf{Nature -- Natural Power} Whenever the cleric is in natural terrain, he gains a \plus2 enhancement bonus to caster level and the improved natural casting ability, as the druid class feature.
\subcf{Protection -- Martyr's Gift} The cleric constantly radiates a \areamed radius emanation of protective energy. Whenever a creature within the area takes damage, the cleric can choose to take half of that damage instead, as the \spell{shield other} spell.
\subcf{Strength -- Might of the Gods} The cleric gains the larger than life ability, as the barbarian class feature. In addition, he may use his Strength as his casting attribute, and to determine the saving throw DC of his channel energy abilities.
\subcf{Travel -- Perfect Stride} The cleric gains perfect stride, as the ranger class feature. He constantly acts as if he were under the effect of a \spell{freedom} spell, except that it does not allow him to act normally underwater.
\subcf{Trickery -- Exemplar of Deceit} The cleric continuously gains the benefits of the \spell{nondetection} spell, with a caster level equal to his cleric level, except that it also protects him from any spells or effects which would prevent him from lying or reveal his lies. If the effect is dispelled or suppressed, he can resume it as a swift action.
\subcf{War -- Warmaster's Favor} The cleric continuously gains the benefits of the \spell{divine favor} spell, with a caster level equal to his cleric level. If the effect is dispelled or suppressed, he can resume it as a swift action.
\subcf{Water -- Water's Flow} At any time, the cleric can transform into a rushing flow of water as a move action that does not provoke attacks of opportunity. As part of the action, he may move up to his movement speed in any direction that water could go. His speed is halved when moving uphill and doubled when moving downhill. He does not provoke attacks of opportunity during this movement, and has physical damage reduction 10 while in this form. At the end of his movement, he regains his normal form.

\par The cleric gains an additional domain mastery at 19th level.

\subsection{Druid}
\begin{dtable*}
\lcaption{The Druid}
\begin{tabularx}{\textwidth}{>{\ccol}p{\levelcol} >{\centering}p{\babcolavg} *{3}{>{\ccol}p{\savecol}} >{\ccol}X *{9}{>{\ccol}p{\spellcol}}}
& & & & & & \multicolumn{9}{c}{\thead{---{}---{}---{}---{}---{}---{}---Spells per Day---{}---{}---{}---{}---{}---}} \\
\thead{Level} & \thead{Base Attack Bonus} & \thead{Fort Save} & \thead{Ref Save} & \thead{Will Save} & \thead{Special} & \thead{1st} & \thead{2nd} & \thead{3rd} & \thead{4th} & \thead{5th} & \thead{6th} & \thead{7th} & \thead{8th} & \thead{9th} \\
1st & \plus0 & \plus3 & \plus0 & \plus1 & Nature sense, natural casting, wild speech
& 3 & \x & \x & \x & \x & \x & \x & \x & \x \\
2nd & \plus1 & \plus4 & \plus1 & \plus2 & Woodland stride
& 4 & \x & \x & \x & \x & \x & \x & \x & \x \\
3rd & \plus2 & \plus5 & \plus1 & \plus3 & Wild aspect
& 5 & \x & \x & \x & \x & \x & \x & \x & \x \\
4th & \plus3 & \plus6 & \plus2 & \plus4 & Venom immunity
& 6 & 3 & \x & \x & \x & \x & \x & \x & \x \\
5th & \plus3 & \plus7 & \plus2 & \plus4 & Wild aspect
& 6 & 4 & \x & \x & \x & \x & \x & \x & \x \\
6th & \plus4 & \plus8 & \plus3 & \plus5 & Wild speech (plants)
& 6 & 5 & 3 & \x & \x & \x & \x & \x & \x \\
7th & \plus5 & \plus9 & \plus3 & \plus6 & Wild aspect
& 6 & 6 & 4 & \x & \x & \x & \x & \x & \x \\
8th & \plus6/\plus1 & \plus10& \plus4 & \plus7 & Improved wild speech
& 6 & 6 & 5 & 3 & \x & \x & \x & \x & \x \\
9th & \plus6/\plus1 & \plus11& \plus4 & \plus7 & Wild aspect
& 6 & 6 & 6 & 4 & \x & \x & \x & \x & \x \\
10th & \plus7/\plus2 & \plus12& \plus5 & \plus8 & Improved natural casting
& 6 & 6 & 6 & 5 & 3 & \x & \x & \x & \x \\
11th & \plus8/\plus3 & \plus13 & \plus5 & \plus9  & Greater wild aspect
& 6 & 6 & 6 & 6 & 4 & \x & \x & \x & \x \\
12th & \plus9/\plus4 & \plus14 & \plus6 & \plus10 & A thousand faces
& 6 & 6 & 6 & 6 & 5 & 3 & \x & \x & \x \\
13th & \plus9/\plus4 & \plus15 & \plus6 & \plus10 & Greater wild aspect
& 6 & 6 & 6 & 6 & 6 & 4 & \x & \x & \x \\
14th & \plus10/\plus5 & \plus16 & \plus7 & \plus11 & Timeless body
& 6 & 6 & 6 & 6 & 6 & 5 & 3 & \x & \x \\
15th & \plus11/\plus6/\plus1 & \plus17 & \plus7 & \plus12 & Greater wild aspect
& 6 & 6 & 6 & 6 & 6 & 6 & 4 & \x & \x \\
16th & \plus12/\plus7/\plus2 & \plus18 & \plus8 & \plus13 & Greater wild speech
& 6 & 6 & 6 & 6 & 6 & 6 & 5 & 3 & \x \\
17th & \plus12/\plus7/\plus2 & \plus19 & \plus8 & \plus13 & Greater wild aspect
& 6 & 6 & 6 & 6 & 6 & 6 & 6 & 4 & \x \\
18th & \plus13/\plus8/\plus3 & \plus20 & \plus9 & \plus14 & Totemic aspect
& 6 & 6 & 6 & 6 & 6 & 6 & 6 & 5 & 3 \\
19th & \plus14/\plus9/\plus4 & \plus21 & \plus9 & \plus15 & Greater wild aspect
& 6 & 6 & 6 & 6 & 6 & 6 & 6 & 6 & 4 \\
20th & \plus15/\plus10/\plus5 & \plus22 & \plus10 & \plus16 & Greater natural casting
& 6 & 6 & 6 & 6 & 6 & 6 & 6 & 6 & 6 \\
\end{tabularx}
\end{dtable*}

\cf{Drd}{Spells} A druid gains additional spells per day and spells known as she increases in level.

\begin{dtable}
    \lcaption{Spells per Day} 
    \centering
    \begin{tabularx}{\columnwidth}{>{\ccol}X *{10}{>{\ccol}p{\spellcol}}}
        \thead{Level} & \thead{1st} & \thead{2nd} & \thead{3rd} & \thead{4th} & \thead{5th} & \thead{6th} & \thead{7th} & \thead{8th} & \thead{9th} \\
        1st & 3 & \x & \x & \x & \x & \x & \x & \x & \x \\
        2nd & 4 & \x & \x & \x & \x & \x & \x & \x & \x \\
        3rd & 5 & \x & \x & \x & \x & \x & \x & \x & \x \\
        4th & 6 & 3 & \x & \x & \x & \x & \x & \x & \x \\
        5th & 6 & 4 & \x & \x & \x & \x & \x & \x & \x \\
        6th & 6 & 5 & 3 & \x & \x & \x & \x & \x & \x \\
        7th & 6 & 6 & 4 & \x & \x & \x & \x & \x & \x \\
        8th & 6 & 6 & 5 & 3 & \x & \x & \x & \x & \x \\
        9th & 6 & 6 & 6 & 4 & \x & \x & \x & \x & \x \\
        10th & 6 & 6 & 6 & 5 & 3 & \x & \x & \x & \x \\
        11th & 6 & 6 & 6 & 6 & 4 & \x & \x & \x & \x \\
        12th & 6 & 6 & 6 & 6 & 5 & 3 & \x & \x & \x \\
        13th & 6 & 6 & 6 & 6 & 6 & 4 & \x & \x & \x \\
        14th & 6 & 6 & 6 & 6 & 6 & 5 & 3 & \x & \x \\
        15th & 6 & 6 & 6 & 6 & 6 & 6 & 4 & \x & \x \\
        16th & 6 & 6 & 6 & 6 & 6 & 6 & 5 & 3 & \x \\
        17th & 6 & 6 & 6 & 6 & 6 & 6 & 6 & 4 & \x \\
        18th & 6 & 6 & 6 & 6 & 6 & 6 & 6 & 5 & 3 \\
        19th & 6 & 6 & 6 & 6 & 6 & 6 & 6 & 6 & 4 \\
        20th & 6 & 6 & 6 & 6 & 6 & 6 & 6 & 6 & 6 \\
    \end{tabularx}
\end{dtable}

\begin{dtable}
\lcaption{Druid Spells Known}
\centering
\begin{tabularx}{\columnwidth}{X *{10}{p{1.1em}}}
& \multicolumn{10}{c}{\thead{---{}---{}---{}---{}---{}---{}---Spells Known---{}---{}---{}---{}---{}---{}---}} \\
\thead{Level} & \thead{1st} & \thead{2nd} & \thead{3rd} & \thead{4th} & \thead{5th} & \thead{6th} & \thead{7th} & \thead{8th} & \thead{9th} \\
1st  & 1 & \x & \x & \x & \x & \x & \x & \x & \x \\
2nd  & 2 & \x & \x & \x & \x & \x & \x & \x & \x \\
3rd  & 3 & \x & \x & \x & \x & \x & \x & \x & \x \\
4th  & 3 & 1 & \x & \x & \x & \x & \x & \x & \x \\
5th  & 4 & 2 & \x & \x & \x & \x & \x & \x & \x \\
6th  & 4 & 2 & 1 & \x & \x & \x & \x & \x & \x \\
7th  & 4 & 3 & 2 & \x & \x & \x & \x & \x & \x \\
8th  & 4 & 3 & 2 & 1 & \x & \x & \x & \x & \x \\
9th  & 4 & 3 & 3 & 2 & \x & \x & \x & \x & \x \\
10th & 4 & 3 & 3 & 2 & 1 & \x & \x & \x & \x \\
11th & 4 & 3 & 3 & 3 & 2 & \x & \x & \x & \x \\
12th & 4 & 3 & 3 & 3 & 2 & 1 & \x & \x & \x \\
13th & 4 & 3 & 3 & 3 & 3 & 2 & \x & \x & \x \\
14th & 4 & 3 & 3 & 3 & 3 & 2 & 1 & \x & \x \\
15th & 4 & 3 & 3 & 3 & 3 & 3 & 2 & \x & \x \\
16th & 4 & 3 & 3 & 3 & 3 & 3 & 2 & 1 & \x \\
17th & 4 & 3 & 3 & 3 & 3 & 3 & 2 & 2 & \x \\
18th & 4 & 3 & 3 & 3 & 3 & 3 & 2 & 2 & 1 \\
19th & 4 & 3 & 3 & 3 & 3 & 3 & 2 & 2 & 2 \\
20th & 4 & 3 & 3 & 3 & 3 & 3 & 2 & 2 & 2
\end{tabularx}
\end{dtable}

\cf{Drd}{Wild Aspect} At 7th and 9th level, a druid learns how to take on a new wild aspect. She may choose from the full list of wild aspects, given below. 

\subcf{Animal}
While embodying the animal aspect, the druid gains an enhancement bonus to any attribute score of her choice equal to the number of animal abilities she possesses.
\begin{wildaspect}
\wilditem Low-Light Vision: The druid gains low-light vision.
\wilditem Scent: The druid gains the scent ability.
\wilditem Natural Attunement: The druid gains a \plus4 enhancement bonus to Creature Handling, Ride, and Survival checks.
\wilditem Animal Affinity: When the druid first acquires this ability, she may choose any wild aspect for which she qualifies. She may learn it and treat it as if it were an animal aspect.
\end{wildaspect}

\subcf{Ape}
While embodying the ape aspect, the druid gains an enhancement bonus to Strength equal to the number of ape abilities she possesses.
\begin{wildaspect}
\wilditem Climb: The druid gains a climb speed equal to her base land speed.
\wilditem Claws: The druid gains two claw attacks which can be used as light natural weapons. A Medium druid deals d6 damage with each claw.
\wilditem Rend: If the druid hits with both claw attacks, she latches on to her opponent's body and tears the flesh. This attack deals damage appropriate to a scimitar appropriate for the druid's size.
\wilditem Improved Grab: When the druid hits a foe with an unarmed strike or natural attack, she may attempt to grapple her foe as a swift action without provoking an attack of opportunity.
\end{wildaspect}

\subcf{Bear}
While embodying the bear aspect, the druid gains an enhancement bonus to Constitution equal to the number of bear abilities she possesses.
\begin{wildaspect}
\wilditem Bite: The druid gains a bite attack which can be used as a heavy natural weapon. A Medium druid deals d8 damage with the bite.
\wilditem Claws: The druid gains two claw attacks which can be used as light natural weapons. A Medium druid deals d6 damage with each claw.
\wilditem
\wilditem Improved Grab: When the druid hits a foe with an unarmed strike or natural attack, she may attempt to grapple her foe as a swift action without provoking an attack of opportunity.
\end{wildaspect}

\subcf{Bull}
While embodying the bull aspect, the druid gains an enhancement bonus to Strength equal to the number of bull abilities she possesses.
\begin{wildaspect}
\wilditem Lope: The druid gains the ability to move on all four legs. When doing so, she gains a \plus20 foot competence bonus to her speed, but she cannot use her hands for anything except moving. When not moving on four legs, her ability to use her hands is unchanged.
\wilditem Gore: The druid gains a gore attack which can be used as a heavy natural weapon. A Medium druid deals d8 damage with a gore.
\wilditem Rush: When the druid hits a foe with an unarmed strike or natural attack on a charge, she may attempt to bull rush her foe as a swift action without provoking an attack of opportunity.
\wilditem
\end{wildaspect}

\subcf{Cat}
While embodying the cat aspect, the druid gains an enhancement bonus to Dexterity equal to the number of cat abilities she possesses.
\begin{wildaspect}
\wilditem Lope: The druid gains the ability to move on all four legs. When doing so, she gains a \plus20 foot competence bonus to her speed, but she cannot use her hands for anything except moving. When not moving on four legs, her ability to use her hands is unchanged.
\wilditem Bite: The druid gains a bite attack which can be used as a heavy natural weapon. A Medium druid deals d8 damage with a bite.
\wilditem Stealth: The druid gains a \plus4 enhancement bonus to Stealth checks.
\wilditem Pounce: During the first round of combat, the druid can make a full attack after charging. She gains no bonus to attack rolls from charging, but still takes the normal penalty to AC if applicable.
\end{wildaspect}

\subcf{Eagle}
While embodying the eagle aspect, the druid gains an enhancement bonus to Charisma equal to the number of eagle abilities she possesses.
\begin{wildaspect}
\wilditem Wings, partial: The druid gains a glide speed equal to her base land speed. While gliding, she cannot use her hands for anything except moving.
\wilditem Talons: The druid gains talons which can be used as a heavy natural weapon. A Medium druid deals d8 damage with her talons.
\wilditem Dive: When the druid hits a foe with an unarmed strike or natural attack on a charge while gliding or flying down, she deals double damage. If she can make multiple attacks on the charge, this effect only applies to the first attack.
\wilditem Wings, full: The druid gains a fly speed equal to her base land speed with average maneuverability. While flying she cannot use her hands for anything except moving. She can only fly for a number of rounds equal 3 \add half her Constitution. After that limit is reached, she must rest for 5 minutes to recuperate.
\end{wildaspect}

\subcf{Fox}
While embodying the fox aspect, the druid gains an enhancement bonus to Intelligence equal to the number of fox abilities she possesses.
\begin{wildaspect}
\wilditem Lope: The druid gains the ability to move on all four legs. When doing so, she gains a \plus20 foot competence bonus to her speed, but she cannot use her hands for anything except moving. When not moving on four legs, her ability to use her hands is unchanged.
\wilditem Bite: The druid gains a bite attack which can be used as a primary natural weapon. A Medium druid deals d8 damage with a bite.
\wilditem Stealth: The druid gains a \plus4 enhancement bonus on Stealth checks.
\wilditem
\end{wildaspect}

\subcf{Owl}
For each owl ability she possesses, the druid gains a \plus1 enhancement bonus to Wisdom  while embodying the owl aspect.
\begin{wildaspect}
\wilditem Wings, partial: The druid gains a glide speed equal to her base land speed. While gliding, she cannot use her hands for anything except moving.
\wilditem Talons: The druid gains talons which can be used as a heavy natural weapon. A Medium druid deals d8 damage with her talons.
\wilditem Senses: The druid gains a \plus4 enhancement bonus on Perception checks.
\wilditem Wings, full: The druid gains a fly speed equal to her base land speed with average maneuverability. While flying, she cannot use her hands for anything except moving. She can fly for a number of rounds equal to 3 \add half her Constitution. After that, she must rest for 5 minutes to recuperate.
\end{wildaspect}

\subcf{Serpent}
While embodying the serpent aspect, the druid gains an enhancement bonus to grapple attacks and saving throw DCs with any poison-based ability she uses equal to the number of serpent abilities she possesses.
\begin{wildaspect}
\wilditem Slither: The druid gains a climb speed equal to half her base land speed. She does not need to use her hands to climb in this way.
\wilditem Bite: The druid gains a bite attack which can be used as a heavy natural weapon. A Medium druid deals d8 damage with a bite.
\wilditem Constrict: After making a successful grapple attack to grapple or damage her foe, the druid may constrict her foe as a swift action. A Medium druid deals d8 \add 1-1/2 her Strength when constricting.
\wilditem Venom: The druid's natural attacks, and any weapons she wields, become coated in poison. The poison deals initial and secondary damage of 1d4 Constitution damage. A Fortitude save (DC 10 \add 1/2 the druid's level \add the druid's Con) negates the damage as normal for poison. This ability lasts for one round per druid level.
\end{wildaspect}

\subcf{Wolf}
While embodying the wolf aspect, the druid gains an circumstance bonus to weapon damage against overwhelmed foes and an enhancement bonus to trip attacks equal to the number of wolf abilities she possesses.
\begin{wildaspect}
\wilditem Lope: The druid gains the ability to move on all four legs. When doing so, she gains a \plus20 foot competence bonus to her speed, but she cannot use her hands for anything except moving. When not moving on four legs, her ability to use her hands is unchanged.
\wilditem Bite: The druid gains a bite attack which can be used as a heavy natural weapon. A Medium druid deals d8 damage with a bite.
\wilditem Trip: When the druid hits a foe with an unarmed strike or natural attack, she may attempt to trip her foe as a swift action without provoking an attack of opportunity.
\wilditem Wolfpack: Foes overwhelmed by the druid increase their overwhelm penalties by 1.
\end{wildaspect}

\cd{Drd}{Wild Speech (Plants) (Ex)} At 6th level, a druid can also converse with plants and plant creatures using her wild speech ability. A regular plant's sense of its surroundings is limited, so it won't be able to give (or recognize) detailed descriptions of creatures or answer questions about events outside its immediate vicinity.

\cf{Drd}{Improved Wild Speech (Su)} At 8th level, anything the druid speaks with with using her wild speech ability must make a Will save to avoid being charmed, as the \spell{charm person} spell, by the druid. The effect lasts for the duration of the conversation, and for 1 hour thereafter. This ability is not mind-affecting, and can affect creatures or even objects of any kind that the druid can converse with. Objects are always considered to fail their Will save.  The druid can choose not to exert this influence.

\cf{Drd}{Improved Natural Casting (Ex)} At 10th level, a druid can cause area spells to originate from up to \rngclose range away from her, as the natural casting ability.

\cf{Drd}{Greater Wild Aspect (Su)} At 11th level, a druid gains the ability to assume aspects of the natural world, including the elements, in addition to those of animals. This ability functions as wild aspect, and assuming a greater wild aspect consumes a use of wild aspect, but the druid may choose from a different list of abilities. Unlike with wild aspects, greater wild aspects must be learned in order within an aspect; a druid cannot gain the air mantle aspect unless she has the endless air aspect. A druid can suppress or resume any greater wild aspect ability as a swift action.

Whenever the druid learns a greater wild aspect, she may choose to learn a wild aspect instead.
\subcf{Air}
\begin{greaterwildaspect}
\wilditem Profusion of Air: The druid constantly exudes good, clean air. She can breathe in any environment, and is immune to \spell{sickening cloud} and similar effects. In addition, she may use her wild speech ability to speak with any natural air.
\wilditem Air Mantle: The druid is surrounded by a mantle of air. Thrown and projectile weapons have a 50\% chance to miss her while this effect is active. Unusually large weapons, such as a giant's boulders, may suffer a decreased miss chance as appropriate to their size.
\wilditemplus Flight: The druid gains a fly speed equal to her land speed, with good maneuverability. She may remain flying for up to 5 rounds at a time. After that, she must land for 1 round before she can fly again.
\end{greaterwildaspect}

\subcf{Earth}
\begin{greaterwildaspect}
\wilditem Earthen Profusion: The druid constantly exudes fresh, solid earth wherever she steps. She gains a \plus4 enhancement bonus to CMD and Fortitude saves against effects that would move her. In addition, she may use her wild speech ability to speak with any natural earth or stone.
\wilditem Earth Mantle: The druid is surrounded by a mantle of earth. She gains a \plus1 enhancement bonus to natural armor class per four druid levels and damage reduction 5/adamantine.
\wilditemplus Earth Glide: The druid gains the earth glide ability, as an earth elemental. She may remain partly within the earth while fighting, granting her cover at no penalty to her own actions.
\end{greaterwildaspect}

\subcf{Fire}
\begin{greaterwildaspect}
\wilditem Flaming Profusion: Wherever the druid moves, she leaves a path of burning flame behind her that lasts until the end of her next turn. A creature who crosses the path takes 1d6 points of fire damage per two druid levels. It can make a Reflex save to halve the damage, with a DC equal to 10 \add the druid's level \add the druid's Charisma. In addition, the druid can her her wild speech ability to speak with any natural fire.
\wilditem Fire Mantle: The druid is surrounded by a mantle of fire. This functions as a warm \spell{fire shield} spell, with a caster level equal to the druid's level.
\wilditemplus Immolation: The druid gains the fire subtype, making her immune to fire but giving her a 50\% vulnerability to cold damage. Whenever she deals fire damage to a creature, she ignites the creature for 5 rounds. An ignited creature takes a \minus2 penalty to attack rolls, saving throws, checks, and AC, and takes d6 damage per round from the fire. If the creature takes a full-round action, it can attempt a Reflex save to put out the fire with a DC of 10 \add the druid's level \add the druid's Charisma. A creature hit by the druid's fire multiple times is not ignited multiple times; only the most recent effect is used.
\end{greaterwildaspect}

\subcf{Plant}
\begin{greaterwildaspect}
\wilditem Profusion of Life: Wherever the druid moves, she leaves a path of small, living plants that entangle foes until the end of her next turn. A creature crossing the path must make a Reflex save with a DC equal to 10 \add the druid's level \add the druid's Charisma or be entangled by the plants and unable to complete the movement. The plants appear on any surface, and will continue to grow if they can survive, though they may die quickly if they appear on inhospitable terrain.
\wilditem Plant Body: The druid's body takes on plantlike characteristics. She gains a \plus1 enhancement bonus to natural armor class per four druid levels and has a 50\% chance to ignore critical hits and sneak attacks.
\wilditemplus Rejuvenation: The druid gains fast healing 5 as long as she remains in sunlight or touches a plant of her size or larger.
\end{greaterwildaspect}

\subcf{Sun}
\begin{greaterwildaspect}
\wilditem Profusion of Light: The druid constantly radiates light in a \arealarge radius. This is treated as true sunlight, not ordinary magical light. Creatures and objects vulnerable to sunlight must make a Fortitude save every round to resist the effect on themselves, with a DC equal to 10 \add the druid's level \add the druid's Charisma.
\wilditem Mantle of Light: The druid glows so brightly that she becomes hard to look at. She gains concealment against all attacks, and any creature attacking her from within the radius of her profusion of light is dazzled for 5 rounds after the attack (no save). She cannot use this ability while suppressing her profusion of light ability, and she cannot use this concealment to hide.
\wilditemplus Piercing Radiance: The druid's illumination radius with her profusion of light ability increases to 100 feet. All visual illusions and shadow effects within the radius are suppressed except those that the druid chooses to allow. Any creature within the radius who attacks the druid is blinded for 1 round after the attack. A successful Fortitude save with a DC of 10 \add the druid's level \add the druid's Charisma prevents the creature from being blinded for the next round.
\end{greaterwildaspect}

\subcf{Water}
\begin{greaterwildaspect}
\wilditem Aqueous Profusion: Wherever the druid moves, she leaves a path of animated water that can grab creatures and cause them to trip. A creature crossing the path must make a Reflex save with a DC equal to 10 \add the druid's level \add the druid's Charisma or fall prone and be unable to complete the movement. In addition, the druid can speak with natural water using her wild speech ability.
\wilditem Watery Mantle: The druid becomes surrounded by an animate mantle of water that reaches out to deflect incoming blows. She gains a \plus10 enhancement bonus to resist grapple attacks and a \plus5 shield bonus.
\wilditemplus Water's Flow: At any time during the duration of this aspect, the druid can transform into a rushing flow of water as a move action that does not provoke attacks of opportunity. As part of the action, she may move up to her movement speed in any direction that water could go. Her speed is halved when moving uphill and doubled when moving downhill. She does not provoke attacks of opportunity during this movement, and has physical damage reduction 10 while in this form. At the end of her movement, she regains her normal form.
\end{greaterwildaspect}

\cf{Drd}{A Thousand Faces (Su)} At 12th level, a druid gains the ability to change
her appearance at will, as if using the \spell{disguise self} spell. This affects the druid's body but not her possessions. It is not an illusory effect, but a minor physical alteration of the druid's appearance, within the limits described for the spell.

\cf{Drd}{Timeless Body (Ex)} After attaining 15th level, a druid no longer takes attribute score penalties for aging and cannot be magically aged. Any penalties she may have already incurred, however, remain in place. Bonuses still accrue, and the druid still dies of old age when her time is up.

\cf{Drd}{Greater Wild Speech (Ex)} At 16th level, the druid can use her wild speech ability to control creatures' actions. If the druid converses with a creature using her wild speech ability, she may spend a standard action and an additional wild speech use to dominate one creature she is speaking with, as the \spell{dominate person} spell. A successful Will save negates this effect. This ability is not mind-affecting, and can affect creatures of any kind that the druid can converse with.

\cf{Drd}{Totemic Aspect (Su)} At 18th level, the druid can choose any one wild aspect (but not greater aspect). She permanently gains the abilities of that aspect, as if she was constantly manifesting it. She may suppress or resume this effect as a swift action. If the druid has multiple abilites from that aspect, she may suppress or resume them each individually.

\cf{Drd}{Greater Natural Casting (Ex)} At 20th level, a druid may cause area spells to originate from any point within \rngmed range of her, as the natural casting ability.

\subsection{Fighter}
\begin{dtable}
\lcaption{The Fighter}
\begin{tabularx}{\columnwidth}{>{\ccol}p{\levelcol} >{\ccol}p{\babcolgood} *{3}{>{\ccol}p{\savecol}} >{\lcol}X}
\thead{Level} & \thead{Base Attack Bonus} & \thead{Fort Save} & \thead{Ref Save} & \thead{Will Save} & \thead{Special} \\
1st & \plus1                         & \plus3 & \plus0 & \plus1 & Armor discipline \\
2nd & \plus2                         & \plus4 & \plus1 & \plus2 & Bonus feat \\
3rd & \plus3                         & \plus5 & \plus1 & \plus3 & Weapon discipline \\
4th & \plus4                         & \plus6 & \plus2 & \plus4 & Adaptive style feat \\
5th & \plus5                         & \plus7 & \plus2 & \plus4 & Combat discipline \\
6th & \plus6/\plus1                  & \plus8 & \plus3 & \plus5 & Bonus feat \\
7th & \plus7/\plus2                  & \plus9 & \plus3 & \plus6 & Improved armor discipline \\
8th & \plus8/\plus3                  & \plus10& \plus4 & \plus7 & Adaptive style feat \\
9th & \plus9/\plus4                  & \plus11& \plus4 & \plus7 & Improved weapon discipline\\
10th & \plus10/\plus5                & \plus12& \plus5 & \plus8 & Battlemaster, bonus feat\\
11th & \plus11/\plus6/\plus1         & \plus13 & \plus5 & \plus9 & Improved combat discipline\\
12th & \plus12/\plus7/\plus2         & \plus14 & \plus6 & \plus10& Adaptive style feat \\
13th & \plus13/\plus8/\plus3         & \plus15 & \plus6 & \plus10& Greater armor discipline \\
14th & \plus14/\plus9/\plus4         & \plus16 & \plus7 & \plus11& Bonus feat, improved adaptive style \\
15th & \plus15/\plus10/\plus5        & \plus17 & \plus7 & \plus12& Greater weapon discipline\\
16th & \plus16/\plus11/\plus6/\plus1 & \plus18 & \plus8 & \plus13& Adaptive style feat \\
17th & \plus17/\plus12/\plus7/\plus2 & \plus19 & \plus8 & \plus13& Greater combat discipline\\
18th & \plus18/\plus13/\plus8/\plus3 & \plus20 & \plus9 & \plus14& Bonus feat, improved battlemaster\\
19th & \plus19/\plus14/\plus9/\plus4 & \plus21 & \plus9 & \plus15& True discipline \\
20th & \plus20/\plus15/\plus10/\plus5& \plus22 & \plus10 & \plus16 & Adaptive style feat, greater adaptive style
\end{tabularx}
\end{dtable}

\cf{Ftr}{Bonus Feats} The fighter gains an additional bonus feat at 6th level and every four fighter levels thereafter (6th, 10th, 14th, and 18th).

\cf{Ftr}{Adaptive Style Feats} The fighter gains an additional adaptive style feat at 8th level and every four fighter levels thereafter (8th, 12th, 16th, and 20th).

\cf{Ftr}{Improved Armor Discipline} At 7th level, a fighter's training in his chosen armor category (or with all armor categories) improves. He reduces the armor check penalty by 4 and decreases the arcane spell failure by 15\%. In addition, he treats his chosen armor (or armors) as if it were one encumbrance category lighter than it is. This does not stack with the effects of armor discipline.
\par This ability means heavy armor is treated as medium, medium armor is treated as light armor, and light armor is treated as being unarmored. Likewise, tower shields are treated as heavy shields (and no longer impose a \minus2 penalty to attack rolls), heavy shields are treated as light shields, and both light shields and bucklers are treated as being unarmored. This can remove the halving of the fighter's Dexterity bonus, if appropriate for the new encumbrance of the fighter's armor.
\par This allows the fighter to qualify for class features using the reduced armor encumbrance category. For example, a fighter 9 / wizard 2 who reduces his encumbrance in light armor could cast without any arcane spell failure in light armor.
\par A fighter who chose a specific armor category gains a \plus2 competence bonus to his dodge modifier while using armor of that category.

\cf{Ftr}{Improved Weapon Discipline} At 9th level, a fighter's training in his chosen weapons improves. He gains a \plus4 competence bonus to resist disarm and sunder attempts when using his chosen weapons. If he chose a specific weapon group, he gains a \plus2 competence bonus to attack rolls with weapons from that group. If he did not, he can apply all weapon group-specific feats he has to any weapon group that he trains with for 8 hours. He retains this benefit for one week after the training.

\cf{Ftr}{Battlemaster} At 10th level, a fighter can improve his allies' combat abilities. As a standard action, he may grant the use one of his combat feats to allies within 30 feet of him who can see and hear him. He can affect a number of allies equal to 1 \add his Intelligence (minimum 1). Affected allies must meet all prerequisites for the granted feat, except that they can ignore any feat prerequisites. The effect lasts for 5 rounds. The fighter can use this ability a number of times per day equal to 3 \add his Charisma.

\cf{Ftr}{Improved Combat Discipline} At 11th level, a fighter's ability to keep fighting despite negative influence improves. He can reduce conditions by two steps instead of one when using combat discipline. For example, a stunned fighter who used combat discipline would instead be staggered.
\par In addition, a fighter may use combat discipline to reduce any penalties he suffers to attribute scores, attack rolls, weapon damage rolls, skill checks, or ability checks that come from negative effects not listed on the combat discipline chart by 2. This cannot be used to reduce the effects of ability damage or drain.

\cf{Ftr}{Greater Armor Discipline} At 13th level, a fighter's training in his chosen armor becomes still greater. He reduces his armor check penalty by 6 and decreases his arcane spell failure by 30\% when using his chosen armor. In addition, he treats his chosen armor as if it were two encumbrance categories lighter than it actually is whenever doing so would be beneficial to him. This does not stack with the benefits of armor discipline or improved armor discipline.
\par A fighter who chose a specific armor category gains a \plus3 competence bonus to his dodge modifier while using armor of that category.

\cf{Ftr}{Improved Adaptive Style} At 14th level, a fighter's ability to adapt to situations improves. He need only spend 1 minute training to change a single adaptive style feat. He may continue training as he wishes, changing one adaptive style feat per minute.

\cf{Ftr}{Greater Weapon Discipline} At 15th level, a fighter's training in his chosen weapons becomes still greater. He increases the critical threat range and critical multiplier of his chosen weapons by 1. This increase applies after and stacks with any other effects that affect critical threat range or critical multiplier. Thus, a fighter using the Heartseeker combat style and wielding a longsword would have a critical threat range of 16-20 (x3), while a similar fighter would have a critical threat range of 18-20 (x5) with a heavy pick.

\cf{Ftr}{Greater Combat Discipline} At 17th level, a fighter's ability to keep fighting despite influence becomes still greater. When using combat discipline, he may ignore any condition listed on the combat discipline chart. In addition, he may use combat discipline to be dazed rather than suffer any non-damaging condition not listed on the chart.

\cf{Ftr}{Improved Battlemaster} At 18th level, the fighter can improve his allies' combat abilities more effectively. When using his battlemaster ability, he can grant two feats at once. In addition, he can use his battlemaster ability as a swift action.

\cf{Ftr}{True Discipline} At 19th level, a fighter's discipline in his chosen area is beyond equal. He must choose either weapon discipline, armor discipline, or combat discipline. Depending on which discipline he chooses, he gains a different bonus.
\subcf{True Weapon Discipline} The fighter can take 10 on the first attack he makes each round and automatically confirms all critical threats while using his chosen weapons.
\subcf{True Armor Discipline} The fighter no longer suffers armor check penalties or arcane spell failure with his chosen armor. In addition, he treats his chosen armor as if it were three encumbrance categories lighter than it actually is.
\par A fighter who chose a specific armor category gains a \plus4 competence bonus to his dodge modifier while using armor of that category.
\subcf{True Combat Discipline} The fighter can use combat discipline to be staggered instead of suffering any nondamaging negative effect with a duration.

\cf{Ftr}{Greater Adaptive Style} At 20th level, a fighter's ability to react to situations is unparalleled. He need only spend a full-round action training to exchange an adaptive style feat. He may continue training as he wishes, changing one adaptive style feat per round.

\subsection{Monk}
\begin{dtable*}
\lcaption{The Monk}
\begin{tabularx}{\textwidth}{>{\ccol}p{\levelcol} >{\ccol}p{\babcolavg} *{3}{>{\ccol}p{\savecol}} >{\lcol}X >{\ccol}p{4.5em} >{\ccol}p{6em}}
\thead{Level} & \thead{Base Attack Bonus} & \thead{Fort Save} & \thead{Ref Save} & \thead{Will Save} & \thead{Special} & \thead{Unarmed Damage} & \thead{Unarmored Speed Bonus} \\
1st & \plus0                    & \plus1 & \plus3 & \plus3    & Enlightened defense, flurry of blows, unarmed strike & 1d6  & \plus0 ft. \\
2nd & \plus1                    & \plus2 & \plus4 & \plus4    & Bonus feat, \ki strike, uncanny dodge                                  & 1d6  & \plus0 ft. \\
3rd & \plus2                    & \plus3 & \plus5 & \plus5    & Bonus feat, still mind, wholeness of body                               & 1d6  & \plus10 ft. \\
4th & \plus3                    & \plus4 & \plus6 & \plus6    & Evasion, slow fall                         & 1d8  & \plus10 ft. \\
5th & \plus3                    & \plus4 & \plus7 & \plus7    & \Ki strike (magic)                                   & 1d8  & \plus10 ft. \\
6th & \plus4                    & \plus5 & \plus8 & \plus8    & Bonus feat, improved uncanny dodge                                          & 1d8  & \plus20 ft. \\
7th & \plus5                    & \plus6 & \plus9 & \plus9    & Diamond body                                         & 1d8  & \plus20 ft. \\
8th & \plus6/\plus1             & \plus7 & \plus10& \plus10   & Tongue of the sun and moon                           & 1d10 & \plus20 ft. \\
9th & \plus6/\plus1             & \plus7 & \plus11& \plus11   & Improved evasion                                     & 1d10 & \plus30 ft. \\
10th & \plus7/\plus2            & \plus8 & \plus12& \plus12   & Greater uncanny dodge, \ki strike (lawful)                   & 1d10 & \plus30 ft. \\
11th & \plus8/\plus3            & \plus9  & \plus13 & \plus13 & Abundant step                                      & 1d10 & \plus30 ft. \\
12th & \plus9/\plus4            & \plus10 & \plus14& \plus14 & Diamond soul                                         & 2d6  & \plus40 ft. \\
13th & \plus9/\plus4            & \plus10 & \plus15 & \plus15 & Quivering palm                                       & 2d6  & \plus40 ft. \\
14th & \plus10/\plus5           & \plus11 & \plus16& \plus16 & Empty step                                           & 2d6  & \plus40 ft. \\
15th & \plus11/\plus6/\plus1    & \plus12 & \plus17 & \plus17 & Timeless body                                        & 2d6  & \plus50 ft. \\
16th & \plus12/\plus7/\plus2    & \plus13 & \plus18 & \plus18 & \Ki strike (adamantine)                              & 2d8  & \plus50 ft. \\
17th & \plus12/\plus7/\plus2    & \plus13 & \plus19 & \plus19 & Moment of perfection                                 & 2d8  & \plus50 ft. \\
18th & \plus13/\plus8/\plus3    & \plus14 & \plus20 & \plus20 & Empty body                                           & 2d8  & \plus60 ft. \\
19th & \plus14/\plus9/\plus4    & \plus15 & \plus21 & \plus21 & Empty soul                                           & 2d8  & \plus60 ft. \\
20th & \plus15/\plus10/\plus5   & \plus16 & \plus22 & \plus22 & \Ki strike (epic), perfect self                      & 2d10 & \plus60 ft. \\
\end{tabularx}
\end{dtable*}

\cf{Mnk}{Unarmed Strike} The monk's unarmed damage continues to increase as she gains levels.

\begin{dtable}
\lcaption{Small or Large Monk Unarmed Damage}
\begin{tabularx}{\columnwidth}{>{\lcol}p{5em} >{\lcol}X >{\lcol}X}
\thead{Level} & \thead{Damage (Small Monk)} & \thead{Damage (Large Monk)} \\
1st-3rd  & 1d4 & 1d8 \\
4th-7th  & 1d6 & 1d10 \\
8th-11th & 1d8 & 2d6 \\
12th-15th & 1d10 & 2d8 \\
16th-19th & 2d6 & 2d10 \\
20th   & 2d8 & 4d6 \\
\end{tabularx}
\end{dtable}

\cf{Mnk}{Bonus Feat} At 6th level, the monk may select any combat maneuver feat that she qualifies for.

\cf{Mnk}{Improved Uncanny Dodge (Ex)} At 6th level and higher, a monk can no longer be overwhelmed as easily; she can react to multiple opponents as easily as she can react to a single attacker. The monk is always treated as being threatened by two fewer creatures than she actually is for the purpose of determining overwhelm penalties.
\par If a character already has uncanny dodge (see above) from a
second class and gains improved uncanny dodge, the character stacks those levels to determine if she should gain greater uncanny dodge, and to determine the minimum level a rogue must be to flank the character.

\cf{Mnk}{Diamond Body (Su)} At 7th level, a monk gains immunity to poisons and diseases of all kinds, including supernatural and magical diseases (such as mummy rot and lycanthropy).

\cf{Mnk}{Tongue of the Sun and Moon (Ex)} A monk of 8th level or higher can speak to and understand the speech of any living creature. This grants her no special ability to speak to or understand creatures that do not speak, such as animals.

\cf{Mnk}{Improved Evasion (Ex)} At 9th level, a monk's evasion ability improves. She still takes no damage on a successful Reflex saving throw against attacks, but henceforth she takes only half damage on a failed save. A helpless monk does not gain the benefit of improved evasion.

\cf{Mnk}{Greater Uncanny Dodge (Ex)} At 10th level and higher, a monk can no longer be overwhelmed, regardless of the number of foes surrounding her.

\cf{Mnk}{Improved Ki Strike (Su)} At 10th level, a monk may add half her Wisdom to damage with unarmed attacks and monk weapons. In addition, those weapons are treated as being lawful for the purpose of dealing damage to creatures with damage reduction.

\cf{Mnk}{Abundant Step (Su)} At 11th level or higher, a monk can slip magically between spaces, as if using the spell \spell{dimension slide}, a number of times per day equal to half her Wisdom. Using this ability is a move action that does not provoke attacks of opportunity. Her caster level for this effect is equal to her monk level.

\cf{Mnk}{Diamond Soul (Ex)} At 12th level, a monk gains spell resistance. A creature with spell resistance may always make a saving throw when a spell is cast on it. The saving throw type is indicated by the spell. If it succeeds, the spell has no effect on it.

\cf{Mnk}{Quivering Palm (Su)} Starting at 13th level, a monk can set up vibrations within the body of another creature that can thereafter be fatal if the monk so desires. She can use this quivering palm attack a number of times per day equal to her Wisdom, but no more than once per round, and she must announce her intent before making her attack roll. Constructs, oozes, plants, undead, incorporeal creatures, and creatures immune to critical hits cannot be affected. Otherwise, if the monk strikes successfully and the target takes damage from the blow, the quivering palm attack succeeds. Thereafter the monk can try to slay the victim at any later time, as long as the attempt is made within a number of days equal to her monk level. To make such an attempt, the monk merely wills the target to die (a free action), and unless the target makes a Fortitude saving throw (DC 10 \add 1/2 the monk's level \add the monk's Wis), it dies. If the saving throw is successful, the target is no longer in danger from that particular quivering palm attack, but it may still be affected by another one at a later time.

\cf{Mnk}{Empty Step (Su)} At 14th level or higher, a monk can assume an ethereal state for brief periods, as if using the \spell{ethereal jaunt} spell, except that the effect only lasts for a single round, a number of times per day equal to her Wisdom. Using this ability is a swift action.

\cf{Mnk}{Ki Strike (Adamantine) (Su)} At 15th level, a monk's unarmed attacks and monk weapons are treated as adamantine weapons for the purpose of dealing damage to creatures with damage reduction and bypassing hardness.

\cf{Mnk}{Timeless Body (Ex)} Upon attaining 15th level, a monk no longer takes penalties to her attribute scores for aging and cannot be magically aged. Any such penalties that she has already taken, however, remain in place. Bonuses still accrue, and the monk still dies of old age when her time is up.

\cf{Mnk}{Moment of Perfection (Su)} At 17th level, a monk can align herself with the universe to achieve a single moment of perfection. She can add her monk level as an enhancement bonus to any one attack roll, opposed skill or ability check, or saving throw, or to her AC against any one attack, as if calling upon the effect of a \spell{moment of prescience} spell with a caster level equal to her monk level. She can use this ability a number of times per day equal to her Wisdom, but she must wait five minutes between each use of this ability.

\cf{Mnk}{Empty Body (Su)} At 18th level, a monk using her empty step ability can stay ethereal for a number of rounds equal to her monk level. She may dismiss the ability and become material as a swift action.

\cf{Mnk}{Empty Soul (Su)} At 19th level, a monk achieves a state of complete emptiness, reacting to all situations without premeditation or thought. She continuously gains the benefits of the \spell{foresight} spell upon her person at all times.

\cf{Mnk}{Ki Strike (Epic) (Su)} At 20th level, a monk's unarmed attacks and monk weapons are treated as epic weapons for the purpose of dealing damage to creatures with damage reduction.

\cf{Mnk}{Perfect Self} At 20th level, a monk becomes a magical creature. She is forevermore treated as an outsider rather than as a humanoid for the purpose of spells and magical effects   whenever doing so is advantageous to her. For instance, \spell{charm person} does not affect her, but she can still be affected by \spell{enlarge person}, and she can still be brought back from the dead.

Additionally, the monk gains damage reduction 10/chaotic, which allows her to ignore the first 10 points of damage from any attack made by a non-chaotic weapon or by any natural attack made by a creature that doesn't have similar damage reduction.

\cf{Mnk}{Slow Fall (Ex)} At 20th level, the monk can use a nearby object to slow her descent and fall any distance without harm.

\subsection{Paladin}
\begin{dtable*}
\lcaption{The Paladin}
\begin{tabularx}{\textwidth}{>{\ccol}p{\levelcol} >{\ccol}p{\babcolgood} *{3}{>{\ccol}p{\savecolpoof}} X *{4}{>{\ccol}p{\spellcolpoof}}}
& & & & & & \multicolumn{4}{c}{\thead{---{}---Spells per Day---{}---}} \\
\thead{Level} & \thead{Base Attack Bonus} & \thead{Fort Save} & \thead{Ref Save} & \thead{Will Save} & \thead{Special} & \thead{1st} & \thead{2nd} & \thead{3rd} & \thead{4th} \\
1st  & \plus1                        & \plus3  & \plus0 & \plus3 & Aura of good, discernment (evil), smite evil & \x & \x & \x & \x \\
2nd  & \plus2                        & \plus4  & \plus1 & \plus4 & Improved smite, lay on hands & \x & \x & \x & \x \\
3rd  & \plus3                        & \plus5  & \plus1 & \plus5 & Aura of courage, bulwark of defense, divine health & \x & \x & \x & \x \\ 
4th  & \plus4                        & \plus6  & \plus2 & \plus6 & Divine grace & 1 & \x & \x & \x \\
5th  & \plus5                        & \plus7  & \plus2 & \plus7 & Discernment (chaos), improved smite & 2 & \x & \x & \x \\
6th  & \plus6/\plus1                 & \plus8  & \plus3 & \plus8 & Aura of resolve, holy ward & 3 & \x & \x & \x \\
7th  & \plus7/\plus2                 & \plus9  & \plus3 & \plus9 & Smite chaos & 3 & \x & \x & \x \\
8th  & \plus8/\plus3                 & \plus10 & \plus4 & \plus10& Improved smite & 3 & 1 & \x & \x \\
9th  & \plus9/\plus4                 & \plus11 & \plus4 & \plus11& Aura of determination, improved bulwark of defense & 3 & 2 & \x & \x \\
10th & \plus10/\plus5                & \plus12 & \plus5 & \plus12& Discernment (lies), pass judgment & 3 & 3 & \x & \x \\
11th & \plus11/\plus6/\plus1         & \plus13 & \plus5 & \plus13& Improved smite & 4 & 3 & \x & \x \\
12th & \plus12/\plus7/\plus2         & \plus14 & \plus6 & \plus14& Aura of protection & 4 & 3 & 1 & \x \\
13th & \plus13/\plus8/\plus3         & \plus15 & \plus6 & \plus15& Forgiving smite & 4 & 3 & 2 & \x \\
14th & \plus14/\plus9/\plus4         & \plus16 & \plus7 & \plus16& Improved smite & 4 & 3 & 3 & \x \\
15th & \plus15/\plus10/\plus5        & \plus17 & \plus7 & \plus17& Aura of warding & 4 & 4 & 3 & \x \\
16th & \plus16/\plus11/\plus6/\plus1 & \plus18 & \plus8 & \plus18& Glory of the martyr & 4 & 4 & 3 & 1 \\
17th & \plus17/\plus12/\plus7/\plus2 & \plus19 & \plus8 & \plus19& Improved smite & 4 & 4 & 3 & 2 \\
18th & \plus18/\plus13/\plus8/\plus3 & \plus20 & \plus9 & \plus20& Greater aura of warding & 4 & 4 & 3 & 3 \\
19th & \plus19/\plus14/\plus9/\plus4 & \plus21 & \plus9 & \plus21& Martyr's retribution & 4 & 4 & 4 & 3 \\
20th & \plus20/\plus15/\plus10/\plus5& \plus22 &\plus10 &\plus22 & Greater smite, improved smite & 4 & 4 & 4 & 4 \\
\end{tabularx}
\end{dtable*}

\cf{Pal}{Spells} The paladin continues to gain new spells known as she gains in level.

\begin{dtable}
\lcaption{Paladin Spells Known}
\begin{tabularx}{\columnwidth}{X *{4}{>{\ccol}X}}
& \multicolumn{4}{c}{\thead{---{}---{}---{}---{}---Spells Known---{}---{}---{}---{}---}} \\
\thead{Level} & \thead{1st} & \thead{2nd} & \thead{3rd} & \thead{4th} \\
1st  & \x & \x & \x & \x \\
2nd  & \x & \x & \x & \x \\
3rd  & \x & \x & \x & \x \\
4th  & 1 & \x & \x & \x \\
5th  & 2 & \x & \x & \x \\
6th  & 2 & \x & \x & \x \\
7th  & 3 & \x & \x & \x \\
8th  & 3 & 1 & \x & \x \\
9th  & 3 & 2 & \x & \x \\
10th & 4 & 2 & \x & \x \\
11th & 4 & 2 & \x & \x \\
12th & 4 & 3 & 1 & \x \\
13th & 4 & 3 & 2 & \x \\
14th & 4 & 3 & 2 & \x \\
15th & 4 & 3 & 2 & \x \\
16th & 4 & 3 & 3 & 1 \\
17th & 4 & 3 & 3 & 2 \\
18th & 4 & 3 & 3 & 2 \\
19th & 4 & 3 & 3 & 2 \\
20th & 4 & 3 & 3 & 3 \\
\end{tabularx}
\end{dtable}

\cf{Pal}{Improved Smiting (Su)} At 8th level, the paladin adds the following improved smiting ability to the list of those that can be selected.
\subcf{Dazing} The paladin's smite shatters her foe's ability to concentrate. A creature struck by the smite must make a Fortitude save to avoid being dazed for one round.
\subcf{Impeding} The paladin's smite traps her foe in place, unable to escape her wrath. A creature struck by the smite must make a Reflex save to avoid having all its movement speeds reduced to 5 feet for one round.
\subcf{Coercing} The paladin's smite forces her foe to join the cause of righteousness, if only for a moment. A creature struck by the smite must make a Will save or else be affected by a \spell{suggestion}, as the spell, of the paladin's choice. The effect lasts for one round.

At 11th level, the paladin adds the following improved smiting ability to the list of those that can be selected.
\subcf{Axiomatic} The paladin's smite is filled with exceptionally lawful divine energy. The attack ignores all damage reduction of chaotic creatures. Aberrations and chaotic outsiders take extra damage equal to twice the paladin's level.

At 14th level, the paladin adds the following improved smiting ability to the list of those that can be selected.
\subcf{Dispelling} The paladin's smite strips away her foe's magical protections. A creature struck by the smite is subject to a targeted \spell{dispel magic} with a bonus equal to the paladin's level.

At 17th level, the paladin adds the following improved smiting ability to the list of those that can be selected.
\subcf{Brilliant} The paladin's smite cannot be turned aside by mortal defenses. The smite is made against the enemy's touch armor class.

\cf{Pal}{Aura of Resolve (Su)} At 6th level, the paladin becomes immune to charm effects. Each ally within 10 feet of her gains a \plus4 enhancement bonus on saving throws against charm effects.

\cf{Pal}{Holy Ward (Sp)} A paladin is first and foremost a defender of her allies, and relies upon \spell{shield other} to defend her allies against harm from threats she cannot block with her skill at arms. A paladin of 6th level or higher who spends at least 10 points of healing from her lay on hands ability in a single round can bestow a \spell{shield other} effect on the healed ally, using her paladin level as her caster level.

\cf{Pal}{Smite Chaos (Su)} At 7th level, the paladin gains the ability to smite chaotic creatures as well as evil creatures with her smite ability, but she must choose which to smite before making the attack. If the paladin attempts to smite a chaotic creature, and that creature is not chaotic, the smite attack deals no damage at all (not even normal weapon damage), but the use of the ability is still spent.

\cf{Pal}{Aura of Determination (Su)} At 9th level, the paladin becomes immune to compulsion effects. Each ally within 10 feet of her gains a \plus4 enhancement bonus on saving throws against compulsion effects.

\cf{Pal}{Discernment (Su)} At 10th level, the paladin may can also identify lies in the range of his discernment ability, as the \spell{discern lies} spell.

\cf{Pal}{Pass Judgment (Su)} At 10th level, the paladin gains the ability to freely pass judgment on those she deems unworthy. As a swift action, she may pass judgment on a creature within 100 feet of her once per day. The creature is treated as being evil, chaotic, or both, at the paladin's discretion, in place of their original alignment. This effect lasts for one day per paladin level, or until the paladin changes her mind about the subject (a free action). This does not change the creature's actions or behavior, but the creature is subject to smite evil or smite chaos, would register as evil under the inspection of a \spell{detect evil} spell, and so on.

No saving throw is allowed against this effect, and it cannot be dispelled, but a \spell{remove curse}, \spell{miracle}, or \spell{wish} spell can remove it. The paladin can use this ability an additional time per day at 13th level and every odd level thereafter. A paladin should be careful when using this ability, as persecution of the innocent can lead overzealous paladins to fall.

\cf{Pal}{Aura of Protection (Su)} At 12th level, the paladin continuously radiates a \spell{magic circle against evil}, as the spell.

\cf{Pal}{Forgiving Smite (Su)} At 13th level, if a paladin smites a creature who is not evil or chaotic, the smite attempt is not wasted.

\cf{Pal}{Glory of the Martyr (Su)} At 14th level, if the paladin dies while fighting evil or protecting her allies, her fallen body erupts in a burst of positive energy, granting all her allies within 100 feet the benefit of a \spell{heal} spell.

\cf{Pal}{Aura of Warding (Su)} At 15th level, the paladin continuously radiates a \spell{lesser globe of invulnerability}, as the spell. The effect travels with the paladin.

\cf{Pal}{Mighty Aura (Su)} At 16th level, the radius of a paladin's auras expands to 20 feet.

\cf{Pal}{Greater Aura of Warding (Su)} At 18th level, the paladin radiates a \spell{globe of invulnerability}, as the spell, instead of a \spell{lesser globe of invulnerability}. The effect continues to travel with the paladin.

\cf{Pal}{Martyr's Retribution (Su)} At 19th level, if the paladin dies while fighting evil or protecting her allies, she can choose to make the explosion of positive energy from her glory of the martyr ability painful to her foes. If she does, her body is almost completely consumed by holy power, preventing her from being raised with \spell{raise dead} and similar effects that require a body. This has two effects. First, a \spell{sunburst} spell immediately takes effect over the area where the paladin fell. Second, a \spell{storm of vengeance} spell begins to take effect, centered on the same area. The spell lasts for 10 rounds, and the lightning strikes focus on the paladin's enemies. Both of these effects harm only the paladin's foes, and do no damage to her allies. However, her allies' vision is still impeded by the \spell{storm of vengeance}.

\cf{Pal}{Greater Smite (Su)} At 20th level, the paladin can apply two improved smiting abilities to every smite attack she makes.

\subsection{Ranger}

\begin{dtable}
\lcaption{The Ranger}
\begin{tabularx}{\columnwidth}{>{\ccol}p{\levelcol} >{\ccol}p{\babcolgood} *{3}{>{\ccol}p{\savecol}} >{\lcol}X}
\thead{Level} & \thead{Base Attack Bonus} & \thead{Fort Save} & \thead{Ref Save} & \thead{Will Save} & \thead{Special} \\
1st  & \plus1                        & \plus3  & \plus1  & \plus1 & Quarry \plus2, Track, wild speech \\
2nd  & \plus2                        & \plus4  & \plus2  & \plus2 & Danger sense, favored terrain \\
3rd  & \plus3                        & \plus5  & \plus3  & \plus3 & Ranger lore \\
4th  & \plus4                        & \plus6  & \plus4  & \plus4 & Low-light vision, tracking expert \\
5th  & \plus5                        & \plus7  & \plus4  & \plus4 & Free stride, tenacious hunter \\
6th  & \plus6/\plus1                 & \plus8  & \plus5  & \plus5 & Favored terrain, ranger lore \\
7th  & \plus7/\plus2                 & \plus9  & \plus6  & \plus6 & Guide \\
8th  & \plus8/\plus3                 & \plus10 & \plus7  & \plus7 & Darkvision, quarry \plus3  \\
9th  & \plus9/\plus4                 & \plus11 & \plus7  & \plus7 & Ranger lore \\
10th & \plus10/\plus5                & \plus12 & \plus8  & \plus8 & Favored terrain (planar) \\
11th & \plus11/\plus6/\plus1         & \plus13 & \plus9  & \plus9 & Hidden hunter\\
12th & \plus12/\plus7/\plus2         & \plus14 & \plus10 & \plus10& Blindsense, advanced lore  \\
13th & \plus13/\plus8/\plus3         & \plus15 & \plus10 & \plus10& Terrain mastery  \\
14th & \plus14/\plus9/\plus4         & \plus16 & \plus11 & \plus11& Favored terrain (planar), quarry \plus4 \\
15th & \plus15/\plus10/\plus5        & \plus17 & \plus12 & \plus12& Advanced ranger lore \\
16th & \plus16/\plus11/\plus6/\plus1 & \plus18 & \plus13 & \plus13& Blindsight  \\
17th & \plus17/\plus12/\plus7/\plus2 & \plus19 & \plus13 & \plus14& Terrain mastery, unerring hunter \\
18th & \plus18/\plus13/\plus8/\plus3 & \plus20 & \plus14 & \plus14& Advanced ranger lore, favored terrain (planar)  \\
19th & \plus19/\plus14/\plus9/\plus4 & \plus21 & \plus15 & \plus15& Perfect stride  \\
20th & \plus20/\plus15/\plus10/\plus5& \plus22 & \plus16 & \plus16& Quarry \plus5, truesight
\end{tabularx}
\end{dtable}

\cf{Rgr}{Favored Terrain (Ex)} At 6th level, and every four levels thereafter, the ranger gains an additional favored terrain.

\cf{Rgr}{Ranger Lore (Ex)} At 6th level, and every 3 levels thereafter, the ranger gains an additional ranger lore ability.

\cf{Rgr}{Guide (Ex)} At 7th level, whenever the ranger is are in his favored terrain, all allies that can see and hear the ranger gain his favored terrain bonuses in that terrain as well.

\cf{Rgr}{Darkvision (Ex)} At 8th level, a ranger's sight improves again, and he gains the ability to see even when there is no light at all. He gains darkvision out to 60 feet, as the dwarf ability. If he already has darkvision, he increases its range by 60 feet.

\cf{Rgr}{Greater Combat Style (Ex)} At 10th level, a ranger's aptitude in combat improves again. He is treated as having the Ambidexterity feat (\pref{Ambidexterity [Combat]}) and the Manyshot feat (\pref{Manyshot [Combat]}), even if he does not have the normal prerequisites for those feats.
\par As before, the benefits of the ranger's chosen style apply only when he wears light or no armor. He loses all benefits of his combat style when wearing medium or heavy armor.

\cf{Rgr}{Favored Terrain (Planar) (Ex)} After 10th level, a ranger may choose any plane as a favored terrain in addition to his normal options whenever he gains a new favored terrain. He is immune to any hostile planar effects from any plane he has chosen as favored terrain. In addition, he gains a \plus2 competence bonus to Knowledge checks relating to the plane and is always treated as trained in Knowledge (planar) for the purpose of such checks.

\cf{Rgr}{Hidden Hunter (Su)} At 11th level, the ranger becomes even more difficult for his quarry to detect. He adds his quarry bonus to his Stealth checks against his quarry. In addition, he continuously benefits from the effect of the \spell{nondectection} spell against all attempts that his quarry makes to detect him magically. The effect uses a caster level equal to his ranger level.

\cf{Rgr}{Advanced Ranger Lore} After 12th level, the ranger can choose an advanced ranger lore in place of a regular ranger lore. All advanced lore abilities are extraordinary abilities unless otherwise indicated. His options for advanced ranger lores are listed below.

\subcf{Camouflage} The ranger can use the Hide skill in any of his favored terrains, even if the terrain does not grant cover or concealment.

\subcf{Combat Style Mastery} The ranger's abilities with traditional ranger combat styles reach their peak. When using a ranged weapon, he can take a move action to study the weak points of a foe within one range increment. If he does, the next attack he makes against that foe, if it is made in the same turn, is made as a ranged touch attack. When wielding two weapons at once, he gains the pounce ability, allowing him to take a full attack action at the end of a charge.

The ranger must have the greater combat style lore to choose this lore. The benefits apply only if the ranger is wearing light or no armor.

\subcf{Greater Combat Style} The ranger's abilities with traditional ranger combat styles improves again. He gains the Two-Weapon Rend and Manyshot feats if he meets the prerequisites. He must have the improved combat style lore to choose this lore. The benefits of this lore only apply if the ranger is wearing light or no armor.

\subcf{Hail of Arrows} A number of times per day equal to 1 \add half the ranger's Constitution, he may take a full-round action to fire a single arrow at every enemy within a \areamed radius. All enemies must be within one range increment of the ranger. This lore can be used with any ranged weapon that the ranger is capable of making a full attack with.

\subcf{Improved Evasion} The ranger's ability to avoid damage improves. He still takes no damage on a successful Reflex saving throw against attacks, but henceforth he takes only half damage on a failed save. A helpless ranger does not gain the benefit of improved evasion.

\subcf{Storm of Blades} A number of times per day equal to 1 \add half the ranger's Constitution, he may take a standard action to make a single melee attack against every enemy he threatens.

\cf{Rgr}{Blindsense (Ex)} At 12th level, a ranger's perceptions are so finely honed that he can sense his enemies without seeing them. He gains the blindsense ability out to 60 feet. This ability allows him to sense the presence and location of objects and foes within 60 feet without seeing them. If he already has the blindsense ability, he increases its range by 60 feet.

\cf{Rgr}{Terrain Mastery (Ex)} At 13th level, a ranger gains a greater degree of mastery over some of his favored terrains. He chooses a single kind of terrain that he has already chosen as a favored terrain. At 17th level, he chooses an additional kind of terrain to master.
\par While in that terrain, his competence bonuses on Perception, Stealth, and Survival checks increase to \plus4. In addition, he gains another ability based on that terrain that is constantly active, whether or not he is currently in the terrain. The options for terrain masteries are given below.
\subcf{Aquatic} The ranger does not suffer penalties for fighting and moving underwater.
\subcf{Cold} The ranger gains cold damage reduction 30.
\subcf{Desert} The ranger becomes immune to fatigue.
\subcf{Forest} The ranger may use his wild speech ability to communicate with plants, as the druid ability.
\subcf{Mountains} The ranger is always treated as if he had a running start when making Jump checks. In addition, he takes only half damage from falling damage.
\subcf{Plains} The ranger gains a \plus10 competence bonus to his land speed.
\subcf{Swamp} The ranger becomes immune to nausea.
\subcf{Underground} The ranger increases the range of his darkvision and blindsense by 60 feet.
\subcf{Urban} The ranger can treat cities as being natural terrain for the purpose of his camouflage and hide in plain sight abilities.

\cf{Rgr}{Blindsight (Ex)} At 16th level, a ranger gains the ability to ``see'' perfectly without his eyes in a 60 foot radius around him. With this ability, he can fight just as well with his eyes closed as with them open. If he already has the blindsight ability, he increases its range by 60 feet.

\cf{Rgr}{Unerring Hunter (Su)} At 17th level, a ranger's ability to hunt down his quarry improves to supernatural levels. Once per day, the ranger may concentrate for a full round to duplicate the effects of the \spell{discern location} spell targeted at his quarry.

\cf{Rgr}{Hide in Plain Sight (Ex)} While in any sort of natural terrain, a ranger of 17th level or higher can use the Hide skill even while being observed. He still needs cover or concealment to hide.

\cf{Rgr}{Perfect Stride (Su)} At 19th level, a ranger's ability to surpass obstacles becomes unparalleled. He constantly acts as if he were under the effect of a \spell{freedom} spell, except that it does not allow him to act normally underwater.

\cf{Rgr}{Truesight (Su)} At 20th level, a ranger's perceptions are accurate enough to defeat even powerful magic. He gains the ability to see all things as they actually are, as the \spell{true seeing} spell, out to a range of 60 feet.

\subsection{Rogue}
\begin{dtable*}
\lcaption{The Rogue}
\begin{tabularx}{\textwidth}{>{\ccol}p{\levelcol} >{\ccol}p{\babcolgood} *{3}{>{\ccol}p{\babcolgood}} X}
\thead{Level} & \thead{Base Attack Bonus} & \thead{Fort Save} & \thead{Ref Save} & \thead{Will Save} & \thead{Special} \\
1st  & \plus0                & \plus0 & \plus3  & \plus0 & Sneak attack \plus1d6 \\
2nd  & \plus1                & \plus1 & \plus4  & \plus1 & Uncanny dodge, skill trick, danger sense \\
3rd  & \plus2                & \plus1 & \plus5  & \plus1 & Ambush attack \plus1d6 \\
4th  & \plus3                & \plus2 & \plus6  & \plus2 & Evasion, combat trick \\
5th  & \plus3                & \plus2 & \plus7  & \plus2 & Skill trick, sneak attack \plus2d6 \\
6th  & \plus4                & \plus3 & \plus8  & \plus3 & Improved uncanny dodge \\
7th  & \plus5                & \plus3 & \plus9  & \plus3 & Ambush attack \plus2d6, combat trick\\
8th  & \plus6/\plus1         & \plus4 & \plus10 & \plus4 & Skill trick \\
9th  & \plus6/\plus1         & \plus4 & \plus11 & \plus4 & Sneak attack \plus3d6 \\
10th & \plus7/\plus2         & \plus5 & \plus12 & \plus5 & Greater uncanny dodge, combat trick \\
11th & \plus8/\plus3         & \plus5 & \plus13 & \plus5 & Advanced skill trick, ambush attack \plus3d6 \\
12th & \plus9/\plus4         & \plus6 & \plus14 & \plus6 & Jack of all trades \\
13th & \plus9/\plus4         & \plus6 & \plus15 & \plus6 & Sneak attack \plus4d6, advanced combat trick \\
14th & \plus10/\plus5        & \plus7 & \plus16 & \plus7 & Advanced skill trick \\
15th & \plus11/\plus6/\plus1 & \plus7 & \plus17 & \plus7 & Ambush attack \plus4d6 \\
16th & \plus12/\plus7/\plus2 & \plus8 & \plus18 & \plus8 & Advanced combat trick \\
17th & \plus12/\plus7/\plus2 & \plus8 & \plus19 & \plus8 & Advanced skill trick, sneak attack \plus5d6 \\
18th & \plus13/\plus8/\plus3 & \plus9 & \plus20 & \plus9 & Master of all trades \\
19th & \plus14/\plus9/\plus4 & \plus9 & \plus21 & \plus9 & Advanced combat trick, ambush attack \plus5d6 \\
20th & \plus15/\plus10/\plus5& \plus10& \plus22 & \plus10& Ambush master, advanced skill trick
\end{tabularx}
\end{dtable*}

\cf{Rog}{Improved Uncanny Dodge (Ex)} At 6th level and higher, a rogue can no longer be overwhelmed as easily; she can react to multiple opponents as easily as she can react to a single attacker. The rogue is always treated as being threatened by two fewer creatures than she actually is for the purpose of determining overwhelm penalties. 

\par If a character already has uncanny dodge (see above) from a second class and gains improved uncanny dodge, the character stacks those levels to determine if she should gain greater uncanny dodge, and to determine the minimum level a rogue must be to sneak attack the character.

\cf{Rog}{Combat Tricks} The rogue gains an additional combat trick at 7th level and every three levels thereafter.

\cf{Rog}{Skill Tricks} The rogue gains an additional skill trick at 8th level and every three levels thereafter.

\cf{Rog}{Greater Uncanny Dodge (Ex)} At 10th level and higher, a rogue can no longer be overwhelmed, regardless of the number of foes surrounding her.

\cf{Rog}{Advanced Skill Trick} At 11th level, and every three levels thereafter, the rogue may choose any of the following advanced skill tricks in addition to her other options for skill tricks. All advanced skill tricks are (Ex) abilities unless otherwise noted.

\subcf{Disable Spell (Su)} The rogue can use Devices to dispel any currently active spell as if it were a magical trap. Doing so requires a full-round action that provokes attacks of opportunity. The DC to disable the spell is equal to 10 \add the caster level of the spell \add the level of the spell. If the spell is not subject to \spell{dispel magic}, it cannot be dispelled using this ability.

\subcf{Exemplar} The rogue must choose one skill that she already has Skill Focus in. The skill must be on the rogue class skill list. She gains a \plus6 competence bonus with the skill.

\subcf{Hide in Plain Sight} The rogue can use the Hide even while being observed. She still needs cover or concealment to hide.

\subcf{Rogue's Luck} Three times per day, the rogue can reroll any skill check. A single roll can never be rerolled more than once.

\subcf{Skill Mastery} The rogue gains the Skill Mastery feat in a number of skills equal to 1 \add half her Intelligence (minimum 1). Each skill must be on the rogue class skill list.

\cf{Rog}{Advanced Combat Trick} At 12th level, and every three levels thereafter, the rogue may choose any of the following advanced combat tricks in addition to her other options for combat tricks. All advanced combat tricks are (Ex) abilities unless otherwise noted.

\subcf{Assassination} To use this ability, the rogue must spend a full round studying a foe has not noticed her and who is not in combat. If she make a melee ambush attack on her next turn against that target, her attack deals maximum damage. If the target becomes aware of her presence before she attacks, this ability has no benefit.

\subcf{Crippling Ambush} A rogue with this trick can ambush attack opponents with such precision that her blows weaken and hamper them. An opponent damaged by one of her ambush attacks takes 2 points of Strength damage. Ability points lost to damage return on their own at the rate of 1 point per day for each damaged ability. If the rogue has no ambush attacks, she may use crippling strike as an ambush attack.

\subcf{Defensive Roll}  The rogue can roll with a potentially lethal blow to take less damage from it than she otherwise would.  A number of times per day equal to half the rogue's Wisdom, when she would be reduced to 0 or fewer hit points by damage in combat (from a weapon or other blow, not a spell or special ability), the rogue can attempt to roll with the damage. To use this ability, the rogue must attempt a Reflex saving throw (DC = damage dealt). If the save succeeds, she takes only half damage from the blow, and the damage is nonlethal; if it fails, she takes full damage. She must be aware of the attack and able to react to it in order to execute her defensive roll -- if she is flat-footed, she can't use this ability. Since this effect would not normally allow a character to make a Reflex save for half damage, the rogue's evasion ability does not apply to the defensive roll.

\subcf{Distant Precision} The rogue has no range limit on her sneak attacks. A rogue must have selected the extended precision combat trick before choosing this trick.

\subcf{Improved Evasion} This talent works like evasion, except that the rogue also takes only half damage on a failed save. A helpless rogue does not gain the benefit of improved evasion.

\subcf{Opportunist}  Once per round, the rogue can make an attack of opportunity against an opponent who has just been struck for damage in melee by another character. This attack counts as one of the rogue's attacks of opportunity for that round.

\subcf{Slippery Mind}  This ability represents the rogue's ability to wriggle free from magical effects that would otherwise control or compel her. If a rogue with slippery mind is affected by a mind-affecting spell or effect and fails her saving throw, she can attempt it again 1 round later at the same DC. She gets only this one extra chance to succeed on her saving throw.

\cf{Rog}{Jack of All Trades (Ex)} At 16th level, a rogue treats all skills as class skills.

\cf{Rog}{Master of All Trades (Ex)} At 18th level, a rogue is treated as having as having at least one skill point in all skills, except for trained-only skills. These ``phantom'' skill points give her ranks in the skills normally, but do not otherwise count as skill points.

\cf{Rog}{Ambush Master (Ex)} A 20th level rogue has achieved such a mastery of tricky combat that she can combine the effects of two different ambush tricks into a single ambush attack.

\subsection{Sorcerer}
\begin{dtable*}
\lcaption{The Sorcerer}
\begin{tabularx}{\textwidth}{>{\ccol}p{\levelcol} >{\ccol}p{7em} *{3}{>{\ccol}p{\savecol}} >{\lcol}X *{9}{>{\ccol}p{\spellcol}}}
& & & & & & \multicolumn{9}{c}{\thead{---{}---{}---{}---{}---{}---{}---Spells per Day---{}---{}---{}---{}---{}---}} \\
\thead{Level} & \thead{Base Attack Bonus} & \thead{Fort Save} & \thead{Ref Save} & \thead{Will Save} & \thead{Special} & \thead{1st} & \thead{2nd} & \thead{3rd} & \thead{4th} & \thead{5th} & \thead{6th} & \thead{7th} & \thead{8th} & \thead{9th} \\
1st & \plus0 & \plus0 & \plus0 & \plus3 & Arcane invocation, Rapid Metamagic, versatile spellcaster
& 3 & \x & \x & \x & \x & \x & \x & \x & \x \\
2nd & \plus1 & \plus1 & \plus1 & \plus4     & Arcane invocation
& 4 & \x & \x & \x & \x & \x & \x & \x & \x \\
3rd & \plus1 & \plus1 & \plus1 & \plus5     & Expanded spell knowledge
& 5 & \x & \x & \x & \x & \x & \x & \x & \x \\
4th & \plus2 & \plus2 & \plus2 & \plus6     & Spellblend
& 6 & 3 & \x & \x & \x & \x & \x & \x & \x \\
5th & \plus2 & \plus2 & \plus2 & \plus7     & Expanded spell knowledge
& 6 & 4 & \x & \x & \x & \x & \x & \x & \x \\
6th & \plus3 & \plus3 & \plus3 & \plus8     & \x
& 6 & 5 & 3 & \x & \x & \x & \x & \x & \x \\
7th & \plus3 & \plus3 & \plus3 & \plus9     & Expanded spell knowledge
& 6 & 6 & 4 & \x & \x & \x & \x & \x & \x \\
8th & \plus4 & \plus4 & \plus4 & \plus10    & Improved spellblend
& 6 & 6 & 5 & 3 & \x & \x & \x & \x & \x \\
9th & \plus4 & \plus4 & \plus4 & \plus11    & Expanded spell knowledge
& 6 & 6 & 6 & 4 & \x & \x & \x & \x & \x \\
10th & \plus5 & \plus5 & \plus5 & \plus12    & \x 
& 6 & 6 & 6 & 5 & 3 & \x & \x & \x & \x \\
11th & \plus5 & \plus5 & \plus5 & \plus13    & Expanded spell knowledge
& 6 & 6 & 6 & 6 & 4 & \x & \x & \x & \x \\
12th & \plus6/\plus1 & \plus6 & \plus6 & \plus14& Versatile spellblend
& 6 & 6 & 6 & 6 & 5 & 3 & \x & \x & \x \\
13th & \plus6/\plus1 & \plus6 & \plus6 & \plus15& Expanded spell knowledge
& 6 & 6 & 6 & 6 & 6 & 4 & \x & \x & \x \\
14th & \plus7/\plus2 & \plus7 & \plus7 & \plus16& \x
& 6 & 6 & 6 & 6 & 6 & 5 & 3 & \x & \x \\
15th & \plus7/\plus2 & \plus7 & \plus7 & \plus17& Expanded spell knowledge
& 6 & 6 & 6 & 6 & 6 & 6 & 4 & \x & \x \\
16th & \plus8/\plus3 & \plus8 & \plus8 & \plus18 & Spellsurge
& 6 & 6 & 6 & 6 & 6 & 6 & 5 & 3 & \x \\
17th & \plus8/\plus3 & \plus8 & \plus8 & \plus19 & Expanded spell knowledge
& 6 & 6 & 6 & 6 & 6 & 6 & 6 & 4 & \x \\
18th & \plus9/\plus4 & \plus9 & \plus9 & \plus20 & \x
& 6 & 6 & 6 & 6 & 6 & 6 & 6 & 5 & 3 \\
19th & \plus9/\plus4 & \plus9 & \plus9 & \plus21 & Expanded spell knowledge
& 6 & 6 & 6 & 6 & 6 & 6 & 6 & 6 & 4 \\
20th & \plus10/\plus5 & \plus10& \plus10& \plus22 & Improved spellsurge
& 6 & 6 & 6 & 6 & 6 & 6 & 6 & 6 & 6 \\
\end{tabularx}
\end{dtable*}

\cf{Sor}{Spells} A sorcerer gains additional spells as he increases in level.

\begin{dtable}
\lcaption{Sorcerer Spells Known}
\begin{tabularx}{\columnwidth}{>{\ccol}X *{10}{>{\ccol}p{\spellcol}}}
& \multicolumn{10}{c}{\thead{---{}---{}---{}---{}---{}---{}---{}---Spells Known---{}---{}---{}---{}---{}---{}---{}---}} \\
\thead{Level} & \thead{1st} & \thead{2nd} & \thead{3rd} & \thead{4th} & \thead{5th} & \thead{6th} & \thead{7th} & \thead{8th} & \thead{9th} \\
1st  & 1 & \x & \x & \x & \x & \x & \x & \x & \x \\
2nd  & 2 & \x & \x & \x & \x & \x & \x & \x & \x \\
3rd  & 3 & \x & \x & \x & \x & \x & \x & \x & \x \\
4th  & 3 & 1 & \x & \x & \x & \x & \x & \x & \x \\
5th  & 4 & 2 & \x & \x & \x & \x & \x & \x & \x \\
6th  & 4 & 2 & 1 & \x & \x & \x & \x & \x & \x \\
7th  & 4 & 3 & 2 & \x & \x & \x & \x & \x & \x \\
8th  & 4 & 3 & 2 & 1 & \x & \x & \x & \x & \x \\
9th  & 4 & 3 & 3 & 2 & \x & \x & \x & \x & \x \\
10th & 4 & 3 & 3 & 2 & 1 & \x & \x & \x & \x \\
11th & 4 & 3 & 3 & 3 & 2 & \x & \x & \x & \x \\
12th & 4 & 3 & 3 & 3 & 2 & 1 & \x & \x & \x \\
13th & 4 & 3 & 3 & 3 & 3 & 2 & \x & \x & \x \\
14th & 4 & 3 & 3 & 3 & 3 & 2 & 1 & \x & \x \\
15th & 4 & 3 & 3 & 3 & 3 & 3 & 2 & \x & \x \\
16th & 4 & 3 & 3 & 3 & 3 & 3 & 2 & 1 & \x \\
17th & 4 & 3 & 3 & 3 & 3 & 3 & 2 & 2 & \x \\
18th & 4 & 3 & 3 & 3 & 3 & 3 & 2 & 2 & 1 \\
19th & 4 & 3 & 3 & 3 & 3 & 3 & 2 & 2 & 2 \\
20th & 4 & 3 & 3 & 3 & 3 & 3 & 2 & 2 & 2
\end{tabularx}
\end{dtable}

\cf{Sor}{Improved Spellblend (Ex)} At 8th level, a sorcerer may combine two spells together. As a full-round action, the sorcerer may cast two spells at once, resolving each spell's effects separately. The spells cast in this way must be at least three spell levels apart, such as a 1st-level spell and a 4th-level spell. In addition, one of the two spells must affect only the sorcerer. Using improved spellblend costs a spell slot of one level higher than the highest level spell being cast.

\cf{Sor}{Versatile Spellblend (Ex)} At 12th level, a sorcerer may combine any two spells together. When using spellblend or improved spellblend, the sorcerer may cast any spells, regardless of whether they affect only the sorcerer. However, using versatile spellblend costs a spell slot of two levels higher than the highest level spell being cast.

\cf{Sor}{Spellsurge (Ex)} At 16th level, a sorcerer may enter a trance-like state once per day in which he can surpass his normal limits. A spellsurge trance lasts for one minute. During the trance, the sorcerer may use his versatile spellblend ability by expending a spell slot of one level lower than the highest level spell being cast. However, the sorcerer is forced to expel the arcane energy welling up inside him, and is forced to use his versatile spellblend ability with all of his actions. The sorcerer can suppress this effect for a round with a DC 30 Will save, but he takes nonlethal damage equal to his caster level if he does so. %what happens if the sorcerer can't act? Does he take damage?

\par At 19th level, a sorcerer can enter a spellsurge trance an additional time per day.

\cf{Sor}{Improved Spellsurge (Ex)} At 20th level, a sorcerer in a spellsurge trance can use his spellblend ability as a standard action instead of as a full-round action.

\subsection{Spellwarped}
\begin{dtable}
    \lcaption{The Spellwarped}
    \begin{tabularx}{\columnwidth}{>{\ccol}p{\levelcol} >{\ccol}p{\babcolavg} *{2}{>{\ccol}p{\savecolpoof}} >{\lcol}X}
        \thead{Level} & \thead{Base Attack Bonus} & \thead{Good Save}\fn{1} & \thead{Normal Saves}\fn{1} & \thead{Special} \\
        1st & \plus0                    & \plus2  & \plus1  & Innate magic, invoke power, spellwarp pool\\
        2nd & \plus1                    & \plus3  & \plus2  & Spellwarped body, surge of power \\
        3rd & \plus2                    & \plus4  & \plus3  & Attuned senses, spellwarped aspect \\
        4th & \plus3                    & \plus5  & \plus4  & Invoke power, resist magic \\
        5th & \plus3                    & \plus6  & \plus4  & Manipulate magic \\
        6th & \plus4                    & \plus7  & \plus5  & Invoke power \\
        7th & \plus5                    & \plus8  & \plus6  & Spellwarped aspect \\
        8th & \plus6/\plus1             & \plus9  & \plus7  & Invoke power \\
        9th & \plus6/\plus1             & \plus10 & \plus7  & Spell resistance\\
        10th & \plus7/\plus2            & \plus11 & \plus8  & Invoke power \\
        11th & \plus8/\plus3            & \plus12 & \plus9  & Spellwarped aspect \\
        12th & \plus9/\plus4            & \plus13 & \plus10 & Invoke power \\
        13th & \plus9/\plus4            & \plus14 & \plus10 & Improved manipulate magic \\
        14th & \plus10/\plus5           & \plus15 & \plus11 & Invoke power \\
        15th & \plus11/\plus6/\plus1    & \plus16 & \plus12 & Spellwarped aspect \\
        16th & \plus12/\plus7/\plus2    & \plus17 & \plus13 & Invoke power \\
        17th & \plus12/\plus7/\plus2    & \plus19 & \plus13 & Mass surge of power \\
        18th & \plus13/\plus8/\plus3    & \plus20 & \plus14 & Invoke power \\
        19th & \plus14/\plus9/\plus4    & \plus21 & \plus15 & Permanent surge of power, spellwarped aspect \\
        20th & \plus15/\plus10/\plus5   & \plus22 & \plus16 & Invoke power \\
    \end{tabularx}
    1 Each spellwarped has a good save determined by his choice of innate magic.
\end{dtable}

\cf{Spl}{Spellwarped Aspect (Su)} At 7th level, and every four levels thereafter, the spellwarped gains an additional spellwarped aspect. He may choose more advanced spellwarped aspects that have level requirements, as indicated below.
\parhead{General}
\subcf{7th -- Accelerated Recovery} The spellwarped regains spellwarp points once per 10 minutes, rather than once per hour.
\subcf{11th -- Improved Attune Senses} The spellwarped can constantly gain the benefit of his Attuned Senses ability. He can toggle his enhanced senses on or off as a swift action.

\parhead{Alteration}
\subcf{7th -- Alter Movement} The spellwarped gains his choice of the Legendary Balance, Legendary Climber, Legendary Leaper, or Legendary Swimmer feats, even if he does not meet the prerequisites. He may select this aspect multiple times, choosing a different bonus feat each time.
\subcf{7th -- Alter Size} When the spellwarped uses his surge of power, he can increase or decrease by a size category, as he chooses. This is a size-affecting effect, and does not stack with other size-affecting effects.

\parhead{Pyromancy}
\subcf{11th -- Improved Flameheart} The spellwarped's flameheart power lasts for 5 rounds.

\parhead{Telekinesis}
\subcf{7th -- Shieldbearer} The spellwarped may wield shields, except tower shields, telekinetically. The shield floats in his square, granting him its AC bonus. He does not need a free hand to wield the shield and suffers no armor check penalty or arcane spell failure from it. The shield follows him as he moves. If it is forcibly removed from his square, he loses control over it and it falls to the ground.
\subcf{11th -- Mind Armory} The spellwarped may control a number of weapons equal to half his Intelligence with his mind blade ability. This does not allow him to make additional attacks per round, but he may attack interchangeably with any weapon he controls. Each weapon threatens an area and contributes to overwhelm penalties, just as with his normal mind blade ability.

\parhead{Temporal}
\subcf{7th -- Accelerate Attack} While his surge of power is active, the spellwarped can make an additional attack at his full attack bonus when making a full attack.

\cf{Spl}{Spell Resistance (Ex)} At 9th level, the magic within the spellwarped allows him to completely ignore other magic, granting him spell resistance. A creature with spell resistance may always make a saving throw when a spell is cast on it. The saving throw type is indicated by the spell. If it succeeds, the spell has no effect on it.

\cf{Spl}{Improved Manipulate Magic (Su)} At 13th level, the spellwarped can use his manipulate magic ability to affect any ally within \rngmed range of him.

\cf{Spl}{Mass Surge of Power (Su)} At 17th level, the spellwarped can share the benefits of his surge of power with his allies. When he uses his surge of power, he can also affect up to five additional creatures within \rngmed range of him.

\cf{Spl}{Permanent Surge of Power (Su)} At 19th level, the spellwarped can maintain the full power of his innate magic without limit. He can gain the effects of his surge of power indefinitely. He may toggle the ability on or off as a swift action at will, without expending spellwarp points. This does not allow him to activate his mass surge of power ability at will, and his allies only gain the benefits for 5 rounds.

\ssecfake{Spellwarped Powers}
As the spellwarped increases in level, he may choose more advanced spellwarped powers, listed below.

\subsubsection{Alteration Powers}
\parhead{8th -- Body Bludgeon} The spellwarped elongates and distorts a part of his body and strikes a foe with it. The foe must be within his reach, as if he were wielding a reach weapon. He must make an attack roll against the foe's AC. If he hits, he deals 1d6 bludgeoning damage per spellwarped level \add his Strength. In addition, whether he hits or misses, he may make a bull rush attack on the creature that does not provoke attacks of opportunity. He need not move with the creature to push it back.
\parhead{8th -- Enlargement} This power functions like the \spell{enlarge person} spell, except that it can affect creatures of any type.
\parhead{10th -- Heal Wounds} As a standard action, the spellwarped can spend two spellwarp points to remove his own injuries by transforming himself into a healthier version of his body. He heals 1d6 points of damage per spellwarped level. This also removes any of the following conditions: blinded, diseased, exhausted, fatigued, nauseated, sickened, and poisoned.
\parhead{12th -- Baleful Polymorph} This attack functions like the \spell{baleful polymorph} spell.
\parhead{14th -- Flight} As a swift action, the spellwarped can grow wings to fly for 5 rounds. His fly speed is equal to his base land speed, and his maneuverability is good. At the end of the duration, the wings are subsumed back into his body. This ability is draining to use, and the spellwarped must wait for five minutes after using it before he can use it again.
\parhead{16th -- Bludgeon the Horde} This attack functions like the body bludgeon attack, except that he may attack all foes within his reach, as if he were wielding a reach weapon. He deals 1d6 bludgeoning damage per two spellwarped levels \add his Strength to each foe.
\parhead{18th -- }
\parhead{20th -- }

\subsubsection{Pyromancy Powers}
\parhead{8th -- Conflagration} The spellwarped release a powerful explosion of flame. All creatures within a \areamed radius spread of him take 1d8 fire damage per two spellwarped levels. A successful Reflex save halves the damage.
\parhead{10th -- Fire Shield} As a standard action, the spellwarped can wreath himself in flame for 5 rounds. Any creature that hits him with its body or a melee weapon takes 1d6 fire damage per two spellwarped levels. Each individual creature can take this damage only once per round.
\parhead{12th -- } 
\parhead{14th -- Flameheart} As a standard action, the spellwarped can become a being of pure fire for 1 round. In this form, he is immune to physical damage and can pass through openings as small as one inch at no movement penalty. However, he cannot attack normally or use any of his items, as they meld into his body. He may invoke any of his spellwarped powers normally. In this form, he can make a touch attack as a standard action to deal 1d6 points of fire damage per spellwarped level.
\parhead{14th -- Firestride} As a move action, the spellwarped can may teleport to any active flame of at least Tiny size within \rngmed range. When he does so, he immolates himself and disappears into a pile of ash before stepping out from the flame unharmed. An ordinary torch is sufficient flame to teleport to, but not a candle.
\parhead{16th -- } 
\parhead{18th -- Phoenix Revival} When the spellwarped takes critical damage, he may spend five spellwarp points as an immediate action, even if the critical damage would be sufficient to kill him. If he does, he ignores the critical damage he just took and dissolves into a pile of ash for 5 rounds. During this time, he can take no actions. If the pile of ash remains intact after 5 rounds, the spellwarped is restored to his normal body, with zero hit points but with all critical damage healed. However, if the pile of ash is dispersed, the spellwarped dies. The ash cannot be harmed by fire damage, and if the pile of ash would take at least 10 points of fire damage during a round, the spellwarped returns one round sooner. The spellwarped may take his normal actions immediately after being restored.
\parhead{20th -- Immolate} The spellwarped consumes the body of a foe within \rngclose range in flames from the inside out. It takes 1d6 points of fire damage per spellwarped level, and if it is bloodied after it takes this damage, it immediately dies. A successful Fortitude save halves the damage and leaves a bloodied creature with 0 hit points.

\subsubsection{Telekinesis Powers}
\parhead{8th -- }
\parhead{10th -- Telekinetic Force} This power functions like the \spell{telekinetic force} spell, using his Intelligence as his casting attribute.
\parhead{12th -- Strangle} The spellwarped crushes the windpipe of a foe within \rngclose range using his mind, dealing 1d6 damage per spellwarped level. A creature bloodied after the damage is dealt is nauseated and immobilized for 1 round. A successful Fortitude save halves the damage and negates the nausea. The spellwarped can maintain concentration on the ability as a standard action to deal additional damage and extend the duration of the nausea by 1 round, spending one spellwarp point per round.
\parhead{14th -- }
\parhead{16th -- } 
\parhead{18th -- }
\parhead{20th -- Mass Strangle} This attack functions like the strangle attack, except that the spellwarped can affect any creatures within a \areasmall radius.

\subsubsection{Temporal Powers}
\parhead{8th -- Pause Time} The spellwarped completely stops time for a single creature within \rngclose range for 5 rounds. A successful Will save negates the effect. The affected creature can take no actions and cannot be moved, damaged, or even affected in any way until the effect ends.
\parhead{10th -- }
\parhead{12th -- Swift Acceleration} This power functions like the brief acceleration power, except that it can be used as a swift action.
\parhead{12th -- Timestream} The spellwarped manipulates time in a \arealarge line that extends out from him for 5 rounds. All creatures and objects that pass through the line are slowed for 1 round, with no saving throw allowed. The spellwarped can exclude his allies from the effect. The timestream is virtually invisible, requiring a DC 30 Perception check to notice in a clear environment, though objects passing through the effect can make it obvious.
\parhead{14th -- }
\parhead{16th -- Flash Step} As a swift action, the spellwarped can accelerate himself to immense speed, allowing him to take a run action. He does not provoke attacks of opportunity for the movement, and is not flat-footed after running. After using this ability, the spellwarped must wait 5 rounds before he can use it again.
\parhead{16th -- Time Reversal} As a swift action, the spellwarped can spend a spellwarp point to create a ``time lock.'' The time lock persists for one round. As a standard action, he can bring a creature backwards through time to the point at which the time lock was created. A Will save negates this effect. An affected creature is perfectly restored to the point immediately after the time lock was created. The effects of any actions that the creature took in the intervening time are undone, any damage it dealt or took is removed, it is restored to its original location, and the creature is restored in all other ways, just as if the intervening time had never occured. The spellwarped cannot reverse time for himself in this way.
\parhead{18th -- Time Stop} As a standard action, the spellwarped can spend three spellwarp points to step into an alternate timestream, causing him to speed up so greatly that all other creatures seem frozen. He can take a single round of actions in this alternate timestream. All creatures he attacks are treated as helpless, but he cannot perform a coup de grace or similar ability; such an act requires more care and precision than is possible with such immense speed. After using this ability, he must wait 5 rounds before he can use it again.
\parhead{20th -- Sever Time} The spellwarped completely stops time for a single creature for 5 rounds. This attack functions like the pause time power, except that no saving throw is allowed.

\subsection{Wizard}
\begin{dtable*}
\lcaption{The Wizard}
\begin{tabularx}{\textwidth}{>{\ccol}p{\levelcol} >{\ccol}p{7em} *{3}{>{\ccol}p{\savecol}} >{\lcol}X *{9}{>{\ccol}p{\spellcol}}}
& & & & & & \multicolumn{9}{c}{\thead{---{}---{}---{}---{}---{}---{}---Spells per Day---{}---{}---{}---{}---{}---}} \\
\thead{Level} & \thead{Base Attack Bonus} & \thead{Fort Save} & \thead{Ref Save} & \thead{Will Save} & \thead{Special} & \thead{1st} & \thead{2nd} & \thead{3rd} & \thead{4th} & \thead{5th} & \thead{6th} & \thead{7th} & \thead{8th} & \thead{9th} \\
1st & \plus0 & \plus0 & \plus0 & \plus3 & Arcane invocation, ritual master
& 3 & \x & \x & \x & \x & \x & \x & \x & \x \\
2nd & \plus1 & \plus1 & \plus1 & \plus4     & Arcane invocation, Scribe Scroll
& 4 & \x & \x & \x & \x & \x & \x & \x & \x \\
3rd & \plus1 & \plus1 & \plus1 & \plus5     & Arcane insight
& 5 & \x & \x & \x & \x & \x & \x & \x & \x \\
4th & \plus2 & \plus2 & \plus2 & \plus6     & Spell sequencer
& 6 & 3 & \x & \x & \x & \x & \x & \x & \x \\
5th & \plus2 & \plus2 & \plus2 & \plus7     & Arcane attunement (1 item), arcane insight
& 6 & 4 & \x & \x & \x & \x & \x & \x & \x \\
6th & \plus3 & \plus3 & \plus3 & \plus8     & \x
& 6 & 5 & 3 & \x & \x & \x & \x & \x & \x \\
7th & \plus3 & \plus3 & \plus3 & \plus9     & Arcane insight
& 6 & 6 & 4 & \x & \x & \x & \x & \x & \x \\
8th & \plus4 & \plus4 & \plus4 & \plus10    & Improved spell sequencer
& 6 & 6 & 5 & 3 & \x & \x & \x & \x & \x \\
9th & \plus4 & \plus4 & \plus4 & \plus11    & Arcane insight
& 6 & 6 & 6 & 4 & \x & \x & \x & \x & \x \\
10th & \plus5 & \plus5 & \plus5 & \plus12    & Arcane attunement (2 items)
& 6 & 6 & 6 & 5 & 3 & \x & \x & \x & \x \\
11th & \plus5 & \plus5 & \plus5 & \plus13    & Arcane insight
& 6 & 6 & 6 & 6 & 4 & \x & \x & \x & \x \\
12th & \plus6/\plus1 & \plus6 & \plus6 & \plus14& Contingency
& 6 & 6 & 6 & 6 & 5 & 3 & \x & \x & \x \\
13th & \plus6/\plus1 & \plus6 & \plus6 & \plus15& Arcane insight
& 6 & 6 & 6 & 6 & 6 & 4 & \x & \x & \x \\
14th & \plus7/\plus2 & \plus7 & \plus7 & \plus16& \x
& 6 & 6 & 6 & 6 & 6 & 5 & 3 & \x & \x \\
15th & \plus7/\plus2 & \plus7 & \plus7 & \plus17& Arcane attunement (3 items), arcane insight
& 6 & 6 & 6 & 6 & 6 & 6 & 4 & \x & \x \\
16th & \plus8/\plus3 & \plus8 & \plus8 & \plus18 & Versatile spell sequencer
& 6 & 6 & 6 & 6 & 6 & 6 & 5 & 3 & \x \\
17th & \plus8/\plus3 & \plus8 & \plus8 & \plus19 & Arcane insight
& 6 & 6 & 6 & 6 & 6 & 6 & 6 & 4 & \x \\
18th & \plus9/\plus4 & \plus9 & \plus9 & \plus20& \x
& 6 & 6 & 6 & 6 & 6 & 6 & 6 & 5 & 3 \\
19th & \plus9/\plus4 & \plus9 & \plus9 & \plus21 & Arcane insight
& 6 & 6 & 6 & 6 & 6 & 6 & 6 & 6 & 4 \\
20th & \plus10/\plus5 & \plus10& \plus10& \plus22 & Arcane attunement (4 items), chain contingency
& 6 & 6 & 6 & 6 & 6 & 6 & 6 & 6 & 6 \\
\end{tabularx}
\end{dtable*}
\cd{Alignment} Any.

\cf{Wiz}{Spells} A wizard gains additional spells per day and spells known as she increases in level.

\begin{dtable}
\lcaption{Wizard Spells Known}
\begin{tabularx}{\columnwidth}{>{\ccol}X *{10}{>{\ccol}p{\spellcol}}}
& \multicolumn{10}{c}{\thead{---{}---{}---{}---{}---{}---{}---{}---Spells Known---{}---{}---{}---{}---{}---{}---{}---}} \\
\thead{Level} & \thead{1st} & \thead{2nd} & \thead{3rd} & \thead{4th} & \thead{5th} & \thead{6th} & \thead{7th} & \thead{8th} & \thead{9th} \\
1st  & 1 & \x & \x & \x & \x & \x & \x & \x & \x \\
2nd  & 2 & \x & \x & \x & \x & \x & \x & \x & \x \\
3rd  & 3 & \x & \x & \x & \x & \x & \x & \x & \x \\
4th  & 3 & 1 & \x & \x & \x & \x & \x & \x & \x \\
5th  & 4 & 2 & \x & \x & \x & \x & \x & \x & \x \\
6th  & 4 & 2 & 1 & \x & \x & \x & \x & \x & \x \\
7th  & 4 & 3 & 2 & \x & \x & \x & \x & \x & \x \\
8th  & 4 & 3 & 2 & 1 & \x & \x & \x & \x & \x \\
9th  & 4 & 3 & 3 & 2 & \x & \x & \x & \x & \x \\
10th & 4 & 3 & 3 & 2 & 1 & \x & \x & \x & \x \\
11th & 4 & 3 & 3 & 3 & 2 & \x & \x & \x & \x \\
12th & 4 & 3 & 3 & 3 & 2 & 1 & \x & \x & \x \\
13th & 4 & 3 & 3 & 3 & 3 & 2 & \x & \x & \x \\
14th & 4 & 3 & 3 & 3 & 3 & 2 & 1 & \x & \x \\
15th & 4 & 3 & 3 & 3 & 3 & 3 & 2 & \x & \x \\
16th & 4 & 3 & 3 & 3 & 3 & 3 & 2 & 1 & \x \\
17th & 4 & 3 & 3 & 3 & 3 & 3 & 2 & 2 & \x \\
18th & 4 & 3 & 3 & 3 & 3 & 3 & 2 & 2 & 1 \\
19th & 4 & 3 & 3 & 3 & 3 & 3 & 2 & 2 & 2 \\
20th & 4 & 3 & 3 & 3 & 3 & 3 & 2 & 2 & 2
\end{tabularx}
\end{dtable}

\cf{Wiz}{Improved Spell Sequencer (Ex)} At 8th level, a wizard gains the ability to create a sequence of two spells which she can cast rapidly later. To create an improved spell sequencer, the wizard must cast two spells, one of which affects only herself. The spells must be at least three levels apart. Neither has any effect immediately. the wizard may later use a full-round action to cast both spells at once.
\par The wizard may have only one spell sequencer or improved spell sequencer active at any time.

\cf{Wiz}{Arcane Attunement (Su)} At 10th level, and every five levels thereafter, the wizard gains the ability to attune herself to an additional item.

\section{Character Advancement}

\begin{dtable}
\lcaption{Character Advancement and Level-Dependent Bonuses}
\begin{tabularx}{\columnwidth}{*{5}{>{\ccol}X}}
  \thead{Character level} & \thead{XP} & \thead{Feats} & \thead{Attribute Increases\fn{1}} \\
1st & 0 & 1st & \x & \x \\
2nd & 2,000 & \x & 1st \\
3rd & 5,000 & 2nd & \x \\
4th & 9,000 & \x & 2nd \\
5th & 15,000 & 3rd & \x \\
6th & 23,000 & \x & 3rd \\
7th & 35,000 & 4th & \x \\
8th & 51,000 & \x & 4th \\
9th & 75,000 & 5th & \x \\
10th & 105,000 & \x & 5th \\
11th & 155,000 & 6th & \x \\
12th & 220,000 & \x & 6th \\
13th & 315,000 & 7th & \x \\
14th & 445,000 & \x & 7th \\
15th & 635,000 & 8th & \x \\
16th & 890,000 & \x & 8th \\
17th & 1,300,000 & 9th & \x \\
18th & 1,800,000 & \x & 9th \\
19th & 2,550,000 & 10th & \x \\
20th & 3,600,000 & \x & 10th
\end{tabularx}
1. The same attribute cannot be increased twice in a row.
\end{dtable}
