\chapter{Equipment}

You begin with 100 gold pieces, and you decide how to spend them.
Assume a character owns at least one outfit of normal clothes. Pick any one of the following clothing outfits: artisan's outfit, entertainer's outfit, explorer's outfit, monk's outfit, peasant's outfit, scholar's outfit, or traveler's outfit.
\section{Wealth And Money}

\subsection{Coins}
The most common coin is the gold piece (gp). A gold piece is worth 10 silver pieces. Each silver piece is worth 10 copper pieces (cp). In addition to copper, silver, and gold coins, there are also platinum pieces (pp), which are each worth 10 gp.

The standard coin weighs about a third of an ounce (fifty to the pound).

\begin{dtable}
\lcaption{Coin Exchange Values}
\begin{tabularx}{\columnwidth}{l c *{4}{>{\ccol}X}}
& & \thead{CP} & \thead{SP} & \thead{GP} & \thead{PP} \\
Copper piece (cp) & = & 1 & 1/10 & 1/100 & 1/1,000 \\
Silver piece (sp) & = & 10 & 1 & 1/10 & 1/100 \\
Gold piece (gp) & = & 100 & 10 & 1 & 1/10 \\
Platinum piece (pp) & = & 1,000 & 100 & 10 & 1
\end{tabularx}
\end{dtable}

\subsection{Wealth Other Than Coins}
Merchants commonly exchange trade goods without using currency. As a means of comparison, some trade goods are detailed below.

\begin{dtable}
\lcaption{Trade Goods}
\begin{tabularx}{\columnwidth}{l >{\lcol}X}
\thead{Cost} & \thead{Item} \\
1 cp & One pound of wheat \\
2 cp & One pound of flour \\
1 sp & One pound of iron, or one chicken \\
5 sp & One pound of tobacco or copper \\
1 gp & One pound of cinnamon, or one goat \\
2 gp & One pound of ginger or pepper, or one sheep \\
3 gp & One pig \\
4 gp & One square yard of linen \\
5 gp & One pound of salt or silver \\
10 gp & One square yard of silk, or one cow \\
15 gp & One pound of saffron or cloves, or one ox \\
50 gp & One pound of gold \\
500 gp & One pound of platinum
\end{tabularx}
\end{dtable}

\subsection{Selling Loot}
In general, a character can sell something for half its listed price.

Trade goods are the exception to the half-price rule. A trade good, in this sense, is a valuable good that can be easily exchanged almost as if it were cash itself.

\section{Weapons}

Weapons are grouped into several interlocking sets of categories. These categories pertain to what weapon group the weapon belongs to (axes, bows, and so on), the weapon's usefulness either in close combat (melee) or at a distance (ranged, which includes both thrown and projectile weapons), its relative encumbrance (light, medium, or heavy), and its size (Small, Medium, or Large).

\subsection{Weapon Groups}

Weapons are organized into thematically related categories called weapon groups. They are described in \trefnp{Weapon Groups}. For example, all axes belong to the ``axes'' weapon group. Some weapons can be found in multiple weapon groups. For example, a dagger is a simple weapon, a light blade, and a thrown weapon.

\begin{dtable!*}
\lcaption{Weapon Groups}
\begin{tabularx}{\textwidth}{l >{\lcol}X >{\lcol}X}
\thead{Group} & \thead{Weapons} & \thead{Exotic Weapons} \\
Armor weapons & Heavy shield (and spiked), light shield (and spiked), spiked armor & \\
Axes & Battleaxe, greataxe, handaxe, throwing axe & Double axe, dwarven urgrosh,  dwarven waraxe \\
Blunt weapons & Club, greatclub, heavy mace, light mace, quarterstaff, sap & \\
Blades, heavy & Falchion, greatsword, longsword, scimitar & Bastard sword, two-bladed sword \\
Blades, light & Dagger, punching dagger, rapier, short sword & Kukri \\
Bows & Longbow, shortbow (and composite) & \\
Crossbows & Heavy crossbow, light crossbow & Hand crossbow, repeating crossbows \\
Flexible Weapons & Flail, heavy flail, nunchaku & Spiked chain, whip \\
Headed weapons & Heavy pick, light hammer, light pick, scythe, sickle, warhammer & Gnome hooked hammer, maul \\
Monk weapons & Kama, nunchaku, quarterstaff, sai, shuriken, siangham & \\
Polearms & Glaive, guisarme, halberd, ranseur & \\
Simple weapons & Club, dagger, light crossbow, quarterstaff, unarmed strike & \\
Spears & Javelin, lance, shortspear, spear, trident & \\
Thrown weapons & Dagger, dart, handaxe, javelin, light hammer, shuriken, sling & Bolas, net \\
\end{tabularx}
\end{dtable!*}

\subsection{Melee and Ranged Weapons}
Melee weapons are used for
making melee attacks, though some of them can be thrown as well.
Ranged weapons are thrown weapons or projectile weapons that are
not effective in melee.

\subsubsection{Weapon Encumbrance:} This is a measure of how much effort it takes to wield a
weapon in combat. A weapon's encumbrance indicates whether a melee weapon, when
wielded by a character of the weapon's size category, is considered a
light weapon, a medium weapon, or a heavy weapon.

\parhead{Light} A light weapon can be used with more finesse than a medium weapon. The wielder can add either his Dexterity or his Strength to attack rolls with light weapons, whichever he prefers. In addition, light weapons are easier to use in the off-hand or while grappling.

An unarmed strike is considered a light weapon unless the character has the Improved Unarmed Strike feat, in which case it can be considered either a light or a medium weapon.

\parhead{Medium} A medium weapon can be used in either the primary hand or the off hand.

A character must have a Strength score of at least 3 to use a medium weapon in one hand. Otherwise, the character must use it in two hands.

\parhead{Heavy} Two hands are required to use a heavy weapon effectively.

A character must have a Strength score of at least 3 to use a heavy weapon effectively. Otherwise, the character is treated as being nonproficient with the weapon.

\subparhead{Double} Double weapons are treated as heavy weapons when only one end at a time is used. When both sides are used at once, one side is treated as being a medium weapon, and the other side is treated as being a light weapon.

\subsubsection{Reach Weapons}
Glaives, guisarmes, lances, longspears, ranseurs, and whips are reach weapons. A reach weapon is a melee weapon that allows its wielder to strike at targets that aren't adjacent to him or her. Most reach double the wielder's natural reach, meaning that a typical Small or Medium wielder of such a weapon can attack a creature 10 feet away, but not a creature in an adjacent square. A typical Large character wielding a reach weapon of the appropriate size can attack a creature 15 or 20 feet away, but not adjacent creatures or creatures up to 10 feet away.

\par Most reach weapons may be ``short hafted'', allowing its wielder to strike adjacent foes but denying them the ability to strike foes farther away. Switching grips like this takes a move action. While short hafting a reach weapon, you take a \minus4 penalty to attack rolls, since the weapon was not designed to be used in that way. A reach weapon can be short hafted unless stated otherwise in its description.

\subsubsection{Double Weapons}
Dire flails, dwarven urgroshes, gnome hooked hammers, orc double axes, quarterstaffs, and two-bladed swords are double weapons. A character can fight with both ends of a double weapon as if fighting with two weapons, just as though the character were wielding a medium weapon and a light weapon.

\par The character can also choose to use a double weapon two handed, attacking with only one end of it. A creature wielding a double weapon in one hand can't use it as a double weapon -- only one end of the weapon can be used in any given round.

\subsubsection{Thrown Weapons}
Daggers, darts, javelins, throwing axes, light hammers, shuriken, and nets are thrown weapons. The wielder applies his or her Strength to damage dealt by thrown weapons (except for splash weapons).

\par A character can also use a thrown weapon in melee, but if the weapon is not designed for melee (such as a javelin or shuriken), the wielder provokes an attack of opportunity when he misses, as if using a weapon with which he is not proficient.

It is possible to throw a weapon that isn't designed to be thrown (that is, a melee weapon that doesn't have a numeric entry in the Range Increment column on \trefnp{Weapons}), but a character who does so takes a \minus4 penalty on the attack roll. Throwing a light or medium weapon is a standard action, while throwing a heavy weapon is a full-round action. Regardless of the type of weapon, such an attack scores a threat only on a natural roll of 20 and deals double damage on a critical hit. Such a weapon has a range increment of 10 feet.

\subsubsection{Projectile Weapons} Light crossbows, slings, heavy crossbows, shortbows, composite shortbows, longbows, composite longbows, hand crossbows, and repeating crossbows are projectile weapons. Most projectile weapons require two hands to use (see specific weapon descriptions). A character gets no Strength bonus on damage rolls with a projectile weapon unless it's a specially built composite shortbow, specially built composite longbow, or sling. If the character has a penalty for low Strength, apply it to damage rolls when he or she uses a bow or a sling.

\subsubsection{Ammunition} Projectile weapons use ammunition: arrows (for bows), bolts (for crossbows), or sling bullets (for slings). When using a bow, a character can draw ammunition as a free action; crossbows and slings require an action for reloading. Generally speaking, ammunition that hits its target is destroyed or rendered useless, while normal ammunition that misses has a 50\% chance of being destroyed or lost.

Although they are thrown weapons, shuriken and darts are treated as
ammunition for the purposes of drawing them, crafting special versions of them, and what happens to them after they are thrown.

\subsubsection{Weapon Size} Every weapon has a size category. This designation indicates the size of the creature for which the weapon was designed.

A weapon's size category isn't the same as its size as an object. Instead, a weapon's size category is keyed to the size of the intended wielder. In general, a light weapon is an object two size categories smaller than the wielder, a medium weapon is an object one size category smaller than the wielder, and a heavy weapon is an object of the same size category as the wielder.

\parhead{Inappropriately Sized Weapons} The measure of how much effort it takes to use a weapon (whether the weapon is designated as a light, medium, or heavy weapon for a particular wielder) is altered by one step for each size category of difference between the wielder's size and the size of the creature for which the weapon was designed. If a weapon's designation would be changed to something other than light, medium, or heavy by this alteration, the creature can't wield the weapon at all.

\subsubsection{Improvised Weapons} Sometimes objects not crafted to be weapons nonetheless see use in combat. Because such objects are not designed for this use, any creature that uses one in combat is considered to be nonproficient with it. If it is used as a melee weapon, the wielder provokes an attack of opportunity when he misses with it. If it is used as a ranged weapon, the wielder takes a \minus4 penalty to attack rolls with it. To determine the size category and appropriate damage for an improvised weapon, compare its relative size and damage potential to the weapon list to find a reasonable match. An improvised weapon scores a threat on a natural roll of 20 and deals double damage on a critical hit. An improvised thrown weapon has a range increment of 10 feet.

\subsection{Weapon Qualities}
Here is the format for weapon entries (given as column headings on Table: Weapons, below).

\parhead{Encumbrance} Describes whether the weapon is a light, medium, heavy, double, or ranged weapon.

\par This cost is the same for a Small or Medium version of the weapon. A Large version costs twice the listed price.
\parhead{Damage} The Damage columns give the damage dealt by the weapon on a successful hit. If two damage ranges are given then the weapon is a double weapon. Use the second damage figure given for the double weapon's extra attack. The damage given is for weapons used by Medium creatures. Weapons used by Small creatures have a damage die that is one size smaller. \trefnp{Weapon Damage and Size} gives weapon damage values for weapons of many sizes.

\begin{dtable}
\lcaption{Weapon Damage and Size}
\begin{tabularx}{\columnwidth}{*{6}{l} >{\lcol}X}
\thead{Medium} & \thead{Tiny} & \thead{Small} & \thead{Large} & \thead{Huge} & \thead{Gargantuan} & \thead{Colossal} \\
1d2  & \x  & 1   & 1d3  & 1d4  & 1d8  & 2d6  \\
1d3  & 1   & 1d2 & 1d4  & 1d6  & 1d10 & 2d8  \\
1d4  & 1d2 & 1d3 & 1d6  & 1d8  & 2d6  & 2d10 \\
1d6  & 1d3 & 1d4 & 1d8  & 1d10 & 2d8  & 4d6  \\
1d8  & 1d4 & 1d6 & 1d10 & 2d6  & 2d10 & 4d8  \\
1d10 & 1d6 & 1d8 & 2d6  & 2d8  & 4d6  & 4d10 \\
\end{tabularx}
\end{dtable}

%''standard'' minimum damage for 
\parhead{Minimum Damage} Some medium and heavy weapons are particularly likely to deal significant damage. If you roll the weapon's minimum damage or lower, reroll the damage. You never reroll the same roll more than once. Each roll made for a critical hit is considered separately.

The minimum damage given is for weapons used by Medium creatures. Weapons used by Small creatures have a minimum damage that is one lower. In general, when a weapon increases in size category, its minimum damage increases by an amount equal to the number of dice rolled for the weapon's new damage. For example, greatsword increases its minimum damage by 2 when it grows from Medium to Large, since a Large greatsword deals 2d6 damage. Similarly, a Small greatsword has a minimum damage 1 lower than a Medium greatsword, since a Small greatsword deals 1d8 damage. Weapons that do not have a minimum damage rating gain a minimum damage rating when increasing in size, provided that the weapon deals at least 1d8 damage at its new size.

\begin{comment}
For example, a Medium greatsword increases its minimum damage by 1 when going increasing to Large, since a Medium greatsword deals \textbf{1}d10 damage. Similarly, a Large greatsword increases its minimum damage by 2 when increasing to Huge, since a Large greatsword deals \textbf{2}d10 damage. Weapons that do not have a minimum damage rating gain a minimum damage rating when increasing in size, provided that the weapon deals at least 1d6 damage at its new size.
\end{comment}

\parhead{Critical} The entry in this column notes how the weapon is used with the rules for critical hits. When your character scores a critical hit, roll the damage two, three, or four times, as indicated by its critical multiplier (using all applicable modifiers on each roll), and add all the results together.

\subparhead{Exception} Bonus damage dice over and above a weapon's normal damage dice are not multiplied when you score a critical hit.
\subparhead{\mult2} The weapon deals double damage on a critical hit.
\subparhead{\mult3} The weapon deals triple damage on a critical hit.
\subparhead{\mult3/\mult4} One head of this double weapon deals triple damage on a critical hit. The other head deals quadruple damage on a critical hit.
\subparhead{\mult4} The weapon deals quadruple damage on a critical hit.
\subparhead{19-20/\mult2} The weapon scores a threat on a natural roll of 19 or 20 (instead of just 20) and deals double damage on a critical hit. (The weapon has a threat range of 19-20.)
\subparhead{18-20/\mult2} The weapon scores a threat on a natural roll of 18, 19, or 20 (instead of just 20) and deals double damage on a critical hit. (The weapon has a threat range of 18-20.)

\parhead{Range Increment} Any attack at less than this distance is not penalized for range. However, each full range increment imposes a cumulative \minus2 penalty on the attack roll. A thrown weapon has a maximum range of five range increments. A projectile weapon can shoot out to ten range increments.
\parhead{Type} Weapons are classified according to the type of damage they deal: bludgeoning, piercing, or slashing. Some monsters may be resistant or immune to attacks from certain types of weapons.

Some weapons deal damage of multiple types. If a weapon is of two types, the damage it deals is not half one type and half another; all of it is both types. Therefore, a creature would have to be immune to both types of damage to ignore any of the damage from such a weapon.

In other cases, a weapon can deal either of two types of damage. In a situation when the damage type is significant, the wielder can choose which type of damage to deal with such a weapon.
\parhead{Cost} This value is the weapon's cost in gold pieces (gp) or silver pieces (sp). The cost includes miscellaneous gear that goes with the weapon.
\parhead{Weight} This column gives the weight of a Medium version of
the weapon. Halve this number for Small weapons, and double it for
Large weapons.

\parhead{Special} Some weapons have special features. See the weapon
descriptions for details.

\subsubsection{Weapon Descriptions}
\begin{dtable!*}
\lcaption{Weapons}
\begin{tabularx}{\textwidth}{l c c c c c >{\lcol}X c c}
\thead{Weapons} & \thead{Encumbrance} & \thead{Dmg} & \thead{Min} & \thead{Critical} & \thead{Range Increment} & \thead{Type\footnotetemp{2}} & \thead{Cost} & \thead{Weight\footnotetemp{1}} \\

Armor weapons &&&&&&&& \\
\tind Shield, heavy & Medium & 1d4 & \x & \mult2 & \x & Bludgeoning & special & special \\
\tind Shield, light & Light & 1d3 & \x & \mult2 & \x & Bludgeoning & special & special \\
\tind Spiked armor & Light & 1d6 & \x & \mult2 & \x & Piercing & special & special \\
\tind Spiked shield, heavy & Medium & \x & 1 & \mult2 & \x & Piercing & special & special \\
\tind Spiked shield, light & Light & \x &  & \mult2 & \x & Piercing & special & special \\

Axes &&&&&&&& \\
\tind Axe, throwing & Light & 1d6 & \x & \mult2 & 10 ft. & Slashing & 8 gp & 2 lb. \\
\tind Battleaxe & Medium & 1d8 & 1 & \mult3 & \x & Slashing & 10 gp & 6 lb. \\
\tind Greataxe & Heavy & 1d10 & 2 & \mult3 & \x & Slashing & 20 gp & 12 lb. \\
\tind Handaxe & Light & 1d6 & \x & \mult3 & \x & Slashing & 6 gp & 3 lb. \\

Blunt weapons &&&&&&& \\
\tind Club & Medium & 1d6 & \x & \mult2 & \x & Bludgeoning & \x & 3 lb. \\
\tind Greatclub & Heavy & 1d10 & 1 & \mult2 & \x & Bludgeoning & 5 gp & 8 lb. \\
\tind Mace, heavy & Medium & 1d8 & 1 & \mult2 & \x & Bludgeoning & 12 gp & 8 lb. \\
\tind Mace, light & Light & 1d6 & \x & \mult2 & \x & Bludgeoning & 5 gp & 4 lb. \\
\tind Morningstar & Medium & 1d8 & 1 & \mult2 & \x & Bludgeoning and piercing & 8 gp & 6 lb. \\
\tind Quarterstaff\fn{5} & Double & 1d6/1d6 &  & \mult2 & \x & Bludgeoning & \x & 4 lb. \\
\tind Sap & Light & 1d6\fn{3} & \x & \mult2 & \x & Bludgeoning & 1 gp & 2 lb. \\

Blades, heavy &&&&&&&& \\
\tind Falchion & Heavy & 1d10 & 2 & \mult3 & \x & Slashing & 75 gp & 8 lb. \\
\tind Greatsword & Heavy & 1d10 & 2 & 19-20/\mult2 & \x & Slashing & 50 gp & 8 lb. \\
\tind Longsword & Medium & 1d8 & 1 & 19-20/\mult2 & \x & Slashing & 15 gp & 4 lb. \\
\tind Scimitar & Medium & 1d8 & 1 & \mult3 & \x & Slashing & 15 gp & 4 lb. \\

Blades, light &&&&&&&& \\
\tind Dagger & Light & 1d4 & \x & 19-20/\mult2 & 10 ft. & Piercing or slashing & 2 gp & 1 lb. \\
\tind Dagger, punching & Light & 1d4 & \x & \mult3 & \x & Piercing & 2 gp & 1 lb. \\
\tind Rapier & Medium & 1d6 & \x & \mult3 & \x & Piercing & 20 gp & 2 lb. \\
\tind Sword, short & Light & 1d6 & \x & 19-20/\mult2 & \x & Piercing or slashing & 10 gp & 2 lb. \\

Bows &&&&&&&& \\
\tind Longbow & Ranged & 1d8 & 1 & \mult3 & 100 ft. & Piercing & 75 gp & 3 lb. \\
\tind Longbow, composite & Ranged & 1d8 & 1 & \mult3 & 100 ft. & Piercing & 100 gp & 3 lb. \\
\tind Shortbow & Ranged & 1d6 & \x & \mult3 & 50 ft. & Piercing & 30 gp & 2 lb. \\
\tind Shortbow, composite & Ranged & 1d6 & \x & \mult3 & 50 ft. & Piercing & 75 gp & 2 lb. \\
\tind Arrows (20) & \x & \x & \x & \x & \x & \x & 1 gp & 3 lb. \\

Crossbows &&&&&&&& \\
\tind Crossbow, heavy & Ranged & 1d10 & 2 & 19-20/\mult2 & 100 ft. & Piercing & 50 gp & 8 lb. \\
\tind Crossbow, light & Ranged & 1d8 & 1 & 19-20/\mult2 & 50 ft. & Piercing & 35 gp & 4 lb. \\
\tind Bolts, crossbow (10) & \x & \x & \x & \x & \x & \x & 1 gp & 1 lb. \\

Flexible weapons &&&&&&&& \\
\tind Flail  & Medium & 1d8 & \x & \mult2 & \x & Bludgeoning & 8 gp & 5 lb. \\
\tind Flail, heavy & Heavy & 1d10 & 1 & \mult2 & \x & Bludgeoning & 15 gp & 10 lb. \\

\end{tabularx}
\end{dtable!*}

\begin{dtable!*}
\begin{tabularx}{\textwidth}{l c c c c c >{\ccol}p{4em} c >{\lcol}X}
\thead{Weapons} & \thead{Encumbrance} & \thead{Cost} & \thead{Dmg (S)} & \thead{Dmg (M)} & \thead{Critical} & \thead{Range Increment} & \thead{Weight\footnotetemp{1}} & \thead{Type\footnotetemp{2}} \\
Headed weapons &&&&&&&& \\
\tind Hammer, light & Light & 1d4 & \x & \mult2 & 20 ft. & Bludgeoning & 1 gp & 2 lb. \\
\tind Pick, heavy & Medium & 1d6 & \x & \mult4 & \x & Piercing & 8 gp & 6 lb. \\
\tind Pick, light & Light & 1d4 & \x & \mult4 & \x & Piercing & 4 gp & 3 lb. \\
\tind Scythe & Heavy & 1d8 & 1 & 19-20/\mult3 & \x & Piercing or slashing & 18 gp & 10 lb. \\
\tind Sickle & Light & 1d6 & \x & \mult3 & \x & Slashing & 6 gp & 2 lb. \\
\tind Warhammer & Medium & 1d8 & 1 & \mult3 & \x & Bludgeoning & 12 gp & 5 lb. \\

Monk weapons &&&&&&&& \\
\tind Kama & Light & 1d6 & \x & \mult2 & \x & Slashing & 2 gp & 2 lb. \\
\tind Nunchaku & Light & 1d6 & \x & \mult2 & \x & Bludgeoning & 2 gp & 2 lb. \\
\tind Quarterstaff\fn{5} & Double & 1d6/1d6 & \x & \mult2 & \x & Bludgeoning & \x & 4 lb. \\
\tind Sai & Light & 1d4 & \x & \mult2 & \x & Piercing & 1 gp & 1 lb. \\
\tind Shuriken (5) & Ranged & 1d2 & \x & \mult2 & 10 ft. & Piercing & 1 gp & 1/2 lb. \\
\tind Siangham & Light & 1d6 & \x & \mult2 & \x & Piercing & 3 gp & 1 lb. \\

Polearms &&&&&&&& \\
\tind Glaive\fn{4} & Heavy & 1d10 & 2 & \mult3 & \x & Slashing & 8 gp & 10 lb. \\
\tind Guisarme\fn{4} & Heavy & 1d10 & 1 & \mult3 & \x & Slashing & 9 gp & 12 lb. \\
\tind Halberd & Heavy & 1d10 & 2 & \mult3 & \x & Piercing or slashing & 10 gp & 12 lb. \\
\tind Ranseur\fn{4} & Heavy & 1d10 & 1 & \mult3 & \x & Piercing & 10 gp & 12 lb. \\

Simple weapons &&&&&&&& \\
\tind Club & Medium & 1d6 & \x & \mult2 & 10 ft. & Bludgeoning & \x & 3 lb. \\
\tind Crossbow, light & Ranged & 1d8 & 1 & 19-20/\mult2 & 80 ft. & Piercing & 35 gp & 4 lb. \\
\tind Dagger & Light & 1d4 & \x & 19-20/\mult2 & 10 ft. & Piercing or slashing & 2 gp & 1 lb. \\
\tind Quarterstaff\fn{5} & Double & 1d6/1d6 & \x & \mult2 & \x & Bludgeoning & \x & 4 lb. \\
\tind Unarmed strike & Light & 1d3\fn{3} & \x & \mult2 & \x & Bludgeoning & \x & \x \\

Spears &&&&&&&& \\
\tind Javelin & Ranged & 1d6 & \x & \mult2 & 30 ft. & Piercing & 1 gp & 2 lb. \\
\tind Lance\fn{4} & Heavy & 1d8 & 2 & \mult3 & \x & Piercing & 10 gp & 10 lb. \\
\tind Longspear\fn{4} & Heavy & 1d8 & 2 & \mult3 & \x & Piercing & 5 gp & 9 lb. \\
\tind Shortspear & Medium & 1d6 & 1 & \mult2 & 20 ft. & Piercing & 1 gp & 3 lb. \\
\tind Spear & Heavy & 1d8 & 2 & \mult3 & 20 ft. & Piercing & 2 gp & 6 lb. \\
\tind Trident & Medium & 1d8 & 2 & \mult2 & 10 ft. & Piercing & 15 gp & 4 lb. \\

Thrown weapons &&&&&&&& \\
\tind Axe, throwing & Light & 1d6 & \x & \mult2 & 10 ft. & Slashing & 8 gp & 2 lb. \\
\tind Dagger & Light & 1d4 & \x & 19-20/\mult2 & 10 ft. & Piercing or slashing & 2 gp & 1 lb. \\
\tind Dart (5) & Ranged & 1d4 & \x & \mult2 & 20 ft. & Piercing & 1 gp & 1/2 lb. \\
\tind Hammer, light & Light & 1d4 & \x & \mult2 & 20 ft. & Bludgeoning & 1 gp & 2 lb. \\
\tind Javelin & Ranged & 1d6 & \x & \mult2 & 30 ft. & Piercing & 1 gp & 2 lb. \\
\tind Shuriken (5) & Ranged & 1d2 & \x & \mult2 & 10 ft. & Piercing & 1 gp & 1/2 lb. \\
\tind Sling & Ranged & 1d4 & \x & \mult2 & 50 ft. & Bludgeoning & 2 gp & 0 lb. \\
\tind Bullets, sling (20) & \x & \x & \x & \x & \x & \x & 1 gp & 5 lb. \\

Weaponlike Spells\fn{6} &&&&&&&&\\

Unarmed weapons &&&&&&&&\\
\tind Claw Sheath & \x & \x & \x & \x & \x & \x & 50 gp & 3 lb. \\
\tind Gauntlet & Light & 1d3 & \x & \mult2 & \x & Bludgeoning & 2 gp & 1 lb. \\
\tind Gauntlet, spiked & Light & 1d4 & \x & \mult2 & \x & Piercing & 5 gp & 1 lb. \\
\tind Unarmed strike & Light & 1d3\fn{3} & \x & \mult2 & \x & Bludgeoning & \x & \x \\
\end{tabularx}
\end{dtable!*}

\begin{dtable!*}
\begin{tabularx}{\textwidth}{l c c c c c >{\ccol}p{4em} c >{\lcol}X}
\thead{Exotic Weapons} & \thead{Encumbrance} & \thead{Cost} & \thead{Dmg (S)} & \thead{Dmg (M)} & \thead{Critical} & \thead{Range Increment} & \thead{Weight\footnotetemp{1}} & \thead{Type\footnotetemp{2}} \\
Armor &&&&&&&& \\
Axes &&&&&&&& \\
\tind Axe, orc double\fn{5} & Double & 1d8/1d8 & 1/1 & \mult3/\mult3 & \x & Slashing & 60 gp & 15 lb. \\
\tind Urgrosh, dwarven\fn{5} & Double & 1d8/1d6 & 1/1 & \mult3 & \x/\mult3 & Slashing or piercing & 50 gp & 12 lb. \\
\tind Waraxe, dwarven & Medium & 1d10 & 1 & \mult3 & \x & Slashing & 30 gp & 8 lb. \\
Blunt weapons &&&&&&&& \\
Blades, heavy &&&&&&&& \\
\tind Sword, bastard & Medium & 1d10 & 2 & 19-20/\mult2 & \x & Slashing & 35 gp & 6 lb. \\
\tind Sword, two-bladed\fn{5} & Double & 1d8/1d8 & 2/2 & 19-20/\mult2 & \x & Slashing & 100 gp & 10 lb. \\
Blades, light &&&&&&&& \\
\tind Kukri & Light & 1d4 &  & 18-20/\mult2 & \x & Slashing & 8 gp & 2 lb. \\
Bows &&&&&&&& \\
Crossbows &&&&&&&& \\
\tind Crossbow, hand & Ranged & 1d4 & \x & 19-20/\mult2 & 30 ft. & Piercing & 100 gp & 2 lb. \\
\tind Crossbow, repeating heavy & Ranged & 1d10 & 2 & 19-20/\mult2 & 120 ft. & Piercing & 400 gp & 12 lb. \\
\tind Crossbow, repeating light & Ranged & 1d8 & 1 & 19-20/\mult2 & 80 ft. & Piercing & 250 gp & 6 lb. \\
\tind Bolts, hand (10) & \x & \x & \x & \x & \x & \x & 1 gp & 1 lb. \\
\tind Bolts, repeating (5) & \x & \x & \x & \x & \x & \x & 1 gp & 1 lb. \\
Flexible weapons &&&&&&&& \\
\tind Chain, spiked\fn{4} & Heavy & 1d8 & 1 & \mult2 & \x & Piercing & 25 gp & 10 lb. \\
\tind Flail, dire\fn{5} & Double & 1d8/1d8 & 1/1 & \mult2/\mult2 & \x & Bludgeoning & 90 gp & 10 lb. \\
\tind Whip\fn{4} & Light & 1d3\fn{3} & \x & \mult2 & \x & Slashing & 1 gp & 2 lb. \\
Headed weapons &&&&&&&& \\
\tind Hammer, gnome hooked\fn{5} & Double & 1d8/1d6 & 1/1 & \mult2/\mult2 & \x & Bludgeoning and piercing & 20 gp & 6 lb. \\
Monk weapons &&&&&&&& \\
Polearms &&&&&&&& \\
Simple weapons &&&&&&&& \\
Spear &&&&&&&& \\
Thrown weapons &&&&&&&& \\
\tind Bolas & Ranged & 1d4\fn{3} & \x & \mult2 & 10 ft. & Bludgeoning & 5 gp & 2 lb. \\
\tind Net & Ranged & \x & \x & 10 ft. & 6 lb. &  & 20 gp & \x \\
Unarmed weapons &&&&&&&&\\
\end{tabularx}
1 Weight figures are for Medium weapons. A Small weapon weighs half as much, and a Large weapon weighs twice as much. \\
2 When two types are given, the weapon is both types if the entry specifies ``and," or either type (player's choice at time of attack) if the entry specifies ``or." \\
3 The weapon deals nonlethal damage rather than lethal damage. \\
4 Reach weapon. \\
5 Double weapon. \\
6 Not actually a weapon group, but can be treated as one for the purpose of feats and abilities (like Weapon Focus).
\end{dtable!*}

Weapons found on Table: Weapons that have special options for the wielder ("you") are described below. Splash weapons are described under Special Substances and Items.
\parhead{Arrows} An arrow used as a melee weapon is treated as a light improvised weapon (\minus4 penalty on attack rolls) and deals damage as a dagger of its size (critical multiplier x2). Arrows come in a leather quiver that holds 20 arrows. An arrow that hits its target is destroyed; one that misses has a 50\% chance of being destroyed or lost.
\parhead{Axe, Orc Double} An orc double axe is a double weapon. You can fight with it as if fighting with two weapons, but if you do, you incur all the normal attack penalties associated with fighting with two weapons, just as if you were using a medium weapon and a light weapon.
\par A creature wielding an orc double axe in one hand can't use it as a double weapon?only one end of the weapon can be used in any given round.
\parhead{Bolas} You can use this weapon to make a ranged trip attack against an opponent.
\parhead{Bolts} A crossbow bolt used as a melee weapon is treated as a light improvised weapon (\minus4 penalty on attack rolls) and deals damage as a dagger of its size (crit \mult2). Bolts come in a wooden case that holds 10 bolts (or 5, for a repeating crossbow). A bolt that hits its target is destroyed; one that misses has a 50\% chance of being destroyed or lost.
\parhead{Bullets, Sling} Bullets come in a leather pouch that holds 10 bullets. A bullet that hits its target is destroyed; one that misses has a 50\% chance of being destroyed or lost.
\parhead{Chain, Spiked} A spiked chain has reach, so you can strike opponents 10 feet away with it. In addition, unlike most other weapons with reach, it can be used against an adjacent foe. However, you do not threaten foes you are not adjacent to. In addition, you cannot attack foes who have cover from you unless they are adjacent to you.
\par You can make trip attacks with the chain. If you are tripped during your own trip attempt, you can drop the chain to avoid being tripped.
\par When using a spiked chain, you get a \plus2 bonus on opposed attack rolls made to disarm an opponent.
\par You cannot short haft a spiked chain.

\parhead{Claw Sheath} A claw sheath is not a weapon in itself, but a covering over a single claw that a creature can use to make claw attacks. Claw sheaths do not grant a creature without claws a claw attack. Normally, a sheath does not improve the claw attack, but magical claw sheaths can be made which grant bonuses to attacks made with the claw. It may be possible to find or craft unusual sheaths for natural weapons other than claws.

\parhead{Crossbow, Hand} You can draw a hand crossbow back by hand. Loading a hand crossbow is a move action that provokes attacks of opportunity.
\par You can shoot, but not load, a hand crossbow with one hand at no penalty. You can shoot a hand crossbow with each hand, but you take a penalty on attack rolls as if attacking with two light weapons.
\par You can fire a crossbow from a prone position without penalty.
\parhead{Crossbow, Heavy} You draw a heavy crossbow back by turning a small winch. Loading a heavy crossbow is a full-round action that provokes attacks of opportunity.
\par Normally, operating a heavy crossbow requires two hands. However, you can shoot, but not load, a heavy crossbow with one hand at a \minus4 penalty on attack rolls. You can shoot a heavy crossbow with each hand, but you take a penalty on attack rolls as if attacking with two medium weapons. This penalty is cumulative with the penalty for one-handed firing.
\par You can fire a crossbow from a prone position without penalty.
\parhead{Crossbow, Light} You draw a light crossbow back by pulling a lever. Loading a light crossbow is a move action that provokes attacks of opportunity.
\par Normally, operating a light crossbow requires two hands. However, you can shoot, but not load, a light crossbow with one hand at a \minus2 penalty on attack rolls. You can shoot a light crossbow with each hand, just as if wielding two light weapons at once. This penalty is cumulative with the penalty for one-handed firing.
\par You can fire a crossbow from a prone position without penalty.
\parhead{Crossbow, Repeating} The repeating crossbow (whether heavy or light) holds 5 crossbow bolts. As long as it holds bolts, you can reload it by pulling the reloading lever (a free action). Loading a new case of 5 bolts is a full-round action that provokes attacks of opportunity.
\par You can fire a repeating crossbow with one hand or fire a repeating crossbow in each hand in the same manner as you would a normal crossbow of the same size. However, you must fire the weapon with two hands in order to use the reloading lever, and you must use two hands to load a new case of bolts.
\par You can fire a crossbow from a prone position without penalty.
\parhead{Dagger} Daggers, including punching daggers, are unusually small, and are considered one size category smaller than normal for a light weapon (that is, three size categories smaller than the creature it is intended for). This typically grants a \plus2 circumstance bonus on Sleight of Hand checks to conceal the dagger. A dagger can be used as a light weapon by a creature one size smaller than the dagger is designed for.
\parhead{Dart} Although they are thrown weapons, darts are treated as ammunition for the purposes of drawing them, crafting special versions of them, and what happens to them after they are thrown.

\parhead{Flail, Dire} A dire flail is a double weapon. You can fight with it as if fighting with two weapons, but if you do, you incur all the normal attack penalties associated with fighting with two weapons, just as if you were using a medium weapon and a light weapon. A creature wielding a dire flail in one hand can't use it as a double weapon -- only one end of the weapon can be used in any given round.
\par When using a dire flail, you get a \plus2 bonus on opposed attack rolls made to disarm an enemy.
\par You can also use this weapon to make trip attacks.
\parhead{Flail or Heavy Flail} With a flail, you get a \plus2 bonus on opposed attack rolls made to disarm an enemy.
\par You can also use this weapon to make trip attacks.
\parhead{Gauntlet} This metal glove lets you deal lethal damage rather than nonlethal damage with unarmed strikes. A strike with a gauntlet is otherwise considered an unarmed attack. The cost and weight given are for a single gauntlet. Medium and heavy armors (except breastplate) come with gauntlets.
\parhead{Gauntlet, Spiked} Your opponent cannot use a disarm action to disarm you of spiked gauntlets. The cost and weight given are for a single gauntlet. An attack with a spiked gauntlet is considered an armed attack.
\parhead{Glaive} A glaive has reach. You can strike opponents 10 feet away with it, but you can't use it against an adjacent foe unless you short haft it.
\parhead{Guisarme} A guisarme has reach. You can strike opponents 10 feet away with it, but you can't use it against an adjacent foe unless you short haft it.
\par You can also use it to make trip attacks. If you are tripped during your own trip attempt, you can drop the guisarme to avoid being tripped.
\parhead{Halberd} If you use a ready action to set a halberd against a charge, you deal double damage on a successful hit against a charging character.
\par You can use a halberd to make trip attacks. If you are tripped during your own trip attempt, you can drop the halberd to avoid being tripped.
\parhead{Hammer, Gnome Hooked} A gnome hooked hammer is a double weapon. You can fight with it as if fighting with two weapons, but if you do, you incur all the normal attack penalties associated with fighting with two weapons, just as if you were using a medium weapon and a light weapon. The hammer's blunt head is a bludgeoning weapon that deals 1d6 points of damage (crit \mult3). Its hook is a piercing weapon that deals 1d4 points of damage (crit \mult4). You can use either head as the primary weapon. The other head is the offhand weapon. A creature wielding a gnome hooked hammer in one hand can't use it as a double weapon -- only one end of the weapon can be used in any given round.
\par You can use a gnome hooked hammer to make trip attacks. If you are tripped during your own trip attempt, you can drop the gnome hooked hammer to avoid being tripped.
\parhead{Javelin} You can throw a javelin using Strength to attack, treating it as a medium thrown weapon. Since it is not designed for melee, you are treated as nonproficient with it and take a \minus4 penalty on attack rolls if you use a javelin as a melee weapon.
\parhead{Kama} The kama is a special monk weapon. This designation gives a monk wielding a kama special options.
\par You can use a kama to make trip attacks. If you are tripped during your own trip attempt, you can drop the kama to avoid being tripped.
\parhead{Lance} A lance deals double damage when used from the back of a charging mount. It has reach, so you can strike opponents 10 feet away with it, but you can't use it against an adjacent foe unless you short haft it.
\par While mounted, you can wield a lance with one hand.
\parhead{Longbow} You need at least two hands to use a bow, regardless of its size. A longbow is too unwieldy to use while you are mounted. If you have a penalty for low Strength, apply it to damage rolls when you use a longbow. If you have a bonus for high Strength, you can apply it to damage rolls when you use a composite longbow (see below) but not a regular longbow.
\parhead{Longbow, Composite} You need at least two hands to use a bow, regardless of its size. You can use a composite longbow while mounted. All composite bows are made with a particular strength rating (that is, each requires a minimum Strength to use with proficiency). If your Strength bonus is less than the strength rating of the composite bow, you can't effectively use it, so you take a \minus2 penalty on attacks with it. The default composite longbow requires a Strength of \plus0 or higher to use with proficiency. A composite longbow can be made with a high strength rating to take advantage of an above-average Strength score; this feature allows you to add your Strength bonus to damage, up to the maximum bonus indicated for the bow. Each point of Strength bonus granted by the bow adds 100 gp to its cost.
\par For purposes of weapon proficiency and similar feats, a composite longbow is treated as if it were a longbow.
\parhead{Longspear} A longspear has reach. You can strike opponents 10 feet away with it, but you can't use it against an adjacent foe unless you short haft it. If you use a ready action to set a longspear against a charge, you deal double damage on a successful hit against a charging character.
\parhead{Net} A net is used to entangle enemies. When you throw a net, you make a ranged touch attack against your target. A net's maximum range is 10 feet. If you hit, the target is entangled. An entangled creature takes a \minus2 penalty on attack rolls and a \minus2 penalty on Dexterity, can move at only half speed, and cannot charge or run. If you control the trailing rope by succeeding on an opposed Strength check while holding it, the entangled creature can move only within the limits that the rope allows. If the entangled creature attempts to cast a spell, it must make a DC 15 add double spell level Concentration check or be unable to cast the spell.
\par An entangled creature can escape with a DC 20 Escape Artist check (a full-round action). The net has 5 hit points and can be burst with a DC 20 Strength check (also a full-round action).
\par A net is useful only against creatures within one size category of you.
\par A net must be folded to be thrown effectively. The first time you throw your net in a fight, you make a normal ranged touch attack roll. After the net is unfolded, you take a \minus4 penalty on attack rolls with it. It takes 2 rounds for a proficient user to fold a net and twice that long for a nonproficient one to do so.
\parhead{Nunchaku} The nunchaku is a special monk weapon. This designation gives a monk wielding a nunchaku special options. With a nunchaku, you get a \plus2 bonus on opposed attack rolls made to disarm an enemy.
\parhead{Quarterstaff} A quarterstaff is a double weapon. You can fight with it as if fighting with two weapons, but if you do, you incur all the normal attack penalties associated with fighting with two weapons, just as if you were using a medium weapon and a light weapon. A creature wielding a quarterstaff in one hand can't use it as a double weapon -- only one end of the weapon can be used in any given round.
\par The quarterstaff is a special monk weapon. This designation gives a monk wielding a quarterstaff special options.
\parhead{Ranseur} A ranseur has reach. You can strike opponents 10 feet away with it, but you can't use it against an adjacent foe unless you short haft it.
\par With a ranseur, you get a \plus2 bonus on opposed attack rolls made to disarm an opponent.
\parhead{Rapier} You can use the Weapon Finesse feat to apply your Dexterity instead of your Strength to attack rolls with a rapier sized for you, even though it isn't a light weapon for you. You can't wield a rapier in two hands in order to apply 1\minus1/2 times your Strength bonus to damage.
\parhead{Sai} With a sai, you get a \plus4 bonus on opposed attack rolls made to disarm an enemy.
\par The sai is a special monk weapon. This designation gives a monk wielding a sai special options.
\parhead{Scythe} A scythe can be used to make trip attacks. If you are tripped during your own trip attempt, you can drop the scythe to avoid being tripped.
\parhead{Shield, Heavy or Light} You can bash with a shield instead of using it for defense. See Armor for details.
\parhead{Shortbow} You need at least two hands to use a bow, regardless of its size. You can use a shortbow while mounted. If you have a penalty for low Strength, apply it to damage rolls when you use a shortbow. If you have a bonus for high Strength, you can apply it to damage rolls when you use a composite shortbow (see below) but not a regular shortbow.
\parhead{Shortbow, Composite} You need at least two hands to use a bow, regardless of its size. You can use a composite shortbow while mounted. All composite bows are made with a particular strength rating (that is, each requires a minimum Strength to use with proficiency). If your Strength bonus is lower than the strength rating of the composite bow, you can't effectively use it, so you take a \minus2 penalty on attacks with it. The default composite shortbow requires a Strength of \plus0 or higher to use with proficiency. A composite shortbow can be made with a high strength rating to take advantage of an above-average Strength score; this feature allows you to add your Strength bonus to damage, up to the maximum bonus indicated for the bow. Each point of Strength bonus granted by the bow adds 75 gp to its cost.
\par For purposes of weapon proficiency and similar feats, a composite shortbow is treated as if it were a shortbow.
\parhead{Shortspear} A shortspear is small enough to wield one-handed.
\parhead{Shuriken} A shuriken is a special monk weapon. This designation gives a monk wielding shuriken special options. A shuriken can't be used as a melee weapon.
\par Although they are thrown weapons, shuriken are treated as ammunition for the purposes of drawing them, crafting special versions of them, and what happens to them after they are thrown.
\par You can throw shuriken from a prone position without penalty.
\parhead{Siangham} The siangham is a special monk weapon. This designation gives a monk wielding a siangham special options.
\parhead{Sickle} A sickle can be used to make trip attacks. If you are tripped during your own trip attempt, you can drop the sickle to avoid being tripped.
\parhead{Sling} Your Strength applies to damage rolls when you use a sling, just as it does for thrown weapons. You can fire, but not load, a sling with one hand. Loading a sling is a move action that requires two hands and provokes attacks of opportunity.
\par You can hurl ordinary stones with a sling, but stones are not as dense or as round as bullets. Thus, such an attack deals damage as if the weapon were designed for a creature one size category smaller than you and you take a \minus1 penalty on attack rolls.
\parhead{Spear} If you use a ready action to set a spear against a charge, you deal double damage on your first successful hit against a charging character. When throwing a spear, you throw it with one hand as if it were a medium weapon.
\parhead{Spiked Armor} You can outfit your armor with spikes, which can deal damage in a grapple or bull rush. See Armor for details.
\parhead{Spiked Shield, Heavy or Light} You can bash with a spiked shield instead of using it for defense. See Armor for details.
\parhead{Strike, Unarmed} A Medium character deals 1d3 points of nonlethal damage with an unarmed strike. A Small character deals 1d2 points of nonlethal damage. A monk or any character with the Improved Unarmed Strike feat can deal lethal or nonlethal damage with unarmed strikes, at her option. The damage from an unarmed strike is considered weapon damage for the purposes of effects that give you a bonus on weapon damage rolls.
\par An unarmed strike is always considered a light weapon. Therefore, you can use the Weapon Finesse feat to apply your Dexterity instead of your Strength to attack rolls with an unarmed strike.
\parhead{Sword, Bastard} A bastard sword is too large to use in one hand without special training; thus, it is an exotic weapon. A character can use a bastard sword two-handed as a martial weapon.
\parhead{Sword, Two-Bladed} A two-bladed sword is a double weapon. You can fight with it as if fighting with two weapons, but if you do, you incur all the normal attack penalties associated with fighting with two weapons, just as if you were using a medium weapon and a light weapon. A creature wielding a two-bladed sword in one hand can't use it as a double weapon -- only one end of the weapon can be used in any given round.
\parhead{Trident} If you use a ready action to set a trident against a charge, you deal double damage on a successful hit against a charging character.
\parhead{Urgrosh, Dwarven} A dwarven urgrosh is a double weapon. You can fight with it as if fighting with two weapons, but if you do, you incur all the normal attack penalties associated with fighting with two weapons, just as if you were using a medium weapon and a light weapon. The urgrosh's axe head is a slashing weapon that deals 1d8 points of damage. Its spear head is a piercing weapon that deals 1d6 points of damage. You can use either head as the primary weapon. The other is the off-hand weapon. A creature wielding a dwarven urgrosh in one hand can't use it as a double weapon -- only one end of the weapon can be used in any given round.
\par If you use a ready action to set an urgrosh against a charge, you deal double damage on your first hit against the charging character. If you set an urgrosh against a charging character, the spear head is the part of the weapon that deals damage.
\parhead{Waraxe, Dwarven} A dwarven waraxe is too large to use in one hand without special training; thus, it is an exotic weapon. A Medium character can use a dwarven waraxe two-handed as a martial weapon, or a Large creature can use it one-handed in the same way. Dwarves do not treat dwarven waraxes as exotic weapons.
\parhead{Whip} A whip deals nonlethal damage. The whip is treated as a light melee weapon with 15 foot reach, though you don't threaten the area into which you can make an attack. In addition, unlike most other weapons with reach, you can use it against foes anywhere within your reach (including adjacent foes).
\par Using a whip provokes an attack of opportunity, just as if you had used a ranged weapon.
\par You can make trip attacks with a whip.
\par When using a whip, you get a \plus2 bonus on opposed attack rolls made to disarm an opponent.

\section{Armor}

Most characters use armor to protect themselves. There are two kinds of armor: body armor, such as chainmail or full plate, and shields.

\subsection{Armor Qualities}
\par Here is the format for armor entries (given as column headings on \trefnp{Armor and Shields}, below).

\parhead{Encumbrance} All armor restricts a character's movement to some degree. The heavier the armor, the more restrictive it is. Shields are divided into light shields (including bucklers), heavy shields, or tower shields, while body armor can be light, medium, or heavy. Many classes are not proficient with heavier kinds of armor.

\parhead{Cost} The cost of the armor for Small or Medium humanoid
creatures. See \trefnp{Armor for Unusual Creatures}, below, for armor prices for other creatures.
\parhead{Armor/Shield Bonus} Body armor improves your armor modifier to AC, while shields improve your shield modifier to AC. Wearing multiple suits of armor or wielding multiple shields does not improve your AC any further.

\parhead{Dexterity Modifier} Medium and heavy body armor limits mobility and agility, halving the character's Dexterity. This halving is applied in addition to the armor's armor check penalty (if any). A Dexterity penalty is not halved.

Even if a character's Dexterity drops to 0 because of armor, this situation does not count as becoming flat-footed.

Your character's encumbrance (the amount of gear he or she carries) may also affect your character's Dexterity bonus.
\subparhead{Shields} Most shields do not affect a character's Dexterity. However, a tower shield halves a character's Dexterity. This is not cumulative with the halving from wearing medium or heavy armor.
\parhead{Armor Check Penalty} Any armor heavier than leather hurts a character's ability to use some skills. An armor check penalty number is the penalty that applies to Acrobatics, Climb, Escape Artist, Jump, Sleight of Hand, and Stealth checks by a character wearing a certain kind of armor. Double the normal armor check penalty is applied to Swim checks. A character's encumbrance (the amount of gear carried, including armor) may also apply an armor check penalty.
\subparhead{Shields} If a character is wearing armor and using a shield, both armor check penalties apply.
\subparhead{Nonproficient with Armor Worn} A character who wears armor and/or uses a shield with which he or she is not proficient takes the armor's (and/or shield's) armor check penalty on attack rolls and on all Strength-based and Dexterity-based ability and skill checks. The penalty for nonproficiency with armor stacks with the penalty for nonproficiency with shields.

\subparhead{Sleeping in Armor} A character who sleeps overnight in medium or heavy body armor is automatically fatigued the next day. He or she takes a \minus2 penalty on Strength and Dexterity and can't charge or run. Sleeping in light armor does not cause fatigue.
\parhead{Arcane Spell Failure} Armor interferes with the gestures that a spellcaster must make to cast an arcane spell that has a somatic component. Arcane spellcasters face the possibility of arcane spell failure if they're wearing armor. Bards can wear light armor without incurring any arcane spell failure chance for their bard spells.

\subparhead{Casting an Arcane Spell in Armor} A character who casts an arcane spell while wearing armor must usually make an arcane spell failure roll. The number in the Arcane Spell Failure Chance column on Table: Armor and Shields is the chance that the spell fails and is ruined. If the spell lacks a somatic component, however, it can be cast with no chance of arcane spell failure.

\subparhead{Shields} If a character is wearing armor and using a shield, add the two numbers together to get a single arcane spell failure chance.
\parhead{Speed} Medium or heavy armor slows the wearer down. The number on Table: Armor and Shields is the character's speed while wearing the armor. Humans, elves, half-elves, and half-orcs have an unencumbered speed of 30 feet.
\par They use the first column. Dwarves, gnomes, and halflings have an unencumbered speed of 20 feet. They use the second column. Remember, however, that a dwarf's land speed remains 20 feet even in medium or heavy armor or when carrying a medium or heavy load.
\subparhead{Shields} Shields do not affect a character's speed.
\subparhead{Flat-Footed} A character's shield modifier to AC is lost when the character is flat-footed.
\parhead{Weight} This column gives the weight of the armor sized for a Medium wearer. Armor fitted for Small characters weighs half as much, and armor for Large characters weighs twice as much.

\subsubsection{Moving in Armor}
The heavier the armor, the more it slows your movement. Light armor has no effect on your movement speed, though it does impose armor check penalties. Running in medium or heavy armor is difficult, and you can run at only three times your movement speed instead of four times your movement speed. Heavy armor slows down your movement speed even further, causing you to move at 2/3 of your normal movement rate.

\subsection{Getting Into And Out Of Armor}
The time required to don armor depends on its type; see Table: Donning Armor.
\parhead{Don} This column tells how long it takes a character to put the armor on. (One minute is 10 rounds.) Readying (strapping on) a shield is only a move action.
\parhead{Don Hastily} This column tells how long it takes to put the armor on in a hurry. The armor check penalty and armor bonus for hastily donned armor are each 1 point worse than normal.
\parhead{Remove} This column tells how long it takes to get the armor off. Loosing a shield (removing it from the arm and dropping it) is only a move action.

\begin{dtable}
\lcaption{Donning Armor}
\begin{tabularx}{\columnwidth}{>{\lcol}X c c c}
\thead{Armor Type} & \thead{Don} & \thead{Don Hastily} & \thead{Remove} \\
Shield (any) & 1 move action & n/a & 1 move action \\
Padded, leather, hide, studded leather, or chain shirt & 1 minute & 5 rounds & 1 minute\fn{1} \\
Breastplate, scale mail, chainmail, banded mail, or splint mail & 4 minutes\fn{1} & 1 minute & 1 minute\fn{1} \\
Half-plate or full plate & 4 minutes\fn{2} & 4 minutes\fn{1} & 1d4\plus1 minutes\fn{1} \\
\end{tabularx}
1 If the character has some help, cut this time in half. A single character doing nothing else can help one or two adjacent characters. Two characters can?t help each other don armor at the same time. \\
2 The wearer must have help to don this armor. Without help, it can be donned only hastily.
\end{dtable}

\begin{dtable!*}
\lcaption{Armor and Shields}
\begin{tabularx}{\textwidth}{>{\lcol}X >{\ccol}p{6em} >{\ccol}p{5em} >{\ccol}p{6em} >{\ccol}p{7em} c c c c}
  &  &  &  &  &  \multicolumn{2}{c}{\x\x Speed \x\x} & & \\
\thead{Armor} & \thead{Armor/Shield Bonus} & \thead{Dex Modifier} & \thead{Armor Check Penalty} & \thead{Arcane Spell Failure Chance} & \thead{(30 ft.)} & \thead{(20 ft.)} & \thead{Cost} & \thead{Weight\fn{1}} \\
Light armor &  &  &  &  &  &  &  &  \\
\tind Padded 				& \plus1 		& 1\mult 	& 0       & 5\%  & 30 ft. & 20 ft. & 5 gp          & 10 lb.\\
\tind Leather 				& \plus2 		& 1\mult 	& \minus1 & 10\% & 30 ft. & 20 ft. & 10 gp        & 15 lb.\\
\tind Studded leather 		& \plus3 		& 1\mult 	& \minus2 & 15\% & 30 ft. & 20 ft. & 25 gp & 20 lb.\\
Medium armor 				&  &  &  &  &  &  &  & \\
\tind Chain shirt 			& \plus4 		& 1/2\mult 	& \minus2 & 20\% & 30 ft. & 20 ft. & 100 gp   & 25 lb.\\
\tind Hide 					& \plus4 		& 1/2\mult 	& \minus4 & 20\% & 30 ft. & 20 ft. & 15 gp           & 25 lb.\\
\tind Scale mail 			& \plus5 		& 1/2\mult 	& \minus4 & 25\% & 30 ft. & 20 ft. & 50 gp     & 30 lb.\\
\tind Chainmail 			& \plus6 		& 1/2\mult 	& \minus5 & 30\% & 30 ft. & 20 ft. & 100 gp     & 40 lb.\\
\tind Breastplate 			& \plus6 		& 1/2\mult 	& \minus4 & 25\% & 30 ft. & 20 ft. & 200 gp   & 30 lb.\\
Heavy armor 				&  &  &  &  &  &  &  & \\
\tind Splint mail 			& \plus7 		& 1/2\mult 	& \minus7 & 40\% & 20 ft. & 15 ft. & 200 gp   & 45 lb.\\
\tind Half-plate 			& \plus8 		& 1/2\mult 	& \minus8 & 40\% & 20 ft. & 15 ft. & 500 gp    & 50 lb.\\
\tind Full plate 			& \plus8 		& 1/2\mult 	& \minus6 & 35\% & 20 ft. & 15 ft. & 1,000 gp  & 50 lb.\\
Shields 					&  &  &  &  &  &  &  & \\
\tind Buckler 				& \plus1 		& \x 		& \minus1 & 5\% & \x & \x & 15 gp             & 5 lb.\\
\tind Shield, light wooden 	& \plus1 		& \x 		& \minus2 & 5\% & \x & \x & 3 gp & 5 lb.\\
\tind Shield, light steel 	& \plus1 		& \x 		& \minus2 & 5\% & \x & \x & 9 gp  & 6 lb.\\
\tind Shield, heavy wooden 	& \plus2 		& \x 		& \minus3 & 15\% & \x & \x & 7 gp & 10 lb.\\
\tind Shield, heavy steel 	& \plus2 		& \x 		& \minus3 & 15\% & \x & \x & 20 gp & 15 lb.\\
\tind Shield, tower 		& \plus4\fn{2} 	& 1/2\mult 	& \minus10 & 50\% & \x & \x & 30 gp       & 45 lb.\\
Extras 						&  &  &  &  &  &  &  & \\
\tind Armor spikes 			& \minus1\fn{3} & \x 		& \minus2 & \x & \x & \x & \plus50 gp & \plus10 lb.\\
\tind Gauntlet, locked 		& \x 			& \x 		& Special & \fn{4} & \x & \x & 8 gp & \plus5 lb.\\
\tind Shield spikes 		& \x 			& \x 		& \minus1 & \x & \x & \x & \plus10 gp & \plus5 lb.\\
\end{tabularx}
1 Weight figures are for armor sized to fit Medium characters. Armor fitted for Small characters weighs half as much, and armor fitted for Large characters weighs twice as much. \\
2 A tower shield can instead grant you cover. See the description. \\
3 Armor spikes reduce the AC bonus granted by the armor they are put on by 1. \\
4 Hand not free to cast spells.
\end{dtable!*}

\subsection{Armor Descriptions}
Any special benefits or accessories to the types of armor found on \trefnp{Armor and Shields} are described below.
\parhead{Armor Spikes} You can have spikes added to your armor, which allow you to deal extra piercing damage (see Table: Weapons) on a successful grapple, bull rush, or overrun attack. If you are not proficient with the spikes, you only deal the damage if you succeed at the attack by 10 or more.
\par An enhancement bonus to a suit of armor does not improve the spikes' effectiveness, but the spikes can be made into magic weapons in their own right.
\parhead{Banded Mail} The suit includes gauntlets.
\parhead{Breastplate} It comes with a helmet and greaves.
\parhead{Buckler} This small metal shield is worn strapped to your forearm. You can use a bow or crossbow without penalty while carrying it. You can also use your shield arm to wield a weapon (whether you are using an off-hand weapon or using your off hand to help wield a heavy weapon), but you take a \minus1 penalty on attack rolls while doing so. If you use a weapon in your off hand, you don't get the buckler's AC bonus for the rest of the round.
\par Unlike most shields, you do not apply your buckler's shield bonus to your touch armor class.
\par You can't bash someone with a buckler.
\parhead{Chain Shirt} A chain shirt comes with a steel cap.
\parhead{Chainmail} The suit includes gauntlets.
\parhead{Full Plate} The suit includes gauntlets, heavy leather boots, a visored helmet, and a thick layer of padding that is worn underneath the armor. Each suit of full plate must be individually fitted to its owner by a master armorsmith, although a captured suit can be resized to fit a new owner at a cost of 200 to 800 (2d4\mult100) gold pieces.
\parhead{Gauntlet, Locked} This armored gauntlet has small chains and braces that allow the wearer to attach a weapon to the gauntlet so that it cannot be dropped easily. It provides a \plus10 bonus on any roll made to keep from being disarmed in combat. Removing a weapon from a locked gauntlet or attaching a weapon to a locked gauntlet is a full-round action that provokes attacks of opportunity.
\par The price given is for a single locked gauntlet. The weight given applies only if you're wearing a breastplate, light armor, or no armor. Otherwise, the locked gauntlet replaces a gauntlet you already have as part of the armor.
\par While the gauntlet is locked, you can't use the hand wearing it for casting spells or employing skills. (You can still cast spells with somatic components, provided that your other hand is free.)
Like a normal gauntlet, a locked gauntlet lets you deal lethal damage rather than nonlethal damage with an unarmed strike.
\parhead{Half-Plate} The suit includes gauntlets.
\parhead{Scale Mail} The suit includes gauntlets.
\parhead{Shield, Heavy, Wooden or Steel} You strap a shield to your forearm and grip it with your hand. A heavy shield is so heavy that you can't use your shield hand for anything else.
\subparhead{Wooden or Steel} Wooden and steel shields offer the same basic protection, though they respond differently to special attacks.
\subparhead{Shield Bash Attacks} You can bash an opponent with a heavy shield, using it as a medium bludgeoning weapon. See \trefnp{Weapons} for the damage dealt by a shield bash.  If you use your shield as a weapon, you lose its AC bonus until your next action (usually until the next round). An enhancement bonus on a shield does not improve the effectiveness of a shield bash made with it, but the shield can be made into a magic weapon in its own right.
\parhead{Shield, Light, Wooden or Steel} You strap a shield to your forearm and grip it with your hand. A light shield's weight lets you carry other items in that hand, although you cannot use weapons or cast spells with it.
\subparhead{Wooden or Steel} Wooden and steel shields offer the same basic protection, though they respond differently to special attacks.
\subparhead{Shield Bash Attacks} You can bash an opponent with a light shield, using it as a light bludgeoning weapon. See \trefnp{Weapons} for the damage dealt by a shield bash. If you use your shield as a weapon, you lose its AC bonus until your next action (usually until the next round). An enhancement bonus on a shield does not improve the effectiveness of a shield bash made with it, but the shield can be made into a magic weapon in its own right.
\parhead{Shield, Tower} This massive wooden shield is nearly as tall as you are. In most situations, it provides the indicated shield bonus to your AC. However, when you take the total defense action with a tower shield, you can treat the shield as a wall along one edge of your square, providing you with total cover. The shield does not, however, provide cover against targeted spells; a spellcaster can cast a spell on you by targeting the shield you are holding. You cannot bash with a tower shield, nor can you use your shield hand for anything else.
\par When employing a tower shield in combat, you take a \minus2 penalty on attack rolls because of the shield's encumbrance.
\parhead{Shield Spikes} When added to your shield, these spikes turn it into a piercing weapon that increases the damage dealt by a shield bash as if the shield were designed for a creature one size category larger than you. You can't put spikes on a buckler or a tower shield. Otherwise, attacking with a spiked shield is like making a shield bash attack (see above).
\par An enhancement bonus on a spiked shield does not improve the effectiveness of a shield bash made with it, but a spiked shield can be made into a magic weapon in its own right.
\parhead{Splint Mail} The suit includes gauntlets.
\parhead{Studded Leather} Only the studs on studded leather are made of metal. If the studs are made from a special material, only the studs are affected, not the whole armor. For example, ironwood studded leather weighs 17.5 pounds, since the weight of the studs is halved.

\subsection{Armor for Unusual Creatures}
Armor and shields for unusually big creatures, unusually little creatures, and nonhumanoid creatures have different costs and weights from those given on \trefnp{Armor and Shields}. Refer to the appropriate line on the table below and apply the multipliers to cost and weight for the armor type in question.
\begin{dtable}
\begin{tabularx}{\columnwidth}{>{\lcol}X c c c c}
  & \multicolumn{2}{c}{\thead{Humanoid}} & \multicolumn{2}{c}{\thead{Nonhumanoid}} \\
\thead{Size} & \thead{Cost} & \thead{Weight} & \thead{Cost} & \thead{Weight} \\
Tiny or smaller\fn{1} & \mult1/2 & \mult1/10 & \mult1 & \mult1/10 \\
Small & \mult1 & \mult1/2 & \mult2 & \mult1/2 \\
Medium & \mult1 & \mult1 & \mult2 & \mult1 \\
Large & \mult2 & \mult2 & \mult4 & \mult2 \\
Huge & \mult4 & \mult4 & \mult8 & \mult4 \\
Gargantuan & \mult8 & \mult8 & \mult16 & \mult8 \\
Colossal & \mult16 & \mult12 & \mult32 & \mult12 \\
\end{tabularx}
1 Divide armor bonus by 2.
\end{dtable}

\subsection{Special Materials}
Armor and shields can be made from special materials, which can alter the properties of the item. The most common special material is ironwood, which is made from wood magically treated using the \spell{ironwood} ritual to be as strong as steel. Ironwood armor functions exactly like  armor of the same type, except that it is made of wood. This causes it to weigh half the normal weight, and allows druids to wear it without penalty. However, it costs twice as much as armor of the same type would normally cost, to a minimum of an additional 200 gp.

\section{Goods And Services}

\begin{dtable!*}
\lcaption{Goods and Services}
\begin{tabularx}{\textwidth}{>{\lcol}X c c >{\lcol}X c c >{\lcol}X c c}
\thead{Item} & \thead{Cost} & \thead{Weight} & \thead{Item} & \thead{Cost} & \thead{Weight} & \thead{Item} & \thead{Cost} & \thead{Weight} \\
\multicolumn{3}{>{\columncolor{tbrown}}c}{\thead{Adventuring Gear}} & \tind Average & 40 gp & 1 lb. & \multicolumn{3}{>{\columncolor{tbrown}}c}{\thead{Special Substances and Items}} \\
Backpack (empty) & 2 gp & 2 lb.\fn{1} & \tind Good & 80 gp & 1 lb. & Acid (flask) & 10 gp & 1 lb. \\
Barrel (empty) & 2 gp & 30 lb. & \tind Amazing & 150 gp & 1 lb. & Alchemist's fire (flask) & 20 gp & 1 lb. \\
Basket (empty) & 4 sp & 1 lb. & Manacles & 15 gp & 2 lb. & Antitoxin (vial) & 50 gp & \x \\
Bedroll & 1 sp & 5 lb.\fn{1} & Manacles, masterwork & 50 gp & 2 lb. & Everburning torch & 110 gp & 1 lb. \\
Bell & 1 gp & \x & Mirror, small steel & 10 gp & 1/2 lb. & Holy water (flask) & 25 gp & 1 lb. \\
Blanket, winter & 5 sp & 3 lb.\fn{1} & Mug/Tankard, clay & 2 cp & 1 lb. & Smokestick & 20 gp & 1/2 lb. \\
Block and tackle & 5 gp & 5 lb. & Oil (1-pint flask) & 1 sp & 1 lb. & Sunrod & 2 gp & 1 lb. \\
Bottle, wine, glass & 2 gp & \x & Paper (sheet) & 4 sp & \x & Tanglefoot bag & 50 gp & 4 lb. \\
Bucket (empty) & 5 sp & 2 lb. & Parchment (sheet) & 2 sp & \x & Thunderstone & 30 gp & 1 lb. \\
Caltrops & 1 gp & 2 lb. & Pick, miner's & 3 gp & 10 lb. & Tindertwig & 1 gp & \x \\
Candle & 1 cp & \x & Pitcher, clay & 2 cp & 5 lb. & \multicolumn{3}{c}{\thead{Tools and Skill Kits}}  \\
Canvas (sq. yd.) & 1 sp & 1 lb. & Piton & 1 sp & 1/2 lb. & Item & Cost & Weight \\
Case, map or scroll & 1 gp & 1/2 lb. & Pole, 10 foot & 5 cp & 8 lb. & Alchemist's lab & 500 gp & 40 lb. \\
Chain (10 ft.) & 30 gp & 2 lb. & Pot, iron & 5 sp & 10 lb. & Artisan's tools & 5 gp & 5 lb. \\
Chalk, 1 piece & 1 cp & \x & Pouch, belt (empty) & 1 gp & 1/2 lb.\fn{1} & Artisan's tools, masterwork & 55 gp & 5 lb. \\
Chest (empty) & 2 gp & 25 lb. & Ram, portable & 10 gp & 20 lb. & Climber's kit & 80 gp & 5 lb.\fn{1} \\
Crowbar & 2 gp & 5 lb. & Rations, trail (per day) & 5 sp & 1 lb.\fn{1} & Disguise kit & 50 gp & 8 lb.\fn{1} \\
Firewood (per day) & 1 cp & 20 lb. & Rope, hempen (50 ft.) & 1 gp & 10 lb. & Healer's kit & 50 gp & 1 lb. \\
Fishhook & 1 sp & \x & Rope, silk (50 ft.) & 10 gp & 5 lb. & Holly and mistletoe & \x & \x \\
Fishing net, 25 sq. ft. & 4 gp & 5 lb. & Sack (empty) & 1 sp & 1/2 lb.\fn{1} & Holy symbol, wooden & 1 gp & \x \\
Flask (empty) & 3 cp & 1-1/2 lb. & Sealing wax & 1 gp & 1 lb. & Holy symbol, silver & 25 gp & 1 lb. \\
Flint and steel & 1 gp & \x & Sewing needle & 5 sp & \x & Hourglass & 25 gp & 1 lb. \\
Grappling hook & 1 gp & 4 lb. & Signal whistle & 8 sp & \x & Magnifying glass & 100 gp & \x \\
Hammer & 5 sp & 2 lb. & Signet ring & 5 gp & \x & Musical instrument, common & 5 gp & 3 lb.\fn{1} \\
Ink (1 oz. vial) & 8 gp & \x & Sledge & 1 gp & 10 lb. & Musical instrument, masterwork & 100 gp & 3 lb.\fn{1} \\
Inkpen & 1 sp & ? & Soap (per lb.) & 5 sp & 1 lb. & Scale, merchant's & 2 gp & 1 lb. \\
Jug, clay & 3 cp & 9 lb. & Spade or shovel & 2 gp & 8 lb. & Spell component pouch & 5 gp & 2 lb. \\
Ladder, 10 foot & 2 sp & 20 lb. & Spyglass & 1,000 gp & 1 lb. & Spellbook, wizard's (blank) & 15 gp & 3 lb. \\
Lamp, common & 1 sp & 1 lb. & Tent & 10 gp & 20 lb.\fn{1} & Thieves' tools & 30 gp & 1 lb. \\
Lantern, bullseye & 12 gp & 3 lb. & Torch & 1 cp & 1 lb. & Thieves' tools, masterwork & 100 gp & 2 lb. \\
Lantern, hooded & 7 gp & 2 lb. & Vial, ink or potion & 1 gp & 1/10 lb. & Tool, masterwork & 50 gp & 1 lb. \\
Lock &   & 1 lb. & Waterskin & 1 gp & 4 lb.\fn{1} & Water clock & 1,000 gp & 200 lb. \\
\tind Very simple & 20 gp & 1 lb. & Whetstone & 2 cp & 1 lb. &  &  &  \\
\end{tabularx}
1 These items weigh one-quarter this amount when made for Small characters. Containers for Small characters also carry one-quarter the normal amount. \\
\x No weight, or no weight worth noting.	
\end{dtable!*}

\begin{dtable!*}
\begin{tabularx}{\textwidth}{>{\lcol}X c c >{\lcol}X c c >{\lcol}X c c}
\thead{Clothing} &  &  & \thead{Mounts and Related Gear} &  &  & \thead{Transport} &  & \\
\thead{Item} & \thead{Cost} & \thead{Weight} & \thead{Item} & \thead{Cost} & \thead{Weight} & \thead{Item} & \thead{Cost} & \thead{Weight} \\
Artisan's outfit & 1 gp & 4 lb.1 & Barding &  &  & Carriage & 100 gp & 600 lb. \\
Cleric's vestments & 5 gp & 6 lb.1 & Medium creature & x2 & x1 & Cart & 15 gp & 200 lb. \\
Cold weather outfit & 8 gp & 7 lb.1 & Large creature & x4 & x2 & Galley & 30,000 gp & \x \\
Courtier's outfit & 30 gp & 6 lb.1 & Bit and bridle & 2 gp & 1 lb. & Keelboat & 3,000 gp & \x \\
Entertainer's outfit & 3 gp & 4 lb.1 & Dog, guard & 25 gp & \x & Longship & 10,000 gp & \x \\
Explorer's outfit & 10 gp & 8 lb.1 & Dog, riding & 150 gp & \x & Rowboat & 50 gp & 100 lb. \\
Monk's outfit & 5 gp & 2 lb.1 & Donkey or mule & 8 gp & \x & Oar & 2 gp & 10 lb. \\
Noble's outfit & 75 gp & 10 lb.1 & Feed (per day) & 5 cp & 10 lb. & Sailing ship & 10,000 gp & \x \\
Peasant's outfit & 1 sp & 2 lb.1 & Horse &  &  & Sled & 20 gp & 300 lb. \\
Royal outfit & 200 gp & 15 lb.1 & Horse, heavy & 200 gp & \x & Wagon & 35 gp & 400 lb. \\
Scholar's outfit & 5 gp & 6 lb.1 & Horse, light & 75 gp & \x & Warship & 25,000 gp & \x \\
Traveler's outfit & 1 gp & 5 lb.1 & Pony & 30 gp & \x & \thead{Spellcasting and Services} &  &  \\
\thead{Food, Drink, and Lodging} &  &  & Warhorse, heavy & 400 gp & \x & Service & \multicolumn{2}{l}{Cost} \\
Item & Cost & Weight & Warhorse, light & 150 gp & \x & Coach cab & \multicolumn{2}{l}{3 cp per mile} \\
Ale &  &  & Warpony & 100 gp & \x & Hireling, trained & \multicolumn{2}{l}{3 sp per day} \\
Gallon & 2 sp & 8 lb. & Saddle &  &  & Hireling, untrained & \multicolumn{2}{l}{1 sp per day} \\
Mug & 4 cp & 1 lb. & Military & 20 gp & 30 lb. & Messenger & \multicolumn{2}{l}{2 cp per mile} \\
Banquet (per person) & 10 gp & \x & Pack & 5 gp & 15 lb. & Road or gate toll & \multicolumn{2}{l}{1 cp} \\
Bread, per loaf & 2 cp & 1/2 lb. & Riding & 10 gp & 25 lb. & Ship's passage & \multicolumn{2}{l}{1 sp per mile} \\
Cheese, hunk of & 1 sp & 1/2 lb. & \thead{Saddle, Exotic} &  &  &  & & \\
Inn stay (per day) &  &  & Military & 60 gp & 40 lb. & Spell, 1st-level & \multicolumn{2}{l}{Caster level x 10 gp\fn{1}} \\
Good & 2 gp & \x & Pack & 15 gp & 20 lb. & Spell, 2nd-level & \multicolumn{2}{l}{Caster level x 20 gp\fn{1}} \\
Common & 5 sp & \x & Riding & 30 gp & 30 lb. & Spell, 3rd-level & \multicolumn{2}{l}{Caster level x 30 gp\fn{1}} \\
Poor & 2 sp & \x & Saddlebags & 4 gp & 8 lb. & Spell, 4th-level & \multicolumn{2}{l}{Caster level x 40 gp\fn{1}} \\
Meals (per day) &  &  & Stabling (per day) & 5 sp & \x & Spell, 5th-level & \multicolumn{2}{l}{Caster level x 50 gp\fn{1}} \\
Good & 5 sp & \x & &  &  & Spell, 6th-level & \multicolumn{2}{l}{Caster level x 60 gp\fn{1}} \\
Common & 3 sp & \x & &  &  & Spell, 7th-level & \multicolumn{2}{l}{Caster level x 70 gp\fn{1}} \\
Poor & 1 sp & \x & &  &  & Spell, 8th-level & \multicolumn{2}{l}{Caster level x 80 gp\fn{1}} \\
Meat, chunk of & 3 sp & 1/2 lb. & &  &  & Spell, 9th-level & \multicolumn{2}{l}{Caster level x 90 gp\fn{1}} \\
Wine &  &  & &  &  &  &  & \\
Common (pitcher) & 2 sp & 6 lb. & &  &  &  &  & \\
Fine (bottle) & 10 gp & 1-1/2 lb. & & & & & & \\
\end{tabularx}
1 See spell description for additional costs. If the additional costs put the spell's total cost above 2,000 gp, that spell is not generally available.
\end{dtable!*}

\subsection{Adventuring Gear}
A few of the pieces of adventuring gear found on Table: Goods and Services are described below, along with any special benefits they confer on the user ("you").
\parhead{Caltrops} A caltrop is a four-pronged iron spike crafted so that one prong faces up no matter how the caltrop comes to rest. You scatter caltrops on the ground in the hope that your enemies step on them or are at least forced to slow down to avoid them. One 2- pound bag of caltrops covers an area 5 feet square.
\par Each time a creature moves into an area covered by caltrops (or spends a round fighting while standing in such an area), it might step on one. The caltrops make an attack roll (base attack bonus \plus0) against the creature. For this attack, the creature's shield, armor, and deflection bonuses do not count. If the creature is wearing shoes or other footwear, it gets a \plus2 armor bonus to AC. If the caltrops succeed on the attack, the creature has stepped on one. The caltrop deals 1 point of damage, and the creature's speed is reduced by one-half because its foot is wounded. This movement penalty lasts for 24 hours, or until the creature is successfully treated with a DC 15 Heal check, or until it receives at least 1 point of magical curing. A charging or running creature must immediately stop if it steps on a caltrop. Any creature moving at half speed or slower can pick its way through a bed of caltrops with no trouble.
\par Caltrops may not be effective against unusual opponents.
\parhead{Candle} A candle dimly illuminates a 5 foot radius and burns for 1 hour.
\parhead{Chain} Chain has hardness 10 and 5 hit points. It can be burst with a DC 26 Strength check.
\parhead{Crowbar} A crowbar it grants a \plus2 circumstance bonus on Strength checks made for such purposes. If used in combat, treat a crowbar as a medium improvised weapon that deals bludgeoning damage equal to that of a club of its size.
\parhead{Flint and Steel} Lighting a torch with flint and steel is a full-round action, and lighting any other fire with them takes at least that long.
\parhead{Grappling Hook} Throwing a grappling hook successfully requires a combat maneuver attack (DC 10, \plus2 per 10 feet of distance thrown).
\parhead{Hammer} If a hammer is used in combat, treat it as a medium improvised weapon that deals bludgeoning damage equal to that of a spiked gauntlet of its size.
\parhead{Ink} This is black ink. You can buy ink in other colors, but it costs twice as much.
\parhead{Jug, Clay} This basic ceramic jug is fitted with a stopper and holds 1 gallon of liquid.
\parhead{Lamp, Common} A lamp clearly illuminates a 15 foot radius, provides shadowy illumination out to a 30 foot radius, and burns for 6 hours on a pint of oil. You can carry a lamp in one hand.
\parhead{Lantern, Bullseye} A bullseye lantern provides clear illumination in a 60 foot cone and shadowy illumination in a 120 foot cone. It burns for 6 hours on a pint of oil. You can carry a bullseye lantern in one hand.
\parhead{Lantern, Hooded} A hooded lantern clearly illuminates a 30 foot radius and provides shadowy illumination in a 60 foot radius. It burns for 6 hours on a pint of oil. You can carry a hooded lantern in one hand.
\parhead{Lock} The DC to open a lock with the Open Lock skill depends on the lock's quality: simple (DC 20), average (DC 25), good (DC 30), or superior (DC 40).
\parhead{Manacles and Manacles, Masterwork} Manacles can bind a Medium creature. A manacled creature can use the Escape Artist skill to slip free (DC 30, or DC 35 for masterwork manacles). Breaking the manacles requires a Strength check (DC 26, or DC 28 for masterwork manacles). Manacles have hardness 10 and 10 hit points.
\par Most manacles have locks; add the cost of the lock you want to the cost of the manacles.
\par For the same cost, you can buy manacles for a Small creature.
\par For a Large creature, manacles cost ten times the indicated amount, and for a Huge creature, one hundred times this amount. Gargantuan, Colossal, Tiny, Diminutive, and Fine creatures can be held only by specially made manacles.
\parhead{Oil} A pint of oil burns for 6 hours in a lantern. You can use a flask of oil as a splash weapon. Use the rules for alchemist's fire, except that it takes a full round action to prepare a flask with a fuse. Once it is thrown, there is a 50\% chance of the flask igniting successfully.
\par You can pour a pint of oil on the ground to cover an area 5 feet square, provided that the surface is smooth. If lit, the oil burns for 2 rounds and deals 1d3 points of fire damage to each creature in the area.
\parhead{Ram, Portable} This iron-shod wooden beam gives you a \plus2 circumstance bonus on Strength checks made to break open a door and it allows a second person to help you without having to roll, increasing your bonus by 2.
\parhead{Rope, Hempen} This rope has 2 hit points and can be burst with a DC 23 Strength check.
\parhead{Rope, Silk} This rope has 4 hit points and can be burst with a DC 24 Strength check. It is so supple that it provides a \plus2 circumstance bonus on Use Rope checks.
\parhead{Spyglass} Objects viewed through a spyglass are magnified to twice their size.
\parhead{Torch} A torch burns for 1 hour, clearly illuminating a 20 foot radius and providing shadowy illumination out to a 40 foot radius. If a torch is used in combat, treat it as a light improvised weapon that deals bludgeoning damage equal to that of a gauntlet of its size, plus 1 point of fire damage. Wielding a torch gives a \plus2 circumstance bonus to dirty trick attempts made to dazzle an opponent. Lighting a torch with an open flame or tindertwig is a standard action, while using flint and steel requires a full-round action.
\parhead{Vial} A vial holds 1 ounce of liquid. The stoppered container usually is no more than 1 inch wide and 3 inches high.

\subsection{Special Substances And Items}
Any of these substances except for the everburning torch and holy water can be made by a character with the Craft (alchemy) skill.
\parhead{Acid} You can throw a flask of acid as a splash weapon. Treat this attack as a ranged touch attack with a range increment of 10 feet. A direct hit deals 1d6 points of acid damage. Every creature within 5 feet of the point where the acid hits takes 1 point of acid damage from the splash.
\parhead{Alchemist's Fire} You can throw a flask of alchemist's fire as a splash weapon. Treat this attack as a ranged touch attack with a range increment of 10 feet.
\par A direct hit deals 1d6 points of fire damage. Every creature within 5 feet of the point where the flask hits takes 1 point of fire damage from the splash. On the round following a direct hit, the target takes an additional 1d6 points of damage. If desired, the target can use a full-round action to attempt to extinguish the flames before taking this additional damage. Extinguishing the flames requires a DC 15 Reflex save. Rolling on the ground provides the target a \plus2 bonus on the save. Leaping into a lake or magically extinguishing the flames automatically smothers the fire.
\parhead{Antitoxin} If you drink antitoxin, you get a \plus5 alchemical bonus on Fortitude saving throws against poison for 1 hour.
\parhead{Everburning Torch} This otherwise normal torch has a continual flame spell cast upon it. An everburning torch clearly illuminates a 20 foot radius and provides shadowy illumination out to a 40 foot radius.
\parhead{Holy Water} Holy water damages undead creatures and evil outsiders almost as if it were acid. A flask of holy water can be thrown as a splash weapon.
\par Treat this attack as a ranged touch attack with a range increment of 10 feet. A flask breaks if thrown against the body of a corporeal creature, but to use it against an incorporeal creature, you must open the flask and pour the holy water out onto the target. Thus, you can douse an incorporeal creature with holy water only if you are adjacent to it. Doing so is a ranged touch attack that does not provoke attacks of opportunity.
\par A direct hit by a flask of holy water deals 2d4 points of damage to an undead creature or an evil outsider. Each such creature within 5 feet of the point where the flask hits takes 1 point of damage from the splash.
\par Temples to good deities sell holy water at cost (making no profit).
\parhead{Smokestick} This alchemically treated wooden stick instantly creates thick, opaque smoke when ignited. The smoke fills a 10 foot cube (treat the effect as a fog cloud spell, except that a moderate or stronger wind dissipates the smoke in 1 round). The stick is consumed after 1 round, and the smoke dissipates naturally.
\parhead{Sunrod} This 1 foot long, gold-tipped, iron rod glows brightly when struck. It clearly illuminates a 30 foot radius and provides shadowy illumination in a 60 foot radius. It glows for 6 hours, after which the gold tip is burned out and worthless.
\parhead{Tanglefoot Bag} When you throw a tanglefoot bag at a creature (as a ranged touch attack with a range increment of 10 feet), the bag comes apart and the goo bursts out, entangling the target and then becoming tough and resilient upon exposure to air. An entangled creature takes a \minus2 penalty on attack rolls and a \minus4 penalty to Dexterity and must make a DC 15 Reflex save or be glued to the floor, unable to move. Even on a successful save, it can move only at half speed. Huge or larger creatures are unaffected by a tanglefoot bag. A flying creature is not stuck to the floor, but it must make a DC 15 Reflex save or be unable to fly (assuming it uses its wings to fly) and fall to the ground. A tanglefoot bag does not function underwater.
\par A creature that is glued to the floor (or unable to fly) can break free by making a DC 17 Strength check or by dealing 15 points of damage to the goo with a slashing weapon. A creature trying to scrape goo off itself, or another creature assisting, does not need to make an attack roll; hitting the goo is automatic, after which the creature that hit makes a damage roll to see how much of the goo was scraped off. Once free, the creature can move (including flying) at half speed. A character capable of spellcasting who is bound by the goo must make a DC 15 Concentration check to cast a spell. The goo becomes brittle and fragile after 2d4 rounds, cracking apart and losing its effectiveness. An application of universal solvent to a stuck creature dissolves the alchemical goo immediately.
\parhead{Thunderstone} You can throw this stone as a ranged attack with a range increment of 20 feet. When it strikes a hard surface (or is struck hard), it creates a deafening bang that is treated as a sonic attack. Each creature within a 10 foot radius spread must make a DC 15 Fortitude save or be deafened for 1 hour. A deafened creature, in addition to the obvious effects, has a 20\% chance to miscast and lose any spell with a verbal component that it tries to cast.
\par Since you don't need to hit a specific target, you can simply aim at a particular 5 foot square. Treat the target square as AC 5.
\parhead{Tindertwig} The alchemical substance on the end of this small, wooden stick ignites when struck against a rough surface. Creating a flame with a tindertwig is much faster than creating a flame with flint and steel (or a magnifying glass) and tinder. Lighting a torch with a tindertwig is a standard action (rather than a full-round action), and lighting any other fire with one is at least a standard action.

\subsection{Tools and Skill Kits}
\parhead{Alchemist's Lab} An alchemist's lab always has the perfect tool for making alchemical items, so it provides a \plus2 circumstance bonus on Craft (alchemy) checks. It has no bearing on the costs related to the Craft (alchemy) skill. Without this lab, a character with the Craft (alchemy) skill is assumed to have enough tools to use the skill but not enough to get the \plus2 bonus that the lab provides.
\parhead{Artisan's Tools} These special tools include the items needed to pursue any craft. Without them, you have to use improvised tools (?2 penalty on Craft checks), if you can do the job at all.
\parhead{Artisan's Tools, Masterwork} These tools serve the same purpose as artisan's tools (above), but masterwork artisan's tools are the perfect tools for the job, so you get a \plus2 circumstance bonus on Craft checks made with them.
\parhead{Climber's Kit} This is the perfect tool for climbing and gives you a \plus2 circumstance bonus on Climb checks.
\parhead{Disguise Kit} The kit is the perfect tool for disguise and provides a \plus2 circumstance bonus on Disguise checks. A disguise kit is exhausted after ten uses.
\parhead{Healer's Kit} It is the perfect tool for healing and provides a \plus2 circumstance bonus on Heal checks. A healer's kit is exhausted after ten uses.
\parhead{Holy Symbol, Silver or Wooden} A holy symbol focuses positive energy. A cleric or paladin uses it as the focus for his spells and as a tool for turning undead. Each religion has its own holy symbol.
\parhead{Unholy Symbols} An unholy symbol is like a holy symbol except that it focuses negative energy and is used by evil clerics (or by neutral clerics who want to cast evil spells or command undead).
\parhead{Magnifying Glass} This simple lens allows a closer look at small objects. It is also useful as a substitute for flint and steel when starting fires. Lighting a fire with a magnifying glass requires light as bright as sunlight to focus, tinder to ignite, and at least a full-round action. A magnifying glass grants a \plus2 circumstance bonus on Appraise checks
involving any item that is small or highly detailed.
\parhead{Musical Instrument, Common or Masterwork} A masterwork instrument grants a \plus2 circumstance bonus on Perform checks involving its use.
\parhead{Spell Component Pouch} A spellcaster with a spell component pouch is assumed to have all the material components and focuses needed for spellcasting, except for those components that have a specific cost, divine focuses, and focuses that wouldn't fit in a pouch.
\parhead{Spellbook, Wizard's (Blank)} A spellbook has 100 pages of parchment, and each spell takes up one page per spell level (one page each for 0-level spells).
\parhead{Thieves' Tools} This kit contains the tools you need to use the Disable Device and Open Lock skills. Without these tools, you must improvise tools, and you take a ?2 circumstance penalty on Disable Device and Open Locks checks.
\parhead{Thieves' Tools, Masterwork} This kit contains extra tools and tools of better make, which grant a \plus2 circumstance bonus on Disable Device and Open Lock checks.
\parhead{Tool, Masterwork} This well-made item is the perfect tool for the job. It grants a \plus2 circumstance bonus on a related skill check (if any). Bonuses provided by multiple masterwork items used toward the same skill check do not stack.
\parhead{Water Clock} This large, bulky contrivance gives the time accurate to within half an hour per day since it was last set. It requires a source of water, and it must be kept still because it marks time by the regulated flow of droplets of water.

\subsection{Clothing}
\parhead{Artisan's Outfit} This outfit includes a shirt with buttons, a skirt or pants with a drawstring, shoes, and perhaps a cap or hat. It may also include a belt or a leather or cloth apron for carrying tools.
\parhead{Cleric's Vestments} These ecclesiastical clothes are for performing priestly functions, not for adventuring.
\parhead{Cold Weather Outfit} A cold weather outfit includes a wool coat, linen shirt, wool cap, heavy cloak, thick pants or skirt, and
boots. This outfit grants a \plus5 circumstance bonus on Fortitude saving throws against exposure to cold weather.
\parhead{Courtier's Outfit} This outfit includes fancy, tailored clothes in whatever fashion happens to be the current style in the courts of the nobles. Anyone trying to influence nobles or courtiers while wearing street dress will have a hard time of it (\minus2 penalty on Charisma-based skill checks to influence such individuals). If you wear this outfit without jewelry (costing an additional 50 gp), you look like an out-of-place commoner.
\parhead{Entertainer's Outfit} This set of flashy, perhaps even gaudy, clothes is for entertaining. While the outfit looks whimsical, its practical design lets you tumble, dance, walk a tightrope, or just run (if the audience turns ugly).
\parhead{Explorer's Outfit} This is a full set of clothes for someone who never knows what to expect. It includes sturdy boots, leather breeches or a skirt, a belt, a shirt (perhaps with a vest or jacket), gloves, and a cloak. Rather than a leather skirt, a leather overtunic may be worn over a cloth skirt. The clothes have plenty of pockets (especially the cloak). The outfit also includes any extra items you might need, such as a scarf or a wide-brimmed hat.
\parhead{Monk's Outfit} This simple outfit includes sandals, loose breeches, and a loose shirt, and is all bound together with sashes. The outfit is designed to give you maximum mobility, and it's made of high-quality fabric. You can hide small weapons in pockets hidden in the folds, and the sashes are strong enough to serve as short ropes.
\parhead{Noble's Outfit} This set of clothes is designed specifically to be expensive and to show it. Precious metals and gems are worked into the clothing. To fit into the noble crowd, every would-be noble also needs a signet ring (see Adventuring Gear, above) and jewelry (worth at least 100 gp).
\parhead{Peasant's Outfit} This set of clothes consists of a loose shirt and baggy breeches, or a loose shirt and skirt or overdress. Cloth wrappings are used for shoes.
\parhead{Royal Outfit} This is just the clothing, not the royal scepter, crown, ring, and other accoutrements. Royal clothes are ostentatious, with gems, gold, silk, and fur in abundance.
\parhead{Scholar's Outfit} Perfect for a scholar, this outfit includes a robe, a belt, a cap, soft shoes, and possibly a cloak.
\parhead{Traveler's Outfit} This set of clothes consists of boots, a wool skirt or breeches, a sturdy belt, a shirt (perhaps with a vest or jacket), and an ample cloak with a hood.

\subsection{Food, Drink, and Lodging}
\parhead{Inn} Poor accommodations at an inn amount to a place on the floor near the hearth. Common accommodations consist of a place on a raised, heated floor, the use of a blanket and a pillow. Good accommodations consist of a small, private room with one bed, some amenities, and a covered chamber pot in the corner.
\parhead{Meals} Poor meals might be composed of bread, baked turnips, onions, and water. Common meals might consist of bread, chicken stew, carrots, and watered-down ale or wine. Good meals might be composed of bread and pastries, beef, peas, and ale or wine.

\subsection{Mounts and Related Gear}
\parhead{Barding, Medium Creature and Large Creature} Barding is a type of armor that covers the head, neck, chest, body, and possibly legs of a horse or other mount. Barding made of medium or heavy armor provides better protection than light barding, but at the expense of speed. Barding can be made of any of the armor types found on Table: Armor and Shields.
\par Armor for a horse (a Large nonhumanoid creature) costs four times as much as armor for a human (a Medium humanoid creature) and also weighs twice as much as the armor found on Table: Armor and Shields (see Armor for Unusual Creatures). If the barding is for a pony or other Medium mount, the cost is only double, and the weight is the same as for Medium armor worn by a humanoid. Medium or heavy barding slows a mount that wears it, as shown on the table below.

\begin{dtable}
\begin{tabularx}{\columnwidth}{>{\lcol}X *{3}{>{\ccol}X}}
 & \multicolumn{3}{c}{\x\x\x Base Speed\x\x\x} \\
Barding & (40 ft.) & (50 ft.) & (60 ft.) \\
Medium & 30 ft. & 35 ft. & 40 ft. \\
Heavy & 30 ft.\fn{1} & 35 ft.\fn{1} & 40 ft.\fn{1} \\
\end{tabularx}
1 A mount wearing heavy armor moves at only triple its normal speed when running instead of quadruple.	
\end{dtable}

\par Flying mounts can't fly in medium or heavy barding.
\par Removing and fitting barding takes five times as long as the figures given on Table: Donning Armor. A barded animal cannot be used to carry any load other than the rider and normal saddlebags.
\parhead{Dog, Riding} This Medium dog is specially trained to carry a Small humanoid rider. It is brave in combat like a warhorse. You take no damage when you fall from a riding dog.
\parhead{Donkey or Mule} Donkeys and mules are stolid in the face of danger, hardy, surefooted, and capable of carrying heavy loads over vast distances. Unlike a horse, a donkey or a mule is willing (though not eager) to enter dungeons and other strange or threatening places.
\parhead{Feed} Horses, donkeys, mules, and ponies can graze to sustain themselves, but providing feed for them is much better. If you have a riding dog, you have to feed it at least some meat.
\parhead{Horse} A horse (other than a pony) is suitable as a mount for a human, dwarf, elf, half-elf, or half-orc. A pony is smaller than a horse and is a suitable mount for a gnome or halfling.
\par Warhorses and warponies can be ridden easily into combat. Light horses, ponies, and heavy horses are hard to control in combat.
\parhead{Saddle, Exotic} An exotic saddle is like a normal saddle of the same sort except that it is designed for an unusual mount. Exotic saddles come in military, pack, and riding styles.
\parhead{Saddle, Military} A military saddle braces the rider, providing a \plus2 circumstance bonus on Ride checks related to staying in the saddle. If you're knocked unconscious while in a military saddle, you have a 75\% chance to stay in the saddle (compared to 50\% for a riding saddle).
\parhead{Saddle, Pack} A pack saddle holds gear and supplies, but not a rider. It holds as much gear as the mount can carry.
\parhead{Saddle, Riding} The standard riding saddle supports a rider.

\subsection{Transport}
\parhead{Carriage} This four-wheeled vehicle can transport as many as four people within an enclosed cab, plus two drivers. In general, two horses (or other beasts of burden) draw it. A carriage comes with the harness needed to pull it.
\parhead{Cart} This two-wheeled vehicle can be drawn by a single horse (or other beast of burden). It comes with a harness.
\parhead{Galley} This three-masted ship has seventy oars on either side and requires a total crew of 200. A galley is 130 feet long and 20 feet wide, and it can carry 150 tons of cargo or 250 soldiers. For 8,000 gp more, it can be fitted with a ram and castles with firing platforms fore, aft, and amidships. This ship cannot make sea voyages and sticks to the coast. It moves about 4 miles per hour when being rowed or under sail.
\parhead{Keelboat} This 50- to 75 foot long ship is 15 to 20 feet wide and has a few oars to supplement its single mast with a square sail. It has a crew of eight to fifteen and can carry 40 to 50 tons of cargo or 100 soldiers. It can make sea voyages, as well as sail down rivers (thanks to its flat bottom). It moves about 1 mile per hour.
\parhead{Longship} This 75 foot long ship with forty oars requires a total crew of 50. It has a single mast and a square sail, and it can carry 50 tons of cargo or 120 soldiers. A longship can make sea voyages. It moves about 3 miles per hour when being rowed or under sail.
\parhead{Rowboat} This 8- to 12 foot long boat holds two or three Medium passengers. It moves about 1-1/2 miles per hour.
\parhead{Sailing Ship} This larger, seaworthy ship is 75 to 90 feet long and 20 feet wide and has a crew of 20. It can carry 150 tons of cargo. It has square sails on its two masts and can make sea voyages. It moves about 2 miles per hour.
\parhead{Sled} This is a wagon on runners for moving through snow and over ice. In general, two horses (or other beasts of burden) draw it. A sled comes with the harness needed to pull it.
\parhead{Wagon} This is a four-wheeled, open vehicle for transporting heavy loads. In general, two horses (or other beasts of burden) draw it. A wagon comes with the harness needed to pull it.
\parhead{Warship} This 100 foot long ship has a single mast, although oars can also propel it. It has a crew of 60 to 80 rowers. This ship can carry 160 soldiers, but not for long distances, since there isn't room for supplies to support that many people. The warship cannot make sea voyages and sticks to the coast. It is not used for cargo. It moves about 2-1/2 miles per hour when being rowed or under sail.

\subsection{Spellcasting And Services}
Sometimes the best solution for a problem is to hire someone else to take care of it.
\parhead{Coach Cab} The price given is for a ride in a coach that transports people (and light cargo) between towns. For a ride in a cab that transports passengers within a city, 1 copper piece usually takes you anywhere you need to go.
\parhead{Hireling, Trained} The amount given is the typical daily wage for mercenary warriors, masons, craftsmen, scribes, teamsters, and other trained hirelings. This value represents a minimum wage; many such hirelings require significantly higher pay.
\parhead{Hireling, Untrained} The amount shown is the typical daily wage for laborers, porters, cooks, maids, and other menial workers.
\parhead{Messenger} This entry includes horse-riding messengers and runners. Those willing to carry a message to a place they were going anyway may ask for only half the indicated amount.
\parhead{Road or Gate Toll} A toll is sometimes charged to cross a well-trodden, well-kept, and well-guarded road to pay for patrols on it and for its upkeep. Occasionally, a large walled city charges a toll to enter or exit (or sometimes just to enter).
\parhead{Ship's Passage} Most ships do not specialize in passengers, but many have the capability to take a few along when transporting cargo. Double the given cost for creatures larger than Medium or creatures that are otherwise difficult to bring aboard a ship.
\parhead{Spell} The indicated amount is how much it costs to get a spellcaster to cast a spell for you. This cost assumes that you can go to the spellcaster and have the spell cast at his or her convenience. If you want to bring the spellcaster somewhere to cast a spell you need to negotiate with him or her, and the default answer is no.
\par The cost given is for a spell with no cost for a material component or focus component. If the spell includes a material component, add the cost of that component to the cost of the spell.
\par If the spell has a focus component (other than a divine focus), add 1/10 the cost of that focus to the cost of the spell.
\par Furthermore, if a spell has dangerous consequences, the spellcaster will certainly require proof that you can and will pay for dealing with any such consequences (that is, assuming that the spellcaster even agrees to cast such a spell, which isn't certain). In the case of spells that transport the caster and characters over a distance, you will likely have to pay for two castings of the spell, even if you aren't returning with the caster.
\par In addition, not every town or village has a spellcaster of sufficient level to cast any spell. In general, you must travel to a small town (or larger settlement) to be reasonably assured of finding a spellcaster capable of casting 1st-level spells, a large town for 2nd-level spells, a small city for 3rd- or 4th-level spells, a large city for 5th- or 6th-level spells, and a metropolis for 7th- or 8th-level spells. Even a metropolis isn't guaranteed to have a local spellcaster able to cast 9th-level spells. Additionally, while you may find spellcasters able to cast some spells, there is no guarantee that they are able to cast the spell you desire. In general, the more common the spell, the more likely you are to find a spellcaster able to cast it.
