\chapter{Feats}\label{Feats}
Feats are special abilities that every character has. Feats can be used to specialize your character particular area, to grant your character new abilities, or to change the way your character does certain things.

\section{Gaining Feats}
Your character gains a feat every odd level: 1st, 3rd, 5th, and so on. Classes can sometimes grant bonus feats as well, which are in addition to these feats which every character gets.

\subsection{Prerequisites}
Some feats have prerequisites. Your character must have the indicated attribute score, class feature, feat, skill, base attack bonus, or other quality designated in order to select or use that feat. A character can gain a feat at the same level at which he or she gains the prerequisite.

A character can't use a feat if he or she has lost a prerequisite.

\section{Types Of Feats}
Some feats are general, meaning that no special rules govern
them as a group. Others belong to particular categories. The categories are given below:
\begin{itemize*}
    \item Bloodline feats allow a character to tap into latent magical abilities in their blood.
    \item Item creation feats allow spellcasters to create magic items of various kinds.
    \item Magic feats affect the way a character casts spells or uses magic.
    \item Metamagic feats let spellcasters cast spells with greater effect, albeit as if the spells were a higher level than they actually are.
    \item Performance feats allow a character to create magical effects through skilled performances.
    \item Skill feats affect a character's mastery of his or her skills.
    \item Spellgift feats grant a non-spellcasting character a limited ability to use spells.
    \item Surge feats allow a character to temporarily gain supernatural abilities.
    \item Combat feats affect a character's prowess in combat.
    \item Combat maneuver feats are combat feats which grant a character particular talent with specific maneuvers.
    \item Combat strike feats allow a character to impose special effects when they strike a foe.
    \item Combat style feats change the way a character fights. A character can only use one style at a time.
\end{itemize*}

\subsection{Bloodline Feats}

Some characters have traces of monstrous blood running in their veins. Most of those will never understand the full potential of their unusual heritage. Bloodline feats allow characters to explore those posibilities by gaining abilites related to their ancestry. Each bloodline feat belongs to a specific type of monster, such as ``dragon''. Some bloodline feats have stronger effects if you have more feats from that heritage.

\subsection{Class Feats}
Class feats improve a character's class features.

\subsection{Item Creation Feats}
An item creation feat lets a spellcaster create magic items. Item creation is described in Rise Advanced.

\subsection{Metamagic Feats}
As a spellcaster's knowledge of magic grows, she can learn to cast spells in ways slightly different from the ways in which the spells were originally designed or learned. Preparing and casting a spell in such a way is harder than normal but, thanks to metamagic feats, at least it is possible. Spells modified by a metamagic feat use a spell slot higher than normal. All effects dependent on spell level (such as the ability to penetrate a \spell{lesser globe of invulnerability}) are calculated according to the spell's modified level.

\parhead{Applying Metamagic Feats} Spellcasters apply metamagic feats on the spot. Therefore, most spellcasters must also take more time to cast a metamagic spell (one enhanced by a metamagic feat) than they do to cast a regular spell. If the spell's normal casting time is a standard action, casting a metamagic version is a full-round action. (This isn't the same as a 1-round casting time.) For a spell with a longer casting time, it takes an extra full-round action to cast the spell. For a spell with a shorter casting time, it takes a standard action to cast the spell.

\par Sorcerers have such an intuitive grasp of magic that they do not need to take extra time to cast spells affected by metamagic feats.

\parhead{Effects of Metamagic Feats on a Spell} In all ways, a metamagic spell operates at its original spell level, even though it is prepared and cast as a higher-level spell.

The modifications made by these feats only apply to spells cast directly by the feat user. A spellcaster can't use a metamagic feat to alter spell-like abilities, spells cast from magic items, or other external sources of magic.

Metamagic feats that eliminate components of a spell don't eliminate the attack of opportunity provoked by casting a spell while threatened. However, casting a spell modified by Quicken Spell does not provoke an attack of opportunity (see \pcref{Quicken Spell}).

Not all metamagic feats cannot be used with all spells. See the specific feat descriptions for the spells that a particular feat can't modify.

\parhead{Multiple Metamagic Feats on a Spell} A spellcaster can apply multiple metamagic feats to a single spell. Changes to its level are cumulative. You can't apply the same metamagic feat more than once to a single spell.

\parhead{Magic Items and Metamagic Spells} With the right item creation feat, you can store a metamagic version of a spell in a scroll, potion, or wand. Level limits for potions and wands apply to the spell's higher spell level (after the application of the metamagic feat). A character doesn't need the metamagic feat to activate an item storing a metamagic version of a spell.

\parhead{Counterspelling Metamagic Spells} Whether or not a spell has been enhanced by a metamagic feat does not affect its vulnerability to counterspelling or its ability to counterspell another spell.

\subsection{Performance Feats}
Performance feats allow a character to use the Perform skill to create magical effects. All Performance feats are also Skill feats.

Each performance has an effect when it is used. Some performances can also be sustained. While you are sustaining a performance, you take a \minus5 penalty to other Perform checks you make. You also cannot cast spells, activate magic items by spell completion (such as scrolls), or activate magic items by magic word (such as wands) while sustaining a performance. You can sustain a performance for a number of minutes equal to 5 \add your Constitution before you must stop performing.

To be affected by a performance feat, a creature must be able to either see or hear the creature giving the performance, depending on whether the performance is visual or auditory. The Perform skill being used determines whether the performance is visual or auditory.

You can use any combination of performance feats you possess a number of times per day equal to the number of performance feats you have \add half your Charisma. If you fail the Perform check, the use of the ability is wasted.

\subsection{Racial feats}
Racial feats improve a character's racial abilities or grant new abilities unique to the character's race.

\subsection{Skill Feats}
Skill feats always affect a character's ability to use skills. Rogues can gain skill feats with their skill trick class feature.

\subsection{Spellgift Feats}
Spellgift feats grant limited spell-like abilities. They may represent the fruits of eldritch pacts, unusual heritage, the exposure to powerful magic, or a particular expression of innate magical ability. You can use any combination of spell-like abilities gained from spellgift feats a number of times per day equal to the number of spellgift feats you possess. The number of spellgift feats you have cannot exceed your Charisma.

If your Charisma is lowered below the number of spellgift feats you have, you are treated as not possessing the feats you do not qualify for, as normal for feat prerequisites. You lose access to the spellgift feats in order by when you acquired them, starting with the most recent feat.

%Design guidelines: effective CL = 1/2 HD, min ability score = effective CL-1
%Uses/day: 3/day for single use abilities, normal duration for buffs (splitting
%duration is already good?)
%Personal range penalizes spell level, but shouldn't make these cheaper.
%Treat personal range spells as one higher level.
\subsection{Surge Feats}
Surge feats grant a character to temporarily increase their abilities to supernatural levels. Each surge feat grants a supernatural ability which can be used a certain number of times per day.

\subsection{Combat Feats}
Combat feats affect a character's combat abilities. Many combat feats require a minimum base attack bonus to acquire.

\subsection{Combat Maneuver Feats}
Combat maneuver feats are combat feats which specifically deal with maneuvers -- pushing opponents around, grappling them, tripping them, or other unusual attacks. Any feat designated as a combat maneuver feat is also considered a combat feat.

\subsection{Combat Style Feats}
Combat style feats grant a character the ability to fight in a particular style, granting them bonuses while fighting in that style. A character can only fight in one style at once. Initiating a style or changing to a different style is a swift action, but a style can be stopped as a free action. Any feat designated as a combat style feat is also considered a combat feat.

\onecolumn
\section{Feat Table}

{\small
\taburowcolors [2] 2{tbrown .. white}
\begin{longtabu}{>{\lcol}p{10em} >{\lcol}p{15em} >{\lcol}X >{\lcol}p{3em}}
%    \taburowcolors{1}{white}{tbrown}
    \thead{General Feats} & \thead{Prerequisites} & \thead{Benefit} & \thead{Page} \\
\featref{Endurance} & Con 3 & Fight normally without hit points & \featpref{Endurance} \\
\featref{Deathless} & Con 5 or base Fortitude defense \plus10 & Immune to death effects & \featpref{Deathless} \\
\featref{Diehard} & Con 3 & Remain conscious after taking critical damage & \featpref{Diehard} \\
\featref{Fearless} & Cha 5 or base Will defense \plus10 & Immune to fear and hostile morale effects & \featpref{Fearless} \\
\featref{Great Fortitude} & \x &  \plus2 bonus to Fortitude defense & \featpref{Great Fortitude} \\
\featref{Iron Will} & \x &  \plus2 bonus to Will defense & \featpref{Iron Will} \\
\featref{Legendary Charisma} & 11th level, Cha 5 & \plus5 to Charisma for some purposes & \featpref{Legendary Charisma} \\
\featref{Legendary Constitution} & 11th level, Con 5 & \plus5 to Constitution for some purposes & \featpref{Legendary Constitution} \\
\featref{Legendary Dexterity} & 11th level, Dex 5 & \plus5 to Dexterity for some purposes & \featpref{Legendary Dexterity} \\
\featref{Legendary Intelligence} & 11th level, Int 5 & \plus5 to Intelligence for some purposes & \featpref{Legendary Intelligence} \\
\featref{Legendary Strength} & 11th level, Str 5 & \plus5 to Strength for some purposes & \featpref{Legendary Strength} \\
\featref{Legendary Wisdom} & 11th level, Wis 5 & \plus5 to Wisdom for some purposes & \featpref{Legendary Wisdom} \\
\featref{Lightning Reflexes} & \x &  \plus2 bonus to Reflex defense & \featpref{Lightning Reflexes} \\
\featref{Mental Fortress} & Cha 9 or base Will defense \plus18 & Immune to hostile mind-affecting effects & \featpref{Mental Fortress} \\
\featref{Perfect Health} & Con 3 or base Fortitude defense \plus6 & Immune to disease, later poison & \featpref{Perfect Health} \\
\featref{Swift} & \x & Increase speed by 5 feet & \featpref{Swift} \\
\featref{Toughness} & \x &  \plus3 hit points \plus1 per level above 3 & \featpref{Toughness} \\

\midrule
\thead{Bloodline Feats} & \thead{Prerequisites} & \thead{Benefit} & \thead{Page} \\
\featref{Celestial Heritage} & Nonevil alignment & Smite evil 1/day & \featpref{Celestial Heritage} \\
\tind \featref{Celestial Body} & Nonevil alignment, Celestial Heritage & Gain physical damage reduction & \featpref{Celestial Body} \\
\tind \featref{Celestial Smiting} & Nonevil alignment, Celestial Heritage & Smite evil more often, more accurately & \featpref{Celestial Smiting} \\
\tind \featref{Celestial Soul} & Nonevil alignment, any three celestial feats & Gain spell resistance against evil & \featpref{Celestial Soul} \\
\featref{Draconic Heritage} & \x & Resist damage from chosen dragon's energy type & \featpref{Draconic Heritage} \\
\tind \featref{Draconic Breath} & Con 3, any three dragon bloodline feats & Gain dragon's breath weapon & \featpref{Draconic Breath} \\
\tind \featref{Draconic Might} & Any three dragon bloodline feats & Gain \plus1 to physical attribute & \featpref{Draconic Might} \\
\tind \featref{Draconic Mind} & Any three dragon bloodline feats & Gain \plus1 to mental attribute & \featpref{Draconic Mind} \\
\tind \featref{Draconic Scales} & Draconic Heritage & Gain natural armor & \featpref{Draconic Scales} \\
\tind \featref{Draconic Senses} & Draconic Heritage & Gain low-light vision, possibly darkvision & \featpref{Draconic Senses} \\
\tind \featref{Draconic Spellpower} & Draconic Heritage & Gain bonus to caster level & \featpref{Draconic Spellpower} \\
\tind \featref{Draconic Voice} & Draconic Heritage & Gain bonus to Intimidate and Persuasion checks & \featpref{Draconic Voice} \\
\tind \featref{Draconic Weapons} & Draconic Heritage & Gain bite attack, possibly claws & \featpref{Draconic Weapons} \\
\tind \featref{Draconic Wings} & Any three dragon bloodline feats & Gain wings to slow falls, glide, eventually fly & \featpref{Draconic Wings} \\
\featref{Elemental Heritage} & \x & Gain defense bonus based on elemental ancestor & \featpref{Elemental Heritage} \\
\tind \featref{Elemental Body} & Any three elemental bloodline feats & 50\% chance to ignore critical hits & \featpref{Elemental Body} \\
\tind \featref{Elemental Force} & Elemental heritage & Unleash element to attack & \featpref{Elemental Force} \\
\tind \featref{Elemental Mastery} & Elemental Heritage & Gain attack bonus in circumstances based on elemental ancestor & \featpref{Elemental Mastery} \\
\tind \featref{Elemental Movement} & Elemental Heritage & Gain movement ability based on elemental ancestor & \featpref{Elemental Movement} \\

\midrule
\thead{Class Feats} & \thead{Prerequisites} & \thead{Benefit} & \thead{Page} \\
\featref{Extra Channeling} & Ability to channel energy & Can channel energy 3 more times per day & \featpref{Extra Channeling} \\
\featref{Extra Invocation} & Ability to use an arcane invocation & Learn a new arcane invocation & \featpref{Extra Invocation} \\
\featref{Extra Rage} & Rage ability & Can rage 1 more time per day & \featpref{Extra Rage} \\
\featref{Extra Smiting}\fn{1} & Smite ability & Can smite 3 more times per day & \featpref{Extra Smiting} \\
\featref{Extra Wild Aspect} & Wild aspect ability & Can use wild aspects 3 more times per day & \featpref{Extra Wild Aspect} \\
\featref{Intense Channeling} & Channel energy 3d6 & \plus1d6 damage or healing with channel energy & \featpref{Intense Channeling} \\
\featref{Mental Discipline} & Combat discipline ability & Use Charisma for combat discipline & \featpref{Mental Discipline} \\
\featref{Selective Channeling} & Channel energy ability & Can exclude two additional creatures & \featpref{Selective Channeling} \\
\featref{Shield Discipline} & Armor discipline ability & Apply armor discipline with shields \\
\featref{Trap Sense} & Danger sense class feature & Automatically detect traps & \featpref{Trap Sense} \\

\midrule
\thead{Item Creation Feats} & \thead{Prerequisites} & \thead{Benefit} & \thead{Page} \\
\featref{Imbue Magic} & Caster level 2nd or Craft (any) 5 ranks & Create magic items & \featpref{Imbue Magic} \\
\featref{Imbuement Admixture} & 2nd level spells, 6th & Combine multiple spells to create items & \featpref{Imbuement Admixture} \\
\featref{Versatile Crafter} & Craft (any) 8 ranks & Craft items from many subschools of magic & \featpref{Versatile Crafter} \\

\midrule
\thead{Magic Feats} & \thead{Prerequisites} & \thead{Benefit} & \thead{Page} \\
\featref{Bardic Spellpower} & Arcane spellcasting ability & Fighter and rogue levels improve some spells & \featpref{Bardic Spellpower} \\
\featref{Battlecaster} & Proficiency with light armor & Reduce arcane spell failure by 10\% & \featpref{Battlecaster} \\
\featref{Combat Casting} & \x &  \plus2 bonus to Concentration checks to cast spells, reroll 1/day & \featpref{Combat Casting} \\
\featref{Craftcaster} & Craft (any) 8 ranks, 2nd level spells & Increase caster level when creating, transforming objects & \featpref{Craftcaster} \\
\featref{Improved Counterspell} & \x &  Counterspell with spell of same school & \featpref{Improved Counterspell} \\
\featref{Legendary Magic} & 5th level spells & Weak creatures are always considered bloodied & \featpref{Legendary Magic} \\
\featref{Magical Synthesis} & Levels in two magical classes & Increase spellcasting ability in two magical classes at once & \featpref{Magical Synthesis} \\
\featref{Mass Transporter} & 5th level Conjuration (Translocation) spell & Transport many creatures at once & \featpref{Mass Transporter} \\
\featref{Mighty Summons} & 2nd level summoning spell & Summoned creatures gain \plus2 Str, \plus2 Con & \featpref{Mighty Summons} \\
\featref{Rapid Metamagic} & Spellcraft 8 ranks, spellcasting ability, one metamagic feat & Apply metamagic effects more quickly & \featpref{Rapid Metamagic} \\
\featref{Residual Awareness} & Divination (Awareness) spell & Gain brief Perception bonus after casting a spell & \featpref{Residual Awareness} \\
\featref{Residual Knowledge} & Divination (Knowledge) spell & Gain brief Knowledge bonus after casting a spell & \featpref{Residual Knowledge} \\
\featref{Residual Beguilement} & 3rd level Enchantment (Beguilement) spell & Gain Persuasion bonus after casting a spell & \featpref{Residual Beguilement} \\
\featref{Residual Shield} & Abjuration (Shielding) spell & Gain benefits of spell briefly after casting on another creature & \featpref{Residual Shield} \\
\featref{Resilient Magic} & 2nd level Abjuration (Negation) spell & Spells persist for 1 round after being dispelled & \featpref{Resilient Magic} \\
\featref{Retributive Counterspell} &  4th level spells & Countered spells rebound on original caster & \featpref{Retributive Counterspell} \\
\featref{Ritual Caster} & Int 3 & Gain ability to perform rituals & \featpref{Ritual Caster} \\
\featref{Spell Focus}\fn{1} & 2nd level spells &  \plus2 caster level with specific type of magic & \featpref{Spell Focus} \\
\tind \featref{Spell Specialization} & 4th level spells, Spell Focus &  \plus2 caster level with specific type of magic, \minus2 penalty with other types. & \featpref{Spell Specialization} \\
\featref{Wardwalker} & 3rd level Abjuration (Warding) spell & Wards you encounter trigger one round late & \featpref{Wardwalker} \\

\midrule
\thead{Metamagic Feats} & \thead{Prerequisites} & \thead{Benefit (Level Adjustment)} & \thead{Page} \\
\featref{Empower Spell} & 2nd level spells & Cast spell as higher level & \featpref{Empower Spell} \\
\featref{Energetic Substitution} & 2nd level spells & Change spell's energy type (\plus1) & \featpref{Energetic Substitution} \\
\featref{Enlarge Spell} & 2nd level spells & Double spell's range (\plus1) & \featpref{Enlarge Spell} \\
\featref{Imbued Spellstrike} & 2nd level spells & Combine spell with weapon attack (\plus1) & \featpref{Imbued Spellstrike} \\
\tind \featref{Improved Imbued Spellstrike} & 3rd level spells, Imbued Spellstrike & Imbue weapon with spell's power up to 5 minutes (\plus2) & \featpref{Improved Imbued Spellstrike} \\
\featref{Quicken Spell} & 3rd level spells & Cast spell as swift action, but lose next action (\plus2) & \featpref{Quicken Spell} \\
\featref{Reach Spell} & 3rd level spells & Cast touch spell at Close range (\plus2) & \featpref{Reach Spell} \\
\featref{Shape Spell} & 3rd level spells & Exclude areas within spell's area (\plus2) & \featpref{Shape Spell} \\
\featref{Silent Spell} & 2nd level spells & Cast spell without verbal components (\plus1) & \featpref{Silent Spell} \\
\featref{Still Spell} & 2nd level spells & Cast spell without somatic components (\plus1) & \featpref{Still Spell} \\
\featref{Sustained Spell} & 2nd level spells & Maintain concentration as swift action (\plus1) & \featpref{Sustained Spell} \\
\featref{Widen Spell} & 4th level spells & Double spell's area (\plus3) & \featpref{Widen Spell} \\

\midrule
\thead{Performance Feats} & \thead{Prerequisites} & \thead{Benefit} & \thead{Page} \\
\featref{Dual Performance} & Perform 14 ranks, any three Performance feats & Sustain two performances at once & \featpref{Dual Performance} \\
\featref{Fascinating Performance} & Perform 4 ranks & Fascinate group of creatures & \featpref{Fascinating Performance} \\
\featref{Freeing Performance} & Perform 12 ranks  & Dispel magical effects on ally & \featpref{Freeing Performance} \\
\featref{Inspire Competence} & Perform 4 ranks  & Grant allies bonuses to skill checks & \featpref{Inspire Competence} \\
\featref{Inspire Courage} & Perform 6 ranks  & Grant allies bonuses to physical attacks and defenses against fear & \featpref{Inspire Courage} \\
\featref{Inspire Greatness} & Perform 12 ranks  & Grant allies bonuses to defenses & \featpref{Inspire Greatness} \\
\featref{Inspire Heroics} & Perform 18 ranks  & Grant allies temporary hit points and extra attack & \featpref{Inspire Heroics} \\
\featref{Inspire Spellpower} & Perform 8 ranks  & Grant allies bonus to caster level & \featpref{Inspire Spellpower} \\
\featref{Rapid Performance} & Perform 14 ranks, any three Performance feats & Perform as a swift action & \featpref{Rapid Performance} \\
\featref{Recover Performance} & Perform 8 ranks, any other Performance feat & Reroll failed performance attempt & \featpref{Recover Performance} \\
\featref{Spellwoven Performance} & Perform 8 ranks, any other Performance feat & Cast some spells while performing & \featpref{Spellwoven Performance} \\
\tind \featref{Suggestive Performance} & Perform 10 ranks, Fascinating Performance & Suggest action to fascinated creatures & \featpref{Suggestive Performance} \\

\midrule
\thead{Racial Feats} & \thead{Prerequisites} & \thead{Benefit} & \thead{Page} \\
\featref{Dwarven Resilience} & Dwarf & \plus2 to defenses against spells & \featpref{Dwarven Resilience} \\
\featref{Focused Mind} & Elf & Use Intelligence to concentrate instead of Constitution & \featpref{Focused Mind} \\
\featref{Giantfighter} & Dwarf, gnome, or halfling & \plus2 to dodge against Large or larger creatures & \featpref{Giantfighter} \\
\featref{Stonecunning} & Dwarf & Gain a sixth sense about stonework & \featpref{Stonecunning} \\

\midrule
\thead{Skill Feats} & \thead{Prerequisites} & \thead{Benefit} & \thead{Page} \\
\featref{Dilettante} & Int 3 & Use some Knowledge skills despite being untrained & \featpref{Dilettante} \\
\featref{Ledgewalker} & Acrobatics 5 ranks & Balance easily on narrow surfaces & \featpref{Ledgewalker} \\
\featref{Legendary Balance} & Acrobatics 13 ranks & Balance on impossible surfaces & \featpref{Legendary Balance} \\
\featref{Legendary Climber} & Climb 13 ranks & Gain climb speed, climb on impossible surfaces & \featpref{Legendary Climber} \\
\featref{Legendary Craftsman} & Craft 13 ranks & Craft items with fewer material components & \featpref{Legendary Craftsman} \\
\featref{Legendary Devicesmith} & Devices 13 ranks & Disable active spell effects & \featpref{Legendary Devicesmith} \\
\featref{Legendary Disguise} & Disguise 13 ranks & Alter magical auras with disguise & \featpref{Legendary Disguise} \\
\featref{Legendary Escapist} & Escape Artist 13 ranks & Escape from magical effects & \featpref{Legendary Escapist} \\
\featref{Legendary Leaper} & Athletics 13 ranks & Make standing jumps without penalty, rebound off objects easily & \featpref{Legendary Leaper} \\
\featref{Legendary Liar} & Bluff 13 ranks & Lies become undetectable by magic & \featpref{Legendary Liar} \\
\featref{Legendary Tumbler} & Acrobatics 13 ranks & Tumble through enemies without provoking & \featpref{Legendary Tumbler} \\
\featref{Open Minded} & \x & Gain two skill points. & \featpref{Open Minded} \\
\featref{Poison Expert} & Int 3, Craft (poison) 8 ranks & Apply poisons more effectively & \featpref{Poison Expert} \\
\featref{Ranged Legerdemain} &  2nd level spells & Use Disable Device or Sleight of Hand at range & \featpref{Ranged Legerdemain} \\
\featref{Scale the Beast} & Climb 8 ranks & Climb on massive creatures & \featpref{Scale the Beast} \\
\featref{Skill Focus}\fn{1} & \x &  \plus3 bonus to checks with selected skill, reroll 1/day & \featpref{Skill Focus} \\
\featref{Skill Mastery}\fn{1} & Any skill 10 ranks & Always take 10 on skill checks & \featpref{Skill Mastery} \\
\featref{Skill Training}\fn{1} & \x & Gain two skills as class skills & \featpref{Skill Training} \\
\featref{Trapfinder} & Perception 5 ranks & Find traps faster & \featpref{Trapfinder} \\
\featref{Track} & \x &  Use Survival skill to track & \featpref{Track} \\
\featref{Veteran's Experience} & Base attack bonus \plus8 & Identify monsters without Knowledge & \featpref{Veteran's Experience} \\

\midrule
\thead{Spellgift Feats} & \thead{Prerequisites} & \thead{Benefit} & \thead{Page} \\
\featref{Arcane Spellgift} & 2nd level, Cha 1 & Gain arcane spell as spell-like ability & \featpref{Arcane Spellgift} \\
\featref{Divine Spellgift} & 2nd level, Cha 1 & Gain divine spell as spell-like ability & \featpref{Divine Spellgift} \\
\featref{Natural Spellgift} & 2nd level, Cha 1 & Gain natural spell as spell-like ability & \featpref{Natural Spellgift} \\
\featref{Potent Spellgift} & Any spellgift feat & \plus2 caster level with spellgifts & \featpref{Potent Spellgift} \\

\midrule
\thead{Surge Feats} & \thead{Prerequisites} & \thead{Benefit} & \thead{Page} \\
\featref{Alacritous Surge} & 16th level, Dex 7 & Briefly gain effects of \spell{haste} & \featpref{Alacritous Surge} \\
\featref{Deathless Surge} & 12th level, Cha 5 & Briefly gain effects of \spell{death ward} & \featpref{Deathless Surge} \\
\featref{Freeing Surge} & 16th level, Dex 7 & Briefly gain effects of \spell{freedom} & \featpref{Freeing Surge} \\
\featref{Healing Surge} & 8th level, Con 3 & Heal damage three times per day & \featpref{Healing Surge} \\
\featref{Prescient Surge} & 12th level, Wis 5 & Roll twice, three times per day & \featpref{Prescient Surge} \\
\featref{Restorative Surge} & 8th level, Con 3 & Heal ability damage three times per day & \featpref{Restorative Surge} \\
\featref{Shattering Surge} & 8th level, Str 3 & \spell{Shatter} struck object three times per day & \featpref{Shattering Surge} \\
\featref{Surge of Speed} & 8th level, Dex 3 & Briefly gain effects of \spell{expeditious retreat} & \featpref{Surge of Speed} \\
\end{longtabu}}

\subsection{Combat Feat Table}

{\small
\taburowcolors [2] 2{tbrown .. white}
\begin{longtabu}{>{\lcol}p{10em} >{\lcol}p{15em} >{\lcol}X >{\lcol}p{6em} >{\lcol}p{3em}}
    \thead{Combat Feats} & \thead{Prerequisites} & \thead{Benefit} & \thead{Feat Type} & \thead{Page} \\
\featref{Armor Familiarity} & Proficiency with armor & Reduce penalties from wearing armor & Equipment & \featpref{Armor Familiarity} \\
\featref{Armor Proficiency (Light)} & \x &  No armor check penalty on physical attacks & Equipment & \featpref{Armor Proficiency (Light)} \\
\tind \featref{Armor Proficiency (Medium)} & Armor Proficiency (light) & No armor check penalty on physical attacks & Equipment & \featpref{Armor Proficiency (Medium)} \\
\tind \tind \featref{Armor Proficiency (Heavy)} & Armor Proficiency (medium) & No armor check penalty on physical attacks & Equipment & \featpref{Armor Proficiency (Heavy)} \\
\featref{Blind-Fight} & \x &  Reroll miss chance for concealment in melee & Awareness & \featpref{Blind-Fight} \\
\featref{Cautious Attack} & \x & Trade damage for maneuver defense & Defense, Style & \featpref{Cautious Attack} \\
\featref{Covering Fire} & \x & Trade ranged attack bonus to penalize struck foes & Precision, Style & \featpref{Covering Fire} \\
\featref{Deadly Aim} & Dex 3 & Trade ranged attack bonus for damage & Precision, Style & \featpref{Deadly Aim} \\
\featref{Defensive Fighting} & \x & Trade attack bonus for defense & Defense, Style & \featpref{Defensive Fighting} \\
\featref{Dodge} & Dex 3 & \plus4 defense against some attacks of opportunity from selected target & Defense, Mobility & \featpref{Dodge} \\
\featref{Exotic Weapon Proficiency}\fn{1} & \x & Don't provoke when attacking with exotic weapons & Equipment & \featpref{Exotic Weapon Proficiency} \\
\featref{Far Shot} & \x & Increase range increment by 50\% or 100\% & Precision & \featpref{Far Shot} \\
\featref{Guardian} & \x & Adjacent allies suffer reduced overwhelm penalties. & Defense & \featpref{Guardian} \\
\featref{Improved Initiative} & \x & \plus2 bonus to initiative checks, reroll 1/day & Reaction & \featpref{Improved Initiative} \\
\featref{Improved Unarmed Strike} & \x &  Considered armed even when unarmed & \x & \featpref{Improved Unarmed Strike} \\
\tind \featref{Deflect Arrows} & Dex 3, Improved Unarmed Strike & Deflect one ranged attacks with your bare hands & Finesse, Reaction & \featpref{Deflect Arrows} \\
\tind \tind \featref{Snatch Arrows} & Dex 5, Deflect Arrows, Improved Unarmed Strike & Catch a deflected ranged attack & Finesse, Reaction & \featpref{Snatch Arrows} \\
\featref{Mounted Archery} & Ride 1 rank & Reduced penalty for ranged attacks while mounted by 4 & Mounted, Precision & \featpref{Mounted Archery} \\
\featref{Mounted Combat} & Ride 1 rank & Negate hits on mount with Ride check & Defense, Mounted & \featpref{Mounted Combat} \\
\featref{Mounted Trample} & Ride 8 ranks & Target cannot avoid mounted overrun & Mobility, Mounted & \featpref{Trample} \\
\featref{Opportunist} & Dex 3 & \plus2 to attack and damage on attacks of opportunity & Finesse, Style & \featpref{Opportunist} \\
\featref{Overpowering Assault} & Str 3 & Trade defense for maneuver bonus & Power, Style & \featpref{Overpowering Assault} \\
\featref{Parry} & Dex 3 & Ready yourself to parry incoming blows & Defense, Reaction & \featpref{Parry} \\
\tind \featref{Shielded Parry} & Dex 3, shield proficiency, Parry & Add shield modifier to parry attempts & Defense, Reaction & \featpref{Shielded Parry} \\
\featref{Point Blank Shot} & \x &  \plus2 bonus to ranged damage within half range increment. & Precision, Style & \featpref{Point Blank Shot} \\
\featref{Precise Shot} & \x & Add half Wis to damage & Precision, Style & \featpref{Precise Shot} \\
\featref{Power Attack} & Str 3 & Trade melee attack bonus for damage & Power, Style & \featpref{Power Attack} \\
\featref{Quick Draw} & \x & Draw weapon as swift action & Reaction & \featpref{Quick Draw} \\
\featref{Ride-By Attack} & Ride 8 ranks & Move before and after a mounted charge & Mobility, Mounted & \featpref{Ride-By Attack} \\
\tind \featref{Spirited Charge} & Ride-By Attack & Double damage with mounted charge & Mounted, Power & \featpref{Spirited Charge} \\
\featref{Shield Proficiency} & \x &  No armor check penalty on physical attacks & Equipment & \featpref{Shield Proficiency} \\
\tind \featref{Tower Shield Proficiency} & Shield Proficiency & No armor check penalty on physical attacks & Equipment & \featpref{Tower Shield Proficiency} \\
\featref{Two-Weapon Fighting} & Dex 3 & Gain \plus2 attack bonus when fighting with two weapons & Finesse & \featpref{Two-Weapon Fighting} \\
\tind \featref{Two-Weapon Defense} & Two-Weapon Fighting & Off-hand weapon grants \plus1 shield defense bonus, later \plus3 & Defense, Finesse & \featpref{Two-Weapon Defense} \\
\featref{Weapon Proficiency}\fn{1} & \x &  Don't provoke when attacking with weapon group & Equipment & \featpref{Weapon Proficiency} \\

\midrule
\thead{Combat Feats} & \thead{Prerequisites} & \thead{Benefit} & \thead{Feat Type} & \thead{Page} \\
\featref{Cleave} & Str 3, base attack bonus \plus4 & Extra melee attack after dropping target & Power & \featpref{Cleave} \\
\featref{Cleaving Smite} & Smite ability, base attack bonus \plus4, Cleave & Extra attack from Cleave is also a smite & Power \\
\featref{Combat Mobility} & Dex 3, base attack bonus \plus4, Dodge & Avoid some attacks of opportunity from selected target & Defense, Mobility & \featpref{Combat Mobility} \\
\tind \featref{Spring Attack} & Dex 3, base attack bonus \plus4, Dodge, Combat Mobility & Move before and after attacks & Mobility, Style & \featpref{Spring Attack} \\
\featref{Distracting Foe} & Base attack bonus \plus4 & Threatened foes suffer Concentration penalties & Style & \featpref{Distracting Foe} \\
\featref{Feign Weakness} & Base attack bonus \plus4 & Provoke attack to render foe flat-footed & Style & \featpref{Feign Weakness} \\
\featref{Improved Dirty Trick} & Base attack bonus \plus4 & \plus2 bonus to dirty trick attacks; dirty tricks last longer & Finesse, Maneuver & \featpref{Improved Dirty Trick} \\
\featref{Improved Disarm} & Base attack bonus \plus4 & \plus2 bonus to disarm attacks; knock foe's weapon away after disarming & Finesse, Maneuver & \featpref{Improved Disarm} \\
\featref{Improved Feint} & Base attack bonus \plus4 & \plus2 bonus to feint attacks; feint lasts longer & Finesse, Maneuver & \featpref{Improved Feint} \\
\featref{Improved Grapple} & Base attack bonus \plus4 & \plus2 bonus to grapple attacks; not flat-footed while grappling & Maneuver, Power & \featpref{Improved Grapple} \\
\featref{Improved Overrun} & Base attack bonus \plus4 & \plus2 bonus to overrun attacks; target can't avoid & Maneuver, Power & \featpref{Improved Overrun} \\
\featref{Improved Shove} & Base attack bonus \plus4 & \plus2 bonus to shove attacks; move target without moving & Maneuver, Power & \featpref{Improved Shove} \\
\featref{Improved Trip} & Base attack bonus \plus4 & \plus2 bonus to trip attacks; tripped foe provokes attacks of opportunity & Finesse, Maneuver & \featpref{Improved Trip} \\
\featref{Intuitive Reaction} & Wis 3, base attack bonus \plus4 & Add half Wisdom to number of attacks of opportunity each round & Reaction & \featpref{Intuitive Reaction} \\
\featref{Knockdown} & Str 3, base attack bonus \plus4 & Shoved foe can be knocked prone & Maneuver, Power & \featpref{Knockdown} \\
\featref{Reveal the Weak Point} & Base attack bonus \plus4 & Sacrifice attack and damage to penalize foe's defenses & Style & \featpref{Reveal the Weak Point} \\
\featref{Riposte} & Dex 3, base attack bonus \plus4, Parry & Foes provoke if you parry very well & Reaction & \featpref{Riposte} \\
\featref{Tactical Analysis} & Int 3, Base attack bonus \plus4 & Identify foe's strengths and weaknesses & Awareness & \featpref{Tactical Analysis} \\
\featref{Threatening Fire} & Base attack bonus \plus4 & Help overwhelm foes with ranged weapons & Precision, Style & \featpref{Threatening Fire} \\
\featref{Unshakeable Stance} & Base attack bonus \plus4 & Trade ability to move for maneuver defense & Defense, Style & \featpref{Unshakeable Stance} \\
\featref{Weapon Focus} & Base attack bonus \plus4 & Special ability with weapon group & Equipment & \featpref{Weapon Focus} \\

\midrule
\thead{Combat Feats} & \thead{Prerequisites} & \thead{Benefit} & \thead{Feat Type} & \thead{Page} \\
\featref{Bewildering Strike} & Base attack bonus \plus8 & Attack bewilders foe & Strike & \featpref{Bewildering Strike} \\
\featref{Cleaving Stride} & Str 3, base attack bonus \plus8, Cleave & Continue movement after dropping target & Mobility, Power & \featpref{Cleaving Stride} \\
\featref{Contingent Counter} & Int 3, base attack bonus \plus8 & Prepare to attack when attack misses you & Reaction & \featpref{Contingent Counter} \\
\featref{Driving Assault} & Str 5, base attack bonus \plus8 & Shoved foe provokes attack from you & Power, Maneuver & \featpref{Driving Assault} \\
\featref{Eye of the Storm} & Base attack bonus \plus8 & You are more difficult to overwhelm & Awareness, Defense & \featpref{Eye of the Storm} \\
\featref{Heartseeker} & Base attack bonus \plus8 & Trade attack bonus to double critical threat range & Precision, Style & \featpref{Heartseeker} \\
\featref{Immobilizing Strike} & Base attack bonus \plus8 & Melee attack prevents foe from moving & Strike & \featpref{Immobilizing Strike} \\
\featref{Inescapable Bulwark of Defense} & Dex 3, Base attack bonus \plus8 & Withdrawing enemies cannot avoid provoking attacks of opportunity & Reaction, Style & \featpref{Inescapable Bulwark of Defense} \\
\featref{Intimidating Strike} & Base attack bonus \plus8, Intimidate 8 ranks & Attack lets you intimidate struck foe & Strike & \featpref{Intimidating Strike} \\
\featref{Intuitive Defense} & Wis 3, base attack bonus \plus8 & Add half Wisdom to defense & Awareness, Defense, Style & \featpref{Intuitive Defense} \\
\featref{Overwhelming Force} & Str 5, base attack bonus \plus8 & Apply full Strength to damage when using two hands & Power & \featpref{Overwhelming Force} \\
\featref{Perfect Shot} & Dex 5, base attack bonus \plus8 & Ignore less than total cover/concealment on ranged attacks & Precision, Style & \featpref{Perfect Shot} \\
\featref{Pierce Wings} & Base attack bonus \plus8 & Ranged attack removes foe's ability to fly & Precision, Strike & \featpref{Pierce Wings} \\
\featref{Predictive Defense} & Int 3, base attack bonus \plus8 & Add half Int to defense & Awareness, Defense, Style & \featpref{Predictive Defense} \\
\featref{Sickening Strike} & Base attack bonus \plus8 & Attack sickens foe & Strike & \featpref{Sickening Strike} \\
\featref{Tactical Prediction} & Int 3, Base attack bonus \plus8 & Predict foe's next action & Awareness & \featpref{Tactical Prediction} \\
\featref{Wall Slam} & Str 5, base attack bonus \plus8 & Slam opponent into wall to deal bonus damage & Maneuver, Power & \featpref{Wall Slam} \\
\featref{Whirlwind Attack} & Dex 5, base attack bonus \plus8 & One melee attack against each opponent within reach & Style & \featpref{Whirlwind Attack} \\

\midrule
\thead{Combat Feats} & \thead{Prerequisites} & \thead{Benefit} & \thead{Feat Type} & \thead{Page} \\
\featref{Confusing Strike} & Base attack bonus \plus12 & Attack confuses bloodied foe & Strike & \featpref{Confusing Strike} \\
\featref{Contingent Attack} & Int 5, base attack bonus \plus12 & Prepare to attack as immediate action & Reaction & \featpref{Contingent Attack} \\
\featref{Legendary Awareness} & Base attack bonus \plus12, any three Awareness feats & Immune to overwhelm & Awareness & \featpref{Legendary Awareness} \\
\featref{Legendary Defense} & Base attack bonus \plus12, any three Defense feats & \plus2 to all defenses & Defense & \featpref{Legendary Defense} \\
\featref{Legendary Finesse} & Base attack bonus \plus12, any three Finesse feats & Add half Dexterity to damage & Finesse & \featpref{Legendary Finesse} \\
\featref{Legendary Maneuver Master} & Base attack bonus \plus12, any three Maneuver feats & Never provoke, deal damage with very successful maneuvers & Maneuver & \featpref{Legendary Maneuver Master} \\
\featref{Legendary Mobility} & Base attack bonus \plus12, any three Mobility feats & Movement does not provoke & Mobility& \featpref{Legendary Mobility} \\
\featref{Legendary Mounted Warrior} & Base attack bonus \plus12, any three Mounted feats & Share damage with mount & Mounted& \featpref{Legendary Mounted Warrior} \\
\featref{Legendary Power} & Base attack bonus \plus12, any three Power feats & Wield weapons as if they were less encumbering & Power& \featpref{Legendary Power} \\
\featref{Legendary Precision} & Base attack bonus \plus12, any three Precision feats & Very accurate hits deal maximum damage & Precision & \featpref{Legendary Precision} \\
\featref{Legendary Reaction} & Retain full defenses when flat-footed & Combat, Reaction & \featpref{Legendary Reaction} \\
\featref{Legendary Style} & Base attack bonus \plus12, any three Style feats & Use two styles at once & \x & \featpref{Legendary Style} \\
\featref{Manyshot} & Dex 7, base attack bonus \plus12 & Shoot two projectiles simultaneously & Precision, Style & \featpref{Manyshot} \\
\featref{Penetrating Strike} & Base attack bonus \plus12 & Attack removes foe's damage reduction & Strike & \featpref{Penetrating Strike} \\
\featref{Two-Weapon Rend} & Str 5, base attack bonus \plus12, Two-Weapon Fighting & Deal Strength damage with both weapons & Power & \featpref{Two-Weapon Rend} \\

\midrule
\thead{Combat Feats} & \thead{Prerequisites} & \thead{Benefit} & \thead{Feat Type} & \thead{Page} \\
\featref{Counterstorm} & Base attack bonus \plus16 & Foes that miss you provoke & Reaction, Style& \featpref{Counterstorm} \\
\featref{Demolishing Strike} & Base attack bonus \plus16 & Attack deals double damage & Strike & \featpref{Demolishing Strike} \\
\featref{Executioner} & Base attack bonus \plus16, any five Reaction feats & Gain free attacks against foes near death & Reaction & \featpref{Executioner} \\
\featref{Killing Blow} & Base attack bonus \plus16 & Attack instantly kills foe without hit points & Strike & \featpref{Killing Blow} \\
\end{longtabu}
\begin{enumerate*}
    \item You can gain this feat multiple times. Each time you do, it has a different effect.
\end{enumerate*}
}%end of small text

\twocolumn

\section{Feat Descriptions}
Here is the format for feat descriptions.

\ssecfake{Feat Name [Type of Feat]}
\parhead{Prerequisite} A minimum attribute score, another feat or feats, a minimum base attack bonus, a minimum number of ranks in one or more skills, or a class level that a character must have in order to acquire this feat. This entry is absent if a feat has no prerequisite. A feat may have more than one prerequisite.
\parhead{Benefit} What the feat enables the character (``you'' in the feat description) to do. If a character has the same feat more than once, its benefits do not stack unless indicated otherwise in the description.
\par In general, characters cannot gain the same feat twice.
\parhead{Normal} What a character who does not have this feat is limited to or restricted from doing. If not having the feat causes no particular drawback, this entry is absent.
\parhead{Special} Additional facts about the feat that may be helpful when you decide whether to acquire the feat.

\feat{Alacritous Surge}{Surge}
\featpre 16th level, Dex 7.
\featben You may gain the benefits of the \spell{haste} spell for up to 5 rounds each day. Activating this ability is a swift action, and it lasts until you will it to stop (a free action) or until the duration expires.

\feat{Analytical Performer}{Skill}
\featpre Perform 4 ranks.
\featben You can treat your Intelligence as your key attribute for the purpose of the Perform skill in place of your Charisma. This allows you to apply your Intelligence to Perform checks in place of your Charisma, and lets you spend Intelligence skill points to gain training in the Perform skill.

\feat{Arcane Spellgift}{Spellgift}
\featpre 2nd level, Cha 1.
\featben You gain the ability to use a non-restricted arcane spell as a spell-like ability once per day. You choose which spell when you gain this feat, and that choice cannot thereafter be changed. Your caster level is equal to your level, and you use your Charisma as your casting attribute.

The spell's level must be no greater than half of your total levels in non-casting classes. In addition, its level cannot exceed your Charisma.
\parhead{Special} You may take this feat multiple times. Each time, you choose a new spell to gain as a spell-like ability. You may use any combination of the spells a number of times per day equal to the number of spellgift feats you have. The number of spellgift feats you have cannot exceed your Charisma.

\feat{Armor Familiarity}{Combat}
Choose one category of armor: light, medium, heavy, or shields.
\parhead{Prerequisite} Proficiency with the chosen armor category.
\parhead{Benefit} You reduce your armor check penalty by 2 and your arcane spell failure by 5\% when using your chosen armor. This effect cannot reduce those penalties below 0.

\feat{Armor Proficiency (Heavy)}{Combat}
\parhead{Prerequisites} Armor Proficiency (light), Armor Proficiency (medium).
\parhead{Benefit} See Armor Proficiency (light).
\parhead{Normal} See Armor Proficiency (light).
\parhead{Special} Fighters, paladins, and clerics automatically have Armor Proficiency (heavy) as a bonus feat. They need not select it.

\feat{Armor Proficiency (Light)}{Combat}
\parhead{Benefit} When you wear a type of armor with which you are proficient, the armor check penalty for that armor applies only to Acrobatics, Climb, Escape Artist, Jump, and Sleight of Hand, and Swim checks. You suffer the armor's normal arcane spell failure.
\parhead{Normal} A character who is wearing armor with which she is not proficient applies its armor check penalty to physical attacks and to all checks that involve moving, including Ride. The character also suffers double the normal arcane spell failure chance for wearing the armor.
\parhead{Special} All characters except wizards, sorcerers, and monks automatically have Armor Proficiency (light) as a bonus feat. They need not select it.

\feat{Armor Proficiency (Medium)}{Combat}
\parhead{Prerequisite} Armor Proficiency (light).
\parhead{Benefit} See Armor Proficiency (light).
\parhead{Normal} See Armor Proficiency (light).
\parhead{Special} Fighters, barbarians, paladins, clerics, druids, and spellwarped automatically have Armor Proficiency (medium) as a bonus feat. They need not select it.

\feat{Bardic Spellpower}{Magic}
\featpre Ability to cast arcane spells.
\featben You can treat your fighter and rogue levels as magical classes when determining your caster level with enchantment and illusion arcane spells.

\feat{Battlecaster}{Magic}
\featpre Proficiency with light armor.
\featben You reduce your chance of arcane spell failure from wearing armor by 10\%.

\feat{Bewildering Strike}{Combat, Strike}
\featpre Base attack bonus \plus8.
\featben When you make a physical attack, you may take a \minus4 penalty to hit. If you deal damage, you make a Will attack against the struck foe to bewilder it for 5 rounds. Your attack bonus is equal to your base attack bonus \add your physical attack attribute (Strength or Dexterity).  A bewildered creature is vulnerable, causing it to take a \minus2 penalty to attacks, defenses, and checks. You cannot make more than one Strike each round.

\feat{Blind-Fight}{Combat}
\parhead{Benefit} In melee, every time you miss because of being unable to see your opponent, you can reroll your miss chance one time to see if you actually hit. In addition, you are not flat-footed against invisible attackers adjacent to you.
\par If you have 10 ranks in Perception, you can automatically pinpoint the location of any invisible creature adjacent to you.
\parhead{Normal} You have a 50\% chance to miss opponents you can't see, and you are flat-footed against them.

\feat{Cautious Attack}{Combat, Style}
\parhead{Benefit} While in this style, you gain a \plus2 bonus to your Maneuver Class. In exchange, you take a \minus2 penalty to damage rolls. At base attack bonus \plus4, and every 4 base attack bonus thereafter, you increase the bonus and penalty by 1. If you do not make a standard attack with a melee weapon during your turn, you cannot use this style.

\feat{Celestial Body}{Bloodline, Celestial}
\parhead{Prerequisites} Nonevil alignment, Celestial Heritage.
\parhead{Benefit} You gain physical damage reduction 2/evil. This damage reduction allows you to ignore the first two points of damage you take each round. If you are hit by an evil-aligned attack, you cannot use your damage reduction for 1 round.

If you have four or more celestial bloodline feats, your damage reduction increases to half your level (if that would be greater).

\feat{Celestial Heritage}{Bloodline, Celestial}
\parhead{Prerequisite} Nonevil alignment.
\parhead{Benefit} You have the blood of a celestial creature in your veins. Once per day, when you make an attack, you can declare the attack to be a smite attack. If you smite an evil creature, you gain a bonus to attack equal to the number of celestial bloodline feats you have, and a bonus to damage equal to your character level.

\feat{Celestial Smiting}{Bloodline, Celestial}
\parhead{Prerequisites} Nonevil alignment, Celestial Heritage.
\parhead{Benefit} You can smite evil with your Celestial Heritage ability a number of times per day equal to the number of celestial bloodline feats you have \add your Charisma. You cannot smite more than once per round.

\feat{Celestial Soul}{Bloodline, Celestial}
\parhead{Prerequisites} Nonevil alignment, any three celestial feats.
\parhead{Benefit} You gain spell resistance against evil spells and spells cast by evil creatures.

\feat{Celestial Spell Conduit}{Bloodline, Celestial}
\parhead{Prerequisites} Nonevil alignment, Celestial Heritage.
\parhead{Benefit} You gain a \plus2 bonus to caster level with Evocation (Channeling) spells and spells from the Good domain. If you have four or more celestial bloodline feats, this bonus increases to \plus4.

\feat{Cleave}{Combat, Power}
\parhead{Prerequisites} Str 3, base attack bonus \plus4.
\parhead{Benefit} If you make a melee attack that knocks a creature unconscious or kills it, you can immediately make an extra melee attack at the same attack bonus against another creature within reach. The extra attack must be made with the same weapon, and it includes all bonuses and penalties that applied to the original attack. You cannot move before making this extra attack. There is no limit to the number of times you can use this feat per round.

\feat{Cleaving Smite}{Combat, Power}
\featpre Smite ability, base attack bonus \plus4, Cleave.
\featben If make a melee smite attack that grants you an extra attack from the Cleave feat, the extra attack is also a smite of the exact same type. This does not consume any additional uses of your smite ability.

\feat{Cleaving Stride}{Combat}
\parhead{Prerequisites} Str 3, base attack bonus \plus8, Cleave.
\parhead{Benefit} If you move to attack a foe, including by charging, and deal enough damage to drop it during your attack, you can continue your movement (if you have any movement remaining) to attack another foe. You may take your extra attack from the Cleave feat before or after continuing your movement.

\feat{Combat Casting}{Magic}
\parhead{Benefit} You gain a \plus2 bonus to Concentration checks made to cast a spell or use a spell-like ability. In addition, once per day you may reroll a Concentration check made to cast a spell.

\feat{Confusing Strike}{Combat, Strike}
\featpre Base attack bonus \plus12.
\featben When you make a physical attack, you may take a \minus6 penalty to hit. If you deal damage, and your foe is bloodied after your damage is dealt, you make a Will attack against the struck foe to confuse it for 1 round. Your attack bonus is equal to your base attack bonus \add your physical attack attribute (Strength or Dexterity).

\confusionexplanation A foe confused by this effect does not automatically attack you on its next turn, even though you attacked it. You cannot make more than one Strike each round.

\feat{Contingent Attack}{Combat, Reaction}
\featpre Int 5, base attack bonus \plus12.
\featben As a full-round action, you may concentrate to prepare a single attack as a contingent action. You may choose any trigger for the contingent action.

Contingent actions depend on a trigger condition. If the trigger condition occurs and you have a contingent action prepared, you may take the contingent action as an immediate action. A contingent action remains prepared until you rest. You may only have one contingent action prepared at a time. If you prepare a new contingent action, it replaces the old action.

\feat{Contingent Counter}{Combat, Reaction}
\featpre Int 3, base attack bonus \plus8.
\featben As a full-round action, you may concentrate to prepare a single attack as a contingent action. The action triggers when an opponent misses you with a melee attack.

Contingent actions depend on a trigger condition. If the trigger condition occurs and you have a contingent action prepared, you may take the contingent action as an immediate action. A contingent action remains prepared until you rest. You may only have one contingent action prepared at a time. If you prepare a new contingent action, it replaces the old action.

\feat{Combat Mobility}{Combat}
\parhead{Prerequisites} Dex 3, Dodge, base attack bonus \plus4.
\parhead{Benefit} When you move, it does not provoke attacks of opportunity from your Dodge target.

\feat{Covering Fire}{Combat, Style}
\parhead{Benefit} While in this style, if you hit a creature with a physical ranged attack, it takes a \minus2 penalty to physical attacks for 1 round. In exchange, you take a \minus2 penalty to damage rolls with physical ranged attacks. At base attack bonus \plus4, and every 4 base attack bonus thereafter, you increase both penalties by 1.

\feat{Counterstorm}{Combat, Reaction, Style}
\featpre Base attack bonus \plus16.
\featben Whenever an opponent misses you with a melee attack, it provokes an attack of opportunity from you.

\feat{Craftcaster}{Magic}
\featpres Craft (any) 8 ranks, ability to cast 2nd level spells.
\featben When casting spells which create or transform a physical substance that you are trained to craft, you gain a \plus1 bonus to your caster level per 5 ranks you have in the appropriate Craft skill.

\feat{Deadly Aim}{Combat, Style}
\parhead{Prerequisites} Dex 3.
\parhead{Benefit} While in this style, you gain a \plus2 bonus to physical ranged damage rolls. In exchange, you take a \minus2 penalty on ranged physical attacks. At base attack bonus \plus4, and every 4 base attack bonus thereafter, the penalty increases by 1 and the bonus increases by 2.

\feat{Deathless}{General}
\featpre Constitution 5 or base Fortitude defense \plus10.
\featben You become immune to death effects.

\feat{Deathless Surge}{Surge}
\featpre 12th level, Cha 5.
\featben You may gain the benefits of the \spell{death ward} spell for up to 5 rounds each day. Activating this ability is a swift action, and it lasts until you will it to stop (a free action) or until the duration expires.

\feat{Defensive Fighting}{Combat, Style}
\parhead{Benefit} While in this style, you gain a \plus2 bonus to your dodge defense modifier. In exchange, you take a \minus2 penalty on physical attacks. At base attack bonus \plus4, and every 4 base attack bonus thereafter, you increase the bonus and penalty by 1. If you do not wield a melee weapon and take the standard attack or total defense action during your turn, you cannot use this style.

\feat{Deflect Arrows}{Combat, Finesse, Reaction}
\parhead{Prerequisites} Dex 3, Improved Unarmed Strike.
\parhead{Benefit} You must have at least one hand free (holding nothing) to use this feat. Once per round when you would normally be hit with a ranged weapon, you may deflect it so that you take no damage from it. You must be aware of the attack and not flatfooted. You can deflect one additional attack at base attack bonus \plus5, \plus10, \plus15, and \plus20, to a maximum number of arrows equal to your Dexterity.
\par Attempting to deflect a ranged weapon doesn't count as an action. Unusually massive ranged weapons and ranged attacks generated by spell effects can't be deflected.

\feat{Demolishing Strike}{Combat, Strike}
\featpre Base attack bonus \plus16.
\featben When you make a physical attack, you may take a \minus8 penalty to hit. If you hit, you deal double damage. You cannot make more than one Strike each round.

\feat{Diehard}{General}
\parhead{Prerequisite} Constitution 3.
\parhead{Benefit} When you take critical damage, you do not automatically fall unconscious. You are still staggered, as normal for having no hit points remaining, and you still die when you take enough critical damage. See \pcref{Dying}, for more details.
\parhead{Normal} A character without this feat who takes critical damage falls unconscious automatically.

\feat{Dilettante}{Skill}
\parhead{Prerequisite} Int 3.
\parhead{Benefit} Choose a number of Knowledge skills equal to your Intelligence. You are treated as trained in those skills, even if you possess no ranks, allowing you to make Knowledge checks in those areas. If your Intelligence increases after taking this feat, you may choose additional Knowledge skills.

\feat{Distracting Foe}{Combat, Style}
\parhead{Prerequisites} Base attack bonus \plus4.
\parhead{Benefit} While in this style, foes you threaten take a \minus2 penalty to Concentration checks. In exchange, you take a \minus2 penalty to physical defenses against melee attacks. At base attack bonus \plus4, and every 4 base attack bonus thereafter, you increase both penalties by 1. If you do not wield a melee weapon, you cannot use this style.

\feat{Divine Spellgift}{Spellgift}
\featpre 2nd level, Cha 1.
\featben You gain the ability to use a divine spell as a spell-like ability once per day. You choose which spell when you gain this feat, and that choice cannot thereafter be changed. Your caster level is equal to your level, and you use your Charisma as your casting attribute.

The spell's level must be no greater than half of your total levels in non-casting classes. In addition, its level cannot exceed your Charisma.
\parhead{Special} You may take this feat multiple times. Each time, you choose a new spell to gain as a spell-like ability. You may use any combination of the spells a number of times per day equal to the number of spellgift feats you have. The number of spellgift feats you have cannot exceed your Charisma.

\feat{Dodge}{Combat}
\parhead{Prerequisite} Dex 3.
\parhead{Benefit} You may designate an opponent as a free action. You receive a \plus4 bonus to your dodge modifier against attacks of opportunity from that opponent that were provoked by movement. In addition, you gain a \plus4 bonus to Acrobatics checks made to tumble against that opponent.

\feat{Draconic Breath}{Bloodline, Dragon}
\parhead{Prerequisites} Con 3, any three dragon bloodline feats.
\parhead{Benefit} You gain a breath weapon based on the type of dragon you chose for the Draconic Heritage feat. The shape of the breath weapon is given on \trefnp{Dragon Types}: either a \arealarge, 5 ft. wide line or a \areamed cone. At 11th level, the size increases to a 100 ft. long, 10 ft. wide line or a \arealarge cone.

When you use your breath weapon, you make a Reflex attack to deal damage to everything in the area. It deals 1d6 damage per two levels. A failed attack deals half damage. Your attack bonus is equal to your level \add your Constitution. After using your breath weapon, you must wait 1d4 rounds before you can use it again.

\feat{Draconic Heritage}{Bloodline, Dragon}
\parhead{Benefit} You have the blood of a dragon in your veins. When you take this feat, choose a type of dragon. You gain damage reduction against the damage type that that dragon's breath weapon deals. The value of the damage reduction is equal to 5 \mtimes the number of dragon bloodline feats that you have. A list of dragons and their associated damage type is given below.

\begin{dtable}
  \lcaption{Dragon Types}
  \begin{tabularx}{\columnwidth}{>{\lcol}X >{\lcol}X >{\lcol}X}
    \thead{Dragon} & \thead{Energy Type} & \thead{Breath Weapon} \\
    Black & Acid & Line \\
    Blue & Electricity & Line \\
    Brass & Fire & Line \\
    Bronze & Electricity & Line \\
    Copper & Acid & Line \\
    Gold & Fire & Cone \\
    Green & Acid & Cone \\
    Red & Fire & Cone \\
    Silver & Cold & Cone \\
    White & Cold & Cone \\
  \end{tabularx}
\end{dtable}

\feat{Draconic Might}{Bloodline, Dragon}
\featpre Any three dragon bloodline feats.
\featben Choose a physical attribute: Strength, Dexterity, or Constitution. You gain a \plus1 bonus to that attribute.

\feat{Draconic Mind}{Bloodline, Dragon}
\featpre Any three dragon bloodline feats.
\featben Choose a mental attribute: Intelligence, Wisdom, or Charisma. You gain a \plus1 bonus to that attribute.

\feat{Draconic Scales}{Bloodline, Dragon}
\parhead{Prerequisite} Draconic Heritage.
\parhead{Benefit} You gain a \plus1 bonus to your natural armor modifier. If you have four or more dragon bloodline feats, this bonus increases to \plus2.

\feat{Draconic Senses}{Bloodline, Dragon}
\parhead{Prerequisite} Draconic Heritage.
\parhead{Benefit} You gain low-light vision. If you already have low-light vision, you can now see four times as well in darkness. If you have four or more dragon bloodline feats, you gain darkvision with a 60 foot range, or the range of your darkvision increases by 60 feet.

\feat{Draconic Spellpower}{Bloodline, Dragon}
\parhead{Prerequisite} Draconic Heritage.
\parhead{Benefit} You gain a \plus1 bonus to caster level with all spells and spell-like abilities. If you have four or more dragon bloodline feats, this bonus increases to \plus2.

\feat{Draconic Voice}{Bloodline, Dragon}
\parhead{Prerequisite} Draconic Heritage.
\parhead{Benefit} You gain a \plus2 bonus to Intimidate and Persuasion checks. If you have four or more dragon bloodline feats, this bonus increases to \plus4.

\feat{Draconic Weapons}{Bloodline, Dragon}
\parhead{Prerequisite} Draconic Heritage
\parhead{Benefit} You gain a bite natural attack that deals d8 damage for a Medium creature. If you have four or more dragon bloodline feats, you also gain a claw natural attack for each hand that deals d6 damage for a Medium creatures.

\feat{Draconic Wings}{Bloodline, Dragon}
\parhead{Prerequisite} Any three dragon bloodline feats.
\parhead{Benefit} You gain wings that sprout from your back. These allow you to slow your fall so you fall no more than 60 feet per round, preventing you from taking falling damage. If you are 8th level, you can use the wings to glide at a rate equal to your base land speed. While gliding, you move forward each round at a rate equal to your base land speed, and you descend at a rate of 10 feet per round.

If you are 14th level, you gain a fly speed equal to your base land speed, though you can only fly for a number of rounds equal to 3 \add half your Constitution. After that limit is reached, you must rest for 5 minutes to recuperate. If you are 20th level, you can fly for any length of time without needing to rest.

\feat{Driving Assault}{Combat, Maneuver, Power}
\featpre Str 5, base attack bonus \plus8.
\featben When you make a shove attack, if you succeed, your opponent provokes an attack of opportunity from you.

\feat{Dual Performance}{Performance, Skill}
\featpres Perform 14 ranks, any three Performance feats.
\featben You can sustain two performances as part of the same swift action. You take a \minus10 penalty to other Perform checks while sustaining two performances at once.

\feat{Dwarven Resilience}{Racial}
\parhead{Prerequisite} Dwarf
\parhead{Benefit} You gain a \plus2 bonus to defenses against poisons, spells, and spell-like effects.

\feat{Elemental Body}{Bloodline, Elemental}
\featpre Any three elemental bloodline feats
\featben You have a 50\% chance to ignore critical hits on you, treating them as regular hits instead.

\feat{Elemental Force}{Bloodline, Elemental}
\featpre Elemental Heritage
\featben Once per day per elemental bloodline feat you possess, you may unleash the power of your element on your foe as an attack action. Air allows you to make a shove attack with a \plus4 bonus, and you use your Constitution in place of your Strength to attack. Earth allows you to make a trip attack with a \plus4 bonus, and you use your Constitution in place of your Strength to attack. Fire allows you to make a touch attack that ignites your foe for 5 rounds if you hit. Water allows you to make a touch attack that dehydrates your foe for 5 rounds if you hit, making it vulnerable.

\feat{Elemental Heritage}{Bloodline, Elemental}
\featben You have the essence of an elemental in your body. When you take this feat, choose a type of elemental to be your elemental ancestor: air, earth, fire, or water. Air and fire elemental heritage grants a \plus2 bonus to your Reflex defense, while earth and water elemental heritage grant a \plus2 bonus to your Fortitude defense.

\feat{Elemental Mastery}{Bloodline, Elemental}
\featpre Elemental Heritage
\featben In circumstances that depend on your elemental ancestor, you gain a \plus1 bonus to physical attacks. Air grants a bonus when you are airborne or fighting airborne creatures. Earth grants a bonus when both you and your foe are standing on unworked earth or stone. Fire grants a bonus when either you or or foe is ignited, or when you are making attacks that deal fire damage. Water grants a bonus when both you and your foe are touching water. If you have four or more elemental bloodline feats, this bonus increases to \plus2.

\feat{Elemental Movement}{Bloodline, Elemental}
\featpre Elemental Heritage
\featben You gain a movement ability based on you choice of elemental ancestor. Air halves the damage you take from falling and improves the maneuverability of any flight abilities you possess by one category. Earth gives you a \plus2 bonus to Maneuver Class against attacks that would force you to move, such as shove and trip attacks. Fire increases your land speed by 5 feet. Water gives you a swim speed equal to your base land speed.

\feat{Empower Spell}{Metamagic}
\parhead{Prerequisite} Ability to cast 2nd level spells.
\parhead{Benefit} When casting a heightened spell, you gain a \plus2 bonus to caster level. A heightened spell uses up a spell slot one level higher than the spell's actual level. Unlike other metamagic feats, you can apply this metamagic feat any number of times, increasing your caster level by 2 each time.

\feat{Endurance}{General}
\featpre Con 3.
\parhead{Benefit} You are not staggered when you have no hit points remaining. You still take critical damage and risk death if you are damaged when you have no hit points.

In addition, wearing medium armor does not make you encumbered. This allows you to sleep in medium armor without penalty, and affects any class features which only function while you are unencumbered.

\parhead{Normal} A character without this feat who has no hit points remaining is staggered. In addition, a character without this feat who sleeps in medium or heavier armor is automatically fatigued the next day.
\parhead{Special} A barbarian automatically gains Endurance as a bonus feat at 3rd level. He need not select it.

\feat{Energetic Substitution}{Metamagic}
\parhead{Prerequisite} Ability to cast 2nd level spells.
\parhead{Benefit} When casting a substituted spell, you can choose what kind of energy damage it deals: cold, fire, or electricity. This can only be applied to spells that originally dealt cold, fire, or electricity damage. A substituted spell uses up a spell slot one level higher than the spell's actual level.

\feat{Enlarge Spell}{Metamagic}
\parhead{Prerequisite} Ability to cast 2nd level spells.
\parhead{Benefit} An enlarged spell has its range doubled. This metamagic can only be applied to spells with a range of \rngclose, \rngmed, or \rnglong. An enlarged spell uses up a spell slot one level higher than the spell's actual level.

\feat{Executioner}{Combat, Reaction}
\parhead{Prerequisites} Base attack bonus \plus16, any five Reaction feats.
\parhead{Benefit} Whenever a foe you threaten becomes staggered by dropping to 0 hit points, it provokes an attack of opportunity from you.

\feat{Eye of the Storm}{Awareness, Combat, Defense}
\parhead{Prerequisite} Base attack bonus \plus8.
\parhead{Benefit} You are considered to be threatened by one fewer creature than you actually are for the purpose of determining overwhelm penalties.

\feat{Exotic Weapon Proficiency}{Combat}
You understand how to use exotic weapons in combat.
\parhead{Benefit} You are proficient with exotic weapons from weapon groups that you are otherwise proficient with. This allows you to defend yourself normally with melee weapons, and to attack with exotic ranged weapons without penalty.
\parhead{Normal} You are not normally proficient with exotic weapons. You cannot defend yourself with melee weapons you are not proficient with, and you take a \minus4 penalty to physical attacks with ranged weapons you are not proficient with.

\feat{Extra Channeling}{Class}
\parhead{Prerequisite} Ability to channel energy
\parhead{Benefit} You can channel energy three more times per day.
\parhead{Normal} Without this feat, a character can channel energy a number of times per day equal to 3 \add half his or her Charisma.

\feat{Extra Invocation}{Class}
\parhead{Prerequisite} Ability to use an arcane invocation.
\parhead{Benefit} You learn a new arcane invocation. You cannot learn invocations from your prohibited schools, if any, with this feat.

\feat{Extra Rage}{Class}
\parhead{Prerequisite} Ability to rage.
\parhead{Benefit} You can rage one more time per day.

\feat{Extra Smiting}{Class}
\parhead{Prerequisite} Smite ability.
\parhead{Benefit} You can smite three more times per day. If you have more than one smite ability, you choose which ability this applies to.
\parhead{Special} You can take more this once if you have more than one smite ability. Its effects do not stack. Each time you take the feat, you choose a different smite ability you have and apply the effects to that ability.

\feat{Extra Wild Aspect}{Class}
\parhead{Prerequisite} Wild aspect ability.
\parhead{Benefit} You gain three extra uses of your wild aspect ability.

\feat{Far Shot}{Combat}
\parhead{Benefit} When you use a projectile weapon, such as a bow, its range increment increases by one-half (multiply by 1-1/2). When you use a thrown weapon, its range increment is doubled.

\feat{Fascinating Performance}{Performance, Skill}
\featpre Perform 4 ranks.
\featben As a standard action, you can make a DC 8 Perform check to fascinate creatures within \rngmed range of you. You can affect one creature per Perform rank you have. Your Perform check is opposed by a creature's Will defense. If you fail against a creature, it is not fascinated. You can sustain the performance as a standard action.

A fascinated creature is sits quietly and observes your performance, taking no other actions, for as long as you play. While fascinated, it takes a \minus4 penalty on checks made as reactions, such as Perception checks and initiative checks. If the creature notices any obvious threat, such as someone aiming a ranged weapon at it, drawing a weapon, or casting a spell, it is no longer fascinated. A fascinated creature's ally may shake it free of the spell as a standard action.

You can use any combination of performance feats you possess a number of times per day equal to the number of performance feats you have \add half your Charisma. If you fail the Perform check, the use of the ability is wasted. This is an enchantment (compulsion), mind-affecting ability.

\feat{Fearless}{General}
\featpre Charisma 5 or base Will defense \plus10.
\featben You are immune to fear effects and hostile morale effects.

\feat{Feign Weakness}{Combat, Style}
\featpre Base attack bonus \plus4.
\featben While in this style, as part of a standard attack, you may provoke an attack of opportunity from one opponent threatening you. You gain a \plus4 bonus to physical defenses against this attack. If the opponent takes the attack of opportunity, they are flat-footed against your next attack, provided that it is made within 1 round.

\feat{Focused Mind}{Racial}
\parhead{Prerequisite} Elf.
\parhead{Benefit} You gain a \plus2 bonus to Will defense. In addition, you can use your Intelligence instead of your Constitution on Concentration checks, such as when you cast spells.

\feat{Freeing Performance}{Performance, Skill}
\featpre Perform 12 ranks.
\featben By spending a minute performing continuously, you can make a DC 28 Perform check to free an ally within \rngmed range from hostile magic. This creates an effect similar to the \spell{break enchantment} spell, with a caster level equal to your Perform ranks.

\feat{Freeing Surge}{Surge}
\featpre 16th level, Dex 7.
\featben You may gain the benefits of the \spell{freedom} spell for up to 5 rounds each day. Activating this ability is a swift action, and it lasts until you will it to stop (a free action) or until the duration expires.

\feat{Giantfighter}{Racial}
\parhead{Prerequisite} Dwarf, gnome, or halfling
\parhead{Benefit} You gain a \plus2 bonus to dodge modifier against creatures of size Large or larger.

\feat{Great Fortitude}{General}
\parhead{Benefit} You gain a \plus2 bonus to your Fortitude defense.

\feat{Guardian}{Combat}
\parhead{Benefit} Allies adjacent to you are considered to be threatened by one fewer creature than they actually are for the purpose of determining overwhelm penalties. You must be wielding a melee weapon and able to attack to gain the benefits of this feat.

\feat{Healing Surge}{Surge}
\featpre 8th level, Con 3.
\featben Three times per day, you may heal 4d6 points of damage as a swift action. This healing increases by 1d6 at 12th level and every 4 levels thereafter. After using this ability, you cannot use it again for 5 rounds.

\feat{Heartseeker}{Combat, Style}
\parhead{Prerequisite} Base attack bonus \plus8.
\parhead{Benefit} While in this style, you double your critical threat range with any weapon you wield. In exchange, you take a \minus4 penalty to all physical attacks which are not critical threats or threat confirmations. This does not stack with any other effects which increase threat range.

\feat{Imbue Magic}{Item Creation}
\parhead{Prerequisite} Caster level 2nd or Craft (any) 6 ranks.
\parhead{Benefit} You can imbue items with magic using your spells or crafting ability. Imbuing an item with magic takes time and material components, as described in the magic item rules found in Rise Advanced.

When you take this feat, you choose one subschool of magic for every 5 ranks that you have in each Craft skill. You can craft items from those subschools. If you later gain additional Craft ranks, you gain new subschools appropriately.

You can also mend a broken magic item if it is one that you could make. Doing so costs a tenth of the raw materials and a quarter of the time it would take to craft that item in the first place. You cannot mend a destroyed magic item.

\feat{Imbued Spellstrike}{Metamagic}
\parhead{Prerequisite} Ability to cast 2nd level spells.
\parhead{Benefit} As part of casting an imbued spellstrike spell, you can make a single attack with a weapon in your hand. If the attack hits, the struck creature is affected by the spell, as if it had been the target, in addition to taking damage from the weapon. The imbuement fades away without effect after 1 round (at the end of your next turn) if you have not struck a foe.

Only spells which affect a single target and have a casting time no shorter than 1 standard action can be channeled in this way. An imbued spellstrike uses up a spell slot one level higher than the spell's actual level.

\feat{Imbuement Admixture}{Item Creation}
\parhead{Prerequisite} 3rd level spells.
\parhead{Benefit} You can blend two spells together to create magic items. This allows you to meet item crafting prerequisites that you could not meet with a single spell.

\feat{Immobilizing Strike}{Combat, Strike}
\featpre Base attack bonus \plus8.
\featben When you make a physical attack, you may take a \minus4 penalty to hit. If you deal damage, you make a Fortitude attack against the struck foe to immobilize it 5 rounds. Your attack bonus is equal to your base attack bonus \add your physical attack attribute (Strength or Dexterity). An immobilized creature can defend itself and attack normally, but it cannot move from its current location. Flying creatures must immediately land, but do not crash. You cannot make more than one Strike each round.

\feat{Improved Shove}{Combat, Maneuver}
\parhead{Prerequisites} Base attack bonus \plus4
\parhead{Benefit} When you shove a creature, you do not need to move with it to move it beyond your reach. You also gain a \plus2 bonus to shove attacks.

\feat{Improved Counterspell}{Magic}
\parhead{Benefit} When counterspelling, you may use any spell of the same school with a spell level at least as high as the target spell.
\parhead{Normal} Without this feat, you may counter a spell only with the same spell or with a spell specifically designated as countering the target spell.

\feat{Improved Dirty Trick}{Combat, Maneuver}
\parhead{Prerequisites} Base attack bonus \plus4
\parhead{Benefit} When you perform a dirty trick on a cretaure successfully, the creature is vulnerable for 1d4 rounds. You also gain a \plus2 bonus to dirty trick attacks.

\feat{Improved Disarm}{Combat, Finesse}
\parhead{Prerequisites} Base attack bonus \plus4
\parhead{Benefit} When you knock an item away from a creature, you can make it land up to 15 feet away in a random direction. You also gain a \plus2 bonus to disarm attacks.

\feat{Improved Feint}{Combat, Finesse}
\parhead{Prerequisites} Base attack bonus \plus4
\parhead{Benefit} When you feint against a creature, it remains flat-footed until it is struck by an attack, provided the attack is made before its next turn. You also gain a \plus2 bonus to feint attacks.
\parhead{Normal} A feinted creature is flat-footed until the next attack is made against it.

\feat{Improved Imbued Spellstrike}{Metamagic}
\parhead{Prerequisite} Ability to cast 3rd level spells, Imbued Spellstrike.
\parhead{Benefit} This metamagic functions like Imbued Spellstrike, except that the imbuement lasts for 5 minutes if you have not struck a foe. If the weapon leaves your hands or if you cast another spell, the imbuement fades away without effect.

An improved imbued spellstrike spell uses up a spell slot two levels higher than the spell's actual level.

\feat{Improved Grapple}{Combat, Maneuver}
\parhead{Prerequisites} Base attack bonus \plus4
\parhead{Benefit} Grappling does not make you flat-footed against other creatures. You also gain a \plus2 bonus to grapple attacks.
\parhead{Normal} While grappling, you are flat-footed against all creatures except the ones you are grappling.

\feat{Improved Initiative}{Combat}
\parhead{Benefit} You gain a \plus2 bonus to initiative checks. In addition, once per day you may reroll an initiative check.

\feat{Improved Overrun}{Combat, Maneuver}
\parhead{Prerequisites} Base attack bonus \plus4
\parhead{Benefit} When you attempt to overrun a foe, the creature may not choose to avoid you unless you let it. You also gain a \plus2 bonus to overrun attacks.
\parhead{Normal} Without this feat, the target of an overrun can choose to avoid you or to block you.

\feat{Improved Trip}{Combat, Maneuver}
\parhead{Prerequisites} Base attack bonus \plus4
\parhead{Benefit} When you successfully trip a foe, it immediately provokes an attack of opportunity from everyone threatening it, including you. These attacks are made as the creature is being tripped, so it does not have penalties for being prone. You also gain a \plus2 bonus to trip attacks.

\feat{Improved Unarmed Strike}{Combat}
\parhead{Benefit} You are considered to be armed even when unarmed -- that is, you do not provoke attacks or opportunity from armed opponents when you attack them while unarmed. However, you still get an attack of opportunity against any opponent who makes an unarmed attack on you.
In addition, your unarmed strikes can deal lethal or nonlethal damage as you choose.
\parhead{Normal} Without this feat, you are considered unarmed when attacking with an unarmed strike, and you can deal only nonlethal damage with such an attack.
\parhead{Special} A monk automatically gains Improved Unarmed Strike as a bonus feat at 1st level. She need not select it.

\feat{Inescapable Bulwark of Defense}{Combat, Style}
\featpre Base attack bonus \plus8.
\featben While in this style, foes that take a withdraw action to move away from you still provoke attacks of opportunity from you. In exchange, you take a \minus4 penalty to physical defenses.

\feat{Inspire Competence}{Performance, Skill}
\featpre Perform 4 ranks.
\featben As a standard action, you can make a DC 8 Perform check to inspire competence in allies within \rngmed range of you. You can affect one creature per two Perform ranks you have. You cannot affect yourself with this performance. An affected creature gains a \plus2 enhancement bonus to skill checks made with a particular skill. This bonus increases by 1 at 8 Perform ranks and every 6 Perform ranks thereafter. You can sustain the performance as a swift action.

You can use any combination of performance feats you possess a number of times per day equal to the number of performance feats you have \add half your Charisma. If you fail the Perform check, the use of the ability is wasted. This is a mind-affecting ability.

\feat{Inspire Courage}{Performance, Skill}
\featpre Perform 6 ranks.
\featben As a standard action, you can make a DC 12 Perform check to inspire courage in allies within \rngmed range of you, including yourself. You can affect one creature per two Perform ranks you have. An affected creatures gains a \plus2 enhancement bonus to physical attacks and defenses against fear. This bonus increases by 1 at 8 Perform ranks and every 6 Perform ranks thereafter. You can sustain the performance as a swift action.

You can use any combination of performance feats you possess a number of times per day equal to the number of performance feats you have \add half your Charisma. If you fail the Perform check, the use of the ability is wasted. This is a mind-affecting ability.

\feat{Inspire Greatness}{Performance, Skill}
\featpre Perform 12 ranks.
\featben As a standard action, you can make a DC 24 Perform check to inspire greatness in allies within \rngmed range of you, including yourself. You can affect one creature per two Perform ranks you have. The affected creatures gain a \plus4 enhancement bonus to special defenses and dodge defense modifier. This bonus increases to \plus5 at 20 Perform ranks.

You can use any combination of performance feats you possess a number of times per day equal to the number of performance feats you have \add half your Charisma. If you fail the Perform check, the use of the ability is wasted. This is a mind-affecting ability.

\feat{Inspire Heroics}{Performance, Skill}
\featpre Perform 18 ranks.
\featben As a standard action, you can make a DC 32 Perform check to inspire heroics in allies within \rngmed range of you, including yourself. You can affect one creature per two Perform ranks you have. All affected creatures gain temporary hit points equal to your check result and can take an extra attack at a \minus5 penalty when making a full attack. This extra attack is not cumulative with similar effects. You can sustain the performance as a swift action.

\feat{Inspire Spellpower}{Performance, Skill}
\featpre Perform 8 ranks.
\featben As a standard action, you can make a DC 16 Perform check to inspire spellpower in allies within \rngmed range of you, including yourself. You can affect one creature per two Perform ranks you have. All affected creatures gain a \plus3 enhancement bonus to their caster level with spells and spell-like abilities. This bonus increases by 1 at 14 Perform ranks and every 6 Perform ranks thereafter. You can sustain the performance as a swift action.

You can use any combination of performance feats you possess a number of times per day equal to the number of performance feats you have \add half your Charisma. If you fail the Perform check, the use of the ability is wasted.

\feat{Intense Channeling}{Class}
\parhead{Prerequisite} Channel energy 3d6.
\parhead{Benefit} You increase the damage dealt or healed by your channel energy ability by 1d6.

\feat{Intimidating Strike}{Combat, Strike}
\featpres Base attack bonus \plus8, Intimidate 8 ranks.
\featben When you make a physical attack, you may take a penalty to hit of \minus4 or greater. If you deal damage and the struck creature is within \rngmed range, you may make an Intimidate check as an immediate action to demoralize it (see \pcref{Demoralize}). You gain a bonus to your Intimidate check equal to the penalty you took to hit. You cannot make more than one Strike each round.

\feat{Intuitive Defense}{Combat, Defense, Style}
\featpre Wis 3, base attack bonus \plus8.
\featben While in this style, you can add half your Wisdom to your physical defenses in place of your Dexterity. This bonus does not apply if you would be denied your Dexterity defense bonus, such as when you are flat-footed.

\feat{Intuitive Reaction}{Combat, Reaction}
\featpres Wisdom 3, base attack bonus \plus4
\featben You may add half your Wisdom to the number of attacks of opportunity you may make each round.
\parhead{Normal} Without this feat, you may make a number of attacks of opportunity each round equal to 1 \add half your Dexterity.

\feat{Iron Will}{General}
\parhead{Benefit} You gain a \plus2 bonus to your Will defense.

\feat{Killing Blow}{Combat, Strike}
\featpre Base attack bonus \plus16.
\featben When you make a physical attack, you may take a \minus8 penalty to hit. If you deal damage, and the struck creature has no hit points remaining after your damage, it dies immediately. This is a death effect. You cannot make more than one Strike each round.

\feat{Knockdown}{Combat, Maneuver, Power}
\featpres Strength 3, base attack bonus \plus4
\featben When you successfully shove an opponent, you can choose to reduce the distance they travel by 5 feet to leave them prone after they are moved.

\feat{Ledgewalker}{Skill}
\featpre Acrobatics 5 ranks.
\featben Using Acrobatics to move along narrow surfaces does not reduce your speed. If you have 10 ranks in Acrobatics, you are not flat-footed while doing so.
\parhead{Normal} You move at half speed while balancing on precarious surfaces, and you are flat-footed while doing so.

\feat{Legendary Awareness}{Awareness, Combat}
\featpre Base attack bonus \plus12, any three Awareness feats.
\featben You cannot be overwhelmed, and never suffer overwhelm penalties. This ability can prevent rogues from sneak attacking you.

\feat{Legendary Balance}{Skill}
\featpre Acrobatics 13 ranks.
\featben You can balance on surfaces that cannot support your weight. The DC is 30 for liquids such as water, 40 for dense gases and raw energy, and 50 for ordinary air. While balancing in this way, you must take a move action each round to continue moving; you cannot remain in the same place in consecutive rounds, or you will fall. The DC increases by 2 for each consecutive round that you spend balancing in this way. You gain a \plus4 bonus on this check per size category smaller than Medium, or a \minus4 penalty per size category larger than Medium.

\feat{Legendary Charisma}{General}
\featpre 11th level, Charisma 5.
\featben You gain a \plus5 bonus to Charisma-based checks.

\feat{Legendary Climber}{Skill}
\featpre Climb 13 ranks.
\featben You gain a climb speed equal to your land speed. This grants several benefits.
\begin{itemize*}
  \item You can always take 10 on Climb checks, even when rushed, endangered, or distracted.
  \item You gain a \plus5 bonus to Climb checks.
  \item A successful Climb check allows you to move a distance equal to your climb speed.
\end{itemize*}

In addition, you can now climb surfaces that are perfectly smooth. The DC is 30 for perfectly smooth vertical surfaces, and 40 to climb on a perfectly smooth ceiling.

\feat{Legendary Constitution}{General}
\featpre 11th level, Constitution 5.
\featben You gain a \plus5 bonus to Constitution-based checks. In addition, you treat your Constitution as being 5 higher for the purpose of feats of endurance such as holding your breath or sprinting.

\feat{Legendary Craftsman}{Skill}
\featpre Craft (any) 13 ranks.
\featben You can craft items with whatever materials you have on hand. When making an item, if your Craft check is 30 or higher, you use half the costly material components to make the item, replacing them with other, less expensive components. For every 10 points by which the check exceeds 30, you halve the required material components again. This only applies once per item; use the highest result achieved while making that item.

This feat does not affect the material components required to craft magical items.

\feat{Legendary Devicesmith}{Skill}
\featpre Devices 13 ranks.
\featben You can disable spell effects on objects or areas as if they were merely complex devices. You can make a Devices check against an active spell effect within 5 feet of you. If your check result is 30 or higher, it may dispel the effect. The DC to dispel the effect is equal to 20 \add the spell's caster level. Failure indicates that the spell is not dispelled.

You must be aware of a spell to disable it, either through the Spellcraft skill or because the effect is noticeable. You cannot disable spell effects on creatures.

\feat{Legendary Dexterity}{General}
\featpre 11th level, Dexterity 5.
\featben You gain a \plus5 bonus to Dexterity-based checks, as well as to initiative.

\feat{Legendary Disguise}{Skill}
\featpre Disguise 13 ranks.
\featben Your disguises can change the magic on a creature. When you make a Disguise check, if the result is 30 or higher, you can decide how that creature and any items on the creature appear when examined by divination spells. For example, you could cause all of its equipment to appear nonmagical, or you could cause it to have a strong aura of good when examined with \spell{detect good}. You cannot create an aura of overwhelming strength with this skill.

Anyone using divination magic on the creature must make a caster level check with a DC equal to your Disguise check in order to perceive the truth. Regardless of the result of the check, the caster is not aware that the check was made.

\feat{Legendary Defense}{Combat, Defense}
\featpre Base attack bonus \plus12, any three Defense feats.
\featben You gain a \plus2 bonus to physical defenses.

\feat{Legendary Escapist}{Skill}
\featpre Escape Artist 13 ranks.
\featben You can attempt to escape from magic itself, slipping hostile spells off of your body so they dissipate harmlessly. As a standard action, you can make an Escape Artist check to throw off magical effects on you. If the result is 30 or higher, it may dispel the effect. The DC to dispel the effect is equal to 20 \add the effect's caster level. Failure indicates that the spell is not dispelled.

You must be aware of a spell to disable it, either through the Spellcraft skill or because the effect is noticeable. You can only dispel spell effects which target you directly, not area effects which include you. If a spell targets multiple creatures, you can only remove its effects on you.

\feat{Legendary Finesse}{Combat, Finesse}
\featpre Base attack bonus \plus12, any three Finesse feats.
\featben You can add half your Dexterity to damage with ranged and melee attacks. This is added in addition to half your Strength (if applicable).

\feat{Legendary Handler}{Skill}
\featpre Creature Handling 13 ranks.
\featben You gain two abilities to help you handle creatures. When pacifying a creature, if your check result is 30 or higher, you may convince it to perform a trick of your choice instead of simply pacifying it, whether or not it knows the trick.

In addition, when training a creature, if your check result is 30 or higher, you can train it in only a hour. The creature remembers the trick just as if you had spent the full amount of time training it.

\feat{Legendary Healer}{Skill}
\featpre Heal 13 ranks.
\featben A number of times per day equal to 3 \add half your Wisdom, you may take a standard action to make a Heal check on a creature you can touch. The creature heals hit points equal to twice your check result. For every twenty hit points you would restore, you can instead cure one point of critical damage.

\feat{Legendary Intelligence}{General}
\featpre 11th level, Intelligence 5.
\featben You gain a \plus5 bonus to Intelligence-based checks. In addition, you can memorize any information perfectly to recall later.

\feat{Legendary Intimidation}{Skill}
\featpre Intimidate 13 ranks.
\featben When you demoralize an opponent, if your check result exceeds 30, the creature is frightened instead of being shaken.

\feat{Legendary Leaper}{Skill}
\featpre Athletics 13 ranks.
\featben You are always treated as having a running start when jumping. When jumping, if your check result is 30 or higher, you take no penalties for making a rebounding jump off of another object while jumping (see \pcref{Rebounding Jump)}.

\feat{Legendary Liar}{Skill}
\featpre Bluff 13 ranks.
\featben Your lies can fool even magic. When you make a Bluff check, if the result is 30 or higher, anyone using magic to reveal or prevent lies must make a caster level check with a DC equal to your Bluff check in order to affect your lie. Regardless of the result of the check, the caster is not aware that the check was made.

\feat{Legendary Magic}{Magic}
\featpre 5th level spells.
\featben If your caster level with a spell is at least twice a creature's level, it is treated as bloodied for the purpose of determining the effects of that spell.

\feat{Legendary Maneuver Master}{Combat}
\featpre Base attack bonus \plus12, any three Maneuver feats.
\featben You never provoke an attack of opportunity for failing a combat maneuver. In addition, if you succeed at a combat maneuver attack by 10 or more, you deal normal damage with the weapon used to perform the maneuver in addition to gaining the successful effects of the maneuver. If the maneuver was performed without a weapon, you deal damage equivalent to an unarmed attack.

\feat{Legendary Mobility}{Combat, Mobility}
\featpre Base attack bonus \plus12, any three Mobility feats.
\featben You do not provoke attacks of opportunity when you move.

\feat{Legendary Mounted Warrior}{Combat, Mounted}
\featpre Base attack bonus \plus12, any three Mounted feats, Ride 10 ranks.
\featben When you take damage, you may choose to have your mount suffer half the damage instead of you (rounded down). Likewise, when your mount takes damage, you may choose to suffer half of that damage instead of your mount (rounded down).

\feat{Legendary Power}{Combat, Power}
\featpres Base attack bonus \plus12, any three Power feats.
\featben You can use weapons as if they were one category less encumbering than they actually are. The weapon encumbrance categories are light, medium, and heavy. For example, you can use a greatsword as a medium weapon in one hand without suffering any penalties.

\feat{Legendary Precision}{Combat, Precison}
\featpre Base attack bonus \plus12, any three Precision feats.
\featben When attacking, if you hit your opponent by 10 or more, you deal maximum damage with your weapon. If the attack is a critical threat, you automatically confirm the threat.

\feat{Legendary Reaction}{Combat, Reaction}
\featpre Base attack bonus \plus12, any three Reaction feats.
\featben You cannot caught be flat-footed, and always retain your Dexterity, dodge, and shield defense modifiers. This ability can prevent rogues from sneak attacking you.

\feat{Legendary Strength}{General}
\featpre 11th level, Strength 5.
\featben You gain a \plus5 bonus to Strength-based checks. In addition, you treat your Strength as being 5 higher for the purpose of determining your carrying and lifting capacity.

\feat{Legendary Style}{Combat}
\featpre Base attack bonus \plus12, any three Style feats.
\featben You may have two styles active at once. Both styles can be changed as part of the same swift action.

\feat{Legendary Swimmer}{Skill}
\featpre Swim 13 ranks.
\featben You gain a swim speed equal to your land speed. This grants several benefits.
\begin{itemize*}
  \item You can always take 10 on Swim checks, even when rushed, endangered, or distracted.
  \item You gain a \plus5 bonus to Swim checks.
  \item A successful Swim check allows you to move a distance equal to your swim speed.
\end{itemize*}

When swimming, if your check result is 30 or higher, you do not suffer any penalties to physical melee attacks, checks, or physical defenses for being underwater. You still suffer the normal penalty with underwater ranged attacks. You can make a Swim check as a swift action to gain this benefit in a round when you do not make a Swim check to move.

\feat{Legendary Tumbler}{Skill}
\featpre Acrobatics 13 ranks.
\featben When tumbling, if your check result is 30 or higher, you gain two benefits. First, you do not provoke attacks of opportunity for that movement. Second, you can tumble through areas occupied by enemies as if they were unoccupied.

\feat{Legendary Wisdom}{General}
\featpre 11th level, Wisdom 5.
\featben You gain a \plus5 bonus to Wisdom-based checks. In addition, once per day you can receive a prophetic insight into future events, as the \spell{divination} ritual.

\feat{Lightning Reflexes}{General}
\parhead{Benefit} You gain a \plus2 bonus to your Reflex defense.

\feat{Magical Synthesis}{Magic}
Choose two magical classes you possess.
\parhead{Prerequisites} Levels in two magical classes.
\parhead{Benefit} When gaining levels in either of your chosen classes, you increase your spellcasting ability with the other class as if your class was nonmagical. See \pcref{Spellcasters and Multiclassing}.

\feat{Manyshot}{Combat, Style}\label{feat:Manyshot}
\parhead{Prerequisites} Dex 7, base attack bonus \plus12.
\parhead{Benefit} While in this style, when you attack with a light thrown weapon or projectile weapon, you may attack with two projectiles or weapons at once. If the attack hits, the first projectile hits. If the attack hits by 5 or more, both projectiles hit. Apply precision-based damage (such as sneak attack) and critical hit damage only once for this attack. Taking advantage of this style requires having multiple weapons in hand or loading two projectiles at once, which may be difficult if you reload slowly.

\feat{Mass Transporter}{Magic}
\featpre Ability to cast a Conjuration (Translocation) spell of at least 5th level.
\featben When casting Conjuration (Translocation) spells and rituals which affect multiple creatures, you can affect a number of creatures equal to your caster level.

\feat{Mental Discipline}{Class}
\featpre Combat discipline class feature.
\featben You may use your Charisma in place of your Constitution to determine the number of times per day you can use your combat discipline ability.
\parhead{Normal} You may use your combat discipline class feature a number of times per day equal to 3 \add his Constitution.

\feat{Mental Fortress}{General}
\featpre Charisma 9 or base Will defense \plus18.
\featben You become immune to hostile mind-affecting effects.

\feat{Mighty Summons}{Magic}
\parhead{Prerequisite} Ability to cast a Conjuration (Summoning) spell of at least 2nd level.
\parhead{Benefit} Each creature you conjure with a summoning spell gains a \plus2 enhancement bonus to Strength and Constitution for the duration of the spell that summoned it.

\feat{Mounted Archery}{Combat}
\featpre Ride 1 rank.
\parhead{Benefit} The penalty you take when using a ranged weapon while mounted is decreased by 4: \minus0 instead of \minus4 if your mount is taking a double move, and \minus4 instead of \minus8 if your mount is running.

\feat{Mounted Combat}{Combat}
\parhead{Prerequisite} Ride 1 rank.
\parhead{Benefit} Once per round when your mount is hit in combat, you may attempt a Ride check as an immediate action to negate the hit. Your check result is used in place of the mount's physical defenses, if it is higher. If that would make the attack fail, the attack fails.

\feat{Mounted Trample}{Combat}
\parhead{Prerequisites} Ride 8 ranks.
\parhead{Benefit} When you attempt to overrun an opponent while mounted, your target may not choose to avoid you. Your mount may make one attack with an appropriate natural weapon (hoof, claw, or other leg-based attack) against any target you knock down, gaining the standard \plus4 bonus to physical melee attacks against prone targets.

\feat{Natural Spellgift}{Spellgift}
\featpre 2nd level, Cha 1.
\featben You gain the ability to use a natural spell as a spell-like ability once per day. You choose which spell when you gain this feat, and that choice cannot thereafter be changed. Your caster level is equal to your level, and you use your Charisma as your casting attribute.

The spell's level must be no greater than half of your total levels in non-casting classes. In addition, its level cannot exceed your Charisma.
\parhead{Special} You may take this feat multiple times. Each time, you choose a new spell to gain as a spell-like ability. You may use any combination of the spells a number of times per day equal to the number of spellgift feats you have. The number of spellgift feats you have cannot exceed your Charisma.

\feat{Nauseating Strike}{Combat, Strike}
\featpre Base attack bonus \plus12.
\featben When you make a physical attack, you may take a \minus6 penalty to hit. If you deal damage, and your foe is bloodied after your damage is dealt, you make a Fortitude attack against the struck foe to nauseate it for 1 round. Your attack bonus is equal to your base attack bonus \add your physical attack attribute (Strength or Dexterity). A nauseated creature can only take a single move action each round. You cannot make more than one Strike each round.

\feat{Open Minded}{Skill}
\parhead{Benefit} You gain two skill points. You may spend these skill points immediately.

\feat{Opportunist}{Combat, Style}
\parhead{Prerequisite} Dex 3.
\parhead{Benefit} While in this style, you gain a \plus2 bonus to attack and damage on attacks of opportunity. In exchange, you take a \minus2 penalty to attacks that are not attacks of opportunity. At base attack bonus \plus4, and every 4 base attack bonus thereafter, you increase the bonus and penalty by 1.

\feat{Overpowering Assault}{Combat, Style}
\parhead{Prerequisite} Str 3.
\parhead{Benefit} While in this style, you take a \minus2 penalty to physical defenses and gain a \plus2 bonus to maneuver attack. At base attack bonus \plus4, and every 4 base attack bonus thereafter, you increase the bonus and penalty by 1. The penalty lasts until the start of your next turn after you end the style.

\feat{Overwhelming Force}{Combat}
\parhead{Prerequisite} Str 5, base attack bonus \plus8
\parhead{Benefit} You add your full Strength to damage when wielding a medium or large melee weapon in two hands.
\parhead{Normal} Without this feat, you add half your Strength to damage.

\feat{Parry}{Combat, Defense, Reaction, Style}
\featpre Dexterity 3.
\featben While in this style, whenever you are attacked and you are not flat-footed against the attack, you may make an attack roll to attempt to parry the attack. You may treat the result of your attack roll as your physical defense against that attack if it would be higher. You can normally make one parry attempt each round. If your base attack bonus is high enough to grant you multiple attacks, you may also make multiple parry attempts. Each parry attempt after the first takes a cumulative \minus5 penalty, just like your attack rolls do.

In exchange, you take a \minus4 penalty to all physical attacks not made to parry.

\feat{Penetrating Strike}{Combat, Strike}
\featpre Base attack bonus \plus12.
\featben When you make a physical attack, you may take a \minus6 penalty to hit. If you deal damage, you make a Fortitude attack against the struck creature. If your attack succeeds, it loses its damage reduction for 5 rounds. Your attack bonus is equal to your base attack bonus \add your physical attack attribute (Strength or Dexterity). You cannot make more than one Strike each round.

\feat{Perfect Health}{General}
\featpre Constitution 3 or base Fortitude defense \plus6.
\featben You become immune to disease, except supernatural diseases such as mummy rot. If you have a Constitution of 7 or a base Fortitude defense of \plus14, you also become immune to poison and supernatural diseases.

\feat{Perfect Shot}{Combat, Style}
\parhead{Prerequisites} Dex 5, Precise Shot, base attack bonus \plus8.
\parhead{Benefit} While in this style, your ranged attacks ignore cover and concealment, except total cover and total concealment. In addition, when you shoot or throw ranged weapons at a grappling opponent, you automatically strike at the opponent you have chosen. In exchange, you must spend a full-round action to make a standard attack.

\feat{Pierce Wings}{Combat, Strike}
\featpre Base attack bonus \plus8.
\featben When you make a physical attack with a ranged weapon, you may take a \minus4 penalty to hit. If you deal damage, you make a Fortitude attack against the struck creature. If your attack succeeds, it loses its ability to fly for 1 round. Your attack bonus is equal to your base attack bonus \add your physical attack attribute (Strength or Dexterity). This only affects creatures who use wings or other physical means to fly, and has no effect on creatures with magical or supernatural flight. You cannot make more than one Strike each round.

\feat{Prescient Surge}{Awareness, Surge}
\featpre 12th level, Wis 5.
\featben Three times per day, you may roll twice on any single physical attack or opposed check. You must decide to use this ability before making the roll. Activating this ability is an immediate action.

\feat{Point Blank Shot}{Combat, Style}
\parhead{Benefit} While in this style, you gain a \plus2 bonus to physical damage rolls with ranged weapons when attacking targets within half of your range increment. In exchange, you take a \minus2 penalty to physical attacks against foes beyond that range. At base attack bonus \plus4, and every 4 base attack bonus thereafter, you increase the bonus and penalty by 1.

\feat{Poison Expert}{Skill}
\featpre Intelligence 3, Craft (poison) 8 ranks.
\featben Whenever you use poisons, you can treat the DC as 10 \add your level \add your Intelligence in place of the poison's normal DC. In addition, you cannot accidentally poison yourself while applying poison to an object.

\feat{Potent Spellgift}{Spellgift}
\featpre Any spellgift feat.
\featben You gain a \plus2 bonus to your caster level with spell-like abilities from spellgift feats.

\feat{Power Attack}{Combat, Style}
\parhead{Prerequisite} Str 3.
\parhead{Benefit} While in this style, you gain a \plus2 bonus to physical melee damage rolls. In exchange, you take a \minus2 penalty to physical melee attacks. At base attack bonus \plus4, and every 4 base attack bonus thereafter, you increase the bonus and penalty by 1. The bonus damage and penalty do not apply to maneuvers or touch attacks. The bonus damage is halved if you are dealing damage with an off-hand weapon or light weapon.

\feat{Precise Shot}{Combat, Style}
\parhead{Benefit} While in this style, you can add half your Wisdom to damage with ranged attacks in addition to half your Strength. Additionally, you can ignore creature when determining whether targets have active cover from you. In exchange, you must spend a full-round action to make a standard attack.

\feat{Predictive Defense}{Combat, Defense, Style}
\featpre Intelligence 3, base attack bonus \plus8.
\featben While in this style, you can add half your Intelligence to your physical defenses in place of your Dexterity. This bonus does not apply if you would be denied your Dexterity defense bonus, such as when you are flat-footed.

\feat{Quick Draw}{Combat}
\parhead{Benefit} You can draw light weapons as a free action and medium weapons as a swift action. You can draw heavy weapons and hidden weapons of any type (see the Sleight of Hand skill) as a move action.
\par A character who has selected this feat may throw light weapons at his full normal rate of attacks (much like a character with a bow).

If you have three or more Reaction feats, you can draw light and medium weapons as an immediate action, and heavy and hidden weapons as a swift action.
\parhead{Normal} Without this feat, you may draw light weapons as a swift action, medium and heavy weapons as a move action, and hidden weapons as a standard action.

\feat{Quicken Spell}{Metamagic}\label{Quicken Spell}
\parhead{Prerequisite} Ability to cast 3rd level spells.
\parhead{Benefit} Casting a quickened spell is a swift action. You can perform another action, even casting another spell, in the same round as you cast a quickened spell. However, casting a quickened spell is mentally exhausting. In the turn after you cast it, you lose your standard action. You may cast only one quickened spell per round. A spell whose casting time is more than 1 standard action cannot be quickened. A quickened spell uses up a spell slot two levels higher than the spell's actual level. Casting a quickened spell doesn't provoke an attack of opportunity.
\parhead{Special} All spellcasters cast a quickened spell as a swift action, even if they would normally increase the casting time of spells with metamagic applied. This is an exception to the general rule that applying metamagic increases the casting time of a spell.

\feat{Ranged Legerdemain}{Skill}
\parhead{Prerequisite} Ability to cast 2nd level spells
\parhead{Benefit} By expending an Evocation (Control) spell of 2nd level or higher, you can use the Disable Device or Sleight of Hand skills at \rngclose range for a number of rounds equal to half the level of the spell slot.

\feat{Rapid Metamagic}{Magic}
\parhead{Prerequisites} Spellcraft 8 ranks, ability to cast spells, one metamagic feat
\parhead{Benefit} When you apply a metamagic feat to a spell, the spell only takes its normal casting time.
\parhead{Normal} Without this feat, applying metamagic takes a standard action (if the spell normally requires less than a standard action), a full-round action (if the spell normally requires a standard action), or an additional full-round action (if the spell takes 1 full round or longer to cast).

\feat{Rapid Performance}{Performance, Skill}
\featpre Perform 14 ranks, any three Performance feats.
\featben You can use performance feats as a swift action by taking a \minus5 penalty on the Perform check to activate the feat.

\feat{Reach Spell}{Metamatic}
\parhead{Prerequisite} Ability to cast 3rd level spells.
\parhead{Benefit} When casting a reach spell, you can use a spell with a range of touch on a target within \rngclose range. You must succeed on a ranged touch attack. A reach spell uses up a spell slot two levels higher than the spell's actual level.

\feat{Recover Performance}{Performance, Skill}
\featpre Perform 8 ranks, any other Performance feat
\featben If you fail the Perform check to use a Performance feat, you can take a move action to reroll the Perform check. If the initial performance was done with the Rapid Performance feat, you take the same \minus5 penalty on the reroll.

\feat{Residual Awareness}{Magic}
\featpre Ability to cast a Divination (Awareness) spell.
\featben Whenever you cast a Divination (Awareness) spell, you gain an enhancement bonus to Perception checks equal to the level of the spell for 5 rounds.

\feat{Residual Beguilement}{Magic}
\featpre Ability to cast an Enchantment (Beguilement) spell of at least 3rd level.
\featben Whenever you cast a Enchantment (Beguilement) spell, you gain an enhancement bonus to Persuasion checks equal to the level of the spell for 5 minutes.

\feat{Residual Knowledge}{Magic}
\featpre Ability to cast a Divination (Knowledge) spell.
\featben Whenever you cast a Divination (Knowledge) spell, you gain an enhancement bonus to Knowledge checks equal to the level of the spell for 5 rounds.

\feat{Residual Shield}{Magic}
\featpre Ability to cast an Abjuration (Shielding) spell.
\featben Whenever you cast an Abjuration (Shielding) spell with a duration of \durshort or longer on another creature, you gain the benefits on yourself for 1 round.

\feat{Resilient Magic}{Magic}
\featpre Ability to cast an Abjuration (Negation) spell of at least 2nd level.
\featben Your spells persist for 1 round after being dispelled if they still have time remaining in their duration.

\feat{Retributive Counterspell}{Magic}
\parhead{Prerequisites} Ability to cast 4th level spells.
\parhead{Benefit} As part of the action to counter a spell, you may expend an Abjuration (Negation) spell of 4th level or higher. If you do, the counterspelled spell is turned back on the caster as if it were affected by the \spell{spell turning} spell. If it cannot be affected by \spell{spell turning}, such as if it is a spell that only affects the caster, it is simply countered as normal.

\feat{Reveal the Weak Point}{Combat, Style}
\featpre Base attack bonus \plus4.
\parhead{Benefit} While in this style, any creature you deal damage to with a physical attack takes a \minus3 penalty to physical defenses for 1 round. In exchange, you take a \minus3 penalty to physical attack and damage. At base attack bonus \plus8, and every 4 base attack bonus thereafter, you increase the bonus and penalty by 1. As normal, penalties from the same source do not stack.

\feat{Ride-By Attack}{Combat}
\featpre Ride 8 ranks.
\parhead{Benefit} When you are mounted and use the charge action, you may move and attack as if with a standard charge and then move again (continuing the straight line of the charge). You do not need to attack from the closest possible space when making a ride-by attack. Your total movement for the round can't exceed the distance you could normally move on a mounted charge. You and your mount do not provoke an attack of opportunity from the opponent that you attack.

\feat{Riposte}{Combat, Reaction}
\featpre Dexterity 3, base attack bonus \plus4, Parry
\featben When you successfully parry an attack from a foe you threaten, if your parry attempt exceeds your opponent's attack roll by 5 or more, your foe provokes an attack of opportunity from you.

\feat{Ritual Caster}{Magic}
\featpre Intelligence 3.
\featben You can learn and perform rituals as if you were an arcane caster with a caster level equal to your character level. The maximum level of ritual that you can learn or perform is equal to half your character level.

\feat{Scale the Beast}{Skill}
\featpre Climb 8 ranks.
\featben As a standard action, you can make a Climb check against the Maneuver Class of a creature adjacent to you. The creature must be three or more size categories larger than you. You gain a \plus4 bonus to the check for every size category by which it exceeds your own. If you succeed, you can climb the creature as if it were a solid object with a Climb DC equal to its Maneuver Class. The creature takes a \minus4 penalty to physical attacks against you, but is not otherwise encumbered by your presence unless it cannot support your weight. It can attempt to remove you by attacking you, or with an appropriate maneuver, such as grappling or shoving.

\feat{Selective Channeling}{Class}
\parhead{Prerequisites} Ability to channel energy.
\parhead{Benefit} You can exclude up to two additional creatures from the effect when you channel energy.
\parhead{Normal} Without this feat, you can exclude a number of creatures from the effect equal to 1 \add half your Wisdom.

\feat{Shape Spell}{Metamagic}
\parhead{Prerequisite} Ability to cast 3rd level spells.
\parhead{Benefit} When casting a shaped spell, you can exclude any number of 5-foot cubes within the spell's area. This allows you to prevent the spell from affecting your allies, while still allowing it to affect your enemies. The area affected by the spell must be contiguous.

Only area spells can be shaped. A shaped spell uses a spell slot two levels higher than the spell's original level.

\feat{Shattering Surge}{Power, Surge}
\featpre 8th level, Str 3.
\featben Three times per day, when you successfully attempt to sunder an object, you may cause it to be affected as if by the targeted version of the \spell{shatter} spell with a caster level equal to your level. Activating this ability is an immediate action.

\feat{Shield Discipline}{Class}
\featpre Armor discipline ability.
\featben You apply the benefits of your armor discipline ability while using shields, as well as while wearing armor. If you reduce the encumbrance category of body armor you wear, you also reduce the encumbrance category of shields you use by the same amount. Reducing a shield's encumbrance category causes tower shields to be treated as heavy shields (and no longer impose a \minus2 penalty to physical attacks), heavy shields to be treated as light shields, and both light shields and bucklers to be treated as being unarmored (allowing you to freely use other items in that hand). As normal, if you use your shield hand for something other than defending yourself with your shield, you do not gain your shield's bonus to physical defenses that round.

\feat{Shield Proficiency}{Combat}
\parhead{Benefit} You can use a shield and take only the standard penalties.
\parhead{Normal} When you are using a shield with which you are not proficient, you take the shield's armor check penalty on physical attacks and on all checks that involve moving, including Ride checks.
\parhead{Special} Barbarians, clerics, druids, fighters, paladins, rangers, and spellwarped automatically have Shield Proficiency as a bonus feat. They need not select it.

\feat{Shielded Parry}{Combat, Defense, Reaction}
\featpres Dexterity 3, shield proficiency, Parry
\featben When readying yourself to parry incoming blows and using a shield, you may add your shield modifier to your attack roll made to parry.

\feat{Sickening Strike}{Combat, Strike}
\featpre Base attack bonus \plus8.
\featben When you make a physical attack, you may take a \minus4 penalty to hit. If you deal damage, you make a Fortitude attack against the struck creature to sicken it for 5 rounds. Your attack bonus is equal to your base attack bonus \add your physical attack attribute (Strength or Dexterity). You cannot make more than one Strike each round.

\feat{Silent Spell}{Metamagic}
\parhead{Prerequisite} Ability to cast 2nd level spells.
\parhead{Benefit} A silent spell can be cast with no verbal components. Spells without verbal components are not affected. A silent spell uses up a spell slot one level higher than the spell's actual level.

\feat{Skill Focus}{Skill}
Choose a skill.
\parhead{Benefit} You gain a \plus3 bonus to all checks involving that skill. In addition, once per day you may reroll a skill check made with that skill.
\parhead{Special} You can gain this feat multiple times. Its effects do not stack. Each time you take the feat, it applies to a new skill.

\feat{Skill Mastery}{Skill}
Choose a skill.
\featpre 10 ranks in the chosen skill.
\featben When making a skill check with your chosen skill, stress and distraction do not prevent you from taking 10, allowing you to automatically succeed. You still cannot take 10 when opposed. See \pcref{Taking 10}.
\parhead{Special} You can gain this feat multiple times. Its effects do not stack. Each time you take the feat, it applies to a new skill.

\feat{Skill Training}{Skill}
\featben Choose any two skills. You treat those skills as class skills.

\feat{Snatch Arrows}{Combat, Finesse, Reaction}
\parhead{Prerequisites} Dex 5, Deflect Arrows, Improved Unarmed Strike.
\parhead{Benefit} When using the Deflect Arrows feat, you may catch the weapon instead of just deflecting it. As an immediate action, you can throw a caught weapon back at the original attacker. If you have an appropriate projectile weapon available, you can also fire caught ammunition.
\par You must have at least one hand free (holding nothing) to use this feat.

\feat{Spell Focus}{Magic}
Choose a school of magic or a spell descriptor.
\parhead{Prerequisite} 2nd level spells.
\parhead{Benefit} You get a \plus2 bonus to your caster level when casting spells and using spell-like abilities from the school of magic or with the spell descriptor you select.
\parhead{Special} You can gain this feat multiple times. Its effects do not stack. Each time you take the feat, it applies to a new school of magic.

\feat{Spell Specialization}{Magic}
Choose a school of magic or a spell descriptor to which you already have applied the Spell Focus feat.
\featpre 4th level spells, Spell Focus.
\featben You gain a \plus2 bonus to your caster level when casting spells and using spell-like abilities from the school of magic or with the spell descriptor you select. In exchange, you take a \minus2 penalty to caster level with all other spells and spell-like abilities.

\feat{Spellwoven Performance}{Performance, Skill}
\featpre Ability to cast spells, Perform 8 ranks, any other Performance feat.
\featben You can cast enchantment and illusion spells while sustaining a performance. The Spellcraft DC to identify those spells as they are cast increases by 10, as the performance disguises the magic.

\feat{Spirited Charge}{Combat}
\parhead{Prerequisites} Ride 1 rank, Ride-By Attack.
\parhead{Benefit} When mounted and using the charge action, you deal double damage with a melee weapon (or triple damage with a lance) on your attack at the end of the charge.%Spirited Charge

\feat{Spring Attack}{Combat, Style}
\parhead{Prerequisites} Dex 3, Dodge, Mobility, base attack bonus \plus4.
\parhead{Benefit} While in this style, you can take a full-round action to move up to your speed and make a standard attack. You must move at least 5 feet after each attack before you can make another attack. You cannot use this style while encumbered.

\feat{Still Spell}{Metamagic}
\parhead{Prerequisite} Ability to cast 2nd level spells.
\parhead{Benefit} A stilled spell can be cast with no somatic components. Spells without somatic components are not affected. A stilled spell uses up a spell slot one level higher than the spell's actual level.

\feat{Stonecunning}{Racial}
\parhead{Prerequisite} Dwarf.
\parhead{Benefit} You gain a \plus2 bonus to Craft and Perception checks related to stone or metal. In addition, if you come within 10 feet of unusual stonework, you can make a Perception check to notice it as if you were actively searching. Finally, you can also intuit depth, sensing your approximate depth underground as naturally as a human can sense which way is up.

\feat{Suggestive Performance}{Performance, Skill}
\featpre Perform 10 ranks, Fascinating Performance.
\featben As a swift action, you can make a DC 20 Perform check to suggest an action to a single creature you are currently fascinating with the Fascinating Performance feat. This functions as the \spell{suggestion} spell, with an attack bonus equal to your Perform modifier. You can affect an additional creature at 12 Perform ranks and every 2 Perform ranks thereafter.

You can use any combination of performance feats you possess a number of times per day equal to the number of performance feats you have \add half your Charisma. If you fail the Perform check, the use of the ability is wasted.

\feat{Surge of Speed}{Surge}
\featpre 8th level, Dex 3.
\featben You may gain the benefits of the \spell{expeditious retreat} spell for up to 5 rounds each day. Activating this ability is a swift action, and it lasts until you will it to stop (a free action) or until the duration expires.

\feat{Sustained Spell}{Metamagic}
\featpre Ability to cast 2nd level spells.
\featben You can maintain concentration on a sustained spell as a swift action instead of as a standard action. If you cast any other spell, you lose the ability to sustain the spell. This only affects spell duration, and has no effect on spells with special effects based on concentration, such as \spell{call lightning}. A sustained spell uses up a spell slot one level higher than the spell's actual level.

\feat{Swift}{General}
\parhead{Benefit} You increase your land speed by 5 feet.

\feat{Tactical Analysis}{Combat}
\parhead{Prerequisite} Intelligence 3, Base attack bonus \plus4.
\parhead{Benefit} You can attempt to identify the strengths and weaknesses of creatures based on your combat experience. As a swift action, you can make a special check with a bonus equal to your base attack bonus \add your Intelligence \add 2 per round you have seen the creature fight. The DC is equal to 10 \add the creature's CR. If you succeed, you learn about the monster's combat abilities as if you had made a successful Knowledge check.

\feat{Tactical Prediction}{Combat}
\parhead{Prerequisites} Intelligence 3, Base attack bonus \plus8
\parhead{Benefit} You can attempt to predict what your opponent will do. As a swift action, you can make a special check with a bonus equal to your base attack bonus \add your Intelligence \add 2 per round you have seen the creature fight. The DC is equal to 15 \add the creature's CR. If you succeed, you learn what the creature is planning to do during its next turn. Of course, it can change its plans, particularly if it hears you tell your allies what it will do.

\feat{Threatening Fire}{Combat, Style}
\parhead{Prerequisites} Base attack bonus \plus4.
\parhead{Benefit} While in this style, when you make a standard attack to attack a creature with a ranged weapon, you are considered to be threatening it for the purpose of determining overwhelm penalties. In exchange, you must direct all your attacks against the same creature when making a standard attack.

\feat{Trapfinder}{Skill}
\featpre Perception 5 ranks.
\featben As a full-round action, you can move up to 10 feet while searching every square within 10 feet of you for traps with the Perception skill (see \pcref{Search}). If you detect a trap partway through your movement, you may immediately stop moving.

\feat{Tower Shield Proficiency}{Combat}
\parhead{Prerequisite} Shield Proficiency.
\parhead{Benefit} You can use a tower shield and suffer only the standard penalties.
\parhead{Normal} A character who is using a shield with which he or she is not proficient takes the shield's armor check penalty on physical attacks and on all checks that involve moving, including Ride checks.
\parhead{Special} Fighters and paladins are automatically proficient with tower shields.

\feat{Toughness}{General}
\parhead{Benefit} You gain \plus3 hit points, \plus1 per level above 3.

\feat{Track}{Skill}\label{Track}
\parhead{Benefit} To find tracks or to follow them for 1 mile requires a successful Survival check. You must make another Survival check every time the tracks become difficult to follow.
You move at half your normal speed (or at your normal speed with a \minus5 penalty on the check, or at up to twice your normal speed with a \minus20 penalty on the check). The DC depends on the surface and the prevailing conditions, as given on the table below:

\begin{dtable}
\begin{tabularx}{\columnwidth}{>{\lcol}X c >{\lcol}X c}
\thead{Surface} & \thead{Survival DC}  & \thead{Surface} & \thead{Survival DC} \\
Very soft ground  & 5  & Firm ground  & 15 \\
Soft ground  & 10  & Hard ground  & 20
\end{tabularx}
\end{dtable}
\par \emph{Very Soft Ground:} Any surface (fresh snow, thick dust, wet mud) that holds deep, clear impressions of footprints.
\par \emph{Soft Ground:} Any surface soft enough to yield to pressure, but firmer than wet mud or fresh snow, in which a creature leaves frequent but shallow footprints.
\par \emph{Firm Ground:} Most normal outdoor surfaces (such as lawns, fields, woods, and the like) or exceptionally soft or dirty indoor surfaces (thick rugs and very dirty or dusty floors). The creature might leave some traces (broken branches or tufts of hair), but it leaves only occasional or partial footprints.
\par \emph{Hard Ground:} Any surface that doesn't hold footprints at all, such as bare rock or an indoor floor. Most streambeds fall into this category, since any footprints left behind are obscured or washed away. The creature leaves only traces (scuff marks or displaced pebbles).
\par Several modifiers may apply to the Survival check, as given on the table below.

\begin{dtable}
\begin{tabularx}{\columnwidth}{>{\lcol}X >{\rcol}p{6em}}
\thead{Condition}  & \thead{Survival DC Modifier} \\
Every three creatures in the group being tracked  & \minus1 \\
Size of creature or creatures being tracked:\footnotetemp{1} &  \\
Fine  & \plus16 \\
Diminutive  & \plus12 \\
Tiny  & \plus8 \\
Small  & \plus4 \\
Medium  & \plus0 \\
Large  & \minus4 \\
Huge  & \minus8 \\
Gargantuan  & \minus12 \\
Colossal  & \minus16 \\
Every 24 hours since the trail was made  & \plus1 \\
Every hour of rain since the trail was made  & \plus1 \\
Fresh snow cover since the trail was made  & \plus10 \\
\thead{Poor visibility:\footnotetemp{2}} &  \\
Overcast or moonless night  & \plus6 \\
Moonlight  & \plus3 \\
Fog or precipitation  & \plus3 \\
Tracked party hides trail (and moves at half speed)  & \plus5
\end{tabularx}
1 For a group of mixed sizes, apply only the modifier for the largest size category. \\
2 Apply only the largest modifier from this category.
\end{dtable}

If you fail a Survival check, you can retry after 1 hour (outdoors) or 10 minutes (indoors) of searching.
\parhead{Normal} Without this feat, you can use the Survival skill to find tracks, but you can follow them only if the DC for the task is 10 or lower. Alternatively, you can use the Search skill to find a footprint or similar sign of a creature's passage using the DCs given above, but you can't use Search to follow tracks, even if someone else has already found them.
\parhead{Special} A ranger automatically has Track as a bonus feat. He need not select it.

\feat{Trap Sense}{Class}
\featpre Danger sense class feature.
\featben Whenever you come within 10 feet of a trap, you receive an immediate Perception check to notice the trap. This check should be made secretly, so you do not know whether you failed to notice a trap.

\feat{Two-Weapon Defense}{Combat}
\parhead{Prerequisites} Dex 3, Two-Weapon Fighting.
\parhead{Benefit} When wielding a double weapon or two weapons (not including natural weapons or unarmed strikes), you gain a \plus1 shield defense bonus. This bonus increases to \plus2 once your base attack bonus reaches \plus8, and to \plus3 at base attack bonus \plus16.
%\par Additionally, if you are proficient with bucklers, you may keep your buckler's shield bonus to AC while you attack with a weapon in your off-hand.
%\parhead{Normal} Without this feat, a character wielding a buckler loses the buckler's shield bonus to AC when using an off-hand weapon to attack.

\feat{Two-Weapon Fighting}{Combat}\label{feat:Two-Weapon Fighting}
You can fight with a weapon in each hand more effectively.
\parhead{Prerequisite} Dex 3.
\parhead{Benefit} You gain a \plus2 bonus to physical attacks when attacking with two weapons at once.

\feat{Two-Weapon Rend}{Combat, Power}\label{Two-Weapon Rend}
\parhead{Prerequisites} Str 5, Two-Weapon Fighting, base attack bonus \plus12.
\parhead{Benefit} When fighting with two weapons at once, you may add half your Strength to damage with your secondary weapon.
\parhead{Normal} Without this feat, you do not apply Strength to damage with your secondary weapon.

\feat{Unshakeable Stance}{Combat, Style}
\parhead{Prerequisites} Base attack bonus \plus4
\parhead{Benefit} While in this style, you gain a \plus3 bonus to Maneuver Class. In exchange, you are unable to move. At base attack bonus \plus8, and every 4 base attack bonus thereafter, you increase the bonus by 1.

\feat{Versatile Crafter}{Item Creation}
\parhead{Prerequisite} Craft (any) 10 ranks.
\parhead{Benefit} You learn how to make items from one subschool for every two ranks you have in each Craft skill. See the Craft skill description for details.
\parhead{Normal} You learn how to make items from one subschool for every five ranks you have in each Craft skill.

\feat{Veteran's Experience}{Skill}
\featpre Base attack bonus \plus8
\featben You can use your battlefield experience in place of learned knowledge to identify monsters. When attempting to identify a monster, you may roll your base attack bonus \add your Intelligence. A successful check gives you the same information as a Knowledge check would.

\feat{Wall Slam}{Combat, Maneuver, Power}
\featpre Str 5, base attack bonus \plus8
\featben If you shove an opponent into a wall or other solid object, he takes d6 damage \add half your Strength and provokes attacks of opportunity from all threatening creatures, including you.

\feat{Wardwalker}{Magic}
\featpre Ability to cast an Abjuration (Warding) spell of at least 3rd level.
\featben When you trigger an Abjuration (Warding) spell, it does not take effect until one round after you triggered it, giving you time to escape.

\feat{Weapon Focus}{Combat}
Choose one weapon group.
\parhead{Prerequisites} Base attack bonus \plus4, proficiency with selected weapon group.
\parhead{Benefit} You gain a special ability based on the weapon group chosen. These special abilities only function when you are making attacks with weapons from your chosen weapon group.
\begin{itemize*}
    \item Armor weapons: When you perform a shield bash, you still benefit from the shield's defense bonus. In addition, armor spikes no longer impose a penalty to your physical defenses.
    \item Axes: You gain a \plus2 bonus to physical attacks against unarmored creatures (creatures without an armor defense bonus).
    \item Blades, heavy: You gain a \plus2 bonus to physical attacks against unarmored creatures (creatures without an armor defense bonus).
    \item Blades, light: If you draw a light blade and attack with it in the same round, your foe is flat-footed against the attack. After the attack, the creature is unaffected by this ability for the rest of the encounter.
    \item Blunt weapons: When you deal damage to a creature, it takes a \minus2 penalty to Will defenses for 1 round. This penalty is not cumulative with itself.
    \item Bows: You can ignore cover (but not total cover) provided by creatures and objects that are at least ten feet away from both you and your target.
    \item Crossbows: The time required for you to reload crossbows is reduced to a free action (for a hand or light crossbow) or a move action (for a heavy crossbow). Reloading a heavy crossbow still provokes an attack of opportunity, but reloading hand and light crossbows does not.
    \item Flexible weapons: You gain a \plus2 bonus to physical attacks against creatures with a shield defense bonus.
    \item Headed weapons: You gain a \plus2 bonus to physical attacks against armored creatures (creatures with an armor defense bonus).
    \item Monk weapons: ??
    \item Polearms: You can switch grips to short haft or stop short hafting a polearm as a swift action, and you take no penalty while short hafting it.
    \item Spears: If a creature charges you, it provokes an attack of opportunity from you when it enters your threatened area.
    \item Thrown weapons: You can defend yourself with a thrown weapon as you throw it, which can prevent you from taking attacks of opportunity for throwing weapons in melee (see \pcref{Thrown Weapons in Melee}).
\end{itemize*}
\parhead{Special} You cannot choose simple weapons when you take this feat.

\feat{Weapon Proficiency}{Combat}
Choose one weapon group.
\parhead{Benefit} You are proficient with weapons from your chosen weapon group. This allows you to defend yourself normally with melee weapons, and to attack with exotic ranged weapons without penalty.
\parhead{Normal} You cannot defend yourself with melee weapons you are not proficient with, and you take a \minus4 penalty to physical attacks with ranged weapons you are not proficient with.
\parhead{Special} You can gain Weapon Proficiency multiple times. Each time you take the feat, it applies to a new weapon group. You cannot choose simple weapons.

\feat{Widen Spell}{Metamagic}
\parhead{Prerequisite} Ability to cast 4th level spells.
\parhead{Benefit} You can alter a burst, emanation, line, or spread shaped spell to increase its area. Any numeric measurements of the spell's area increase by 100\%. A widened spell uses up a spell slot three levels higher than the spell's actual level.
\par Spells that do not have an area of one of these four sorts are not affected by this feat.

\feat{Whirlwind Attack}{Combat, Style}
\parhead{Prerequisites} Dex 5, base attack bonus \plus8.
\parhead{Benefit} While in this style, as a full-round action, you can make a single melee attack against each opponent you threaten. In exchange, you take a \minus3 penalty to attack.

\section{Other Feat Rules}

\subsection{Bonus Feats}
Some class features and abilities grant a character bonus feats. Unless otherwise specified, the character must still meet any prerequisites for the feat. If the character does not meet the prerequisites at the time the bonus feat is granted, the character does not gain the feat. If the character later meets the prerequisites, the character immediately gains the benefit of the bonus feat.

If a character gains a feat as a bonus feat that he or she has already acquired through other means, the character may select instead any other feat for which she qualifies.

\subsection{Retraining Feats}
At every even level, your character can choose to retrain an old feat in exchange for a new feat. You can only retrain feats for other feats you could have acquired at the time you took the original feat. Thus, you cannot retrain feats gained through class features which give you a specific feat, since there were no other feats you could have taken. However, a 6th level fighter can retrain his 2nd level fighter bonus feat for any other combat feat that he qualified for at 2nd level.
