\ExplSyntaxOn

\newcommand{\tempderp}{''}

\def\replace#1#2#3{%
 \def\tmp##1#2{##1#3\tmp}%
   \tmp#1\stopreplace#2\stopreplace}
\def\stopreplace#1\stopreplace{}

\DeclareDocumentCommand{\createspell}{m m o m}
{
    \DeclareDocumentCommand{#1}{s o}
    {
        \IfBooleanTF{##1}
        {
            \spellhead[##2]*{#2}[#3]~#4
        }
        {
            \spellhead[##2]{#2}[#3]~#4
        }
    }
}

\newcommand{\makemacro}[2]
{
    \lowercase{
        \expandafter
        \newcommand\csname#1\endcsname
    }{#1:~#2}
}

\ExplSyntaxOff

\createspell{\SLablateimpact}{Ablate Impact}
    {Immediately reduce damage from physical attacks.}
\createspell{\SLablativeshield}{Ablative Shield}
    {Immediately reduce damage from spells.}
\createspell{\SLacidarrowgreater}{Acid Arrow}[Greater]
    {Fling acid to deal lingering damage and stagger.}
\createspell{\SLacidarrow}{Acid Arrow}
    {Fling acid to deal lingering damage}
\createspell{\SLacidfog}{Acid Fog}
    {Acidic fog deals damage each round.}
\createspell{\SLagony}{Agony}
    {Creature suffers penalties due to debilitating pain.}
\createspell{\SLaid}{Aid}
    {Grant ally temporary hit points and immunity to fear.}
\createspell{\SLairwalk}{Air Walk}
    {Ally treads on air as if solid.}
\createspell{\SLalignweapon}{Align Weapon}
    {Weapon becomes \plus2 and chaotic.}
\createspell{\SLalignweapon}{Align Weapon}
    {Weapon becomes \plus2 and evil.}
\createspell{\SLalignweapon}{Align Weapon}
    {Weapon becomes \plus2 and good.}
\createspell{\SLalignweapon}{Align Weapon}
    {Weapon becomes \plus2 and lawful.}
\createspell{\SLanimalgrowth}{Animal Growth}
    {Enlarges multiple animals.}
\createspell{\SLantilifeshell}{Antilife Shell}
    {Zone blocks living creatures from entering.}
\createspell{\SLantimagicfield}{Antimagic Field}
    {Emanation negates magic.}
\createspell{\SLaqueousblade}{Aqueous Blade}
    {Weapon transforms to attack Reflex defense.}
\createspell{\SLassimilate}{Assimilate}
    {Foe is absorbed into your body.}
\createspell{\SLassimilate}{Assimilate}
    {Incorporate creature into your own body.}
\createspell{\SLaversion}{Aversion}
    {Compel creature to avoid something.}
\createspell{\SLbane}{Bane}
    {Foe is impaired.}
\createspell{\SLbane}{Bane}
    {Nearby enemies take \minus2 to attack.}
\createspell{\SLbanishment}{Banishment}
    {Banish multiple extraplanar creatures. }
\createspell{\SLbarkskin}{Barkskin}
    {Grants \plus2 bonus to AC and DR 2/adamantine.}
\createspell{\SLbestowcurse}{Bestow Curse}
    {\minus6 to an attribute score; \minus4 on attack rolls, saves, and checks; or 25\% chance of losing each action.}
\createspell{\SLbestowcurse}{Bestow Curse}
    {\minus6 to an attribute; \minus4 on attack rolls, saves, checks, and AC; or 25\% chance of losing each action.}
\createspell{\SLblacktentacles}{Black Tentacles}
    {Tentacles grapple all within small spread.}
\createspell{\SLbladebarrier}{Blade Barrier}
    {Wall of blades deals 6d6 damage.}
\createspell{\SLblasphemy}{Blasphemy}
    {Kills, paralyzes, nauseatees, or sickens nonevil subjects in large radius.}
\createspell{\SLbless}{Bless}
    {Nearby allies gain \plus2 to attack.}
\createspell{\SLblindnessdeafness}{Blindness/Deafness}
    {Creature is blinded or deafened.}
\createspell{\SLblink}{Blink}
    {You randomly vanish and reappear for short duration.}
\createspell{\SLblur}{Blur}
    {Subject is harder to see.}
\createspell{\SLburninghands}{Burning Hands}
    {1d6 fire damage in medium cone.}
\createspell{\SLburninghands}{Burning Hands}
    {1d6 fire damage in small cone.}
\createspell{\SLcalllightninggreater}{Call Lightning}[Greater]
    {As \spell{call lightning}, but 5d8 damage per bolt.}
\createspell{\SLcalllightninggreater}{Call Lightning}[Greater]
    {As call lightning, but 5d8 damage per bolt.}
\createspell{\SLcalllightning}{Call Lightning}
    {Calls down lightning bolts (3d8 per bolt) from sky.}
\createspell{\SLcalmemotions}{Calm Emotions}
    {Calms creatures, negating emotion effects.}
\createspell{\SLchainlightning}{Chain Lightning}
    {5d10 damage; secondary targets take half damage.}
\createspell{\SLchangestaff}{Changestaff}
    {Your staff becomes a treant on command.}
\createspell{\SLchaoshammer}{Chaos Hammer}
    {Bewilders, deals 8d6 damage to nonchaotic creature.}
\createspell{\SLcharmmonster}{Charm Monster}
    {Makes monster believe it is your ally.}
\createspell{\SLcharmpersonmass}{Charm Person}[Mass]
    {As \spell{charm person}, but affects multiple creatures.}
\createspell{\SLcharmperson}{Charm Person}
    {Makes one person your friend.}
\createspell{\SLclenchedfist}{Clenched Fist}
    {Large hand provides cover or attacks your foes.}
\createspell{\SLcloakofchaos}{Cloak of Chaos}
    {Protect multiple creatures against lawful foes. }
\createspell{\SLcloakofchaos}{Cloak of Chaos}
    {Protect multiple creatures against lawful foes.}
\createspell{\SLcloudkill}{Cloudkill}
    {Fog deals Con damage each round.}
\createspell{\SLcolorspray}{Color Spray}
    {Creatures in medium cone are bewildered.}
\createspell{\SLcombattransformation}{Combat Transformation}
    {You gain combat bonuses.}
\createspell{\SLcommandmass}{Command}[Mass]
    {As \spell{command}, but affects multiple creatures.}
\createspell{\SLcommand}{Command}
    {One subject obeys selected command for 1 round.}
\createspell{\SLconeofcoldgreater}{Cone of Cold}[Greater]
    {8d6 cold damage and fatigue in large cone.}
\createspell{\SLconeofcold}{Cone of Cold}
    {5d6 cold damage and fatigue in medium cone.}
\createspell{\SLconfusionmass}{Confusion}[Mass]
    {Multiple creatures act randomly.}
\createspell{\SLconfusion}{Confusion}
    {Subject acts randomly.}
\createspell{\SLcontrolwater}{Control Water}
    {Raises or lowers bodies of water.}
\createspell{\SLcreatesound}{Create Sound}
    {Figment sounds.}
\createspell{\SLcreepingdoom}{Creeping Doom}
    {Swarms of centipedes attack at your command.}
\createspell{\SLcripple}{Cripple}
    {Subject cannot use its limbs.}
\createspell{\SLcrushingdespair}{Crushing Despair}
    {Subjects in medium cone are demoralized.}
\createspell{\SLcrushinghand}{Crushing Hand}
    {Large hand provides cover or crushes your foes.}
\createspell{\SLcurecriticalwoundsmass}{Cure Critical Wounds}[Mass]
    {Cures 8d6 damage for many creatures.}
\createspell{\SLcurecriticalwounds}{Cure Critical Wounds}
    {Cures 6d6 damage.}
\createspell{\SLcurecriticalwounds}{Cure Critical Wounds}
    {Cures 6d8 damage.}
\createspell{\SLcurelightwoundsmass}{Cure Light Wounds}[Mass]
    {Cures 5d6 damage for many creatures.}
\createspell{\SLcurelightwounds}{Cure Light Wounds}
    {Cures 2d6 damage.}
\createspell{\SLcuremoderatewoundsmass}{Cure Moderate Wounds}[Mass]
    {Cures 6d6 damage for many creatures.}
\createspell{\SLcuremoderatewounds}{Cure Moderate Wounds}
    {Cures 4d6 damage.}
\createspell{\SLcureseriouswoundsmass}{Cure Serious Wounds}[Mass]
    {Cures 7d6 damage for many creatures.}
\createspell{\SLcureseriouswounds}{Cure Serious Wounds}
    {Cures 6d6 damage.}
\createspell{\SLdancinglights}{Dancing Lights}
    {Creates torches or other lights.}
\createspell{\SLdarkvision}{Darkvision}
    {See 60 ft. in total darkness.}
\createspell{\SLdaylight}{Daylight}
    {Large radius of bright light.}
\createspell{\SLdeathknell}{Death Knell}
    {Gain temporary HP when subject dies.}
\createspell{\SLdeathknell}{Death Knell}
    {Kill dying creature and gain temporary hp, more for killing powerful creatures.}
\createspell{\SLdeathwardmass}{Death Ward}[Mass]
    {As \spell{death ward}, but affects multiple creatures.}
\createspell{\SLdeathward}{Death Ward}
    {Grants immunity to death spells and negative energy effects.}
\createspell{\SLdeepslumber}{Deep Slumber}
    {Subject falls asleep for a long time.}
\createspell{\SLdeflection}{Deflection}
    {Gain \plus2 to physical defenses.}
\createspell{\SLdelaydamage}{Delay Damage}
    {Take half damage now and half later. }
\createspell{\SLdelayedblastfireball}{Delayed Blast Fireball}
    {6d6 fire damage in medium radius; you can postpone blast for 5 rounds.}
\createspell{\SLdelaypoison}{Delay Poison}
    {Quickly stops poison from harming subject.}
\createspell{\SLdestruction}{Destruction}
    {Kills subject and destroys remains.}
\createspell{\SLdetectalignment}{Detect Alignment}
    {Reveals creatures, spells, or objects of selected alignment.}
\createspell{\SLdictum}{Dictum}
    {Kills, paralyzes, stuns, or staggers nonlawful subjects in 40 ft. cube.}
\createspell{\SLdimensionalanchor}{Dimensional Anchor}
    {Bars extradimensional movement.}
\createspell{\SLdimensiondoormass}{Dimension Door}[Mass]
    {Teleport you and multiple allies within 1,000 feet.}
\createspell{\SLdimensiondoor}{Dimension Door}
    {Teleports you within 1,000 feet.}
\createspell{\SLdimensionslide}{Dimension Slide}
    {Teleports nearby creature short distance.}
\createspell{\SLdiscernlies}{Discern Lies}
    {Reveals deliberate falsehoods.}
\createspell{\SLdiscernvulnerability}{Discern Vulnerability}
    {Quickly find weaknesses in foe's defenses.}
\createspell{\SLdiscordantsong}{Discordant Song}
    {Confuse creatures with sustained music.}
\createspell{\SLdisintegrate}{Disintegrate}
    {Deals 12d8 damage, leaving only dust.}
\createspell{\SLdismissal}{Dismissal}
    {Forces a creature to return to native plane.}
\createspell{\SLdispelmagicgreater}{Dispel Magic}[Greater]
    {As dispel magic, but up to \plus20 on check.}
\createspell{\SLdispelmagic}{Dispel Magic}
    {Cancels magical spells and effects.}
\createspell{\SLdispelmagic}{Dispel Magic}
    {Cancels spells and magical effects.}
\createspell{\SLdisplacement}{Displacement}
    {Attacks miss subject 50\% of the time.}
\createspell{\SLdisruptingweapon}{Disrupting Weapon}
    {Melee weapon destroys undead.}
\createspell{\SLdivinefavorgreater}{Divine Favor}[Greater]
    {You gain \plus2 Str and attack and damage bonus.}
\createspell{\SLdivinefavor}{Divine Favor}
    {You gain \plus2 on attack and damage rolls.}
\createspell{\SLdominatemonster}{Dominate Monster}
    {As dominate person, but any creature.}
\createspell{\SLdominateperson}{Dominate Person}
    {Controls humanoid telepathically.}
\createspell{\SLdrainlife}{Drain Life}
    {Deals 10d6 damage and steals life.}
\createspell{\SLearthenblade}{Earthen Blade}
    {Create magical weapon from the ground.}
\createspell{\SLearthglide}{Earth Glide}
    {Creature glides through earth. }
\createspell{\SLearthquake}{Earthquake}
    {Intense tremor shakes large radius.}
\createspell{\SLearthspikemass}{Earthspike}[Mass]
    {Spikes from the ground impale multiple targets.}
\createspell{\SLearthspike}{Earthspike}
    {Spike from the ground deals damage. }
\createspell{\SLearthspike}{Earthspike}
    {Spike from the ground impales target.}
\createspell{\SLearthspike}{Earthspike}
    {[Mass] Spikes from the ground deal damage to multiple targets.}
\createspell{\SLearthspull}{Earth's Pull}
    {Subject feels gravity more strongly.}
\createspell{\SLelementalswarm}{Elemental Swarm}
    {Summons multiple elementals.}
\createspell{\SLenergyconversion}{Energy Conversion}
    {As \spell{greater resist energy}, plus you can fire rays of absorbed energy.}
\createspell{\SLenergydrain}{Energy Drain}
    {Subject gains six negative levels.}
\createspell{\SLenervation}{Enervation}
    {Inflict three negative levels.}
\createspell{\SLenfeeblement}{Enfeeblement}
    {Inflicts \minus4 penalty to physical attribute.}
\createspell{\SLenfeeblement}{Enfeeblement}
    {Target takes \minus4 to physical attribute.}
\createspell{\SLenlargepersonmass}{Enlarge Person}[Mass]
    {Multiple humanoid creatures double in size.}
\createspell{\SLenlargeperson}{Enlarge Person}
    {Humanoid creature doubles in size.}
\createspell{\SLentangle}{Entangle}
    {Plants entangle everyone in small spread.}
\createspell{\SLentanglinggrowth}{Entangling Growth}
    {New plants grow and entangle everyone in medium spread.}
\createspell{\SLentropicshield}{Entropic Shield}
    {Ranged attacks against subject have 50\% miss chance.}
\createspell{\SLetherealjaunt}{Ethereal Jaunt}
    {You become ethereal for short duration.}
\createspell{\SLetherealness}{Etherealness}
    {Travel to Ethereal Plane with companions.}
\createspell{\SLexpeditiousretreat}{Expeditious Retreat}
    {Your speed doubles.}
\createspell{\SLfaeriefire}{Faerie Fire}
    {Outlines subjects in small radius with light, canceling concealment, invisibility, and the like.}
\createspell{\SLfalsereality}{False Reality}
    {As \spell{persistent image}, but affects massive area.}
\createspell{\SLfear}{Fear}
    {Subjects within medium cone flee for short duration.}
\createspell{\SLfeatherfall}{Feather Fall}
    {Objects or creatures fall slowly.}
\createspell{\SLfeeblemind}{Feeblemind}
    {Subject's Intelligence drops to \minus9.}
\createspell{\SLfingerofdeath}{Finger of Death}
    {Kills one subject.}
\createspell{\SLfireball}{Fireball}
    {Small radius of fire deals 3d6 damage.}
\createspell{\SLfireseeds}{Fire Seeds}
    {Acorns and berries become grenades and bombs.}
\createspell{\SLfireshield}{Fire Shield}
    {Creatures attacking you take damage; you're protected from cold.}
\createspell{\SLfireshield}{Fire Shield}
    {Creatures attacking you take damage; you're protected from heat or cold.  \spellhead{Resist Energy}[Greater] As \spell{resist energy}, but all energy types.}
\createspell{\SLfireshield}{Fire Shield}
    {Creatures attacking you take damage; you're protected from heat or cold.}
\createspell{\SLfirestorm}{Fire Storm}
    {Deals 8d6 fire damage to enemies in large spread.}
\createspell{\SLflameblade}{Flame Blade}
    {Wield fire as scimitar.}
\createspell{\SLflamestrike}{Flame Strike}
    {Smite foes with divine fire for 5d6 damage.}
\createspell{\SLfly}{Fly}
    {Subject flies at speed of 30 ft.}
\createspell{\SLfogcloud}{Fog Cloud}
    {Medium cylinder of fog obscures vision.}
\createspell{\SLforcecage}{Forcecage}
    {Cube or cage of force imprisons all inside.}
\createspell{\SLforeseeprobability}{Foresee Probability}
    {Know creature's next die results.}
\createspell{\SLforesightgreater}{Foresight}[Greater]
    {Powerful ``sixth sense'' warns of impending danger.}
\createspell{\SLforesight}{Foresight}
    {Powerful ``sixth sense'' warns of impending danger.}
\createspell{\SLforesight}{Foresight}
    {``Sixth sense'' warns of impending danger.}
\createspell{\SLforget}{Forget}
    {A creature forgets something.}
\createspell{\SLfreedommass}{Freedom}[Mass]
    {As \spell{freedom}, but affects multiple creatures.}
\createspell{\SLfreedom}{Freedom}
    {Subject moves normally despite impediments.}
\createspell{\SLfreezingsphere}{Freezing Sphere}
    {Freezes water or deals cold damage.}
\createspell{\SLgaseousform}{Gaseous Form}
    {Subject becomes insubstantial and can fly slowly.}
\createspell{\SLgate}{Gate}
    {Connects two planes for travel.}
\createspell{\SLgentledescent}{Gentle Descent}
    {Subject can glide.}
\createspell{\SLghoultouch}{Ghoul Touch}
    {Sicken or nauseate creauture.}
\createspell{\SLgiantvermin}{Giant Vermin}
    {Turns centipedes, scorpions, or spiders into giant vermin.}
\createspell{\SLglitterdust}{Glitterdust}
    {Outlines concealed and invisible creatures in small radius.}
\createspell{\SLgraspinghand}{Grasping Hand}
    {Hand provides cover, pushes, or grapples.}
\createspell{\SLgrease}{Grease}
    {Makes 10 ft. square or one object slippery.}
\createspell{\SLgreateracidarrow}{Greater Acid Arrow}
    {Fling acid to deal lingering damage and stagger}
\createspell{\SLgustofwind}{Gust of Wind}
    {Blows away or knocks down smaller creatures in large line.}
\createspell{\SLharm}{Harm}
    {Touch deals 12d8 damage and 4 Con damage.}
\createspell{\SLhastemass}{Haste}[Mass]
    {Many subjects move faster, get extra attack.}
\createspell{\SLhaste}{Haste}
    {Subject moves faster, gets extra attack.}
\createspell{\SLheal}{Heal}
    {Cures 12d8 damage, all diseases and mental conditions.}
\createspell{\SLheatmetal}{Heat Metal}
    {Make metal so hot it damages those who touch it.}
\createspell{\SLheroismgreater}{Heroism}[Greater]
    {Gives combat bonuses, temporary hit points.}
\createspell{\SLheroism}{Heroism}
    {Gives \plus2 on attacks, checks, and saves.}
\createspell{\SLholdmonster}{Hold Monster}
    {As \spell{hold person}, but any creature.}
\createspell{\SLholdmonster}{Hold Monster}
    {As hold person, but any creature.}
\createspell{\SLholdmonster}{Hold Monster}
    {[Mass] As \spell{hold monster}, but affects multiple creatures.}
\createspell{\SLholdpersonmass}{Hold Person}[Mass]
    {As \spell{hold person}, but affects multiple creatures.}
\createspell{\SLholdperson}{Hold Person}
    {Paralyzes one humanoid for short duration.}
\createspell{\SLholyaura}{Holy Aura}
    {Protect multiple creatures against evil foes. }
\createspell{\SLholysmite}{Holy Smite}
    {Blinds and deals 8d6 damage to nongood creature.}
\createspell{\SLholyword}{Holy Word}
    {Kills, paralyzes, blinds, or deafens nongood subjects in large radius.}
\createspell{\SLholyword}{Holy Word}
    {Kills, paralyzes, slows, or deafens nongood subjects in large radius.}
\createspell{\SLhorridwilting}{Horrid Wilting}
    {Deals 8d6 damage to many foes.}
\createspell{\SLhypnoticpattern}{Hypnotic Pattern}
    {Fascinates creatures in small radius.}
\createspell{\SLicestorm}{Ice Storm}
    {Hail deals 3d4 damage in small cylinder.}
\createspell{\SLicestorm}{Ice Storm}
    {Hail deals 4d4 damage in small cylinder.}
\createspell{\SLimplosion}{Implosion}
    {Kills one creature/round.}
\createspell{\SLimprisonment}{Imprisonment}
    {Entombs subject beneath the earth.}
\createspell{\SLinertialshield}{Inertial Shield}
    {Subject gains damage reduction.}
\createspell{\SLinflictcriticalwoundsmass}{Inflict Critical Wounds}[Mass]
    {Deals 8d6 damage to many creatures.}
\createspell{\SLinflictcriticalwounds}{Inflict Critical Wounds}
    {Deals 8d6 damage.}
\createspell{\SLinflictcriticalwounds}{Inflict Critical Wounds}
    {[Mass] Deals 8d6 damage to many creatures.}
\createspell{\SLinflictlightwoundsmass}{Inflict Light Wounds}[Mass]
    {Deals 5d6 damage to many creatures.}
\createspell{\SLinflictlightwounds}{Inflict Light Wounds}
    {Deals 2d6 damage.}
\createspell{\SLinflictlightwounds}{Inflict Light Wounds}
    {[Mass] Deals 5d6 damage to many creatures.}
\createspell{\SLinflictmoderatewoundsmass}{Inflict Moderate Wounds}[Mass]
    {Deals 6d6 damage to many creatures.}
\createspell{\SLinflictmoderatewounds}{Inflict Moderate Wounds}
    {Deals 4d6 damage.}
\createspell{\SLinflictseriouswoundsmass}{Inflict Serious Wounds}[Mass]
    {Deals 7d6 damage to many creatures.}
\createspell{\SLinflictseriouswounds}{Inflict Serious Wounds}
    {Deals 6d6 damage.}
\createspell{\SLinsanity}{Insanity}
    {Subject indefinitely acts randomly.}
\createspell{\SLinterposinghand}{Interposing Hand}
    {Hand provides cover against one opponent at a time.}
\createspell{\SLinvestmagic}{Invest Magic}
    {All of subject's equipment is magical.}
\createspell{\SLinvisibilitygreater}{Invisibility}[Greater]
    {As \spell{invisibility}, but subject can attack and stay invisible.}
\createspell{\SLinvisibilitymass}{Invisibility}[Mass]
    {As \spell{invisibility}, but affects multiple creatures.}
\createspell{\SLinvisibility}{Invisibility}
    {Subject invisible until it attacks.}
\createspell{\SLinvisibility}{Invisibility}
    {Subject is invisible until it attacks.}
\createspell{\SLironbody}{Iron Body}
    {Your body becomes living iron.}
\createspell{\SLirresistibledance}{Irresistible Dance}
    {Forces subject to dance.}
\createspell{\SLknock}{Knock}
    {Opens locked or magically sealed door.}
\createspell{\SLlevitate}{Levitate}
    {Subject moves up and down at your direction.}
\createspell{\SLlifebreakercurse}{Lifebreaker Curse}
    {Target takes permanent damage.}
\createspell{\SLlifeseekingmissile}{Lifeseeking Missile}
    {Homing missiles deal 3d10 damage.}
\createspell{\SLlifeseekingmissile}{Lifeseeking Missile}
    {Homing missiles unerringly deal 3d10 damage.}
\createspell{\SLlightningbolt}{Lightning Bolt}
    {Large line of electricity deals 3d6 damage.}
\createspell{\SLlightningbolt}{Lightning Bolt}
    {Large line of electricity deals 3d8 damage.}
\createspell{\SLlinkvitality}{Link Vitality}
    {Two targets share each others' damage and healing.}
\createspell{\SLlocatecreaturegreater}{Locate Creature}[Greater]
    {Senses direction toward distant visualized creatures.}
\createspell{\SLlocatecreature}{Locate Creature}
    {Senses direction toward visualized creature.}
\createspell{\SLlocatecreature}{Locate Creature}
    {[Greater] Senses direction toward distant visualized creatures.}
\createspell{\SLlocateentity}{Locate Entity}
    {Locates creatures or objects within 1 mile.}
\createspell{\SLlocateobjectgreater}{Locate Object}[Greater]
    {Senses direction toward distant visualized objects.}
\createspell{\SLlocateobjectlesser}{Locate Object}[Lesser]
    {Senses direction toward visualized object.}
\createspell{\SLlocateobject}{Locate Object}
    {Senses direction toward visualized object.}
\createspell{\SLlongstrider}{Longstrider}
    {Increases your speed.}
\createspell{\SLlongstrider}{Longstrider}
    {Your speed increases by 10 ft.}
\createspell{\SLmagearmor}{Mage Armor}
    {Armor made of magical force gives \plus2 AC.}
\createspell{\SLmagehand}{Mage Hand}
    {5-pound telekinesis.}
\createspell{\SLmagesdisjunction}{Mage's Disjunction}
    {Dispels all magic.}
\createspell{\SLmagiccircleagainstchaos}{Magic Circle against Chaos}
    {Protect allies in a \areamed radius and hedge out opposing outsiders.}
\createspell{\SLmagiccircleagainstevil}{Magic Circle against Evil}
    {Protect allies in a \areamed radius and hedge out opposing outsiders.}
\createspell{\SLmagiccircleagainstgood}{Magic Circle against Good}
    {Protect allies in a \areamed radius and hedge out opposing outsiders.}
\createspell{\SLmagicfanggreater}{Magic Fang}[Greater]
    {All subject's natural weapons get \plus2 bonus.}
\createspell{\SLmagicfang}{Magic Fang}
    {One natural weapon of subject creature becomes \plus2 weapon.}
\createspell{\SLmagicmissile}{Magic Missile}
    {Force missiles unerringly deal 2d4 damage.}
\createspell{\SLmagicvestment}{Magic Vestment}
    {Armor or shield becomes \plus2.}
\createspell{\SLmagicweapon}{Magic Weapon}
    {Weapon becomes \plus2.}
\createspell{\SLmajorimage}{Major Image}
    {As silent image, plus sound, smell and thermal effects.}
\createspell{\SLmanipulateprobability}{Manipulate Probability}
    {Control creature's next die results.}
\createspell{\SLmaze}{Maze}
    {Traps subject in extradimensional maze.}
\createspell{\SLmeldintostone}{Meld into Stone}
    {You and your gear merge with stone.}
\createspell{\SLmeteorswarm}{Meteor Swarm}
    {Massive hail of meteors deals 9d8 damage.}
\createspell{\SLminorimage}{Minor Image}
    {As \spell{silent image}, plus some sound.}
\createspell{\SLmirrorimagegreater}{Mirror Image}[Greater]
    {Creates regenerating decoy duplicates of you.}
\createspell{\SLmirrorimage}{Mirror Image}
    {Creates decoy duplicates of you.}
\createspell{\SLmislead}{Mislead}
    {Turns you invisible and creates illusory double.}
\createspell{\SLmissilestorm}{Missile Storm}
    {Swarm of missiles strike many creatures.}
\createspell{\SLmomentofpresciencegreater}{Moment of Prescience}[Greater]
    {Roll twice with large bonus for single attack, save, or opposed check.}
\createspell{\SLmomentofpresciencelesser}{Moment of Prescience}[Lesser]
    {Roll twice for single attack, save, or opposed check.}
\createspell{\SLmomentofprescience}{Moment of Prescience}
    {Roll twice with bonus for single attack, save, or opposed check.}
\createspell{\SLmomentofprescience}{Moment of Prescience}
    {[Lesser] Roll twice for single attack, save, or opposed check.}
\createspell{\SLobscuringmist}{Obscuring Mist}
    {Fog surrounds you.}
\createspell{\SLorderswrath}{Order's Wrath}
    {Bewilders and deals 8d6 damage to nonlawful creature.}
\createspell{\SLpersistentimage}{Persistent Image}
    {As \spell{major image}, but no concentration required.}
\createspell{\SLphantasmalkiller}{Phantasmal Killer}
    {Fearsome illusion can frighten target to death.}
\createspell{\SLphantasmalmaze}{Phantasmal Maze}
    {Subject acts as if affected by \spell{maze}.}
\createspell{\SLphantasmalwound}{Phantasmal Wound}
    {Subject believes it is grievously wounded.}
\createspell{\SLphantasmalwound}{Phantasmal Wound}
    {Subject believes it is near death.}
\createspell{\SLphasedoor}{Phase Door}
    {Creates an invisible passage through wood or stone.}
\createspell{\SLpoison}{Poison}
    {Touch deals 1d6 Con damage repeatedly.}
\createspell{\SLpolarray}{Polar Ray}
    {Ranged touch attack slows and deals 16d6 cold damage.}
\createspell{\SLpowerwordblind}{Power Word Blind}
    {Blinds bloodied creature.}
\createspell{\SLpowerwordconfuse}{Power Word Confuse}
    {Bloodied subject acts randomly.}
\createspell{\SLpowerwordkill}{Power Word Kill}
    {Kills bloodied creature.}
\createspell{\SLpowerwordkill}{Power Word Kill}
    {Kills one bloodied creature.}
\createspell{\SLpowerwordstun}{Power Word Stun}
    {Stuns a creature.}
\createspell{\SLpowerwordstun}{Power Word Stun}
    {Stuns bloodied creature.}
\createspell{\SLprecognitiongreater}{Precognition}[Greater]
    {See into the future to gain combat bonuses and extra attack.}
\createspell{\SLprecognitionlesser}{Precognition}[Lesser]
    {See into the future to gain attack and damage bonus.}
\createspell{\SLprecognition}{Precognition}
    {See into the future to gain combat bonuses.}
\createspell{\SLprismaticbeam}{Prismatic Beam}
    {Beam of multicolored light has random effects.}
\createspell{\SLprismaticspray}{Prismatic Spray}
    {Cone of multicolored light has random effects.}
\createspell{\SLprismaticspray}{Prismatic Spray}
    {Rays hit subjects with a variety of effects.}
\createspell{\SLprismaticstorm}{Prismatic Storm}
    {Sphere of multicolored light has random effects.}
\createspell{\SLprismaticwall}{Prismatic Wall}
    {Multicolored wall has random effects.}
\createspell{\SLprismaticwall}{Prismatic Wall}
    {}
\createspell{\SLprohibitiongreater}{Prohibition}[Greater]
    {Forcibly prevent creatures from taking specific action.}
\createspell{\SLprohibition}{Prohibition}
    {Punish creatures that take specific action.}
\createspell{\SLprohibition}{Prohibition}
    {[Greater] Forcibly prevent creatures from taking specific action.}
\createspell{\SLprojectimage}{Project Image}
    {Illusory double can talk and cast spells.}
\createspell{\SLprotectionfromalignment}{Protection from Alignment}
    {\plus2 to special defenses, \plus4 against chosen alignment.}
\createspell{\SLprotectionfromenergygreater}{Protection from Energy}[Greater]
    {As \spell{protection from energy}, but against all energy types.}
\createspell{\SLprotectionfromenergy}{Protection from Energy}
    {Reduce damage from one kind of energy.}
\createspell{\SLreadmindgreater}{Read Mind}[Greater]
    {Unavoidably learn a creature's surface thoughts.}
\createspell{\SLreadmindmass}{Read Mind}[Mass]
    {Learn surface thoughts from multiple creatures.}
\createspell{\SLreadmind}{Read Mind}
    {Learn a creature's surface thoughts.}
\createspell{\SLreducepersonmass}{Reduce Person}[Mass]
    {Reduces several creatures.}
\createspell{\SLreduceperson}{Reduce Person}
    {Humanoid creature halves in size.}
\createspell{\SLregenerate}{Regenerate}
    {Touched creature heals 10 hit points/round.}
\createspell{\SLrepelmetalorstone}{Repel Metal or Stone}
    {Pushes away metal and stone in large line.}
\createspell{\SLrepulsion}{Repulsion}
    {Creatures can't approach you.}
\createspell{\SLresilientsphere}{Resilient Sphere}
    {Force globe protects but traps one creature or object.}
\createspell{\SLresistenergygreater}{Resist Energy}[Greater]
    {As \spell{resist energy}, but all energy types.}
\createspell{\SLresistenergy}{Resist Energy}
    {Ignores 10 (or more) points of damage/attack from specified energy type.}
\createspell{\SLresistenergy}{Resist Energy}
    {Ignores first 10 (or more) points of damage/attack from specified energy type.}
\createspell{\SLretributivebrilliance}{Retributive Brilliance}
    {Subject can dazzle or blind attacker once.}
\createspell{\SLretributiveshield}{Retributive Shield}
    {Grant DR 8 that reflects damage at foes.}
\createspell{\SLretributiveshield}{Retributive Shield}
    {Grants DR 8 that reflects damage at foes.}
\createspell{\SLretrieveally}{Retrieve Ally}
    {Teleport willing creature next to you.}
\createspell{\SLretrievegreater}{Retrieve}[Greater]
    {Teleport an item you can see to your hand.}
\createspell{\SLretrieveobject}{Retrieve Object}
    {Teleport an unattended item you can see to your hand.}
\createspell{\SLrevelation}{Revelation}
    {Grant target vision of one of three futures.}
\createspell{\SLreversegravity}{Reverse Gravity}
    {Objects and creatures fall upward.}
\createspell{\SLrevivify}{Revivify}
    {Restores recently dead body to life.}
\createspell{\SLrighteousmight}{Righteous Might}
    {Your size increases, and you gain combat bonuses.}
\createspell{\SLsanctuary}{Sanctuary}
    {Opponents can't attack subject until it attacks.}
\createspell{\SLscintillatingpattern}{Scintillating Pattern}
    {Twisting colors bewilder foes.}
\createspell{\SLscorchingray}{Scorching Ray}
    {Ray (or rays) deal 4d6 fire damage.}
\createspell{\SLseaoffog}{Sea of Fog}
    {200 ft. radius of fog obscures vision.}
\createspell{\SLseaoffog}{Sea of Fog}
    {500 ft. cylinder of fog obscures vision.}
\createspell{\SLsearinglight}{Searing Light}
    {Ray deals 6d6 damage and blinds, more against undead.}
\createspell{\SLseeinvisibility}{See Invisibility}
    {Reveals invisible creatures or objects.}
\createspell{\SLshadowbody}{Shadow Body}
    {You become a living shadow.}
\createspell{\SLshadowpuppet}{Shadow Puppet}
    {Step into Plane of Shadow and create shadow duplicate in your stead.}
\createspell{\SLshadowumbra}{Shadow Umbra}
    {Shadowy aura has a 50\% chance to absorb attacks. }
\createspell{\SLsharepain}{Share Pain}
    {Split damage with willing subject.}
\createspell{\SLshieldoffaith}{Shield of Faith}
    {Grants \plus2 or higher shield bonus.}
\createspell{\SLshieldoflaw}{Shield of Law}
    {Protect multiple creatures against chaotic foes. }
\createspell{\SLshieldoflaw}{Shield of Law}
    {Protect multiple creatures against lawful foes.}
\createspell{\SLshillelagh}{Shillelagh}
    {Cudgel or quarterstaff becomes \plus2 weapon.}
\createspell{\SLshockinggrasp}{Shocking Grasp}
    {Touch deals 2d6 electricity damage.}
\createspell{\SLshockinggrasp}{Shocking Grasp}
    {Touch delivers 2d6 electricity damage.}
\createspell{\SLshoutgreater}{Shout}[Greater]
    {Yell deals 7d6 sonic damage in large cone; stuns creatures, damages objects.}
\createspell{\SLshout}{Shout}
    {Deafens all within medium cone and deals 4d6 sonic damage.}
\createspell{\SLshrinkitem}{Shrink Item}
    {Object shrinks to one-sixteenth size.}
\createspell{\SLsilence}{Silence}
    {Subject cannot make sound.}
\createspell{\SLsilentimage}{Silent Image}
    {Creates minor illusion of your design.}
\createspell{\SLskysmite}{Skysmite}
    {Unerring lightning deals 12d6 damage.}
\createspell{\SLskysmite}{Skysmite}
    {Unerring lightning deals 6d6 damage.}
\createspell{\SLsleepmass}{Sleep}[Mass]
    {Tire multiple creatures, possibly putting them to sleep.}
\createspell{\SLsleep}{Sleep}
    {Tire a creature, possibly putting it to sleep.}
\createspell{\SLslowmass}{Slow}[Mass]
    {Many subjects take only one action/round.}
\createspell{\SLslow}{Slow}
    {Subject takes only one action/round.}
\createspell{\SLsolidfog}{Solid Fog}
    {Fog blocks vision and slows movement.}
\createspell{\SLsoundburst}{Sound Burst}
    {2d6 sonic damage in a small radius.}
\createspell{\SLspellmessage}{Message}
    {Whispered conversation at distance.}
\createspell{\SLspellresistance}{Spell Resistance}
    {Subject gains spell resistance.}
\createspell{\SLspelltheftgreater}{Spelltheft}[Greater]
    {As \spell{greater dispel magic}, but you gain the effects of dispelled spells.}
\createspell{\SLspelltheft}{Spelltheft}
    {As \spell{dispel magic}, but you gain the effects of dispelled spells.}
\createspell{\SLspellturning}{Spell Turning}
    {Reflect 1d4\plus6 spell levels back at caster.}
\createspell{\SLspiderclimb}{Spider Climb}
    {Grants ability to walk on walls and ceilings.}
\createspell{\SLspiritualweapon}{Spiritual Weapon}
    {Magic weapon attacks on its own.}
\createspell{\SLspiritualweapon}{Spiritual Weapon}
    {Magical weapon attacks on its own.}
\createspell{\SLstampede}{Stampede}
    {Summon bison to trample foes.}
\createspell{\SLstinkingcloud}{Stinking Cloud}
    {Fog nauseates creatures inside.}
\createspell{\SLstoneskin}{Stoneskin}
    {\plus3 AC, ignore 8 damage per round.}
\createspell{\SLstormlord}{Stormlord}
    {Air protects you from attacks. }
\createspell{\SLstormofvengeance}{Storm of Vengeance}
    {Storm rains acid, lightning, and hail.}
\createspell{\SLstriptheflesh}{Strip the Flesh}
    {Rend foe's skin from its body.}
\createspell{\SLsuggestionmass}{Suggestion}[Mass]
    {As \spell{suggestion}, but shorter and affects multiple creatures.}
\createspell{\SLsuggestion}{Suggestion}
    {Compels subject to follow stated course of action.}
\createspell{\SLsummonmonsteriii}{Summon Monster III}
    {Calls extraplanar creature to fight for you.}
\createspell{\SLsummonmonsterii}{Summon Monster II}
    {Calls extraplanar creature to fight for you.}
\createspell{\SLsummonmonsteriv}{Summon Monster IV}
    {Calls extraplanar creature to fight for you.}
\createspell{\SLsummonmonsterix}{Summon Monster IX}
    {Calls extraplanar creature to fight for you.}
\createspell{\SLsummonmonsteri}{Summon Monster I}
    {Calls extraplanar creature to fight for you.}
\createspell{\SLsummonmonsterviii}{Summon Monster VIII}
    {Calls extraplanar creature to fight for you.}
\createspell{\SLsummonmonstervii}{Summon Monster VII}
    {Calls extraplanar creature to fight for you.}
\createspell{\SLsummonmonstervi}{Summon Monster VI}
    {Calls extraplanar creature to fight for you.}
\createspell{\SLsummonmonsterv}{Summon Monster V}
    {Calls extraplanar creature to fight for you.}
\createspell{\SLsummonnaturesallyiii}{Summon Nature's Ally III}
    {Calls creature to fight.}
\createspell{\SLsummonnaturesallyii}{Summon Nature's Ally II}
    {Calls creature to fight.}
\createspell{\SLsummonnaturesallyiv}{Summon Nature's Ally IV}
    {Calls creature to fight.}
\createspell{\SLsummonnaturesallyix}{Summon Nature's Ally IX}
    {Calls creature to fight.}
\createspell{\SLsummonnaturesallyi}{Summon Nature's Ally I}
    {Calls creature to fight.}
\createspell{\SLsummonnaturesallyviii}{Summon Nature's Ally VIII}
    {Calls creature to fight.}
\createspell{\SLsummonnaturesallyvii}{Summon Nature's Ally VII}
    {Calls creature to fight.}
\createspell{\SLsummonnaturesallyvi}{Summon Nature's Ally VI}
    {Calls creature to fight.}
\createspell{\SLsummonnaturesallyv}{Summon Nature's Ally V}
    {Calls creature to fight.}
\createspell{\SLsummonnaturesarmy}{Summon Nature's Army}
    {Calls one creature/level to fight.}
\createspell{\SLsummonswarm}{Summon Swarm}
    {Summons swarm of bats, rats, or spiders.}
\createspell{\SLsunbeam}{Sunbeam}
    {Beam dazzles and deals 5d6 damage.}
\createspell{\SLsunburst}{Sunburst}
    {Blinds all within large radius, deals 8d8 damage.}
\createspell{\SLtelekinesis}{Telekinesis}
    {Moves object, attacks creature, or hurls object or creature.}
\createspell{\SLtelekineticforce}{Telekinetic Force}
    {Move things with your mind.}
\createspell{\SLtelekineticmaneuver}{Telekinetic Maneuver}
    {Telekinetically dirty trick, disarm, grapple, shove, trip your target.}
\createspell{\SLtelekineticthrust}{Telekinetic Thrust}
    {Hurl objects with the force of your mind.}
\createspell{\SLtelepathy}{Telepathy}
    {Grants ability to communicate mentally. }
\createspell{\SLtemporalstasis}{Temporal Stasis}
    {Puts subject into suspended animation.}
\createspell{\SLtimestop}{Time Stop}
    {You act freely for 1d3\plus1 rounds.}
\createspell{\SLtotemicmindgreater}{Totemic Mind}[Greater]
    {Subject gains \plus4 to Int, Wis, or Cha.}
\createspell{\SLtotemicmindmass}{Totemic Mind}[Mass]
    {As \spell{totemic mind}, but affects multiple creatures.}
\createspell{\SLtotemicmind}{Totemic Mind}
    {Subject gains \plus2 to Int, Wis, or Cha.}
\createspell{\SLtotemicpowergreater}{Totemic Power}[Greater]
    {Subject gains \plus4 to Str, Dex, or Con.}
\createspell{\SLtotemicpowermass}{Totemic Power}[Mass]
    {As \spell{totemic power}, but affects multiple creatures.}
\createspell{\SLtotemicpower}{Totemic Power}
    {Subject gains \plus2 to Str, Dex, or Con.}
\createspell{\SLtouchofidiocy}{Touch of Idiocy}
    {Subject takes \minus4 penalty to Int, Wis, and Cha.}
\createspell{\SLtransmuteanyobject}{Transmute Any Object}
    {Transforms objects into new forms.}
\createspell{\SLtransmutefleshandstone}{Transmute Flesh and Stone}
    {Turns subject creature into statue, or restores petrified creature.}
\createspell{\SLtreeshape}{Tree Shape}
    {You look exactly like a tree for extended duration.}
\createspell{\SLtremorsense}{Tremorsense}
    {``See'' through the ground by concentrating.}
\createspell{\SLtrueseeing}{True Seeing}
    {Lets you see all things as they really are.}
\createspell{\SLtrueseeing}{True Seeing}
    {\M Lets you see all things as they really are.}
\createspell{\SLtruestrike}{True Strike}
    {\plus20 on your next attack roll.}
\createspell{\SLunholyaura}{Unholy Aura}
    {Protect multiple creatures against good foes. }
\createspell{\SLunholyblight}{Unholy Blight}
    {Sickens and deals 8d6 damage to good creature.}
\createspell{\SLunlivingeyes}{Unliving Eyes}
    {Grants lifesight ability.}
\createspell{\SLunlivingheart}{Unliving Heart}
    {Gain 5 temporary HP, treated as undead.}
\createspell{\SLventriloquism}{Ventriloquism}
    {Throws voice for medium duration.}
\createspell{\SLwailofthebanshee}{Wail of the Banshee}
    {Kills many creatures.}
\createspell{\SLwailofthebanshee}{Wail of the Banshee}
    {Scream kills many creatures. }
\createspell{\SLwalloffire}{Wall of Fire}
    {Passing through wall deals 4d6 damage.}
\createspell{\SLwalloffire}{Wall of Fire}
    {Passing through wall deals 5d6 fire damage.}
\createspell{\SLwallofforce}{Wall of Force}
    {Wall is immune to damage.}
\createspell{\SLwallofstone}{Wall of Stone}
    {Creates a stone wall that can be shaped.}
\createspell{\SLwallofthorns}{Wall of Thorns}
    {Thorns damage anyone who tries to pass.}
\createspell{\SLwaterwalk}{Water Walk}
    {Subject treads on water as if solid.}
\createspell{\SLwavesofexhaustion}{Waves of Exhaustion}
    {Creatures in medium cone become exhausted.  }
\createspell{\SLwavesofexhaustion}{Waves of Exhaustion}
    {Creatures in medium cone become exhausted.}
\createspell{\SLwavesoffatigue}{Waves of Fatigue}
    {Creatures in large cone become fatigued.}
\createspell{\SLweb}{Web}
    {Fills medium radius spread with sticky spiderwebs.}
\createspell{\SLweird}{Weird}
    {As \spell{phantasmal killer}, but affects multiple creatures.}
\createspell{\SLweird}{Weird}
    {As \spell{phantasmal killer}, but affects multiple creatures}
\createspell{\SLwindstrikegreater}{Windstrike}[Greater]
    {Wind moves and bludgeons target for 10d6 damage.}
\createspell{\SLwindstrike}{Windstrike}
    {Wind bludgeons target for 4d6 damage.}
\createspell{\SLwordofchaos}{Word of Chaos}
    {Kills, confuses, stuns, or deafens nonchaotic subjects in large radius.}
\createspell{\SLwordofrecall}{Word of Recall}
    {Teleports you back to designated place.}
\createspell{\SLzephyrbladegreater}{Zephyr Blade}[Greater]
    {Melee weapon can strike from a distance for full damage.}
\createspell{\SLzephyrblade}{Zephyr Blade}
    {Melee weapon can strike from a short distance for half damage.}
