\section{Armor}

\subsection{Magic Armor and Shield Special Ability Descriptions}

A list of additional abilities for magic armor and shields is given here.

\begin{dtable}
\lcaption{Unusual Armor Special Abilities}
\begin{tabularx}{\columnwidth}{>{\lcol}X >{\lcol}X}
\thead{Special Ability} & \thead{Bonus} \\
Ghost touch & \plus1 bonus \\
Glamered & \plus1 bonus \\
Invulnerability & \plus3 bonus \\
Etherealness  & \plus5 bonus \\
\end{tabularx}
\end{dtable}

\begin{dtable}
\lcaption{Unusual Shield Special Abilities}
\begin{tabularx}{\columnwidth}{>{\lcol}X >{\lcol}X}
\thead{Special Ability} & \thead{Bonus} \\
Arrow catching & \plus1 bonus \\
Bashing & \plus1 bonus \\
Blinding & \plus1 bonus \\
Ghost touch & \plus1 bonus \\
Arrow deflection & \plus2 bonus \\
Animated & \plus4 bonus \\
Reflecting  & \plus5 bonus \\
\end{tabularx}
\end{dtable}

\armdescription{Animated}{4}{14th}{Strong transmutation}{Standard (command)} Upon command, an animated shield floats within 2 feet of the wielder, protecting her as if she were using it herself but freeing up both her hands. Only one shield can protect a character at a time. A character with an animated shield still takes any penalties associated with shield use, such as armor check penalty, arcane spell failure chance, and nonproficiency. In addition, a character using an animated shield suffers a \minus2 penalty to attack rolls as the shield gets in the way of the character's attacks.

\prereq{Craft Magic Arms and Armor, animate objects}

\armdescription{Arrow Catching}{1}{6th}{Faint abjuration}{\x{} and Standard (command)} A shield with this ability attracts ranged weapons to it. It grants a \plus1 enhancement bonus to the wielder's deflection modifier against ranged weapons because projectiles and thrown weapons veer toward it. Additionally, any projectile or thrown weapon aimed at a target within 5 feet of the shield's wearer diverts from its original target and targets the shield's bearer instead. (If the wielder has total cover relative to the attacker, the projectile or thrown weapon is not diverted.) Additionally, those attacking the wearer with ranged weapons ignore any miss chances that would normally apply. Projectiles and thrown weapons that have an enhancement bonus higher than the shield's base AC bonus are not diverted to the wearer (but the shield's increased AC bonus still applies against these weapons). The wielder can activate or deactivate this ability with a command word.

\prereq{Craft Magic Arms and Armor, entropic shield}

\armdescription{Arrow Deflection}{2}{8th}{Faint abjuration}{\x} This shield protects the wielder as if he had the Deflect Arrows feat. Once per round, when he would normally be struck by a ranged weapon, the shield deflects the weapon. He must be aware of the attack and not flat-footed. Attempting to deflect a ranged weapon doesn't count as an action. Exceptional ranged weapons, such as boulders hurled by giants or acid arrows, can't be deflected.

\prereq{Craft Magic Arms and Armor, \my{protection from arrows}}

\armdescription{Bashing}{1}{6th}{Faint transmutation}{\x} A shield with this special ability is designed to perform a shield bash. A bashing shield deals damage as if it were a weapon of two size categories larger (a Medium light shield thus deals 1d6 points of damage and a Medium heavy shield deals 1d8 points of damage). The shield acts as a \plus1 weapon when used to bash. (Only light and heavy shields can have this ability.)

\prereq{Craft Magic Arms and Armor, bull's strength}

\armdescriptiondc{Blinding}{1}{6th}{Faint evocation}{Standard (command)}{12} A shield with this ability flashes with a brilliant light up to twice per day upon command of the wielder. Anyone within 20 feet except the wielder must make a Reflex save or be blinded for 1d4 rounds.

\prereq{Craft Magic Arms and Armor, \my{daylight}}

\armdescription{Etherealness}{5}{16th}{Strong transmutation}{Standard (command)} On command, this ability allows the wearer of the armor to become ethereal (as the ethereal jaunt spell) once per day. The character can remain ethereal for \my{up to 1 minute}, and once he returns to normal, he cannot become ethereal again that day.

\prereq{Craft Magic Arms and Armor, ethereal jaunt}

\armdescription{Ghost Touch}{1}{6th}{Faint transmutation}{\x} This armor or shield seems almost translucent. Both its enhancement bonus and its armor bonus count against the attacks of incorporeal creatures. It can be picked up, moved, and worn by incorporeal creatures at any time. Incorporeal creatures gain the armor or shield's enhancement bonus against both corporeal and incorporeal attacks, and they can still pass freely through solid objects.

\prereq{Craft Magic Arms and Armor, \my{blink}}

\armdescription{Glamered}{1}{6th}{Faint illusion}{\x} A suit of armor with this ability appears normal. Upon command, the armor \my{appears to change} shape and form to assume the appearance of a normal set of clothing. The armor retains all its properties (including weight \my{and sound}) when glamered. Only a \spell{true seeing} spell or similar magic reveals the true nature of the armor when disguised.

\prereq{Craft Magic Arms and Armor, disguise self}

\armdescription{Invulnerability}{5}{12th}{Moderate abjuration}{\x} This suit of armor grants the wearer damage reduction of \my{5/{-}}.

\prereq{Craft Magic Arms and Armor, \my{inertial barrier}}

\armdescription{Reflecting}{5}{14th}{Strong abjuration}{Immediate (command)} This shield seems like a mirror. Its surface is completely reflective. Once per day, it can be called on to reflect a spell back at its caster exactly like the \spell{spell turning} spell.

\prereq{Craft Magic Arms and Armor, spell turning}

\section{Weapons}

A list of weapon special abilities is given here.

\begin{dtable}
\lcaption{Melee Weapon Special Abilities}
\begin{tabularx}{\columnwidth}{>{\lcol}X >{\lcol}X}
\thead{Special Ability} & \thead{Bonus} \\
Bane & \plus1 bonus \\
Energetic & \plus1 bonus \\
Keen\fn{1} & \plus1 bonus \\
Throwing & \plus1 bonus \\
Anarchic & \plus2 bonus \\
Axiomatic & \plus2 bonus \\
Disruption\fn{2} & \plus2 bonus \\
Energetic Burst & \plus2 bonus \\
Holy & \plus2 bonus \\
Unholy & \plus2 bonus \\
Speed & \plus4 bonus \\
\end{tabularx}
1 Piercing or slashing weapons only.
\end{dtable}

\begin{dtable}
\lcaption{Ranged Weapon Special Abilities}
\begin{tabularx}{\columnwidth}{>{\lcol}X >{\lcol}X}
\thead{Special Ability} & \thead{Bonus} \\
Bane & \plus1 bonus \\
Distance & \plus1 bonus \\
Energetic & \plus1 bonus \\
Returning & \plus1 bonus \\
Anarchic & \plus2 bonus \\
Axiomatic & \plus2 bonus \\
Energetic Burst & \plus2 bonus \\
Holy & \plus2 bonus \\
Returning, greater & \plus2 bonus \\
Unholy & \plus2 bonus \\
Speed & \plus4 bonus \\
\end{tabularx}
\end{dtable}

\begin{dtable}
\lcaption{Unusual Melee Weapon Special Abilities}
\begin{tabularx}{\columnwidth}{>{\lcol}X >{\lcol}X}
\thead{Special Ability} & \thead{Bonus} \\
Defending & \plus1 bonus \\
Ghost touch & \plus1 bonus \\
Ki Focus & \plus1 bonus \\
Merciful & \plus1 bonus \\
Thundering & \plus1 bonus \\
Disruption\fn{1} & \plus2 bonus \\
Mighty cleaving & \plus2 bonus \\
Spell storing & \plus2 bonus \\
Brilliant energy & \plus4 bonus \\
Dancing & \plus4 bonus \\
Spell storing, greater & \plus4 bonus \\
Vorpal\fn{2} & \plus5 bonus \\
Shielding & Variable bonus
\end{tabularx}
1 Bludgeoning weapons only. \\
2 Piercing or slashing weapons only.
\end{dtable}

\begin{dtable}
\lcaption{Unusual Ranged Weapon Special Abilities}
\begin{tabularx}{\columnwidth}{>{\lcol}X >{\lcol}X}
\thead{Special Ability} & \thead{Bonus} \\
Ghost touch & \plus1 bonus \\
Merciful & \plus1 bonus \\
Thundering & \plus1 bonus \\
Seeking & \plus2 bonus \\
Spell storing & \plus2 bonus \\
Wounding & \plus2 bonus \\
Brilliant energy & \plus4 bonus \\
Spell storing, greater & \plus4 bonus \\
\end{tabularx}
\end{dtable}

\subsection{Unusual Magic Weapon Special Ability Descriptions}

In addition to enhancement bonuses, weapons can have one or more of the special abilities detailed below. A weapon with a special ability must have at least a \plus1 enhancement bonus.

\armdescription{Anarchic}{2}{10th}{Moderate evocation [chaotic]}{\x} An anarchic weapon is infused with the power of chaos. This makes the weapon chaos-aligned and thus allows it to bypass the corresponding damage reduction. It deals double bonus damage against lawful foes. In addition, on a 20, the struck target is confused for 1 round if it is lawful. It bestows one negative level on any lawful creature attempting to wield it. The negative level remains as long as the weapon is in hand and disappears when the weapon is no longer wielded. This negative level cannot be overcome in any way (including \spell{restoration} rituals) while the weapon is wielded. Bows, crossbows, and slings so crafted bestow the chaotic power upon their ammunition.

\prereq{Craft Magic Arms and Armor, chaos hammer, creator must be chaotic}

\armdescription{Axiomatic}{2}{10th}{Moderate evocation [lawful]}{\x} An axiomatic weapon is infused with the power of law. This makes the weapon law-aligned and thus allows it to bypass the corresponding damage reduction. It deals double bonus damage against chaotic foes. In addition, on a 20, the struck target is dazed for 1 round if it is chaotic. It bestows one negative level on any chaotic creature attempting to wield it. The negative level remains as long as the weapon is in hand and disappears when the weapon is no longer wielded. This negative level never results in actual level loss, but it cannot be overcome in any way (including restoration spells) while the weapon is wielded. Bows, crossbows, and slings so crafted bestow the lawful power upon their ammunition.

\prereq{Craft Magic Arms and Armor, order's wrath, creator must be lawful}

\armdescription{Bane}{1}{6th}{Faint conjuration}{\x} A bane weapon excels at attacking one type or subtype of creature. Against its designated foe, it deals double bonus damage. Bows, crossbows, and slings so crafted bestow the bane quality upon their ammunition. A list of possible foes is described on the following table.

\begin{dtable}
\begin{tabularx}{\columnwidth}{>{\lcol}X >{\lcol}X}
\thead{Designated Foe} & \thead{Designated Foe}\\
Aberrations & Animals \\
Constructs & Dragons \\
Elementals & Fey \\
Giants & Humanoids, civilized \\
Humanoids, savage & Magical beasts \\
Monstrous humanoids & Oozes \\
Outsiders, inner planes & Outsiders, outer planes \\
Plants & Undead \\
Vermin & \\
\end{tabularx}
\end{dtable}

\prereq{Craft Magic Arms and Armor, summon monster (any)}

\armdescription{Brilliant Energy}{4}{16th}{Strong conjuration, evocation, and transmutation}{\x} A brilliant energy weapon has its significant portion transformed into \my{pure energy}, although this does not modify the item's weight. It always gives off light as a torch (20-foot radius). A brilliant energy weapon deals damage with all five energy types at once: acid, cold, electricity, fire, and sonic. A struck target uses the lowest of its energy resistances and damage reduction to avoid damage from the weapon, allowing it to effectively ignore such defences unless the target resists every type of elemental damage and has damage reduction. In addition, if the target has any vulnerabilities to energy damage, a brilliant energy weapon deals extra damage as if it was dealing damage of that type.

\prereq{Craft Magic Arms and Armor, gaseous form, sunbeam or sunburst, five spells of 3rd level or higher with the following descriptors: acid, cold, electricity, fire, and sonic}

\armdescription{Corrosive}{1}{6th}{Faint conjuration [acid]}{Swift (command)} Upon command, a corrosive weapon is sheathed in acid. The acid does not harm the wielder. The effect remains until another command is given. A corrosive weapon deals double bonus damage against objects. In addition, on a 20, the weapon deals only acid damage, ignores damage reduction, and sickens the struck target. Bows, crossbows, and slings so crafted bestow the corrosive property upon their ammunition.

\prereq{Craft Magic Arms and Armor, acid arrow}

\armdescription{Dancing}{4}{16th}{Strong transmutation}{Standard (command)} As a standard action, a dancing weapon can be loosed to attack on its own. It fights for 4 rounds using the \my{attack bonus} of the one who loosed it and then drops. While dancing, it cannot make attacks of opportunity, and the person who activated it is not considered armed with the weapon. In all other respects, it is considered wielded or attended by the creature for all maneuvers and effects that target items. While dancing, it takes up the same space as the activating character and can attack \my{foes it would threaten if wielded normally}. The dancing weapon accompanies the person who activated it everywhere, whether she moves by physical or magical means. If the wielder who loosed it has an unoccupied hand, she can grasp it while it is attacking on its own as a \my{swift} action; when so retrieved the weapon can't dance (attack on its own) again until 4 rounds after it started dancing.

\prereq{Craft Magic Arms and Armor, animate objects}

\armdescription{Defending}{1}{6th}{Faint abjuration}{Free (mental)} A defending weapon allows the wielder to transfer some or all of the sword's enhancement bonus to his AC as an enhancement bonus to his dodge modifier when fighting defensively, using Combat Expertise, or taking the total defense action. As a free action, the wielder chooses how to allocate the weapon's enhancement bonus when taking one of those three actions, and the effect to AC lasts until \my{the start of his} next turn.

\prereq{Craft Magic Arms and Armor, \my{lesser precognition}}

\armdescriptiondc{Disruption}{2}{12th}{Moderate necromancy}{\x}{14} A weapon of disruption is the bane of all undead. Any bloodied undead creature struck in combat must succeed on a Will save or be destroyed. Healthy undead creatures suffer no ill effect. A weapon of disruption must be a bludgeoning weapon. (If you roll this property randomly for a piercing or slashing weapon, reroll.)

\prereq{Craft Magic Arms and Armor, disrupting weapon}

\armdescription{Distance}{1}{6th}{Faint divination}{\x} This property can only be placed on a ranged weapon. A weapon of distance has double the range increment of other weapons of its kind.

\prereq{Craft Magic Arms and Armor, farsight}

\armdescription{Flaming}{1}{6th}{Faint evocation [fire]}{Swift (command)} Upon command, a flaming weapon is sheathed in fire. The fire does not harm the wielder. The effect remains until another command is given. All damage dealt by a flaming weapon is fire damage. In addition, on a 20, the weapon ignores damage reduction and ignites the struck target for 5 rounds.
\par An ignited creature takes a \minus2 penalty to attack rolls, checks, saving throws, and armor class, and takes d6 damage per round from the fire. If the creature takes a full-round action, it can attempt a DC 15 Reflex save to put out the flames. This action provokes attacks of opportunity. Dropping prone gives a \plus4 circumstance bonus on this save.
\par Projectile weapons so crafted bestow the flaming property upon their ammunition.

\prereq{Craft Magic Arms and Armor, flame blade}

\armdescription{Force}{2}{8th}{Moderate evocation [force]}{Swift (command)} Upon command, a freezing weapon is sheathed in force energy. The force does not harm the wielder. The effect remains until another command is given. All damage dealt by a force weapon is force damage. In addition, on a 20, the weapon ignores damage reduction and knocks the target struck prone.
\par Projectile weapons so crafted bestow the force property upon their ammunition.

\prereq{Craft Magic Arms and Armor, magic missile or spiritual weapon}

\armdescription{Freezing}{1}{6th}{Faint evocation [cold]}{Swift (command)} Upon command, a freezing weapon is sheathed in icy cold. The cold does not harm the wielder. The effect remains until another command is given. All damage dealt by a freezing weapon is cold damage. In addition, on a 20, the weapon ignores damage reduction and fatigues the target struck for 5 rounds.
\par Projectile weapons so crafted bestow the freezing property upon their ammunition.

\prereq{Craft Magic Arms and Armor, ray of frost}

\armdescription{Ghost Touch}{1}{6th}{Faint transmutation}{\x} A ghost touch weapon deals damage normally against incorporeal creatures, regardless of its bonus. (An incorporeal creature's 50\% chance to avoid damage does not apply to attacks with ghost touch weapons.) The weapon can be picked up and moved by an incorporeal creature at any time. A manifesting ghost can wield the weapon against corporeal foes. Essentially, a ghost touch weapon counts as either corporeal or incorporeal at any given time, whichever is more beneficial to the wielder.

\prereq{Craft Magic Arms and Armor, \my{blink}}

\armdescription{Holy}{2}{10th}{Moderate evocation [good]}{\x} A holy weapon is imbued with holy power. This power makes the weapon good-aligned and thus allows it to bypass the corresponding damage reduction. It deals double bonus damage against evil foes. In addition, on a 20, the struck target is blinded for 2 rounds. It bestows one negative level on any evil creature attempting to wield it. The negative level remains as long as the weapon is in hand and disappears when the weapon is no longer wielded. This negative level never results in actual level loss, but it cannot be overcome in any way (including restoration spells) while the weapon is wielded. Bows, crossbows, and slings so crafted bestow the holy power upon their ammunition.

\prereq{Craft Magic Arms and Armor, holy smite, creator must be good}

\armdescription{Keen}{1}{6th}{Faint transmutation}{\x} This ability doubles the threat range of a weapon. Only piercing or slashing weapons can be keen. This benefit doesn't stack with any other effect that expands the threat range of a weapon (such as the Improved Critical feat).

\prereq{Craft Magic Arms and Armor, magic weapon}

\armdescription{Ki Focus}{1}{6th}{Faint transmutation}{\x} The magic weapon serves as a channel for the wielder's \ki, allowing her to use her special \ki attacks through the weapon as if they were unarmed attacks. These attacks include the monk's stunning attack and quivering palm, as well as the Stunning Fist feat. Only melee weapons can have the \ki focus ability.

\prereq{Craft Magic Arms and Armor, creator must be a monk}

\armdescription{Merciful}{1}{6th}{Faint necromancy}{\x} \my{All damage the weapon deals} is nonlethal damage. Bows, crossbows, and slings so crafted bestow the merciful effect upon their ammunition. This effect can be suppressed or resumed with a command word as a swift action.

\prereq{Craft Magic Arms and Armor, cure light wounds}

\armdescription{Mighty Cleaving}{2}{8th}{Moderate evocation}{\x} A mighty cleaving weapon allows a wielder to make one cleave attempt in a round, as if he had the Cleave feat, except that it only cleaves once per round. If the wielder has the Cleave feat, this attack is granted in addition to the wielder's normal extra attack granted by the feat..

\prereq{Craft Magic Arms and Armor, \my{bull's strength}}

\armdescription{Returning}{1}{6th}{Faint conjuration}{\x} This special ability can only be placed on a weapon that can be thrown. A returning weapon teleports back to the creature that threw it. It returns to the thrower just before the creature's next turn (and is therefore ready to use again in that turn).

Catching a returning weapon when it comes back is a free action. If the character can't catch it, the weapon drops to the ground in the square from which it was thrown.

\prereq{Craft Magic Arms and Armor, lesser conjure weapon}

\armdescription{Returning, Greater}{2}{10th}{Moderate conjuration}{\x} This functions as the returning ability, except that the weapon teleports back to the creature that threw it immediately after the attack is resolved, allowing it to make multiple attacks in the same round with the same throwing weapon.

\prereq{Craft Magic Arms and Armor, conjure weapon, greater}

\armdescription{Shocking}{1}{6th}{Faint evocation [electricity]}{Swift (command)} Upon command, a shocking weapon is sheathed in electricity. The electricity does not harm the wielder. The effect remains until another command is given. All damage dealt by a shocking weapon is electricity damage. In addition, on a 20, the weapon ignores damage reduction and staggers the struck target for 2 rounds. Projectile weapons crafted bestow the shocking property upon their ammunition.

\prereq{Craft Magic Arms and Armor, shocking grasp}

\armdescription{Speed}{4}{10th}{Moderate transmutation}{\x} When making a full attack action, the wielder of a speed weapon may make one extra attack with it. The attack uses the wielder's full base attack bonus, plus any modifiers appropriate to the situation. (This benefit is not cumulative with similar effects, such as a \spell{haste} spell.)

A character who is dual-wielding with a speed weapon may not attack with both weapons on the attack granted by the speed property.

\prereq{Craft Magic Arms and Armor, haste}

\armdescription{Throwing}{1}{4th}{Faint transmutation}{\x} This ability can only be placed on a melee weapon. A melee weapon crafted with this ability gains a range increment of 10 feet and can be thrown without penalty by a wielder proficient in its normal use.

\prereq{Craft Magic Arms and Armor, magic stone}

\armdescription{Seeking}{2}{10th}{Moderate divination}{\x} Only ranged weapons can have the seeking ability. The weapon veers toward its target, negating any miss chances that would otherwise apply, such as from concealment. (The wielder still has to aim the weapon at the right square. Arrows mistakenly shot into an empty space, for example, do not veer and hit invisible enemies, even if they are nearby.)

\prereq{Craft Magic Arms and Armor, \my{precognition or} true seeing}

\armdescription{Shielding}{1 to \plus5}{3 \mtimes bonus}{Faint abjuration (shielding) [force] (\plus1, \plus2); moderate (\plus3, \plus4; or strong (\plus5)}{\x} A shielding weapon projects a shield of force akin to the \spell{shield} spell when wielded. This grants the wielder a shield bonus to AC of \plus1 to \plus5, depending on the power of the shielding weapon. This shield bonus is considered separate from any other shield the wielder might possess, and never stacks with other shield bonuses.

\armdescription{Spell Storing}{2}{10th}{Moderate transmutation (plus aura of stored spell}{Immediate (command)} A spell storing weapon allows a spellcaster to store a single targeted spell of up to 3rd level in the weapon. (The spell must have a casting time of \my{no more than} 1 standard action.) Any time the weapon strikes a creature and the creature takes damage from it, the weapon can immediately cast the spell on that creature as an \my{immediate} action if the wielder desires. (This special ability is an exception to the general rule that casting a spell from an item takes at least as long as casting that spell normally.) Once the spell has been cast from the weapon, a spellcaster can cast any other targeted spell of up to 3rd level into it. The weapon magically imparts to the wielder the name of the spell currently stored within it. A randomly rolled spell storing weapon has a 50\% chance to have a spell stored in it already.

\prereq{Craft Magic Arms and Armor}

\armdescription{Spell Storing, Greater}{4}{14th}{Strong transmutation (plus aura of stored spell)}{Immediate (command)} A greater spell storing weapon functions as a spell storing weapon, except that it allows a spellcaster to store a single targeted spell of up to 6th level in the weapon.

\prereq{Craft Magic Arms and Armor}

\armdescription{Thundering}{1}{6th}{Faint evocation [sonic]}{Swift (command)} Upon command, a thundering weapon is sheathed in sonic energy. The energy does not harm the wielder. The effect remains until another command is given. All damage dealt by a thundering weapon is sonic damage. In addition, on a 20, the weapon ignores damage reduction and deafens the struck target for 5 rounds. Bows, crossbows, and slings so crafted bestow the thundering property upon their ammunition.

\prereq{Craft Magic Arms and Armor, \my{shatter or sound burst}}

\armdescription{Unholy}{2}{10th}{Moderate evocation [evil]}{\x} An unholy weapon is imbued with unholy power. This power makes the weapon evil-aligned and thus allows it to bypass the corresponding damage reduction. It deals double bonus damage against good foes. In addition, on a 20, the struck target is sickened for 5 rounds if it is good. It bestows one negative level on any good creature attempting to wield it. The negative level remains as long as the weapon is in hand and disappears when the weapon is no longer wielded. This negative level never results in actual level loss, but it cannot be overcome in any way (including restoration spells) while the weapon is wielded. Bows, crossbows, and slings so crafted bestow the unholy power upon their ammunition.

\prereq{Craft Magic Arms and Armor, unholy blight, creator must be evil}

\armdescription{Vorpal}{5}{18th}{Strong necromancy and transmutation}{\x} This potent and feared ability allows the weapon to sever the heads of those it strikes. A vorpal weapon has a \plus1 enhancement bonus to its critical multiplier and a doubled critical threat range, as the keen property. In addition, upon a roll of natural 20 (followed by a successful roll to confirm the critical hit), the weapon severs the opponent's head (if it has one) from its body \my{if it is bloodied after the damage dealt by the attack}. Some creatures, such as many aberrations and all oozes, have no heads. Most other creatures, however, die when their heads are cut off. A vorpal weapon must be a slashing weapon.

\prereq{Craft Magic Arms and Armor, circle of death \my{or finger of death, greater magic weapon}}

\armdescription{Wounding}{2}{8th}{Moderate necromancy}{\x} A wounding weapon deals 1 point of Constitution damage from blood loss when it hits a creature. A critical hit does not multiply the Constitution damage. Creatures immune to critical hits or that do not bleed (such as plants or constructs) are immune to the Constitution damage dealt by this weapon.

\prereq{Craft Magic Arms and Armor, \my{vampiric touch}}