\section{Scrolls}

\parhead{Scroll Mishaps} When a mishap occurs, the spell on the scroll has a reversed or harmful effect. Possible mishaps are given below.
\begin{itemize*}
\item A surge of uncontrolled magical energy deals 1d6 points of damage per spell level to the scroll user.
\item Spell strikes the scroll user or an ally instead of the intended target, or a random target nearby if the scroll user was the intended recipient.
\item Spell takes effect at some random location within spell range.
\item Spell's effect on the target is contrary to the spell's normal effect.
\item The scroll user suffers some minor but bizarre effect related to the spell in some way. Most such effects should last only as long as the original spell's duration, or 2d10 minutes for instantaneous spells.
\item Some innocuous item or items appear in the spell's area.
\item Spell has delayed effect. Sometime within the next 1d12 hours, the spell activates. If the scroll user was the intended recipient, the spell takes effect normally. If the user was not the intended recipient, the spell goes off in the general direction of the original recipient or target, up to the spell's maximum range, if the target has moved away.
\end{itemize*}

\parhead{Scroll Origins} Some classes are more likely to make scrolls than others. When determining the origin of a scroll

\section{Staffs}
%Staffs are spell level * caster level * 500 per spell for a 50 charge staff.

A staff is a long shaft of wood that stores several spells. Unlike wands, which can contain a wide variety of spells, each staff is of a certain kind and holds specific spells. A staff has 50 charges when created.

\parhead{Physical Description} A typical staff is 4 feet to 7 feet long and 2 inches to 3 inches thick, weighing about 5 pounds. Most staffs are wood, but a rare few are bone, metal, or even glass. (These are extremely exotic.) Staffs often have a gem or some device at their tip or are shod in metal at one or both ends. Staffs are often decorated with carvings or runes. A typical staff is like a walking stick, quarterstaff, or cudgel. It has AC 7, 10 hit points, hardness 5, and a break DC of 24.

\parhead{Activation} Staffs use the spell trigger activation method, so casting a spell from a staff is usually a standard action that doesn't provoke attacks of opportunity. (If the spell being cast, however, has a longer casting time than 1 standard action, it takes that long to cast the spell from a staff.) To activate a staff, a character must hold it forth in at least one hand (or whatever passes for a hand, for nonhumanoid creatures).

\begin{dtable}
\lcaption{Staffs}
\begin{tabularx}{\columnwidth}{>{\lcol}X l l}
\thead{Staff} & \thead{Market Price} & \thead{Item Level} \\
Illumination & 6,000 gp & 8th \\
Swarming insects & 17,000 gp & 12th \\
Charming & 14,7500 gp & 11th \\
Healing & 27,750 gp & 14th \\
Size alteration & 30,000 gp & 15th \\
Abjuration & 21,500 gp & 13th \\
Conjuration & 21,500 gp & 13th \\
Enchantment & 21,500 gp & 13th \\
Evocation & 21,500 gp & 13th \\
Illusion & 21,500 gp & 13th \\
Necromancy & 21,500 gp & 13th \\
Transmutation & 21,500 gp & 13th \\
Divination & 21,500 gp & 13th \\
Fire & 21,750 gp & 13th \\
Frost & 23,500 gp & 14th \\
Earth and stone & 28,500 gp & 16th \\
Woodlands & 50,000 gp & 18th \\
Life & 34,000 gp & 16th \\
Passage & 49,000 gp & 18th \\
Power & 57,000 gp & 19th
\end{tabularx}
\end{dtable}

\subsection{Staff Descriptions}

Staffs use the wielder's attribute score and relevant feats to set the DC for saves against their spells. Unlike with other sorts of magic items, the wielder can use his caster level when activating the power of a staff if it's higher than the caster level of the staff.

This means that staffs are far more potent in the hands of a powerful spellcaster. Because they use the wielder's attribute score to set the save DC for the spell, spells from a staff are often harder to resist than ones from other magic items, which use the minimum attribute score required to cast the spell. Not only are aspects of the spell dependant on caster level (range, duration, and so on) potentially higher, but spells from a staff are harder to dispel and have a better chance of overcoming a target's spell resistance.

Unlike wands, staffs typically have 50 charges when first created.

Furthermore, a staff can hold a spell of any level, unlike a wand, which is limited to spells of 4th level or lower. The minimum caster level of a staff is 8th. Standard staffs are described below.

\tooldescription{Abjuration}{21,500}{13th}{12th}{Moderate abjuration}{Standard (spell trigger)} Usually carved from the heartwood of an ancient oak or other large tree, this staff allows use of the following spells:
\begin{itemize*}
\item Shield (1 charge)
\item Resist energy (1 charge)
\item Dispel magic (1 charge)
\item Lesser globe of invulnerability (2 charges)
\item Spell resistance (3 charges)
\end{itemize*}

\prereq{Craft Staff, dispel magic, lesser globe of invulnerability, resist energy, shield, \my{spell resistance}}

\tooldescription{Charming}{14,750}{11th}{10th}{Moderate enchantment}{Standard (spell trigger)} Made of twisting wood ornately shaped and carved, this staff allows use of the following spells:
\begin{itemize*}
\item Charm person (1 charge)
\item Charm monster (2 charges)
\end{itemize*}

\prereq{Craft Staff, charm person, charm monster}

\tooldescription{Conjuration}{21,500}{13th}{12th}{Moderate conjuration}{Standard (spell trigger)} This staff is usually made of ash or walnut and bears ornate carvings of many different kinds of creatures. It allows use of the following spells:
\begin{itemize*}
\item Obscuring mist (1 charge)
\item Acid arrow (1 charge)
\item Stinking cloud (1 charge)
\item Summon monster IV (2 charges)
\item Cloudkill \my{(3 charges)}
\end{itemize*}

\prereq{Craft Staff, \my{acid arrow}, cloudkill, \my{obscuring mist}, stinking cloud, \my{wall of ice}}

\tooldescription{Divination}{21,500}{13th}{12th}{Moderate divination}{Standard (spell trigger)} Made from a supple length of willow, often with a forked tip, this staff allows use of the following spells:
\begin{itemize*}
\item Farsight (1 charge)
\item Locate object (1 charge)
\item Detect thoughts (1 charge)
\item True cast (2 charges)
\item Greater precognition (3 charges)
\end{itemize*}

\prereq{Craft Staff, \my{detect thoughts}, \my{farsight}, locate creature, locate object, \my{true cast}}

\tooldescription{Earth and Stone}{28,500}{16th}{12th}{Moderate transmutation}{Standard (spell trigger)} This staff is topped with a fist-sized emerald that gleams with smoldering power. It allows the use of the following spells:
\begin{itemize*}
\item Soften earth and stone (1 charge)
\item Stone shape (1 charges)
\item Passwall (2 charges)
\item Transmute mud and rock (2 charges)
\end{itemize*}

\prereq{Craft Staff, passwall, soften earth and stone, stone shape, transmute mud and rock}

\tooldescription{Enchantment}{21,500}{13th}{12th}{Moderate enchantment}{Standard (spell trigger)} Often made from applewood and topped with a clear crystal, this staff allows use of the following spells:
\begin{itemize*}
\item Attraction (1 charge)
\item Hideous laughter (1 charge)
\item Charm person (1 charge)
\item Suggestion (2 charges)
\item Hold monster (3 charges)
\end{itemize*}

\prereq{Craft Staff, \my{attraction, charm person,} hideous laughter, \my{hold monster, suggestion}}

\tooldescription{Evocation}{21,500}{13th}{12th}{Moderate evocation}{Standard (spell trigger)} Usually very smooth and carved from hickory, willow, or yew, this staff allows use of the following spells:
\begin{itemize*}
\item Magic missile (1 charge)
\item Shatter (1 charge)
\item Fireball (1 charge)
\item Shout (2 charges)
\item Chain lightning (3 charges)
\end{itemize*}

\prereq{Craft Staff, chain lightning, fireball, magic missile, shatter, \my{shout}}

\tooldescription{Fire}{21,750}{13th}{10th}{Moderate evocation}{Standard (spell trigger)} Crafted from bronzewood with brass bindings, this staff allows use of the following spells:
\begin{itemize*}
\item Burning hands (1 charge)
\item Flaming sphere (1 charge)
\item Fireball (1 charge)
\item Scorching ray (1 charge)
\item Wall of fire (2 charges)
\end{itemize*}

\prereq{Craft Staff, burning hands, fireball, \my{flaming sphere, scorching ray,} wall of fire}

\tooldescription{Frost}{23,500}{14th}{12th}{Moderate conjuration and evocation}{Standard (spell trigger)} Tipped on either end with a glistening diamond, this rune-covered staff allows use of the following spells:
\begin{itemize*}
\item Ray of frost (1 charge)
\item Sleet storm (1 charge)
\item Ice storm  (2 charges)
\item Wall of ice (2 charges)
\item Cone of cold (3 charges)
\end{itemize*}

\prereq{Craft Staff, cone of cold, ice storm, \my{ray of frost, sleet storm,} wall of ice}

\tooldescription{Illusion}{21,500}{13th}{12th}{Moderate illusion}{Standard (spell trigger)} This staff is made from ebony or other dark wood and carved into an intricately twisted, fluted, or spiral shape. It allows use of the following spells:
\begin{itemize*}
\item Ventriloquism (1 charge)
\item Silent image (1 charge)
\item Invisibility (1 charge)
\item Major image (2 charges)
\item Seeming (3 charges)
\end{itemize*}

\prereq{Craft Staff, \my{invisibility}, major image, \my{seeming, silent image, ventriloquism}}

\tooldescription{Illumination}{6,000}{8th}{8th}{Moderate evocation [Light]}{Standard (spell trigger)}
\parhead{Illumination} This staff is usually sheathed in silver and decorated with \my{stars}. It allows use of the following spells:
\begin{itemize*}
\item Light (1 charge)
\item Daylight (1 charge)
\item Searing light (2 charges)
\end{itemize*}

\prereq{Craft Staff, dancing lights, daylight, flare, \my{searing light}}

\tooldescription{Life}{34,000}{16th}{14th}{Strong necromancy (Healing) [Positive]}{Standard (spell trigger)} Made of thick oak shod in gold, this staff allows use of the following spells:
\begin{itemize*}
\item Cure serious wounds (1 charge)
\item Link vitality (1 charge)
\item Death ward (2 charges)
\item Cure light wounds, mass (2 charges)
\item Heal (3 charges)
\end{itemize*}

\prereq{Craft Staff, \my{cure serious wounds, death ward, link vitality, mass cure light wounds,} heal}

\tooldescription{Necromancy}{21,500}{13th}{12th}{Moderate necromancy}{Standard (spell trigger)} This staff is made from ebony or other dark wood and carved with the images of bones and skulls. It allows use of the following spells:
\begin{itemize*}
\item Ray of enfeeblement (1 charge)
\item Ghoul touch (1 charge)
\item Vampiric touch (1 charge)
\item Enervation (2 charges)
\item Waves of fatigue (3 charges)
\end{itemize*}

\prereq{Craft Staff, enervation, ghoul touch, \my{ray of enfeeblement, vampiric touch}, waves of fatigue}

\tooldescription{Passage}{49,000}{18th}{18th}{Strong conjuration and transmutation}{Standard (spell trigger)} This potent item allows use of the following spells:
\begin{itemize*}
\item Dimension door (1 charge)
\item Passwall (2 charges)
\item Phase door (2 charges)
\item Dimension door, mass (3 charges)
\end{itemize*}

\prereq{Craft Staff, dimension door, mass dimension door, passwall, phase door}

%+2000 gp for weapon
%+3rd level spell for smite (4,500)
%Retributive strike is ``free''
\tooldescription{Power}{57,000}{19th}{16th}{Strong abjuration, enchantment, evocation, and necromancy}{Standard (spell trigger) and swift (command word)} The staff of power is a very potent magic item. It is usually topped with a glistening gem, its shaft straight and smooth. It has the following powers:
\begin{itemize*}
\item Magic missile (1 charge)
\item Ray of enfeeblement (1 charge)
\item Levitate (1 charge)
\item Lightning bolt (1 charge)
\item Fireball (1 charge)
\item Cone of cold (2 charges)
\item Hold monster (2 charges)
\item Wall of force (in a 10-ft.-diameter hemisphere around the caster only) (2 charges)
\item Globe of invulnerability (3 charges)
\end{itemize*}

The staff is also a \plus2 quarterstaff, and its wielder may use it to smite opponents. If 1 charge is expended (as a \my{swift} action), the staff causes double damage (x3 on a critical hit) for 1 round.

A staff of power can be used for a retributive strike, requiring it to be broken by its wielder. (If this breaking of the staff is purposeful and declared by the wielder, it can be performed as a standard action that does not require the wielder to make a Strength check.) All charges currently in the staff are instantly released in a 30-foot radius. All within 2 squares of the broken staff take points of damage equal to 8 \mtimes the number of charges in the staff, those 3 or 4 squares away take 6 \mtimes the number of charges in damage, and those 5 or 6 squares distant take 4 \mtimes the number of charges in damage. All those affected can make DC \my{25} Reflex saves to reduce the damage by half.

The character breaking the staff has a 50\% chance of traveling to another plane of existence, but if he does not, the explosive release of spell energy destroys him. Only certain items, including the staff of the magi and the staff of power, are capable of being used for a retributive strike.

After all charges are used up from the staff, it remains a \plus2 quarterstaff. (Once empty of charges, it cannot be used for a retributive strike.)

\prereq{Craft Staff, Craft Magic Arms and Armor, magic missile, ray of enfeeblement, levitate, fireball, lightning bolt, cone of cold, hold monster, wall of force, globe of invulnerability}

\tooldescription{Size Alteration}{20,000}{13th}{10th}{Moderate transmutation}{Standard (spell trigger)}
\parhead{Size Alteration} Stout and sturdy, this staff of dark wood allows use of the following spells:
\begin{itemize*}
\item Reduce person (1 charge)
\item Enlarge person (1 charge)
\item Shrink item (1 charge)
\item Reduce person, mass (2 charges)
\end{itemize*}

\prereq{Craft Staff, enlarge person, reduce person, mass reduce person, shrink item}

\tooldescription{Swarming Insects}{11,500}{10th}{10th}{Moderate conjuration}{Standard (spell trigger)} Made of twisted dark wood with dark spots resembling crawling insects (which occasionally seem to move), this staff allows use of the following spells:
\begin{itemize*}
\item Summon swarm (1 charge)
\item Insect plague (3 charges)
\end{itemize*}

\prereq{Craft Staff, insect plague, summon swarm}

\tooldescription{Transmutation}{21,500}{13th}{12th}{Moderate transmutation}{Standard (spell trigger)} This staff is generally carved from or decorated with petrified wood and allows use of the following spells:
\begin{itemize*}
\item Expeditious retreat (1 charge)
\item Spider climb (1 charge)
\item Enlarge person (1 charge)
\item Haste (2 charges)
\item Baleful polymorph (3 charges)
\end{itemize*}

\prereq{Craft Staff, baleful polymorph, \my{enlarge person}, expeditious retreat, \my{haste, spider climb}}

\tooldescription{Woodlands}{35,500}{16th}{16th}{Strong conjuration, divination, enchantment, and transmutation}{Standard (spell trigger)} Appearing to have grown naturally into its shape, this oak, ash, or yew staff allows use of the following spells:
\begin{itemize*}
\item Speak with animals (1 charge)
\item Charm animal (1 charge)
\item Barkskin (1 charge)
\item Spike stones (2 charges)
\item Wall of thorns (3 charges)
\item Summon nature's ally VI (3 charges)
\item Animate plants (4 charges)
\end{itemize*}

The staff may be used as a weapon, functioning as a \plus2 quarterstaff. The staff of the woodlands also allows its wielder to pass without trace at will, with no charge cost. These two attributes continue to function after all the charges are expended.

\prereq{Craft Staff, Craft Magic Arms and Armor, animate plants, barkskin, charm animal, pass without trace, speak with animals, summon nature's ally VI, wall of thorns}