\section{Rings}

A full list of magical rings is given here.

\begin{dtable}
\lcaption{Rings}
\begin{tabularx}{\columnwidth}{>{\lcol}X l l}
\thead{Ring} & \thead{Market Price} & \thead{Item Level} \\
Protection \plus1 & 1,000 gp & 3rd \\
Feather falling & 2,000 gp & 5th \\
Force shield & 2,000 gp & 5th \\
Climbing & 2,500 gp & 5th \\
Jumping & 2,500 gp & 5th \\
Swimming & 2,500 gp & 5th \\
Sustenance & 3,000 gp & 6th \\
Counterspells & 4,000 gp & 6th \\
Protection \plus2 & 4,000 gp & 6th \\
Arcane mastery (I) & 5,000 gp & 7th \\
Animal friendship & 6,000 gp & 8th \\
Ram & 6,000 gp & 8th \\
Energy resistance, minor & 8,000 gp & 9th \\
Evasion & 8,000 gp & 9th \\
Water walking & 8,000 gp & 9th \\
Protection \plus3 & 9,000 gp & 9th \\
Chameleon power & 10,000 gp & 10th \\
Climbing, improved & 10,000 gp & 10th \\
Jumping, improved & 10,000 gp & 10th \\
Swimming, improved & 10,000 gp & 10th \\
Freedom & 12,000 gp & 10th \\
Mind shielding & 12,000 gp & 10th \\
Invisibility & 13,500 gp & 11th \\
X-ray vision & 13,500 gp & 11th \\
Friend shield (pair) & 14,000 gp & 11th \\
Energy resistance & 16,000 gp & 12th \\
Protection \plus4 & 16,000 gp & 12th \\
Spell storing, minor & 18,000 gp & 12th \\
Arcane mastery (II) & 20,000 gp & 13th \\
Blinking & 24,000 gp & 14th \\
Energy resistance, major & 24,000 gp & 14th \\
Telekinesis & 25,000 gp & 14th \\
Protection \plus5 & 25,000 gp & 14th \\
Regeneration & 36,000 gp & 16th \\
Spell turning & 36,000 gp & 16th \\
Arcane mastery (III) & 45,000 gp & 17th \\
Spell storing & 50,000 gp & 18th \\
\end{tabularx}
\end{dtable}

\subsubsection{Ring Descriptions}

Standard rings are described below.

%*.75 for 24 hour restriction
\itemdescription{Animal Friendship}{6,000}{8th}{Ring}{8th}{Moderate enchantment}{Standard (command)} On command, this ring affects an animal as if the wearer had cast \spell{charm animal}. An animal who successfully saves against this spell cannot be affected by it for another 24 hours.

\prereq{Forge Ring, charm animal}

\itemdescriptionmulti{Arcane Mastery}{3}{
\thead{Type:} & \thead{I} & \thead{II} & \thead{III} \\
\thead{Price:} & 5,000 gp & 20,000 gp & 45,000 gp \\
\thead{Level:} & 7th & 13th & 17th \\
\thead{Caster Level:} & 6th & 10th & 14th \\
\thead{Aura:} & Faint & Moderate & Strong}{Universal}{Ring}{\x} This special ring comes in \my{three} kinds (arcane mastery I, arcane mastery II, and arcane mastery III), all of them useful only to arcane spellcasters. The wearer's arcane spells per day are doubled for one specific spell level. A ring of arcane mastery I doubles 1st-level spells, a ring of arcane mastery II doubles 2nd-level spells, and a ring of arcane mastery III doubles 3rd-level spells. Bonus spells from high attribute scores are not doubled.

\prereq{Forge Ring, creator must cast arcane spells}

\itemdescription{Blinking}{24,000}{14th}{Ring}{10th}{Moderate transmutation}{Standard (command)} \my{When activated}, this ring makes the wearer blink, as with the \spell{blink} spell, \my{for 1 minute}. \my{This ability functions three times per day.}

\prereq{Forge Ring, blink}

\itemdescription{Chameleon Power}{10,000}{10th}{Ring}{8th}{Moderate illusion}{Swift (mental)} \my{When activated}, the wearer of this ring can gain the ability to magically blend in with the surroundings. This provides a \plus10 \my{enhancement} bonus on her Hide checks.

\prereq{Forge Ring, \my{blur}}

\itemdescription{Climbing}{2,500}{5th}{Ring}{4th}{Faint transmutation}{\x} This ring is actually a magic leather cord that ties around a finger. It continually grants the wearer a \plus5 \my{enhancement} bonus on Climb checks.

\prereq{Forge Ring, creator must have 5 ranks in the Climb skill}

\itemdescription{Climbing, Improved}{10,000}{10th}{Ring}{8th}{Moderate transmutation}{\x} As climbing, except it grants a \plus10 \my{enhancement} bonus on its wearer's Climb checks.

\prereq{Forge Ring, creator must have 10 ranks in the Climb skill}

%Price: As original
\itemdescription{Counterspells}{4,000}{7th}{Ring}{8th}{Moderate abjuration}{\x} This ring might seem to be a ring of spell storing upon first examination. However, while it allows a single spell of 1st through 6th level to be cast into it, that spell cannot be cast out of the ring again. Instead, should that spell ever be cast upon the wearer, the spell is immediately countered, as a counterspell action, requiring no action (or even knowledge) on the wearer's part. Once so used, the spell cast within the ring is gone. A new spell (or the same one as before) may be placed in it again.

\prereq{Forge Ring, \my{dispel magic}}

\itemdescription{Energy Resistance}{16,000}{12th}{Ring}{12th}{Moderate abjuration}{\x{} and standard (command)} As minor energy resistance, except that it grants 20 points of resistance.

\prereq{Forge Ring, resist energy}

\itemdescription{Energy Resistance, Major}{24,000}{14th}{Ring}{16th}{Strong abjuration}{\x{} and standard (command)} As minor energy resistance, except that it grants 30 points of resistance.

\prereq{Forge Ring, resist energy}

\itemdescription{Energy Resistance, Minor}{8,000}{9th}{Ring}{8th}{Moderate abjuration}{\x{} and standard (command)} This reddish iron ring continually protects the wearer from damage from one type of energy - acid, cold, electricity, fire, or sonic. Each time the wearer would normally take such damage, subtract the ring's resistance value from the damage dealt. The wearer can change what kind of energy the ring resists by speaking a command word as a standard action.

A minor ring of energy resistance grants 10 points of resistance.

\prereq{Forge Ring, resist energy}

\itemdescription{Evasion}{8,000}{9th}{Ring}{6th}{Faint transmutation}{Immediate (mental)} \my{Twice per day,} this ring grants the wearer the ability to avoid damage as if she had evasion. \my{If} she makes a Reflex saving throw \my{within 1 round after activating the ring} to determine whether she takes half damage, a successful save results in no damage.

\prereq{Forge Ring, \my{ablative barrier, cat's grace}}

\itemdescription{Feather Falling}{2,000}{5th}{Ring}{4th}{Faint transmutation}{\x} This ring is crafted with a feather pattern all around its edge. It acts exactly like a \spell{feather fall} spell, activated immediately if the wearer falls more than 5 feet.

\prereq{Forge Ring, feather fall}

\itemdescription{Force Shield}{2,000}{5th}{Ring}{4th}{Faint abjuration and evocation}{Swift (command)} An iron band, this simple ring generates a shield-sized (and shield-shaped) \my{barrier} of force that stays with the ring and can be wielded by the wearer as if it were a heavy shield (\plus2 AC). This special creation has no armor check penalty or arcane spell failure chance since it is weightless and encumbrance-free, \my{though it is still considered to be armor for the purpose of class features and abilities}. It can be activated and deactivated at will.

\prereq{Forge Ring, \my{floating disk, shield}}

\itemdescription{Freedom}{12,000}{10th}{Ring}{10th}{Moderate abjuration}{Swift (command)} This gold ring allows the wearer to act as if under the effect of a \spell{freedom} spell for up to one minute per day. The rounds need not be used sequentially.

\prereq{Forge Ring, \my{freedom}}

%Price: (6000 for shield other + 0.75*0.75*6000 for status)*0.75 for mated pair
\itemdescription{Friend Shield}{(two rings) 14,000}{11th}{Ring}{8th}{Moderate abjuration and divination}{Standard (command)} These curious rings always come in pairs. A friend shield ring without its mate is useless. Either wearer of one of a pair of the rings can, at any time, command his or her ring to cast a shield other spell with the wearer of the mated ring as the recipient. This effect has no range limitation. In addition, the wearer of one of the rings is constantly aware of the position and condition of the wearer of the mated ring, as the \spell{status} spell.

\prereq{Forge Ring, shield other, \my{status}}

\itemdescription{Invisibility}{13,500}{11th}{Ring}{8th}{Moderate illusion}{Standard (command)} By activating this simple silver ring, the wearer can benefit from \spell{invisibility}, as the spell. \my{The effect lasts for 8 minutes. This item functions three times per day.}

\prereq{Forge Ring, invisibility}

%Price: Not sure why
\itemdescription{Mind Shielding}{12,000}{10th}{Ring}{8th}{Moderate abjuration (shielding)}{\x} This ring is usually of fine workmanship and wrought from heavy gold. The wearer is continually immune to detect thoughts, discern lies, and any attempt to magically discern her alignment.

\prereq{Forge Ring, nondetection, \my{undetectable alignment}}

\itemdescriptiontwok{Protection}{Abjuration (shielding)}{Ring}{\x} This ring offers continual magical protection in the form of a bonus to saving throws of \plus1 to \plus5, depending on the kind of ring.

\prereq{Forge Ring, protection from chaos/good/evil/law}

%Price: Not sure why
\itemdescription{Ram}{6,000}{8th}{Ring}{8th}{Moderate evocation}{Standard (command)} The ring of the ram is an ornate ring forged of hard metal, usually iron or an iron alloy. It has the head of a ram as its device.

The wearer can command the ring to give forth a ramlike force, manifested by a vaguely discernible shape that resembles the head of a ram or a goat. This force strikes a single target, dealing 1d6 points of \my{force} damage if 1 charge is expended, 2d6 points if 2 charges are used, or 3d6 points if 3 charges (the maximum) are used. Treat this as a ranged attack with a \my{30-foot maximum range} and no penalties for distance.

The force of the blow is considerable, and those struck by the ring are subject to a bull rush. The ram has a \my{bonus on bull rush attacks equal to 5 \add character level of wielder} The ram gains a \my{\plus2 circumstance bonus} on the bull rush attempt if 2 charges are expended, or a \my{\plus5 circumstance bonus} if 3 charges are expended.

In addition to its attack mode, the ring of the ram also has the power to open doors as if it were a character with Strength \my{20}. If 2 charges are expended, the effect is equivalent to a character with Strength \my{24}. If 3 charges are expended, the effect is that of a character with Strength \my{30}.

A newly created ring has 50 charges. When all the charges are expended, the ring becomes a nonmagical item.

\prereq{Forge Ring, bull's strength, \my{knock, and either telekinetic maneuver or telekinesis}}

%Priced based on Vigor
\itemdescription{Regeneration}{36,000}{16th}{Ring}{18th}{Strong necromancy}{\x} This white gold ring continually allows a living wearer to heal \my{3 points of damage per round}. If the wearer loses a limb, an organ, or any other body part while wearing this ring, the ring regenerates it as the spell. In either case, only damage taken while wearing the ring is regenerated.

\prereq{Forge Ring, regenerate}

%Priced based on spell slot x2 for not needing to be able to use the spell.
\itemdescription{Spell Storing, Minor}{18,000}{12th}{Ring}{12th}{Moderate evocation}{Standard (command); see text} A minor ring of spell storing contains up to three levels of spells that the wearer can cast by saying a command word. Each spell has a caster level equal to the minimum level needed to cast that spell. The user need not provide any material components or focus, or pay an XP cost to cast the spell, and there is no arcane spell failure chance for wearing armor (because the ring wearer need not gesture). The activation time for the ring is same as the casting time for the relevant spell, with a minimum of 1 standard action.

For a randomly generated ring, treat it as a scroll to determine what spells are stored in it. If you roll a spell that would put the ring over the three-level limit, ignore that roll; the ring has no more spells in it. (Not every newly discovered ring need be fully charged.)

A spellcaster can cast any spells into the ring, so long as the total spell levels do not add up to more than three. Metamagic versions of spells take up storage space equal to their spell level modified by the metamagic feat. A spellcaster can use a scroll to put a spell into the minor ring of spell storing. Spells that cannot be cast on other targets, such as those with a range of personal, may not be cast into a ring of spell storing.

The ring magically imparts to the wearer the names of all spells currently stored within it.

\prereq{Forge Ring, imbue with spell ability}

\itemdescription{Spell Storing}{50,000}{18th}{Ring}{16th}{Strong evocation}{Standard (command); see text} As the minor ring of spell storing, except it holds up to five levels of spells.

\prereq{Forge Ring, imbue with spell ability}

\itemdescription{Spell Turning}{36,000}{16th}{Ring}{16th}{Strong abjuration}{Standard (command)} \my{Once} per day on command, this simple platinum band automatically reflects the next nine levels of spells cast at the wearer, exactly as if \spell{spell turning} had been cast upon the wearer.

\prereq{Forge Ring, spell turning}

%Price: not sure why
\itemdescription{Sustenance}{3,000}{6th}{Ring}{4th}{Faint conjuration}{\x} This ring continually provides its wearer with life-sustaining nourishment. The ring also refreshes the body and mind, so that its wearer needs only sleep 2 hours per day to gain the benefit of 8 hours of sleep. This does not affect how much rest the wearer must get to regain spells. The ring must be worn for a full week before it begins to work. If it is removed, the owner must wear it for another week to reattune it to himself.

\prereq{Forge Ring, create food and water}

\itemdescription{Swimming}{2,500}{5th}{Ring}{4th}{Faint transmutation}{\x} This silver ring has a wave pattern etched into the band. It continually grants the wearer a \plus5 competence bonus on Swim checks.

\prereq{Forge Ring, creator must have 5 ranks in the Swim skill}

\itemdescription{Swimming}{10,000}{10th}{Ring}{8th}{Moderate transmutation}{\x} As swimming, except it grants a \plus10 competence bonus on its wearer's Swim checks.

\prereq{Forge Ring, creator must have 10 ranks in the Swim skill}

\itemdescriptiondc{Telekinesis}{25,000}{14th}{Ring}{14th}{Strong evocation}{Standard (command)}{15} \my{Three times per day,} this ring allows the caster to use the spell \spell{telekinesis} on command.

\prereq{Forge Ring, telekinesis}

\itemdescription{Water Walking:}{8,000}{9th}{Ring}{6th}{Faint transmutation}{\x} This ring, set with an opal, allows the wearer to continually utilize the effects of the spell \spell{water walk}.

\prereq{Forge Ring, water walk}

%Price: As clair/claud, *0.75 for being weaker
\itemdescription{X-Ray Vision}{13,500}{11th}{Ring}{8th}{Moderate divination}{Standard (command)} On command, this ring gives its possessor the ability to see into and through solid matter. Vision range is 20 feet, with the viewer seeing as if he were looking at something in normal light even if there is no illumination. X-ray vision can penetrate 1 foot of stone, 1 inch of common metal, or up to 3 feet of wood or dirt. Thicker substances or a thin sheet of lead blocks the vision.

Using the ring is physically exhausting, causing the wearer 1 point of Constitution damage per minute after the first 10 minutes of use in a single day.

\prereq{Forge Ring, \my{clairaudience/clairvoyance}}

\section{Rods}

Rods are scepterlike devices that have unique magical powers and do not usually have charges. Anyone can use a rod.

\parhead{Physical Description} Rods weigh approximately 5 pounds.

They range from 2 feet to 3 feet long and are usually made of iron or some other metal. (Many, as noted in their descriptions, can function as light maces or clubs due to their sturdy construction.)

These sturdy items have AC 9, 10 hit points, hardness 10, and a break DC of 27.

\parhead{Activation} Details relating to rod use vary from item to item. See the individual descriptions for specifics.

\begin{dtable}
\lcaption{Rods}
\begin{tabularx}{\columnwidth}{>{\lcol}X l l}
\thead{Rod} & \thead{Market Price} & \thead{Item Level}\\
Cancellation & 3,000 gp & 6th \\
Flame extinguishing & 4,500 gp & 7th \\
Immovable & 5,000 gp & 7th \\
Metal and mineral detection & 6,000 gp & 8th \\
Enemy detection & 9,000 gp & 9th \\
Wonder & 9,000 gp & 10th \\
Withering & 10,000 gp & 10th \\
Python & 13,500 gp & 11th \\
Splendor & 15,500 gp & 14th \\
Negation & 18,000 gp & 12th \\
Thunder and lightning & 20,000 gp & 13th \\
Viper & 24,000 gp & 14th \\
Security & 27,000 gp & 14th \\
Rulership & 32,000 gp & 15th \\
Absorption & 50,000 gp & 18th \\
\end{tabularx}
\end{dtable}

\subsubsection{Rod Descriptions}

Although all rods are generally scepterlike, their configurations and abilities run the magical gamut. Standard rods are described below.

%Price: ad hoc
\itemdescription{Absorption}{50,000}{18th}{\x}{16th}{Strong abjuration}{\x} This rod acts as a magnet, drawing spells or spell-like abilities into itself. The magic absorbed must be a single-target spell or a ray directed at either the character possessing the rod or her gear. The rod then nullifies the spell's effect and stores its potential until the wielder releases this energy in the form of spells of her own. She can instantly detect a spell's level as the rod absorbs that spell's energy. Absorption requires no action on the part of the user if the rod is in hand at the time.

A running total of absorbed (and used) spell levels should be kept. The wielder of the rod can use captured spell energy to cast any spell she knows without expending a spell slot. The only restrictions are that the levels of spell energy stored in the rod must be equal to or greater than the level of the spell the wielder wants to cast, that any material components required for the spell be present, and that the rod be in hand when casting. For casters such as sorcerers who do not prepare spells, the rod's energy can be used to cast any spell of the appropriate level or levels that they know.

A rod of absorption absorbs a maximum of fifty spell levels and can thereafter only discharge any remaining potential it might have. The rod cannot be recharged. The wielder knows the rod's remaining absorbing potential and current amount of stored energy.

To determine the absorption potential remaining in a newly found rod, roll d\% and divide the result by 2. Then roll d\% again: On a result of 71-100, half the levels already absorbed by the rod are still stored within.

\prereq{Craft Rod, \my{spelltheft}, spell turning}

\itemdescriptiondc{Cancellation}{3,000}{6th}{\x}{12th}{Moderate abjuration}{\x}{16} This rod is a bane to magic items, for its touch \my{suppresses all magical properties in an item for 1 hour}. The item touched must make a Will save to prevent the rod from \my{suppressing} it. If a creature is holding it at the time, then the item can use the holder's Will save bonus in place of its own if the holder's is better. In such cases, contact is made by making a melee touch attack roll. Upon \my{suppressing} an item, the rod itself becomes brittle and cannot be used again.

A rod of cancellation can also be used to negate the effects of certain spells which create effects similar to magic items. This is noted in the appropriate spell descriptions.

\prereq{Craft Rod, \my{greater dispel magic}}

\itemdescription{Enemy Detection}{9,000}{9th}{\x}{10th}{Moderate divination}{Standard (command)} This device pulses in the wielder's hand and points in the direction of any creature or creatures hostile to the bearer of the device (nearest ones first). These creatures can be invisible, ethereal, hidden, disguised, or in plain sight. Detection range is \my{50 feet}. If the bearer of the rod concentrates for a full round, the rod pinpoints the location of the nearest enemy and indicates how many enemies are within range. The rod can be used three times each day, each use lasting up to 10 minutes.

\prereq{Craft Rod, \my{detect thoughts}}

%Price: Pyrotechnics *0.75 for weaker, *0.75 for many charges/day
\itemdescription{Flame Extinguishing}{4,500}{7th}{\x}{6th}{Moderate transmutation}{Standard (manipulation)} This rod can extinguish Medium or smaller nonmagical fires with simply a touch (a standard action). For the rod to be effective against other sorts of fires, the wielder must expend 1 or more of the rod's charges.

Extinguishing a Large or larger nonmagical fire, or a magic fire of Medium or smaller (such as that of a flaming weapon or a \spell{burning hands} spell), expends 1 charge. Continual magic flames, such as those of a weapon or a fire creature, are suppressed for \my{1 minute} and flare up again after that time. To extinguish an instantaneous fire spell, the rod must be within the area of the effect and the wielder must have used a ready action, effectively countering the entire spell.

When applied to Large or larger magic fires, such as those caused by \spell{fireball}, \spell{flame strike}, or \spell{wall of fire}, extinguishing the flames expends 2 charges from the rod.

If the device is used upon a fire creature (a melee touch attack), it deals 6d6 points of damage to the creature. This use requires 3 charges.

A rod of flame extinguishing has 10 charges when found. Spent charges are renewed every day, so that a wielder can expend up to 10 charges in any 24-hour period.

\prereq{Craft Rod, pyrotechnics}

\itemdescription{Immovable Rod}{5,000}{7th}{\x}{8th}{Moderate evocation and transmutation}{Move (manipulation)} This rod is a flat iron bar with a small button on one end. When the button is pushed (a move action), the rod does not move from where it is, even if staying in place defies gravity. Thus, the owner can lift or place the rod wherever he wishes, push the button, and let go. Several immovable rods can even make a ladder when used together (although only two are needed). An immovable rod can support up to 8,000 pounds before falling to the ground. If a creature pushes against an immovable rod, it must make a DC 30 Strength check to move the rod up to 10 feet in a single round.

\prereq{Craft Rod, \my{floating disk,} levitate}

\begin{comment}
\parhead{Lordly Might} This rod has functions that are spell-like, and it can also be used as a magic weapon of various sorts. It also has several more mundane uses. The rod of lordly might is metal, thicker than other rods, with a flanged ball at one end and six studlike buttons along its length. (Pushing any of the rod's buttons is equivalent to drawing a weapon.) It weighs 10 pounds.

The following spell-like functions of the rod can each be used once per day.
\begin{itemize*}
\item Hold person upon touch, if the wielder so commands (Will DC 14 negates). The wielder must choose to use this power and then succeed on a melee touch attack to activate the power. If the attack fails, the effect is lost.
\item Fear upon all enemies viewing it, if the wielder so desires (10-foot maximum range, Will DC 16 partial). Invoking this power is a standard action.
\item Deal 2d4 hit points of damage to an opponent on a successful touch attack (Will DC 17 half ) and cure the wielder of a like amount of damage. The wielder must choose to use this power before attacking, as with hold person.
\end{itemize*}

The following weapon functions of the rod have no limit on the number of times they can be employed.
\begin{itemize*}
\item In its normal form, the rod can be used as a \plus2 light mace.
\item When button 1 is pushed, the rod becomes a \plus1 flaming longsword. A blade springs from the ball, with the ball itself becoming the sword's hilt. The weapon lengthens to an overall length of 4 feet.
\item When button 2 is pushed, the rod becomes a \plus4 battleaxe. A wide blade springs forth at the ball, and the whole lengthens to 4 feet.
\item When button 3 is pushed, the rod becomes a \plus3 shortspear or \plus3 longspear. The spear blade springs forth, and the handle can be lengthened up to 12 feet (wielder's choice), for an overall length of from 6 feet to 15 feet. At its 15-foot length, the rod is suitable for use as a lance.
\end{itemize*}

The following other functions of the rod also have no limit on the number of times they can be employed.
\begin{itemize*}
\item Climbing pole/ladder. When button 4 is pushed, a spike that can anchor in granite is extruded from the ball, while the other end sprouts three sharp hooks. The rod lengthens to anywhere between 5 and 50 feet in a single round, stopping when button 4 is pushed again. Horizontal bars three inches long fold out from the sides, 1 foot apart, in staggered progression. The rod is firmly held by the spike and hooks and can bear up to 4,000 pounds. The wielder can retract the pole by pushing button 5.
\item The ladder function can be used to force open doors. The wielder plants the rod's base 30 feet or less from the portal to be forced and in line with it, then pushes button 4. The force exerted has a Strength of \plus12.
\item When button 6 is pushed, the rod indicates magnetic north and gives the wielder a knowledge of his approximate depth beneath the surface or height above it.
\end{itemize*}

Strong enchantment, evocation, necromancy, and transmutation; CL 19th; Craft Rod, Craft Magic Arms and Armor, inflict light wounds, bull's strength, flame blade, hold person, fear; Price 70,000 gp.
\end{comment}

%75% price due to metal/mineral caveat
\itemdescription{Metal and Mineral Detection}{6,000}{8th}{\x}{8th}{Moderate divination}{Full-round (mental)} This rod pulses in the wielder's hand and points to the largest mass of metal within \my{500 feet}. However, the wielder can concentrate on a specific metal or mineral. If the specific mineral is within \my{500 feet}, the rod points to any places it is located, and the rod wielder knows the approximate quantity as well. If more than one deposit of the specified metal or mineral is within range, the rod points to the largest cache first. Each operation requires a full-round action. This effect is blocked by even a thin sheet of lead (though it can be used to detect lead).

\prereq{Craft Rod, locate object}

\begin{comment}
\subsubsection{Metamagic Rods}
%preferred item level: minimum level needed to cast a spell of the modified level? No. Too high at higher levels.
%preferred item level: minimum level needed to cast a spell of (avg of highest spell levels affected by rod + spell level modifier)?
%Preferred price: L1 3,000 (6th), N1 8,000 (9th), G1 18,000 (12th)? <-too high at higher levels
%Preferred price: L2 12,000, N2 35,000, G2 xxx
%Preferred price: L3 17,500, N3 50,000, G3 xxx
%Preferred price: L4 25,000, N4 xxx. *1.5 for quicken because swift.
%Pricing structure: (2x spell slot of highest possible modified level - 2x spell slot of avg level affected) / 3 (2 uses/day)
%= L1 5,250 N1 16,500 G1 33,000
%= L2 9,250 N2 23,000 G2
Metamagic rods hold the essence of a metamagic feat. Using a metamagic rod, a spellcaster can cast a spell as if affected by the appropriate metamagic feat, including increasing the spell slot required to cast the spell.  All the rods described here are use-activated (but casting spells in a threatened area still draws an attack of opportunity). A caster may only use one metamagic rod on any given spell, but it is permissible to combine a rod with metamagic feats possessed by the rod's wielder.

Possession of a metamagic rod does not confer the associated feat on the owner, only the ability to use the given feat a specified number of times per day. \my{A spellcaster} must still take a full-round action when using a metamagic rod, just as if using a metamagic feat he possesses.

\parhead{Lesser and Greater Metamagic Rods} Normal metamagic rods can be used with spells of 6th level or lower. Lesser rods can be used with spells of 3rd level or lower, while greater rods can be used with spells of 9th level or lower.

\itemdescription{Metamagic, Empower}{(Lesser) 8,000 gp (9th); (normal) 15,000 gp (11th); (greater) 24,500}{14th}{\x}{(Lesser) 10th; (normal) 14th; (greater) 18th}{(Lesser) moderate (no school); (normal, greater) strong (no school)}{\x} The wielder can cast up to three spells per day that are empowered as though using the Empower Spell feat.

\prereq{Craft Rod, Empower Spell}

\itemdescription{Metamagic, Enlarge}{(Lesser) 4,500 gp (7th); (normal) 10,000 gp (10th); (greater) 18,000}{12th}{\x}{(Lesser) 8th; (normal) 12th; (greater) 16th}{(Lesser) faint (no school); (normal) moderate (no school); (greater) strong (no school)}{\x}  The wielder can cast up to three spells per day that are enlarged as though using the Enlarge Spell feat.

\prereq{Craft Rod, Enlarge Spell}

\itemdescription{Metamagic, Extend}{(Lesser) 4,500 gp (7th}; (normal) 10,000 gp (10th); (greater) 18,000}{12th}{\x}{(Lesser) 8th; (normal) 12th; (greater) 16th}{(Lesser) faint (no school); (normal) moderate (no school); (greater) strong (no school)}{\x} The wielder can cast up to three spells per day that are extended as though using the Extend Spell feat.

\prereq{Craft Rod, Extend Spell}

\itemdescription{Metamagic, Maximize}{(Lesser) 12,500 gp (11th); (normal) 21,000 gp  (13th); 32,000}{15th}{\x}{(Lesser) 8th; (normal) 12th; (greater) 16th}{(Lesser) faint (no school); (normal) moderate (no school); (greater) strong (no school)}{\x} The wielder can cast up to three spells per day that are maximized as though using the Maximize Spell feat.

Strong (no school); CL 17th; Craft Rod, Maximize Spell feat; Price 14,000 gp (lesser), 54,000 gp (normal), 121,500 gp (greater).

\itemdescription{Metamagic, Quicken}{(Lesser) 18,000 gp (11th); (normal) 21,000 gp  (13th); 32,000}{15th}{\x}{(Lesser) 8th; (normal) 12th; (greater) 16th}{(Lesser) faint (no school); (normal) moderate (no school); (greater) strong (no school)}{\x}
\parhead{Metamagic, Quicken} The wielder can cast up to three spells per day that are quickened as though using the Quicken Spell feat.

Strong (no school); CL 17th; Craft Rod, Quicken Spell; Price 35,000 gp (lesser), 75,500 gp (normal), 170,000 gp (greater).

\parhead{Metamagic, Silent} The wielder can cast up to three spells per day without verbal components as though using the Silent Spell feat.

Strong (no school); CL 17th; Craft Rod, Silent Spell; Price 3,000 gp (lesser), 11,000 gp (normal), 24,500 gp (greater).
\end{comment}

%Price/2 for magic items only (major restriction)
\itemdescription{Negation}{18,000}{12th}{\x}{14th}{Strong abjuration}{Standard (command)} This device negates the spell or spell-like function or functions of magic items. The wielder points the rod at the magic item, and a pale gray beam shoots forth to touch the target device, attacking as a ray (a ranged touch attack). The ray functions as a greater dispel magic spell, except it only affects magic items. To negate instantaneous effects from an item, the rod wielder needs to have used a ready action. The dispel check uses the rod's caster level \my{(14th)}. The target item gets no saving throw, although the rod can't negate artifacts (even minor artifacts). The rod can function three times per day.

\prereq{Craft Rod, \my{greater dispel magic}}

%Priced as summon nature's ally V 1/day, *0.75 for specific form, +1000 for weapon
\itemdescription{Python}{10,500}{10th}{\x}{12th}{Moderate transmutation}{Standard (manipulation)}
\parhead{Python} This rod is longer than normal rods. It is about 4 feet long and weighs 10 pounds. It strikes as a \plus1/\my{masterwork} quarterstaff. If the user throws the rod to the ground (a standard action), it grows to become a giant constrictor snake by the end of the round. The python obeys all commands of the owner. (In animal form, it retains the \plus1 enhancement bonus on attacks and damage possessed by the rod form.) The serpent returns to rod form (a full-round action \my{for the snake}) whenever the wielder desires, or whenever it moves farther than 100 feet from the owner. If the snake form is slain, it returns to rod form and cannot be activated again for three days.

\prereq{Craft Rod, Craft Magic Arms and Armor, baleful polymorph}

%Price: Mass charm, /4 for 500 minute restriction, ignore specialist modifier (not very specialist). DC is lower than normal.
\itemdescriptiondc{Rulership}{32,000}{15th}{\x}{18th}{Strong enchantment}{\x}{14} This rod looks like a royal scepter worth at least 5,000 gp in materials and workmanship alone. The wielder can command the obedience and fealty of creatures within 120 feet when she activates the device (a standard action). Creatures totaling 300 Hit Dice can be ruled, but creatures with Intelligence scores of 12 or higher are entitled to a Will save to negate the effect. Ruled creatures obey the wielder as if she were their absolute sovereign. Still, if the wielder gives a command that is contrary to the nature of the creatures commanded, the magic is broken. The rod can be used for 500 total minutes before crumbling to dust. This duration need not be continuous.

\prereq{Craft Rod, mass charm monster}

\itemdescription{Security}{27,000}{14th}{\x}{14th}{Strong conjuration}{Standard (command)} This item creates an \my{extradimensional} space, a pocket paradise. There the rod's possessor and as many as \my{100} other creatures can stay in complete safety for a period of time, up to \my{100} days divided by the number of creatures affected. All fractions are rounded down.

In this pocket paradise, creatures don't age, and natural healing take place at twice the normal rate. Fresh water and food (fruits and vegetables only) are in abundance. The climate is comfortable for all creatures involved.

Activating the rod (a standard action) causes the wielder and all creatures touching the rod to be transported instantaneously to the paradise. Members of large groups can hold hands or otherwise maintain physical contact, allowing all connected creatures in a circle or a chain to be affected by the rod. Unwilling creatures \my{cannot be transported}. If \my{a creature is unwilling}, other creatures beyond that point in a chain can still be affected by the rod.

When the rod's effect expires or is dispelled, all the affected creatures instantly reappear in the location they occupied when the rod was activated. If something else occupies the space that a traveler would be returning to, then his body is displaced a sufficient distance to provide the space required for reentry. The rod's possessor can dismiss the effect whenever he wishes before the maximum time period expires, but the rod can only be activated once per week.

\prereq{Craft Rod, \my{create food and water, endure elements, plane shift}}

% 7,500 for tent, 8,000 for cha, 4000 for garb. All related.
\itemdescription{Splendor}{15,500}{12th}{\x}{12th}{Moderate conjuration, enchantment, and transmutation}{Standard (command)} The possessor of this rod gains a \plus4 enhancement bonus to her Charisma score for as long as she holds or carries the item. Once per day, the rod creates and garbs her in clothing of the finest fabrics, plus adornments of furs and jewels.

Apparel created by the magic of the rod remains in existence for 12 hours. However, if the possessor attempts to sell or give away any part of it, to use it for a spell component, or the like, all the apparel immediately disappears. The same applies if any of it is forcibly taken from her.

The value of noble garb created by the rod ranges from 7,000 to 10,000 gp (1d4\plus6 \mtimes 1,000 gp) - 1,000 gp for the fabric alone, 5,000 gp for the furs, and the rest for the jewel trim (maximum of twenty gems, maximum value 200 gp each).

In addition, the rod has a second special power, usable once per week. Upon command, it creates a palatial tent - a huge pavilion of silk 60 feet across. Inside the tent are temporary furnishings and food suitable to the splendor of the pavilion and sufficient to entertain as many as one hundred persons. The tent and its trappings last for one day. At the end of that time, the tent and all objects associated with it (including any items that were taken out of the tent) disappear.

\prereq{Craft Rod, eagle's splendor, fabricate, major creation}

%Price: Weapon: 5000
% Thunder: 1st level spell *2 for immediate *0.75 for only usable after hit = 3000
% Lightning: 3rd level spell, 1/day = 4,500
% Thunder and Lightning: 5th level spell, 1/week: 10,000
\itemdescriptiondc{Thunder and Lightning}{20,000}{13th}{\x}{12th}{Moderate evocation}{Immediate (command) or standard (command)}{13} Constructed of iron set with silver rivets, this rod has the properties of a \my{\plus2 shocking} light mace. Its other magical powers are as follows.
\begin{itemize*}
\item Thunder: Once per day, when the rod strikes a foe, you can say a command word to cause thunder to rumble out from the rod. If the foe is bloodied, it becomes stunned; a healthy foe is staggered instead (Fortitude negates). The effect lasts for 1 round. Activating this effect requires an immediate action.
\item Lightning: Once per day as a standard action, the wielder can cause the rod to shoot out a line of lightning, as the \spell{lightning bolt} spell (10d6 points of electricity damage, Reflex half) to a range of 100 feet.
\item Thunderbolt: Once per week as a standard action, the wielder of the rod can combine the thunder described above with \my{the lightning bolt described above}. The \my{thunder} affects all within 10 feet of the bolt.
\end{itemize*}

\prereq{Craft Rod, Craft Magic Arms and Armor, lightning bolt, \my{sound burst}}

%Price: 9000 for weapon, 16000 for poison
\itemdescriptiondc{Viper}{24,000}{14th}{\x}{12th}{Moderate necromancy}{Standard (command)}{14}
\parhead{Viper} This rod strikes as a \my{\plus3} heavy mace. Once per day, upon command, the head of the rod becomes that of an actual serpent for 10 minutes. During this period, any successful strike with the rod deals its usual damage and also poisons the creature hit. The poison deals 1d10 points of Constitution damage immediately (Fortitude negates) and another 1d10 points of Constitution damage 1 minute later (Fortitude negates).

\prereq{Craft Rod, Craft Magic Arms and Armor, poison}

%Price: 13500 for contagion, 1000 for weapon
\itemdescriptiondc{Withering}{10,000}{10th}{\x}{10th}{Moderate necromancy}{\x}{13} A rod of withering acts as a \plus1 light mace that deals no hit point damage \my{when used to attack}. Instead, the wielder deals 1d4 points of Strength damage and 1d4 points of \my{Dexterity} damage to any creature she hits with the rod. If she scores a critical hit, the damage from that hit is permanent ability drain. In either case, the defender negates the effect with a Fortitude save.

\prereq{Craft Rod, Craft Magic Arms and Armor, contagion}

%Price: As 3rd level spell (based on rough power level of effects) /2 for random effects
\itemdescriptiondc{Wonder}{9,000}{9th}{\x}{10th}{Moderate (all schools)}{Standard (command)}{14}
\parhead{Wonder} A rod of wonder is a strange and unpredictable device that randomly generates any number of weird effects each time it is used. (Activating the rod is a standard action.) Typical powers of the rod include the following. Unless otherwise specified, any effects of the rod wear off after the normal duration of the spell or after 24 hours have passed, whichever comes first.
\begin{dtable}
\begin{tabularx}{\columnwidth}{l >{\lcol}X}
\thead{d\%} & \thead{Wondrous Effect} \\
01-05 & \spell{Slow} creature pointed at for 1 minute (Will negates). \\
06-10 & \spell{Faerie fire} surrounds the target. \\
11-15 & Deludes wielder for 1 round into believing the rod functions as indicated by a second die roll (no save). \\
16-20 & \spell{Gust of wind} as the spell. \\
21-25 & Wielder learns target's surface thoughts (as with \spell{detect thoughts}) for 1d4 rounds (Will negates). \\
26-30 & \spell{Stinking cloud} at close (30 ft.) range (Fortitude negates). \\
31-33 & Heavy rain falls for 1 round in large (50 ft.) radius centered on rod wielder. \\
34-36 & Summon an animal, as \spell{summon nature's ally} - a rhino (01-25 on d\%), elephant (26-50), or mouse (51-100). \\
37-46 & \spell{Lightning bolt}, 6d6 damage (Reflex half). \\
47-49 & Stream of 600 large butterflies pours forth and flutters around for 2 rounds, blinding everyone (including wielder) within 25 ft. (Reflex negates). \\
50-53 & \spell{Enlarge person} if within medium (100 ft.) range of rod (Fortitude negates). \\
54-58 & \spell{Darkness}, centered 30 ft. away from rod. \\
59-62 & Grass grows in 160-sq.-ft. area before the rod, or grass existing there grows to ten times normal size. \\
63-65 & Turn ethereal any nonliving object of up to 1,000 lb. mass and up to 30 cu. ft. in size. \\
66-69 & Reduce wielder, as \spell{reduce person} (no save). \\
70-79 & \spell{Fireball} at target or 100 ft. straight ahead, 6d6 damage (Reflex half). \\
80-84 & \spell{Invisibility} covers rod wielder. \\
85-87 & Leaves grow from target if within medium (100 ft.) range of rod. \\
88-90 & 10-40 gems, value 1 gp each, shoot forth in a 30-ft.-long stream. Each gem deals 1 point of damage to any creature in its path: Roll 5d4 for the number of hits and divide them among the available targets. The gems disappear after 1 hour. \\
91-95 & Shimmering colors dance and play over a 40-ft.-by-30-ft. area in front of rod. Creatures therein are blinded for 1d6 rounds (Fortitude negates). \\
96-97 & Wielder (50\% chance) or target (50\% chance) turns permanently blue, green, or purple (no save). \\
98-100 & \spell{Transmute flesh and stone}, reversing the target's current state, if target is within medium (100 ft.) range (Fortitude negates).
\end{tabularx}
\end{dtable}

\prereq{Craft Rod, confusion, creator must be chaotic}
