\chapter{Noncore Changes}

\section{General Changes}
%5-level prestige classes should generally require 6th level. 10-level prestige classes should generally require 5th level.

\subsection{Classes}
\begin{itemize*}
\item Hit Points: d4 and d6 HD become 4 HV. d8 HD becomes 5 HV, d10 HD becomes 6 HV, and d12 HD becomes 7 HV.
\item Skill Points: 2 \add Int skill points becomes 2 skill points. 4\add skill points becomes 4 skill points, 6\add skill points becomes 8 skill points, and 8\add skill points becomes 12 skill points.
\end{itemize*}

\subsection{Feats}
\begin{itemize*}
\item {[}Ambush] feats no longer require sacrificing sneak attack dice. They give rogues extra options on their ambush attacks.
\item {[}Wild] feats do not exist.
\item Any feat that required Point Blank Shot requires Precise Shot instead.
\end{itemize*}

\subsection{Spells}
\begin{itemize*}
\item All non-costly spell components and focuses are removed.
\item Assassin-only spells have no verbal components.
\item Paladin-only spells have no somatic components.
\item Ranger-only spells have no verbal components.
\end{itemize*}

\section{Monster Manual}
\begin{itemize*}
\item Djinni: Wind Walk is self-only.
\item Nightmare: Astral Projection and Etherealness are self-only.
\item Outsiders: Creatures with outsider racial hit dice get automatic proficiency with any weapon groups they are described as wielding. Players with outsider HD get proficiency with any weapon group of their choice.
\item Undead are no longer immune to sneak attacks.
\item A creature with only secondary natural attacks can make a single attack with one secondary natural attack as if it were a primary attack, taking no penalty and adding its full Strength as bonus damage.
\item Multiattack requires base attack bonus \plus6.
\end{itemize*}

\section{Complete Adventurer}
\subsection{Prestige Classes}
\subsubsection{Animal Lord}
\cd{Requirements} Base attack bonus \plus4 instead of \plus5.
\cd{Animal Bond} The bonus is a competence bonus.
\cd{First Totem} The bonus is a competence bonus.
\cd{Wild Aspect} An animal lord's class levels stack with his druid levels, if any, when determining his wild aspect ability. This replaces the listed values for wild aspect uses per day. An animal lord adds the wild aspect ability appropriate to his totem to the list of wild aspects he can take.
\subcf{Sharklord} The bite damage increases to 1d10 at 7th level.
\subsubsection{Bloodhound}
\cd{Requirements} Does not require Move Silently or Endurance. Requires 8 ranks in Survival and Skill Focus (Survival).
\cd{Mark} The skill bonuses are competence bonuses.
\cd{Tenacious Pursuit} The bonuses are competence bonuses.
\cd{Crippling Strike} As with the rogue class feature, only one crippling strike attack can be made per round.
\cd{Freedom of Movement} The effect lasts for a total time per day of one round per two class levels plus the bloodhound's Wisdom.
\subsubsection{Daggerspell Mage}
\cd{Requirements} Concentration 8 ranks, Weapon Focus (light blade), Two-Weapon Fighting, 2nd level arcane spellcasting, Sneak attack \plus1d6.
\subsubsection{Daggerspell Shaper}
Does not exist.
\subsubsection{Dread Pirate}
\cd{Seamanship} The bonus is a competence bonus.
\cd{Fearsome Reputation} The bonus is a circumstance bonus.
\cd{Rally the Crew} The bonus is an enhancement bonus. It is activated as a swift action. The value of the bonus and the number of uses per day increases at 5th, 7th, and 9th level. It gives no bonus to saving throws.
\cd{Sneak Attack} The amount of extra damage increases by 1d6 at 5th, 7th, and 9th level.
\cd{Steady Stance} The bonus is a circumstance bonus.
\cd{Motivate the Scum} The bonus is an enhancement bonus.
\subsubsection{Dungeon Delver}
\cd{Requirements} Disable Device 8 ranks, Hide 4 ranks, Move Silently 4 ranks, Search 8 ranks, Alertness, Blind-Fight, trapfinding.
\cd{Deep Survival} The bonus is a competence bonus.
\cd{Trap Sense} Does not exist.
\cd{Augury} Activated as a swift action.
\cd{Darkstalker} At 5th level, when the dungeon delver hides, creatures with blindsense, blindsight, scent, or tremorsense do not automatically detect him. They must make Spot and Listen checks to detect him as if they did not have those abilities.
\cd{Blindsense} This is a constant ability that does not need to be activated. It is gained at 7th level.
\cd{Blindsight} At 10th level, the dungeon delver gains constant blindsight out to 30 feet.
\subsubsection{Exemplar}
\cd{Requirements} 10 ranks in any three skills, Skill Focus (any).
\cd{Skill Artistry} The bonus is a \plus10 competence bonus. It can apply to any skill in which the exemplar has at least 10 ranks.
\cd{Bonus Feat} The feat can be any [Skill] feat.
\cd{Sustaining Presence} The bonus does not apply to Concentration checks.
\cd{Persuasive Performance} (Su) ability. This does not mimic a Diplomacy check. Each viewer is affected by a \spell{charm monster} spell with a caster level equal to your character level and a saving throw DC equal to your skill check.
\cd{Perfect Self} Does not exist.
\cd{Perfect Mind} At 10th level, an exemplar's mind becomes virtually immune to outside influence when she concentrates on her chosen area of focus. Whenever she would make a Will save, she can make a skill check instead and use the result in place of the Will save.
\subsubsection{Fochlucan Lyrist}
\cd{Requirements} Knowledge (nature) 8 ranks), Perform (any) 8 ranks, Speak Language (Druidic), any alignment with a neutral component, ability to cast 2nd level arcane and divine spells, bardic knowledge, inspire competence.
\cd{Core Features} 3/4 base attack bonus.
\subsubsection{Ghost-Faced Killer}
\cd{Requirements} Base attack bonus \plus4, Hide 4 ranks, Intimidate 8 ranks, Move Silently 4 ranks, Power Attack
\cd{Ghost Step} A ghost-faced killer can use this ability one additional time per every two class levels gained above 1st.
\cd{Sudden Strike} The bonus damage increases by 1d6 every even level.
\cd{Frightful Attack} Can be used a number of times per day equal to half the ghost-faced killer's class level \add his Charisma. The Will save DC against dying of fear increases by the damage bonus from Power Attack on the frightful attack. Creatures with Hit Values that exceed the ghost-faced killer's character level \add his Charisma are shaken for one round per class level instead of dying.
\subsubsection{Highland Stalker}
Does not exist.
\subsubsection{Maester}
\cd{Spells per Day/Spells Known} Increases magic level and caster level at every level.
\cd{Bonus Feats} Gained at 2nd and 4th level.
\cd{Improved Crafting} At 5th level, a maester no longer gains any negative levels from crafting magic items.
\subsubsection{Master of Many Forms}
Does not exist.
\subsubsection{Nightsong Enforcer}
\cd{Requirements} Base attack bonus \plus4, Hide 8 ranks, Listen 4 ranks, Move Silently 8 ranks, Spot 4 ranks, Alertness, evasion.
\cd{Combat Tricks} Nightsong enforcer levels stack with rogue levels for determining when the nightsong enforcer gains combat tricks.
\cd{Skill Teamwork} The bonus is an enhancement bonus. The nightsong enforcer only grants a bonus to skills that she is trained with.
\subsubsection{Nightsong Infiltrator}
\cd{Requirements} Disable Device 4 ranks, Hide 8 ranks, Move Silently 8 ranks, Search 4 ranks, Stealthy, evasion.
\cd{Skill Tricks} Nightsong enforcer levels stack with rogue levels for determining when the nightsong enforcer gains skill tricks.
\cd{Teamwork Trap Sense} Renamed to Teamwork Danger Sense. She gains a \plus4 competence bonus to initiative checks that increases by \plus1 at 4th, 7th, and 10th level. If the infiltrator has danger sense from other classes, her infiltrator levels stack with her levels in those other classes. All allies within 30 feet of the infiltrator also gain half that bonus as an enhancement bonus.
\cd{Trapfinding} As the rogue class feature.
\cd{Steady Stance} The bonus is a circumstance bonus.
\cd{Teamwork Infiltration} The bonus is an enhancement bonus.
\cd{Break Away} The bonus is a circumstance bonus to her dodge modifier.
\cd{Grant Move Action} This ability can be used a number of times per day equal to the infiltrator's Intelligence. At 9th level, the ability can be used as a move action.
\cd{Skill Mastery} Applies to Disable Device, Hide, Move Silently, and Search checks.
\subsubsection{Ollam}
\cd{Requirements} Knowledge (history) 8 ranks, Knowledge (any other) 8 ranks, Perform (oratory) 4 ranks.
\cd{Spells per Day/Spells Known} Advances casting at every level.
\cd{Inspire Competence} The bonus is an enhancement bonus.
\cd{Inspire Resilience} The bonus is an enhancement bonus.
\subsubsection{Shadowbane Inquisitor}
\cd{Requirements} Lawful good, base attack bonus \plus4, Gather Information 4 ranks, Knowledge (religion) 4 ranks, Sense Motive 8 ranks, Investigator, detect evil class feature or ability to cast detect evil as a spell, turn undead, sneak attack \plus1d6
\cd{Sacred Stealth} The bonus is an enhancement bonus.
\cd{Smite} The shadowbane inquisitor's levels stack with paladin levels for the purpose of the smite evil and improved smiting abilities. This replaces the original benefit.
\cd{Merciless Purity} Instead of gaining a bonus to saving throws, the inquisitor gains temporary hit points equal to twice her character level.
\cd{Righteous Fervor} The bonus is an enhancement bonus equal to the inquisitor's Charisma bonus (minimum \plus1).
\cd{Burning Light} Deals 1d6 damage per two turning levels. Creatures vulnerable to bright light take 1d6 damage per turning level instead.
\subsubsection{Shadowbane Stalker}
\cd{Requirements} Lawful good, Gather Information 8 ranks, Search 4 ranks, Sense Motive 4 ranks, detect evil class feature or ability to cast detect evil as a spell, turn undead, sneak attack\plus1d6
\cd{Sacred Stealth} The bonus is an enhancement bonus.
\cd{Sacred Defense} The bonus is an enhancement bonus to your dodge modifier.
\cd{Sacred Strike} The bonus is a circumstance bonus.
\subsubsection{Shadowmind}
Does not exist.
\subsubsection{Spymaster}
\cd{Requirements} Requires Deceitful instead of Skill Focus (Bluff).
\subsubsection{Streetfighter}
\cd{Requirements} Base attack bonus \plus5, Bluff 4 ranks, Intimidate 4 ranks, Knowledge (local) 4 ranks, Blind-Fight
\cd{Always Ready} Renamed to danger sense, which functions as the rogue class feature.
\cd{Combat Tricks} Levels in streetfighter stack with rogue levels for the purpose of gaining combat tricks.
\cd{Improved Dirty Trick} Gained at 2nd level.
\cd{Stand Tough} Gained at 3rd level. Can be used a number of times per day equal to the streetfighter's Constitution (minimum 1).
\cd{Sneak Attack} Gained at 2nd and 4th level.
\cd{Uncanny Dodge} Gained at 4th level.
\cd{Defensive Roll} Gained at 5th level. Can be used a number of times per day equal to the streetfighter's Constitution or Intelligence, whichever is higher.
\subsubsection{Tempest}
Does not exist.
\subsubsection{Thief-Acrobat}
\cd{Fast Acrobatics} If the thief-acrobat already has the Fast Acrobatics skill trick, he gains a different skill trick from the rogue list instead.
\cd{Kip Up} If the thief-acrobat already has the Kip Up skill trick, he gains a different skill trick from the rogue list instead.
\cd{Steady Stance} The bonus is a circumstance bonus.
\cd{Agile Fighting} The bonus is a competence bonus to your dodge modifier.
\cd{Defensive Roll} At 5th level, this can be used a number of times per day equal to the streetfighter's Dexterity or Intelligence, whichever is higher.
\subsubsection{Vigilante}
\cd{Requirements} Requires Investigator instead of Alertness.
\cd{Smite the Guilty} Can be used a number of times per day equal to half the vigilante's level \add his Charisma.
\cd{Quick Hide} The bonus is a competence bonus.
\subsubsection{Virtuoso}
\cd{Requirements} Requires 8 ranks in Perform instead of 10.
\cd{Spellcasting:} Progresses casting at every level.
\cd{Bardic Music (move):} At 1st level, the virtuoso can initiate bardic music or a virtuoso performance as a move action instead of as a standard action.
\cd{Lyrical Freedom} At 2nd level, the virtuoso gains lyrical freedom, as the bard ability.
\cd{Combine Songs} At 4th level, the virtuoso gains combine songs, as the bard ability.
\cd{Bardic Music (swift):} At 6th level, the virtuoso can initiate bardic music or a virtuoso performance as a swift action instead of as a move or standard action.
\subsubsection{Wild Plains Outrider}
\cd{Requirements} Ride 8 ranks, Mounted Combat, animal companion large enough to serve as a mount, or a paladin's special mount.

\subsection{Feats}
\begin{itemize*}
\item Appraise Magic Value: [Skill] feat. No material components are required.
\item Ascetic Knight: Your paladin and monk levels stack when determining the extra damage and saving throw DC of your smite evil ability.
\item Ascetic Mage: Prerequisites: \Ki strike (magic), ability to cast 1st level arcane spells.
\item Brachiation: [Skill] feat.
\item Brutal Throw: [Combat] feat.
\item Combat Intuition: The Sense Motive check is made as a swift action. The attack bonus is a circumstance bonus.
\item Danger Sense: [Combat] feat. You must decide to reroll before the first round of the encounter starts.
\item Death Blow: [Combat] feat.
\item Deft Opportunist: Does not exist.
\item Devoted Inquisitor: When you successfully use your sneak attack ability and your smite evil ability against the same foe in a single attack, you increase the DC of your smite evil ability by 2. This replaces the dazing ability.
\item Devoted Performer: Your paladin and monk levels stack when determining the extra damage and saving throw DC of your smite evil ability.
\item Devoted Tracker: Your paladin and ranger levels stack when determining the extra damage and saving throw DC of your smite evil ability.
\item Disguise Spell: Onlookers must match or exceed your check result with a Spot check if the spell has somatic components, or with a Listen check if the spell has verbal components. A disguised spell can be identified by a Spellcraft check, but only by someone who realizes you're casting a spell.
\item Dual Strike: Does not exist.
\item Expert Tactician: [Combat] feat.
\item Extra Music: Use the core feat.
\item Extraordinary Concentration: [Skill] feat.
\item Extraordinary Spell Aim: [Skill] feat. If the spell's area is an emanation from you, you cannot exclude yourself from its area.
\item Goad: [Combat] feat.
\item Hear the Unseen: [Skill] feat.
\item Jack of All Trades: [Skill] feat.
\item Leap Attack: [Combat] feat. You gain extra damage with the Power Attack feat equal to the penalty you take on the attack roll.
\item Open Minded: Use the core feat.
\item Oversized Two-Weapon Fighting: [Combat] feat.
\item Power Throw: [Combat] feat. You can use the Power Attack feat with thrown weapons.
\item Quick Reconnoiter: [Skill] feat. The bonus is a competence bonus.
\item Razing Strike: The attack and damage bonuses are circumstance bonuses.
\item Staggering Strike: [Ambush] feat. Prerequisite: Sneak attack \plus3d6.
\item Versatile Performer: [Skill] feat.
\end{itemize*}
\subsection{Spells}
\begin{itemize*}
\item Arrowmind: Lasts for 1 round.
\item Balancing Lorecall: The bonus is an enhancement bonus.
\item Critical Strike: The bonus is an circumstance bonus.
\item Daggerspell Stance: The bonus is an enhancement bonus.
\item Dissonant Chord: Creatures that fail the Fortitude save are also deafened for 1 minute.
\item Embrace the Wild: Does not exist.
\item Focusing Chant: Gives a \plus2 enhancement bonus.
\item Grave Strike: Does not exist.
\item Harmonic Chorus: The bonus is an enhancement bonus.
\item Healthful Rest: The rate at which the subjects recover from all ailments, including ability damage, is also doubled. This stacks with the use of the Heal skill.
\item Improvisation: The bonus is an circumstance bonus.
\item Insidious Rhythm: The Concentration DC is equal Insidious Rhythm's save DC \add double the level of the spell being cast.
\item Insightful Feint: The bonus is a circumstance bonus, and it applies to feint checks.
\item Instant Locksmith: The bonus is a circumstance bonus.
\item Instant Search: The spell allows you to search every square within 10 feet of you with a single check. The bonus is a circumstance bonus.
\item Joyful Noise: Evocation spell.
\item Listening Lorecall: The bonus is an enhancement bonus. 10 or more ranks in Listen grants blindsense out to 60 feet instead of blindsight.
\item Nature's Favor: The bonus is an enhancement bonus.
\item Nightstalker's Transformation: The bonuses are enhancement bonuses. You gain proficiency with crossbows, light blades, and blunt weapons.
\item Shadow Form: The bonuses are enhancement bonuses.
\item Tactical Precision: The bonus is a circumstance bonus. Allies get the bonus if two affected allies overwhelm the same creature.
\item War Cry: The bonuses are enhancement bonuses.
\item Wraithstrike: Applies to the next attack you make, not all attacks in the round.
\end{itemize*}


\section{Complete Arcane}
\subsection{Prestige Classes}
\subsubsection{Acolyte of the Skin}
\cd{Requirements} Any nongood alignment, Knowledge (the planes) 8 ranks, ability to cast 2nd level spells, must have made peaceful contact with an evil outsider, must undergo the Ritual of Bonding.
\cd{Wear Fiend} The bonuses are competence bonuses.
\cd{Poison} The caster level is equal to the acolyte's caster level. Can be used an additional time per day at 4th, 7th, and 10th level (instead of 5th level).
\cd{Spells per Day/Spells Known} Increases magic level and caster level at every level.
\cd{Fiendish Glare} Can be used an additional time per day at 6th and 9th level.
\cd{Skin Adaptation} The bonus is a competence bonus.
\cd{Glare of the Pit} Can be used an additional time per day at 10th level.
\cd{Summon Fiend} Gained at 8th level.
\subsubsection{Alienist}
\cd{Requirements} Able to cast at least one summoning spell of 2nd level or higher instead of 3rd level or higher.
\cd{Alien Blessing} The bonus is a competence bonus. It increases to \plus2 at 5th level and to \plus3 at 9th level.
\cd{Mad Certainty} Grants one additional hit point per HV. This applies to future HV gained.
\cd{Insane Certainty} Grants one additional hit point per HV. This applies to future HV gained.
\subsubsection{Argent Savant}
\cd{Requirements} Knowledge (arcana) 4 ranks, Spellcraft 8 ranks, able to cast at least three spells with the force descriptor, at least one of which must be 3rd level or higher.
\cd{Expanded Spell Knowledge} At 1st, 3rd, and 5th level, the argent savant adds a spell to their spell list from the list of specialist-only spells. The spell must be a force spell.
\cd{Force Specialization} Gained at 2nd level. If the spell does not deal damage in dice, this adds 2 damage. The bonuses are competence bonuses.
\cd{Force Armor} Gained at 4th level.
\cd{Ablate Force} Gained at 1st level. Reduces the damage by twice the argent savant's class level.
\cd{Ablate Impact} A 5th level argent savant gains damage reduction 5/magic.
\cd{Unbind Force} Can also be used to counterspell force spells.
\subsubsection{Blood Magus}
\cd{Requirements} Does not require Great Fortitude.
\cd{Spells per Day/Spells Known} Increases magic level and caster level at every level.
\cd{Blood Component} Also increases save DC by 1. The bonuses are circumstance bonuses.
\cd{Death Knell} Can be used an additional time per day at 5th, 7th, and 9th level.
\cd{Bloodseeking Spell} Increases damage by one point per die of damage, or by 3 if the spell does not do dice of damage.
\cd{Thicker Than Water} Gained at 2nd level. Increases by 1 at every even level.
\cd{Awaken Blood} Gained at 7th level. Can be used an additional time per day at 9th level.
\cd{Infusion} Gained at 8th level.
\cd{Greater Blood Component} A 9th level blood magus can increase a spell's caster level and save DC by 2 by taking 3 points of damage as a free action. The bonuses are circumstance bonuses.
\cd{Bloodwalk} Can be used a number of times per day equal to the blood magus's Constitution.
\subsubsection{Effigy Master}
Does not exist.
%\cd{Requirements} Craft (leatherworking, metalworking, or woodworking) 8 ranks, Knowledge (arcana) 4 ranks, Spellcraft 4 ranks, Craft Wondrous Item, 2nd level spellcasting.
%\cd{Spells per Day/Spells Known} Increases magic level and caster level at every level.
\subsubsection{Elemental Savant}
\cd{Requirements} Knowledge (the planes) 8 ranks, Spellcraft 8 ranks, Energy Substitution (acid, cold, electricity, or fire), able to cast at least three spells that have one of the following descriptors in common: acid, cold, electricity, or fire. One of the spells must be at least 2nd level. Must have made peaceful contact with an elemental or outsider that has an elemental subtype (air, earth, fire, or water).
\cd{Spells per Day/Spells Known} Increases magic level and caster level at every level.
\cd{Elemental Specialty} Does not exist.
\cd{Energy Penetration} Does not exist.
\cd{Energy Power} Gained at 1st level. When casting spells from her chosen element, the elemental savant gains a \plus1 competence bonus to her caster level. This bonus increases by \plus1 at 5th and 9th level.
\cd{Immunity to Sleep} Does not exist; see Elemental Resilience.
\cd{Elemental Resilience} At 2nd level, an elemental savant becomes immune to sleep. At 4th level, she becomes immune to paralysis. At 6th level, she becomes immune to poison. At 8th level, she becomes immune to stunning.
\cd{Energy Focus} Gained at 3rd level. Increases by \plus1 at 6th and 9th level. The bonus is a competence bonus.
\cd{Darkvision} Gained at 5th level.
\cd{Elemental Perfection} Does not grant immunity to critical hits or overwhelm penalties.
\subsubsection{Enlightened Fist}
\cd{Requirements} Concentration 8 ranks, Spellcraft 4 ranks, Combat Casting, Improved Unarmed Strike, Stunning Fist, ability to cast 1st level arcane spells.
\cd{Spells per Day/Spells Known} Increases magic level and caster level at every level except 1st.
\cd{Ki Strike} As the monk ability.
\cd{Fist of Energy} Activated as a swift action.
\cd{Arcane Rejuvenation} Heals damage equal to five times the spell's level.
\subsubsection{Fatespinner}
\cd{Requirements} Knowledge (arcana) 8 ranks, Profession (gambler) 4 ranks, able to cast 3rd-level arcane spells, including at least one divination spell of 1st level or higher.
\cd{Spells per Day/Spells Known} Increases magic level and caster level at every level.
\cd{Spin Fate} Each day, a fatespinner can use a number of points equal to twice his fatespinner class level. He cannot spend more than his fatespinner class level in points in a single turn.
\cd{Seal Fate} Activated as a swift action.
\subsubsection{Geometer}
\cd{Requirements} Decipher Script 8 ranks, Knowledge (arcana) 4 ranks, Search 4 ranks, Spellcraft 8 ranks, Scribe Scroll, ability to cast 3rd-level arcane spells.
\cd{Glyph of Warding} A geometer adds \spell{glyph of warding} to his spell list. He can learn and cast the spell just like any other spell.
\cd{Book of Geometry} Does not exist.
\cd{Sigilsight} Gained at 2nd level. The bonus is a competence bonus.
\cd{Simple Spellglyph} At 3rd level, the geometer gains a \plus4 circumstance bonus to Concentration checks made to cast a spell from a spellglyph.
\cd{Pass Sigil} Gained at 4th level.
\cd{Powerful Spellglyph} The bonus is a circumstance bonus.
\cd{Greater Glyph of Warding} A geometer adds \spell{greater glyph of warding} to his spell list. He can learn and cast the spell just like any other spell.
\subsubsection{Green Star Adept}
\cd{Requirements} Base attack bonus \plus4, Knowledge (arcana) 8 ranks, Knowledge (local) 4 ranks, Great Fortitude, ability to cast 1st level arcane spells. Must also drink the starmetal infusion.
\cd{Improved Caster Level} Does not exist.
\cd{Starmetal Dependency} Does not exist.
\cd{Starmetal Rigor} Does not decrease Dexterity.
\cd{Unnatural Metabolism} The bonuses are circumstance bonuses.
\cd{Fortification} Does not exist.
\cd{Emerald Perfection} The adept becomes a living construct instead of a construct. He becomes immune to poison, paralysis, disease, nausea, and energy drain. He heals half the normal damage from spells from the healing subschool. When his hit points are less than 0, but not below his Constitution score in negative hit points, he is inert. He is unconscious and helpless, and cannot perform any actions. However, he is automatically considered stable. These effects all of the normal effects.
\subsubsection{Initiate of the Sevenfold Veil}
Does not exist.
\subsubsection{Mage of the Arcane Order}
Does not exist.
\subsubsection{Master Transmogrifist}
Does not exist.
\subsubsection{Mindbender}
\cd{Requirements} Requires the ability to cast 3rd level arcane spells instead of arcane caster level 5th.
\cd{Telepathy} Gained at 3rd level.
\cd{Mindread} Gained at 1st level. It can be used twice per day at 1st level, and an additional time per day at 4th, 7th, and 10th level.
\subsubsection{Seeker of the Song}
\cd{Requirements} Knowledge (arcana) 10 ranks, Perform (any one) 10 ranks, Skill Focus (Perform [any one]), bardic music ability.
\cd{Rapture of the Song} The AC bonus is a competence bonus to dodge modifier. The bonus to saving throws is a competence bonus.
\subsubsection{Sublime Chord}
Does not exist.
\subsubsection{Suel Arcanamach}
\cd{Requirements} Requires 4 ranks in Spellcraft instead of 5.
\cd{Class Features} Provides full base attack bonus progression.
\subsubsection{Wayfarer Guide}
\cd{Spells per Day/Spells Known} Increases magic level and caster level at every level.
\subsubsection{Wild Mage}
Does not exist (under construction).
\subsection{Feats}
\begin{itemize*}
\item Arcane Mastery: Requires caster level 6th. You can take 10 with this feat even during combat.
\item Arcane Preparation: Does not exist.
\item Chain Spell: Can be used on any spell that targets a single creature or object. Targets are allowed saving throws to avoid damage if the original spell allowed a saving throw, but no extra saving throw is granted if the original spell did not allow a saving throw. Note that ray spells are not valid for Chain Spell, since they do not specify a single target.
\item Craft Contingent Spell: Does not exist.
\item Delay Spell: Uses a spell slot one level higher than the spell's actual level.
\item Extra Spell: This feat only gives you access to spells which are on your spell list.
\item Mage Slayer: [Combat] feat. Creatures with spell-like abilities that you threaten cannot use them defensively.
\item Nonlethal Substitution: The nonlethal damage is still of the appropriate energy type, so it is affected normally by resistances and vulnerabilities. A nonlethal spell uses a spell slot of the spell's normal level.
\item Pierce Magical Concealment: [Combat] feat.
\item Pierce Magical Protection: [Combat] feat. This also pierces and dispels any bonuses to Armor Class granted by spell-like abilities.
\item Persistent Spell: This can only apply to spells which have a duration based on caster level. Fixed duration spells cannot be persisted. "Fixed range" refers specifically to spells like Detect Magic and Arcane Sight, which are effectively personal but technically affect a fixed range away from you. It does not affect spells that affect things a fixed distance away from you. Even if other metamagic effects change how a spell can be cast, those effects cannot make a spell Persistable if it was not originally Persistable. If Persisted, a spell cannot also be Extended.
\item Sculpt Spell: Possible areas include a Small (10 ft. radius) cylinder that is 20 feet high, a Medium (20 ft.) cone, a Small (10 ft. radius) spread, or a Long (50 ft.) line.
\end{itemize*}

\section{Complete Champion}
\subsection{Feats}
\begin{itemize*}
\item Battle Blessing: Does not exist.
\item Awesome Smite: Does not exist.
\item Bestial Charge: Does not exist.
\item Fragile Construct: You can also use this ability to reduce the damage reduction of constructs by the same amount. The bonus to sunder attempts is a competence bonus.
\item Holy Potency: The bonuses are circumstance bonuses.
\item Holy Warrior: The bonus is a competence bonus.
\item Imbued Healing: The bonuses are enhancement bonuses.
\item Mitigate Suffering: Does not stack with itself when used on the same target repeatedly.
\item Protective Ward: The bonus is an enhancement bonus to deflection modifier.
\item Retrieve Spell: Only requires one other divine feat as a prerequisite.
\item Spiritual Counter: Costs 1 \add spell level turn or rebuke attempts.
\item Spontaneous Domains: Does not exist.
\item Touch of Healing: Does not exist.
\item Umbral Shroud: Gives a 20\% miss chance. The Will save DC is equal to 10 \add the highest-level darkness spell you have \add your casting attribute. The secondary benefit is a \plus1 caster level with darkness spells.
\end{itemize*}
\subsubsection{Devotion Feats}
Devotion or [Domain] feats do not exist.

\section{Complete Divine}
\subsection{Prestige Classes}
\subsubsection{Black Flame Zealot}
\cd{Requirements} Hide 8 ranks, Knowledge (religion) 4 ranks, Move Silently 8 ranks, Iron Will, ability to cast 2nd-level divine spells, sneak attack \plus1d6.
\cd{Spells per Day/Spells Known} Increases magic level and caster level at every level.
\cd{Sacred Flame} Activated as a swift action.
\subsubsection{Blighter}
Does not exist.
\subsubsection{Church Inquisitor}
\cd{Requirements} Skill requirements: Knowledge (religion) 8 ranks, Sense Motive 8 ranks.
\subsubsection{Consecrated Harrier}
\cd{Requirements} Skill requirements: Bluff 4 ranks, Gather Information 4 ranks, Sense Motive 4 ranks, Survival 4 ranks.
\cd{Blessing of Scripture} Also applies to Gather Information checks. The bonus is a circumstance bonus.
\subsubsection{Contemplative}
\cd{Class Features} The class is a five level class.
\cd{Divine Health} A contemplative is also immune to poisons.
\cd{Bonus Domain} Gained at 4th level.
\cd{Divine Body} Does not exist. See Divine Health.
\cd{Divine Soul} Gained at 5th level.
\cd{Eternal Body} Gained at 5th level.
\cd{Mystic Union} Does not exist.
\subsubsection{Divine Crusader}
\cd{Requirements} Alignment must match chosen deity, base attack bonus \plus7, Knowledge (religion) 8 ranks, Weapon Focus (in chosen deity's favored weapon group)
\cd{Domain Power} Gained at 1st level. As the cleric ability, but only for the divine crusader's chosen domain.
\cd{Channeled Domain Power} Gained at 3rd level. As the cleric ability, but only for the divine crusader's chosen domain. A divine crusader without the ability to turn undead gains 3 \add Charisma uses of positive (or negative) energy that can be used to power channeled domain abilities, but not to turn or rebuke undead.
\cd{Resistance to Electricity} Does not exist.
\cd{Greater Domain Power} Gained at 5th level. As the cleric ability, but only for the divine crusader's chosen domain..
\cd{Weapon Specialization} Does not exist.
\cd{Resistance to Acid} Does not exist.
\cd{Greater Channeled Domain Power} Gained at 7th level. As the cleric ability, but only for the divine crusader's chosen domain.
\cd{Darkvision} Gained at 6th level.
\cd{Domain Mastery} Gained at 9th level. As the cleric ability, but only for the divine crusader's chosen domain.
\subsubsection{Divine Oracle}
\cd{Scry Bonus} The bonus is a competence bonus. Increases by \plus1 at 5th and 9th level.
\cd{Domain Power} Gained at 1st level. As the cleric ability, but only for the Oracle domain.
\cd{Trap Sense} Replaced with danger sense, as the rogue ability.
\cd{Channeled Domain Power} Gained at 3rd level. As the cleric ability, but only for the Oracle domain. A divine oracle without the ability to turn undead gains 3 \add Charisma uses of positive (or negative) energy that can be used to power channeled domain abilities, but not to turn or rebuke undead.
\cd{Divination Enhancement} Does not exist (see the Oracle domain).
\cd{Greater Domain Power} Gained at 5th level. As the cleric ability, but only for the Oracle domain.
\cd{Improved Uncanny Dodge} As the barbarian ability.
\cd{Greater Channeled Domain Power} Gained at 7th level. As the cleric ability, but only for the Oracle domain.
\cd{Greater Uncanny Dodge} Gained at 8th level.
\cd{Domain Mastery} Gained at 9th level. As the cleric ability, but only for the Oracle domain.
\subsubsection{Entropomancer}
\cd{Requirements} Concentration 4 ranks, Knowledge (arcana) 4 ranks, Great Fortitude, ability to cast 3rd-level spells
\cd{Shard of Entropy} Lasts for a number of rounds equal to the entropomancer's class level \add his Wisdom.
\cd{Spells per Day/Spells Known} Can progress arcane casting.
\cd{Entropic Field} Lasts for a number of rounds equal to the entropomancer's class level \add his Wisdom.
\subsubsection{Evangelist}
\cd{Requirements} Skill Requirements: Bluff, Diplomacy, or Intimidate 8 ranks, Perform (oratory) 8 ranks, Skill Focus (Bluff, Diplomacy, or Intimidate).
\cd{Great Orator} Inspire Hope gives an enhancement bonus.
\cd{Fast Talk} Does not exist.
\cd{Righteous Conviction} At 2nd level, the evangelist adds his Charisma to his saving throws against mind-affecting spells and effects.
\subsubsection{Geomancer}
\cd{Ley Lines} The bonus is a circumstance bonus.
\cd{Drift} The bonuses are competence bonuses unless otherwise stated.
\cd{Stage 2 Drifts} The coat of white fur gives a \plus4 circumstance bonus to Hide checks in snowy areas.
\cd{Stage 3 Drifts} Eyes as sharp as an eagle or owl grants circumstance bonuses. The hawklike talons do not grant Combat Finesse.
\cd{Stage 4 Drifts} The rage similar to a wolverine makes geomancer levels stack with levels in classes that provide rage for the purpose of the temporary hit points gained. Pouncing like a leopard grants a charge, if it is made in the first round of combat.
\cd{Stage 5 Drifts} The unicorn horn grants a circumstance bonus. Feathered or batlike wings grant a fly speed equal to your base land speed. Curling into a spiny ball is done as part of taking a total defense action. It grants a \plus4 circumstance bonus to your natural armor in addition to the normal bonus from the total defense action. Becoming as graceful as a pixie also grants a \plus1 competence bonus to your dodge modifier. If your teeth drip poison, you can combine the benefits of that drift with any bite attack you already have (such as from other drifts). Having your skin become tree bark grants you a \plus2 competence bonus to your natural armor modifier.
\subsubsection{Holy Liberator}
\cd{Requirements} Skill Requirements: Diplomacy 4 ranks, Sense Motive 4 ranks.
\cd{Spells per Day} Increased by 1 at each spell level.
\cd{Smite Evil} As the paladin ability.
\cd{Improved Smiting} As the paladin ability. 
\cd{Aura of Resolve} As the paladin ability.
\cd{Divine Grace} As the paladin ability.
\cd{Aura of Determination} Gained at 7th level. As the paladin ability.
\cd{Celestial Companion} As the paladin ability, but with the celestial template.
\subsubsection{Hospitaler}
Does not exist.
\subsubsection{Pious Templar}
Does not exist.
\subsubsection{Radiant Servant of Pelor}
\cd{Requirements} Heal 8 ranks, Knowledge (religion) 4 ranks, Extra Turning, ability to cast 2nd level divine spells, access to the Sun domain, must worship Pelor.
\cd{Extra Greater Turning} Does not exist.
\cd{Greater Domain Power} Gained at 1st level. As the cleric ability, but only for the Sun domain.
\cd{Greater Channeled Domain Power} Gained at 4th level. As the cleric ability, but only for the Sun domain.
\cd{Empower Healing} Gained at 3rd level. Applies when casting any spell from the healing subschool.
\cd{Aura of Warding} Gained at 2nd level. The radiant servant gains a \plus2 competence bonus to Will saving throws. All allies within 10 feet of him gain a \plus2 circumstance bonus to Will saving throws.
\cd{Bonus Domain} Does not exist.
\cd{Maximize Healing} Applies when casting any spell from the healing subschool.
\cd{Positive Energy Burst} Deals 1d6 points of damage per turning level. The save DC is equal to 10 \add half turning level \add Cha.
\cd{Domain Mastery} Gained at 9th level. As the cleric ability, but only for the Sun domain.
\cd{Supreme Healing} Applies when casting any spell from the healing subschool.
\subsubsection{Rainbow Servant}
Does not exist.
\subsubsection{Sacred Exorcist}
\cd{Requirements} Knowledge (religion) 8 ranks, Knowledge (the planes) 8 ranks, ability to cast \spell{magic circle against evil}
\cd{Resist Possession} The bonuses are circumstance bonuses.
\cd{Chosen Foe} The bonuses are circumstance bonuses.
\cd{Greater Domain Power} Gained at 5th level. As the cleric ability, but only for the Good domain. The sacred exorcist need not have the Good domain.
\cd{Consecrated Presence} Gained at 6th level.
\cd{Domain Mastery} Gained at 10th level. As the cleric ability, but only for the Good domain. The sacred exorcist need not have the Good domain.
\subsubsection{Sacred Fist}
\cd{AC Bonus} Does not exist. See \ki ward.
\cd{Ki Ward} As the monk ability. Sacred fist levels stack with monk levels for the purpose of determining the AC bonus.
\cd{Sacred Flames} Activated as a swift action.
\cd{Blindsense} Extends to a radius of 30 feet.
\cd{Inner Armor} Activated as a swift action. The effect lasts for 1 minute. The bonuses are enhancement bonuses.
\subsubsection{Seeker of the Misty Isle}
\cd{Spells per Day/Spells Known} Increases magic level and caster level at every level.
\cd{Corellon's Perception} Checks to rehear sounds can be made as a swift action.
\subsubsection{Shining Blade of Heironeous}
\cd{Adaptations} Most deities can have their own equivalent of the shining blade class, which may grant different abilities.
\cd{Requirements} Lawful good, base attack bonus \plus5, base Will save \plus3, Knowledge (religion) 8 ranks, ability to cast 1st-level divine spells, must worship Heironeous.
\cd{Spells per Day/Spells Known} Increases magic level and caster level at every level.
\cd{Shock Blade} Activated as a swift action. The effect lasts for 1 minute. Can be used a number of times per day equal to half the shining blade's class level \add his Charisma.
\cd{Holy Blade} Activated as a swift action. The effects last for 1 minute. Can be used a number of times per day equal to half the shining blade's class level \add his Charisma.
\cd{Brilliant Blade} Activated as a swift action. The effects last for 1 minute. Can be used a number of times per day equal to half the shining blade's class level \add his Charisma.
\cd{Infinite Blade} The shining blade can use his brilliant blade ability at will.
\subsubsection{Stormlord}
\cd{Requirements} Base Fortitude save \plus3, Endurance, Great Fortitude, Weapon Focus (spears), ability to cast 2nd-level divine spells, must have been hit by lightning, whether magical or mundane, and survived.
\cd{Enhanced Javelins} The bonus increases to \plus3 at 10th level.
\cd{Elemental Conflagration} Renamed to storm of elemental fury.
\subsubsection{Temple Raider of Olidammara}
\cd{Requirements} Any chaotic alignment, Disable Device 4 ranks, Knowledge (religion) 2 ranks, Search 8 ranks, trapfinding, must worship Olidammara.
\cd{Spells per Day} Increased by 1 at each spell level.
\cd{Trapfinding} Does not exist.
\cd{Trap Sense} Does not exist.
\cd{Intuitive Trapsmith} Gained at 1st level. A temple raider of Olidammara can add his Wisdom instead of his Intelligence to Search and Disable Device checks.
\cd{Sneak Attack} Gained at 1st level. As the rogue ability. Increases at 4th, 7th, and 10th level.
\cd{Danger Sense} Gained at 2nd level. As the rogue ability. Improves by \plus1 at 5th and 8th level.
\cd{Skill Mastery} Gained at 8th level.
\cd{Greater Uncanny Dodge} Gained at 9th level. As the rogue ability.
\cd{Luck Domain} Does not exist.
\cd{Luck of Olidammara} At 10th level, a temple raider of Olidammara can reroll any attack roll, check, or saving throw he makes as a free action. He can use this ability a number of times per day equal to his Wisdom. He must use this reroll before determining whether the roll succeeds or fails.
\subsubsection{Ur-Priest}
Does not exist.
\subsubsection{Void Disciple}
\cd{Requirements} Spellcraft 8 ranks, Skill Focus (Spellcraft), Spell Penetration, ability to cast 2nd-level spells.
\cd{Spells per Day/Spells Known} Increases magic level and caster level at every level.
\cd{Moment of Clarity} The effect lasts for 1 minute. Only combat and skill feats can be granted with this ability. If ranks are bestowed, the affected character acts as if they had one more skill point in the skill than they actually have. At 11th level, this can be activated as a swift action.
\cd{Reach Through the Void} At 6th level, you can use your void abilities that require touching the target at close (30 ft.) range as a ranged touch attack. If the attack misses, the use of the ability is wasted.
\cd{Altering the Course} The bonus is a circumstance bonus. At 11th level, this can be used an additional time per day.
\cd{Void Release} The effect lasts for 1 minute. Activated as a standard action.
\cd{Void Strike} Can be used once per minute.
\subsubsection{Warpriest}
Does not exist (under construction).
%\cd{Requirements} Base attack bonus \plus5, Combat Casting, ability to cast 1st level divine spells, ability to turn or rebuke undead.
%\cd{Rally} The bonus is a circumstance bonus.

\subsection{Feats}
\begin{itemize*}
\item Arcane Disciple: Does not exist.
\item Augment Healing: The bonus is a competence bonus.
\item Disciple of the Sun: Does not exist.
\item Divine Metamagic: Cannot raise a spell's effective spell slot level above the spell level that you could normally cast.
\item Divine Spell Power: Increases caster level by \plus2 for the next divine spell cast.
\item Domain Focus: Applies to all spells from that domain, regardless of how they are cast.
\item Domain Spontaneity: Does not exist.
\item Elemental Healing: Affects creatures in a Large (50 ft.) burst.
\item Empower Turning: Does not exist.
\item Glorious Weapons: Activated as a swift action. It affects all allies in a Large (50 ft.) burst.
\item Profane Boost: Activated as a swift action. It affects all creatures in a Large (50 ft.) burst.
\item Quicken Turning: Activated as a swift action. Using a quickened turning costs two uses of turn or rebuke undead.
\item Reach Spell: The spell becomes a ray.
\item Sacred Boost: Activated as a swift action. It affects all creatures in a Large (50 ft.) burst.
\item Sacred Healing: Affects creatures in a Large (50 ft.) burst.
\item True Believer: The feat allows you to roll twice and take the better result for a single saving throw instead of giving a bonus.
\end{itemize*}

\section{Complete Mage}
\subsection{Feats}
\begin{itemize*}
\item Acidic Splatter: Close (30 ft.) range.
\item Alacritous Cogitation: Does not exist.
\item Dimensional Jaunt: You must also have line of effect to your destination. Scrying does not count as providing line of sight.
\item Dimensional Reach: You must also have line of effect to the item.
\item Drowning Glance: The creature must successfully save to be immune to further uses of the drowning glance.
\item Elemental Adept: Does not exist.
\item Fey Heritage: Gives a \plus2 competence bonus to Bluff and Hide checks. If you have 10 ranks in one of those skills, the bonus for that skill increases to \plus4. Does not give a bonus against enchantment effects.
\item Fey Legacy: Gives invisibility instead of dimension door, and summon nature's ally IV instead of summon nature's ally V.
\item Fey Power: The bonus is a competence bonus.
\item Fey Presence: Give charm person instead of charm monster.
\item Fiendish Heritage: Gives a \plus2 competence bonus to Bluff and Hide checks. If you have 10 ranks in one of those skills, the bonus for that skill increases to \plus4. This replaces the feat's original effect.
\item Fiendish Legacy: Gives suggestion instead of teleport, and summon monster IV instead of summon monster V.
\item Fiendish Power: The bonus is a competence bonus.
\item Fiendish Presence: Gives darkness instead of suggestion.
\item Hurricane Breath: The bull rush check is made with a bonus equal to your caster level with air spells \add the level of your highest-level air spell available.
\item Insightful Divination: The bonuses are enhancement bonuses.
\item Invisible Needle: Does not exist.
\item Magic Sensitive: Close (30 ft.) range.
\item Melodic Casting: Remove the second paragraph.
\item Minor Shapeshift: Does not exist.
\item Piercing Evocation: The amount of untyped damage is equal to your caster level with the spell.
\item Ranged Recall: Requires Precise Shot instead of Point Blank Shot.
\item Sickening Grasp: This feat give you the ability to make a touch attack as a standard action to gain the effects of the feat. Its benefits do not apply to all melee touch attacks you make.
\item Touch of Distraction: Does not exist.
\item Winter's Blast: The cone extends 20 feet out.
\end{itemize*}

\section{Complete Scoundrel}
\subsection{Feats}
\begin{itemize*}
\item Ascetic Stalker: Does not exist. %Prerequisites: \Ki power, \ki ward.
\item Better Lucky than Good: Can be used as an immediate action.
\item Chord of Distraction: Expends two uses of bardic music when used.
\item Daredevil Athlete: The bonus is a circumstance bonus.
\item Daring Outlaw: Does not exist. %Change if add Swashbuckler back in
\item Daring Warrior: Does not exist. %Change if add Swashbuckler back in
\item Deadly Defense: [Combat] feat.
\item Disemboweling Strike: Does not exist.
\item Fortuitous Strike: Can be used as an immediate action.
\item Good Karma: You do not take extra damage from the attack if it hits you.
\item Martial Stalker: Does not exist. %Prerequisites: \Ki power, weapon discipline.
\item Master Spellthief: Does not exist. %Change if add spellthief back in
\item Merciful Strike: Does not exist.
\item Mind Drain: Does not exist. %Change if add psionics back in
\item Psithief: Does not exist. %Needs psionics
\item Psychic Luck: Does not exist. %Needs psionics
\item Savvy Rogue: Does not exist.
\item Swift Ambusher: Does not exist. %Needs scout
\item Third Time's the Charm: Does not exist.
\item Victor's Luck: Can be used as an immediate action.
\item Warning Shout: The bonus is a circumstance bonus.
\end{itemize*}

\section{Complete Warrior}
\subsection{Prestige Classes}
\subsubsection{Bear Warrior}
The class is now a 5 level prestige class.
\cd{Requirements} Base attack bonus \plus8 instead of \plus7, Improved Grapple and Toughness instead of Power Attack, Survival 6 ranks.
\cd{Rage} Bear warrior levels stack with barbarian levels for determining uses of rage per day.
\cd{Bear Form} The bear warrior replaces all of the normal benefits from raging with the benefits appropriate to the bear form taken. Every bear form has a base land speed of 40 feet.
\cd{Black Bear Form} \plus8 Strength, \plus2 Dexterity, \plus4 Constitution, and \plus2 Will saves. These are circumstance bonuses. He gains a \plus2 racial bonus to natural armor, which does not stack with other racial bonuses to natural armor. He also gains two claw attacks (d6) and a bite attack (d8).
\cd{Brown Bear Form} Gained at 3rd level. \plus12 Strength, \plus2 Dexterity, \plus6 Constitution, and \plus3 Will saves. These are circumstance bonuses. He gains a \plus4 racial bonus to natural armor. He also becomes Large and gains improved grab, two claw attacks (d8), and a bite attack (2d6).
\cd{Dire Bear Form} Gained at 5th level. \plus16 Strength, \plus2 Dexterity, \plus8 Constitution, and \plus4 Will saves. These are circumstance bonuses. He gains a \plus6 racial bonus to natural armor. He also becomes Large and gains improved grab, two claw attacks (d10), and a bite attack (2d8).
\cd{Scent} Gained at 2nd level.
\cd{Speak with Bears} At 2nd level, the bear warrior can speak with any kind of bear that he can turn into. If he can turn into a brown bear, he can also speak with polar bears.
\cd{Command Bears} At 4th level, the bear warrior can command any kind of bear that he can turn into. If he can turn into a brown bear, he can also speak with polar bears.
\cd{Bear's Endurance} At 4th level, the bear warrior gains a \plus2 inherent bonus to Constitution.
\subsubsection{Bladesinger}
\cd{Requirements} Does not require ranks in Balance or Dodge.
\subsubsection{Cavalier}
\cd{Requirements} Base attack bonus \plus5, Ride 8 ranks, Mounted Combat, Ride-by Attack, Spirited Charge
\cd{Courtly Knowledge} Does not exist.
\cd{Mounted Weapon Bonus} The bonuses are competence bonuses.
\cd{Ride Bonus} The bonus is equal to the cavalier's class level.
\cd{Full Mounted Attack} Does not exist.
\cd{Mounted Defense} At 7th level, the cavalier can make two Ride checks per round with the Mounted Combat feat to avoid a hit on his mount.
\subsubsection{Darkwood Stalker}
Does not exist.
\subsubsection{Dervish}
\cd{Requirements} Base attack bonus \plus5, Perform (dance) 4 ranks, Tumble 4 ranks, Combat Expertise, Dodge, Mobility, Spring Attack
\cd{AC Bonus} The bonus is a competence bonus to the dervish's dodge modifier. The bonus increases to \plus2 at 4th level and to \plus3 at 8th level. It does not function if the dervish is using a shield.
\cd{Dervish Dance} While in a dervish dance, as a full-round action, the dervish can move up to her full normal movement speed and take a full attack. She may split her attacks as she chooses between any targets that she moves adjacent to during the round. The dervish is subject to attacks of opportunity while dancing, but may tumble normally as part of her move. A dervish prevented from completing her move may finish her full attack against any targets she can reach without moving. This replaces the normal benefits of a dervish dance.
\par The normal restrictions on entering a dervish dance still apply.
\cd{Slashing Blades} Does not exist.
\cd{Movement Mastery} Gained at 6th level.
\cd{Fast Movement} The bonus is a competence bonus.
\cd{Whirlwind Attack} Gained at 2nd level. If the dervish already has Whirlwind Attack, she may choose any other combat feat for which she qualifies. While in a dervish dance, the dervish may perform a whirlwind attack at any point during her movement, replacing her regular full attack as the feat normally would.
\cd{Spring Attack} Does not exist.
\cd{Flurry of Slashes} At 5th level, while in a dervish dance, the dervish can take an extra attack each round at her full base attack bonus. However, if she does so, she takes a \minus2 penalty on all attacks made during that round. This does not stack with any other ability which grants extra attacks.
\cd{Improved Reaction} Does not exist.
\cd{Reaping Dance} At 8th level, a dervish gains the benefit of the Great Cleave feat while performing a dervish dance, even if she does not meet the prerequisites for the feat.
\subsubsection{Drunken Master}
\cd{Core Features} The drunken master has a HV of 6 and full base attack bonus progression.
\cd{Requirements} Tumble 8 ranks, Great Fortitude, Improved Unarmed Strike
\cd{Drink like a Demon} Each drink represents a mug of ale, a glass of win, or a corresponding amount of stronger alcohol. Each drink lasts for 1 minute. The bonus is a circumstance bonus that stacks with itself.
\cd{Improvised Weapons} The drunken master takes no penalty to attack rolls with improvised weapons. This replaces the normal effects of this ability.
\cd{Improved Feint} Gained at 2nd level. If the drunken master already has Improved Feint, he may choose any other combat feat for which he qualifies.
\cd{Swaying Waist} At 2nd level, a drunken master gains a \plus1 competence bonus to his dodge modifier. This bonus increases by \plus1 at every even level. The drunken master gains this bonus as long as he is wearing no armor or light armor, including a buckler or light shield.
\cd{Improved Improvised Weapons} Gained at 3rd level. The drunken master also gains a \plus1 competence bonus to attack rolls with improvised weapons, and deals damage with improvised weapons as if the weapons were one size category larger.
\cd{Improved Bull Rush} Gained at 4th level. If the drunken master already has Improved Bull Rush, he may choose any other combat feat for which he qualifies.
\cd{AC Bonus} Does not exist.
\cd{Stagger} Gained at 5th level. The drunken master may use the Tumble skill while moving at full speed as part of this charge. This replaces the existing Tumble ability.
\cd{Greater Improvised Weapons} Gained at 6th level. The drunken master gains a \plus2 competence bonus to attack rolls with improvised weapons and deals damage with improvised weapons as if the weapons were two size categories larger.
\cd{Improved Drink like a Demon} At 6th level, the drunken master can consume two drinks with a single move action.
\cd{Improved Grapple} Does not exist.
\cd{For Medicinal Purposes} Gained at 7th level. This ability can be used any number of times per day.
\cd{Corkscrew Rush} Gained at 8th level. The DC is equal to 10 \add 1/2 base attack bonus \add Constitution. Bloodied opponents are stunned, and healthy opponents are staggered.
\cd{Superior Improvised Weapons} The drunken master gains a \plus3 competence bonus to attack rolls with improvised weapons and deals damage with improvised weapons as if the weapons were three size categories larger.
\cd{Breath of Flame} Activated as a standard action. The drunken master can consume up to his Constitution in drinks as part of the action, dealing 3d10 damage per drink consumed. The drunken master can consume a number of drinks per day in this way equal to twice his class level.
\subsubsection{Exotic Weapon Master}
Renamed to Weapon Master.
\cd{Requirements} Base attack bonus \plus6, Weapon Focus, Weapon Specialization, Improved Critical.
\cd{Exotic Weapon Stunt} Renamed to Weapon Stunt. All weapon stunts do not require exotic weapons. Instead, they work with the weapon groups used to gain access to this class. Characters with the improved weapon discipline class feature who chose to focus in all weapon groups may use weapon stunts with any weapon that meets the requirements for the stunt.
\begin{itemize*}
\item Close-Quarters Ranged Combat: This can only be used with one-handed ranged weapons, such as slings or thrown daggers.
\item Double Weapon Defense: Gives Two-Weapon Defense when using double weapons.
\item Exotic Reach: Renamed to uncanny reach. Does not work with spiked chains.
\item Exotic Sunder: Renamed to uncanny sunder.
\item Flurry of Strikes: Does not work with spiked chains.
\item Show Off: The effect lasts for 1 minute.
\item Throw Exotic Weapon: Renamed to uncanny throw.
\item Twin Exotic Weapon Fighting: Renamed to Twin Weapon Fighting. Can be used when wielding the same light weapon in each hand. 
\item Uncanny Blow: Does not exist.
\end{itemize*}
\subsubsection{Eye of Gruumsh}
Does not exist yet.
\subsubsection{Frenzied Berserker}
\cd{Requirements} Base attack bonus \plus9, Cleave, Destructive Rage, Intimidating Rage, Power Attack.
\cd{Frenzy} The bonuses are circumstance bonuses. Attempting to end the frenzy requires a swift action.
\cd{Diehard} If the frenzied berserker already has Diehard, she may choose any other combat feat for which he qualifies.
\cd{Supreme Cleave} Once per round, the frenzied berserker can move five feet between attacks when using the Cleave or Great Cleave feat. This movement provokes attacks of opportunity as normal.
\cd{Improved Power Attack} The frenzied berserker gains an additional \plus1 circumstance bonus to damage for every \minus1 penalty she takes on her melee attack rolls with the Power Attack feat.
\cd{Supreme Power Attack} The frenzied berserker gains an additional \plus2 circumstance bonus to damage for every \minus1 penalty she takes on her melee attack rolls with the Power Attack feat.
\subsubsection{Gnome Giant-Slayer}
Does not exist.
\subsubsection{Halfling Outrider}
\cd{Core Features} The class is a five level class.
\cd{Requirements} Halfling, base attack bonus \plus5, Listen 4 ranks, Ride 8 ranks, Spot 4 ranks, Mounted Combat
\cd{AC Bonus} The bonus is a competence bonus to the halfling outrider's dodge modifier.
\cd{Mount} Halfling outrider class levels stack with either paladin, druid, or ranger levels for determining the characteristics of a paladin's special mount or of an animal companion. The choice of class must be made when the first level of halfling outrider is gained.
\cd{Defensive Riding} The bonuses are circumstance bonuses. The mount gains a bonus to Reflex saves instead of Will saves.
\cd{Stand on Mount} Gained at 3rd level.
\cd{Evasion} Gained at 4th level.
\cd{Quick Turn} Gained at 5th level.
\subsubsection{Hulking Hurler}
Does not exist yet.
%\cd{Requirements} Brutal Throw, Power Attack, and Power Throw instead of the original feats.
\subsubsection{Hunter of the Dead}
Does not exist.
\subsubsection{Invisible Blade}
\cd{Requirements} Bluff 8 ranks, Improved Feint, Weapon Focus (light blade)
\cd{Dagger Sneak Attack} As the rogue ability, but with daggers, kukris, and punching daggers only.
\cd{Unfettered Defense} This ability can also be used in light armor.
\cd{Tricky Feint} At 2nd level, the invisible blade can use a Bluff check instead of a combat maneuver check to feint in combat. The Bluff check includes any bonuses that would apply to feint attempts, such as the \plus4 competence bonus from Improved Feint.
\cd{Bleeding Wound} Gained at 3rd level. This ability is used as an ambush attack.
\cd{Feint Mastery} Applies to all checks made to feint in combat, whether using a Bluff check or a combat maneuver check.
\cd{Uncanny Feint} At 5th level, the invisible blade can feint as a swift action. This replaces the normal effect of this ability.
\subsubsection{Justicar}
\cd{Requirements} Lawful alignment, base attack bonus \plus5, Gather Information 8 ranks, Search 4 ranks, Sense Motive 4 ranks, Survival 8 ranks, Investigator, Track.
\cd{Weakening Strike} As the original crippling strike ability. This can be used only once per round. The justicar must choose which attack is a weakening strike before rolling the attack.
\cd{Crippling Strike} At 6th level, the justicar can deal 2 points of Strength damage when making a weakening strike.
\cd{Improved Nonlethal Strike} At 9th level, the nonlethal damage the justicar deals becomes more punishing. It is treated as lethal damage for the purpose of healing the damage.
\subsubsection{Kensai}
Does not exist yet.
\subsubsection{Knight of the Chalice}
Does not exist.
\subsubsection{Knight Protector}
\cd{Requirements} Lawful neutral or lawful good alignment, base attack bonus \plus5, Diplomacy 4 ranks, proficiency with heavy armor, Combat Reflexes, Iron Will.
\cd{Core Features} The class is a five level class, and has good Fortitude saves.
\cd{Bulwark of Defense} At 1st level, any squares the knight protector threatens are treated as difficult terrain for his foes.
\cd{Defensive Stance} The bonus is a circumstance bonus, and applies to all adjacent allies. The maximum number of points the knight protector can transfer increases by 1 at 3rd and 5th level.
\cd{No Mercy} Gained at 2nd level. A second extra attack is gained at 4th level.
\cd{No Escape} At 3rd level, any creature hit by an attack of opportunity from a knight protector must immediately stop moving. Flying creatures are not affected by this ability.
\cd{Retributive Attack} Gained at 4th level. A knight protector can make three retributive attacks per day.
\cd{No Surrender} At 5th level, if a knight protector would be killed or knocked unconscious by hit point loss, he may make a Will save with a DC equal to the number of negative hit points he has to remain alive and conscious. He does not take damage for acting, but continue to lose hit points for bleeding every turn unless he stabilizes. He must repeat this save every round on his turn to remain alive and conscious. If he fails his save, he falls unconscious, and if his current negative hit points exceeds his Constitution score, he dies as normal.
\subsubsection{Master Thrower}
\cd{Requirements} Base attack bonus \plus5, Precise Shot, Quick Draw, Weapon Focus (thrown).
\cd{Quick Draw} Does not exist.
\cd{Point Blank Shot} At 1st level, the master thrower gains Point Blank Shot as a bonus feat. If he already has Point Blank Shot, he may choose any other combat feat for which he qualifies.
\cd{Thrown Weapon Tricks}
\begin{itemize*}
\item Defensive Throw: When fighting defensively, the master thrower's thrown attacks do not provoke attacks of opportunity.
\item Doubletoss: When throwing two weapons at once with two-weapon fighting, the off-hand weapon gains the master thrower's full Strength bonus to damage.
\item Palm Throw: When throwing two weapons at once with two-weapon fighting, the effect of palm throw only applies to the main hand. If the master thrower has both palm throw and doubletoss, the effect can be used with both hands.
\end{itemize*}
\cd{Critical Focus} The master thrower gains Improved Critical (thrown).

\subsubsection{Master of the Unseen hand}
\cd{Improved Caster Level} Does not exist.
\cd{Spells per Day/Spells Known} Increases magic level and caster level at every level after 1st.

\subsubsection{Mindspy}
\cd{Anticipate} The AC bonus is based on Intelligence bonus, not Charisma bonus, and is not limited by class level.
\cd{Improved Detect Thoughts} At 2nd level, the mindspy gains a competence bonus equal to her class level to the saving throw DC of any \spell{detect thoughts} spells or abilities she uses.
\cd{Multiple Surface Thoughts} A mindspy can simultaneously detect the surface thoughts of a number of creatures equal to her mindspy class level.
\cd{Persistent Mindscan} At 5th level, each opponent targeted by the mindspy's \spell{detect thoughts} must make a new saving throw every round to resist the effect.

\subsubsection{Nature's Warrior}
Does not exist.

\subsubsection{Occult Slayer}
\cd{Requirements} Base attack bonus \plus5, Knowledge (arcana) 2 ranks, Spellcraft 4 ranks, Weapon Focus (any).
\cd{Magical Defense} The bonus is a competence bonus.
\cd{Nondetection Cloak} The caster level equals twice the occult slayer's class level.
\cd{Weapon Bond} The damage increases to 2d6 at 4th level.
\cd{Blank Thoughts} The ability can be suppressed and renewed as a swift action.

\subsubsection{Order of the Bow Initiate}
Does not exist yet.

\subsubsection{Purple Dragon Knight}
Does not exist yet.

\subsubsection{Rage Mage}
\cd{Core Features} Full base attack bonus.
\cd{Rage} Levels in rage mage stack with levels in classes that provide the rage ability for the purpose of determining the temporary hit points gained from raging and for determining the uses of rage per day.
\cd{Spell Rage} A rage mage can cast spells while in a rage, as long as the spell's casting time is no more than 1 full round and the spell's level is not higher than her rage mage class level. She cannot cast spells from the divination, enchantment, or illusion schools while in a rage. While in a spell rage, the rage mage uses her base attack bonus as her caster level.
\par This replaces the normal effects of this ability, as well as the uses per day given. For a rage mage, there is no distinction between rages and spell rages.
\cd{Overcome Spell Failure} At 2nd level, a rage mage ignores arcane spell failure with light armor. At 5th level, she ignores arcane spell failure with medium armor.
\cd{Empowering Fury} At 3rd level, a rage mage can empower one spell while raging without adjusting the spell's level or casting time. She may use this ability once during each rage. The spell's level must be at least two levels lower than her rage mage class level.
\cd{Angry Spell} At 4th level, a rage mage can use her Strength instead of her normal casting ability to determine the saving throw DC of her spells while raging.
\cd{Maximizing Fury} At 6th level, a rage mage can maximize one spell while raging without adjusting the spell's level or casting time. She may use this ability once during each rage. The spell's level must be at least three levels lower than her rage mage class level.
\cd{Furious Casting} At 8th level, a rage mage no longer needs somatic components to cast spells, and no longer suffers any chance of arcane spell failure. Her anger and words alone are sufficient to cast spells.
\cd{Spell Fury} Gained at 9th level, and renamed to quickening fury. A rage mage can quicken one spell while raging without adjusting the spell's level. She may use this ability once during each rage. The spell's level must be at least four levels lower than her rage mage class level.

\subsubsection{War Chanter}
\cd{Requirements} Nonlawful alignment, base attack bonus \plus4, Perform (any) 8 ranks, Endurance, Improved Initiative.
\cd{War Chanter Music} War chanter levels also stack with bard levels for the purpose of determining the bonuses gained by inspire courage. War chanter music abilities require Perform (any) instead of Perform (sing) or Perform (oratory).
\cd{Inspire Toughness} Grants 10 temporary hit points \plus1 per class level and the Endurance feat. At 5th level, the temporary hit points granted double and affected allies gain the Diehard feat. At 9th level, affected creatures gain fast healing 3.
\cd{Battle Leader} At 1st level, the war chanter becomes more effective in combat while leading his allies. While performing bardic music or war chanter music, the war chanter gains a \plus1 competence bonus to his attack rolls and dodge modifier. This bonus increases by \plus1 at 3rd, 6th, and 9th level.
\cd{Singing Shout} At 2nd level, the war chanter with 11 or more ranks in Perform can use her music to grant all allies in a large (50 foot) radius a \plus4 enhancement bonus to Strength. This effect lasts for one minute.
\cd{Bardic Music (Move)} At 3rd level, the war chanter can initiate her bardic music or war chanter music as a move action.
\cd{Inspire Recklessness} Gained at 4th level. The bonus is a circumstance bonus.
\cd{Bardic Music (Swift)} At 6th level, the war chanter can initiate her bardic music or war chanter music as a swift action.
\cd{Inspire Awe} Affected creatures are panicked if bloodied and shaken if healthy.
\cd{Improved Singing Shout} At 8th level, allies affected by Singing Shout also gain a \plus4 enhancement bonus to Dexterity and Constitution.
\cd{Inspire Legion} Does not grant a bonus to damage.

\subsection{Feats}
\subsubsection{General Feats}
\begin{itemize*}
\item Arcane Strike: Activated as a swift action. The bonuses are circumstance bonuses.
\item Arterial Strike: [Ambush] feat. Prerequisite: Sneak attack \plus1d6.
\item Axiomatic Strike: As a swift action, you may expend a use of Stunning Fist to deal an extra 2d6 points of damage against chaotic foes with all of your attacks with monk weapons for one round.
\item Clever Wrestling: Does not require Improved Unarmed Strike. The bonus applies to Escape Artist checks as well.
\item Close Quarters Fighting: [Combat] feat.
\item Defensive Strike: [Combat] feat. If your Dodge target attacks and misses you while you are using the total defense action, you get an attack of opportunity against that opponent. You get no bonus if that opponent does not attack you or if it attacks and hits you.
\item Defensive Throw: [Combat] feat.
\item Extend Rage: Use the core feat instead.
\item Extra Rage/Smiting/Stunning: Use the core feat instead.
\item Eyes in the Back of Your Head: You reduce your overwhelm penalties by 2. You can still be sneak attacked while overwhelmed, even if you take no penalties.
\item Faster Healing: You heal twice as fast as normal.
\item Favored Power Attack: When you use the Power Attack feat against a favored enemy, you deal extra damage equal to the penalty you take on the attack.
\item Fists of Iron: [Combat] feat. As a swift action, you may expend a use of Stunning Fist to deal an extra 1d6 points of damage with all of your unarmed attacks for 1 round. The bonus is a circumstance bonus.
\item Flick of the Wrist: [Combat] feat.
\item Flying Kick: [Combat] feat. Does not require Power Attack. Requires 5 ranks in Jump.
\item Freezing the Lifeblood: [Combat] feat. Slows healthy creatures and paralyzes bloodied creatures. The attack is made as a touch attack.
\item Greater Two-Weapon Defense: Does not exist.
\item Hamstring: Does not exist.
\item Hold the Line: [Combat] feat.
\item Improved Buckler Defense: Does not exist.
\item Improved Combat Expertise: Does not exist.
%\item Improved Favored Enemy: Does not exist.
\item Improved Mounted Archery: [Combat] feat.
\item Improved Rapid Shot: [Combat] feat. Requires Precise Shot instead of Point Blank Shot.
\item Improved Two-Weapon Defense: Does not exist.
\item Improved Weapon Familiarity: Does not exist.
\item Instantaneous Rage: You can enter a rage as an immediate action.
\item Intimidating Rage: You can demoralize an opponent as a swift action.
\item Karmic Strike: [Combat] feat. You can only make the attacks of opportunity against your Dodge target.
\item Pain Touch: [Combat] feat. Sickens healthy creatures and nauseates bloodied creatures, but can affect creatures of any size category.
\item Phalanx Fighting: [Combat] feat. The bonus is a circumstance bonus.
\item Pin Shield: [Combat] feat. When attacking with two weapons, you may give up your off-hand attack. If you do, you momentarily pin your opponent's shield with your off-hand weapon, and your opponent gains no benefit from her shield until the end of your action. This feat can only be used against opponents within one size category of you.
\item Power Critical: Does not exist.
\item Prone Attack: [Combat] feat.
\item Ranged Disarm: [Combat] feat.
\item Ranged Pin: [Combat] feat.
\item Ranged Sunder: [Combat] feat.
\item Rapid Stunning: [Combat] feat.
\item Roundabout Kick: [Combat] feat. Does not require Power Attack.
\item Sharp-Shooting: [Combat] feat. Your targets gain no bonus from cover. This has no effect against foes with no cover or total cover.
\item Shield Charge: [Combat] feat.
\item Shield Slam: [Combat] feat. Can be used as a full attack action or as part of a charge action.
\item Swarmfighting: [Combat] feat. The bonus is a circumstance bonus.
\item Throw Anything: [Combat] feat.
\item Weakening Touch: [Combat] feat. The attack is made as a touch attack.
\end{itemize*}
\subsubsection{Divine Feats}
\begin{itemize*}
\item Divine Cleansing: As a standard action, you can spend one of your turn or rebuke attempts to grant all allies (including yourself) within a 50-foot burst a second saving throw against an ongoing spell, spell-like ability, or supernatural ability affecting the ally that allows a Fortitude save. If the ally is subject to more than one ongoing magic effect, the target chooses which effect he or she makes a saving throw against. If the subject succeeds at the saving throw on the second attempt, the effect ends immediately.
\item Divine Might: This is activated as a swift action.
\item Divine Resistance: Does not require Divine Cleansing. This is activated as an immediate action.
\item Divine Shield: This lasts for one minute.
\item Divine Vigor: The speed bonus is an enhancement bonus. The effect grants two temporary hit points per turning level.
\item Sacred Vengeance: Activated as a swift action.
\end{itemize*}
\subsubsection{Tactical Feats}
All tactical feats are [Combat] feats.
\begin{itemize*}
\item Combat Brute: Momentum Swing: The penalty on the attack roll does not need to be \minus5 or worse. The bonus damage is equal to the penalty taken on the attack.
\item Elusive Target: Diverting Defense: You must be overwhelmed by foes on opposite sides of you, not flanked.
\item Formation Expert: Step into the Breach is used as an immediate action.
\item Giantbane: Duck Underneath: The Tumble check is made as a move action. Climb Aboard: The Climb check is made as a move action against a DC equal to the creature's CMD. You gain a \plus4 circumstance bonus on the check per size category of difference between your size and the creature's size.
\item Raptor School: Falcon's Feathers: To use this maneuver, you must be wearing a cloak. When you feint, you can whip the cloak around you in a distracting fashion. If you successfully feint against an opponent, you gain a \plus2 circumstance bonus to your dodge modifier against the next attack that opponent makes against you in the next round. Hawk's Eye: The bonus is a circumstance bonus.
\item Shock Trooper: Heedless Charge: The penalty on the attack roll does not need to be \minus5 or worse. You can assign up to half of the attack roll penalty from Power Attack from your Armor Class instead.
\end{itemize*}
\subsubsection{Weapon Style Feats}
All weapon style feats are [Combat] feats.
\subsection{Spells}
\section{Draconomicon}
\subsection{Feats}
\subsubsection{Dragon Feats}
Any individual use of a breath weapon can have a number of metabreath feats used on it (including multiple uses of the same metabreath feat) equal to your Constitution.
\begin{itemize*}
\item Adroit Flyby Attack: Applies when flying and taking a standard action, not just when taking the attack action.
\item Large and In Charge: Instead of an opposed Strength check, make a bull rush attack that does not provoke attacks of opportunity.
\item Improved Multiattack requires base attack bonus \plus11.
\end{itemize*}
\subsubsection{General Feats}
\begin{itemize*}
\item Clever Wrestling: [Combat] feat. Does not require Improved Unarmed Strike.
\item Clever Sidestep: [Combat] feat. Does not require Improved Unarmed Strike.
\item Close-Quarters Fighting: [Combat] feat.
\item Dragon Cohort: Does not exist.
\item Dragon Hunter: The bonuses are circumstance bonuses.
\item Dragon Hunter Bravery: The bonus is a circumstance bonus.
\item Dragon Hunter Defense: The bonus is a circumstance bonus.
\item Dragon Steed: Does not exist.
\item Dragon Wild Shape: Does not exist.
\item Dragonbane: Instead of the normal benefits, it causes all attacks you make to deal an extra 1d6 damage against dragons.
\item Dragonfoe: The bonuses are circumstance bonuses. Instead of penalizing saving throws, you get a \plus2 circumstance bonus to your saving throw DC against dragons.
\item Dragonfriend: The bonuses are circumstance bonuses.
\item Dragonsong: The bonus is a competence bonus.
\item Dragonthrall: The bonuses are circumstance bonuses.
\item Overhead Thrust: [Combat] feat.
\end{itemize*}
\section{Dragon Magic}
\subsection{Feats}
\begin{itemize*}
\item Armor of Scales: The bonuses are enhancement bonuses.
\item Blue Dragon Lineage: Each orb deals 2d6 points of electricity damage.
\item Brass Dragon Lineage: Primary effect is exhausted, bloodied effect is sleep. Lasts for a number of rounds equal to the spell's level.
\item Draconic Heritage: Does not grant bonuses to saving throws.
\item Draconic Senses: The bonus is a competence bonus.
\item Dragonfire Inspiration: Instead of its normal benefit, this doubles the enhancement bonus to damage rolls, but it converts the bonus damage to be fire damage. For example, a 6th level bard would grant a \plus2 enhancement bonus to attack and a \plus4 enhancement bonus to damage (which would be dealt as fire damage). 
\item Dragontouched: The bonus to skill checks is a competence bonus. There is no bonus against paralysis and sleep effects.
\item Gold Dragon Lineage: The bonus is a circumstance bonus.
\item Initiate of Bahamut: Your cleric levels and the levels in the class that allow you to smite evil stack for the purpose of determining the extra damage and save DC of your smite evil ability.
\item Initiate of Falazure: You add double the spell's level to the damage healed. The bonus is a competence bonus.
\item Initiate of Hlal: The bonus is a circumstance bonus.
\item Initiate of Io: The bonus is a circumstance bonus.
\item Initiate of Tamara: You add double the spell's level to the damage healed. The bonus is a competence bonus.
\item Initiate of Tiamat: The bonus is a circumstance bonus.
%B \item Silver Dragon Lineage: Primary effect is slowed, bloodied effect is paralyzed. Lasts for a number of rounds equal to the spell's level.
\item Slayer of Dragons: The bonuses are circumstance bonuses.
\item Words of Draconic Power: The bonus to caster level from Potent Words is an enhancement bonus. It gives no bonus to saves.
\end{itemize*}

\section{Dungeon Master's Guide}
\subsection{NPC Classes}
\subsubsection{Skirmisher}
\cd{Base Attack Bonus} Average.
\cd{Saving Throws} Good Reflex.
\cd{Alignment} Any.
\cd{Hit Value} 5.
\cf{Skr}{Class Skills} The skirmisher's class skills (and the key ability for each skill) are Climb (Str), Jump (Str), Swim (Str), Balance (Dex), Escape Artist (Dex), Hide (Dex), Move Silently (Dex), Ride (Dex), Tumble (Dex), Use Rope (Dex), Craft (Int), Knowledge (dungeoneering) (Int), Knowledge (geography) (Int), Search (Int), Heal (Wis), Listen (Wis), Profession (Wis), Spot (Wis), Survival (Wis), Gather Information (Cha).
\cf{Skr}{Skill Points at 1st Level} 8.
\cf{Skr}{Weapon and Armor Proficiency} A skirmisher is proficient with simple weapons, any two weapon groups, light armor, and shields (except tower shields).

\section{Dungeonscape}
\subsection{Feats}
\begin{itemize*}
\item Combat Tinkering: [Skill] feat. The bonuses are circumstance bonuses.
\item Quick Reconnoiter: [Skill] feat. The bonus is a competence bonus.
\item Trap Engineer: [Skill] feat. The bonuses are circumstance bonuses.
\item Trap Sensitivity: Requires danger sense \plus5.
\item Tunnel Fighting: The bonuses are circumstance bonuses.
\end{itemize*}

\section{Ghostwalk}
\subsection{Classes}
\subsubsection{Eidolon}
\cd{Class Skills} All class skills that the ghost has from its other classes.
\subsubsection{Eidoloncer}
\cd{Class Skills} All class skills that the ghost has from its other classes.
\subsection{Feats}
\begin{itemize*}
\item Agony Touch: Does not have a restriction on the number of uses per day. A Fort save negates (DC 10 \add 1/2 HV \add Cha). The damage is dealt in addition to the damage dealt by the Corrupting Touch feat.
\item Circle Magic: Does not exist.
\item Control Visage: The bonus is a competence bonus. 
\item Corpse Malevolence: The body's HP includes its constitution bonus.
\item Corrupting Touch: Deals 1d6 \add Cha damage.
\item Dancing Blade: Does not exist.
\item Diehard: Use the core feat.
\item Divine Energy Focus: Does not exist. Use Improved Turning instead.
\item Education: The bonus is a competence bonus.
\item Empower Turning: Does not exist.
\item Enervating Touch: Does not have an alignment restriction.
\item Ethereal Sidestep: You must have line of sight and line of effect to your destination.
\item Extra Favored Enemy: Does not exist.
\item Extra Spell: As the Complete Arcane version.
\item Extra Wild Shape: Does not exist.
\item Fast Wild Shape: Does not exist.
\item Foe Hunter: The bonus is a circumstance bonus.
\item Forceful Staff Style: Also requires Stunning Fist.
\item Freezing Touch: Deals 1d10 cold damage.
\item Frightful Moan: Shakes healthy creatures, panics bloodied creatures.
\item Full Manifestation: Activated as a swift action.
\item Ghost Glide: You gain a 20 foot glide speed instead of a fly speed.
\item Ghost Smiting: Does not exist.
\item Ghost-Touch Spell: Damaging spells with additional effects not relating to damage do not affect non-ghosts.
\item Gift of Tongues: The bonus is a competence bonus. Gives a bonus to Bluff instead of Innuendo.
\item Green Bond: The bonus is a competence bonus. Gives a bonus to Survival instead of Wilderness Lore.
\item Haunting Appearance: Does not require Improved Control Visage.
\item Heighten Turning: Does not exist.
\item Horrific Appearance: Does not have an alignment restriction. Does not require Improved Control Visage.
\item Improved Control Visage: Mimics the \spell{disguise self} spell.
\item Improved Deflection: The bonus is a competence bonus.
\item Improved Ghost Flight: Also improves flight maneuverability to perfect.
\item Improved Turn Resistance: The bonus is a competence bonus.
\item Incorporeal Form: Activated as a swift action.
\item Incorporeal Target Fighting: [Combat] feat.
\item Militia: Does not exist.
\item Natural Spell: Does not exist.
\item Nauseating Touch: Sickens healthy creatures, nauseates bloodied creatures.
\item Power Lunge: [Combat] feat.
\item Quicken Turning: As the Complete Divine feat.
\item Rend Ghost: Deals 2d8 damage to ghosts.
\item Saddleback: Does not exist.
\item Sculpt Ghost Body: The bonus is an enhancement bonus. Multiple bonuses do not stack with themselves.
\item Sherem-Lar Sorcery: Does not exist.
\item Survivor: Does not exist.
\item Touch Attack Specialization: As the Complete Arcane version.
\item Twin Sword Style: Does not exist. Use Two Weapon Defense instead.
\end{itemize*}

\section{Heroes of Battle}
\subsection{Feats}
\begin{itemize*}
\item Ballista Proficiency: [Combat] feat.
\item Block Arrow: [Combat] feat.
\item Coordinated Shot: [Combat] feat. Does not require Point Blank Shot.
\item Courageous Rally: Does not exist.
\item Extra Followers: Does not exist.
\item Guerilla Scout: The bonus is a competence bonus. It only takes two skill points to master Spot and Listen, rather than three. If you already spent two skill points to specialize in Spot and Listen, you master them instead.
\item Guerilla Warrior: The bonus is a competence bonus. It only takes two skill points to master Hide and Move Silently, rather than three. If you already spent two skill points to specialize in Spot and Listen, you master them instead.
\item Improved Cohort: Does not exist.
\item Inspirational Leadership: Does not exist.
\item Mounted Mobility: [Combat] feat. Renamed Mounted Dodge. Provides the benefits of Dodge to you and your mount.
\item Natural Leader: Does not exist.
\item Plunging Shot: [Combat] feat.
\item Practiced Cohort: Does not exist.
\item Shield Wall: The bonus is a competence bonus to your shield modifier.
\item Veteran Knowledge: Does not exist.
\end{itemize*}

\section{Libris Mortis}
\subsection{Feats}
\begin{itemize*}
\item Bolster Resistance: The bonus is an enhancement bonus.
\item Corrupted Wild Shape: Does not exist.
\item Daunting Presence: [Combat] feat.
\item Deadly Chill: The bonus is an enhancement bonus.
\item Divine Accuracy: Activated as a standard action.
\item Empower Turning: Does not exist.
\item Enduring Life: The bonus is a circumstance bonus.
\item Fell Animate: The zombies last for one minute before falling apart. A fell animated spell uses up a spell slot two levels higher than the spell's actual level.
\item Fell Drain: The negative level disappears after one minute.
\item Ghost Scarred: The bonuses are circumstance bonuses.
\item Hardened Flesh: The bonus is an enhancement bonus.
\item Heighten Turning: Does not exist.
\item Improved Paralysis: The bonus is a competence bonus.
\item Improved Turn Resistance: The bonus is a competence bonus.
\item Lifebond: The bonuses are enhancement bonuses.
\item Necromantic Presence: The bonus is an enhancement bonus.
\item Necropotent: The bonus is a circumstance bonus, and applies to a chosen weapon group.
\item Nimble Bones: The bonuses are enhancement bonuses.
\item Quicken Manifestation: Activated as a swift action.
\item Quicken Turning: Activated as a swift action. Using a quickened turning costs two uses of turn or rebuke undead.
\item Sacred Vengeance: Activated as a swift action.
\item Spell Drain: Does not exist.
\item Tomb-Born Resilience: The bonuses are circumstance bonuses.
\item Undead Leadership: Does not exist.
\item Unquenchable Flame of Life: The bonus is a circumstance bonus.
\end{itemize*}

\section{Lords of Madness}
\subsection{Feats}
\begin{itemize*}
\item Aberration Banemagic: The bonuses are circumstance bonuses.
\item Aberration Blood: The bonuses are competence bonuses.
\item Aberration Wild Shape: Does not exist.
\item Bestial Hide: The bonus is a competence bonus.
\item Darkstalker: Requires Hide 12 ranks, Move Silently 12 ranks, Stealthy as a prerequisite.
\item Deepspawn: The bonus is a competence bonus.
\item Inhuman Reach: Does not exist.
\item Inhuman Vision: The bonus is a competence bonus.
\item Music of the Outer Spheres: The bonuses from Vitalize Aberration are enhancement bonuses. The effects last for as long as the affected creatures hear the bard sing and for 5 rounds thereafter.
\item Ocular Spell: Uses a spell slot three levels higher than the spell's actual level.
\item Parrying Shield: Does not exist.
\item Scavenging Gullet: The bonus is a circumstance bonus.
\item Starspawn: Requires two other aberrant feats as a prerequisite, not just one.
\item Thrall Bred: The bonuses are enhancement bonuses.
\item Warped Mind: Gives a competence bonus equal to the number of aberrant feats you possess. Does not grant power points.
\item Wild Talent: Does not exist.
\end{itemize*}

\section{Magic of Incarnum}
Does not exist.

\section{Miniatures Handbook}
\subsection{Feats}
\begin{itemize*}
\item Battlefield Inspiration: Activated as a swift action.
\item Danger Sense: [Combat] feat. You must decide to reroll before the first round of the encounter starts.
\item Daunting Presence: [Combat] feat.
\item Deft Opportunist: Does not exist.
\item Distracting Attack: [Combat] feat.
\item Double Hit: Does not exist.
\item Energy Affinity: Renamed to Energy Substitution. Use the prerequisites and benefits from the Complete Arcane feat.
\item Extra Domain Spell: Does not exist.
\item Foe Specialist: The bonus is a circumstance bonus.
\item Goad: [Combat] feat.
\item Greater Powerful Charge: [Combat] feat.
\item Hurling Charge: [Combat] feat.
\item Improved Shieldmate: [Combat] feat.
\item Mage Slayer: As the Complete Arcane feat.
\item Martial Throw: [Combat] feat.
\item Mounted Casting: [Skill] feat. The bonus is a circumstance bonus, and applies only to Concentration checks provoked from riding the horse.
\item Powerful Charge: [Combat] feat. Uses the new weapon damage progression for size increases (1d8, 1d10, 2d6, 2d8, etc.)
\item Pushback: [Combat] feat.
\item Reckless Charge: [Combat] feat.
\item Second Wind: Activated as a swift action. Heals half the character's hit points.
\item Sidestep: Does not exist.
\end{itemize*}

\section{Player's Handbook II}
\subsection{Classes}
\subsubsection{Beguiler}
\cd{Weapon Proficiencies} Simple weapons and any two other weapon groups. This replaces their proficiencies with specific weapons.
\cd{Trapfinding} As the rogue ability.
\cd{Cloaked Casting} The bonuses are circumstance bonuses.
\cd{Advanced Learning} Acquired at 3rd level and every odd level thereafter. Allows access to both the generalist and specialist spell lists.
\subsubsection{Dragon Shaman}
\cd{Core Features} Full base attack bonus.
\cd{Draconic Aura} The bonus provided by the draconic aura is an enhancement bonus. It starts at \plus1 and increases by an additional \plus1 at every odd level.
\subsubsection{Knight}
Does not exist.
\subsection{Feats}
\subsubsection{General Feats}
\begin{itemize*}
\item Acrobatic Strike: [Combat] feat. The bonus is a circumstance bonus.
\item Active Shield Defense: [Combat] feat. You take no penalty to attacks of opportunity when taking the total defense action.
\item Adaptable Flanker: Does not exist.
\item Agile Shield Fighter: [Combat] feat. This functions as the Two-Weapon Fighting feat, but only when combining a shield bash and an armed attack.
\item Arcane Flourish: The bonus is an enhancement bonus.
\item Arcane Thesis: The bonus is a competence bonus. The reduction in spell level is applied only once per spell, regardless of the number of metamagic feats used, and is treated as a competence bonus.
\item Armor Specialization: [Combat] feat.
\item Battle Dancer: [Combat] feat. The bonus is a circumstance bonus.
\item Bounding Assault: [Combat] feat.
\item Brutal Strike: [Combat] feat.
\item Combat Acrobat: [Skill] feat.
\item Combat Tactician: [Combat] feat.
\item Cometary Collision: [Combat] feat.
\item Crossbow Sniper: [Combat] feat.
\item Crushing Strike: [Combat] feat. Uses the benefit of the Driving Attack feat. Requires a full attack action instead of a full-round action. You make the bull rush attempt using your normal bull rush modifier, adding the damage dealt on the attack as a circumstance bonus.
\item Deadeye Shot: [Combat] feat. Does not require Point Blank Shot.
\item Defensive Sweep: [Combat] feat.
\item Driving Attack: [Combat] feat. Uses the benefit of the Crushing Strike feat. The bonus is a circumstance bonus.
\item Elven Spell Lore: Requires Spellcraft 12 ranks. The bonus is a competence bonus. You can change the damage type of the spell when casting the spell. You can only choose from the energy damage types (acid, cold, electricity, and fire). When cast, the spell has the subtype appropriate to its new damage type.
\item Fade Into Violence: [Skill] feat. The bonus if your opponent is larger is a circumstance bonus.
\item Fiery Fist: [Combat] feat. Requires base attack bonus \plus4. Adds 1d10 extra damage. Can be chosen for the monk's 6th level bonus feat.
\item Fiery \Ki Defense: [Combat] feat. Requires base attack bonus \plus4. Deals 1d10 damage. Can be selected for the monk's 6th level bonus feat, and does not require any base attack bonus when chosen in this way.
\item Flay: [Combat] feat.
\item Grenadier: [Combat] feat.
\item Hindering Opportunist: [Combat] feat.
\item Intimidating Strike: [Skill] feat.
\item Keen-eared Scout: [Skill] feat. The bonus is a circumstance bonus.
\item \Ki Blast: [Combat] feat. Requires base attack bonus \plus4. 
\item Leap of the Heavens: [Skill] feat. Requires Jump 9 ranks. The bonus is a circumstance bonus.
\item Lunging Strike: [Combat] feat. Requires a standard action.
\item Master Manipulator: [Skill] feat.
\item Melee Evasion: [Combat] feat. Requires base attack bonus \plus4. Can be used when fighting defensively or when using the Combat Expertise feat. Allows you to make a Reflex save with a circumstance bonus equal to the AC bonus from fighting defensively or from the Combat Expertise feat. The result is used as your normal AC and touch AC against that single attack if the result is higher than your AC would be otherwise.
\item Melee Weapon Mastery: [Combat] feat. You apply your Weapon Focus and Weapon Specialization bonuses when using any melee weapons of the chosen damage type. This replaces the normal benefit from the feat.
\item Overwhelming Assault: [Combat] feat. The bonus is a circumstance bonus.
\item Penetrating Shot: [Combat] feat.
\item Ranged Weapon Mastery: [Combat] feat. You apply your Weapon Focus and Weapon Specialization bonuses when using any ranged weapons of the chosen damage type. This replaces the normal benefit from the feat.
\item Rapid Blitz: [Combat] feat.
\item Robilar's Gambit: [Combat] feat. The bonus is a circumstance bonus. The stance lasts until you leave it (a free action).
\item Shield Sling: [Combat] feat.
\item Shield Specialization: [Combat] feat.
\item Shield Ward: Does not exist.
\item Short Haft: Use the core version.
\item Slashing Flurry: [Combat] feat. Ignore the standard attack version. When making a full attack, the penalty is \minus2 instead.
\item Spectral Skirmisher: [Combat] feat.
\item Stalwart Defense: [Combat] feat.
\item Steadfast Determination: Does not exist.
\item Telling Blow: [Combat] feat. If you critical when you could normally deliver precision damage to the target, you deal double your precision damage (regardless of the critical range of the weapon).
\item Trophy Collector: [Skill] feat. The bonuses are circumstance bonuses.
\item Tumbling Feint: [Combat] feat. Does not require Bluff ranks. Grants a \plus5 circumstance bonus to your next feint check against that opponent.
\item Two-Weapon Pounce: [Combat] feat.
\item Two-Weapon Rend: [Combat] feat.
\item Vatic Gaze: Requires a Spellcraft check, not a Sense Motive check. The DC is 10 \add the target's caster level.
\item Versatile Unarmed Strike: [Combat] feat.
\item Vexing Flanker: [Combat] feat. You gain a \plus2 circumstance to melee attack rolls against opponents suffering overwhelm penalties.
\item Wanderer's Diplomacy: [Skill] feat. Social agility allows you to use Bluff like Diplomacy, and does not alter attitudes.
\item Water Splitting Stone: [Combat] feat. Requires base attack bonus \plus4. Allows you to ignore up to 5 points of damage reduction instead of its normal benefit.
\item Weapon Supremacy: Does not exist.
\end{itemize*}
\subsubsection{Ceremony Feats}
Ceremony feats are [Skill] feats. Most bonuses are enhancement bonuses. Vengeful Bonds of Brotherhood gives a circumstance bonus.
\subsubsection{Combat Form Feats}
Combat Form feats are [Combat] feats.
\begin{itemize*}
\item Combat Defense: If you have three or more combat form feats, you gain a \plus1 competence bonus to your dodge modifier against all opponents.
\item Combat Focus: The bonus is a circumstance bonus.
\item Combat Stability: You gain a \plus4 circumstance bonus to your CMD. If you have three or more combat form feats, the bonus granted by this feat increases to \plus8.
\item Combat Strike: The bonus is a circumstance bonus. You lose the benefits of the forms immediately after you finish your turn.
\item Combat Vigor: While you are maintaining your combat focus, you gain three temporary hit points per round. If you have three or more combat form feats, you gain six temporary hit points each round instead.
\end{itemize*}
\subsubsection{Divine Feats}
\begin{itemize*}
\item Divine Armor: Requires turning level 5th instead of caster level 5th. Activated as an immediate action.
\item Divine Fortune: Requires turning level 5th instead of caster level 5th. The bonus is a circumstance bonus.
\item Divine Justice: After a minute passes without being damaged, you can mark a different foe.
\item Profane Aura: Requires turning level 9th instead of caster level 9th. The bonus is an enhancement bonus.
\item Sacred Healing: The bonuses are enhancement bonuses.
\item Sacred Radiance: The bonus is an enhancement bonus.
\end{itemize*}
\subsubsection{Heritage Feats}
\begin{itemize*}
\item Celestial Sorcerer Lance: The line is a Large (50 ft.) line.
\item Infernal Sorcerer Heritage: The bonus is a circumstance bonus.
\item Infernal Sorcerer Howl: The cone is a Medium (20 ft.) cone.
\end{itemize*}
\subsubsection{Tactical Feats}
Tactical feats are [Combat] feats.
\begin{itemize*}
\item Blood-Spiked Charger: Spiked Rebuke functions when using the Combat Expertise feat. The bonus is a circumstance bonus. Spiked Slam does not provoke attacks of opportunity. It deals extra damage equal to twice your Strength bonus.
\item Combat Cloak Expert: Cloak Defense grants a \plus1 competence bonus to your dodge modifier. When fighting defensively, using Combat Expertise, or taking the total defense action, you gain a \plus2 competence bonus instead. The maneuver requires a free hand to perform. Whirling Cloak requires a free hand to perform.
\item Combat Panache: Sneering Glower can be used on the same turn that you hit your opponent.
\item Einhander: Narrow Profile grants a \plus1 competence bonus to your dodge modifier. When fighting defensively, using Combat Expertise, or taking the total defense action, you gain a \plus2 competence bonus instead. Off-Hand Balance grants a \plus2
\end{itemize*}

\section{Races of the Dragon}
\subsection{Racial Substitution Levels}
\subsubsection{Dragonblood Sorcerer}
\begin{itemize*}
\item 7th level: The list of spells known grants Heroism instead of Fly for good sorcerers, and Ray of Enfeeblement instead of Charm Person for evil sorcerers. This replaces the expanded spell knowledge normally gained at this level.
\end{itemize*}

\section{Races of Stone}
\subsection{Prestige Classes}
\subsubsection{Shadowcraft Mage}
\begin{itemize*}
\item Extended Illusion: Gained at 1st level. In addition, any figments which require concentration persist for a number of rounds equal to the shadowcraft mage's class level after the shadowcraft mage ceases to concentrate on them.
\item Cloak of Shadow: Gained at 2nd level. Grants a 20\% miss chance that does not scale with level. A shadowcraft mage can dismiss or resume this effect as a swift action.
\item Shadow Illusion: Can mimic only general sorcerer/wizard spells. Shadow illusions cast from silent image, minor image, and major image are 25\% real. Shadow illusions cast from persistent image and programmed image are 50\% real.
\par The shadow illusion only lasts for the normal duration of the figment spell used. When the duration expires, the shadow illusion is dismissed.
\item Silent Illusion: Gained at 4th level.
\item Powerful Shadow Magic: Increases effect strength by 25\%, to a maximum of 100\%.
\end{itemize*}