%no adequate fear spell
\begin{spellsection}{Symbol of Fear}{7}
\spellinfo{Abjuration/Enchantment (Emotion, Warding) [Fear, Mind-Affecting, Trap]}{Arcane}
\spellatk{Will negates}
\spelleffect This spell functions like \spell{symbol of death}, except that the triggering creature instead suffers the effects of the \spell{terror} spell.
\spellnotes Magic traps such as \spell{symbol of terror} can be detected with the Perception skill and disabled with the Devices skill. The DC in each case is 25 \add spell level, or 32 for \spell{symbol of terror}.
\ritualcost*{7}
\end{spellsection}

\begin{spellsection}{Prestidigitation}{1}
\spellinfo{Universal}{Arcane}
\spelltime{1 minute}
\spelltwocol{\spelltgt{See text}}{\spellrng{Personal/\rngclose}}
\spelldur 1 hour
\spelleffect This ritual enables you to perform simple magical effects for 1 hour on objects or creatures within \rngclose range of you by concentrating (a standard action). The effects are minor and have severe limitations. A \spell{prestidigitation} can slowly lift 1 pound of material. It can color, clean, or soil items in a 1-foot cube each round. It can chill, warm, or flavor 1 pound of nonliving material. It cannot deal damage or affect the concentration of spellcasters. \spell{Prestidigitation} can create small objects, but they look crude and artificial. The materials created by a prestidigitation spell are extremely fragile, and they cannot be used as tools, weapons, or spell components.
\spellnotes This ritual lacks the power to duplicate spell effects. Any actual change to an object (beyond just moving, cleaning, or soiling it) persists only 1 hour. Creatures and attended objects, such as the clothes a creature is wearing, cannot be affected unless the creature is willing. Spell resistance does not apply to you, but it does apply to any objects or creatures you try to affect.
\ritualcost*{1}
\end{spellsection}


\begin{spellsection}{Miracle}{9}
    \begin{spellheader}
    \end{spellheader}
    \begin{spellcontent}
        \spellinfo{Evocation (Channeling)}{Divine}
        \begin{spelltargetinginfo}
            \spelltwocol{\spelltgteffarea{See text}}{\spellrng{See text}}
        \end{spelltargetinginfo}
        \begin{spelleffects}
            \spellatk{See text}
            \spelleffect You don't so much cast a miracle as request one. You state what you would like to have happen and request that your deity (or the power you pray to for spells) intercede.
            \par A miracle can do any of the following things.
            \begin{itemize}
                \item Duplicate any cleric spell of 8th level or lower (including spells to which you have access because of your domains). 
                \item Duplicate any other spell of 7th level or lower.
                \item Undo the harmful effects of certain spells, such as feeblemind or insanity.
                \item Have any effect whose power level is in line with the above effects.
            \end{itemize}
            \par Alternatively, a cleric can make a very powerful request. Examples of especially powerful miracles of this sort could include the following.
            \begin{itemize}
                \item Swinging the tide of a battle in your favor by raising fallen allies to continue fighting.
                \item Moving you and your allies, with all your and their gear, from one plane to another through planar barriers to a specific locale with no chance of error.
                \item Protecting a city from an earthquake, volcanic eruption, flood, or other major natural disaster.
            \end{itemize}
            \par In any event, a request that is out of line with the deity's (or alignment's) nature is refused.
            \spelldur See text
        \end{spelleffects}
    \end{spellcontent}
    \begin{spellfooter}
        \spellnotes If you request a miracle, your deity (or the power you pray to) will expect something of you in return. You must cast \spell{commune}commune to learn what this is within 24 hours, or you will lose the ability to cast any cleric spells other than \spell{commune}. For more moderate miracles, you may be required to offer 25,000gp worth of incense and gems. For especially powerful miracles, or multiple moderate miracles, you may geased with a task to complete.
        \par When a miracle spell duplicates a spell with a material component that costs more than 5,000 gp, you must provide that component.
    \end{spellfooter}
\end{spellsection}
