
\feat{Elemental Body}{Bloodline, Elemental}
\featpre Any three elemental bloodline feats.
\featben You have a 50\% chance to ignore critical hits on you, treating them as regular hits instead.

\feat{Elemental Force}{Bloodline, Elemental}
\featpre Elemental Heritage.
\featben Once per day per elemental bloodline feat you possess, you may unleash the power of your element on your foe as an attack action.
Air allows you to make a shove attack with a \plus4 bonus, and you use your Constitution in place of your Strength to attack.
Earth allows you to make a trip attack with a \plus4 bonus, and you use your Constitution in place of your Strength to attack.
Fire allows you to make a touch attack that causes your foe to be \ignited for 5 rounds if you hit.
Water allows you to make a touch attack that dehydrates your foe for 5 rounds if you hit, making it \vulnerable.

\feat{Elemental Heritage}{Bloodline, Elemental}
\featben You have the essence of an elemental in your body.
When you take this feat, choose a type of elemental to be your elemental ancestor: air, earth, fire, or water.
Air and fire elemental heritage grants a \plus2 bonus to your Reflex defense, while earth and water elemental heritage grant a \plus2 bonus to your Fortitude defense.

\feat{Elemental Mastery}{Bloodline, Elemental}
\featpre Elemental Heritage.
\featben In circumstances that depend on your elemental ancestor, you gain a \plus1 bonus to physical attacks.
Air grants a bonus when you are airborne or fighting airborne creatures.
Earth grants a bonus when both you and your foe are standing on unworked earth or stone.
Fire grants a bonus when either you or your foe is ignited, or when you are making attacks that deal fire damage.
Water grants a bonus when both you and your foe are touching water.
If you have four or more elemental bloodline feats, this bonus increases to \plus2.

\feat{Elemental Movement}{Bloodline, Elemental}
\featpre Elemental Heritage.
\featben You gain a movement ability based on you choice of elemental ancestor.
Air halves the damage you take from falling and improves the maneuverability of any flight abilities you possess by one category.
Earth gives you a \plus2 bonus to maneuver defense against attacks that would force you to move, such as shove and trip attacks.
Fire increases your land speed by 5 feet.
Water gives you a swim speed equal to your base land speed.

\feat{Focused Mind}{Racial}
\featpre Elf.
\featben You gain a \plus2 bonus to Mental defense.
In addition, you can use your Intelligence instead of your Constitution on Concentration checks, such as when you cast spells.

\feat{Alacritous Surge}{Surge}
\featpres 16th level, Dex 7.
\featben You may gain the benefits of the \spell{haste}
spell for up to 5 rounds each day.
Activating this ability is a swift action, and it lasts until you will it to stop (a free action) or until the duration expires.

\feat{Deathless Surge}{Surge}
\featpres 12th level, Wil 5.
\featben You may gain the benefits of the \spell{death ward}
spell for up to 5 rounds each day.
Activating this ability is a swift action, and it lasts until you will it to stop (a free action) or until the duration expires.

\feat{Freeing Surge}{Surge}
\featpres 16th level, Dexterity 7.
\featben You may gain the benefits of the \spell{freedom}
spell for up to 5 rounds each day.
Activating this ability is a swift action, and it lasts until you will it to stop (a free action) or until the duration expires.

\feat{Healing Surge}{Surge}
\featpres 8th level, Constitution 3.
\featben Three times per day, you may heal 4d6 points of damage as a swift action.
This healing increases by 1d6 at 12th level and every 4 levels thereafter.
After using this ability, you cannot use it again for 5 rounds.

\feat{Prescient Surge}{Awareness, Surge}
\featpres 12th level, Wisdom 5.
\featben Three times per day, you may roll twice on any single physical attack or opposed check.
You must decide to use this ability before making the roll.
Activating this ability is an immediate action.

\feat{Shattering Surge}{Power, Surge}
\featpres 8th level, Strength 3.
\featben Three times per day, when you successfully attempt to disarm an object, you may cause it to be affected as if by the targeted version of the \spell{shatter} spell with a spellpower equal to your level.
Activating this ability is an immediate action.

\feat{Surge of Speed}{Surge}
\featpres 8th level, Dexterity 3.
\featben You may gain the benefits of the \spell{expeditious retreat}
spell for up to 5 rounds each day.
Activating this ability is a swift action, and it lasts until you will it to stop (a free action) or until the duration expires.

