\section{Rituals}

\small

\spellsection{Alarm}{1}
\spellinfo{Abjuration (Warding) [Trap]}{Arcane, Divine}
\spelltime{1 minute}
\spelltwocol{\spellzone{\arealarge radius}}{\spellrng{\rngclose}}
\spelldur{\durext \dismissable}
\spellspecial When you perform this ritual, you choose whether to create an audible or mental alarm. You may also specify a password.
\spellline
\begin{spelltrigger}{A creature (Tiny or larger) enters the area without speaking the password}
    \spelleffect[Audible Alarm] The sound of a hand bell rings in the area. It is typically clearly identifiable up to 100 feet away, and audible up to 500 feet away.
    \spelleffect[Mental Alarm] You receive a single mental ``ping''. This awakens you from normal sleep but does not otherwise disturb concentration.
\end{spelltrigger}
\spellnotes A \spell{silence} spell or similar effect can prevent an audible alarm from being heard. This ritual can be made permanent with a \spell{permanency} ritual.
\ritualcost*{1}

\spellsection{Alter Magic Aura}{2}
\spellinfo{Illusion (Glamer) [Magic]}{Arcane}
\spelltime{1 minute}
\spellrng{Touch}
\spelldur{One year \dismissable}
\begin{spelltarget}{One magical object (Large or smaller)}[Magic vs. Will]
    \spellsuccess This ritual alters one of the target's magic auras (see \pcref{Spellcraft}). You can decrease the caster level of the aura by up to half your caster level. Alternately, you can increase the caster level of the aura, up to a maximum of your caster level.

    In addition, you can change the school, subschool, and descriptors of the aura.
\end{spelltarget}
\spellnotes If the target is examined with \spell{identify} or a similar effect, the true aura is revealed.

\par This ritual has no effects on artifacts.
\ritualcost*{2}

\spellsection{Alter Self}{2}
\spellinfo{Transmutation (Polymorph)}{Arcane}
\spelltime{1 minute}
\spelldur{\durlong}
\spellsr{Yes}
\begin{spelltarget}{You}
    \spelleffect When you perform this ritual, you make a Disguise check to disguise yourself. You gain a \plus10 enhancement bonus on the check, and you take no penalties for emulating a different gender or race.
\end{spelltarget}
\ritualcost{2}

\spellsection{Analyze Dweomer}{5}
\spelldesc{You infallibly discern the magical properties of a magic item.}
\spellinfo{Divination (Knowledge)}{Arcane}
\spelltime{1 minute}
\spellrng{Touch}
\begin{spelltarget}{One object}
    \spelleffect You learn the target's magic properties, as \spell{identify}. This ritual reveals the exact properties of cursed items and artifacts.
\end{spelltarget}
\ritualcost{5}

\spellsection{Animate Dead}{3}
\spelldesc{You bind a fragment of a dead creature's soul to its corpse, reanimating it as an undead skeleton or zombie.}
\spellinfo{Necromancy (Soul, Vitalism) [Evil, Negative]}{Arcane, Divine}
\spelltime{1 minute}
\spellrng{Touch}
\begin{spelltarget}{One or more corpses}
    \spellspecial The combined levels of all targets cannot exceed your caster level.
    \spelleffect The target becomes an undead creature that obeys your spoken commands. You choose whether to create a skeleton or a zombie. Creating a zombie require a mostly intact corpse, including most of the flesh. Creating a skeleton only requires a mostly intact skeleton. If a skeleton is made from an intact corpse, the flesh quickly falls off the animated bones.
\end{spelltarget}
\spellnotes The undead you create remain under your control indefinitely. No matter how many times you use this ritual, however, you can control only 4 levels worth of undead creatures per caster level. If you exceed this number, all the newly created creatures fall under your control, and any excess undead from previous castings become uncontrolled. You choose which creatures are released.

Once destroyed, an undead creature never be animated again, even if the corpse is intact.
\ritualcost{3}[in black onyx gems.]

\spellsection{Animal Messenger}{2}
\spelldesc{You compel a Tiny animal to go to a spot you designate.}
\spellinfo{Enchantment (Compulsion) [Mind-Affecting]}{Arcane, Nature}
\spelltime{1 minute; see text}
\spellrng{\rnglong}
\spelldur{One week}
\spellsr{Yes}
\begin{spelltarget}{One Tiny animal}
    \spelleffect As soon as you begin performing this ritual, the target animal approaches you and awaits your bidding. You can mentally impress on the animal a certain place well known to you or an obvious landmark. The directions must be simple, because the animal depends on your knowledge and can't find a destination on its own. During the casting of the ritual, you can attach some small item or note to the messenger. The animal then goes to the designated location and waits there, straying only to gather food and water as necessary, until the duration of the ritual expires. During this period of waiting, the messenger allows others to approach it and remove any scroll or token it carries.
\end{spelltarget}
\spellnotes The most common use for this ritual is to get an animal to carry a message to your allies. The animal cannot be one tamed or trained by someone else.
\spellfocus{Food desirable to the animal}
\ritualcost{2}

\spellsection{Antipathy}{9}
\spelldesc{You fill nearby creatures of a particular kind with an overpowering revulsion, compelling them to leave the area.}
\spellinfo{Enchantment (Emotion) [Mind-Affecting]}{Arcane, Nature}
\spelltime{1 hour}
\spelltwocol{\spellzone{1 mile radius}}{\spellrng{\rngclose}}
\spelldur{One week \dismissable}
\spellsr{Yes}
\spellspecial When you perform this ritual, choose an alignment (chaotic, good, evil, or lawful) or creature type.
\begin{spelltargets}{All creatures in the area of the chosen alignment or type}l[Magic vs. Will]
    \spelleffect The target takes a \minus4 penalty to attacks, saves, and checks as long as it remains in the area.
    \spellsuccess The target leaves the area as soon as physically possible, using any means necessary. It will not willingly return to the area as long as the spell is in effect.
    \spellfailure The target feels a desire to leave the area, though it is not compelled to act on that desire immediately.
\end{spelltargets}
\ritualcost{9}

\spellsection{Arcane Eye}{4}
\spellinfo{Divination (Scrying)}{Arcane}
\spelltime{1 minute}
\spellrng{\rngmed}
\spelldur{\durlong \dismissable}
\spellline
\spelleffect This ritual creates a magic sensor. As a standard action, you can concentrate to see through the sensor as if you were in its location. While concentrating, you can move the sensor up to 30 feet per round in any direction.

You can freely control and see through the sensor even if it moves beyond the spell's range, or out of your line of effect.
\spellnotes \sensorspellnotes
\ritualcost*{4}

\spellsection{Arcane Lock}{2}
\spellinfo{Abjuration/Transmutation (Alteration, Warding)}{Arcane}
\spelltime{1 minute}
\spellrng{Touch}
\spelldur{Permanent}
\begin{spelltarget}{One closable object, such as a door or box (Large or smaller)}
    \spelleffect The target object is magically locked. It can be unlocked with a Devices check against a DC equal to 30 \add your caster level. The DC to break it open forcibly increases by 10.

    You can freely pass your own \ritual{arcane lock}, as if the object were not locked.
\end{spelltarget}
\spellnotes A \ritual{knock} spell suppresses the effect for 10 minutes, but does not dispel the \spell{arcane lock}.
\ritualcost*{2}

\spellsection{Arcane Mark}{1}
\spelldesc{You inscribe your personal sigil on a creature or object.}
\spellinfo{Universal}{Arcane}
\spelltime{1 minute}
\spellrng{\rngtouch}
\spelldur{Permanent}
\spellspecial When you learn this ritual, choose a personal rune or mark. It can consist of no more than six characters, and must fit within one square foot.
\begin{spelltarget}{One creature or object}
    \spelleffect Your personal rune or mark is written on the target. The target is not harmed by the writing. You can choose whether the writing will be visible or invisible.
\end{spelltarget}
\spellnotes If an \spell{arcane mark} is placed on a creature, normal wear gradually causes the effect to fade in about a month. You must place your mark on an object before \spell{instant summons} can be cast on it.
\ritualcost*{1}

\spellsection{Atonement}{5}
\spellinfo{Evocation (Channeling)}{Divine, Nature}
\spelltime{1 hour}
\spellrng{Touch}
\spellsr{Yes}
\begin{spelltarget}{One creature}
    \spelleffect If the target has been punished for a significant transgression against its alignment, deity, or organization, this ritual can undo that punishment. The exact effects of this atonement depend on the nature of the punishment the creature is suffering.

    The cost of the ritual depends on the circumstances of the act that provoked the punishment. If the creature's act was unintentional, or under magical influence, no material components are required. If the creature's act was intentional, but the creature seeks atonement intentionally, the normal cost must be paid. If the creature's act was intentional, and the creature does not seek atonement willingly, the material component cost is doubled.
\end{spelltarget}
\ritualcost{5}

\spellsection{Augury}{2}
\spellinfo{Divination (Knowledge)}{Divine}
\spelltime{1 minute}
\spellspecial When you perform this ritual, you state an action that a creature (or group of creatures) could take.
\spelleffect You learn whether a particular action is likely to bring good or bad results for you in the immediate future. The ritual provides one of four results:
\begin{itemize}
    \item Weal (if the action will probably bring good results).
    \item Woe (for bad results).
    \item Weal and woe (for both).
    \item No response (for actions that don't have especially good or bad results).
\end{itemize}

This ritual does not describe the future with certainty. It describes which result is most probable. The more unambiguous the action's effects, the more likely the ritual is to be correct.
\spellnotes This ritual can only foresee events roughly half an hour into the future. Thus, the result might not take into account the long-term consequences of an action.
\ritualcost*{2}

\spellsection{Awaken}{7}
\spellinfo{Transmutation (Augment, Imbuement)}{Nature}
\spelltime{24 hours}
\spellrng{Touch}
\spellsr{Yes}
\begin{spelltarget}{One animal}
    \spelleffect The target becomes sentient. Its Intelligence becomes 1d6-3. Its type changes from animal to magical beast. It gains the ability to speak and understand one language that you know.
\end{spelltarget}
\ritualcost{7}

\spellsection{Binding}{3}
\spelldesc{You create an invisible cage designed to hold a creature inside.}
\spellinfo{Abjuration (Negation)}{Arcane, Divine}
\spelltime{1 hour}
\spelltwocol{\spellzone{\areasmall radius}}{\spellrng{\rngclose}}
\spelldur{24 hours}
\spellsr{Yes}
\spellspecial When you perform this ritual, you may cast \spell{dimensional anchor}.
\spellline
\spelleffect This ritual inscribes a magic circle of ritual components on the ground, denoting the edges of the area. If the circle is broken, the ritual's effects end immediately.
\begin{spelltrigger}{A creature other than you enters the area}
    \begin{spelltarget}{The entering creature}[Magic vs. Will]
        \spellsuccess The target is unable to escape the area physically, alter the circle in any way, or use abilities that have effects which extend outside the area. If it uses any spells or abilities, the effects stop at the edge of the area, as if the edge of the area was an impassable barrier.

        If you cast \spell{dimensional anchor} while performing the ritual, the target is also affected by \spell{dimensional anchor}, preventing it from escaping the area magically.
        \spellfailure The target can leave the area, break the circle, and otherwise act normally.
    \end{spelltarget}
\end{spelltrigger}

\spellnotes You can perform this ritual on an existing \spell{binding} to set its remaining duration to 24 hours, rather than to create a new \spell{binding}. You must make a new attack against all creatures in the area each time you reset the duration in this way.

The magic circle is obvious, but a DC 16 Spellcraft check is required to identify that the circle belongs to a \spell{binding} ritual. Alternately, the magic of the trap can be identified with a DC 28 Perception check.
\ritualcost*{3}

\spellsection{Bless Water}{1}
\spelldesc{You imbue water with holy power.}
\spellinfo{Evocation (Channeling) [Good]}{Divine}
\spelltime{1 minute}
\spellrng{Touch}
\spellsr{Yes}
\begin{spelltarget}{One pint of water}
    \spelleffect The target becomes holy water.
\end{spelltarget}
\spellnotes Large bodies of water can be made holy by performing this ritual multiple times. Holy water can be thrown as a splash weapon, dealing 2d4 points of damage to a struck undead creature or an evil outsider.
\ritualcost*{1}

\spellsection{Break Enchantment}{5}
\spellinfo{Abjuration (Negation) [Magic]}{Arcane, Divine}
\spelltime{1 minute}
\spellrng{\rngclose}
\begin{spelltarget}{One creature or object}[Caster level vs. Special]
    \spelleffect Magical effects on the target are dispelled, as a targeted \spell{dispel magic}. This ritual can also remove effects that are immune to \spell{dispel magic}, such as curses, if they are 6th level or lower.
\end{spelltarget}
\par If the effect comes from some permanent magic item, \ritual{break enchantment} does not remove the curse from the item, but it does frees the victim from the item's effects.
\ritualcost{5}

\spellsection{Clairaudience/Clairvoyance}{3}
\spellinfo{Divination (Scrying)}{Arcane, Divine}
\spelltime{1 minute}
\spellrng{\rngext \rngunrestricted}
\spelldur{\durmed \dismissable}
\spellline
\spelleffect This ritual creates a magic sensor. You don't need line of sight or line of effect to create the sensor, but the destination must be known -- a place familiar to you or an obvious one. As a standard action, you can concentrate to see and hear through the sensor as if you were in its location. While concentrating, you can rotate the sensor to see and hear in any direction.
\spellnotes \sensorspellnotes
\ritualcost*{3}

\spellsection{Clone}{7}
\spellinfo{Conjuration (Creation)}{Arcane}
\spelltime{24 hours}
\spellrng{Touch}
\spelleffect This ritual creates an inert duplicate of a creature. The duplicate body is identical to the creature's original body, but it has no soul, and rots if not preserved. It can be used in place of the creature's original body for the purpose of spells and effects which animate or raise the dead, such as \spell{animate dead} or \spell{raise dead} (see \pcref{Bringing Back the Dead}).
\ritualcost{7}[in ritual components, and a piece of flesh (not hair, nails, scales, or the like) with a volume of at least 1 cubic inch that was taken from the original creature's living body. The piece of flesh need not be fresh, but it must be kept from rotting.]

\spellsection{Commune}{5}
\spellinfo{Divination (Communication) [Planar]}{Divine}
\spelltime{10 minutes}
\spelldur{\durmed or until expended}
\spellline
\spelleffect You contact your deity -- or agents thereof -- and ask questions that can be answered by a simple yes or no. You are allowed one such question per caster level. The answers given are correct within the limits of the entity's knowledge. ``Unclear'' is a legitimate answer, because powerful beings of the Outer Planes are not necessarily omniscient. In cases where a one-word answer would be misleading or contrary to the deity's interests, a short phrase (five words or less) may be given as an answer instead.
\spellnotes A cleric without a deity contacts a philosophically aligned deity.
\ritualcost*{5}

\spellsection{Commune with Nature}{5}
\spelldesc{You become one with nature, attaining knowledge of the surrounding territory.}
\spellinfo{Divination (Knowledge)}{Nature}
\spelltime{10 minutes}
\spellrng{10 miles or 1,000 feet; see text}
\spellline
\spelleffect You instantly gain knowledge of as many as three facts from among the following subjects: the ground or terrain, plants, minerals, bodies of water, people, general animal population, presence of woodland creatures, presence of powerful unnatural creatures, or even the general state of the natural setting.
\par In outdoor settings, the ritual operates in a 10 mile radius. In natural underground settings -- caves, caverns, and the like -- the radius is limited to 1,000 feet.
\spellnotes The ritual does not function where nature has been replaced by construction or settlement, such as in dungeons and towns.
\ritualcost*{5}

\spellsection{Comprehend Languages}{2}
\spelldesc{You can understand any language.}
\spellinfo{Divination (Communication)}{Arcane, Divine, Nature}
\spelltime{1 minute}
\spelldur{\durlong}
\begin{spelltarget}{You}
    \spelleffect You gain the ability to understand any language. This does not grant you any additional ability to speak or write in other languages.
\end{spelltarget}
\spellnotes This ritual can be foiled by certain obscuring magic (such as the \spell{secret page} and \spell{illusory script} rituals). It does not decipher codes or reveal messages concealed in otherwise normal text. You may be unable to understand dead or extremely obscure languages.

This ritual can be made permanent with a \spell{permanency} ritual.
\ritualcost{2}

\spellsection{Consecrate}{2}
\spelldesc{You bless an area with holy power, disrupting undead.}
\spellinfo{Evocation (Channeling) [Good]}{Divine}
\spelltime{1 minute}
\spelltwocol{\spellzone{\areamed radius}}{\spellrng{\rngclose}}
\spelldur{24 hours}
\spellline
\spelleffect Undead cannot be created within or summoned into the area.
\begin{spelltarget}*{All undead creatures in the area}
    \spelleffect The target is \vulnerable.
\end{spelltarget}
\ritualcost{2}

\spellsection{Contact Other Plane}{5}
\spellinfo{Divination (Knowledge) [Planar]}{Arcane}
\spelltime{10 minutes}
\spelldur{\durmed or until expended}
\spelleffect You send your mind to another plane of existence (an Elemental Plane or some plane farther removed) in order to receive advice and information from powers there. You are allowed one such question per caster level. The answers given are correct within the limits of the entity's knowledge. ``Unclear" is a legitimate answer, because even powerful beings of other planes are not necessarily omniscient. In cases where a one-word answer would be misleading or contrary to the entity's interests, a short phrase (five words or less) may be given as an answer instead.
\ritualcost{5}

\spellsection{Contagion}{3}
\spelldesc{You infect your foe with a contagious disease.}
\spellinfo{Necromancy (Flesh)}{Arcane, Divine, Nature}
\spelltime{10 minutes}
\spellrng{10 miles \rngunrestricted}
\spellsr{Yes}
\begin{spelltarget}{One creature}[Magic vs. Fortitude]
    \spellsuccess The target contracts a disease selected from the table below, which strikes immediately (no incubation period). The disease's attack bonus for the initial and subsequent incubation periods is equal to your magic attack bonus.
    \begin{dtable}
        \begin{tabularx}{\columnwidth}{l X}
            \thead{Disease} & \thead{Damage} \\
            Blinding sickness & 1d4 Str\footnotetemp{1} \\
            Cackle fever & 1d6 Wis \\
            Filth fever & 1d3 Dex and 1d3 Con \\
            Mindfire & 1d6 Int \\
            Red ache & 1d6 Str \\
            Shakes & 1d6 Dex \\
            Slimy doom & 1d6 Con
        \end{tabularx}
        1 Each time a victim takes 3 or more Strength damage from blinding sickness, he or she must make another Fortitude save or be permanently blinded.	 
    \end{dtable}

\end{spelltarget}
\ritualcost{3}[in ritual components, and one ounce of fresh flesh or blood that belongs to the target. In general, blood is not considered ``fresh'' after one hour, while flesh takes up between a day and a week to decay, depending on the conditions.]

\spellsection{Continual Flame}{2}
\spellinfo{Illusion (Figment) [Light]}{Arcane, Divine}
\spelltime{1 minute}
\spellrng{Touch}
\spelldur{Permanent}
\begin{spelltarget}{One object}
    \spelleffect The target glows like a torch. The effect looks like a regular flame, but it creates no heat and doesn't use oxygen. A \spell{continual flame} can be covered and hidden, but not smothered or quenched.
\end{spelltarget}
\ritualcost{2}

\spellsection{Control Weather}{7}
\spellinfo{Evocation (Control) [Air]}{Arcane, Divine, Nature}
\spelltime{10 minutes; see text}
\spellzone{2 mile radius cylinder centered on you}
\spelldur{\durext}
\spelleffect This ritual changes the weather in the area. The weather begins to form as you perform the ritual. You can call forth weather appropriate to the climate and season of the area you are in. For example, you can normally create a thunderstorm, but not if the season is winter or you are in a desert.
\par You control the general tendencies of the weather, such as the direction and intensity of the wind. You cannot control specific applications of the weather -- where lightning strikes, for example, or the exact path of a tornado.Contradictory conditions are not possible simultaneously.
\spellnotes \spell{Control weather} can do away with atmospheric phenomena (naturally occurring or otherwise) as well as create them.
\ritualcost*{7}

\spellsection{Create Food and Water}{2}
\spellinfo{Conjuration (Creation)}{Arcane, Divine, Nature}
\spelltime{10 minutes}
\spellrng{\rngclose}
\spellline
\spelleffect This ritual creates food and drink. The food is sufficient to sustain three creatures per caster level for 24 hours. The food that this ritual creates is simple fare of your choice -- highly nourishing, if rather bland. You can create food suitable to sustain any kind of creature, if you know what it eats.

Food created by this ritual decays and becomes inedible within 24 hours. The water created by this ritual is just like clean rain water, and does not go bad as the food does.
\spellnotes The food can be kept fresh for another 24 hours by performing a \ritual{purify food and water} ritual on it.
\ritualcost*{2}

\spellsection{Create Magic Aura}{1}
\spellinfo{Illusion (Glamer) [Magic]}{Arcane}
\spelltime{1 minute}
\spellrng{Touch}
\spelldur{One year \dismissable}
\begin{spelltarget}{One location or nonmagical object (Large or smaller)}
    \spelleffect The target gains a magical aura (see \pcref{Spellcraft}). You can choose the school, subschool, and descriptors of the aura. You can also choose the caster level, up to a maximum of your caster level.
\end{spelltarget}
\spellnotes If the target is examined with \spell{identify} or a similar effect, the false nature of the aura is revealed.
\ritualcost*{1}

\spellsection{Create Object}{3}
\spellinfo{Conjuration (Creation)}{Arcane}
\spelltime{1 minute}
\spellrng{\rngclose}
\spelldur{\durext}
\spelleffect When you perform this ritual, you make a Craft check to craft an object. The object appears out of thin air, without any raw materials. The object must be made of nonliving, vegetable matter, such as wood or cloth. The volume of the object created cannot exceed 1 cubic foot per caster level.
\spellnotes This ritual cannot create poisons, alchemical substances, and other reactive items.
\ritualcost*{3}

\spellsectioncomma{Create Object}{Greater}{6}
\spellinfo{Conjuration (Creation)}{Arcane}
\spellrng{\rngclose}
\spelltime{10 minutes}
\spelldur{See text}
\spelleffect This ritual creates an object, as \spell{create object}, except that it can also create mineral-based objects, such as stone or metal. The duration of the created item varies with its relative hardness and rarity, as indicated on the following table.
\begin{dtable}
    \begin{tabularx}{\columnwidth}{>{\lcol}X l}
        \thead{Hardness and Rarity Examples} & Duration \\
        Vegetable matter & One month \\
        Stone, crystal, base metals & One week \\
        Precious metals & 24 hours \\
        Gems & One hour \\
        Rare metal\fn{1} & 1 minute \\
    \end{tabularx}
    1 Includes adamantine, alchemical silver, and mithral. You can't use this ritual to create a cold iron item.
\end{dtable}
\ritualcost{5}

\spellsection{Create Water}{1}
\spelldesc{You create water to ease the thirst of you and your companions.}
\spellinfo{Conjuration (Creation) [Water]}{Divine}
\spellrng{\rngclose}
\spelltime{1 minute}
\spelleffect This ritual creates wholesome, drinkable water. It creates 2 gallons per caster level. The water can be created at multiple locations within the ritual's range, allowing you to fill multiple small water containers.
\spellnotes Conjuration spells can't create substances or objects within a creature. Water weighs about 8 pounds per gallon.
\ritualcost*{1}

\spellsection{Create Greater Undead}{8}
\spellinfo{Necromancy (Soul, Vitalism) [Evil, Negative]}{Arcane, Divine}
\spelltime{1 hour}
\spellrng{\rngclose}
\begin{spelltarget}{One intact corpse}
    \spelleffect The target becomes an undead creature that obeys your spoken commands. The type of creature you can create depends on your caster level, as shown on \trefnp{Create Greater Undead}.
\end{spelltarget}

\begin{dtable}
    \lcaption{Create Greater Undead}
    \begin{tabularx}{\columnwidth}{*{2}{>{\lcol}X}}
        Caster Level & Undead Created \\
        16th--19th     & Wraith \\
        20th--24th     & Spectre \\
        25th or higher & Devourer \\
    \end{tabularx}
\end{dtable}

\spellnotes As \spell{animate dead}.
\ritualcost*{8}[in black onyx gems.]

\spellsection{Create Undead}{5}
\spellinfo{Necromancy (Soul, Vitalism) [Evil, Negative]}{Arcane, Divine}
\spelltime{1 hour}
\spellrng{\rngclose}
\begin{spelltarget}{One intact corpse}
    \spelleffect The target becomes an undead creature that obeys your spoken commands. The type of creature you can create depends on your caster level, as shown on \trefnp{Create Undead}.
\end{spelltarget}

\begin{dtable}
    \begin{tabularx}{\columnwidth}{*{2}{>{\lcol}X}}
        \thead{Caster Level} & \thead{Undead Created} \\
        10th--12th     & Ghoul \\
        13th--16th     & Ghast \\
        17th--20th     & Mummy \\
        21st or higher & Mohrg \\
    \end{tabularx}
\end{dtable}
\spellnotes As \spell{animate dead}.
\ritualcost*{5}[in black onyx gems.]

\spellsection{Curse Water}{1}
\spellinfo{Evocation (Channeling) [Evil]}{Divine}
\spelltime{1 minute}
\spellrng{Touch}
\spellsr{Yes}
\begin{spelltarget}{One pint of water}
    \spelleffect The target becomes unholy water. 
\end{spelltarget}
\spellnotes Large bodies of water can be made unholy by performing this ritual multiple times. Unholy water can be thrown as a splash weapon, dealing 2d4 points of damage to a struck good outsider.
\ritualcost*{1}

\spellsection{Darkness}{2}
\spelldesc{You create a dark aura around an object of your choosing, keeping light without.}
\spellinfo{Illusion (Glamer) [Darkness]}{Arcane, Divine}
\spelltime{1 minute}
\spellrng{\rngtouch}
\spelltgt{One object}
\spellemanation{\areamed radius from the target}
\spelldur{\durlong \dismissable}
\spellline
\spelleffect Light within or passing through the area is dimmed to be no greater than shadowy illumination.
\spellnotes Any effect which blocks light also blocks this spell's emanation.
\ritualcost*{2}

\spellsection{Demand}{8}
\spellinfo{Divination/Enchantment (Communication, Compulsion) [Mind-Affecting]}{Arcane}
\spellrng{Unlimited \rngunrestricted}
\spellsr{Yes}
\spellspecial You can target any creature with whom you are familiar, regardless of its location.
\begin{spelltarget}{One creature}[Magic vs. Will]
    \spelleffect You send the target a short verbal message, as \spell{sending}.
    \spellsuccess The target is compelled to obey the message, as \spell{suggestion}.
\end{spelltarget}
\spellnotes As \spell{sending}.
\spellnotes The contact itself is not mind-affecting -- only the \spell{suggestion} effect.
\ritualcost*{8}

\spellsection{Desecrate}{2}
\spellinfo{Evocation (Channeling) [Evil]}{Divine}
\spelltime{1 minute}
\spelltwocol{\spellzone{\areamed radius}}{\spellrng{\rngclose}}
\spelldur{24 hours}
\spellline
\spelleffect Undead created within or summoned into the area gain \plus1 hit point per level of the undead creature.
\begin{spelltarget}*{All undead creatures in the area}
    \spelleffect The target gains a \plus2 enhancement bonus to physical attacks, all checks, and special defenses.
\end{spelltarget}
\ritualcost{2}

\spellsection{Detect Scrying}{4}
\spellinfo{Divination (Awareness) [Magic]}{Arcane}
\spelltime{1 minute}
\spellemanation{\arealarge radius centered on you}
\spelldur{\durext}
\spellline
\spelleffect You automatically locate all magical sensors within the area, and immediately become aware of any attempt to observe you by means of a divination (scrying) spell or effect.

\par When you detect a scrying attempt, you make an opposed caster level check against the scrying creature (d20 \add caster level). If you win, you get a visual image of the scrier and an accurate sense of his or her direction and distance from you.
\ritualcost{4}

\spellsection{gDisguise Self}{1}
\spellinfo{Illusion (Glamer) [Unreal]}{Arcane}
\spelltime{1 minute}
\spelldur{\durlong \dismissable}
\begin{spelltarget}{You}
    \spelleffect When you perform this ritual, you make a Disguise check. You gain a \plus10 enhancement bonus on the check, and you can freely alter the appearance of your clothes and equipment, regardless of their original form.
\end{spelltarget}
\spellnotes Creatures can identify the illusion, as \spell{silent image}.
\ritualcost*{1}

\spellsectioncomma{Disguise Self}{Greater}{3}
\spellinfo{Illusion (Glamer) [Unreal]}{Arcane}
\spelltime{10 minutes}
\spelldur{\durext \dismissable}
\begin{spelltarget}{You}
    \spelleffect You alter your appearance, as \spell{disguise self}. As a standard action, you can concentrate to alter your appearance, just as if you had performed the \spell{disguise self} ritual again.
\end{spelltarget}
\spellnotes Creatures can identify the illusion, as \spell{silent image}.
\ritualcost{3}

\spellsection{Discern Location}{8}
\spellinfo{Divination (Knowledge)}{Arcane, Divine, Nature}
\spelltime{10 minutes}
\spellrng{Unlimited \rngunrestricted}
\spellspecial You can target any creature or object you are familiar with, regardless of its location. To find a creature, you must have seen it once or have an item which once belonged to it. To find an object, you must have seen it once.
\begin{spelltarget}{One creature or object}
    \spelleffect You learn the target's location (place, name, business name, building name, or the like), community, county (or similar political division), country, continent, and the plane of existence where the target lies.
\end{spelltarget}
\ritualcost{8}

\spellsection{Dimensional Lock}{5}
\spelldesc{You create a shimmering emerald field that completely blocks extradimensional travel.}
\spellinfo{Abjuration (Negation)}{Arcane, Divine}
\spelltime{10 minutes}
\spelltwocol{\spellzone{\arealarge radius}}{\spellrng{\rngmed}}
\spelldur{Thirty days}
\spellsr{Yes}
\spellline
\spelleffect Extradimensional travel into or out of the spell's area is impossible. All Conjuration (Translocation) and Conjuration (Summoning) effects are prohibited, as well as \spell{astral projection} and similar spell-like abilities.
\spellnotes This ritual does not interfere with the movement of creatures already in ethereal or astral form when the ritual is finished, nor does it block extradimensional perception. Also, this not prevent summoned creatures from disappearing at the end of a summoning spell.
\ritualcost*{5}

\spellsection{Divination}{4}
\spellinfo{Divination (Knowledge)}{Divine}
\spelltime{10 minutes}
\spellline
\spelleffect When you perform this ritual, you ask a question concerning a specific goal, event, or activity that is to occur within one week. You receive a useful piece of advice in reply to your question. The advice can be as simple as a short phrase, or it might take the form of a cryptic rhyme or omen. 

This ritual does not describe the future with certainty. It provides advice which is likely to be correct. The more unambiguous the question, the more likely the ritual is to be correct.
\spellnotes This ritual can only foresee events roughly a week into the future. Thus, the advice might not take into account the long-term consequences of an action.
\ritualcost*{4}

\spellsection{Dream}{5}
\spellinfo{Divination/Illusion (Communication, Phantasm) [Mind-Affecting]}{Arcane, Divine, Nature}
\spelltime{10 minutes}
\spellrng{Unlimited \rngunrestricted}
\spellsr{Yes}
\spellspecial You can target any creature that you can unambiguously identify, regardless of its location.
\begin{spelltarget}{One creature}
    \spelleffect You send the target a dream that it experiences the next time it falls asleep. The dream can be of any length, and the target remembers it perfectly after waking. If the dream lasts longer than 10 minutes, you must extend the duration of the ritual to match the length of the dream. If the target wakes up before the dream message is complete, it will continue dreaming of the remainder of the message the next time it falls asleep.
\end{spelltarget}

\spellnotes This communication is entirely one-way. You learn nothing about the target's dreams, and it is unable to ask questions of you in its dream.

While the target is awake, before the dream is delivered, the magical aura from this effect can be detected and dispelled on the target.
\ritualcost*{5}

\spellsection{Emancipation}{8}
\spellinfo{Abjuration (Negation)}{Arcane, Divine}
\spelltime{1 minute}
\spellrng{\rngmed or see text}
\spellsr{Yes}
\begin{spelltarget}{One creature}
    \spelleffect The target is freed from all spells and effects that restrict its actions, including binding, charms, entangle, daze, domination, grappling, imprisonment, \spell{maze}, nausea, paralysis, petrification, pinning, sleep, slow, stun, \spell{temporal stasis}, and \spell{web}. To free a creature from \spell{imprisonment} or \spell{maze}, you must know its name and background, and you must cast this spell at the spot where it was entombed or banished into the maze.
\end{spelltarget}
\ritualcost{8}

\spellsection{Endure Elements}{1}
\spellinfo{Abjuration (Shielding)}{Arcane, Divine, Nature}
\spelltime{1 minute}
\spellrng{Touch}
\spelldur{\durext}
\spellsr{Yes}
\begin{spelltarget}{One creature or object}
    \spelleffect The target suffers no harm from being in a hot or cold environment. It can exist comfortably in conditions between \minus50 and 140 degrees Fahrenheit. Its equipment, if any, is also protected.
\end{spelltarget}
\spellnotes This ritual does not protect the target from fire or cold damage.
\ritualcost*{1}

\spellsection{Erase}{1}
\spellinfo{Transmutation (Alteration)}{Arcane}
\spelltime{1 minute}
\spellrng{Touch}
\begin{spelltarget}{One scroll, page, parchment, or other text with writing in ink}
    \spelleffect You erase writing on the target, leaving it as if it had never been written on. You can choose to erase all of the writing, or only specific words.

    To erase magical writing, you must succeed on a caster level check against a DC equal to 11 \add the caster level of the magic.
\end{spelltarget}
\spellnotes This ritual can remove \spell{explosive runes}, a \spell{glyph of warding}, a \spell{sepia snake sigil}, or an \spell{arcane mark}, but not \spell{illusory script} or a \spellindirect{symbol of death}{symbol} spell. It is possible to perform this ritual without reading the writing to be erased, though the ritual grants no special protection against activating such traps.
\ritualcost*{1}

\spellsection{Explosive Runes}{3}
\spelldesc{You trace mystic runes that explode when read.}
\spellinfo{Abjuration (Warding) [Force, Traps]}{Arcane}
\spelltime{1 minute}
\spellrng{Touch}
\spellfocus{One object with writing on it (Small or smaller)}
\spelldur{Thirty days or until discharged \dismissable}
\spellsr{Yes}
\spellline
\begin{spelltrigger}{A creature reads the focus object while within 5 feet of it}
    \spellburst{\areasmall radius centered on the focus}
    \begin{spelltargets}*{Everything in the area}l[Magic vs. Reflex]
        \spellsuccess 3d6 force damage \add d6 per four caster levels above 6th.
        \spellfailure As above, but half damage.
    \end{spelltargets}
\end{spelltrigger}
\spellnotes The primary target receives no defense against the explosion when the runes are read.

Magic traps such as \spell{explosive runes} can be detected with the Perception skill and disabled with the Devices skill. The DC is 25 \add spell level, or DC 28 for \spell{explosive runes}.
\ritualcost*{3}

\spellsection{Fabricate}{5}
\spellinfo{Transmutation (Alteration)}{Arcane, Nature}
\spelltime{10 minutes}
\spellrng{\rngclose}
\begin{spelltargets}{One or more nonmagical objects; see text}
    \spelleffect When you perform this ritual, you make a Craft check to transform the targets into a new item (or items) made of the same materials. You gain a \plus10 enhancement bonus on the check, and you require none of the tools or time expenditure that would normally be necessary.

    The total size of all targets combined must be Large size or smaller. 
\end{spelltargets}
\ritualcost{5}[in ritual components, and the original objects to be transformed.]

\spellsection{False Vision}{5}
\spellinfo{Illusion (Glamer)}{Arcane}
\spelltime{10 minutes}
\spelltwocol{\spellzone{\arealarge radius}}{\spellrng{\rngclose}}
\spelldur{\durext \dismissable}
\spellline
\spelleffect Any divination (scrying) spell used to view anything within the area instead views a figment of your design, as \spell{major image}. As a standard action, you can concentrate to alter the image within the area.
\spellnotes Creatures can identify the illusion, as \spell{silent image}.
\ritualcost*{5}

\spellsection{Fertility/Infertility}{3}
\spellinfo{Transmutation (Alteration)}{Divine}
\spelltime{1 hour}
\spellzone{1/2 mile radius centered on you}
\spelldur{One year}
\spellspecial This spell has two versions. Its effects depend on which version is chosen.
\subspell{Fertility}: Normal plants within the area are twice as productive.
\par \subspell{Infertility}: Normal plants within the area are half as productive.
\spellnotes You may designate places within the area that are not affected.
\ritualcost*{3}

\spellsection{Find the Path}{6}
\spellinfo{Divination (Knowledge)}{Arcane, Divine, Nature}
\spelltime{1 minute}
\spellrng{Touch}
\spelldur{\durext or until discharged \dismissable}
\spellsr{Yes}
\begin{spelltarget}{One creature}
    \spelleffect When you perform this ritual, you unambiguously specify a location. The target knows exactly what direction it must travel to reach that destination by the most direct physical route. It is not always led in the exact direction of the destination -- if there is an impassable obstacle between the target and the destination, the ritual will direct the target around the obstacle, rather than through it.

    The guidance provided by this ritual adjusts to match whatever the target's current physical capabilities are, including flight and other unusual movement modes. It does not see into the future, and changing circumstances may cause the most direct path to change over time. The guidance does not consider hostile creatures, traps, and other passable dangers which may impede progress.

    When the destination is reached, the ritual's effect ends.
\end{spelltarget}

\spellnotes If this ritual is performed inside a \spell{maze}, it allows the target to immediately leave the maze.

\ritualcost*{6}

\spellsection{Find Traps}{2}
\spelldesc{You grant your ally an intuitive insight into the workings of traps, allowing her to easily spot danger ahead.}
\spellinfo{Divination (Knowledge)}{Arcana, Divine, Nature}
\spellrng{Touch}
\spelldur{\durmed}
\spellsr{Yes}
\begin{spelltarget}{One creature}
    \spelleffect The target gains a bonus on Perception checks made to find traps equal to one-half your caster level. In addition, as a full-round action, she may move up 10 feet while searching every square within 10 feet of her for traps with the Perception skill (see \pcref{Search}). If she detects a trap partway through her movement, she may immediately stop moving.
\end{spelltarget}
\spellnotes This ritual does not grant any ability to disable any traps found. See \pcref{Devices}.
\ritualcost*{2}

\spellsection{Fire Trap}{3}
\spelldesc{You create a trap that erupts in a fiery explosion when an intruder opens the item that the trap protects.}
\spellinfo{Abjuration/Evocation (Energy, Warding) [Fire]}{Arcane, Nature}
\spelltime{10 minutes}
\spellrng{Touch}
\spellfocus{One openable object (Large or smaller)}
\spelldur{Thirty days or until discharged \dismissable}
\spellsr{Yes}
\spellspecial When you perform this ritual, you can specify a password.
\spellline
\begin{spelltrigger}{A creature opens the focus object without saying the password}
    \spellburst{\areasmall radius centered on the focus}
    \begin{spelltargets}{Everything within the area}l[Magic vs. Reflex]
        \spellsuccess 3d6 fire damage \add d6 per four levels after 6th
        \spellfailure As above, but half damage.
    \end{spelltargets}
\end{spelltrigger}
\spellnotes \firespellnotes

\par Magic traps such as \spell{fire trap} can be detected with the Perception skill and disabled with the Devices skill. The DC is 25 \add spell level, or DC 28 for \spell{fire trap}.
\ritualcost{3}

\spellsection{Floating Disk}{1}
\spellinfo{Evocation [Force]}{Arcane}
\spelltime{1 minute}
\spellrng{\rngmed}
\spelldur{\durext \dismissable}
\spellline
\spelleffect This ritual creates a slightly concave, circular plane of force that can carry loads for you. The disk floats 3 feet above the ground at all times and remains level. It is 4 feet in diameter and 1 inch deep at its center. It can hold 100 pounds of weight per caster level.

As a swift action, you can command the disk to move up to 50 feet. If you move farther than 30 feet from the disk, it automatically follows you as if you had commanded it. If it remains farther than 30 feet from you for two consecutive rounds, it the disk disappears, dropping its contents.
\spellnotes The disk can hold up to two gallons of liquid.
\ritualcost*{1}

\spellsection{Forbiddance}{8}
\spellinfo{Abjuration/Evocation (Power, Warding)}{Divine}
\spelltime{1 hour}
\spelltwocol{\spellzone{Up to ten 50 ft. cubes}}{\spellrng{\rngext}}
\spelldur{One year}
\spellsr{Yes}
\spellspecial When you perform this ritual, choose an alignment (chaotic, good, evil, or lawful). You may also specify a password.
\spellline
\spelleffect Extraplanar travel within the area is forbidden, as \spell{dimensional lock}.
\begin{spelltrigger}{A creature of the chosen alignment enters the area without saying the password}
    \begin{spelltarget}*{The entering creature}[Magic vs. Will]
        \spellsuccess 8d6 divine damage \add d6 per caster level above 16th.
    \end{spelltarget}
\end{spelltrigger}
\spellnotes A successful dispel attempt only affects one 50-foot cube.
\par You can't have multiple overlapping \spell{forbiddance} rituals. In such a case, the more recent effect stops at the boundary of the older effect.
\ritualcost{8}

\spellsection{Gate}{9}
\spellinfo{Conjuration (Creation, Translocation) [Planar]}{Arcane, Divine}
\spelltime{1 minute}
\spellrng{\rngmed}
\spelldur{Concentration (up to 5 rounds); see text}
\spellline
\spelleffect This ritual creates an interdimensional connection between your plane of existence and a plane you specify, allowing travel between those two planes in either direction.
\par The gate itself is a circular disk between 5 and 20 feet in diameter, oriented in the direction you desire when it comes into existence (typically vertical and facing you). It is a two-dimensional window looking into the plane you specified when casting the spell, and anyone or anything that moves through is shunted instantly to the other side.
\par The \spell{gate} has a front and a back. Creatures moving through the gate from the front are transported to the other plane; creatures moving through it from the back are not.
\par A \spell{gate} spell functions much like a \spell{plane shift} spell, except that the gate opens precisely at the point you desire (a creation effect). 
\spellnotes A \spell{gate} cannot be opened to another point on the same plane. This ritual works only for interplanar travel.
\ritualcost*{9}

\spellsection{Geas/Quest}{6}
\spellinfo{Enchantment (Compulsion) [Language-Dependent, Mind-Affecting, Sound-Dependent]}{Arcane,  Divine}
\spelltime{10 minutes}
\spellrng{\rngmed}
\spelldur{Thirty days, one week, or until discharged \dismissable}
\spellsr{Yes}
\begin{spelltarget}{One creature}
    \spelleffect When you perform this ritual, you specify a task the target must perform, or an activity the target must refrain from. The target is compelled to obey your command.

    If the target is charged to perform a task, it is still able to eat, sleep, and otherwise function normally without performing the task at all times. In general, it should spend at least 12 hours a day performing the task. However, it cannot use its free time to work against the completion of the task.

    \par If the target is prevented from obeying the command for 24 hours, it takes a \minus2 penalty to all attributes. Each day, another \minus2 penalty accumulates, up to a total of \minus8. No attribute can be reduced to less than \minus9 by this effect. All attribute penalties are removed 24 hours after the subject resumes obeying the command.
\end{spelltarget}
\spellnotes If the target is unable to understand the command, the ritual has no effect. The target is able to follow the literal meaning of the command rather than its intended meaning, potentially allowing it to subvert poorly worded instructions.

\par This ritual cannot be dispelled with \spell{dispel magic}. A \spell{remove curse} spell only ends the effect if its caster level is higher than this ritual's caster level.

\par Sorcerers and wizards usually refer to this spell as geas, while clerics call the same spell quest.
\ritualcost*{6}

\spellsection{Gentle Repose}{2}
\spelldesc{You preserve the remains of a dead creature so that they do not decay.}
\spellinfo{Transmutation (Temporal)}{Arcane, Divine, Nature}
\spelltime{10 minutes}
\spellrng{Touch}
\spelldur{Thirty days \dismissable}
\spellsr{Yes}
\begin{spelltarget}{Nonmagical object touched}
    \spelleffect Time does not pass for the target, preventing it from decaying or spoiling. This can extend the time a poison or similar item lasts before becoming inert. If used on a corpse, this effectively extends the time limit on raising that creature from the dead (see \spell{raise dead}) and similar effects that require a fresh body. Additionally, this can make transporting a fallen comrade more pleasant.
\end{spelltarget}
\ritualcost{2}

\spellsection{Glyph of Warding}{3}
\spelldesc{You weave a tracery of faintly glowing lines in the air, forming a warding sigil. When the spell is completed, the glyph and tracery become nearly invisible.}
\spellinfo{Abjuration (Warding) [Trap]}{Divine}
\spelltime{10 minutes}
\spellrng{Touch}
\spellfocus{One location}
\spellzone{\areasmall radius limit centered on the focus}
\spelldur{Thirty days or until discharged}
\spellsr{Yes}
\spellspecial When you perform this ritual, you can specify a password. You must also choose an energy type (acid, cold, electricity, or fire).
\spellline
\spelleffect This ritual creates an invisible, intangible glyph floating in the air at the focus location. A creature that can see invisible objects can perceive the glyph, which is takes up approximately one square foot. If the creature can read magic, such as with the \ritual{read magic} ritual, it can identify the energy type of the glyph.
\begin{spelltrigger}{A creature enters the area without speaking the password}
    \spellburst{\areasmall radius centered on the focus}
    \begin{spelltargets}*{Everything in the area}[Magic vs. Reflex]
        \spellsuccess 3d6 damage \add d6 per four caster levels after 6th. The damage is of the energy type chosen when the ritual was performed.
        \spellfailure As above, but half damage.
    \end{spelltargets}
\end{spelltrigger}
\spellnotes Magic traps such as \spell{glyph of warding} can be detected with the Perception skill and disabled with the Devices skill. The DC is 25 \add spell level, or DC 28 for \spell{glyph of warding}.
\ritualcost*{3}

\spellsection{Glyph of Warding, Greater}{7}
\spellinfo{Abjuration [Barrier]}{Divine}
\spelltime{10 minutes}
\spellrng{Touch}
\spellfocus{One location}
\spellzone{\arealarge radius limit centered on the focus}
\spellspecial When you perform this ritual, you can specify a password. You must also choose an energy type (acid, cold, electricity, or fire).
\spellline
\spelleffect This spell creates an invisible glyph, as \spell{glyph of warding}.
\begin{spelltrigger}{A creature enters the area without speaking the password}
    \spellburst{\arealarge radius centered on the focus}
    \begin{spelltargets}*{Everything in the area}[Magic vs. Reflex]
        \spellsuccess 6d6 damage \add d6 per four caster levels after 12th. The damage is of the energy type chosen when the ritual was performed.
        \spellfailure As above, but half damage.
    \end{spelltargets}
\end{spelltrigger}
\spellnotes Magic traps such as \spell{greater glyph of warding} can be detected with the Perception skill and disabled with the Devices skill. The DC is 25 \add spell level, or DC 32 for \spell{greater glyph of warding}.
\ritualcost*{7}

\spellsection{Hallow}{7}
\spelldesc{You make an area holy.}
\spellinfo{Evocation (Power) [Good]}{Divine}
\spelltime{24 hours}
\spelltwocol{\spellzone{\arealarge radius}}{\spellrng{\rngclose}}
\spelldur{One year}
\spellline
\spelleffect The area becomes holy. This has several effects.
\begin{itemize}
    \item A \spellindirect{magic circle against alignment}{magic circle against evil} effect fills the area.
    \item Any dead body interred in the area cannot be turned into an undead creature.
    \item While performing the ritual, you can cast another spell or perform another ritual you know. The spell or ritual functions in the entire area as long as this ritual lasts, regardless of its normal area or duration. You may designate whether the effect applies to all creatures, or to all creatures that have a specific faith or alignment. The spells and rituals which can be fixed to a \spell{hallow} ritual are as follows: \spell{aid}, \spell{bane}, \spell{bless}, \spell{comprehend languages}, \spell{darkness}, \spell{daylight}, \spell{death ward}, \spell{dimensional anchor}, \spell{endure elements}, \spell{freedom}, \spell{protection from energy}, \spell{resist energy}, \spell{silence}, \spell{tongues}, and \spell{zone of truth}.
\end{itemize}
\spellnotes An area can receive only one \spell{hallow} spell (and its associated spell effect) at a time. If an area is \spellindirect{unhallow}{unhallowed}, it cannot be hallowed.
\ritualcost{7}

\spellsection{Hallucinatory Terrain}{4}
\spellinfo{Illusion (Glamer)}{Arcane}
\spelltime{10 minutes}
\spelltwocol{\spellzone{Ten 30 ft. cubes}}{\spellrng{\rnglong}}
\spelldur{24 hours \dismissable}
\spelleffect You make natural terrain in the area look, sound, and smell like some other sort of natural terrain. Structures, equipment, and creatures within the area are not hidden or changed in appearance.
\spellnotes Creatures can identify the illusion, as \spell{silent image}.
\ritualcost*{4}

\spellsection{Heroes' Feast}{6}
\spelldesc{You bring forth a great feast, including a magnificent table, chairs, service, and food and drink, re-enacting the celebrations of ancient heroes. After your allies consume the ambrosial food and nectar-like bevarage, they are restored in body and mind.}
\spellinfo{Conjuration/Enchantment (Creation, Emotion)}{Arcane, Divine, Nature}
\spelltwocol{\spelltime{10 minutes}}{\spellrng{\rngclose}}
\spelldur{1 hour plus \durext; see text}
\spellline
\spelleffect This ritual creates a feast that takes 1 hour to consume. The feast feeds a number of Medium or smaller creatures equal to your caster level. Large creatures count as two Medium creatures, Huge creatures count as two Large creatures, and so on.
\begin{spelltargets}*{All creatures partaking of the feast}
    Immediately after starting the feast, the target is cured of all disease, fear, poison, sickness, and nausea. After the feast is complete, the target becomes immune to poison and fear effects for 12 hours, and gains temporary hit points equal to 20 \add 1 per two caster levels above 12th.
\end{spelltargets}

\spellnotes If the feast is interrupted for any reason, the ritual is ruined and all of its effects are ended.
\ritualcost*{6}

\spellsection{Identify}{1}
\spellinfo{Divination (Knowledge)}{Arcane, Divine}
\spelltime{1 hour}
\spellrng{Touch}
\begin{spelltarget}{One object}
    \spelleffect You learn all of the target's magic properties, including how to activate any functions it has.
\end{spelltarget}
\spellnotes If used on a cursed item, this ritual only reveals the properties the item appears to have, not the properties of the curse. This ritual does not function when used on an artifact.
\ritualcost*{1}

\spellsection{Illusory Script}{4}
\spelldesc{You write a message woven with a hidden magical command, compelling any viewer except the message's intended recipient to obey you.}
\spellinfo{Enchantment/Illusion (Compulsion, Glamer) [Mind-Affecting, Trap]}{Arcane}
\spelltime{1 minute or longer; see text}
\spellrng{Touch}
\spelldur{One year or until discharged \dismissable}
\spellsr{Yes}
\spellspecial When you perform this ritual, you must specify a password. You must also specify a \spell{suggestion}.
\begin{spelltarget}[Primary]{One object}
    \spelleffect You write text onto the target object. The text is unintelligible once written. Any creature that attempts to read the text and speaks the password is able to see the text as it was originally written.
\end{spelltarget}
\begin{spelltrigger}{A creature reads the primary target without speaking the password}
    \begin{spelltarget}*[Secondary]{The reading creature}[Magic vs. Will]
        \spellsuccess The target is compelled to obey your chosen instructions, as \spell{suggestion}.
    \end{spelltarget}
\end{spelltrigger}
\spellnotes Performing this ritual takes ten times as long as it would normally take to write the same amount of text, to a minimum of 1 minute. A \spell{true seeing} spell can reveal the hidden message.
\ritualcost*{4}

\spellsection{Illusory Wall}{3}
\spellinfo{Illusion (Figment) [Unreal]}{Arcane}
\spelltime{1 minute}
\spellrng{\rngclose}
\spelldur{One year}
\spellline
\spelleffect A figment of your design appears, as \spell{silent image}, except that the image must fit within a 10 foot square.
\spellnotes This ritual is usually used to create illusions of walls, floors, or ceilings. Creatures can identify the illusion, as \spell{silent image}.
\ritualcost*{3}

\spellsection{Instant Refuge}{7}
\spellinfo{Conjuration/Transmutation (Imbuement, Translocation) [Teleportation]}{Divine}
\spelltime{10 minutes}
\spellrng{Touch}
\spelldur{One year or until discharged}
\spellspecial When you perform this ritual, you must specify a command word.
\begin{spelltarget}{One object}
    \spelleffect As a standard action, a creature holding the target object can speak the command word. This teleports the creature and its equipment to your current location, and discharges the ritual.
\end{spelltarget}
\spellnotes If you are on another plane, speaking the command word has no effect and does not discharge the ritual.
\ritualcost*{7}

\spellsection{Instant Retrieval}{7}
\spellinfo{Conjuration/Transmutation (Imbuement, Translocation) [Teleportation]}{Arcane}
\spelltime{10 minutes}
\spellrng{Touch}
\spelldur{One year or until discharged}
\spellspecial When you perform this ritual, you must specify a command word.
\begin{spelltarget}{One object (Medium or smaller)}
    \spelleffect As a standard action, you can speak the command word to teleport the target object into your hand. The object returns to you regardless of its location. If the object is attended, it is not transported, but you know who has the object and roughly where that creature was located when the summons occurred.
\end{spelltarget}
\spellnotes The command word has no effect if said by any other creature. If the object is on another plane, speaking the command word has no effect and does not discharge the ritual.
\ritualcost{7}

\spellsection{Invisibility Purge}{2}
\spelldesc{You surround the touched object or creature with a mobile sphere of power that reveals invisible objects and creatures.}
\spellinfo{Abjuration (Negation)}{Arcane, Divine}
\spelltime{1 minute}
\spellrng{Touch}
\spellfocus{One object or creature}
\spellemanation{\arealarge radius centered on the focus}
\spelldur{\durlong \dismissable}
\spellline
\spelleffect All forms of invisibility are suppressed in the area, causing everything invisible to become visible.
\ritualcost{2}

\spellsection{Ironwood}{3}
\spellinfo{Transmutation (Alteration)}{Nature}
\spelltime{1 hour}
\spellrng{Touch}
\begin{spelltarget}{One wooden object weighing up to 50 pounds}
    \spelleffect The target is transformed into ironwood. While remaining natural wood in almost every way, ironwood is as strong, heavy, and resistant to fire as steel. Spells that affect metal or iron do not function on ironwood. Spells that affect wood do affect ironwood, although ironwood does not burn.
\end{spelltarget}
\spellnotes By performing this ritual multiple times in succession, you may transform wooden objects too heavy to be affected by a single casting of the ritual. Ironwood armor and weapons created through this spell are as durable as their normal steel counterparts, and are freely usable by druids.
\ritualcost*{3}

\spellsection{Legend Lore}{5}
\spellinfo{Divination (Knowledge)}{Arcane}
\spelltime{1 hour; see text}
\spelldur{See text}
\spelleffect This ritual brings to your mind legends about an important person, place, or thing. The amount of information, and the speed of the ritual, depends on how much information you have about the subject.

If the person or thing is at hand, or if you are in the place in question, you learn complete and specific legends immediately after the ritual is complete.

If you only know detailed information about the subject, you learn legends that are either complete or specific, but not both, 1d10 days after the ritual is complete.

If you only know rumors about the subject, you learn incomplete and vague legends 1d10 weeks after the ritual is complete.

\spellnotes If the person, place, or thing is not of legendary importance, you gain no information. In general, characters who are 11th level and higher are ``legendary," as are the sorts of creatures they contend with, the major magic items they wield, and the places where they perform their key deeds.
\ritualcost{5}

\spellsection{Light}{1}
\spellinfo{Illusion (Figment) [Light]}{Arcane, Divine, Nature}
\spelltime{1 minute}
\spellrng{\rngtouch}
\spelldur{\durlong \dismissable}
\begin{spelltarget}{One creature or object}
    \spelleffect The target glows like a torch, shedding bright light in a \areamed radius (and dim light for an additional 20 feet).

    As a swift action, you can suppress or intensify the light, preventing the object from shedding light or causing it to shed light in up to a \arealarge radius (and dim light for an additional 50 feet). Either effect lasts for 1 round.
\end{spelltarget}
\ritualcost{1}

\spellsection{Liveoak}{6}
\spellinfo{Transmutation (Animation)}{Divine}
\spelltime{10 minutes}
\spellrng{Touch}
\spelltgt{One tree (Huge or smaller)}
\spelldur{Thirty days}
\spellspecial When you perform this ritual, you must specify an action and a triggering condition. The condition must be something that a typical human in the target's place could detect.
\begin{spelltrigger}{The triggering condition occurs}
    \spelleffect The target tree animates for 5 minutes, functioning as a treant. The treant takes whatever action was specified. The actions it can take are limited, and are usually restricted to attacking foes.

    At the end of the 5 minutes, the treant tries to return to its original location and take root again. If unable to do so, it takes root where it is. After animating, 24 hours must pass before the target can animate again.
\end{spelltrigger}
\ritualcost{6}

\spellsection{Mage's Magnificent Mansion}{7}
\spellinfo{Conjuration (Creation, Translocation) [Planar]}{Sor/Wiz}
\spelltime{10 minutes}
\spelltwocol{\spellzone{Up to ten 10-foot cubes}}{\spellrng{\rngclose}}
\spelldur{One week \dismissable}
\spellspecial When you perform this ritual, you may designate any number of creatures you can see.
\spellline
\spelleffect This ritual creates an extradimensional dwelling that has a single entrance on the plane from which the ritual was performed. The entry point looks like a faint shimmering in the air that is 5 feet wide and 10 feet high. Only those you designate may enter the mansion, and the portal is shut and made invisible behind you when you enter. You may open it again from your own side at will. Once observers have passed beyond the entrance, they are in a magnificent foyer with numerous chambers beyond. The atmosphere is clean, fresh, and warm.
\par You can create any floor plan you desire to the limit of the spell's effect. The place is furnished and contains sufficient foodstuffs to serve a nine-course banquet to a dozen people per caster level. A staff of near-transparent servants (as many as two per caster level), liveried and obedient, wait upon all who enter. The servants function as the \spell{unseen servant} ritual, except that they are visible and can go anywhere in the mansion.
\spellnotes Since the place can be entered only through its special portal, outside conditions do not affect the mansion, nor do conditions inside it pass to the plane beyond.
\ritualcost*{7}

\spellsection{Mage's Private Sanctum}{5}
\spellinfo{Abjuration (Ward)}{Arcane}
\spelltime{10 minutes}
\spelltwocol{\spellzone{\arealarge radius}}{\spellrng{\rngclose}}
\spelldur{\durext \dismissable}
\spellline
\spelleffect Anyone looking into the area from outside sees only a dark, foggy mass. Darkvision cannot penetrate it. No sounds, no matter how loud, can escape the area, so nobody can eavesdrop from outside. Those inside can see out normally.

Divination (Awareness) and Divination (Scrying) spells cannot perceive anything within the area. The ward prevents speech between those inside and those outside (because it blocks sound), but it does not prevent magical communication.
\spellnotes This ritual does not prevent creatures or objects from moving into and out of the area. It can be made permanent with a \spell{permanency} ritual.
\ritualcost*{5}

\spellsection{Magic Mouth}{1}
\spellinfo{Illusion (Figment)}{Arcane}
\spelltime{1 standard action}
\spellrng{\rngclose}
\spelltgt{One creature or object}
\spelldur{Thirty days or until discharged \dismissable}
\spellsr{Yes}
\spellspecial When you perform this ritual, you must specify a triggering condition. The condition must be something that a typical human in the target's place could detect. You must also specify a message of twenty-five words or less.
\begin{spelltrigger}{The triggering condition occurs}
    \spelleffect The target appears to grow an enchanted mouth, and it speaks the chosen message aloud.
\end{spelltrigger}
\spellnotes The mouth's pronunciation is too inexact to activate effects that require command words. This ritual can be made permanent with a \spell{permanency} ritual.
\ritualcost*{1}

\spellsection{Mark of Justice}{5}
\spellinfo{Necromancy (Life) [Curse]}{Divine}
\spelltime{10 minutes}
\spellrng{Touch}
\spelltgt{One creature}
\spelldur{Permanent \dismissable; see text}
\spellsr{Yes}
\spellspecial When you perform this ritual, you specify a prohibited activity.
\begin{spelltrigger}{The target performs the prohibited activity}
    \spelleffect The target is cursed, as \spell{bestow curse}.
\end{spelltrigger}
\spellnotes \cursespellnotes
\ritualcost{5}

\spellsection{Mending}{1}
\spellinfo{Transmutation (Alteration)}{Arcane, Divine, Nature}
\spelltime{1 minute}
\spellrng{Touch}
\spellsr{Yes}
\begin{spelltarget}{One object}
    \spelleffect The target is healed for 2d6 damage \add d6 per two caster levels above 2nd. 
\end{spelltarget}
\spellnotes This ritual has no effect on broken or destroyed items. 
\ritualcost*{1}

\spellsectioncomma{Mending}{Greater}{3}
\spellinfo{Transmutation (Alteration)}{Arcane, Divine, Nature}
\spelltime{1 minute}
\spellrng{Touch}
\spellsr{Yes}
\begin{spelltarget}{One object}
    \spelleffect The target is healed for 6d6 damage \add d6 per two caster levels above 2nd. For every 10 points of healing granted by this ritual, it can instead cure 1 point of critical damage.
\end{spelltarget}
\spellnotes This ritual has no effect on destroyed items.

\spellsection{Mind Blank}{9}
\spelldesc{The target is protected from all effects that detect or influence emotions or thoughts.}
\spellinfo{Abjuration (Shielding)}{Arcane}
\spelltime{1 standard action}
\spellrng{\rngclose}
\spelldur{\durext}
\spellsr{Yes}
\begin{spelltarget}{One creature}
    \spelleffect The target is immune to all mind-affecting effects, even beneficial ones. In addition, it is immune to Divination (Awareness) and Divination (Scrying) effects. In the case of scrying that views an area the creature is in, such as \spell{arcane eye}, the spell works, but the creature simply isn't detected. Scrying attempts that are targeted specifically at the target do not work at all.
\end{spelltarget}
\ritualcost{9}

\spellsection{Mount}{1}
\spellinfo{Conjuration (Summoning)}{Arcane}
\spelltime{1 minute}
\spellrng{\rngclose}
\spelldur{\durext \dismissable}
\spellline
\spelleffect You summon a light horse or a pony (your choice) to serve you as a mount. The steed serves willingly and well. The mount comes with a bit and bridle and a riding saddle.
\ritualcost{1}

\spellsection{Move Earth}{6}
\spellinfo{Transmutation (Alteration) [Earth]}{Nature}
\spelltime{Ten minutes per cube}
\spellrng{\rnglong}
\spellzone{Up to ten 10-foot cubes, none more than 10 feet below the ground}
\spellline
\spelleffect This ritual moves dirt, clay, loam, and sand in the area. This can collapse embankments, move hillocks, shift dunes, and so forth. However, it cannot move rock formations. 

This ritual does not violently break the surface of the ground. The terrain begins moving as soon as the ritual is started, and finishes moving when the ritual is complete. It moves in wavelike crests and troughs, with the earth reacting with glacierlike fluidity until the desired result is achieved. Trees, structures, rock formations, and such are mostly unaffected except for changes in elevation and relative topography.
\spellnotes This ritual cannot be used for tunneling, and is generally too slow to trap or bury creatures. Its primary use is for digging or filling moats or for adjusting terrain contours before a battle. It has no effect on earth creatures.
\ritualcost*{6}

\spellsection{Neutralize Poison}{4}
\spellinfo{Transmutation (Alteration) [Healing]}{Divine, Nature}
\spelltime{1 minute}
\spellrng{Touch}
\spelldur{\durext \dismissable}
\spellsr{Yes}
\begin{spelltarget}{One creature or object (Medium or smaller)}
    \spelleffect The target becomes immune to poison, and poison currently in the target is neutralized. Any poison-based abilities the target has, such as a creature's venom, are suppressed.
\end{spelltarget}
\spellnotes This does not reverse instantaneous effects from poison, such as hit point damage, temporary ability damage, or effects that don't go away on their own.
\ritualcost*{4}

\spellsection{Nightmare}{6}
\spellinfo{Divination/Illusion (Communication, Phantasm) [Mind-Affecting]}{Arcane, Divine, Nature}
\spelltime{10 minutes}
\spellrng{Unlimited \rngunrestricted}
\spellsr{Yes}
\spellspecial You can target any creature that you can unambiguously identify, regardless of its location.
\begin{spelltarget}{One creature}[Magic vs. Will]
    \spelleffect You send the target a dream, as \spell{dream}, except that the dream is a nightmare.
    \spellsuccess The target is unable to sleep restfully. It regains no hit points for resting, and is \fatigued when it wakes up. It can still regain the use of its spells and other abilities.
    \spellspecial Your attack is modified based on how well you know the subject and what sort of physical connection (if any) you have to that creature, as shown on \tref{Scrying Connection Modifiers}.
\end{spelltarget}
\spellnotes As \spell{dream}.
\ritualcost*{6}

\spellsection{Nondetection}{3}
\spelldesc{You protect your ally from detection by divination spells.}
\spellinfo{Abjuration (Shielding)}{Arcane, Divine}
\spelltime{1 minute}
\spellrng{Touch}
\spelldur{\durext \dismissable}
\spellsr{Yes}
\begin{spelltarget}{One creature or object}
    \spelleffect The target gains spell resistance against Divination (Awareness) and Divination (Scrying) spells equal to 10 \add your caster level. If you are the target, the spell resistance granted is equal to 15 \add your caster level.
\end{spelltarget}
\ritualcost{3}

\spellsection{Overland Flight}{6}
\spellinfo{Transmutation (Imbuement)}{Arcane}
\spelltime{10 minutes}
\spellrng{Touch}
\spelldur{\durext}
\begin{spelltarget}{One creature}
    \spelleffect As a standard action, the target can concentrate to gain a 30 foot fly speed with good maneuverability for 1 round. When not concentrating, the target falls at only 60 feet per round, preventing it from taking any damage from landing.

    The target can use this spell for long-distance movement, concentrating to fly each round. However, it cannot take a forced march. This means it can typically cover 30 miles in an ten-hour period of flight.
\end{spelltarget}
\spellnotes An \unencumbered creature with a fly speed can fly through the air. See \pcref{Flying}, for more details.
\ritualcost*{6}

\spellsection{Pass Without Trace}{2}
\spellinfo{Transmutation (Imbuement)}{Nature}
\spelltime{1 minute}
\spellrng{Touch}
\spelldur{\durlong \dismissable}
\spellsr{Yes}
\begin{spelltarget}{One touched creature}
    \spelleffect The target can move through any type of terrain and leave neither footprints nor scent. Tracking the subjects is virtually impossible by nonmagical means; the DC is increased by 20.
\end{spelltarget}
\ritualcost{2}

\spellsection{Passwall}{5}
\spellinfo{Transmutation (Alteration)}{Arcane, Divine, Nature}
\spelltime{10 minutes}
\spelltwocol{\spellzone{One 5-foot cube per three caster levels}}{\spellrng{Touch}}
\spelldur{\durext \dismissable}
\spellline
\spelleffect This ritual creates a passage through wooden, plaster, or stone walls, but not through other materials. The material within the area simply ceases to exist for the duration of the spell. If the wall's thickness is more than the depth of the passage created, then a single \spell{passwall} makes a niche or short tunnel. Several \spell{passwall} spells can then form a continuing passage to breach very thick walls. When \spell{passwall} ends, creatures within the passage are ejected out the nearest exit.
\spellnotes If someone dispels the \spell{passwall} or you dismiss it, creatures in the passage are ejected out of the closest exit.
\ritualcost*{5}

\spellsection{Permanency}{5}
\spellinfo{Universal}{Arcane}
\spelltime{10 minutes}
\spellrng{See text}
\spelltgt{See text}
\spelldur{Permanent; see text}
\spellline
\spelleffect This ritual makes the duration of certain other spells and rituals permanent. You first cast the desired spell or perform the desired ritual, and then follow it with the \spell{permanency} ritual. Depending on the effect, you must be of a minimum caster level and must expend a specific gp value of diamond dust as a material component.
\par You can make the following effects permanent in regard to yourself.
\begin{dtable}
    \begin{tabularx}{\columnwidth}{>{\lcol}X >{\lcol}X l}
        \thead{Spell} & \thead{Minimum Caster Level} & \thead{GP Cost} \\
        \spell{Arcane sight} & 14th & 3,750 gp \\
        \spell{Comprehend languages} & 12th & 2,500 gp \\
        \spell{Darkvision} & 12th & 2,500 gp \\
        \spell{Read magic} & 10th & 1,250 gp \\
        \spell{See invisibility} & 12th & 2,500 gp \\
        \spell{Tongues} & 14th & 3,750 gp
    \end{tabularx}
\end{dtable}
You cannot cast these spells on other creatures. This application of permanency can be dispelled only by a caster of higher level than you were when you performed the \spell{permanency} ritual.

In addition to personal use, permanency can be used to make the following effects permanent on yourself, another creature, or an object (as appropriate).
\begin{dtable}
    \begin{tabularx}{\columnwidth}{>{\lcol}X >{\lcol}X l}
        \thead{Spell} & \thead{Minimum Caster Level} & \thead{GP Cost} \\
        Magic fang & 10th & 1,250 gp \\
        Magic fang, greater & 14th & 3,750 gp \\
        Reduce person & 10th & 1,250 gp \\
        Telepathic bond* & 14th & 3,750 gp
    \end{tabularx}
    *Only bonds two creatures per casting of \spell{permanency}.
\end{dtable}
Additionally, the following effects can only be rendered permanent on objects or areas.
\begin{dtable}
    \lcaption{Spells Subject To Permanency}
    \begin{tabularx}{\columnwidth}{>{\lcol}X >{\lcol}X l}
        \thead{Spell} & \thead{Minimum Caster Level} & \thead{GP Cost} \\
        Alarm & 10th & 1,250 gp \\
        Animate objects & 18th & 6,250 gp \\
        Create sound & 10th & 1,250 gp \\
        Dancing lights & 10th & 1,250 gp \\
        Gust of wind & 10th & 1,250 gp \\
        Invisibility & 14th & 3,750 gp \\
        Mage's private sanctum & 18th & 6,250 gp \\
        Magic mouth & 10th & 1,250 gp \\
        Phase door & 22nd & 8,750 gp \\
        %Prismatic sphere & 17th & 11,250 gp \\
        %Prismatic wall & 16th & 10,000 gp \\
        Shrink item & 14th & 3,750 gp \\
        Solid fog & 16th & 5,000 gp \\
        Stinking cloud & 14th & 3,750 gp \\
        Symbol of death & 22nd & 8,750 gp \\
        Symbol of destruction & 22nd & 8,750 gp \\
        Symbol of terror & 22th & 8,750 gp \\
        Symbol of insanity & 20th & 7,500 gp \\
        Symbol of pain & 18th & 6,250 gp \\
        Symbol of persuasion & 16th & 5,000 gp \\
        Symbol of sleep & 22nd & 8,750 gp \\
        Wall of fire & 18th & 6,250 gp \\
        Wall of force & 18th & 6,250 gp \\
        Wall of ice & 16th & 5,000 gp \\
        Web & 12th & 2,500 gp
    \end{tabularx}
\end{dtable}
\spellnotes Spells cast on other creatures, objects, or locations (not on you) are vulnerable to \spell{dispel magic} as normal.
\spellmat{See tables above.}
%\ritualcost{5}

\spellsection{Permanent Image}{6}
\spellinfo{Illusion (Figment) [Unreal]}{Arcane}
\spelltwocol{\spellzone{\arealarge radius}}{\spellrng{\rngmed}}
\spelldur{Permanent \dismissable}
\spellline
\spelleffect A figment of your design appears within the area, as \spell{major image}.
\ritualcost{6}

\spellsection{Phantom Steed}{3}
\spelldesc{You create a quasi-real horselike creature to serve you or one of your allies. It has a black head and body, gray mane and tail, and smoke-colored, insubstantial hooves that make no sound.}
\spellinfo{Illusion/Transmutation (Imbuement, Shadow)}{Arcane}
\spelltime{1 minute}
\spellrng{\rngclose}
\spelldur{\durext \dismissable}
\spellline
\spelleffect This ritual creates a Large, horselike creature that can only be ridden by you or one person you designate. On its body, it bears what seems to be a saddle, bit, and bridle sized perfectly for its intended rider. 
It moves at a speed of 10 feet per two caster levels. It can bear its rider's weight plus up to 10 pounds per caster level.

The phantom steed cannot fight. Its physical defenses are all equal to 10 \add half your caster level, and it has 10 hit points \add 1 per caster level. If it loses all its hit points, the phantom steed disappears.

\par As your caster level increases, the phantom steed gains additional abilities, as noted below.

\par 8th level: The mount can ride over sandy, muddy, or even swampy ground without difficulty or decrease in speed.

\par 12th level: The mount can use \spell{water walk} at will (as the spell, no action required to activate this ability).

\par 16th level: The mount can use \spell{air walk} at will (as the spell, no action required to activate this ability) for up to 1 round at a time, after which it falls to the ground.

\par 20th level: The mount can fly at its speed (good maneuverability) by concentrating, as the \spell{overland flight} spell.
\ritualcost{3}

\spellsection{Phase Door}{7}
\spellinfo{Conjuration (Creation/Translocation) [Planar]}{Arcane, Divine}
\spelltime{10 minutes}
\spelltwocol{\spellzone{One 5-foot cube per three caster levels}}{\spellrng{Touch}}
\spelldur{\durext or until expended}
\spellspecial When you perform this ritual, you may specify a triggering condition. The condition must be something that a typical human could detect.
\spellline
\spelleffect This ritual creates an ethereal passage through wooden, plaster, or stone walls, but not other materials. One phase door offering entrance to the passage appears at each end of the area. The entrances are invisible and inaccessible to all creatures except you, and any creatures who meet the triggering condition.

The passage can be used a number of times equal to half your caster level before the ritual is expended. A creature using the phase door enters the Ethereal Plane when it enters the passage, and reappears on the Matereal Plane when it leaves the passage. Creatures inside the passage cannot leave the boundaries of the passage on the Ethereal Plane. The door does not allow light, sound, or spell effects through it, nor can you see through it without using it.

\spellnotes A \spell{true seeing} spell or similar magic reveals the presence of a phase door but does not allow its use. If either end of the passage is dispelled, the passage is destroyed, and anyone within the passage is harmlessly ejected from the closest exit. This ritual can be made permanent with a \spell{permanency} ritual.
\ritualcost*{7}

\spellsection{Planar Binding}{7}
\spellinfo{Abjuration/Conjuration (Translocation) [Planar]}{Arcane}
\spellspecial This ritual functions like \spell{lesser planar binding}, except that it can call up to three extraplanar creatures whose combined level does not exceed 12.
\ritualcost{7}

\spellsectioncomma{Planar Binding}{Greater}{9}
\spellinfo{Abjuration/Conjuration (Translocation) [Planar] [see text for lesser planar binding]}{Arcane}
\spellspecial This ritual functions like \spell{lesser planar binding}, except that it can call up to three extraplanar creatures whose combined level does not exceed 18.
\ritualcost{9}

\spellsectioncomma{Planar Binding}{Lesser}{5}
\spelldesc{You attempt to lure a creature from another plane into a trap so you can compel it to perform a service for you.}
\spellinfo{Abjuration/Conjuration (Translocation) [Planar] [see text]}{Arcane}
\spelltime{2 hours}
\spellrng{See text}
\spellsr{Yes}
\spellspecial When you perform this ritual, you must specify a type of extraplanar creature, or the proper name of a specific extraplanar creature. You may also cast \spell{dimensional anchor}.
\spellline
\spelleffect This ritual creates a trap to hold a creature, as \spell{binding}. In addition, it calls a specified extraplanar creature into the binding.
\begin{spelltarget}{One extraplanar creature}[Magic vs. Will]
    \spellspecial If you specified a type of extraplanar creature, the target is a random creature of that type. If you named a specific extraplanar creature, the target is that ccreature.
    \spellsuccess The target is teleported from its original location into the \spell{binding}. You must then make a separate attack to determine whether the creature is bound. If the creature is not bound, it typically attacks you.

    If the creature is bound, you can attempt to negotiate with it (see \pcref{Persuasion}), or to magically compel it into servitude with rituals like \spellindirect{geas/quest}{geas}
    \spellfailure The target learns your identity and that you attempted to bind it, but is otherwise unaffected.
\end{spelltarget}
\spellnotes This ritual grants no special ability to return the creature to its home plane after binding it. If you use this ritual to call an air, chaotic, earth, evil, fire, good, lawful, or water creature, it gains the appropriate descriptor.
\ritualcost*{5}

\spellsection{Plane Shift}{6}
\spellinfo{Conjuration (Translocation) [Planar]}{Arcane, Divine}
\spelltime{1 minute}
\spellrng{Touch}
\spellsr{Yes}
\spellspecial When you perform this ritual, you specify a destination on another plane adjacent to your current plane. Precise accuracy is nigh impossible, and the actual destination is usually 1d100 miles away from the intended destination.
\begin{spelltargets}{You and up to five willing creatures}
    \spelleffect The target is teleported to the destination.
\end{spelltargets}
\spellnotes The Astral Plane connects to every plane, but transit from other planes is usually more limited. From the Material Plane, you can reach the Ethereal Plane, the Plane of Shadow, or the Astral Plane.
\ritualcost*{7}

\begin{comment}
\spellsection{Prestidigitation}{1}
\spellinfo{Universal}{Arcane}
\spelltime{1 minute}
\spelltwocol{\spelltgt{See text}}{\spellrng{Personal/\rngclose}}
\spelldur{1 hour}
\spellsr{Yes}
\spelleffect This ritual enables you to perform simple magical effects for 1 hour on objects or creatures within \rngclose range of you by concentrating (a standard action). The effects are minor and have severe limitations. A \spell{prestidigitation} can slowly lift 1 pound of material. It can color, clean, or soil items in a 1-foot cube each round. It can chill, warm, or flavor 1 pound of nonliving material. It cannot deal damage or affect the concentration of spellcasters. \spell{Prestidigitation} can create small objects, but they look crude and artificial. The materials created by a prestidigitation spell are extremely fragile, and they cannot be used as tools, weapons, or spell components.
\spellnotes This ritual lacks the power to duplicate spell effects. Any actual change to an object (beyond just moving, cleaning, or soiling it) persists only 1 hour. Creatures and attended objects, such as the clothes a creature is wearing, cannot be affected unless the creature is willing. Spell resistance does not apply to you, but it does apply to any objects or creatures you try to affect.
\ritualcost*{1}
\end{comment}

\spellsection{Programmed Image}{7}
\spellinfo{Illusion (Figment)}{Arcane}
\spelltwocol{\spellzone{\arealarge radius}}{\spellrng{\rngmed}}
\spelldur{One year or until triggered, then \durmed}
\spellspecial When you perform this ritual, you must specify a triggering condition. The condition must be something that a typical human could detect within the area.
\begin{spelltrigger}{The triggering condition occurs within the area}.
    \spelleffect A figment of your design appears within the area, as \spell{major image}.
\end{spelltrigger}
\spellnotes Creatures can identify the illusion, as \spell{silent image}.
\ritualcost{7}

\spellsection{Prying Eyes}{5}
\spellinfo{Conjuration (Creation)}{Arcane}
\spelltwocol{\spelltime{1 minute}}{\spellrng{One mile}}
\spelldur{\durext; see text \dismissable}
\spelleffect You create a number of semitangible, visible magical orbs (called ``eyes") equal to your caster level. These eyes move out, scout around, and return as you direct them when casting the spell. Each eye can see 120 feet (normal vision only) in all directions.
\par While the individual eyes are quite fragile, they're small and difficult to spot. Each eye is a Fine construct, about the size of a small apple, that has 1 hit point, a physical defense of 18 (\plus8 size bonus), flies at a speed of 30 feet with perfect maneuverability, and has a \plus16 Stealth modifier. It has a Perception modifier equal to your caster level and is subject to illusions, darkness, fog, and any other factors that would affect your ability to receive visual information about your surroundings. An eye traveling through darkness must find its way by touch.
\par When you create the eyes, you specify instructions you want them to follow in a command of no more than twenty-five words. Any knowledge you possess is known by the eyes as well.
\par In order to report their findings, the eyes must return to your hand. Each replays in your mind all it has seen during its existence. It takes an eye 1 round to replay 1 hour of recorded images. After relaying its findings, an eye disappears.
\par If an eye ever gets more than 1 mile away from you, it instantly ceases to exist. However, your link with the eye is such that you won't know if the eye was destroyed because it wandered out of range or because of some other event.
\par The eyes exist for up to 1 hour per caster level or until they return to you.
\spellnotes \spellindirect{dispel magic}{Dispel magic} can destroy eyes. If an eye is sent into darkness, it could hit a wall or similar obstacle and destroy itself.
\ritualcost*{5}

\spellsectioncomma{Prying Eyes}{Greater}{8}
\spellinfo{Conjuration/Divination (Awareness, Creation)}{Arcane}
\spellrng{10 miles}
\spelleffect This spell functions like \spell{prying eyes}, except that the eyes can range farther from you and they can see all things as they actually are, as the \spell{true seeing} spell.
\ritualcost{8}

\spellsection{Purify Food and Drink}{1}
\spellinfo{Transmutation (Alteration)}{Divine, Nature}
\spelltwocol{\spelltgt{5 cu. ft. of contaminated food and water}}{\spellrng{Touch}}
\spellsr{Yes}
\spellattack{Fortitude negates (object)}
\spelleffect This spell makes spoiled, rotten, poisonous, or otherwise contaminated food and water pure and suitable for eating and drinking. This spell does not prevent subsequent natural decay or spoilage. Unholy water and similar food and drink of significance is spoiled by \spell{purify food and drink}, but the spell has no effect on creatures of any type or magical liquids, such as potions.
\spellnotes Water weighs about 8 pounds per gallon. One cubic foot of water contains roughly 8 gallons and weighs about 60 pounds.
\ritualcost*{1}

\spellsection{Raise Dead}{5}
\spellinfo{Necromancy (Life, Soul) [Healing]}{Divine}
\spelltime{1 minute}
\spelltwocol{\spelltgt{Dead creature touched}}{\spellrng{Touch}}
\spellattack{None; see text}
\spelleffect You restore life to a creature that has been dead for no longer than thirty days. The subject's soul must be free and willing to return. If the subject's soul is not willing to return, the spell does not work; therefore, a subject that wants to return receives no saving throw.
\par Coming back from the dead is an ordeal. The subject gains a negative level. If the subject was 1st level when it died, it takes 2 points of Constitution drain instead. This negative level or Constitution drain lasts for thirty days, or until the subject next gains a level. A character who died with spells prepared has a 50\% chance of losing any given spell upon being raised, in addition to losing spells for losing a level. A spellcasting creature that doesn't prepare spells (such as a sorcerer) has a 50\% chance of losing any given unused spell slot as if it had been used to cast a spell, in addition to losing spell slots for losing a level.
\par A raised creature has 0 hit points and 1 critical damage. Any attribute damage is removed. Normal poison and normal disease are cured in the process of raising the subject, but magical diseases and curses are not undone. While the spell closes mortal wounds and repairs lethal damage of most kinds, the body of the creature to be raised must be whole. Otherwise, missing parts are still missing when the creature is brought back to life. None of the dead creature's equipment or possessions are affected in any way by this spell.
\spellnotes A creature who has been turned into an undead creature or killed by a death effect can't be raised by this spell. Constructs, elementals, outsiders, and undead creatures can't be raised. The spell cannot bring back a creature that has died of old age.
\spellmat{Diamonds worth a total of least 2,500 gp.}
\ritualcost{5}

\spellsection{Read Magic}{1}
\spellinfo{Divination (Knowledge)}{Arcane, Divine}
\spelltwocol{\spelltgt{You}}{\spellrng{Personal}}
\spelldur{\durlong}
\spelleffect You gain the ability to decipher magical inscriptions on objects -- books, scrolls, weapons, and the like -- that would otherwise be unintelligible. This deciphering does not normally invoke the magic contained in the writing, although it may do so in the case of a cursed scroll. Furthermore, once the spell is cast and you have read the magical inscription, you are thereafter able to read that particular writing without recourse to the use of \spell{read magic}. The spell allows you to identify a \spell{glyph of warding} with a DC 13 Spellcraft check, a \spell{greater glyph of warding} with a DC 16 Spellcraft check, or any \spellindirect{symbol of death}{symbol} spell with a Spellcraft check (DC 10 \add spell level).
\spellnotes \spell{Read magic} can be made permanent with a \spell{permanency} ritual.
\ritualcost*{1}

\spellsection{Reincarnate}{4}
\spellinfo{Conjuration/Necromancy (Creation, Soul)}{Nature}
\spelltime{1 hour}
\spelltwocol{\spelltgt{Dead creature touched}}{\spellrng{Touch}}
\spellattack{None; see text}
\spelleffect With this spell, you bring back a dead creature in another body, provided that its death occurred no more than one week before the casting of the spell and the subject's soul is free and willing to return. If the subject's soul is not willing to return, the spell does not work; therefore, a subject that wants to return receives no saving throw.
\par Since the dead creature is returning in a new body, all physical ills and afflictions are repaired. The condition of the remains and the cause of death, except death from old age, is not a factor. So long as some small portion of the creature's body still exists, it can be reincarnated, but the portion receiving the spell must have been part of the creature's body at the time of death. The magic of the spell creates an entirely new body for the soul to inhabit from the natural elements at hand. The body ages to match the age of the original creature over the course of the ritual.
\par A reincarnated creature is identical to the original creature in all respects, including physical abilities, except for its race. First eliminate the subject's racial adjustments (since it is no longer of his previous race) and then apply the adjustments found below to its remaining attribute scores. The subject gains a permanent negative level. (If this reduction would put its effective level at 0, it can't be reincarnated). This negative level lasts for thirty days, or until the subject next gains a level.
\par It's possible for the change in the subject's attributes to make it difficult for it to pursue its previous character class. If this is the case, the subject is well advised to become a multiclass character.
\par For a humanoid creature, the new incarnation is determined using the following table. For nonhumanoid creatures, a similar table of creatures of the same type should be created.
\par Any effect that would prevent a resurrection spell from bringing a creature back from the dead also prevents reincarnate from bringing the creature back. Constructs, elementals, outsiders, and undead creatures can't be reincarnated. The spell cannot bring back a creature who has died of old age.
\begin{dtable}
    \begin{tabularx}{\columnwidth}{l >{\lcol}X l l l}
        d\% & Incarnation & Str & Dex & Con \\
        01 & Bugbear & \plus4 & \plus2 & \plus2 \\
        02--13 & Dwarf & \plus0 & \plus0 & \plus2 \\
        14--25 & Elf & \plus0 & \plus2 & \minus2 \\
        26 & Gnoll & \plus4 & \plus0 & \plus2 \\
        27--38 & Gnome & \minus2 & \plus0 & \plus2 \\
        39--42 & Goblin & \minus2 & \plus2 & \plus0 \\
        43--52 & Half-elf & \plus0 & \plus0 & \plus0 \\
        53--62 & Half-orc & \plus2 & \plus0 & \plus0 \\
        63--74 & Halfling & \minus2 & \plus2 & \plus0 \\
        75--89 & Human & \plus0 & \plus0 & \plus0 \\
        90--93 & Kobold & \minus4 & \plus2 & \minus2 \\
        94 & Lizardfolk & \plus2 & \plus0 & \plus2 \\
        95--98 & Orc & \plus4 & \plus0 & \plus0 \\
        99 & Troglodyte & \plus0 & \minus2 & \plus4 \\
        100 & Other & ? & ? & ?
    \end{tabularx}
\end{dtable}
\par The reincarnated creature gains all abilities associated with its new form, including forms of movement and speeds, natural armor, natural attacks, extraordinary abilities, and the like, but it doesn't automatically speak the language of the new form.
\spellnotes A wish or a miracle spell can restore a reincarnated character to his or her original form.
\spellmat{Rare oils and unguents worth a total of least 500 gp, spread over the remains.}
\ritualcost{4}

\spellsection{Remove Blindness/Deafness}{3}
\spellinfo{Necromancy (Life) [Healing, Positive]}{Divine, Nature}
\spelltime{1 minute}
\spelltwocol{\spelltgt{Creature touched}}{\spellrng{Touch}}
\spellsr{Yes}
\spelleffect This ritual cures blindness or deafness (your choice), whether the effect is normal or magical in nature. The spell does not restore ears or eyes that have been lost, but it repairs them if they are damaged.
\ritualcost{3}

\spellsection{Remove Curse}{4}
\spellinfo{Necromancy}{Arcane, Divine, Nature}
\spelltime{1 minute}
\spelltwocol{\spelltgt{Creature or item touched}}{\spellrng{Touch}}
\spellsr{Yes}
\spelleffect This ritual removes all curses on an object or a creature. If an item is inherently cursed, the spell does not remove that curse, though it typically enables the creature afflicted with any such cursed item to remove and get rid of it. Certain special curses may not be removed by this spell or may only be removed by a caster of a certain level or higher.
\ritualcost{4}

\spellsection{Remove Disease}{3}
\spellinfo{Transmutation (Alteration) [Healing]}{Divine, Nature}
\spelltime{1 minute}
\spelltwocol{\spelltgt{Creature touched}}{\spellrng{Touch}}
\spellsr{Yes}
\spelleffect This ritual cures all diseases that the subject is suffering from. The spell also kills parasites, including green slime and others. Certain special diseases may not be cured by this spell or may be cured only by a caster of a certain level or higher.
\spellnotes This ritual does not prevent reinfection after a new exposure to the same disease at a later date.
\ritualcost*{3}

\spellsection{Restoration}{4}
\spellinfo{Necromancy (Life) [Healing, Positive]}{Divine}
\spelleffect This ritual functions like \spell{lesser restoration}, except that it is more effective. It cures all temporary ability damage, restores all points permanently drained from a single attribute (your choice if more than one is drained), eliminates any fatigue or exhaustion suffered by the target, and removes one negative level.
\spellnotes This ritual does not restore negative levels or Constitution drain acquired due to death.
\spellmat{Diamond dust worth 50 gp that is sprinkled over the target.}
\ritualcost{4}

\spellsectioncomma{Restoration}{Greater}{7}
\spellinfo{Necromancy (Life) [Healing, Positive]}{Divine}
\spelltime{10 minutes}
\spelleffect This ritual functions like \spell{lesser restoration}, except that it is more effective. It dispels all magical effects penalizing the creature's abilities, cures all temporary ability damage, and restores all points permanently drained from all attribute scores. It also eliminates fatigue and exhaustion, and removes all forms of insanity, confusion, and similar mental effects. Finally, it removes all negative levels afflicting the healed creature. 
\spellnotes This ritual does not restore negative levels or Constitution drain acquired due to death.
\spellmat{Diamond dust worth 250 gp that is sprinkled over the target.}
\ritualcost{7}

\spellsectioncomma{Restoration}{Lesser}{2}
\spellinfo{Necromancy (Life) [Healing, Positive]}{Divine}
\spelltime{1 minute}
\spelltwocol{\spelltgt{Creature touched}}{\spellrng{Touch}}
\spellsr{Yes}
\spelleffect This ritual dispels any magical effects reducing one of the subject's attribute scores or cures up to 5 points of temporary ability damage to one of the subject's attribute scores. It also eliminates any fatigue or exhaustion suffered by the character.
\spellnotes This ritual does not restore permanent ability drain.
\ritualcost*{2}

\spellsection{Resurrection}{7}
\spellinfo{Necromancy (Life, Soul)}{Clr}
\spelltime{10 minutes}
\spelleffect This ritual functions like \spell{raise dead}, except that you are able to restore life and complete strength to any deceased creature.
\par The condition of the remains is not a factor. So long as some small portion of the creature's body still exists, it can be resurrected, but the portion receiving the spell must have been part of the creature's body at the time of death. (The remains of a creature hit by a \spell{disintegrate} spell count as a small portion of its body.) The creature can have been dead no longer than 10 years per caster level.
\par Upon completion of the spell, the creature is immediately restored to full hit points, vigor, and health, with no loss of prepared spells. The subject gains a negative level. If the subject was 1st level when it died, it takes 2 points of Constitution drain instead. This negative level or Constitution drain lasts for thirty days, or until the subject next gains a level.
\spellnotes You can resurrect someone killed by a death effect or someone who has been turned into an undead creature and then destroyed. You cannot resurrect someone who has died of old age. Constructs, elementals, outsiders, and undead creatures can't be resurrected.
\spellmat{A sprinkle of holy water and diamonds worth a total of at least 5,000 gp.}
\ritualcost{7}

\spellsection{Screen}{8}
\spellinfo{Illusion (Figment, Glamer) [Unreal]}{Arcane}
\spelltime{10 minutes}
\spelltwocol{\spellzone{30 ft. cube/level}}{\spellrng{\rngclose}}
\spelldur{\durext}
\spellattack{None or Will disbelief (if interacted with); see text}
\spelleffect This spell combines several elements to create a powerful protection from scrying and direct observation. When casting the spell, you dictate what will and will not be observed in the spell's area. The illusion created must be stated in general terms. Once the conditions are set, they cannot be changed.
\par Attempts to scry the area automatically detect the image stated by you with no save allowed. Sight and sound are appropriate to the illusion created.
\par Direct observation may allow a save (as per a normal illusion), if there is cause to disbelieve what is seen. Even entering the area does not cancel the illusion or necessarily allow a save, assuming that the beings concealed by the illusion take care to stay out of the way of observers and the observers do not directly interact with the illusion.
\ritualcost{8}

\spellsection{Sculpt Sound}{4}
\spellinfo{Illusion (Glamer)}{Arcane}
\spelltime{10 minutes}
\spelltwocol{\spelllimit{\areamed radius}}{\spellrng{\rngmed}}
\spelltgts{Five creatures or objects within the area}
\spelldur{\durext \dismissable}
\spellsr{Yes}
\spellattack{Will negates (object)}
\spelleffect You change the sounds that creatures or objects make. You can deaden sounds that exist or transform sounds into other sounds, but you cannot create new sounds where none existed or amplify the volume of existing sounds. All affected creatures or objects must have their sounds altered in the same way.

At any point during the spell's duration, you can concentrate on the effect (a standard action) to change the sound for all affected creatures within your range. If you do so, the remaining duration is reduced by 1 hour, and the effect is dismissed for any affected creature outside your range.
\spellnotes You can change the qualities of sounds but cannot create words with which you are unfamiliar yourself. A spellcaster whose voice is changed or quieted dramatically is treated as deafened when casting spells (20\% chance of failure)
\ritualcost*{4}

\spellsection{Scrying}{5}
\spellinfo{Divination (Scrying)}{Arcane, Divine, Nature}
\spelltwocol{\spelltime{1 hour}}{\spellrng{See text}}
\spelldur{\durmed \dismissable}
\spellsr{Yes}
\spellattack{Will negates}
\spelleffect \par You can see and hear some creature, which may be at any distance. If the subject succeeds on a Will save, the scrying attempt simply fails. The difficulty of the save depends on how well you know the subject and what sort of connection (if any) you have to that creature. Furthermore, if the subject is on another plane, it gets a \plus5 bonus on its Will save.

\begin{dtable}
    \lcaption{Scrying Modifiers}
    \begin{tabularx}{\columnwidth}{>{\lcol}X >{\lcol}p{4em}}
        \thead{Knowledge} & \thead{Attack Modifier} \\
        None\footnotetemp{1} & \minus10 \\
        Secondhand (you have heard of the subject) & \minus5 \\
        Firsthand (you have met the subject) & \plus0 \\
        Familiar (you know the subject well) & \plus \\
        \thead{Connection} & \thead{Attack Modifier} \\
        Likeness or picture & \plus2 \\
        Possession or garment & \plus4 \\
        Body part, lock of hair, bit of nail, etc. & \plus10
    \end{tabularx}
    1 You must have some sort of connection to a creature you have no knowledge of.
\end{dtable}

\par If the save fails, you can see and hear the subject and the subject's immediate surroundings (approximately 10 feet in all directions of the subject). If the subject moves, the sensor follows at a speed of up to 150 feet.
\par As with all divination (scrying) spells, the sensor has your full visual acuity, including any magical effects. In addition, the following spells have a 5\% chance per caster level of operating through the sensor: \spell{detect chaos}, \spell{detect evil}, \spell{detect good}, \spell{detect law}, \spell{message}, \spell{read magic}, and \spell{tongues}.
\par If the save succeeds, it is immune to any further scrying attempts by you for the next 24 hours.
\mfx{Arcana Focus:} A mirror of finely wrought and highly polished silver costing not less than 500 gp. The mirror must be at least 2 feet by 4 feet.
\mfx{Religion Focus:} A holy water font costing not less than 50 gp.
\mfx{Nature Focus:} A natural pool of water.
\ritualcost{5}

\spellsectioncomma{Scrying}{Greater}{7}
\spellinfo{Divination (Scrying)}{Arcane, Divine, Nature}
\spelltime{1 minute}
\spelldur{\durext}
\spelleffect This ritual functions like \spell{scrying}, except that it can be cast more quickly and lasts longer. Additionally, all of the following spells function reliably through the sensor: \spell{comprehend languages}, \spell{detect alignment}, \spell{message}, \spell{read magic}, and \spell{tongues}.
\ritualcost{7}

\spellsection{Secret Chest}{4}
\spellinfo{Conjuration (Teleportation) [Object-Affecting, Planar]}{Arcane}
\spelltime{10 minutes}
\spelltwocol{\spelltgt{One chest and up to 1 cu. ft. of goods/caster level}}{\spellrng{See text}}
\spelldur{Sixty days or until discharged}
\spelleffect You hide a chest on the Ethereal Plane for as long as sixty days and can retrieve it at will. The chest can contain up to 1 cubic foot of material per caster level (regardless of the chest's actual size, which is about 3 feet by 2 feet by 2 feet). If any living creatures are in the chest, there is a 75\% chance that the spell simply fails. Once the chest is hidden, you can retrieve it by concentrating (a standard action), and it appears next to you.
\par The chest must be exceptionally well crafted and expensive, constructed for you by master crafters. The cost of such a chest is never less than 500 gp. Once it is constructed, you must make a tiny replica (of the same materials and perfect in every detail), so that the miniature of the chest appears to be a perfect copy. (The replica costs 25 gp.) You can have only one pair of these chests at any given time. The chests are nonmagical and can be fitted with locks, wards, and so on, just as any normal chest can be.
\par To hide the chest, you cast the ritual while touching both the chest and the replica. The chest vanishes into the Ethereal Plane. You need the replica to recall the chest. After sixty days, there is a cumulative chance of 5\% per day that the chest is irretrievably lost. If the miniature of the chest is lost or destroyed, there is no way, not even with a wish spell, that the large chest can be summoned back, although an extraplanar expedition might be mounted to find it.
\par Living things in the chest eat, sleep, and age normally, and they die if they run out of food, air, water, or whatever they need to survive.
\spellfocus{The chest (which costs 500 gp) and its replica (which costs 25 gp).}
\ritualcost{4}

\spellsection{Secret Page}{3}
\spellinfo{Transmutation (Alteration)}{Arcane}
\spelltime{10 minutes}
\spelltwocol{\spelltgt{Page touched, up to 3 sq. ft. in size}}{\spellrng{Touch}}
\spelldur{Permanent}
\spelleffect This ritual alters the contents of a page so that they appear to be something entirely different. The text of a spell can be changed to show even another spell. \spell{Explosive runes} or \spell{sepia snake sigil} can be cast upon the secret page.
\par A \spell{comprehend languages} spell alone cannot reveal a \spell{secret page}'s contents. You are able to reveal the original contents by speaking a special word. You can then peruse the actual page, and return it to its secret page form at will. You can also remove the spell by double repetition of the special word.
\spellnotes A Spellcraft check reveals dim magic on the page in question but does not reveal its true contents. \spell{True seeing} reveals the presence of the hidden material but does not reveal the contents. A \spell{secret page} spell can be dispelled, and the hidden writings can be destroyed by means of an \spell{erase} spell.
\ritualcost*{3}

\spellsection{Secure Shelter}{4}
\spellinfo{Conjuration (Creation, Summoning)}{Arcane}
\spelltwocol{\spelltime{10 minutes}}{\spellrng{\rngclose}}
\spelldur{\durext \dismissable}
\spelleffect You conjure a sturdy cottage or lodge made of material that is common in the area where the spell is cast. The floor is level, clean, and dry. In all respects the lodging resembles a normal cottage, with a sturdy door, two shuttered windows, and a small fireplace.
\par The shelter has no heating or cooling source (other than natural insulation qualities). Therefore, it must be heated as a normal dwelling, and extreme heat adversely affects it and its occupants. The dwelling does, however, provide considerable security otherwise -- it is as strong as a normal stone building, regardless of its material composition. The dwelling resists flames and fire as if it were stone. It is impervious to normal missiles (but not the sort cast by siege engines or giants).
\par The door, shutters, and even chimney are secure against intrusion, the former two being locked and the latter secured by an iron grate at the top and a narrow flue. In addition, these three areas are protected by an alarm spell. Finally, an \spell{unseen servant} is conjured to provide service to you for the duration of the shelter.
\par The secure shelter contains rude furnishings -- eight bunks, a trestle table, eight stools, and a writing desk.
\ritualcost{4}

\spellsection{Seeming}{4}
\spellinfo{Illus (Glamer) [Unreal]}{Arcane}
\spelltwocol{\spelllimit{\areamed radius}}{\spellrng{\rngclose}}
\spelltgts{One creature per caster level in the area}
\spelldur{\durlong \dismissable}
\spelleffect This spell functions like \spell{disguise self}, except that it affects multiple creatures. Affected creatures resume their normal appearances if slain.
\spellsr{Yes}
\ritualcost{4}

\spellsection{Sending}{4}
\spellinfo{Divination (Communication)}{Arcane, Divine}
\spelltime{10 minutes}
\spellrng{Unlimited \rngunrestricted}
\spellsr{Yes}
\spellspecial You can target any creature with whom you are familiar, regardless of its location. You must have seen the creature at least once.
\begin{spelltarget}{One creature}
    \spelleffect You send the target a short verbal message. The message must be twenty-five words or less, and speaking the message must not take longer than five rounds.

    After receiving the message, the target has five rounds to reply with a similarly restricted message. After that time, the sending is complete.
\end{spelltarget}
\spellnotes If the creature in question is not on the same plane of existence as you are, there is a 10\% chance that the sending does not arrive. (Local conditions on other planes may worsen this chance considerably.)
\ritualcost*{4}

\spellsection{Sepia Snake Sigil}{4}
\spellinfo{Abjuration/Transmutation (Temporal, Warding) [Trap]}{Arcane}
\spelltime{10 minutes}
\spellrng{Touch}
\spellfocus{One object with writing on it (Small or smaller)}
\spelldur{Thirty days or until discharged}
\spellline
\begin{spelltrigger}{A creature reads the focus object while within 5 feet of it}
    \begin{spelltarget}*{The reading creature}[Magic vs. Reflex]
        \spelldur{Thirty days}
        \spellsuccess The target is \slowed. In addition, a sigil of a snake appears on its forehead, marking its condition.
    \end{spelltarget}
\end{spelltrigger}
\spellnotes Magic traps such as \spell{sepia snake sigil} can be detected with the Perception skill and disabled with the Devices skill. The DC in each case is 25 \add spell level, or 28 for \spell{sepia snake sigil}.
\ritualcost{3}

\spellsection{Sequester}{7}
\spellinfo{Abjuration/Transmutation (Shielding, Temporal)}{Arcane}
\spelltime{10 minutes}
\spelltwocol{\spelltgt{One willing creature or object (up to ten 2 ft. cubes) touched}}{\spellrng{Touch}}
\spelldur{Thirty days \dismissable}
\spellsr{No or Yes (object)}
\spellattack{None or Will negates (object)}
\spelleffect The subject is invisible and cannot be detected by divination spells. However, it also becomes comatose and is effectively in a state of suspended animation until the spell wears off or is dispelled.
\spellnotes The Will save prevents an attended or magical object from being sequestered. There is no save to see the sequestered creature or object or to detect it with a divination spell. This ritual does not prevent the subject from being discovered through other means, such as touch.
\ritualcost*{7}

\spellsection{Shadow Walk}{6}
\spellinfo{Conjuration/Illusion (Shadow, Translocation) [Planar]}{Arcane}
\spelltime{1 standard action}
\spelltwocol{\spelltgts{You and up to four touched creatures}}{\spellrng{Touch}}
\spelldur{\durext \dismissable}
\spellsr{Yes}
\spellattack{Will negates}
\spelleffect To use this spell, you must be in an area of shadowy illumination. You and any other subjects are then transported along a coiling path of shadowstuff to the edge of the Material Plane where it borders the Plane of Shadow. The effect is largely illusory, but the path is quasi-real. You can take more than one creature along with you, but all must be touching each other.
\par In the region of shadow, you move at a rate of 50 miles per hour, moving normally on the borders of the Plane of Shadow but much more rapidly relative to the Material Plane. Thus, you can use this spell to travel rapidly by stepping onto the Plane of Shadow, moving the desired distance, and then stepping back onto the Material Plane.
\par Because of the blurring of reality between the Plane of Shadow and the Material Plane, you can't make out details of the terrain or areas you pass over during transit, nor can you predict perfectly where your travel will end. It's impossible to judge distances accurately, making the spell virtually useless for scouting or spying. Furthermore, when the spell effect ends, you are shunted 1d10\mtimes 100 feet in a random horizontal direction from your desired endpoint. If this would place you within a solid object, you are shunted an additional 1d10\mtimes 1,000 feet in the same direction. If this would still place you within a solid object, you (and any creatures with you) are shunted to the nearest empty space available, but the strain of this activity renders each creature fatigued (no save).

\par Any creatures touched by you when \spell{shadow walk} is cast also make the transition to the borders of the Plane of Shadow.
\par They may opt to follow you, wander off through the plane, or stumble back into the Material Plane (50\% chance for either of the latter results if they are lost or abandoned by you). Creatures unwilling to accompany you into the Plane of Shadow receive a Will saving throw, negating the effect if successful.
\ritualcost{6}

\spellsection{Shambler}{9}
\spellinfo{Conjuration/Transmutation (Animation, Creation)}{Natural}
\spelltime{1 hour}
\spelltwocol{\spelllimit{\areamed radius}}{\spellrng{\rngmed}}
\spelldur{Seven days or seven months \dismissable; see text}
\spelleffect The shambler spell creates 1d4\plus2 shambling mounds. The creatures willingly aid you in combat or battle, perform a specific mission, or serve as bodyguards. The creatures remain with you for seven days unless you dismiss them. If the shamblers are created only for guard duty, however, the duration of the spell is seven months. In this case, the shamblers can only be ordered to guard a specific site or location. Shamblers summoned to guard duty cannot move outside the spell's range, which is measured from the point where each first appeared. 
\par The shamblers have resistance to fire as normal shambling mounds do only if the terrain is rainy, marshy, or damp.
\ritualcost{9}

\spellsection{Shape Metal}{4}
\spellinfo{Transmutation (Alteration)}{Arcane, Divine, Nature}
\spelleffect This ritual functions like \spell{shape wood}, except that you make a Craft (metal) check, and you shape metal instead of wood. 
\ritualcost{4}

\spellsection{Shape Stone}{3}
\spellinfo{Transmutation (Alteration) [Earth]}{Arcane, Divine, Nature}
\spelleffect This ritual functions like \spell{shape wood}, except that you make a Craft (stone) check, and you shape stone instead of wood.
\ritualcost{3}

\spellsection{Shape Wood}{2}
\spellinfo{Transmutation (Alteration)}{Arcane, Divine, Nature}
\spelltime{10 minutes}
\spelltwocol{\spelltgt{Nonmagical wood touched, affecting up to 10 cubic feet}}{\spellrng{Touch}}
\spellsr{Yes}
\spellattack{No}
\spelleffect When you perform this ritual, you make a Craft (wood) check to change the shape of an existing piece of wood, or part of a larger piece of wood, to suit your purposes. You gain a \plus10 enhancement bonus on the check, and you need no additional tools.
\ritualcost{2}

\spellsection{Snare}{3}
\spellinfo{Transmutation (Alteration) [Trap]}{Nature}
\spelltime{1 minute}
\spelltwocol{\spelltgt{Touched nonmagical circle of vine, rope, or thong with a 2 ft. diameter \add 2 ft./level}}{\spellrng{Touch}}
\spelldur{Permanent until triggered or broken}
\spelleffect This ritual enables you to make a snare that functions as a magic trap. The snare can be made from any supple vine, a thong, or a rope. When you cast \spell{snare} upon it, the cordlike object blends with its surroundings (Perception DC 23 to locate). One end of the snare is tied in a loop that contracts around one or more of the limbs of any creature stepping inside the circle.
\par If a strong and supple tree is nearby, the snare can be fastened to it. The spell causes the tree to bend and then straighten when the loop is triggered, dealing 1d6 points of damage to the creature trapped and lifting it off the ground by the trapped limb or limbs. If no such tree is available, the cordlike object tightens around the creature, dealing no damage but causing it to be entangled.
\par The snare is magical. To escape, a trapped creature must make a DC 23 Escape Artist check or a DC 23 Strength check that is a full-round action. The snare has a physical defense of 2 and 5 hit points. A successful escape from the snare breaks the loop and ends the spell.
\ritualcost{3}

\spellsection{Soul Bind}{9}
\spellinfo{Necromancy (Soul)}{Arcane, Divine}
\spelltime{1 minute}
\spellcmp{Verbal, Somatic, Focus}
\spelltwocol{\spelltgt{Corpse}}{\spellrng{\rngclose}}
\spelldur{Permanent}
\spelleffect You draw the soul from a newly dead body and imprison it in a black sapphire gem. The subject must have been dead for no longer than 5 rounds before the ritual is started. The soul, once trapped in the gem, cannot be returned through \spell{clone}, \spell{raise dead}, \spell{reincarnate}, \spell{resurrection}, \spell{true resurrection}, or even a \spell{miracle} or a \spell{wish}. Only by destroying the gem or dispelling the spell on the gem can one free the soul (which is then still dead).
\spellfocus{A black sapphire of at least 500 gp value for every level possessed by the creature whose soul is to be bound. If the gem is not valuable enough, it shatters when the binding is attempted. (While creatures have no concept of level as such, the value of the gem needed to trap an individual can be researched. Remember that this value can change over time as creatures gain more levels.)}
\ritualcost{9}

\spellsection{Speak with Dead}{3}
\spellinfo{Divination/Necromancy (Communication, Flesh) [Language-Dependent]}{Arcane, Divine}
\spelltime{10 minutes}
\spelltwocol{\spelltgt{One dead creature}}{\spellrng{\rngclose}}
\spelldur{10 minutes \dismissable}
\spellattack{Will negates; see text}
\spelleffect You grant the semblance of life and intellect to a corpse, allowing it to answer several questions that you put to it. You may ask one question per two caster levels. Unasked questions are wasted if the duration expires. The corpse's knowledge is limited to what the creature knew during life, including the languages it spoke (if any). Answers are usually brief, cryptic, or repetitive. If the creature's alignment was different from yours, the corpse gets a Will save to resist the spell as if it were alive.
\par If the corpse has been subject to \spell{speak with dead} within the past week, the new spell fails. You can cast this spell on a corpse that has been deceased for any amount of time, but the body must be mostly intact to be able to respond. A damaged corpse may be able to give partial answers or partially correct answers, but it must at least have a mouth in order to speak at all.
\par This spell does not let you actually speak to the person (whose soul has departed). It instead draws on the imprinted knowledge stored in the corpse. The partially animated body retains the imprint of the soul that once inhabited it, and thus it can speak with all the knowledge that the creature had while alive. The corpse, however, cannot learn new information. Indeed, it can't even remember being questioned.
\spellnotes This spell does not affect a corpse that has been turned into an undead creature.
\ritualcost*{3}

\spellsection{Symbol of Death}{7}
\spellinfo{Abjuration/Necromancy (Life, Warding) [Death, Trap]}{Arcane}
\spelltime{10 minutes}
\spelltwocol{\spelltgt{One object}}{\spellrng{Touch}}
\spelldur{Permanent or until discharged}
\spellsr{Yes}
\spellattack{Fortitude negates}
\spelleffect This ritual allows you to scribe a potent rune of power upon a surface. When it is triggered, the triggering creature suffers the effects of a \spell{finger of death} spell. It can be set to trigger in response to any visual cue that takes place within \rngclose range of the symbol which can be ``seen'' by the symbol. Darkness does not stop the symbol's sight, but physical obstacles or illusions block it. Once the ritual is performed, the triggering conditions cannot be changed.
\par When scribing a \spell{symbol of death}, you can specify a password or phrase. Anyone speaking the password cannot trigger the rune so long as the creature remains within \rngmed range of the rune. If the creature leaves that range and returns later, it must say the password again.
\spellnotes The symbol is considered to have a Perception modifier equal to your caster level. Magic traps such as \spell{symbol of death} can be detected with the Perception skill and disabled with the Devices skill. The DC in each case is 25 \add spell level, or 32 for \spell{symbol of death}.
\par \spellindirect{read magic}{Read magic} allows you to identify a \spell{symbol of death} with a DC 17 Spellcraft check. Of course, if the symbol of death is set to be triggered by reading it, this will trigger the symbol.
\par \spellindirect{dispel magic}{Dispel magic} can remove a \spell{symbol of death}, but an \spell{erase} spell has no effect. Destruction of the surface where a \spell{symbol of death} is inscribed destroys the symbol but also triggers it on the closest creature within \rngclose range of the symbol.
\par \spell{Symbol of death} can be made permanent with a \spell{permanency} spell. A permanent \spell{symbol of death} that is disabled becomes inactive for 10 minutes, then can be triggered again as normal.
\ritualcost{7}

\spellsection{Symbol of Destruction}{7}
\spellinfo{Abjuration/Necromancy (Flesh, Warding) [Death, Trap]}{Divine}
\spelleffect This spell functions like \spell{symbol of death}, except that the triggering creature instead suffers the effects of the \spell{destruction} spell.
\spellnotes Magic traps such as \spell{symbol of destruction} can be detected with the Perception skill and disabled with the Devices skill. The DC in each case is 25 \add spell level, or 32 for \spell{symbol of destruction}.
\ritualcost*{7}

\spellsection{Symbol of Fear}{7}
\spellinfo{Abjuration/Enchantment (Emotion, Warding) [Fear, Mind-Affecting, Trap]}{Arcane}
\spellsr{Yes}
\spellattack{Will negates}
\spelleffect This spell functions like \spell{symbol of death}, except that the triggering creature instead suffers the effects of the \spell{terror} spell.
\spellnotes Magic traps such as \spell{symbol of terror} can be detected with the Perception skill and disabled with the Devices skill. The DC in each case is 25 \add spell level, or 32 for \spell{symbol of terror}.
\ritualcost*{7}

\spellsection{Symbol of Insanity}{6}
\spellinfo{Abjuration/Enchantment (Compulsion, Warding) [Mind-Affecting, Trap]}{Arcane}
\spellsr{Yes}
\spellattack{Will negates}
\spelleffect This spell functions like \spell{symbol of death}, except that the triggering creature instead suffers the effects of the \spell{insanity} spell.
\spellnotes Magic traps such as \spell{symbol of insanity} can be detected with the Perception skill and disabled with the Devices skill. The DC in each case is 25 \add spell level, or 31 for \spell{symbol of insanity}.
\ritualcost*{6}

\spellsection{Symbol of Pain}{5}
\spellinfo{Abjuration/Necromancy (Flesh, Warding) [Trap]}{Arcane}
\spelleffect This spell functions like \spell{symbol of death}, except that the triggering creature instead suffers the effects of the \spell{crippling pain} spell.
\spellnotes Magic traps such as \spell{symbol of pain} can be detected with the Perception skill and disabled with the Devices skill. The DC in each case is 25 \add spell level, or 30 for \spell{symbol of pain}.
\ritualcost*{5}

\spellsection{Symbol of Persuasion}{4}
\spellinfo{Abjuration/Enchantment (Compulsion, Warding) [Language-Dependent, Mind-Affecting, Sound-Dependent, Trap]}{Arcane}
\spellsr{Yes}
\spellattack{Will negates}
\spelleffect This spell functions like \spell{symbol of death}, except that the triggering creature is instead affected by the \spell{suggestion} spell. You choose the suggestion when you perform the ritual, and it cannot thereafter be changed.
\spellnotes Magic traps such as \spell{symbol of persuasion} can be detected with the Perception skill and disabled with the Devices skill. The DC in each case is 25 \add spell level, or 29 for \spell{symbol of persuasion}.
\ritualcost*{4}

\spellsection{Symbol of Sleep}{7}
\spellinfo{Abjuration/Enchantment (Compulsion, Warding) [Mind-Affecting, Trap]}{Arcane}
\spellsr{Yes}
\spellattack{Will negates}
\spelleffect This spell functions like \spell{symbol of death}, except that the triggering creature instead suffers the effects of the \spell{deep slumber} spell.
\spellnotes Magic traps such as \spell{symbol of sleep} can be detected with the Perception skill and disabled with the Devices skill. The DC in each case is 25 \add spell level, or 32 for \spell{symbol of sleep}.
\ritualcost*{7}

\spellsection{Sympathetic Vibration}{5}
\spellinfo{Evocation (Energy) [Sonic]}{Arcane}
\spelltime{10 minutes}
\spelltwocol{\spelltgt{One freestanding structure}}{\spellrng{Touch}}
\spelldur{Up to 1 minute}
\spellsr{Yes}
\spellattack{None; see text}
\spelleffect By attuning yourself to a freestanding structure such you can create a damaging vibration within it. Once it begins, the vibration deals 2d10 points of damage per round to the target structure. (Hardness has no effect on the spell's damage.) You can choose at the time of casting to limit the duration of the spell; otherwise it lasts for 1 minute. If the spell is cast upon a target that is not freestanding, the surrounding stone dissipates the effect and no damage occurs.
\spellnotes This ritual cannot affect creatures (including constructs). Since a structure is an unattended object, it gets no saving throw, and typically no spell resistance, to resist the effect.
\ritualcost*{5}

\spellsection{Sympathy}{9}
\spellinfo{Enchantment (Emotion) [Mind-Affecting]}{Arcane, Nature}
\spelltime{1 hour}
\spelltwocol{\spelltgt{One location (up to a 10 ft. cube/level) or one object}}{\spellrng{\rngclose}}
\spelldur{One week \dismissable}
\spellsr{Yes}
\spellattack{Will partial}
\spelleffect You cause an object or location to emanate magical vibrations that attract either a specific kind of intelligent creature or creatures of a particular alignment, as defined by you. The particular kind of creature to be affected must be named specifically. A creature subtype is not specific enough. Likewise, the specific alignment must be named.
\par Creatures of the specified kind or alignment feel elated and pleased to be in the area or desire to touch or to possess the object. The compulsion to stay in the area or touch the object is overpowering. If the save is successful, the creature is released from the enchantment, but a subsequent save must be made 1d6\mult10 minutes later. If this save fails, the affected creature attempts to return to the area or object.
\ritualcost{9}

\spellsection{Telepathic Bond}{3}
\spelldesc{You forge a mental link binding two allies together.}
\spellinfo{Divination/Transmutation (Communication, Imbuement)}{Arcane}
\spelltime{10 minutes}
\spelltwocol{\spelltgts{You and one willing creature, or two willing creatures}}{\spellrng{\rngclose}}
\spelldur{\durext \dismissable}
\spellsr{Yes}
\spelleffect The subjects can communicate mentally through telepathy. The communication is instantaneous across any distance within the same plane.
\spellnotes No special influence is established as a result of the bond. This ritual can be made permanent with a \spell{permanency} ritual.
\ritualcost*{3}

\spellsection{Telepathic Bond, Mass}{6}
\spellinfo{Divination/Transmutation (Communication, Imbuement)}{Arcane}
\spelltgts{You plus up to five willing creatures in a \areamed radius}
\spelleffect This ritual functions like \spell{telepathic bond}, except that it links multiple creatures together into the same bond. Each affected creature can communicate with all other creatures, either privately or to the group as a whole. If desired, you may leave yourself out of the bond forged. This decision must be made at the time of casting.
\spellnotes This ritual can be made permanent with a \spell{permanency} ritual. If you perform this ritual multiple times, you may link each casting of the ritual together such that all subjects may telepathically communicate with each other. 
\ritualcost*{6}

\spellsection{Teleport}{6}
\spellinfo{Conjuration (Translocation) [Teleportation]}{Arcane}
\spelltime{1 minute; see text}
\spelltwocol{\spelltgts{You and up to four touched Medium creatures; see text}}{\spellrng{Personal and touch}}
\spellsr{Yes}
\spellattack{None and Will negates (object)}
\spelleffect This spell instantly transports you and the subjects to a designated destination, which may be as distant as 100 miles away. Each creature can bring along objects whose weight doesn't exceed the creature's maximum load. All creatures to be transported must be in contact with one another, and at least one of those creatures must be in contact with you. After completing the spell, you and anyone bring with you must skip their next actions as they adjust to their new surroundings.
\par By doubling the casting time of the ritual, you can take along an additional Medium creature. You can repeat this process, doubling the new casting time each time, up to a maximum casting time of just over an hour (allowing you to take a total of ten Medium creatures in addition to yourself).
\par You can also bring creatures larger than Medium. A Large creature counts as two Medium creatures, a Huge creature counts as two Large creatures, and so forth.
\par This ritual is unusually difficult to perform correctly. You must have some clear idea of the location and layout of the destination. The clearer your mental image, the more likely the teleportation works. Areas of strong physical or magical energy may make teleportation more hazardous or even impossible. As part of the ritual, you make an Intelligence check against DC 10. The following categories apply modifiers:
\begin{itemize}
    \item ``Very familiar" is a place where you have been very often and where you feel at home. You gain a \plus5 bonus.
    \item ``Studied carefully" is a place you know well, either because you can currently see it, you've been there often, or you have used other means (such as scrying) to study the place for at least one hour. No bonuses or penalties apply.
    \item ``Seen casually" is a place that you have seen more than once but with which you are not very familiar. You take a \minus5 penalty.
    \item ``Viewed once" is a place that you have seen once, possibly using magic. You take a \minus10 penalty.
    \item ``False destination" is a place that does not truly exist or if you are teleporting to an otherwise familiar location that no longer exists as such or has been so completely altered as to no longer be familiar to you. You automatically fail.
\end{itemize}

Success means you arrive on target. Failure means you arrive a random distance away from your intended destination in a random direction. The distance off target is equal to 1d10\mult1d10\% of the distance that you would have traveled to your intended destination. Failure by 10 or more means you arrive in a completely different area within range that is visually or thematically similar to your intended destination.
\spellnotes This ritual is incapable of interplanar travel. Generally, you appear in the closest similar place within range. If no such area exists within the spell's range, the ritual simply fails instead.

Only objects held or in use (attended) by another person receive saving throws. You do not apply spell resistance against this ritual. 
\ritualcost{6}

\spellsection{Teleport Object}{7}
\spellinfo{Conjuration (Translocation) [Teleportation]}{Arcane}
\spelltwocol{\spelltgt{One touched object of up to 50 lb./level and 3 cu. ft./level}}{\spellrng{Touch}}
\spellsr{Yes}
\spellattack{Will negates (object)}
\spelleffect This spell functions like \spell{teleport}, except that it teleports an object, not you. Creatures and magical forces (such as a \spell{delayed blast fireball}) cannot be teleported.
\ritualcost{7}

\spellsection{Teleport, Greater}{8}
\spellinfo{Conjuration (Translocation) [Teleportation]}{Arcane}
\spelleffect This ritual functions like \spell{teleport}, except that you can teleport up to 1,000 miles, and you cannot arrive off target. In addition, you need not have seen the destination, provided that you have an accurate and specific description of the place to which you are teleporting. If you attempt to teleport with insufficient information, with misleading information, or to an invalid location (such as inside a solid object), you simply disappear and reappear in your original location.
\ritualcost{8}

\spellsection{Teleportation Circle}{9}
\spellinfo{Conjuration (Translocation) [Teleportation, Trap]}{Arcane}
\spelltime{10 minutes}
\spellrng{Touch}
\spelldur{\durext or until discharged \dismissable}
\spellsr{Yes}
\spellattack{Will negates}
\spelleffect You create an invisible \areasmall radius circle on the floor or other horizontal surface that teleports, as \spell{greater teleport}, any creature who stands on it to a designated spot. Once you designate the destination for the circle, you can't change it. The ritual fails if the destination is not a valid destination.
\par The circle can teleport a number of creatures equal to your caster level before it ceases to function.
\spellnotes Magic traps such as \spell{teleportation circle} can be detected with the Perception skill and disabled with the Devices skill. The DC in each case is 25 \add spell level, or 34 for \spell{teleportation circle}.
\ritualcost*{9}

\spellsection{Tiny Hut}{3}
\spellinfo{Evocation (Control)}{Arcane}
\spelltime{1 minute}
\spellzone{\areamed radius centered on you}
\spelldur{\durext \dismissable}
\spelleffect You create an unmoving, opaque sphere of any color you desire around your location. Half the sphere projects above the ground, and the lower hemisphere passes through the ground. As many as nine other Medium creatures can fit into the field with you; they can freely pass into and out of the hut without harming it. However, if you remove yourself from the hut, the spell ends.
\par The temperature inside the hut is 70\degree F if the exterior temperature is between 0\degree and 100\degree F. An exterior temperature below 0\degree or above 100\degree lowers or raises the interior temperature on a 1-degree-for-1 basis. The hut also provides protection against the elements, such as rain, dust, and sandstorms. The hut withstands any wind of less than hurricane force, but a hurricane (75\add mph wind speed) or greater force destroys it.
\par The interior of the hut is a hemisphere. Although the force field is opaque from the outside, it is transparent from within. Missiles, weapons, and most spell effects can pass through the hut without affecting it, although the occupants cannot be seen from outside the hut (they have total concealment).
\ritualcost{3}

\spellsection{Tongues}{4}
\spellinfo{Divination (Communication)}{Arcane, Divine, Nature}
\spelltwocol{\spelltgt{Creature touched}}{\spellrng{Touch}}
\spelldur{\durlong}
\spellsr{Yes}
\spelleffect This spell grants the creature touched the ability to speak and understand the language of any intelligent creature, whether it is a racial tongue or a regional dialect. The subject can speak only one language at a time, although it may be able to understand several languages. \spell{Tongues} does not enable the subject to speak with creatures who don't speak.
\spellnotes \spell{Tongues} may be unable to translate dead or extremely obscure languages. It does not predispose any creature addressed toward the subject in any way. \spell{Tongues} can be made permanent with a \spell{permanency} spell.
\ritualcost*{4}

\spellsection{Transport via Plants}{6}
\spellinfo{Conjuration (Translocation) [Teleportation]}{Nature}
\spelleffect This ritual functions like \spell{teleport}, except that both the starting and ending points must be living, Medium or larger plants. You and any other creatures you bring with you step into one plant and out of the other plant.
\begin{comment}
You can enter any normal plant (Medium or larger) and pass any distance to a plant of the same kind in a single round, regardless of the distance separating the two. The entry plant must be alive. The destination plant need not be familiar to you, but it also must be alive. If you are uncertain of the location of a particular kind of destination plant, you need merely designate direction and distance and the transport via plants spell moves you as close as possible to the desired location. If a particular destination plant is desired but the plant is not living, the spell fails and you are ejected from the entry plant.
\par After completing the spell, you and anyone bring with you can't take any actions until the next turn that each creature would have.
\par You can bring along objects as long as their weight doesn't exceed your maximum load. You may also bring one additional willing Medium or smaller creature (carrying gear or objects up to its maximum load) or its equivalent per three caster levels. Use the following equivalents to determine the maximum number of larger creatures you can bring along: A Large creature counts as two Medium creatures, a Huge creature counts as two Large creatures, and so forth. All creatures to be transported must be in contact with one another, and at least one of those creatures must be in contact with you.
\par You can't use this spell to travel through plant creatures.
\par The destruction of an occupied plant slays you and any creatures you have brought along, and ejects the bodies and all carried objects from the tree.
\end{comment}
\ritualcost{6}

\spellsection{Trap the Soul}{8}
\spellinfo{Necromancy (Life, Soul)}{Arcane}
\spelltwocol{\spelltgt{One trigger object; see text}}{\spellrng{Touch}}
\spelldur{Permanent; see text}
\spellsr{Yes}
\spelleffect You imbue an object with the power to trap a creature's life and soul. The object must be inscribed with the creature's name. It holds the trapped creature indefinitely or until it is broken, which allows the creature's body to reform. Outsiders can be compelled to perform a single service for the creature releasing them; other creatures can simply go free.
\par For a creature's soul to be trapped, it must intentionally pick up or accept the object while \bloodied. If it does so, its life force is automatically transferred to the gem without the benefit of a save, though spell resistance applies normally. If it picks up the object while healthy, or if any other creature picks up the object, nothing happens, though magical inspection can reveal the existence of the trap.
\spellmat{In addition to the trigger object, a gem must be used. The gem must be worth at least 500 gp value for every level possessed by the creature to be trapped. When the ritual is performed, the gem shatters into dust which is embedded into the trigger object. If the gem is not valuable enough, the creature will suffer no ill effect when it picks up the trigger object, though the ritual's effect will still be discharged, rending the trigger object useless. (While creatures have no concept of level as such, the value of the gem needed to trap an individual can be researched. Remember that this value can change over time as creatures gain more levels.)}
\spellfocus{A special trigger object, prepared as described above, is needed.}
\ritualcost{8}

\spellsection{Tree Stride}{5}
\spellinfo{Conjuration (Translocation) [Teleportation]}{Nature}
\spelltime{1 minute}
\spelltwocol{\spelltgt{You}}{\spellrng{Personal}}
\spelleffect This ritual functions like \spell{teleport}, except that it only affects you, and both the starting and ending points must be Large or larger trees.
\begin{comment}
You gain the ability to enter trees and move from inside one tree to inside another tree. The first tree you enter and all others you enter must be of the same kind, must be living, and must have girth at least equal to yours. By moving into an oak tree (for example), you instantly know the location of all other oak trees within transport range (see below) and may choose whether you want to pass into one or simply step back out of the tree you moved into. You may choose to pass to any tree of the appropriate kind within the transport range as shown on the following table.
\begin{dtable}
    \begin{tabularx}{\columnwidth}{>{\lcol}X >{\lcol}X}
        Type of Tree & Transport Range \\
        Oak, ash, yew & 3,000 feet \\
        Elm, linden & 2,000 feet \\
        Other deciduous & 1,500 feet \\
        Any coniferous & 1,000 feet \\
        All other trees & 500 feet
    \end{tabularx}
\end{dtable}
\par You may move into a tree up to one time per caster level (passing from one tree to another counts only as moving into one tree). The spell lasts until the duration expires or you exit a tree. Each transport is a full-round action.
\par You can, at your option, remain within a tree without transporting yourself, but you are forced out when the spell ends. If the tree in which you are concealed is chopped down or burned, you are slain if you do not exit before the process is complete.
\end{comment}
\ritualcost{5}

\spellsection{True Resurrection}{9}
\spellinfo{Necromancy (Life, Soul)}{Divine}
\spelltime{1 hour}
\spelleffect This spell functions like \spell{raise dead}, except that you can resurrect a creature that has been dead for as long as 10 years per caster level. This spell can even bring back creatures whose bodies have been destroyed, provided that you unambiguously identify the deceased in some fashion (reciting the deceased's time and place of birth or death is the most common method).
\par Upon completion of the spell, the creature is immediately restored to full hit points, vigor, and health, with no negative levels, constitution drain, or loss of prepared spells.
\par You can revive someone killed by a death effect or someone who has been turned into an undead creature and then destroyed. This spell can also resurrect elementals or outsiders, but it can't resurrect constructs or undead creatures.
\spellnotes Even \spell{true resurrection} can't restore to life a creature who has died of old age.
\spellmat{A sprinkle of holy (or unholy) water and diamonds worth a total of at least 10,000 gp.}
\ritualcost{9}

\spellsection{Undetectable Alignment}{2}
\spellinfo{Abjuration (Shielding)}{Arcane}
\spelltime{1 minute}
\spelltwocol{\spelltgt{One creature or object}}{\spellrng{\rngclose}}
\spelldur{\durext \dismissable}
\spellsr{Yes}
\spellattack{Will negates (object)}
\spelleffect This ritual conceals the alignment of an object or a creature from all forms of divination.
\ritualcost{2}

\spellsection{Unhallow}{5}
\spellinfo{Evocation (Power) [Evil]}{Divine}
\spelltime{24 hours}
\spelltwocol{\spellzone{\arealarge radius}}{\spellrng{\rngclose}}
\spelldur{Instantaneous/1 year}
\spellsr{See text}
\spellattack{See text}
\spelleffect Unhallow makes a particular site, building, or structure an unholy site. This has three major effects.
\par First, the site or structure is guarded by a \spell{magic circle against good} effect.
\par Second, any dead body interred in an unhallowed site costs half the normal material components to raise as an undead creature.
\par Finally, you may choose to fix a single spell effect to the unhallowed site. The spell effect lasts for one year and functions throughout the entire site, regardless of its normal duration and area or effect. You may designate whether the effect applies to all creatures, creatures that share your faith or alignment, or creatures that adhere to another faith or alignment. At the end of the year, the chosen effect lapses, but it can be renewed or replaced simply by casting unhallow again.
\par Spell effects that may be tied to an unhallowed site include \spell{aid}, \spell{bane}, \spell{bless}, \spell{cause fear}, \spell{darkness}, \spell{daylight}, \spell{death ward}, \spell{detect evil}, \spell{dimensional anchor}, \spell{discern lies}, \spell{endure elements}, \spell{freedom of movement}, \spell{invisibility purge}, \spell{protection from energy}, \spell{remove fear}, \spell{resist energy}, \spell{silence}, \spell{tongues}, and \spell{zone of truth}.
\par Saving throws and spell resistance might apply to these spells' effects. (See the individual spell descriptions for details.)
\spellnotes An area can receive only one hallow or unhallow spell (and its associated spell effect) at a time. If an area is hallowed, it cannot be unhallowed.
\spellmat{Herbs, oils, and incense worth at least 500 gp, plus 500 gp per level of the spell to be tied to the unhallowed area.}
\ritualcost{5}

\spellsection{Unseen Servant}{1}
\spellinfo{Conjuration/Evocation (Creation, Control)}{Arcane}
\spelltwocol{\spelltime{1 minute}}{\spellrng{\rngmed}}
\spelldur{\durlong \dismissable}
\spelleffect An unseen servant is an invisible, mindless, shapeless force that performs simple tasks at your command. It can run and fetch things, open unstuck doors, and hold chairs, as well as clean and mend. The servant can perform only one activity at a time, but it repeats the same activity over and over again if told to do so as long as you remain within range. It can open only normal doors, drawers, lids, and the like. It has an effective Strength score of 2 (so it can lift 20 pounds or drag 100 pounds). It can trigger traps and such, but it can exert only 20 pounds of force, which is not enough to activate certain pressure plates and other devices. It can't perform any task that requires a skill check with a DC higher than 10 or that requires a check using a skill that can't be used untrained. Its speed is 15 feet.
\par The servant cannot attack in any way. It cannot be killed, but it dissipates if it takes 6 points of damage from area attacks. (It gets no saves against attacks.) If you attempt to send it beyond the spell's range (measured from your current position), the servant ceases to exist.
\ritualcost{1}

\spellsection{Vision}{7}
\spellinfo{Divination (Knowledge)}{Arcane}
\spelltime{1 minute}
\spelleffect This ritual functions like \spell{legend lore}, except that it works more quickly but produces some strain on you. You pose a question about some person, place, or object, then perform the ritual. If the person or object is at hand or if you are in the place in question, you receive a vision about it by succeeding on a caster level check (1d20 \add caster level) against DC 20. If only detailed information on the person, place, or object is known, the DC is 25, and the information gained is incomplete. If only rumors are known, the DC is 30, and the information gained is vague. After the spell is complete, you are exhausted.
\par If you perform this ritual while fatigued or exhausted, you must make a Fortitude save (DC 15 if fatigued or DC 20 if exhausted) or immediately fall unconscious for ten minutes. When you wake up, you remember the information (if any).
\ritualcost{7}

\spellsection{Water Breathing}{3}
\spellinfo{Transmutation (Imbuement)}{Arcane, Divine, Nature}
\spelltime{10 minutes}
\spelltwocol{\spelltgt{Living creatures touched}}{\spellrng{Touch}}
\spelldur{\durext \dismissable; see text}
\spellsr{Yes}
\spelleffect The transmuted creatures can breathe water freely. Divide the duration evenly among all the creatures you touch.
\spellnotes The ritual does not make creatures unable to breathe air.
\ritualcost*{3}

\spellsection{Whispering Wind}{2}
\spellinfo{Divination (Communication) [Air]}{Arcane, Nature}
\spelltime{1 minute}
\spellrng{10 miles}
\spelldur{\durext or until discharged}
\spelleffect You send a message or sound on the wind to a designated spot. The whispering wind travels to a specific location within range that is familiar to you, provided that it can find a way to the location. A whispering wind is as gentle and unnoticed as a zephyr until it reaches the location. It then delivers its whisper-quiet message or other sound. Note that the message is delivered regardless of whether anyone is present to hear it. The wind then dissipates.
\par You can prepare the spell to bear a message of no more than twenty-five words, cause the spell to deliver other sounds for 1 round, or merely have the whispering wind seem to be a faint stirring of the air. You can likewise cause the whispering wind to move as slowly as 1 mile per hour or as quickly as 1 mile per 10 minutes.
\spellnotes This spell cannot speak verbal components, use command words, or activate magical effects.
\ritualcost*{2}

\spellsection{Wind Walk}{6}
\spellinfo{Transmutation (Polymorph) [Air]}{Divine, Nature}
\spelltime{1 standard action}
\spelltwocol{\spelltgts{You and up to four touched creatures}}{\spellrng{Touch}}
\spelldur{\durext \dismissable; see text}
\spellsr{Yes}
\spelleffect \par You alter the substance of your body to a cloudlike vapor (as the \spell{gaseous form} spell) and move through the air, possibly at great speed. You can take other creatures with you, each of which acts independently.
\par Normally, a wind walker flies at a speed of 10 feet with perfect maneuverability. If desired by the subject, a magical wind wafts a wind walker along at up to 600 feet per round (60 mph) with poor maneuverability. Wind walkers are not invisible but rather appear misty and translucent. If fully clothed in white, they are likely to be mistaken for clouds, fog, vapors, or the like.
\par A wind walker can regain its physical form as desired and later resume the cloud form. Each change to and from vaporous form takes 5 rounds, which counts toward the duration of the ritual (as does any time spent in physical form). As noted above, you can dismiss the ritual, and you can even dismiss it for individual wind walkers and not others.
\par For the last minute of the ritual's duration, a wind walker in cloud form automatically descends 60 feet per round (for a total of 600 feet), though it may descend faster if it wishes. This descent serves as a warning that the ritual is about to end.
\ritualcost{6}

\spellsection{Zone of Truth}{2}
\spellinfo{Enchantment (Inhibition) [Mind-Affecting]}{Arcane, Divine}
\spelltime{1 minute}
\spelltwocol{\spellzone{\areamed radius}}{\spellrng{\rngmed}}
\spelldur{\durmed}
\spellsr{Yes}
\spellattack{Will negates}
\spelleffect Creatures within the area (or those who enter it) can't speak any deliberate and intentional lies. Each potentially affected creature is allowed a save to avoid the effects when the spell is cast or when the creature first enters the area. Affected creatures are aware of this enchantment. Therefore, they may avoid answering questions to which they would normally respond with a lie, or they may be evasive as long as they remain within the boundaries of the truth. Creatures who leave the area are free to speak as they choose.
\ritualcost{2}
