\chapter{Spells}

An \M{} or \F{} appearing at the end of a spell's name in the spell lists denotes a spell with a material or focus component, respectively, that is not normally included in a spell component pouch.

\parhead{Order of Presentation} In the spell lists and the spell descriptions that follow them, the spells are presented in alphabetical order by name except for those belonging to certain spell chains.
When a spell's name begins with ``lesser,'' ``greater,'' or ``mass,'' the spell description is alphabetized under the second word of the spell name instead.

\parhead{Hit Values} The term ``Hit Values" is used synonymously with ``character levels" for effects that affect a number of Hit Values of creatures. Creatures with Hit Values only from their race, not from classes, have character levels equal to their Hit Values.

\parhead{Caster Level} A spell's power often depends on caster level. Every spellcaster has a caster level which is primarily determined by its class level. A creature with no classes who uses spells has a caster level equal to its Hit Values unless otherwise specified. The word ``level'' in the spell lists that follow always refers to casting power.

\parhead{Enhancement Bonuses} If a spell or magical effect provides a numerical bonus, it is an enhancement bonus unless otherwise stated. Enhancement bonuses do not stack with each other; only the highest bonus applies.

\parhead{Targets, Subjects, Creatures, and Characters} All of these words refer to whatever is affected by the spell. ``Targets" and ``Subjects'' are used for spells that target individual creatures or objects. Spells with instantaneous duration typically refer to ``targets", while spells that have a duration typically refer to ``subjects''. ``Creatures" and ``characters" are used interchangeably, and are typically used with spells that do not affect specific targets.

\parhead{Restricted Spells} Some spells on the sorcerer/wizard spell list have their names written in italics. This indicates that the spell is a restricted spell. Restricted spells can only be learned by sorcerer and wizard class features, and are not generally available; see the class descriptions for more details.

\small

\section{Bard Spells}

\subsection{1st-level Bard Spells}
\begin{spelllist}
\spellhead{Attraction} Subject has an attraction you specify. 
\spellhead{Cause Fear} One creature flees for short duration.
\spellhead{Comprehend Languages} You understand all spoken and written languages.
\spellhead{Create Sound} Figment sounds.
\spellhead{Cure Light Wounds} Cures 2d8 damage.
\spellhead{Dancing Lights} Creates torches or other lights.
\spellhead{Detect Secret Doors} Reveals hidden doors within large cone.
\spellhead{Disguise Self} Changes your appearance.
\spellheadc{Dispel Magic}{Lesser} Cancels magical spells and effects on a single target.
\spellhead{Expeditious Retreat} Your speed doubles.
\spellhead{Grease} Makes 10 ft. square or one object slippery.
\spellhead{Light} Object shines like a torch.
\spellhead{Message} Whispered conversation at distance.
\spellhead{Mage Hand} 5-pound telekinesis.
\spellhead{Sleep} Tire a creature, possibly putting it to sleep.
\spellhead{Summon Monster I} Calls extraplanar creature to fight for you.
\spellhead{Ventriloquism} Throws voice for medium duration.
\end{spelllist}

\subsection{2nd-level Bard Spells}
\begin{spelllist}
\spellhead{Aversion} Subject has aversion you specify.
\spellhead{Blur} Attacks miss subject 20\% of the time.
\spellhead{Calm Emotions} Calms creatures, negating emotion effects.
\spellhead{Charm Person} Makes one person your friend. 
\spellhead{Cure Moderate Wounds} Cures 4d8 damage.
\spellhead{Darkness} Object sheds supernatural shadow in medium radius.
\spellhead{Hideous Laughter} Subject loses actions for short duration.
\spellhead{Hold Person} Paralyzes one humanoid for short duration.
\spellhead{Hypnotic Pattern} Fascinates creatures in small radius.
\spellhead{See Invisibility} Reveals invisible creatures or objects.
\spellhead{Shatter} Sonic vibration damages objects or crystalline creatures in small radius.
\spellhead{Silence} Negates sound in medium radius.
\spellhead{Silent Image} Creates minor illusion of your design.
\spellhead{Slow} Subject takes only one action/round.
\spellhead{Sound Burst} Deals 2d6 sonic damage in small radius; may stun.
\spellhead{Summon Monster II} Calls extraplanar creature to fight for you.
\spellhead{Totemic Mind} Subject gains \plus2 to Int, Wis, or Cha for medium duration.
\end{spelllist}

\subsection{3rd-level Bard Spells}
\begin{spelllist}
\spellhead{Confusion} Subject acts randomly for short duration.
\spellhead{Crushing Despair} Subjects in medium cone are demoralized.
\spellhead{Cure Serious Wounds} Cures 6d8 damage.
\spellhead{Daze} Subject is bewildered and may lose an action.
\spellhead{Detect Thoughts} Allows ``listening" to surface thoughts.
\spellhead{Dispel Magic} Cancels magical spells and effects.
\spellhead{Glibness} You gain \plus30 bonus on Bluff checks, and your lies can escape magical discernment.
\spellhead{Good Hope} Subjects gain \plus2 on attack and damage, 10 \plus1/level temporary hp.
\spellhead{Heroism} Gives \plus2 on attacks and saves.
\spellhead{Invisibility} Subject is invisible for short duration or until it attacks.
\spellhead{Minor Image} As silent image, plus some sound.
\spellhead{Sculpt Sound} Creates new sounds or changes existing ones.
\spellhead{Suggestion} Compels subject to follow stated course of action.
\spellhead{Summon Monster III} Calls extraplanar creature to fight for you.
\spellhead{Tongues} Speak and understand any language.
\spellhead{Zone of Silence} Keeps eavesdroppers from overhearing conversations.
\end{spelllist}

\subsection{4th-level Bard Spells}
\begin{spelllist} 
\spellhead{Charm Monster} Makes monster believe it is your ally.
\spellhead{Cure Critical Wounds} Cures 8d10 damage.
\spellhead{Displacement} Attacks miss subject 50\% of the time.
\spellhead{Fear} Subjects within medium cone flee for short duration.
\spellhead{Hold Monster} As hold person, but any creature.
\spellhead{Major Image} As silent image, plus sound, smell and thermal effects.
\spellhead{Mind Fog} Fog imposes \minus10 penalty to Wis checks and Will saves.
\spellhead{Modify Memory} Changes 5 minutes of subject's memories.
\spellhead{Rainbow Pattern} Lights fascinate creatures in a small radius.
\spellhead{Shadow Conjuration} Mimics conjuring below 4th level, but only 25\% real.
\spellhead{Shout} Deafens all within medium cone and deals 1d6/level sonic damage.
\spellheadc{Sleep}{Mass} Tire multiple creatures, possibly putting them to sleep.
\spellhead{Summon Monster IV} Calls extraplanar creature to fight for you.
\spellhead{Totemic Mind} Subject gains \plus4 to Int, Wis, or Cha for medium duration, or \plus2 to each.%
\end{spelllist}

\subsection{5th-level Bard Spells}
\begin{spelllist}
\spellheadc{Cure Light Wounds}{Mass} Cures 5d8 damage for many creatures.
\spellhead{Dominate Person} Controls humanoid telepathically.
\spellheadc{Dispel Magic}{Greater} As dispel magic, but up to \plus20 on check.
\spellheadc{Heroism}{Greater} Gives immunity to fear and bonus on attacks, saves, and checks.
\spellhead{Mislead} Turns you invisible and creates illusory double.
\spellhead{Invisibility Sphere} Makes everyone within small radius invisible.
\spellhead{Persistent Image} As major image, but no concentration required.
\spellhead{Seeming} Changes appearance of group of creatures.
\spellhead{Shadow Evocation} Mimics evocation of lower than 5th level, but only 25\% real.
\spellhead{Summon Monster V} Calls extraplanar creature to fight for you.
\spellheadc{Totemic Mind}{Mass} As \spell{Totemic mind}, but affects multiple creatures.
\end{spelllist}

\subsection{6th-level Bard Spells}
\begin{spelllist}
\spellhead{Analyze Dweomer}\F Reveals magical aspects of subject.
\spellheadc{Charm Person}{Mass} As \spell{charm person}, but affects multiple creatures.
\spellheadc{Cure Moderate Wounds}{Mass} Cures 6d8 damage for many creatures.
\spellheadc{Hold Person}{Mass} As \spell{hold person}, but affects multiple creatures.
\spellhead{Irresistible Dance} Forces subject to dance.
\spellhead{Project Image} Illusory double can talk and cast spells.
\spellheadc{Shout}{Greater} Yell deals 6d6 damage in large cone; stuns creatures, damages objects.
\spellhead{Song of Discord} Forces targets in small spread to attack each other.
\spellheadc{Suggestion}{Mass} As \spell{suggestion}, but shorter and affects multiple creatures.
\spellhead{Summon Monster VI} Calls extraplanar creature to fight for you.
\end{spelllist}

\section{Cleric Spells}

\subsection{1st-level Cleric Spells}
\begin{spelllist}
\spellhead{Bane} Nearby enemies are shaken.
\spellhead{Bless} Nearby allies gain \plus2 to attack.
\spellhead{Cause Fear} One creature flees for short duration.
\spellhead{Command} One subject obeys selected command for 1 round.
\spellhead{Cure Light Wounds} Cures 2d8 damage.
\spellhead{Delay Poison} Stops poison from harming subject for extended duration.
\spellhead{Detect Chaos/Evil/Good/Law} Reveals creatures, spells, or objects of selected alignment.
\spellhead{Detect Poison} Detects poison in large cone.
\spellhead{Detect Undead} Reveals undead within large cone.
\spellheadc{Dispel Magic}{Lesser} Cancels magical spells and effects on a single target.
\spellhead{Entropic Shield} Ranged attacks against you have 20\% miss chance.
\spellhead{Magic Weapon} Weapon becomes \plus2.
\spellhead{Magic Vestment} Armor or shield becomes \plus2.
\spellhead{Inflict Light Wounds} Touch deals 2d8 damage.
\spellhead{Light} Object shines like a torch.
\spellhead{Obscuring Mist} A Fog surrounds you.
\spellhead{Protection from Chaos/Evil/Good/Law} \plus2 to saves and counter opposing mind control.
\spellhead{Sanctuary} Opponents can't attack you, and you can't attack.
\spellhead{Shield of Faith} Aura grants \plus2 or higher deflection bonus.
\spellhead{Summon Monster I} Calls extraplanar creature to fight for you.
\end{spelllist}

\subsection{2nd-level Cleric Spells}
\begin{spelllist}
\spellhead{Align Weapon} Weapon becomes \plus2 and good, evil, lawful, or chaotic.
\spellhead{Comprehend Languages} You understand all spoken and written languages.
\spellhead{Cure Moderate Wounds} Cures 4d8 damage.
\spellhead{Daylight} Large radius of bright light.
\spellhead{Hold Person} Paralyzes one humanoid for short duration.
\spellhead{Inflict Moderate Wounds} Touch attack, 4d8 damage.
\spellhead{Locate Object} Senses direction toward object (specific or type).
\spellhead{Divine Favor} You gain \plus2 on attack and damage rolls.
\spellhead{Resist Energy} Ignores 10 (or more) points of damage/attack from specified energy type.
\spellhead{Share Pain} Willing subject takes half of your damage.
\spellhead{Shield Other}\F You take half of subject's damage.
\spellhead{Silence} Negates sound in medium radius.
\spellhead{Spiritual Weapon} Magic weapon attacks on its own.
\spellhead{Summon Monster II} Calls extraplanar creature to fight for you.
\spellhead{Totemic Mind} Subject gains \plus2 to Int, Wis, or Cha for medium duration.
\spellhead{Totemic Power} Subject gains \plus2 to Str, Dex, or Con for medium duration.
\spellhead{Zone of Truth} Subjects within medium emanation cannot lie.
\end{spelllist}

\subsection{3rd-level Cleric Spells}
\begin{spelllist}
\spellhead{Aid} Ally gains \plus2 attack bonus, 10 \plus1/level temporary hp.
\spellhead{Bestow Curse} \minus6 to an attribute score; \minus4 on attack rolls, saves, and checks; or 25\% chance of losing each action.
\spellhead{Blindness/Deafness} Makes subject blinded or deafened.
\spellhead{Contagion} Infects subject with chosen disease.
\spellhead{Cure Serious Wounds} Cures 6d8 damage.
\spellhead{Dimensional Anchor} Bars extradimensional movement.
\spellhead{Discern Lies} Reveals deliberate falsehoods.
\spellhead{Dispel Magic} Cancels spells and magical effects.
\spellhead{Inflict Serious Wounds} Touch attack, 6d8 damage.
\spellhead{Invisibility Purge} Dispels invisibility within large radius.
\spellhead{Protection from Energy} Absorb 10 points/level of damage from one kind of energy.
\spellhead{Searing Light} Ray deals 6d6 damage and blinds, more against undead.
\spellhead{Share Pain, Forced} Unwilling subject takes half of your damage.
\spellhead{Summon Monster III} Calls extraplanar creature to fight for you.
\end{spelllist}

\subsection{4th-level Cleric Spells}
\begin{spelllist}
\spellhead{Air Walk} Subject treads on air as if solid (climb at 45-degree angle).
\spellhead{Cure Critical Wounds} Cures 8d8 damage.
\spellhead{Death Ward} Grants immunity to death spells and negative energy effects.
\spellhead{Dismissal} Forces a creature to return to native plane.
\spellhead{Divine Power} You gain attack and damage bonus, \plus4 Str, and 2 hp/level.
\spellhead{Freedom} Subject moves normally despite impediments.
\spellhead{Inflict Critical Wounds} Touch attack, 8d8 damage.
\spellhead{Invest Magic} All of subject's equipment is magical.
\spellhead{Poison} Touch deals 1d10 Con damage, repeats in 1 min.
\spellheadc{Resist Energy}{Greater} As \spell{resist energy}, but all energy types.
\spellhead{Spell Immunity} Subject is immune to one spell per four levels.
\spellhead{Summon Monster IV} Calls extraplanar creature to fight for you.
\spellhead{Tongues} Speak and understand any language.
\end{spelllist}

\subsection{5th-level Cleric Spells}
\begin{spelllist}
\spellheadc{Command}{Mass} As \spell{command}, but affects multiple creatures.
\spellheadc{Cure Light Wounds}{Mass} Cures 5d8 damage for many creatures.
\spellhead{Disrupting Weapon} Melee weapon destroys undead.
\spellhead{Flame Strike} Smite foes in small radius with divine fire (8d8 damage).
\spellheadc{Inflict Light Wounds}{Mass} Deals 5d8 damage to many creatures.
\spellhead{Magic Circle against Chaos/Evil/Good/Law} Protect allies in a \areamed radius and hedge out opposing outsiders.
\spellhead{Righteous Might} Your size increases, and you gain combat bonuses.
\spellhead{Revivify} Restores recently dead body to life.
\spellhead{Slay Living} Touch attack kills subject.
\spellhead{Summon Monster V} Calls extraplanar creature to fight for you.
\end{spelllist}

\subsection{6th-level Cleric Spells}
\begin{spelllist}
\spellhead{Banishment} Banishes extraplanar creatures in medium radius around you.
\spellhead{Blade Barrier} Wall of blades deals 6d8 damage.
\spellheadc{Cure Moderate Wounds}{Mass} Cures 6d8 damage for many creatures.
\spellheadc{Dispel Magic}{Greater} As dispel magic, but up to \plus20 on check.
\spellhead{Harm} Deals 12d8 damage to touched target.
\spellhead{Heal} Cures 12d8 damage, all diseases and mental conditions.
\spellheadc{Inflict Moderate Wounds}{Mass} Deals 6d8 damage to many creatures.
\spellheadc{Protection from Energy}{Greater}
\spellhead{Summon Monster VI} Calls extraplanar creature to fight for you.
\spellheadc{Totemic Mind}{Mass} As \spell{Totemic mind}, but affects multiple creatures.
\spellheadc{Totemic Power}{Mass} As \spell{Totemic power}, but affects multiple creatures.
\spellhead{True Seeing}\M Lets you see all things as they really are.
\spellhead{Undeath to Death}\M Destroys 2 HV/level undead.
\spellhead{Word of Recall} Teleports you back to designated place.
\end{spelllist}

\subsection{7th-level Cleric Spells}
\begin{spelllist}
\spellhead{Antilife Shell} Small, immobile emanation hedges out living creatures.
\spellhead{Blasphemy} Kills, paralyzes, weakens, or dazes nonevil subjects in large radius.
\spellheadc{Cure Serious Wounds}{Mass} Cures 7d8 damage for many creatures.
\spellhead{Destruction}\F Stuns and deals 14d8 damage to subject, destroys remains.
\spellhead{Dictum} Kills, paralyzes, slows, or deafens nonlawful subjects in 60 ft. cube.
\spellheadc{Hold Person}{Mass} As \spell{hold person}, but affects multiple creatures.
\spellhead{Holy Word} Kills, paralyzes, blinds, or deafens nongood subjects in large radius.
\spellheadc{Inflict Serious Wounds}{Mass} Deals 7d8 damage to many creatures.
\spellhead{Summon Monster VII} Calls extraplanar creature to fight for you.
\spellhead{Word of Chaos} Kills, confuses, stuns, or deafens nonchaotic subjects in large radius.
\end{spelllist}

\subsection{8th-level Cleric Spells}
\begin{spelllist}
\spellhead{Antimagic Field} Negates magic in small emanation from you.
\spellhead{Cloak of Chaos}\F \plus4 to AC, \plus4 resistance, and SR 10 against lawful spells.
\spellheadc{Cure Critical Wounds}{Mass} Cures 8d8 damage for many creatures.
\spellheadc{Death Ward}{Mass} As \spell{death ward}, but affects multiple creatures.
\spellhead{Earthquake} Intense tremor shakes large radius.
\spellhead{Energy Drain} Subject gains 2d4 negative levels.
\spellheadc{Freedom}{Mass} As \spell{freedom}, but affects multiple creatures.
\spellhead{Holy Aura}\F \plus4 to AC, \plus4 resistance, and SR 10 against evil spells.
\spellheadc{Inflict Critical Wounds}{Mass} Deals 8d8 damage to many creatures.
\spellhead{Regenerate} Touched creature heals 10 hit points/round.
\spellhead{Shield of Law}\F \plus4 to AC, \plus4 resistance, and SR 10 against chaotic spells.
\spellheadc{Spell Immunity}{Greater} As spell immunity, but up to 8th-level spells.
\spellhead{Summon Monster VIII} Calls extraplanar creature to fight for you.
\spellhead{Unholy Aura}\F \plus4 to AC, \plus4 resistance, and SR 10 against good spells.
\end{spelllist}

\subsection{9th-level Cleric Spells}
\begin{spelllist}
\spellheadc{Heal}{Mass} As heal, but heals 9d8 for many subjects.
\spellhead{Implosion} Kills one creature/round.
\spellhead{Miracle}\M Requests a deity's intercession.
\spellhead{Storm of Vengeance} Storm rains acid, lightning, and hail.
\spellhead{Summon Monster IX} Calls extraplanar creature to fight for you.
\end{spelllist}

\section{Cleric Domains}  

\begin{comment}
\subsection{Example Domain}
\parhead{Domain Power}
\parhead{Channelled Domain Power}
\parhead{Greater Domain Power}
\parhead{Greater Channelled Domain Power}
\parhead{Domain Mastery}
\begin{spelllist}
\spellhead[1]{}
\spellhead[1]{}
\spellhead[2]{}
\spellhead[2]{}
\spellhead[3]{}
\spellhead[3]{}
\spellhead[4]{}
\spellhead[4]{}
\spellhead[5]{}
\spellhead[5]{}
\spellhead[6]{}
\spellhead[6]{}
\spellhead[7]{}
\spellhead[7]{}
\spellhead[8]{}
\spellhead[8]{}
\spellhead[9]{}
\spellhead[9]{}
\end{spelllist}
\end{comment}

\subsection{Air Domain}

\subsubsection{Air Domain Spells}
\begin{spelllist}
\spellhead[1]{Feather Fall} Objects or creatures fall slowly.
\spellhead[1]{Gust of Wind} Blows away or knocks down smaller creatures in large line.
\spellhead[2]{Windstrike} Wind bludgeons target for 4d6 damage.
\spellhead[2]{Zephyr Blade} Melee weapon can strike from a short distance for half damage.
\spellhead[3]{Call Lightning} Calls down lightning bolts (3d8 per bolt) from sky.
\spellhead[3]{Gaseous Form} Subject becomes insubstantial and can fly slowly.
\spellhead[4]{Air Walk} Subject treads on air as if solid (climb at 45-degree angle).
\spellhead[4]{Summon Monster V*} Calls creature to fight.
\spellhead[5]{}
\spellheadc[5]{Windstrike}{Greater} Wind moves and bludgeons target for 10d6 damage.
\spellhead[6]{Chain Lightning} 10d6 damage; 1 secondary bolt/level each dealing half damage.
\spellheadc[6]{Zephyr Blade}{Greater} Melee weapon can strike from a distance for full damage.
\spellheadc[7]{Call Lightning}{Greater} As \spell{call lightning}, but 5d8 damage per bolt.
\spellhead[7]{Summon Monster VIII*} Calls creature to fight.
\spellhead[8]{Reverse Gravity} Objects and creatures fall upward.
\spellhead[9]{Elemental Swarm*} Summons multiple elementals.
\spellhead[9]{Storm of Vengeance} Storm rains acid, lightning, and hail.
\end{spelllist}
*Summons air elementals only

\subsection{Chaos Domain}

\subsubsection{Chaos Domain Spells}
\begin{spelllist}
\spellhead[1]{Entropic Shield} 
\spellhead[1]{Protection from Law} \plus2 to saves and counter opposing mind control.
\spellhead[2]{Align Weapon} Weapon becomes \plus2 and chaotic.
\spellhead[2]{Hideous Laughter} Subject loses actions for short duration.
\spellhead[3]{Confusion} Subject acts randomly for short duration.
\spellhead[3]{Summon Monster III*} Calls extraplanar creature to fight for you.
\spellhead[4]{Chaos Hammer} Confuses and deals 3d8 damage to lawful creatures.
\spellhead[5]{Animate Objects} Objects attack your foes.
\spellhead[5]{Magic Circle against Law} Protect allies in a \areamed radius.
\spellhead[6]{Insanity} Subject suffers continuous \spell{confusion}
\spellhead[6]{Summon Monster VI*} Calls extraplanar creature to fight for you.
\spellhead[7]{Prismatic Spray} Rays hit subjects with a variety of effects.
\spellhead[7]{Word of Chaos} Kills, confuses, stuns, or deafens nonchaotic subjects in large radius.
\spellhead[8]{Cloak of Chaos}\F \plus4 to AC, \plus4 resistance, SR 10 against lawful spells.
\spellhead[8]{Prismatic Wall} Wall's colors have array of effects.
\spellhead[9]{Irresistible Dance} Forces subject to dance.
\spellhead[9]{Summon Monster IX*} Calls extraplanar creature to fight for you.
\end{spelllist}
*Summon chaotic creatures only

\subsection{Death Domain}
\subsubsection{Death Domain Spells}
\begin{spelllist}
\spellhead[1]{Ray of Enfeeblement} Ray inflicts 1d6 \plus1 per two levels Str penalty.
\spellhead[1]{Ray of Clumsiness} Ray inflicts 1d6 \plus1 per two levels Dex penalty.
\spellhead[2]{Crush Life} Target takes 4d6 unavoidably.
\spellhead[2]{Death Knell} Kill dying creature and gain temporary hp, more for killing powerful creatures.
\spellhead[3]{Contagion} Infects subject with chosen disease.
\spellhead[3]{Vampiric Touch} Touch deals 6d8 damage; caster gains half damage as temporary hp.
\spellhead[4]{Death Ward} Grants immunity to death spells and negative energy effects.
\spellhead[4]{Enervation} Subject gains 1d4 negative levels.
\spellheadc[5]{Crush Life}{Greater} Target takes 10d6 damage unavoidably and gains a negative level.
\spellhead[5]{Slay Living} Touch attack kills subject.
\spellhead[6]{Circle of Death}\M Kills 2/level HV of creatures.
\spellhead[6]{Harm} Deals 12d8 damage to touched target.
\spellhead[7]{Finger of Death} Kills one subject.
\spellhead[7]{} 
\spellheadc[8]{Death Ward}{Mass} As \spell{death ward}, but affects multiple creatures.
\spellhead[8]{Energy Drain} Subject gains 2d4 negative levels.
\spellhead[9]{Power Word Kill} Kills bloodied creature.
\spellhead[9]{Wail of the Banshee} Kills one creature/level.
\end{spelllist}

\subsection{Destruction Domain}

\subsubsection{Destruction Domain Spells}
\begin{spelllist}
\spellhead[1]{Burning Hands} 1d8 fire damage in small cone.
\spellhead[1]{Shocking Grasp} Touch delivers 2d6 electricity damage.
\spellhead[2]{Shatter} Sonic vibration damages objects or crystalline creatures in small radius.
\spellhead[2]{Warp Wood} Bends wood (shaft, handle, door, plank).
\spellhead[3]{Contagion} Infects subject with chosen disease.
\spellhead[3]{Lightning Bolt} Large line of electricity deals 3d8 damage.
\spellhead[4]{Shout} Deafens all within medium cone and deals 4d6 sonic damage.
\spellhead[4]{}
\spellhead[5]{Flame Strike} Smite foes in small radius with divine fire (5d6 damage).
\spellhead[5]{}
\spellhead[6]{Disintegrate} Makes one creature or object vanish.
\spellhead[7]{Destruction}\F Stuns and deals 14d8 damage to subject, destroys remains.
\spellheadc[7]{Shout}{Greater} Yell deals 6d6 sonic damage in large cone; stuns creatures, damages objects.
\spellhead[8]{Earthquake} Intense tremor shakes 5 ft./level radius.
\spellhead[8]{Fire Storm} Deals 8d6 fire damage to enemies in large spread.
\spellhead[9]{Implosion} Kills one creature/round.
\spellhead[9]{Meteor Swarm} Four exploding spheres each deal 3d8 fire damage.
\end{spelllist}

\begin{comment}
\subsection{Dragon Domain}
\parhead{Domain Power} Gain a bite attack or two claw attacks.
\parhead{Channelled Domain Power} You can rebuke (but not command) dragons as an evil cleric rebukes undead.
\parhead{Greater Domain Power} You gain low-light vision and darkvision out to 60 feet. If you already had low-light vision, your vision range doubles. If you already had darkvision, its range increases by 60 feet.
\parhead{Greater Channelled Domain Power} You gain a frightful presence for 1 minute, as the dragon ability. The save DC is 10 \add 1/2 turning level \add Charisma.
\parhead{Domain Mastery}

\begin{spelllist}
\spellheadc[1]{Burning Hands}\fn{1} 1d8 fire damage.
\spellhead[1]{Command} One subject obeys selected command for 1 round.
\spellhead[2]{Comprehend Languages} You understand all spoken and written languages.
\spellhead[2]{Locate Object} Senses direction toward object (specific or type).
\spellheadc[3]{Lightning Bolt}\fn{1} Large line of electricity deals 3d8 damage.
\spellhead[3]{Tongues} Speak and understand any language.
\spellhead[4]{Fly} Subject flies at speed of 60 ft.
\spellhead[4]{Suggestion} Compels subject to follow stated course of action.
\spellheadc[5]{Command}{Greater} As \spell{command}, but affects multiple creatures.
\spellheadc[5]{Cone of Cold}\fn{1} 5d8 cold damage and brief slow in large cone.
\spellhead[6]{Locate Entity} Locates creatures or objects within 1 mile.
\spellhead[6]{True Seeing}\M Lets you see all things as they really are.
\spellhead[7]{Overland Flight} You fly at a speed of 40 ft. and can hustle over long distances.
\spellhead[7]{Power Word Blind} Blinds bloodied creature.
\spellhead[8]{Power Word Stun} Stuns bloodied creature.
\spellheadc[8]{Suggestion}{Mass} As \spell{suggestion}, plus one subject/level.
\spellhead[9]{Power Word Kill} Kills bloodied creature.
\spellhead[9]{}
\end{spelllist}
1 Used as a breath attack, requiring no somatic components.
\end{comment}

\subsection{Earth Domain}

\subsubsection{Earth Domain Spells}
\begin{spelllist}
\spellhead[1]{Earth's Pull} Subject feels gravity more strongly.
\spellhead[1]{Tremorsense} ``See'' through the ground by concentrating.
\spellhead[2]{}
\spellhead[2]{Soften Earth and Stone} Turns stone to clay or dirt to sand or mud.
\spellhead[3]{Meld into Stone} You and your gear merge with stone.
\spellhead[3]{Shape Stone} Sculpts stone into any shape.
\spellhead[4]{Spike Stones} Creatures in area take 1d8 damage per 5 feet moved, may be slowed.
\spellhead[4]{Summon Monster V*} Calls extraplanar creature to fight for you.
\spellhead[5]{Stoneskin} \plus5 AC, ignore 10 damage per round.
\spellhead[5]{Wall of Stone} Creates a stone wall that can be shaped.
\spellhead[6]{Transmute Flesh and Stone} Turns subject creature into statue, or restores petrified creature.
\spellhead[7]{Earthquake} Intense tremor shakes large radius.
\spellhead[7]{Summon Monster VIII*} Calls extraplanar creature to fight for you.
\spellhead[8]{Iron Body} Your body becomes living iron.
\spellhead[8]{}
\spellhead[9]{Elemental Swarm*} Summons multiple elementals.
\spellhead[9]{Imprisonment} Entombs subject beneath the earth.
\end{spelllist}
*Summons earth elementals only.

\subsection{Evil Domain}

\subsubsection{Evil Domain Spells}
\begin{spelllist}
\spellhead[1]{Bane} Nearby enemies are shaken.
\spellhead[1]{Protection from Good} \plus2 to saves and counter opposing mind control.
\spellhead[2]{Align Weapon} Weapon becomes \plus2 and evil.
\spellhead[2]{Death Knell} Kill dying creature and gain temporary hp, more for killing powerful creatures.
\spellhead[3]{Share Pain, Forced} Unwilling subject takes half of your damage.
\spellhead[4]{Summon Monster III*} Calls extraplanar creature to fight for you.
\spellhead[4]{Enervation} Subject gains 1d4 negative levels.
\spellhead[4]{Unholy Blight} Sickens and deals 3d8 damage to good creatures.
\spellhead[5]{}
\spellhead[5]{Magic Circle against Good} Protect allies in a \areamed radius and hedge out opposing outsiders.
\spellhead[6]{Harm} Deals 12d10 damage to touched target.
\spellhead[6]{Summon Monster VI*} Calls extraplanar creature to fight for you.
\spellhead[7]{Blasphemy} Kills, paralyzes, weakens, or dazes nonevil subjects in large radius.
\spellhead[7]{}
\spellhead[8]{Energy Drain} Subject gains 2d4 negative levels.
\spellhead[8]{Unholy Aura}\F \plus4 to AC, \plus4 resistance, SR 10 against good spells.
\spellhead[9]{Assimilate} Incorporate creature into your own body.
\spellhead[9]{Summon Monster IX*} Calls extraplanar creature to fight for you.
\end{spelllist}
*Summons evil creatures only

\subsection{Fire Domain}

\subsubsection{Fire Domain Spells}
\begin{spelllist}
\spellhead[1]{Burning Hands} 1d6 fire damage in medium cone.
\spellhead[1]{}
\spellhead[2]{Flame Weapon} Flaming weapon deals 1d8 \add casting attribute value damage.
\spellhead[2]{Scorching Ray} Ray (or rays) deal 4d6 fire damage.
\spellhead[3]{}
\spellhead[3]{Fireball} small radius of fire deals 3d6 damage.
\spellhead[4]{Fire Shield} Creatures attacking you take damage; you're protected from heat or cold.
\spellhead[4]{Summon Monster V*} Calls extraplanar creature to fight for you.
\spellhead[5]{Flame Strike} Smite foes in small radius with divine fire (1d6/level damage).
\spellhead[5]{Wall of Fire} Passing through wall deals 5d6 fire damage.
\spellhead[6]{Delayed Blast Fireball} 7d6 fire damage; you can postpone blast for 5 rounds.
\spellhead[6]{Fire Seeds} Acorns and berries become grenades and bombs.
\spellhead[7]{}
\spellhead[7]{Summon Monster VIII*} Calls extraplanar creature to fight for you.
\spellhead[8]{Fire Storm} Deals 8d6 fire damage to enemies in large spread.
\spellhead[9]{Elemental Swarm**} Summons multiple elementals.
\spellhead[9]{Meteor Swarm} Four exploding spheres each deal 3d8 fire damage.
\end{spelllist}
*Summons fire elementals only.

\subsection{Good Domain}
\subsubsection{Good Domain Spells}
\begin{spelllist}
\spellhead[1]{Bless} Allies gain d6 \plus1/level temporary hp.
\spellhead[1]{Protection from Evil} \plus2 to saves and counter opposing mind control.
\spellhead[2]{Align Weapon} Weapon becomes \plus2 and good.
\spellhead[2]{Shield Other}\F You take half of subject's damage.
\spellhead[3]{Aid} Ally gains \plus2 attack bonus, 10 \plus1/level temporary hp.
\spellhead[3]{Summon Monster III*} Calls extraplanar creature to fight for you.
\spellhead[4]{Death Ward} Grants immunity to death spells and negative energy effects.
\spellhead[4]{Holy Smite} Blinds and deals 4d6 damage to evil creatures.
\spellhead[5]{Magic Circle against Evil} Protect allies in a \areamed radius and hedge out opposing outsiders.
\spellhead[5]{Righteous Might} Your size increases, and you gain combat bonuses.
\spellhead[6]{Heal} Cures 12d8 damage, all diseases and mental conditions.
\spellhead[6]{Summon Monster VI*} Calls extraplanar creature to fight for you.
\spellhead[7]{Holy Word}\F Kills, paralyzes, slows, or deafens nongood subjects in large radius.
\spellhead[7]{}
\spellheadc[8]{Death Ward}{Mass} As \spell{death ward}, but affects multiple creatures.
\spellhead[8]{Holy Aura} \plus4 to AC, \plus4 resistance, and SR 10 against evil spells.
\spellheadc[9]{Heal}{Mass} As heal, but heals 9d8 for many subjects.
\spellhead[9]{Summon Monster IX*} Calls extraplanar creature to fight for you.
\end{spelllist}
*Summons good creatures only.

\subsection{Knowledge Domain}
\subsubsection{Knowledge Domain Spells}
\begin{spelllist}
\spellhead[1]{Comprehend Languages}
\spellhead[1]{}
\spellhead[2]{Locate Object}
\spellhead[2]{}
\spellhead[3]{Detect Thoughts}
\spellhead[3]{Tongues}
\spellhead[4]{Locate Creature}
\spellheadc[4]{Moment of Prescience}{Lesser}
\spellhead[5]{True Seeing}
\spellhead[6]{Analyze Dweomer} 
\spellhead[6]{Locate Entity}
\spellheadc[7]{Arcane Sight}{Greater}
\spellhead[7]{Moment of Prescience}
\spellheadc[8]{Detect Thoughts}{Greater}
\spellhead[8]{}
\spellhead[9]{Foresight}
\spellhead[9]{Revelation}
\end{spelllist}

\subsection{Law Domain}

\subsubsection{Law Domain Spells}
\begin{spelllist}
\spellhead[1]{Command} One subject obeys selected command for 1 round.
\spellhead[1]{Protection from Chaos} \plus2 to saves and counter opposing mind control.
\spellhead[2]{Align Weapon} Weapon becomes \plus2 and lawful.
\spellhead[2]{Zone of Truth} Subjects within range cannot lie.
\spellhead[3]{Discern Lies} Reveals deliberate falsehoods.
\spellhead[3]{Summon Monster III*} Calls extraplanar creature to fight for you.
\spellhead[4]{Hold Monster} As \spell{hold person}, but any creature.
\spellhead[4]{Order's Wrath} Dazes and deals 3d8 damage to chaotic creatures.
\spellheadc[5]{Command}{Greater} As \spell{command}, but affects multiple creatures.
\spellhead[5]{Magic Circle against Chaos} Protect allies in a \areamed radius and hedge out opposing outsiders.
\spellhead[6]{True Seeing}\M Lets you see all things as they really are.
\spellhead[6]{Summon Monster VI*} Calls extraplanar creature to fight for you.
\spellhead[7]{Dictum} Kills, paralyzes, slows, or deafens nonlawful subjects in 60 ft cube.
\spellhead[7]{Power Word Stun} Stuns bloodied creature.
\spellhead[8]{}
\spellhead[8]{Shield of Law}\F \plus4 to AC, \plus4 resistance, and SR 10 against chaotic spells.
\spellhead[9]{Imprisonment} Entombs subject beneath the earth.
\spellhead[9]{Summon Monster IX*} Calls extraplanar creature to fight for you.
\end{spelllist}
*Summons lawful creatures only.

\subsection{Life Domain}

\subsubsection{Life Domain Spells}
\begin{spelllist}
\spellhead[1]{Bless} Allies in large radius gain 5 \plus1/level temporary hp (max 10).
\spellhead[1]{Cure Light Wounds} Cures 2d8 damage.
\spellhead[2]{}
\spellhead[2]{Cure Moderate Wounds} Cures 4d10 damage.
\spellhead[3]{Cure Serious Wounds} Cures 6d10 damage.
\spellhead[3]{Link Vitality} Two targets share each others' damage and healing.
\spellhead[4]{Death Ward} Grants immunity to death spells and negative energy effects.
\spellhead[4]{Cure Critical Wounds} Cures 8d10 damage.
\spellheadc[5]{Cure Light Wounds}{Mass} Cures 5d8 damage for many creatures.
\spellhead[5]{}
\spellheadc[6]{Cure Moderate Wounds}{Mass} Cures 6d8 damage for many creatures.
\spellhead[6]{Heal} Cures 10 damage/level, all diseases and mental conditions.  
\spellheadc[7]{Cure Serious Wounds}{Mass} Cures 7d8 damage for many creatures.
\spellhead[7]{Regenerate} Touched creature heals 10 hit points/round.
\spellheadc[8]{Cure Critical Wounds}{Mass} Cures 8d8 damage for many creatures.
\spellheadc[8]{Death Ward}{Mass} As \spell{death ward}, but affects multiple creatures.
\spellheadc[9]{Heal}{Mass} As heal, but heals 9d8 for many subjects.
\spellhead[9]{Miracle} Requests a deity's intercession.
\end{spelllist}

\subsection{Magic Domain}

\subsubsection{Magic Domain Spells}
\begin{spelllist}
\spellheadc[1]{Dispel Magic}{Lesser} Cancels magical spells and effects on a single target.
\spellhead[1]{Magic Weapon} Weapon becomes \plus2.
\spellhead[2]{Ablative Shield}
\spellhead[2]{Spelltheft}{Lesser} As \spell{lesser dispel magic}, but you gain the effects of dispelled spells.
\spellhead[3]{Dimensional Anchor} Bars extradimensional movement.
\spellhead[3]{Dispel Magic} Cancels magical spells and effects.
\spellhead[4]{Ablative Fortress}
\spellhead[4]{Spell Immunity} Subject is immune to one spell per four levels.
\spellhead[5]{Spell Resistance} Subject gains SR 10.
\spellhead[5]{Spelltheft} As \spell{dispel magic}, but you gain the effects of dispelled spells.
\spellhead[6]{Analyze Dweomer}\F Reveals magical aspects of subject.
\spellheadc[6]{Dispel Magic}{Greater} As dispel magic, but up to \plus20 on check.
\spellhead[7]{Antimagic Field} Negates magic in small emanation from you.
\spellhead[7]{Spell Turning} Reflect 1d4\plus6 spell levels back at caster.
\spellhead[8]{Protection from Spells} Subject gains SR 10 and \plus5 to saves vs spells.
\spellheadc[8]{Spell Immunity}{Greater} As spell immunity, but up to 8th-level spells.
\spellhead[9]{Mage's Disjunction} Dispels all magic in medium radius.
\spellhead[9]{}
\end{spelllist}
\subsection{Nature Domain}

\subsubsection{Nature Domain Spells}
\begin{spelllist}
\spellhead[1]{Detect Animals or Plants} Detects animals or plants in a large cone.
\spellhead[1]{Entangle} Plants entangle everyone in small spread.
\spellhead[2]{Barkskin} Grants \plus2 bonus to AC and DR 2/adamantine.
\spellhead[2]{Reduce Animal} Shrinks one willing animal.
\spellheadc[3]{Magic Fang}{Greater} One natural weapon of subject gets \plus1/three levels on attack and damage (max \plus5).
\spellhead[3]{Summon Nature's Ally IV} Calls creature to fight.
\spellhead[4]{Giant Vermin} Turns centipedes, scorpions, or spiders into giant vermin.
\spellhead[4]{}
\spellhead[5]{Animate Plants} Plants animate and fight for you.
\spellhead[5]{Wall of Thorns} Thorns damage anyone who tries to pass.
\spellhead[6]{Animal Growth} Enlarges multiple animals.
\spellhead[6]{Summon Nature's Ally VI} Calls creature to fight.
\spellhead[7]{Antilife Shell} Small stationary emanation hedges out living creatures.
\spellhead[7]{Fire Seeds} Acorns and berries become grenades and bombs.
\spellhead[8]{Changestaff} Your staff becomes a treant on command.
\spellhead[8]{Summon Nature's Army} Calls one creature/level to fight.
\spellhead[9]{Stampede} Summon bison to stampede foes.
\spellhead[9]{Summon Nature's Ally IX} Calls creature to fight.
\end{spelllist}

\subsection{Protection Domain}
\subsubsection{Protection Domain Spells}
\begin{spelllist}
\spellhead[1]{Sanctuary} Opponents can't attack you, and you can't attack.
\spellhead[1]{Shield of Faith} Aura grants \plus2 or higher deflection bonus.
\spellhead[2]{Resist Energy} Ignores 10 (or more) points of damage/attack from specified energy type.
\spellhead[2]{Shield Other}\F You take half of subject's damage.
\spellhead[3]{Redirection} Redirect attacks on you to other targets.
\spellhead[3]{Protection from Energy} Absorb 10 points/level of damage from one kind of energy.
\spellhead[4]{Death Ward} Grants immunity to death spells and negative energy effects.
\spellhead[4]{Spell Immunity} Subject is immune to one spell per four levels.
\spellhead[5]{Spell Resistance} Subject gains SR 10.
\spellhead[5]{Stoneskin} \plus5 AC, ignore 10 damage per round.
\spellheadc[6]{Shield Other}{Greater} You take all of subject's damage.
\spellhead[6]{Repulsion} Creatures can't approach you.
\spellhead[7]{Energy Conversion} As \spell{greater resist energy}, plus you can fire rays of absorbed energy.
\spellhead[7]{Spell Turning} Reflect 1d4\plus6 spell levels back at caster.
\spellheadc[8]{Death Ward}{Mass} As \spell{death ward}, but affects multiple creatures.
\spellheadc[8]{Spell Immunity}{Greater} As spell immunity, but up to 8th-level spells.
\spellhead[9]{Foresight} ``Sixth sense" warns of impending danger.
\spellhead[9]{}
\end{spelllist}

\subsection{Strength Domain}

\subsubsection{Strength Domain Spells}
\begin{spelllist}
\spellhead[1]{Longstrider} Increases your speed.
\spellhead[1]{Jump} Subject gets bonus on Jump checks.
\spellhead[2]{Totemic Power} Subject gains \plus2 to Str, Dex, or Con for medium duration.
\spellhead[2]{}
\spellhead[3]{Enlarge Person} Humanoid creature doubles in size.
\spellhead[4]{Divine Power} You gain attack and damage bonus, \plus6 Str, and 2 hp/level.
\spellhead[4]{Shout} Deafens all within medium cone and deals 1d6/level sonic damage.
\spellhead[5]{Stoneskin} \plus5 AC, ignore 10 damage per round.
\spellhead[5]{Righteous Might} Your size increases, and you gain combat bonuses.
\spellhead[6]{Forceful Hand} Hand pushes creatures away.
\spellheadc[6]{Enlarge Person}{Mass} Enlarges several creatures.
\spellhead[7]{Grasping Hand} Large hand provides cover, pushes, or grapples.
\spellheadc[7]{Shout}{Greater} Yell deals 1d6/level sonic damage in large cone; stuns creatures, damages objects.
\spellhead[8]{Iron Body} Your body becomes living iron.
\spellhead[8]{Crushing Hand} Large hand provides cover, pushes, or crushes your foes.
\spellhead[9]{Clenched Fist} Large hand provides cover, pushes, or attacks your foes.
\spellhead[9]{} 
\end{spelllist}

\begin{comment}
\subsection{SUN DOMAIN}
\parhead{Domain Power} You are immune to being dazzled or blinded by bright light.
\parhead{Channelled Domain Power} You can perform an improved turning or rebuking against undead. Undead that are turned or rebuked also catch on fire. Until they break the turning or rebuking effect, the undead do not attempt to put out the flames.
\parhead{Greater Domain Power} You can continuously radiate a \spell{daylight} effect, as the spell. You can suppress or resume this ability as a swift action.
\parhead{Greater Channelled Domain Power} You perform a greater turning or rebuking against undead. The greater turning is like a normal turning except that the undead creatures that would be turned or rebuked are destroyed instead.
\parhead{Domain Mastery} 

\subsubsection{Sun Domain Spells}
\begin{spelllist}
\spellhead[1]{Daylight} Large radius of bright light.
\spellhead[2]{Flaming Sphere} Creates rolling ball of fire, 2d6 damage, lasts short duration.
\spellhead[2]{Blindness/Deafness \fn{1}}
\spellhead[3]{Plant Growth}
\spellhead[3]{Searing Light} Ray deals 1d8/ two levels, more against undead.
\spellhead[4]{Death Ward} Grants immunity to death spells and negative energy effects.
\spellhead[4]{Disrupting Weapon} Melee weapon destroys undead.
\spellhead[5]{Wall of Fire} Passing through wall deals 5d6 damage
\spellhead[5]{Flame Strike} Smite foes in small radius with divine fire (1d6/level damage).
\spellhead[6]{Fire Seeds} Acorns and berries become grenades and bombs.
\spellhead[6]{Sunbeam} Multiple beams blind and deal 5d6 damage in large line.
\spellhead[7]{Prismatic Spray} Rays hit subjects with variety of effects.
\spellhead[7]{Sunburst} Blinds all within large radius, deals 10d6 damage.
\spellhead[8]{Fire Storm} Deals 1d6/level fire damage in shapeable burst.
\spellhead[8]{Prismatic Wall} Wall's colors have array of effects.
\spellhead[9]{Meteor Swarm} Four exploding spheres each deal 6d6 fire damage.
\spellhead[9]{Prismatic Sphere} As prismatic wall, but surrounds on all sides.
\end{spelllist}
1. Blindness only
\end{comment}

\subsection{Travel Domain}

\subsubsection{Travel Domain Spells}
\begin{spelllist}
\spellhead[1]{Feather Fall} Objects or creatures fall slowly.
\spellhead[1]{Longstrider} Increases your speed.
\spellhead[2]{Resist Energy} Ignores 10 (or more) points of damage/attack from specified energy type.
\spellhead[2]{Spider Climb} Grants ability to walk on walls and ceilings.
\spellhead[3]{Dimension Slide} Teleports you very short distance.
\spellhead[3]{Gaseous Form} Subject becomes insubstantial and can fly slowly.
\spellhead[4]{Air Walk} Subject treads on air as if solid (climb at 45-degree angle).
\spellhead[4]{Freedom} Subject moves normally despite impediments.
\spellhead[5]{Dimension Door} Teleports you within Long range.
\spellhead[5]{Passwall} Creates passage through wood or stone wall.
\spellhead[6]{Repulsion} Creatures can't approach you.
\spellhead[6]{Word of Recall} Teleports you back to designated place.
\spellhead[7]{Ethereal Jaunt} You become ethereal for short duration.
\spellhead[7]{Phase Door} Creates an invisible passage through wood or stone.
\spellheadc[8]{Freedom}{Mass} As \spell{freedom}, but affects multiple creatures.
\spellheadc[8]{Dimension Door}{Mass} Teleports you and one ally/level within Long range.
\spellhead[9]{Etherealness} Travel to Ethereal Plane with companions.
\spellhead[9]{Gate} Connects two planes for travel.
\end{spelllist}

\subsection{Trickery Domain}
\subsubsection{Trickery Domain Spells}
\begin{spelllist}
\spellhead[1]{Create Sound} Figment sounds.
\spellhead[1]{Ventriloquism} Throws voice for medium duration.
\spellhead[2]{Darkness} Object sheds supernatural shadow in medium radius.
\spellhead[2]{Silence} Negates sound in medium radius.
\spellhead[3]{Confusion} Subject acts randomly for short duration.
\spellhead[3]{Invisibility} Subject invisible for short duration or until it attacks.
\spellhead[4]{}
\spellhead[4]{Phantasmal Killer} Fearsome illusion can frighten target to death.
\spellhead[5]{Mind Fog} Fog imposes \minus5 penalty to Wisdom.
\spellhead[5]{Phantom Maze} Subject acts as if affected by \spell{maze}.
\spellhead[6]{Mislead} Turns you invisible and creates illusory double.
\spellhead[6]{Seeming} Changes appearance of group of creatures.
\spellheadc[7]{Invisibility}{Greater} As \spell{invisibility}, but affects any within 100 ft.
\spellhead[7]{}
\spellheadc[8]{Confusion}{Mass} Subjects in small radius act randomly for short duration.
\spellhead[8]{Reverse Gravity} Objects and creatures fall upward.
\spellhead[9]{Maze} Traps subject in extradimensional maze.
\spellhead[9]{Weird} As \spell{phantasmal killer}, but affects multiple creatures
\end{spelllist}

\subsection{War Domain}
\subsubsection{War Domain Spells}
\begin{spelllist}
\spellhead[1]{Bane} Nearby enemies are shaken.
\spellhead[1]{Bless} Nearby allies gain \plus2 to attack.
\spellhead[2]{Divine Favor} You gain \plus2 on attack and damage rolls.
\spellhead[2]{Spiritual Weapon} Magical weapon attacks on its own.
\spellhead[3]{Aid} Ally gains \plus2 attack bonus, 10 \plus1/level temporary hp.
\spellhead[3]{}
\spellhead[4]{Invest Magic} All of subject's equipment is magical.
\spellhead[4]{Divine Power} You gain attack and damage bonus, \plus6 Str, and 2 hp/level.
\spellhead[5]{Flame Strike} Smite foes in small radius with divine fire (5d6 damage).
\spellhead[5]{Waves of Fatigue} Creatures in large cone become fatigued.
\spellhead[6]{Blade Barrier} Wall of blades deals 6d6 damage.
\spellhead[6]{}
\spellhead[7]{}
\spellhead[7]{}
\spellhead[8]{Fire Storm} Deals 8d6 fire damage to enemies in large spread.
\spellhead[8]{Waves of Exhaustion} Creatures in medium cone become exhausted.  
\spellhead[9]{}
\spellhead[9]{Storm of Vengeance} Storm rains acid, lightning, and hail.
\end{spelllist}

\subsection{Water Domain}
\subsubsection{Water Domain Spells}
\begin{spelllist}
\spellhead[1]{Obscuring Mist} A Fog surrounds you.
\spellhead[1]{Ray of Frost} Icy ray deals 2d8 cold damage and hinders movement.
\spellhead[2]{Fog Cloud} Fog obscures vision.
\spellhead[2]{}
\spellhead[3]{Control Water} Raises or lowers bodies of water.
\spellhead[3]{Water Walk} Subject treads on water as if solid.
\spellhead[4]{}
\spellhead[4]{Summon Nature's Ally IV*} Calls creature to fight.
\spellhead[5]{Ice Storm} Hail deals damage in small cylinder.
\spellhead[5]{Wall of Ice} Create plane of ice or hemisphere can trap creatures inside.
\spellhead[6]{Cone of Cold} 5d8 cold damage and brief slow in large cone.
\spellhead[6]{Solid Fog} Fog blocks vision and slows movement.
\spellhead[7]{Freezing Sphere} Freezes water or deals cold damage.
\spellhead[7]{Summon Nature's Ally VII*} Calls creature to fight.
\spellhead[8]{Horrid Wilting} Fatigues and deals 8d8 damage to living creatures.
\spellhead[8]{Polar Ray} Ranged touch attack slows and deals 16d8 cold damage.
\spellhead[9]{Elemental Swarm*} Summons multiple elementals.
\spellhead[9]{Storm of Vengeance} Storm rains acid, lightning, and hail.
\end{spelllist}
*Summons water elementals or aquatic creatures only.


\section{Druid Spells}

\subsection{1st-level Druid Spells}
\begin{spelllist}
\spellhead{Cure Light Wounds} Cures 2d8 damage.
\spellhead{Delay Poison} Stops poison from harming subject for extended duration.
\spellhead{Detect Animals or Plants} Detects animals or plants in large cone.
\spellhead{Detect Poison} Detects poison in large cone.
\spellhead{Detect Snares and Pits} Reveals natural or primitive traps in large cone.
\spellhead{Earth's Pull} Subject feels gravity more strongly.
\spellhead{Entangle} Plants entangle everyone in small spread.
\spellhead{Gust of Wind} Blows away or knocks down smaller creatures in large line.
\spellhead{Jump} Subject gets bonus on Jump checks.
\spellhead{Light} Object shines like a torch.
\spellhead{Longstrider} Your speed increases by 10 ft.
\spellhead{Magic Fang} One natural weapon of subject creature becomes \plus2 weapon.
\spellhead{Obscuring Mist} A Fog surrounds you.
\spellhead{Shillelagh} Cudgel or quarterstaff becomes \plus2 weapon.
\spellhead{Spike Growth} Creatures in area take 1d6 damage per 5 ft. moved, may be slowed.
\spellhead{Summon Nature's Ally I} Calls creature to fight.
\spellhead{Tremorsense} ``See'' through the ground by concentrating.
\end{spelllist}

\subsection{2nd-level Druid Spells}
\begin{spelllist}
\spellhead{Barkskin} Grants \plus2 bonus to AC and DR 2/adamantine.
\spellheadc{Cone of Cold}{Lesser} 2d6 cold damage and brief fatigue in medium cone.
\spellhead{Cure Moderate Wounds} Cures 4d10 damage.
\spellheadc{Dispel Magic}{Lesser} Cancels magical spells and effects on a single target.
\spellhead{Faerie Fire} Outlines subjects in small radius with light, canceling concealment, invisibility, and the like.
\spellhead{Flame Weapon} Flaming weapon deals 1d8 \add casting attribute value damage.
\spellhead{Heat Metal} Make metal so hot it damages those who touch it.
\spellhead{Resist Energy} Ignores 10 (or more) points of damage/attack from specified energy type.
\spellhead{Shape Wood} Rearranges wooden objects to suit you.
\spellhead{Soften Earth and Stone} Turns stone to clay or dirt to sand or mud.
\spellhead{Spider Climb} Grants ability to walk on walls and ceilings.
\spellhead{Summon Nature's Ally II} Calls creature to fight.
\spellhead{Summon Swarm} Summons swarm of bats, rats, or spiders.
\spellhead{Totemic Mind} Subject gains \plus2 to Int, Wis, or Cha for medium duration.
\spellhead{Totemic Power} Subject gains \plus2 to Str, Dex, or Con for medium duration.
\spellhead{Tree Shape} You look exactly like a tree for extended duration.
\spellhead{Windstrike} Wind bludgeons target for 4d6 damage.
\spellhead{Warp Wood} Bends wood (shaft, handle, door, plank).
\spellhead{Zephyr Blade} Melee weapon can strike from a short distance for half damage.
\end{spelllist}

\subsection{3rd-level Druid Spells}
\begin{spelllist}
\spellhead{Call Lightning} Calls down lightning bolts (3d8 per bolt) from sky.
\spellhead{Contagion} Infects subject with chosen disease.
\spellhead{Control Water} Raises or lowers bodies of water.
\spellhead{Cure Serious Wounds} Cures 6d10 damage.
\spellhead{Fog Cloud} Small cylinder of fog obscures vision.
\spellhead{Lightning Bolt} Large line of electricity deals 3d6 damage.
\spellheadc{Magic Fang}{Greater} All subject's natural weapons get \plus2 bonus.
\spellhead{Meld into Stone} You and your gear merge with stone.
\spellhead{Poison} Touch deals 1d10 Con damage, repeats in 1 min.
\spellhead{Protection from Energy} Absorb 10 points/level of damage from one kind of energy.
\spellhead{Shape Stone} Sculpts stone into any shape.
\spellhead{Snare} Creates a magic booby trap.
\spellhead{Summon Nature's Ally III} Calls creature to fight.
\spellhead{Water Walk} Subject treads on water as if solid.
\end{spelllist}

\subsection{4th-level Druid Spells}
\begin{spelllist}
\spellhead{Air Walk} Subject treads on air as if solid (climb at 45-degree angle).
\spellhead{Cure Critical Wounds} Cures 8d10 damage.
\spellhead{Dispel Magic} Cancels spells and magical effects.
\spellhead{Entangling Growth} New plants grow and entangle everyone in large spread.
\spellhead{Freedom} Subject moves normally despite impediments.
\spellhead{Giant Vermin} Turns centipedes, scorpions, or spiders into giant vermin.
\spellhead{Ice Storm} Hail deals damage in small cylinder.
\spellheadc{Resist Energy}{Greater} As \spell{resist energy}, but all energy types.
\spellhead{Spike Stones} Creatures in area take 1d8 damage per 5 feet moved, may be slowed.
\spellhead{Summon Nature's Ally IV} Calls creature to fight.
\spellhead{Wall of Ice} Create plane of ice or hemisphere can trap creatures inside.
\end{spelllist}

\subsection{5th-level Druid Spells}
\begin{spelllist}
\spellhead{Adapt Body} Your body automatically adapts to hostile environments.
\spellheadc{Animal Affinity}{Mass} As \spell{animal affinity}, but affects multiple creatures.
\spellhead{Animate Plants} Plants animate and fight for you.
\spellhead{Baleful Polymorph} Transforms subject into harmless animal.
\spellheadc{Call Lightning}{Greater} As call lightning, but 5d8 damage per bolt.
\spellhead{Cone of Cold} 5d6 cold damage and fatigue in medium cone.
\spellheadc{Cure Light Wounds}{Mass} Cures 5d8 damage for many creatures.
\spellhead{Stoneskin} \plus5 AC, ignore 10 damage per round.
\spellhead{Summon Nature's Ally V} Calls creature to fight.
\spellhead{Wall of Fire} Passing through wall deals 4d6 damage.
\spellhead{Wall of Stone} Creates a stone wall that can be shaped.
\spellhead{Wall of Thorns} Thorns damage anyone who tries to pass.
\spellheadc{Windstrike}{Greater} Wind moves and bludgeons target for 10d6 damage.
\end{spelllist}

\subsection{6th-level Druid Spells}
\begin{spelllist}
\spellhead{Animal Growth} Enlarges multiple animals.
\spellhead{Antilife Shell} Small radius emanation hedges out living creatures.
\spellheadc{Cure Moderate Wounds}{Mass} Cures 5d8 damage for many creatures.
\spellheadc{Dispel Magic}{Greater} As dispel magic, but up to \plus20 on check.
\spellhead{Fire Seeds} Acorns and berries become grenades and bombs.
%\spellhead{Liveoak} Oak becomes treant guardian.
\spellheadc{Protection from Energy}{Greater}
\spellhead{Solid Fog} Fog blocks vision and slows movement.
\spellhead{Summon Nature's Ally VI} Calls creature to fight.
\spellheadc{Totemic Mind}{Mass} As \spell{totemic mind}, but affects multiple creatures.
\spellheadc{Totemic Power}{Mass} As \spell{totemic power}, but affects multiple creatures.
\spellheadc{Zephyr Blade}{Greater} Melee weapon can strike from a distance for full damage.
\end{spelllist}

\subsection{7th-level Druid Spells}
\begin{spelllist}
\spellheadc{Call Lightning}{Greater} As \spell{call lightning}, but 5d8 damage per bolt.
\spellhead{Changestaff} Your staff becomes a treant on command.
\spellhead{Creeping Doom} Swarms of centipedes attack at your command.
\spellheadc{Cure Serious Wounds}{Mass} Cures 7d8 damage for many creatures.
\spellhead{Heal} Cures 12d8 damage, all diseases and mental conditions.
\spellhead{Oak Body} Your body becomes as hard as oak. 
\spellhead{Summon Nature's Ally VII} Calls creature to fight.
\spellhead{Sunbeam} Beam blinds and deals 7d6 damage.
\spellhead{Wind Walk} You and your allies turn vaporous and travel fast.
\end{spelllist}

\subsection{8th-level Druid Spells}
\begin{spelllist}
\spellheadc{Cure Critical Wounds}{Mass} Cures 8d8 damage for many creatures.
\spellhead{Earthquake} Intense tremor shakes large radius.
\spellhead{Fire Storm} Deals 8d6 fire damage to enemies in large spread.
\spellheadc{Freedom}{Mass} As \spell{freedom}, but affects multiple creatures.
\spellhead{Regenerate} Touched creature heals 10 hit points/round.
\spellhead{Repel Metal or Stone} Pushes away metal and stone in large line.
\spellhead{Sea of Fog} 500 ft. cylinder of fog obscures vision.
\spellhead{Summon Nature's Ally VIII} Calls creature to fight.
\spellhead{Summon Nature's Army} Calls one creature/level to fight.
\spellhead{Sunburst} Blinds all within large radius, deals 8d8 damage.
\end{spelllist}

\subsection{9th-level Druid Spells}
\begin{spelllist}
\spellhead{Elemental Swarm} Summons multiple elementals.
\spellhead{Stampede} Summon bison to stampede foes.
\spellhead{Storm of Vengeance} Storm rains acid, lightning, and hail.
\spellhead{Summon Nature's Ally IX} Calls creature to fight.
\end{spelllist}

\section{Paladin Spells}  

\subsection{1st-level Paladin Spells}
\begin{spelllist}
\spellhead{Bless} Nearby allies gain \plus2 to attack.
\spellhead{Bless Weapon} Weapon becomes \plus2 and good.
\spellhead{Cure Light Wounds} Cures 2d8 damage.
\spellhead{Delay Poison} Stops poison from harming subject for extended duration.
\spellhead{Detect Chaos} Reveals creatures, spells, or objects of chaotic alignment.
\spellhead{Magic Vestment} Armor or shield becomes \plus2.
\spellhead{Protection from Chaos/Evil} \plus2 to saves and counter opposing mind control.
\end{spelllist}

\subsection{2nd-level Paladin Spells}
\begin{spelllist}
\spellhead{Cure Moderate Wounds} Cures 4d10 damage.
\spellhead{Daylight} Large radius of bright light.
\spellhead{Divine Favor} You gain \plus2 on attack and damage rolls.
\spellhead{Resist Energy} Ignores 10 (or more) points of damage/attack from specified energy type.
\spellhead{Shield Other}\F You take half of subject's damage.
\spellhead{Zone of Truth} Subjects within small emanation cannot lie.
\end{spelllist}

\subsection{3rd-level Paladin Spells}
\begin{spelllist} 
\spellhead{Aid} Ally gains \plus2 attack bonus, 10 \plus1/level temporary hp.
\spellhead{Cure Serious Wounds} Cures 6d10 damage.
\spellhead{Discern Lies} Reveals deliberate falsehoods.
\spellhead{Heal Mount} As \spell{heal} on warhorse or other special mount.
\spellhead{Protection from Energy} Absorb 10 points/level of damage from one kind of energy.
\end{spelllist}

\subsection{4th-level Paladin Spells}
\begin{spelllist}
\spellhead{Cure Critical Wounds} Cures 8d10 damage.
\spellhead{Death Ward} Grants immunity to death spells and negative energy effects.
\spellhead{Divine Power} You gain attack and damage bonus, \plus4 Str, and 2 hp/level.
\spellhead{Holy Sword} Weapon becomes \plus5 and holy.
\spellhead{Magic Circle against Chaos/Evil} Protect allies in a \areamed radius and hedge out opposing outsiders.
\spellheadc{Resist Energy}{Greater} As \spell{resist energy}, but all energy types. 
\end{spelllist}

\section{Sorcerer/Wizard Spells}
\subsection{1st-level Sorcerer/wizard Spells}
\begin{swspelllist}
\spellheadc[Abjur]{Dispel Magic}{Lesser} Cancels magical spells and effects on a single target.
\spellhead{Mage Armor} Armor made of magical force gives \plus2 AC.
\spellhead{Protection from Chaos/Evil/Good/Law} \plus2 to saves and counter opposing mind control.
\spellhead{Shield} Invisible disc gives \plus2 to AC.
\spellheadrestricted{Resist Impact} Resist one attack against you.
\spellheadrestricted{Sanctuary} Opponents can't attack you, and you can't attack.

\spellhead[Conj]{Grease} Makes 10 ft. square or one object slippery.
\spellhead{Obscuring Mist} A Fog surrounds you.
\spellhead{Summon Monster I} Calls extraplanar creature to fight for you.
\spellheadrestricted{Retrieve} Teleport an unattended item you can see to your hand.
\spellheadrestricted{Summon Monster II} Calls extraplanar creature to fight for you.

\spellhead[Div]{Detect Poison} Detects poison in large cone.
\spellhead{Detect Secret Doors} Reveals hidden doors in large cone.
\spellhead{Detect Undead} Reveals undead within large cone.
\spellhead{Message} Whispered conversation at distance.
\spellheadrestricted{Farsight} Doubles range of vision.
\spellheadrestrictedc{Precognition}{Lesser} See into the future to gain attack bonus.

\spellhead[Ench]{Command} One subject obeys selected command for 1 round.
\spellhead{Sleep} Tire a creature, possibly putting it to sleep.
\spellheadrestricted{Attraction} Subject has an attraction you specify.
\spellheadrestricted{Cause Fear} One creature flees for short duration.

\spellhead[Evoc]{Burning Hands} 1d8 fire damage in medium cone.
\spellhead{Mage Armor} Armor made of magical force gives \plus2 AC.
\spellhead{Mage Hand} 5-pound telekinesis.
\spellhead{Magic Missile} Missile unerringly deals d10 damage.
\spellhead{Shocking Grasp} Touch deals 2d6 electricity damage.
\spellheadrestricted{Feather Fall} Objects or creatures fall slowly.
\spellheadrestricted{Gust of Wind} Blows away or knocks down smaller creatures in large line.

\spellhead[Illus]{Color Spray} Creatures in medium cone are bewildered.
\spellhead{Dancing Lights} Creates torches or other lights.
\spellhead{Light} Object shines like a torch.
\spellhead{Ventriloquism} Throws voice for medium duration.
\spellheadrestricted{Create Sound} Figment sounds.
\spellheadrestricted{Disguise Self} Changes your appearance.

\spellhead[Necro]{Ray of Enfeeblement} Ray inflicts 1d6 \plus1 per two levels Str penalty.
\spellhead{Ray of Clumsiness} Ray inflicts 1d6 \plus1 per two levels Dex penalty.
\spellheadrestricted{Chill Touch} One touch/level deals 1d6 damage \add casting attribute value and possibly 1 Str damage.
\spellheadrestricted{False Life} Gain 10 temporary hp \plus1/level.

\spellhead[Trans]{Magic Vestment} Armor or shield becomes \plus2.
\spellhead{Magic Weapon} Weapon becomes \plus2.
\spellhead{Jump} Subject gets bonus on Jump checks.
\spellheadrestricted{Backbiter} Weapon attacks its wielder when used.
\spellheadrestricted{Expeditious Retreat} Your speed doubles.
\end{swspelllist}

\subsection{2nd-level Sorcerer/wizard Spells} 
\begin{swspelllist}
\spellhead[Abjur]{Resist Energy} Ignores first 10 (or more) points of damage/attack from specified energy type.
\spellhead{Share Pain} Willing subject takes half of your damage.
\spellhead{Shield Other} You take half of subject's damage.
\spellheadrestricted{Ablative Shield} Immediately reduce damage from foes' spells.
\spellheadrestrictedc{Spelltheft}{Lesser} As \spell{lesser dispel magic}, but you gain the effects of dispelled spells.

\spellhead[Conj]{Acid Arrow} Ranged touch attack deals 3d6 damage, more over time.
\spellhead{Glitterdust} Outlines concealed and invisible creatures in small radius.
\spellhead{Summon Monster II} Calls extraplanar creature to fight for you.
\spellheadrestricted{Dissipating Touch} Touch deals 4d8 damage.
\spellheadrestricted{Summon Monster III} Calls extraplanar creature to fight for you.

\spellhead[Div]{Arcane Sight} Magical auras become visible to you.
\spellhead{Comprehend Languages} You understand all spoken and written languages.
\spellhead{Darkvision} See 60 ft. in total darkness.
\spellhead{Locate Object} Senses direction toward object (specific or type).
\spellhead{See Invisibility} Reveals invisible creatures or objects.
\spellheadrestricted{}
\spellheadrestricted{Precognition} See into the future to gain attack and saving throw bonus.

\spellhead[Ench]{Hold Person} Paralyzes one humanoid for short duration.
\spellhead{Hypnotic Pattern} Fascinates creatures in small radius.
\spellhead{Touch of Idiocy} Subject takes \minus4 penalty to Int, Wis, and Cha.
\spellheadrestricted{Aversion} Subject has aversion you specify.
\spellheadrestricted{Charm Person} Makes one person your friend.

\spellheadc[Evoc]{Cone of Cold}{Lesser} 2d6 cold damage and brief fatigue in medium cone.
\spellhead{Scorching Ray} Ray (or rays) deal 6d6 fire damage.
\spellhead{Shatter} Sonic vibration damages objects or crystalline creatures.
\spellheadrestricted{Knock} Opens locked or magically sealed door.
\spellheadrestricted{Levitate} Subject moves up and down at your direction.

\spellhead[Illus]{Blur} Attacks miss subject 20\% of the time.
\spellhead{Darkness} Object sheds supernatural shadow in medium radius.
\spellhead{Hypnotic Pattern} Fascinates creatures in small radius.
\spellheadrestricted{Mirror Image} Creates decoy duplicates of you (1d4 \plus1 per three levels, max 8).
\spellheadrestricted{Silent Image} Creates minor illusion of your design.

\spellhead[Necro]{Death Knell} Kill dying creature and gain temporary hp, more for killing powerful creatures.
\spellhead{Ghoul Touch} Paralyze one subject, which exudes stench that makes those nearby sickened.
\spellhead{Share Pain} Willing subject takes half of your damage.
\spellhead{Shield Other} You take half of subject's damage.
\spellheadrestricted{Crush Life} Target takes 4d6 unavoidably.
\spellheadrestrictedc{Transfer Suffering}{Lesser} Give 2d6 of your damage to touched subject.

\spellhead[Trans]{Spider Climb} Grants ability to walk on walls and ceilings.
\spellhead{Totemic Mind} Subject gains \plus2 to Int, Wis, or Cha for medium duration.
\spellhead{Totemic Power} Subject gains \plus2 to Str, Dex, or Con for medium duration.
\spellhead{Vestments of the Mage} Reduce spell failure on armor or shield by 10\%.
\spellheadrestricted{Reduce Person} Humanoid creature halves in size.
\spellheadrestricted{Slow} Subject takes only one action/round.
\end{swspelllist}

\subsection{3rd-level Sorcerer/wizard Spells} 
\begin{swspelllist}
\spellhead[Abjur]{Dimensional Anchor} Bars extradimensional movement.
\spellhead{Dispel Magic} Cancels magical spells and effects.
\spellhead{Protection from Energy} Absorb 10 points/level of damage from one kind of energy.
\spellhead{Share Pain, Forced} Unwilling subject takes half of your damage.
\spellheadrestricted{Invisibility Purge} Suppresses invisibility within large radius.
\spellheadrestricted{Redirection} Redirect attacks on you to other targets.

\spellhead[Conj]{Dimensional Anchor} Bars extradimensional movement.
\spellhead{Fog Cloud} Small cylinder of fog obscures vision.
\spellhead{Summon Monster III} Calls extraplanar creature to fight for you.
\spellhead{Web} Fills medium radius spread with sticky spiderwebs.
\spellheadrestricted{Dimension Slide} Teleports you very short distance.
\spellheadrestricted{Summon Monster IV} Calls extraplanar creature to fight for you.

\spellhead[Div]{Detect Thoughts} Allows ``listening" to surface thoughts.
\spellhead{Telepathic Bond} Mental link lets two creatures communicate.
\spellheadrestrictedc{Moment of Prescience}{Lesser} Roll twice for single attack, save, or opposed check.
\spellheadrestricted{Tongues} Speak and understand any language.

\spellhead[Ench]{Confusion} Subject acts randomly for short duration.
\spellhead{Daze} Subject is bewildered and may lose an action.
\spellheadrestricted{Crushing Despair} Subjects in medium cone are demoralized.
\spellheadrestricted{Heroism} Gives \plus2 on attacks and saves.

\spellhead[Evoc]{Fireball} Small radius of fire deals 3d6 damage.
\spellhead{Lightning Bolt} Large line of electricity deals 3d6 damage.
\spellheadrestricted{Interposing Hand} Hand provides cover against one opponent at a time.
\spellheadrestricted{Telekinetic Force} Move an object with the sustained force of your mind.

\spellhead[Illus]{Invisibility} Subject is invisible for medium duration or until it attacks.
\spellhead{Rainbow Pattern} Lights fascinate 24 HV of creatures.
\spellheadrestricted{Minor Image} As \spell{silent image}, plus some sound.
\spellheadrestricted{Phantom Steed} Shadow horse appears for extended duration.

\spellhead[Necro]{Bestow Curse} \minus6 to an attribute score; \minus4 on attack rolls, saves, checks, and AC; or 25\% chance of losing each action.
\spellhead{Blindness/Deafness} Makes subject blinded or deafened.
\spellhead{Contagion} Infects subject with chosen disease.
\spellhead{Ray of Exhaustion} Ray makes subject exhausted.
\spellhead{Share Pain, Forced} Unwilling subject takes half of your damage.
\spellheadrestricted{Link Vitality} Two targets share each others' damage and healing.
\spellheadrestricted{Vampiric Touch} Touch deals 6d8 damage; caster gains half damage as temporary hp.

\spellhead[Trans]{Phantom Steed} Shadow horse appears for extended duration.
\spellhead{Shape Stone} Sculpts stone into any shape.
\spellhead{Shrink Item} Object shrinks to one-sixteenth size.
\spellhead{Telepathic Bond} Mental link lets two creatures communicate.
\spellheadrestricted{Enlarge Person} Humanoid creature doubles in size.
\spellheadrestricted{Gaseous Form} Subject becomes insubstantial and can fly slowly.
\end{swspelllist}

\subsection{4th-level Sorcerer/wizard Spells} 
\begin{swspelllist}
\spellhead[Abjur]{Dismissal} Forces a creature to return to native plane.
\spellhead{Fire Shield} Creatures attacking you take damage; you're protected from heat or cold.
\spellheadc{Globe of Invulnerability}{Lesser} Stops 1st- through 3rd-level spell effects.
\spellheadc{Resist Energy}{Greater} As \spell{resist energy}, but all energy types.
\spellhead{Spell Immunity} Subject is immune to one spell per four levels.
\spellheadrestricted{Ablative Fortress} Immediately reduce damage from foes' spells in area.
\spellheadrestricted{Inertial Shield} Gain physical DR 10.

\spellhead[Conj]{Blink} You randomly vanish and reappear for short duration.
\spellhead{Dismissal} Forces a creature to return to native plane.
\spellhead{Ice Storm} Hail deals 4d6 damage in small cylinder.
\spellhead{Summon Monster IV} Calls extraplanar creature to fight for you.
\spellhead{Wall of Ice} Create plane of ice or hemisphere that can trap creatures inside.
\spellheadrestricted{Dimension Door} Teleports you within long range.
\spellheadrestricted{Summon Monster V} Calls extraplanar creature to fight for you.

\spellhead[Div]{Locate Creature} Indicates direction to familiar creature.
\spellheadc{Moment of Prescience}{Lesser} Roll twice for single attack, save, or opposed check.
\spellhead{Tongues} Speak and understand any language.
\spellheadrestricted{Discern Vulnerability} Quickly find weaknesses in foe's defenses.
\spellheadrestrictedc{Precognition}{Greater} See into the future to gain attack, damage, dodge, and saving throw bonus.

\spellhead[Ench]{Hold Monster} As hold person, but any creature.
\spellhead{Phantasmal Killer} Fearsome illusion can frighten target to death.
\spellheadc{Sleep}{Mass} Tire multiple creatures, possibly putting them to sleep.
\spellheadrestricted{}
\spellheadrestricted{}

\spellhead[Evoc]{Fire Shield} Creatures attacking you take damage; you're protected from heat or cold.
\spellhead{Ice Storm} Hail deals 4d6 damage in small cylinder.
\spellhead{Shout} Deafens all within medium cone and deals 4d6 sonic damage.
\spellhead{Wall of Ice} Create plane of ice or hemisphere that can trap creatures inside.
\spellheadrestricted{Resilient Sphere} Force globe protects but traps one creature or object.
\spellheadrestricted{Telekinetic Maneuver} Telekinetically bull rush, disarm, grapple, or trip your target.

\spellhead[Illus]{Displacement} Attacks miss subject 50\% of the time.
\spellhead{Phantasmal Killer} Fearsome illusion can frighten target to death.
\spellheadrestricted{Major Image} As silent image, plus sound, smell and thermal effects.
\spellheadrestricted{Shadow Conjuration} Mimics conjuration below 4th level, but only half real.

\spellhead[Necro]{Death Ward} Grants immunity to death spells and negative energy effects.
\spellhead{Enervation} Subject gains 1d4 negative levels.
\spellheadrestricted{Transfer Suffering} Give 8d8 points of your damage to touched subject.
\spellheadrestricted{}

\spellhead[Trans]{Fly} Subject flies at speed of 60 ft.
\spellhead{Invest Magic} All of subject's equipment is magical.
\spellheadrestricted{Haste} Subject moves faster, gets extra attack.
\end{swspelllist}

\subsection{5th-level Sorcerer/wizard Spells} 
\begin{swspelllist}
\spellhead[Abjur]{Globe of Invulnerability} As \spell{lesser globe of invulnerability}, plus 4th-level spell effects.
\spellhead{Magic Circle against Chaos/Evil/Good/Law} Protect allies in a \areamed radius and hedge out opposing outsiders.
\spellhead{Spell Resistance} Subject gains SR 10.
\spellheadrestricted{Spelltheft} As \spell{dispel magic}, but you gain the effects of dispelled spells.
\spellheadrestricted{Retributive Shield} Gain DR 5/- that reflects damage at foes.

\spellheadc[Conj]{Acid Arrow}{Greater} As acid arrow, but 10d6 damage initially and more over time.
\spellheadc{Glitterdust}{Greater} Dazzles and outlines concealed and invisible creatures in small radius.
\spellhead{Mind Fog} Fog imposes \minus10 penalty to Wis checks and Will saves.
\spellhead{Stinking Cloud} Fog nauseates creatures inside.
\spellhead{Summon Monster V} Calls extraplanar creature to fight for you.
\spellheadrestrictedc{Retrieve}{Greater} Teleport an item you can see to your hand.
\spellheadrestricted{Summon Monster VI} Calls extraplanar creature to fight for you.

\spellhead[Div]{Discern Vulnerability} Quickly find weaknesses in foe's defenses.
\spellheadc{Precognition}{Greater} See into the future to gain attack, damage, dodge, and saving throw bonus.
\spellheadrestricted{True Seeing}\M Lets you see all things as they really are.
\spellheadrestricted{True Strike} \plus20 on your next attack roll.

\spellhead[Ench]{Feeblemind} Subject's Int and Cha drop to 1.
\spellhead{Mind Fog} Fog imposes \minus5 penalty to Wisdom.
\spellhead{Suggestion} Compels subject to follow stated course of action.
\spellheadrestricted{Charm Monster} Makes monster believe it is your ally.
\spellheadrestricted{Fear} Subjects within medium cone flee for short duration.

\spellhead[Evoc]{Chain Lightning} 10d6 damage; 1 secondary bolt/level each deals half damage.
\spellhead{Cone of Cold} 5d6 cold damage and fatigue in medium cone.
\spellhead{Wall of Fire} Passing through wall deals 5d6 damage.
\spellheadrestricted{Telekinetic Thrust} Hurl objects with the force of your mind.
\spellheadrestricted{Wall of Force} Wall is immune to damage.

\spellhead[Illus]{Invisibility Sphere} Makes everyone within small radius invisible.
\spellhead{Phantom Maze} Subject acts as if affected by \spell{maze}.
\spellheadrestricted{Seeming} Changes appearance of group of creatures.
\spellheadrestricted{Shadow Evocation} Mimics evocation below 5th level, but only half real.

\spellhead[Necro]{Crippling Pain} Creature suffers penalties due to debilitating pain.
\spellhead{Waves of Fatigue} Creatures in large cone become fatigued.
\spellheadrestrictedc{Crush Life}{Greater} Target takes 10d6 damage unavoidably and gains a negative level.
\spellheadrestricted{}

\spellhead[Trans]{Combat Transformation}\M You gain combat bonuses.
\spellhead{Passwall} Creates passage through wood or stone wall.
\spellhead{Stoneskin} \plus5 AC, ignore 10 damage per round.
\spellhead{Wall of Stone} Creates a stone wall that can be shaped.
\spellheadrestricted{Animate Objects} Objects attack your foes.
\spellheadrestricted{Baleful Polymorph} Transforms subject into harmless animal.
\end{swspelllist}

\subsection{6th-level Sorcerer/wizard Spells} 
\begin{swspelllist}
\spellhead[Abjur]{Banishment} Banishes extraplanar creatures in medium radius around you.
\spellheadc{Dispel Magic}{Greater} As dispel magic, but up to \plus20 on check.
\spellheadc{Protection from Energy}{Greater} As \spell{protection from energy}, but against all energy types.
\spellheadrestricted{Repulsion} Creatures can't approach you.
\spellheadrestrictedc{Shield Other}{Greater} You take all of subject's damage.

\spellhead[Conj]{Black Tentacles} Tentacles grapple all within small spread.
\spellhead{Solid Fog} Fog blocks vision and slows movement.
\spellhead{Summon Monster VI} Calls extraplanar creature to fight for you.
\spellheadrestricted{Phase Door} Creates an invisible passage through wood or stone.
\spellheadrestricted{Summon Monster VII} Calls extraplanar creature to fight for you.

\spellhead[Div]{Locate Entity} Locates creatures or objects within 1 mile.
\spellhead{Nightmare} Sends vision dealing 1d10 damage, fatigue.
\spellheadc{Telepathic Bond}{Mass} Mental link lets allies communicate.
\spellhead{True Seeing}\M Lets you see all things as they really are.
\spellheadrestricted{Analyze Dweomer}\F Reveals magical aspects of subject.
\spellheadrestricted{Moment of Prescience} As lesser moment of prescience, with \plus1/2 levels circumstance bonus.

\spellhead[Ench]{Insanity} Subject suffers continuous \spell{confusion}.
\spellhead{Power Word Confuse} Confuses a creature.
\spellheadrestricted{Dominate Person} Controls humanoid telepathically.
\spellheadrestrictedc{Heroism}{Greater} Gives immunity to fear and bonus on attacks, saves, and checks.

\spellhead[Evoc]{Delayed Blast Fireball} 6d8 fire damage in small radius; you can postpone blast for 5 rounds.
\spellhead{Mage's Sword}\F Floating magic blade strikes opponents.
\spellheadrestricted{Forceful Hand} Hand pushes creatures away.
\spellheadrestricted{Telekinesis} Moves object, attacks creature, or hurls object or creature.

\spellheadc[Illus]{Invisibility}{Greater} As \spell{invisibility}, but subject can attack and stay invisible.
\spellhead{Mislead} Turns you invisible and creates illusory double.
\spellheadrestricted{Persistent Image} As \spell{major image}, but no concentration required.
\spellheadrestricted{}

\spellhead[Necro]{Circle of Death}\M Kills 2/level HV of creatures.
\spellhead{Undeath to Death}\M Destroys 2/level HV of undead.
\spellheadrestricted{Harm} Deals 12d8 damage to touched target.
\spellheadrestrictedc{Share Pain}{Greater} Willing subject takes all of your damage.

\spellhead[Trans]{Black Tentacles} Tentacles grapple all within small spread.
\spellhead{Disintegrate} Makes one creature or object vanish.
\spellheadc{Telepathic Bond}{Mass} Mental link lets allies communicate.
\spellheadc{Totemic Mind}{Mass} As \spell{totemic mind}, but affects multiple creatures.
\spellheadc{Totemic Power}{Mass} As \spell{Totemic power}, but affects multiple creatures.
\spellheadrestrictedc{Reduce Person}{Mass} Reduces several creatures.
\spellheadrestricted{Transmute Flesh and Stone} Turns subject creature into statue, or restores petrified creature.
\end{swspelllist}

\subsection{7th-level Sorcerer/wizard Spells}
\begin{swspelllist}
\spellhead[Abjur]{Energy Conversion} As \spell{greater resist energy}, plus you can fire rays of absorbed energy.
\spellhead{}
\spellheadrestricted{Antimagic Field} Negates magic in small emanation from you.
\spellheadrestricted{Spell Turning} Reflect 1d4\plus6 spell levels back at caster.

\spellhead[Conj]{Banishment} Banishes 2 HV/level of extraplanar creatures.
\spellhead{Cloudkill} Fog deals Con damage each round.
\spellhead{Ethereal Jaunt} You become ethereal for short duration.
\spellhead{Summon Monster VII} Calls extraplanar creature to fight for you.
\spellheadrestrictedc{Dimension Door}{Mass} Teleport you and one ally/level within long range.
\spellheadrestricted{Summon Monster VIII} Calls extraplanar creature to fight for you.

\spellheadc[Div]{Arcane Sight}{Greater} As arcane sight, but also reveals magic effects on creatures and objects.
\spellhead{Moment of Prescience} As lesser moment of prescience, with \plus1/2 levels circumstance bonus.
\spellheadrestricted{Foresight} ``Sixth sense" warns of impending danger.
\spellheadrestricted{}

\spellhead[Ench]{Deep Slumber} Subject falls asleep for a long time.
\spellheadc[Ench]{Hold Person}{Mass} As \spell{hold person}, but affects multiple creatures.
\spellheadrestrictedc{Charm Person}{Mass} As \spell{charm person}, but any within medium radius.
\spellheadrestricted{Terror} Subject is unavoidably afraid.

\spellhead[Evoc]{Energy Conversion} As \spell{greater resist energy}, plus you can fire rays of absorbed energy.
\spellhead{Freezing Sphere} Freezes water or deals 7d6 cold damage in medium radius.
\spellhead{Prismatic Spray} Rays hit subjects with variety of effects.
\spellheadc{Shout}{Greater} Yell deals 7d6 sonic damage in large cone; stuns creatures, damages objects.
\spellheadrestricted{Forcecage}\M Cube or cage of force imprisons all inside.
\spellheadrestricted{Grasping Hand} Hand provides cover, pushes, or grapples.

\spellheadc[Illus]{Invisibility}{Mass} As \spell{invisibility}, but affects any within 100 ft.
\spellheadrestricted{Project Image} Illusory double can talk and cast spells.
\spellheadrestrictedc{Shadow Conjuration}{Greater} As shadow conjuration, but up to 6th level.

\spellhead[Necro]{Power Word Blind} Blinds bloodied creature.
\spellhead{Strip the Flesh} Rend foe's skin from its body.
\spellheadrestricted{Finger of Death} Kills one subject.
\spellheadrestrictedc{Link Vitality}{Mass} As \spell{link vitality}, but affects one creature/level.

\spellhead[Trans]{Oak Body} Your body becomes as hard as oak.
\spellhead{Overland Flight} You fly at a speed of 40 ft. and can hustle over long distances.
\spellheadrestrictedc{Enlarge Person}{Mass} Enlarges several creatures.
\spellheadrestrictedc{Slow}{Mass} Many subjects take only one action/round.

\spellhead[Univ]{Limited Wish}\M Alters reality -- within spell limits.
\end{swspelllist}

\subsection{8th-level Sorcerer/wizard Spells} 
\begin{swspelllist}
\spellhead[Abjur]{Protection from Spells}\M\F Confers spell resistance 10 and \plus5 to saves vs. spells.
\spellheadc{Spell Immunity}{Greater} As spell immunity, but up to 8th-level spells.
\spellheadrestrictedc{Spelltheft}{Greater} As \spell{greater dispel magic}, but you gain the effects of dispelled spells.

\spellhead[Conj]{Acid Fog} Medium cylinder of solid fog deals acid damage.
\spellhead{Sea of Fog} 200 ft. radius of fog obscures vision.
\spellhead{Summon Monster VIII} Calls extraplanar creature to fight for you.
\spellheadrestricted{Maze} Traps subject in extradimensional maze.
\spellheadrestricted{Summon Monster IV} Calls extraplanar creature to fight for you.

\spellheadc[Div]{Detect Thoughts}{Greater} As \spell{detect thoughts}, but longer and without concentration.
\spellhead{Foresight} ``Sixth sense" warns of impending danger.
\spellheadrestricted{Revelation} Grant target vision of one of three futures.

\spellheadc[Ench]{Confusion}{Mass} Subjects in medium radius act randomly for short duration.
\spellhead{Power Word Stun} Stuns a creature.
\spellhead{Scintillating Pattern} Twisting colors confuse, stun, or render unconscious.
\spellheadc{Suggestion}{Mass} As \spell{suggestion}, but shorter and affects multiple creatures.
\spellheadrestricted{Dominate Monster} As dominate person, but any creature.

\spellheadc[Evoc]{Cone of Cold}{Greater} 8d6 cold damage and fatigue in large cone.
\spellhead{Polar Ray} Ranged touch attack slows and deals 16d6 cold damage.
\spellhead{Prismatic Wall} Wall's colors have array of effects.
\spellheadrestricted{Crushing Hand} Large hand provides cover or crushes your foes.

\spellhead[Illus]{Scintillating Pattern} Twisting colors confuse, stun, or render unconscious.
\spellhead{Shadow Body} You become a living shadow.
\spellheadrestrictedc{Shadow Evocation}{Greater} As shadow evocation, but up to 7th level.

\spellhead[Necro]{Energy Drain} Subject gains 2d4 negative levels.
\spellhead{Waves of Exhaustion} Creatures in medium cone become exhausted.
\spellheadrestrictedc{Crush Life}{Mass} Deals 8d6 damage unavoidably to many foes.

\spellhead[Trans]{Iron Body} Your body becomes living iron.
\spellhead{Reverse Gravity} Objects and creatures fall upward.
\spellhead{Shadow Body} You become a living shadow.
\spellhead{Temporal Stasis}\M Puts subject into suspended animation.
\spellheadrestrictedc{Haste}{Mass} Many subjects move faster, get extra attack.
\end{swspelllist}

\subsection{9th-level Sorcerer/wizard Spells} 
\begin{swspelllist}
\spellhead[Abjur]{Emancipation} Releases creature from imprisonment, many other impediments.
\spellheadrestricted{Mage's Disjunction} Dispels all magic.

\spellhead[Conj]{Etherealness} Travel to Ethereal Plane with companions.
\spellhead{Imprisonment} Entombs subject beneath the earth.
\spellhead{Shadow Puppet} Step into Plane of Shadow and create shadow duplicate in your stead.
\spellhead{Summon Monster IX} Calls extraplanar creature to fight for you.
\spellheadrestricted{}

\spellhead[Div]{Revelation} Grant target vision of one of three futures.
\spellheadrestrictedc{Moment of Prescience}{Greater} As lesser moment of prescience, but with \plus1/level circumstance bonus.

\spellheadc[Ench]{Daze}{Mass} As \spell{daze}, but affects multiple creatures.
\spellheadc{Hold Monster}{Mass} As \spell{hold monster}, but affects multiple creatures.
\spellhead{Weird} As \spell{phantasmal killer}, but affects multiple creatures
\spellheadrestricted{Irresistible Dance} Forces subject to dance.

\spellhead[Evoc]{Meteor Swarm} Four exploding spheres each deal 6d6 fire damage in large radius.
\spellhead{Prismatic Sphere} As \spell{prismatic wall}, but surrounds on all sides.
\spellheadrestricted{Clenched Fist} Large hand provides cover or attacks your foes.

\spellhead[Illus]{Weird} As \spell{phantasmal killer}, but affects multiple creatures
\spellheadrestricted{Shadow Puppet} Step into Plane of Shadow and create shadow duplicate in your stead.

\spellhead[Necro]{Assimilate} Incorporate creature into your own body.
\spellhead{Power Word Kill} Kills one bloodied creature.
\spellheadrestricted{Wail of the Banshee} Kills many creatures.

\spellhead[Trans]{Alter Any Object} Transforms objects into new forms.
\spellhead{Assimilate} Incorporate creature into your own body.
\spellhead{Imprisonment} Entombs subject beneath the earth.
\spellheadrestricted{Time Stop} You act freely for 1d4\plus1 rounds.

\spellhead[Univ]{Wish}\M As \spell{limited wish}, but with fewer limits.
\end{swspelllist}

\section{Arcane Invocation List}
\begin{swspelllist}
\spellhead[Abjur]{Ablative Aura} Damage foes who attack a shielded ally.
\spellhead{Bestow Protection} Protect ally from physical or magical attacks.
\spellhead[Conj]{Acid Orb} Propel a small orb of acid at a foe for d4 damage.
\spellhead{Conjure Projectile} Summon projectiles that hit a foe for d6 damage.
\spellhead[Div]{Premonition} Grant ally temporary attack bonus.
\spellhead{Twist Fate} Know foe's future action and apply penalty of choice.
\spellheadc[Ench]{Confusion}{Lesser} Touched foe is briefly confused.
\spellhead{Distract} Foe is bewildered.
\spellhead[Evoc]{Magic Ray} Fire ray of magical energy.
\spellheadc{Telekinesis}{Lesser} Attack with weapon telekinetically.
\spellhead[Illus]{False Foe} Create illusionary creature that acts like an ally.
\spellhead{Phantom Injury} Foe believes it is grievously wounded.
\spellhead[Necro]{Draining Touch} Steal life force from touched foe.
\spellhead{Exhaustion} Briefly exhaust foe.
\spellhead[Trans]{Imbue Weapon} Weapon deals d8 extra damage when it next hits.
\spellheadc{Slow}{Lesser} Subject only takes one action next round.
\end{swspelllist}

\section{Arcane Ritual List}
\subsection{1st-level Arcane Rituals}
\begin{rituallist}
\spellhead[Arcana]{Alarm} Wards large emanation for 24 hours.
\spellhead{Erase} Mundane or magical writing vanishes.
\spellhead{Floating Disk} Creates 3 ft. diameter horizontal disk that holds 100 lb./level.
\spellhead{Identify} Determines properties of magic item.
\spellhead{Unseen Servant} Invisible force obeys your commands.
\spellhead[Craft]{Mending} Makes minor repairs on an object.
\spellhead[Engineering]{Mending} Makes minor repairs on an object.
\spellhead[Forgery]{Erase} Mundane or magical writing vanishes.
\spellhead[Nature]{Mount} Summons riding horse for 24 hours.
\spellhead[Spell]{Magic Aura} Alters object's magic aura.
\spellhead{Read Magic} Read magical writing.
\spellhead[Survival]{Endure Elements} Exist comfortably in hot or cold environments.
\spellhead{Create Food and Water}
\spellhead[None]{Arcane Mark} Inscribes a personal rune (visible or invisible).
\spellhead{Prestidigitation} Performs minor tricks.
\end{rituallist}

\subsection{2nd-level Arcane Rituals}
\begin{rituallist}
\spellhead[Arcana]{Arcane Lock} Magically locks a portal or chest.
\spellhead{Continual Flame} Makes a permanent, heatless torch.
\spellhead{Find Traps} Notice traps better.
\spellhead{Magic Mouth} Speaks once when triggered.
\spellhead{Whispering Wind} Sends a short message 1 mile/level.
\spellhead[Eng]{Find Traps} Notice traps better.
\spellhead[Heal]{Gentle Repose} Preserves one corpse.
\spellhead[Nature]{Whispering Wind} Sends a short message 1 mile/level.
%\spellhead[Planes]{Rope Trick} As many as eight creatures hide in extradimensional space.
%\spellhead[Spell]{Misdirection} Misleads divinations for one creature or object.
\spellhead[Spell]{Undetectable Alignment} Conceals alignment for 24 hours.
\end{rituallist}

\subsection{3rd-level Arcane Rituals}
\begin{rituallist}
\spellhead[Arcana]{Explosive Runes} Deals 1d6/level damage in a small radius when read.
\spellhead{Clairaudience/Clairvoyance} Hear or see at a distance for 10 minutes.
\spellhead{Fire Trap} Opened object deals 1d4/level fire damage.
\spellhead{Illusory Wall} Wall, floor, or ceiling looks real, but anything can pass through.
\spellhead{Secret Page} Changes one page to hide its real content.
\spellhead{Sepia Snake Sigil} Creates text symbol that immobilizes reader.
\spellheadc[Craft]{Mending}{Greater} Repairs an object.
\spellheadc[Eng]{Mending}{Greater} Repairs an object.
%\spellhead[Craft]{Object Reading} Learn details about an object's previous owners.
%\spellhead[Engineering]{Object Reading} Learn details about an object's previous owners.
\spellhead[Forgery]{Secret Page} Changes one page to hide its real content.
\spellhead[Nature]{Fire Trap} Opened object deals 1d4/level fire damage.
\spellhead{Water Breathing} Subjects can breathe underwater.
\spellhead[Relig]{Animate Dead} Creates undead skeletons and zombies.
\spellhead{Speak with Dead} Corpse answers one question/two levels.
\spellhead[Spell]{Nondetection} Hides subject from divination, scrying.
%\spellhead{Obscure Object} Masks object against scrying.
\spellhead[Survival]{Tiny Hut} Creates shelter for ten creatures.
\spellhead{Water Breathing} Subjects can breathe underwater.
\end{rituallist}

\subsection{4th-level Arcane Rituals}
\begin{rituallist}
\spellhead[Arcana]{Arcane Eye} Invisible floating eye moves 30 ft./round.
\spellheadc{Geas}{Lesser} Commands subject of 7 HV or less.
\spellhead{Hallucinatory Terrain} Makes one type of terrain appear like another (field into forest, or the like).
%\spellhead{Mage's Faithful Hound} Phantom dog can guard, attack.
\spellhead{Remove Curse} Frees object or person from curse.
\spellhead{Secure Shelter} Creates sturdy cottage.
\spellhead{Symbol of Persuasion} Trap compels triggering creature to follow \spell{suggestion}.
\spellhead[Craft]{Minor Creation} Creates one cloth or wood object.
\spellhead[Eng]{Minor Creation} Creates one cloth or wood object.
\spellhead[Forgery]{Illusory Script} Only intended reader can decipher.
\spellhead{Secure Shelter} Creates sturdy cottage.
\spellhead[Heal]{Remove Curse} Frees object or person from curse.
\spellhead[Linguistics]{Illusory Script} Only intended reader can decipher.
\spellhead[Spell]{Detect Scrying} Alerts you of magical eavesdropping.
\spellhead[Relig]{Divination} Provides useful advice for specific proposed actions.
\spellheadc{Geas}{Lesser} Commands subject of 7 HV or less.
\spellhead{Hallucinatory Terrain} Makes one type of terrain appear like another (field into forest, or the like).
\end{rituallist}

\subsection{5th-level Arcane Rituals}
\begin{rituallist}
\spellhead[Arcana]{Dream} Sends message to anyone sleeping.
\spellhead{Mage's Faithful Hound} Phantom dog can guard, attack.
%\spellhead{Mirage Arcana} As hallucinatory terrain, plus structures.
\spellhead{Prying Eyes} 1d4 \plus1/level floating eyes scout for you.
\spellhead{Sending} Delivers short message anywhere, instantly.
\spellhead{Symbol of Pain} Trap fills triggering creature with pain.
\spellhead[Eng]{Fabricate} Transforms raw materials into finished items.
\spellhead{Major Creation} As minor creation, plus stone and metal.
\spellhead[Local]{Legend Lore}\F Lets you learn tales about a person, place, or thing.
\spellhead[Nature]{Scrying}\F Spies on subject from a distance.
\spellhead[Planes]{Contact Other Plane} Lets you ask question of extraplanar entity.
\spellhead{Dimensional Lock} Teleportation and interplanar travel blocked for 30 days.
\spellheadc{Planar Binding}{Lesser} Traps extraplanar creature of 6 HV or less until it performs a task.
\spellhead[Relig]{Scrying}\F Spies on subject from a distance.
\spellhead{Secret Chest}\F Hides expensive chest on Ethereal Plane; you retrieve it at will.
\spellhead{Sending} Delivers short message anywhere, instantly.
\spellhead[Spell]{Break Enchantment} Frees subjects from enchantments, alterations, curses, and petrification.
\spellhead{False Vision} Fools scrying with an illusion.
\spellhead{Mage's Private Sanctum} Prevents anyone from viewing or scrying an area for 24 hours.
\spellhead{Permanency} Makes certain spells permanent.
\end{rituallist}

\subsection{6th-level Arcane Rituals}
\begin{rituallist}
\spellhead[Arcana]{Geas/Quest} As lesser geas, plus it affects any creature.
\spellhead{Nightmare} Sends vision dealing 1d10 damage, fatigue.
\spellhead{Permanent Image} Includes sight, sound, and smell.
\spellhead{Symbol of Insanity} Trap drives triggering creature insane.
\spellhead{Teleport} Instantly transports you as far as 100 miles/level.
\spellhead[Eng]{Move Earth} Digs trenches and build hills.
\spellhead[Geography]{Find the Path} Shows most direct way to a location.
\spellhead[Nature]{Move Earth} Digs trenches and build hills.
\spellhead[Relig]{Create Undead} Creates ghouls, ghasts, mummies, or mohrgs.
\spellhead{Geas/Quest} As lesser geas, plus it affects any creature.
\spellhead[Planes]{Planar Binding} As lesser planar binding, but up to 12 HV.
\spellhead{Shadow Walk} Step into shadow to travel rapidly.
\end{rituallist}

\subsection{7th-level Arcane Rituals}
\begin{rituallist}
%\spellhead[Arcana]{Binding} Utilizes an array of techniques to imprison a creature.
\spellhead[Arcana]{Clone}\F Duplicate awakens when original dies.
\spellhead{Instant Summons} Prepared object appears in your hand.
\spellhead{Programmed Image} As major image, plus triggered by event.
\spellheadc{Scrying}{Greater} As scrying, but faster and longer.
\spellhead{Sequester} Subject is invisible to sight and scrying; renders creature comatose.
\spellhead{Symbol of Death} Trap kills triggering creature.
\spellhead{Symbol of Sleep} Trap compels triggering creature to fall asleep.
\spellhead{Symbol of Terror} Trap frightens triggering creature.
\spellhead{Teleport Object} As teleport, but affects a touched object.
\spellhead[Eng]{Mordenkainen's Magnificent Mansion}\F Door leads to extradimensional mansion.
\spellhead[Local]{Vision} As legend lore, but quicker and strenuous.
\spellheadc[Nature]{Scrying}{Greater} As scrying, but faster and longer.
\spellhead[Planes]{Mordenkainen's Magnificent Mansion}\F Door leads to extradimensional mansion.
\spellhead{Plane Shift}\F As many as eight subjects travel to another plane.
\spellheadc[Relig]{Scrying}{Greater} As scrying, but faster and longer.
\spellhead[Spell]{Dimensional Lock} Teleportation and interplanar travel blocked for 30 days.
\end{rituallist}

\subsection{8th-level Arcane Rituals}
\begin{rituallist}
\spellheadc{Teleport}{Greater} As teleport, but no range limit and no off-target arrival.
\spellhead[Div]{Discern Location} Reveals exact location of creature or object.
\spellheadc{Prying Eyes}{Greater} As prying eyes, but eyes have true seeing.
\spellheadc{Scrying}{Greater} As scrying, but faster and longer.
\spellhead[Ench]{Antipathy} Object or location affected by spell repels certain creatures.
\spellhead{Demand} As \spell{sending}, plus you can send \spell{suggestion}.
\spellhead{Sympathy}\F Object or location attracts certain creatures.
\spellhead[Illus]{Screen} Illusion hides area from vision, scrying.
\spellhead[Relig]{Create Greater Undead}\M Creates shadows, wraiths, spectres, or devourers.
\end{rituallist}

\subsection{9th-level Arcane Rituals}
\begin{rituallist}
\spellhead[Abjur]{Mind Blank} Subject is immune to mental/emotional magic and scrying.
\spellhead[Conj]{Teleportation Circle} Circle teleports any creature inside to designated spot.
\spellhead[Necro]{Soul Bind}\F Traps newly dead soul to prevent resurrection.
\end{rituallist}

\section{Divine Ritual List}
\subsection{1st-level Divine Rituals}
\begin{rituallist}
\spellhead[Craft]{Mending} Makes minor repairs on an object.
\spellhead[Eng]{Mending} Makes minor repairs on an object.
\spellhead[Nature]{Pass without Trace} One subject/level leaves no tracks.
\spellhead[Relig]{Bless Water} Makes holy water.
\spellhead{Curse Water} Makes unholy water.
\spellhead[Spellcraft]{Read Magic} Read magical writing.
\spellhead[Survival]{Create Water} Creates 2 gallons/level of pure water.
\spellhead{Endure Elements} Exist comfortably in hot or cold environments.
\spellhead{Pass without Trace} One subject/level leaves no tracks.
\spellhead{Purify Food and Drink} Purifies 1 cu. ft./level of food or water.
\end{rituallist}

\subsection{2nd-level Divine Rituals}
\begin{rituallist}
\spellhead{Animal Messenger} Sends a Tiny animal to a specific place.
\spellhead{Augury}\M\F Learns whether an action will be good or bad.
\spellhead{Consecrate} Fills area with positive energy, making undead weaker.
\spellhead{Continual Flame} Makes a permanent, heatless torch.
\spellhead{Desecrate} Fills area with negative energy, making undead stronger.
\spellhead{Find Traps} Notice traps better.
\spellhead{Gentle Repose} Preserves one corpse.
\spellheadc{Restoration}{Lesser} Dispels magical ability penalty or repairs 1d4 ability damage.
\spellhead{Undetectable Alignment} Conceals alignment for 24 hours.
\spellhead{Water Breathing} Subjects can breathe underwater.
\spellhead{Whispering Wind} Sends a short message 1 mile/level.
\spellhead[Survival]{Create Food and Water}
\end{rituallist}

\subsection{3rd-level Divine Rituals}
\begin{rituallist}
\spellhead{Animate Dead} Creates undead skeletons and zombies.
\spellhead{Fire Trap} Opened object deals 1d4/level fire damage.
\spellhead{Glyph of Warding}\M Inscription harms those who pass it.
\spellhead{Helping Hand} Ghostly hand leads subject to you.
\spellhead{Ironwood} Magic wood is strong as steel.
\spellheadc{Mending}{Greater} Repairs an object.
\spellhead{Obscure Object} Masks object against scrying.
\spellhead{Remove Disease} Cures all diseases affecting subject.
\spellhead{Remove Blindness/Deafness} Cures normal or magical conditions.
\spellhead{Remove Curse} Frees object or person from curse.
\spellhead{Speak with Dead} Corpse answers one question/two levels.
\end{rituallist}

\subsection{4th-level Divine Rituals}
\begin{rituallist}
\spellhead{Divination} Provides useful advice for specific proposed actions.
\spellhead{Neutralize Poison} Immunizes subject against poison, detoxifies venom in or on subject.
\spellheadc{Planar Ally}{Lesser} Exchange services with a 6 HV extraplanar creature.
\spellhead{Reincarnate} Brings dead subject back in a random body.
\spellhead{Restoration} Restores level and attribute score drains.
\spellhead{Sending} Delivers short message anywhere, instantly.
\end{rituallist}

\subsection{5th-level Divine Rituals}
\begin{rituallist}
\spellhead{Atonement}\F\M Removes burden of misdeeds from subject.
\spellhead{Break Enchantment} Frees subjects from enchantments, alterations, curses, and petrification.
\spellhead{Commune} Deity answers one yes-or-no question/level.
\spellhead{Commune with Nature} Learn about terrain for 1 mile/level.
\spellhead{Dimensional Lock} Teleportation and interplanar travel blocked for one day/level.
\spellhead{Mark of Justice} Designates action that will trigger curse on subject.
\spellhead{Raise Dead} Restores life to subject who died as long as thirty days ago.
\spellhead{Scrying}\F Spies on subject from a distance.
\spellhead{Tree Stride} Step from one tree to another far away.
\spellhead{Unhallow} Designates location as unholy.
\end{rituallist}

\subsection{6th-level Divine Rituals}
\begin{rituallist}
\spellhead{Create Undead} Create ghouls, ghasts, mummies, or mohrgs.
\spellhead{Find the Path} Shows most direct way to a location.
\spellhead{Geas/Quest} As \spell{lesser geas}, plus it affects any creature.
\spellheadc{Glyph of Warding}{Greater} As \spell{glyph of warding}, but up to 10d10 damage or 6th level spell.
\spellhead{Heroes' Feast} Food for one creature/level cures and grants combat bonuses.
\spellhead{Move Earth} Digs trenches and builds hills.
\spellhead{Planar Ally} As lesser planar ally, but up to 12 HV.
\spellhead{Plane Shift}\F As many as eight subjects travel to another plane.
\spellhead{Stone Tell} Talk to natural or worked stone.
\spellhead{Transport via Plants} Move instantly from one plant to another of the same kind.
\spellhead{Wind Walk} You and your allies turn vaporous and travel fast.
\end{rituallist}

\subsection{7th-level Divine Rituals}
\begin{rituallist}
\spellhead{Hallow} Designates location as holy.
\spellhead{Refuge} Alters item to transport its possessor to you.
\spellheadc{Restoration}{Greater} As restoration, plus other conditions.
\spellhead{Resurrection} Fully restore dead subject.
\spellhead{Symbol of Destruction} Trap kills triggering creature.
\end{rituallist}

\subsection{8th-level Divine Rituals}
\begin{rituallist}
\spellhead{Awaken} Animal or tree gains human intellect
\spellhead{Create Greater Undead} Create shadows, wraiths, spectres, or devourers.
\spellhead{Discern Location} Reveals exact location of creature or object.
\spellhead{Forbiddance} Blocks planar travel, damages creatures of different alignment.
\spellheadc{Planar Ally}{Greater} As \spell{lesser planar ally}, but up to 18 HV.
\spellheadc{Scrying}{Greater} As \spell{scrying}, but faster and longer.
\end{rituallist}

\subsection{9th-level Divine Rituals}
\begin{rituallist}
\spellhead{Antipathy} Object or location affected by spell repels certain creatures.
%\spellhead{Astral Projection} Projects you and companions onto Astral Plane.
\spellhead{Shambler} Summons 1d4\plus2 shambling mounds to fight for you.
\spellhead{Soul Bind}\F Traps newly dead soul to prevent resurrection.
\spellhead{Sympathy} Object or location attracts certain creatures.
\spellhead{True Resurrection} As resurrection, plus remains aren't needed.
\end{rituallist}
