\chapter{Alternate Class Features}

\section{Barbarian}

\cf{Bbn}{Spellsundering Strike (Su)} At 7th level, a barbarian can put such force into an attack that it can break magical effects on the target. The fact that most barbarians (and even many wizards) have no idea how this works is no impediment to its effectiveness.
\par A barbarian may make a spellsundering strike a number of times per day equal to his Constitution as part of an attack action. Anything hit by a spellsundering strike becomes the target of a targeted \spell{dispel magic} effect, with a bonus equal to the barbarian's class level (not capped at \plus10). However, the barbarian cannot dispel any effect with a caster level higher than the damage he dealt with the spellsundering strike (though the damage need not be lethal).

\section{Fighter}

\subsection{Unfettered Discipline}
Some fighters cast off the trappings of armor and shields, relying entirely on their agility and reactions to guide them through combat. In doing so, they become inhumanly perceptive  
These benefits replace all of the fighter's armor discipline class features, as well as his armor and shield proficiencies.

\cf{Ftr}{Unfettered Discipline (Ex)} At 1st level, a fighter learns to defend himself intuitively. While unarmored and unencumbered, he may add his Wisdom to his AC and MC.

This bonus to AC applies even against touch attacks or when the fighter is flat-footed. He loses this bonus when he is helpless, when he wears any armor, when he uses a shield, or when he carries a medium or heavy load.

\cf{Ftr}{Improved Unfettered Discipline (Ex)} At 7th level, a fighter's ability to fight without restriction improves. While unarmored and unencumbered, he does not provoke attacks of opportunity when he moves. In addition, he gains a \plus10 foot competence bonus to his land speed.

\cf{Ftr}{Greater Unfettered Discipline (Ex)} At 13th level, a fighter's ability to fight without restriction becomes still greater. While unarmored and unencumbered, he gains the benefits of the Counterstorm feat, even if he does not meet the prerequisites. In addition, his land speed bonus improves to \plus20.

\cf{Ftr}{True Unfettered Discipline (Ex)} At 19th level, the fighter may choose to gain the true unfettered discipline ability with his true discipline class feature. While the fighter is unarmored and unencumbered, he gains a \plus10 circumstance bonus to his dodge modifier to AC against the first attack that each creatures makes against him each round. In addition, his land speed bonus improves to \plus30.

\subsection{Ki Master} 
Some fighters seek more than mere mental discipline. Instead, these sage warriors, commonly called monks, seek inner perfection. In doing so, they gain access to a powerful energy, called \ki, that flows through all living creatures. By tapping into this power, they can perform superhuman feats.

These benefits replace all of the fighter's combat discipline class features, and two of his weapon groups.

\cf{Ftr}{\Ki Fists} The fighter's unarmed strikes deal lethal damage, and deal damage as if the fighter was two size categories larger (d6 damage for a Medium creature). 

\cf{Ftr}{\Ki Power} At 5th level, the fighter's weapons become empowered by his \ki energy. He treats all weapons he wields as if they were \plus1 weapons, gaining a \plus1 enhancement bonus to attack and damage. This bonus increases by 1 at 8th level and every 3 levels thereafter.
