\section{Aberrations}

    \begin{monsection}{Aboleth}{12}[3]
        \begin{spellcontent}
            \begin{spelltargetinginfo}
                \pari \mb{Senses} Darkvision 50 ft., Awareness \plus17
                \pari \mb{Movement} 10 ft.\ land speed, 60 ft.\ swim speed
                \pari \mb{Size} Huge; \mb{Reach} 20 ft.
            \end{spelltargetinginfo}
            \begin{spelleffects}
                \pari \mb{HP} 504; \mb{Defenses} AD 33, Fort 42, Ref 22, Ment 40
                \pari \mb{Attacks} Tentacle \plus18 (2d10\plus10); \mb{Strikes} 3
                \pari \mb{Special Attacks} Slime \plus15; \mb{Power} 15
                \pari \mb{Alacrity} Two actions per round
                \pari \mb{Attributes} Str 15, Dex 0, Con 15, Int 12, Per 8, Wil 16
            \end{spelleffects}
        \end{spellcontent}
        \begin{spellfooter}
            \pari \mb{Level} Aberration 12 [Adept, Behemoth]
            \pari \mb{Encounter} Underground; solitary
            \pari \mb{Abilities} Natural armor \plus4
        \end{spellfooter}
    \end{monsection}
    \parhead{Traits} Armored Carapace, Increased Size, Magical Ability, Magical Ability, Magical Strike, Myriad Magical Abilities, Titanic Defenses.

    \parhead{Slime}[\glossterm{Magical}] As an immediate action, when an aboleth strikes a creature with a tentacle, it can secrete slime onto the struck creature.
    If it does, it makes a \glossterm{Power} vs. Fortitude attack against the struck creature.
    Success means that target is \sickened for 2 rounds.
    Critical success means the target is also \nauseated for 1 round.
    \parhead{Psionics}[\glossterm{Magical}] As a standard action, an aboleth can concentrate to cause the effects of one of the following spells.
    Its spellpower is equal to its power.
    \begin{itemize}
        \item The \textit{create image} spell, augmented to affect all senses.
        \item The \textit{dominate person} spell, augmented to affect creatures of any type.
        \item The \textit{phantasmal killer} spell.
        \item The \textit{invisibility} spell, augmented to affect Huge creatures.
        \item The \textit{invulnerability} spell.
    \end{itemize}
