\section{Outsiders}

    \begin{monsection}{Arrowhawk}{3}
        \begin{spellcontent}
            \begin{spelltargetinginfo}
                \pari \mb{Senses} Awareness \plus8, darkvision 50 ft.
                \pari \mb{Movement} 60 ft.\ fly speed (perfect)
                \pari \mb{Size} Small; \mb{Reach} 5 ft.
            \end{spelltargetinginfo}
            \begin{spelleffects}
                \pari \mb{HP} 24; \mb{Defenses} AD 21, Fort 12, Ref 21, Wil 16
                \pari \mb{Attacks} Bite \plus11 (1d6\plus2, 1d6 electricity); \mb{Strikes} 1
                \pari \mb{Special Attacks} Electricity ray \plus5 (4d6 electricity)
                \pari \mb{Attributes} Str 2, Dex 6, Con 2, Int 0, Per 6, Wil 2
            \end{spelleffects}
        \end{spellcontent}
        \begin{spellfooter}
            \pari \mb{Languages} Auran
            \pari \mb{Levels} Slayer 3
            \pari \mb{Encounter} Plane of Air; solitary or clutch (2-4)
            \pari \mb{Abilities} Natural armor \plus2, Flight, Damaging Ray, Natural Energy
        \end{spellfooter}
    \end{monsection}

    \parhead{Electricity Ray} As a standard action, an arrowhawk can fire a ray of electricity from its tail at a creature or object within \rngmed range.
    It makes a \plus5 vs. Reflex attack against the target.
    Success means the target takes 3d10 electricity damage.
    Critical success deals double damage.
    Failure deals half damage.

    \begin{monsection}[Bebelith]{Demon}{11}
        \begin{spellcontent}
            \begin{spelltargetinginfo}
                \pari \mb{Senses} Awareness \plus{16}, darkvision 50 ft.
                \par Telepathy (100 ft. range, willing only)
                \pari \mb{Movement} 50 ft.\ land speed, 25 ft. climb speed
                \pari \mb{Size} Huge; \mb{Reach} 10 ft.
            \end{spelltargetinginfo}
            \begin{spelleffects}
                \pari \mb{HP} 237; \mb{Defenses} AD 27, Fort 39, Ref 21, Ment 22
                \pari \mb{Attacks} Bite +16 (2d10+7, poison); \mb{Strikes} 3
                \par Claw +16 (2d8+7, rend)
                \pari \mb{Special Attacks} Poison \plus14 vs Fort (6d6, 13d6) [Immediate]
                \par Rend +14 vs. Fort (7d10) [Immediate]
                \par Web +14 vs. Ref (immobilized 2 rounds) [Swift]
                \pari \mb{Attributes} Str 14, Dex 4, Con 13, Int 1, Per 5, Wil 2
            \end{spelleffects}
        \end{spellcontent}
        \begin{spellfooter}
            \pari \mb{Languages} Abyssal (hearing only)
            \pari \mb{Levels} Behemoth 5, Slayer 6
            \pari \mb{Encounter} The Abyss; solitary
            \pari \mb{Abilities} Natural armor \plus4, Behemoth Size, Natural Venom, Rend, Spit Web, Brute Force, Durable
        \end{spellfooter}
    \end{monsection}

    \parhead{Poison} If a bebelith damages a target with its bite attack, it can inject poison as an immediate action.
    The poison's potency is 14.
    The primary effect makes the target \sickened and deals 7d6 poison damage.
    The terminal effect makes the target \nauseated for 1 round and deals 14d8 poison damage.

    \parhead{Rend} If a bebelith hits the same target with two claw attacks in the same round, it can attempt to rend as an immediate action.
    It makes a \plus14 vs. Fortitude attack against the target.
    Success deals 7d10 bludgeoning damage.

    \parhead{Web} As a swift action, a bebelith can spit a web at a creature or object within \rngmed range.
    It makes a \plus14 vs. Reflex attack against the target.
    Success means the target is \immobilized for 2 rounds.

    As a standard action, the target can make a grapple attack or Escape Artist check to escape the webbing.
    The DR to escape is 21.
