\chapter{Traits}\label{Traits}

\section{Trait Limitations}
    The number of traits you have from a particular class cannot exceed the levels you have in that class.

\section{Trait Types}

    \parhead{Active} An active trait grants a monster a new active ability to use. Most active traits can be used as a standard action. A monster with too many active traits will have difficulty using them all in combat, and may feel unfocused. A monster without active traits may feel boring and repetitive to fight.

    \parhead{Augment} An augment trait improves a monster's existing attacks or special abilities. Most augment traits are passive abilities or are activated as an immediate action. A monster with too many augment traits may be overly specialized in a single ability, and not be interesting to fight. A monster without augment traits may be less effective with its attacks and special abilities.

    \parhead{Body} A body trait alters a monster's body. Most body traits are passive abilities. A monster with too many body traits may be bizarre or confusing to fight.

    \parhead{Defense} A defense trait improves a monster's defenses or heals it. Most defensive traits are passive abilities that do not require activation. A monster with too many defense traits may take a long time to kill without posing any challenge. A monster without defense traits may die too quickly to use its abilities.

    \parhead{Mobility} A mobility trait improves a monster's ability to move around a battlefield. A monster with too many mobility traits may have redundant ways to move. A monster without mobility traits may not be able to effectively attack mobile characters.

\section{Trait Tables}
    \onecolumn

    \begin{longtabuwrapper}
        \begin{longtabu}{>{\lcol}p{10em} >{\lcol}p{15em} >{\lcol}X >{\lcol}p{8em} >{\lcol}p{3em}}
            \lcaption{Traits}\\
            \tb{General Traits}\label{General Traits} & \tb{Prerequisites} & \tb{Benefit} & \tb{Trait Types} & \tb{Page} \\
            \featref{Ability Focus} & 6th level & Gain \plus4 with ability & Augment & \featpref{Ability Focus} \\
            \featref{Incorporeal} & 4th level & Become incorporeal & Body & \featpref{Incorporeal} \\
            \featref{Increased Size} & 6th level & Increase size with level & Body & \featpref{Increased Size} \\
            \featref{Innate Magic} & 6th level & Continuously gain spell effect & Augment & \featpref{Innate Magic} \\
            \featref{Manylegged} & \tdash & Gain many legs & Body & \featpref{Manylegged} \\
            \featref{Natural Grab} & Natural weapon, combat prowess 4 & Natural weapon can grapple & Augment & \featpref{Natural Grab} \\
            \featref{Natural Shove} & Natural weapon & Natural weapon can shove & Augment & \featpref{Natural Shove} \\
            \featref{Natural Trip} & Natural weapon & Natural weapon can trip & Augment & \featpref{Natural Grab} \\
            \featref{Natural Venom} & Natural weapon & Natural weapon becomes poisonous & Augment & \featpref{Natural Venom} \\
            \featref{Resist Magic} & 8th level & Gain spell resistance & Defense & \featpref{Resist Magic} \\
            \featref{Spellfeeder} & 8th level & Heal from spells cast nearby & Defense & \featpref{Spellfeeder} \\
            \featref{Spit Web} & \tdash & Gain ability to spit immobilizing webs & Active & \featpref{Spit Web} \\
            \featref{Superior Senses} & \tdash & Gain extraordinary sense & Augment & \featpref{Superior Senses} \\

            \midrule
            \tb{Adept Traits}\label{Adept Traits} & \tb{Prerequisites} & \tb{Benefit} & \tb{Trait Types} & \tb{Page} \\
            \featref{Draining Touch} & \tdash & Gain ability to deal damage with a touch & Active & \featpref{Draining Touch} \\
            \featref{Humanoid Form} & 4th level & Gain ability to change into humanoid form & Body & \featpref{Humanoid Form} \\
            \featref{Magical Ability} & \tdash & Gain ability to mimic a spell's effects & Active & \featpref{Magical Ability} \\
            \featref{Magical Retribution} & 8th level & Deal damage to attackers & Defense & \featpref{Magical Retribution} \\
            \featref{Magical Strike} & 4th level, Natural weapon & Can make an attack that imposes spell's effects & \tdash & \featpref{Magical Strike} \\
            \featref{Natural Magic} & 8th level, Natural weapon & Natural weapon causes a spell's effects & Augment & \featpref{Natural Magic} \\
            \featref{Myriad Magical Abilities} & 6th level & Gain many abilities that mimic spell effects & Active & \featpref{Myriad Magical Abilities} \\
            \featref{Petrifying Gaze} & 4th level & Gain gaze attack & Active & \featpref{Petrifying Gaze} \\
            \featref{Spellcaster} & \tdash & Gain ability to cast spells & Active & \featpref{Spellcaster} \\

            \midrule
            \tb{Behemoth Traits}\label{Behemoth Traits} & \tb{Prerequisites} & \tb{Benefit} & \tb{Trait Types} & \tb{Page} \\
            \featref{Defensive} & \tdash & Increase defense & \tdash & \featpref{Defensive} \\
            \featref{Resist Damage} & \tdash & Gain damage reduction & Defense & \featpref{Resist Damage} \\
            \featref{Supreme Durability} & 6th level & Gain bonus hit points & Augment & \featpref{Supreme Durability} \\
            \featref{Retributive Spikes} & 4th level & Deal damage to melee attackers & Defense & \featpref{Retributive Spikes} \\
            \featref{Titanic Body} & 10th level & Increase size by one size category & Body & \featpref{Titanic Body} \\

            \midrule
            \tb{Slayer Traits}\label{Slayer Traits} & \tb{Prerequisites} & \tb{Benefit} & \tb{Trait Types} & \tb{Page} \\
            \featref{Boulder Toss} & Slayer 4, Strength 5 & Throw objects incredible distances & Active & \featpref{Boulder Toss} \\
            \featref{Breath Weapon: Cone} & \tdash & Gain cone-shaped breath weapon & Active & \featpref{Breath Weapon: Cone} \\
            \featref{Breath Weapon: Line} & \tdash & Gain line-shaped breath weapon & Active & \featpref{Breath Weapon: Line} \\
            \featref{Brute Force} & \tdash & Increase damage of physical attacks & Augment & \featpref{Brute Force} \\
            \featref{Burrower} & \tdash & Gain burrow speed & Mobility & \featpref{Burrower} \\
            \featref{Damaging Aura} & \tdash & Gain aura that damages nearby foes & Augment & \featpref{Damaging Aura} \\
            \featref{Damaging Ray} & \tdash & Gain ability to fire damaging ray & Active & \featpref{Damaging Ray} \\
            \featref{Multistrike} & 6th level, three hands & Attack all threatened foes & Active & \featpref{Multistrike} \\
            \featref{Natural Energy} & 4th level, Natural weapon & Natural weapon deals energy damage & Augment & \featpref{Natural Energy} \\
            \featref{Rend} & Two strikes, Strength 5, two natural weapons & Deal bonus damage after hitting with two weapons & Augment & \featpref{Rend} \\
        \end{longtabu}
    \end{longtabuwrapper}

    \twocolumn

\section{Trait Descriptions}

    \feat{Ability Focus}{Augment}
    \featpre 6th level.
    \parhead{Choice} Choose an ability you have with an attack or effect that depends on your \glossterm{power}.
    \featben You gain a \plus4 bonus to your power for the purpose of your chosen ability.

    \feat{Boulder Toss}{Active, Slayer}
    \featpre 4th level, slayer template, Strength 5.
    \featben You can throw improvised objects, such as rocks, with great accuracy and power.
    You gain proficiency with improvised thrown weapons, and your \glossterm{range increment} with all thrown weapons becomes 50 feet.

    At 10 Strength, and every 5 Strength thereafter, your range increment with thrown weapons increases by 50 feet.

    \feat{Breath Weapon: Cone}{Active, Magical, Slayer}
    \featpre Slayer template.
    \parhead{Choice} Choose a damage type when you gain this trait.
    \featben You gain a breath weapon attack.
    As a standard action, you can make a \glossterm{Power} vs. Reflex attack against all creatures within a \areamed cone-shaped burst.
    Success deals 1d8 damage of your chosen damage type per two power.
    Critical success deals double damage.
    Failure deals half damage.
    After using this ability, you cannot use it again for 1d4 rounds.

    At 8th level, the area increases to a \arealarge cone.
    At 16th level, the damage increases to 1d10 damage per two power.
    At 20th level, the area increases to a \areahuge cone.

    \feat{Breath Weapon: Line}{Active, Magical, Slayer}
    \featpre Slayer template.
    \parhead{Choice} Choose a damage type when you gain this trait.
    \featben You gain a breath weapon attack.
    As a standard action, you can make a \glossterm{Power} vs. Reflex attack against all creatures within a \arealarge, 5 ft.\ wide line-shaped burst.
    Success deals 1d8 damage of your chosen damage type per two power.
    Critical success deals double damage.
    Failure deals half damage.
    After using this ability, you cannot use it again for 1d4 rounds.

    At 6th level, the area increases to \arealarge, 10 ft.\ wide line.
    At 10th level, the area increases to \areahuge, 10 ft.\ wide line.
    At 14th level, the area increases to \areahuge, 20 ft.\ wide line.
    At 18th level, the damage increases to 1d10 damage per two levels.

    \feat{Brute Force}{Augment, Slayer}
    \featpre Slayer template.
    \featben You gain a \plus2 bonus to damage with physical attacks.

    At 6th level, and every 4 levels thereafter, this bonus increases by 1.

    \feat{Burrower}{Mobility, Slayer}
    \featpre Slayer template.
    \featben You gain a burrow speed equal to half your land speed.
    You can only burrow through loose earth and dirt, and you cannot construct usable tunnels while burrowing.

    At 6th level, and every 4 levels thereafter, you can gain one of the following abilities.
    \begin{itemize}
        \item Your burrow speed because equal to your land speed.
        \item You gain the ability to construct usable tunnels while burrowing.
            You burrow at half speed while doing so.
        \item You gain the ability to burrow through solid rock and objects of similar durability.
        \item If you have the ability to burrow through solid rock, you can burrow through iron and objects of similar durability.
    \end{itemize}

    \feat{Damaging Aura}{Augment, Magical}
    \featpre Slayer template.
    \parhead{Choice} Choose a damage type.
    \featben At the end of every round, enemies within a 5-ft.\ radius emanation from you take damage equal to your \glossterm{power} of the chosen damage type.
    At 6th level, the area increases to a \areasmall radius emanation.
    At 12th level, the area increases to a \areamed radius emanation.
    At 18th level, the area increases to a \arealarge radius emanation.

    \feat{Damaging Ray}{Active, Magical, Slayer}
    \featpre Slayer template.
    \parhead{Choice} Choose a damage type when you gain this trait.
    \featben As a standard action, you can fire an ray at a creature or object within \rngclose range.
    You make a \glossterm{Power} vs. Reflex attack against the target.
    Success deals 1d6 damage of your chosen damage type per power.
    Critical success deals double damage.
    Failure deals half damage.

    \feat{Defensive}{Behemoth, Defense}
    \parhead{Choice} Choose a defense: Armor, Fortitude, Reflex, or Mental.
    \featben If you chose Armor defense, you gain a \plus1 bonus to Armor defense.
    Otherwise, you gain a \plus2 bonus to your chosen defense.

    At 6th level, and every 4 levels thereafter, this defense bonus increases by \plus1 if you chose Armor defense, or by \plus2 otherwise.

    \parhead{Special} You may take this trait multiple times.
    Each time, you choose a new defense.

    \feat{Draining Touch}{Active, Adept, Magical}
    \featpre Adept template.
    \featben If you touch a creature, it takes 1d6 life damage \add 1 per two \glossterm{power}.
    You heal hit points equal to half the damage dealt in this way.
    In addition, you become proficient with touching creatures, gaining the normal \plus4 accuracy bonus for proficiency.

    \feat{Incorporeal}{Body}
    \featpre 4th level.
    \featben You become \glossterm{incorporeal}.
    You no longer have a land speed, and gain a fly speed with perfect maneuverability.
    Your fly speed is equal to what your land speed would be.

    \feat{Increased Size}{Body}
    \featpre 4th level.
    \featben Your size increases by one size category.
    At 12th level, and every eight levels thereafter, you can increase your size by one additional size category.
    This effect stacks with most other effects that increase your size.

    If you have the behemoth template, you can also increase your size by one additional size category at 8th and 16th level.

    \feat{Innate Magic}{Augment, Magical}
    \featpre 4th level.
    \parhead{Choice} Choose a spell
    \featben You continuously gain the benefit of a spell effect.
    The spell must be a \glossterm{targeted spell} with a duration of \durshort or longer, and its level must not exceed half your level.
    If the effect is discharged or otherwise ended, it automatically resumes its effect after 5 minutes.
    This is a supernatural ability, and cannot be dispelled.

    \feat{Humanoid Form}{Adept, Body, Magical}
    \featpre 4th level, adept template.
    \featben As a standard action, you can spend an adept point to change your shape.
    When you change your shape, you can assume a humanoid form.
    You choose the appearance and nature of your humanoid form when you gain this trait, and are not able to change it freely.
    Your humanoid form resembles your choice of the common humanoid races: dwarf, elf, gnome, half-elf, half-orc, halfling, or human.
    Your form has the size and shape of the race you mimic, but you do not gain any abilities of the chosen race.
    While in humanoid form, you cannot use any special abilities that require physical features of your normal form.
    This includes natural weapons, poisons, Armor defense bonuses, and similar abilities.

    In addition to assuming the humanoid form completely, you can also assume a hybrid form.
    In a hybrid form, you retain physical features of your normal form, allowing you the full use of your special abilities.
    However, a DR 20 Awareness check can identify unusual physical features that reveal your inhuman nature.

    In both humanoid and hybrid forms, you can wield weapons and wear equipment like other humanoids.
    Changing your form with this ability is not a magical effect, does not have a duration, and cannot be dispelled.
    You revert to your true form when killed.

    \feat{Magical Ability}{Active, Adept, Magical}
    \featpre Adept template.
    \parhead{Choice} Choose a spell.
    You may also choose any number of augments to apply to the spell, increasing its level appropriately.
    The spell's total level must be no greater than half your level (minimum 1).
    \featben As a standard action, you can concentrate to cause your chosen spell's effect, including any augments.
    Its spellpower is equal to your \glossterm{power}.

    When using this ability, you cannot apply additional augments to the spell, apply metamagic feats, or use similar abilities, since you are not actually casting the spell.

    \parhead{Special} You may take this trait multiple times.
    Each time, you choose a new spell.

    \feat{Magical Retribution}{Adept, Defense, Magical}
    \featpre 8th level, adept template.
    \parhead{Choice} Choose a damage type.
    \featben Whenever a creature within \rngmed range of you attacks you, the attacking creature takes 1d6 damage per two \glossterm{power} of your chosen damage type.
    Any individual creature can only take this damage once per round.
    This is a supernatural ability.

    \feat{Magical Strike}{Active, Adept, Magical}
    \featpre 4th level, adept template.
    \parhead{Choice} Choose a \glossterm{targeted spell} with a level no higher than half your level.
    The spell cannot deal damage.
    \featben As a standard action, you can make a single \glossterm{strike}.
    If you hit with the strike, you deal damage normally.
    In addition, the struck target is affected by your chosen spell.
    If you miss with the strike, the target suffers no effects.

    \feat{Manylegged}{Body}
    \featben You have an extraordinary number of legs.
    This grants you the benefits given below.

    \feat{Multistrike}{Active, Slayer}
    \featpres 6th level, slayer template, at least three hands or hand-like appendages.
    \featben As a standard action, you can make one \glossterm{strike} against all creatures you threaten.
    For each strike per round you can make beyond two, you can make one additional strike against all creatures in this way, up to a maximum number of strikes equal to the number of hand-like appendages you have.

    \feat{Myriad Magical Abilities}{Active, Adept, Magical}
    \featpre 6th level, adept template.
    \parhead{Choice} Choose up to three spells.
    You may also choose any number of augments to apply to the spell, increasing its level appropriately.
    The total level of each spell individually must be no greater than half your level, \sub 2.
    \featben As a standarda ction, you can concentrate to cause the effect of any one of your chosen spells, as if you had cast the spell.
    Its spellpower is equal to your \glossterm{power}.

    When using this ability, you cannot apply additional augments to the spell, apply metamagic feats, or use similar abilities, since you are not actually casting the spell.

    \parhead{Special} You may take this trait multiple times.
    Each time, you choose up to three new spells.

    \feat{Natural Grab}{Augment}
    \featpre At least one \glossterm{natural weapon}, combat prowess 4.
    \parhead{Choice}
    \featben Choose one of your natural weapons.
    When you make a successful attack with that weapon, you can attempt to \glossterm{grapple} the struck creature as an immediate action.

    \feat{Natural Energy}{Augment, Slayer}
    \featpre Slayer, Natural weapon, 4th level.
    \parhead{Choice} Choose a natural weapon you have, and a type of \glossterm{energy damage}.
    \featben Your chosen natural weapon deals 1d6 extra damage of your chosen energy type on every successful attack.

    At 8th level, and every 4 levels thereafter, this damage increases by one dice increment.

    \feat{Natural Magic}{Adept, Augment, Magical}
    \featpres Natural weapon, adept level 8.
    \parhead{Choice} Choose one of your natural weapons, and a spell.
    The spell must be a \glossterm{targeted spell}.
    You may also choose any number of augments to apply to the spell, increasing its level appropriately.
    The spell's total level must be no greater than half your level, \sub 3.
    \featben When you make a successful attack with your chosen weapon, you can inflict your chosen spell's effects on the struck creature as an immediate action.
    Its spellpower is equal to your \glossterm{power}.

    When using this ability, you cannot apply additional augments to the spell, apply metamagic feats, or use similar abilities, since you are not actually casting the spell.

    \parhead{Special} You may take this trait multiple times.
    Each time, you choose a new spell.

    \feat{Natural Shove}{Augment}
    \featpre At least one \glossterm{natural weapon}.
    \parhead{Choice} Choose one of your natural weapons.
    \featben When you make a successful attack with your chosen weapon, you can attempt to \glossterm{shove} the struck creature as an immediate action.

    \feat{Natural Trip}{Augment}
    \featpre At least one \glossterm{natural weapon}.
    \parhead{Choice} Choose one of your natural weapons.
    \featben When you make a successful attack with your chosen weapon, you can attempt to \glossterm{trip} the struck creature as an immediate action.

    \feat{Natural Venom}{Augment}
    \featpre At least one \glossterm{natural weapon}.
    \parhead{Choice} Choose one of your natural weapons, and a poison effect from the options below.
    Some poison effects have level prerequisites, as described in the name of the ability.
    \begin{itemize}
        \item Damaging Poison: Primary effect 1d6 poison damage per two potency, terminal effect 1d8 poison damage per potency.
        \item Sickening Poison: Primary effect \sickened, terminal effect \nauseated for 2 rounds.
        \item Paralyzing Poison (6th): Primary effect \immobilized, terminal effect \paralyzed for 2 rounds.
    \end{itemize}
    \featben Whenever you deal physical damage with your chosen weapon, you can inject your chosen poison into the damaged creature as an immediate action.
    The potency of the poison is equal to your \glossterm{power}.

    At 6th level, and every 6 levels thereafter, this ability improves, and you gain one of the following benefits.
    \begin{itemize}
        \item Choose a different natural weapon and apply a single type of poison to it.
        \item Choose a natural weapon that already has poison and apply a new type of poison to it.
            If one natural weapon has multiple poisons, you can inject all poisons as part of the same immediate action.
        \item Choose a natural weapon that already has poison.
            You do not need to spend an immediate action to apply poison with that weapon.
            You still cannot poison more than one creature per round in this way.
    \end{itemize}

    \feat{Petrifying Gaze}{Active, Adept, Magical}
    \featpre Adept 6.
    \featben You gain a gaze attack, allowing you to harm a creature by gazing at it.
    As a standard action, you can focus your gaze on a creature within \rngclose range.
    You must be able to see the target, but you do not need \glossterm{line of effect} to it.
    You make a \glossterm{Power} vs. Fortitude attack against the target.
    Success means it is \slowed for 2 rounds.
    Critical success means it is \petrified.

    \feat{Rend}{Augment, Slayer}
    \featpres Slayer template, two strikes, Strength 5, two natural weapons.
    \featben If you hit the same target twice in the same round with two different natural weapons, you can rend it as an immediate action.
    If you do, you make a \glossterm{Power} vs. Fortitude attack against the target.
    Success deals 1d10 slashing damage per two power.
    Failure deals half damage.

    \feat{Resist Damage}{Defense}
    \parhead{Choice} Choose a type of damage to resist: physical damage or energy damage.
    \featben You gain damage reduction against your chosen damage type equal to your \glossterm{power}.

    At 10th level, this damage reduction becomes equal to twice your \glossterm{power}.

    \feat{Resist Magic}{Defense}
    \featpre 8th level.
    \featben You gain \glossterm{magic resistance} equal to 10 \add your \glossterm{power}.

    \feat{Retributive Spikes}{Behemoth, Defense}
    \featpre 4th level, behemoth template.
    \featben Whenever a creature attacks you with a physical melee attack, the attacking creature takes 1d6 piercing damage per two \glossterm{power}.
    Any individual creature can only take this damage once per round.

    \feat{Spellcaster}{Active, Adept}
    \featpre Adept template.
    \parhead{Choice} Choose a type of spellcasting: arcane, divine, or nature.
    \featben You can cast spells from your chosen magic source.
    You know a number of spells equal to half your adept level \add 1.
    The maximum spell level you can learn and cast is equal to half your level (minimum 1).
    These spells can be of any level.
    You do not have spell slots, and may freely cast any spells you know.
    Your spellpower is equal to your \glossterm{power}.
    \parhead{Special} You may take this trait multiple times.
    Each time, you choose a new type of spellcasting.
    You track your spells known separately for each type of spellcasting you have.

    \feat{Spellfeeder}{Defense, Magical}
    \featpre 8th level.
    \featben Whenever a spell is cast within \rngmed range of you, you heal hit points equal to your \glossterm{power}.
    In addition, whenever you resist a spell with spell resistance, you regain hit points equal to twice your \glossterm{power}.
    This ability does not grant you spell resistance if you do not have it.

    \feat{Spit Web}{Active}
    \featben As a standard action, you can spit a web (or similar substance) at a creature or object within \rngmed range.
    You make a \glossterm{Power} vs. Reflex attack against the target.
    Success means the target is \immobilized for 2 rounds.
    As a standard action, the target can make a grapple attack or Escape Artist check to escape the webbing.
    The DR to escape is equal to 10 \add your power.

    If you are 8th level or higher, you can use this ability as a swift action.

    \feat{Superior Senses}{Augment}
    \featpre 6th level.
    \featben You gain up to two of the following senses.
    \begin{itemize}
        \item Blindsense (200 ft.\ range)
        \item Darkvision (200 ft.\ range)
        \item Lifesense (200 ft.\ range)
        \item Tremorsense (200 ft.\ range)
        \item Lifesight (50 ft.\ range)
        \item Tremorsight (50 ft.\ range)
        \item Truesight (50 ft.\ range)
    \end{itemize}
    \parhead{Special} You may take this trait multiple times.
    Each time, you choose a new sense.

    \feat{Supreme Durability}{Augment, Behemoth}
    \featpre 6th level.
    \featben You gain bonus hit points equal to five times your \glossterm{power}.

    \feat{Titanic Body}{Behemoth, Body}
    \featpre 10th level, behemoth template.
    \featben Your size increases by one size category.
    This stacks with other effects that increase your size.
