\section{Undead}

    \begin{monsection}{Allip}{4}
        \begin{spellcontent}
            \begin{spelltargetinginfo}
                \pari \mb{Senses} Awareness \plus 1, darkvision 50 ft.
                \pari \mb{Movement} 30 ft.\ fly speed (perfect)
                \pari \mb{Size} Medium; \mb{Reach} 5 ft.
            \end{spelltargetinginfo}
            \begin{spelleffects}
                \pari \mb{HP} hp; \mb{Defenses}
                \pari \mb{Special Defenses} Incorporeal
                \pari \mb{Attacks} Touch \plus9 (1d6+3, draining); \mb{Strikes} 1
                \pari \mb{Special Attack} Babble \plus4
                \pari \mb{Attributes} Str \tdash, Dex 5, Con \tdash, Int 1, Per 2, Wil 7
            \end{spelleffects}
        \end{spellcontent}
        \begin{spellfooter}
            \pari \mb{Levels} Adept 4
            \pari \mb{Environment} Any
            \pari \mb{Organization} Solitary
            \pari \mb{Traits} [3] Incorporeal, Draining Touch, Magical Ability
        \end{spellfooter}
    \end{monsection}
    \parhead{Adept Points} An allip has three adept points.
    They return one hour after being spent.
    \parhead{Babble} As a standard action, an allip can spend an adept point to babble incoherently.
    It makes a Level vs. Mental attack against all creatures within an \areamed radius of it.
    Success against a target means it is \disoriented for 2 rounds.
    Critical success means it is \confused for 2 rounds instead.
    This is a \glossterm{Auditory}, \glossterm{Compulsion}, \glossterm{Mind} effect. 

    \parhead{Draining Touch} If an allip touches a creature, the touched creature takes 1d6 damage \add 1 per two Willpower.
    The allip heals hit points equal to half the damage dealt in this way.
    In addition, it is proficient with its touch attack, gaining the normal \plus4 accuracy bonus for proficiency.

    \parhead{Incorporeal} Allips are \glossterm{incorporeal}.
