\chapter{Feats}\label{Feats}

Feats are special abilities that every character has.
Feats can be used to specialize your character particular area, to grant your character new abilities, or to change the way your character does certain things.

\section{Gaining Feats}
Your character gains a feat every odd level: 1st, 3rd, 5th, and so on.
Classes can sometimes grant bonus feats as well, which are in addition to these feats which every character gets.

\subsection{Prerequisites}
Some feats have prerequisites.
Your character must have the indicated attribute score, ability, feat, skill, combat prowess, or other quality designated in order to select or use that feat.
A character can gain a feat at the same level at which he or she gains the prerequisite.

A character can't use a feat if he or she has lost a prerequisite.

\section{Types Of Feats}
Some feats are general, meaning that no special rules govern
them as a group.
Others belong to particular categories.
The categories are given below:
\begin{itemize}
    \item Bloodline feats allow a character to tap into latent magical abilities in their blood.
    \item Item creation feats allow spellcasters to create magic items of various kinds.
    \item Magic feats affect the way a character casts spells or uses magic.
    \item Metamagic feats let spellcasters cast spells with greater effect, albeit as if the spells were a higher level than they actually are.
    \item Performance feats allow a character to create magical effects through skilled performances.
    \item Skill feats affect a character's mastery of his or her skills.
    \item Spellgift feats grant a non-spellcasting character a limited ability to use spells.
    %\item Surge feats allow a character to temporarily gain supernatural abilities.
    \item Combat feats affect a character's prowess in combat.
    \item Combat maneuver feats are combat feats which grant a character particular talent with specific maneuvers.
    \item Combat strike feats allow a character to impose special effects when they strike a foe.
    \item Combat style feats change the way a character fights.
        A character can only use one style at a time.
    %\item Bane feats provide characters with bonuses against specific kinds of foes.
        %You can only gain the benefits of a single Bane feat against a particular creature.
        %If more than such feat applies, choose which one to gain the benefits from.
\end{itemize}

\subsection{Bloodline Feats}

Some characters have traces of monstrous blood running in their veins.
Most of those will never understand the full potential of their unusual heritage.
Bloodline feats allow characters to explore those posibilities by gaining abilites related to their ancestry.
Each bloodline feat belongs to a specific type of monster, such as ``dragon''.
Some bloodline feats have stronger effects if you have more feats from that heritage.

\subsection{Class Feats}
Class feats improve a character's class abilities, or grant new special abilities related to a class.

\subsection{Item Creation Feats}
An item creation feats let a spellcaster create magic items.

\subsection{Metamagic Feats}
Metamagic feats change properties of spells you cast.
Spells modified by a metamagic feat are treated as being higher level than normal.
They require higher level spell slots to cast, and cannot be cast unless you have access to normal spells of the spell's modified level.
All effects dependent on spell level are calculated according to the spell's modified level.
Metamagic feats cannot be used to enhance spell-like abilities, magic item abilities, or other effects which are not spells.

\parhead{Combining Metamagic Feats}
You can apply multiple metamagic feats to a single spell.
Changes to the spell's level are cumulative.
You can't apply the same metamagic feat more than once to a single spell.

\parhead{Magic Items and Metamagic Spells}
With the right item creation feat, you can store a metamagic version of a spell in a scroll, potion, or wand.
Level limits for potions and wands apply to the spell's higher spell level (after the application of the metamagic feat).
A character doesn't need the metamagic feat to activate an item storing a metamagic version of a spell.

\subsection{Performance Feats}
Performance feats allow a character to use the Perform skill to create magical effects.
All Performance feats are also Skill feats.

Each performance has an effect when it is used.
Most performances continue their effects if you continue performing.

All performance feats are \glossterm{Auditory}, \glossterm{Speech}, or \glossterm{Visual} effects, depending on the nature of the performance used to activate the feat.

You can use any combination of performance feats you possess a number of times per day equal to twice the number of performance feats you have.
If you fail the Perform check, the use of the ability is wasted.

\subsection{Racial Feats}\label{Racial Feats}
Racial feats improve a character's racial abilities or grant new abilities unique to the character's race.

\subsection{Skill Feats}
Skill feats improve a character's skills or grant them new abilities related to the skills they already have.

\subsection{Spellgift Feats}
Spellgift feats grant limited spell-like abilities.
They may represent the fruits of eldritch pacts, unusual heritage, the exposure to powerful magic, or a particular expression of innate magical ability.
You can use any combination of spell-like abilities gained from spellgift feats a number of times per day equal to the number of spellgift feats you possess.
The number of spellgift feats you have, plus the levels you have in spellcasting classes, cannot exceed your Willpower.

If your Willpower is lowered below this value, or you gain additional levels in spellcasting classes, you are treated as not possessing the spellgift feats you do not qualify for, as normal for feat prerequisites.
You lose access to the spellgift feats in order by when you acquired them, starting with the most recent feat.

\subsection{Combat Feats}
Combat feats affect a character's combat abilities.
Many combat feats require a minimum combat prowess to acquire.

\subsection{Combat Maneuver Feats}
Combat maneuver feats are combat feats which specifically deal with maneuvers -- pushing opponents around, grappling them, tripping them, or other unusual attacks.
Any feat designated as a combat maneuver feat is also considered a combat feat.

\subsection{Style Feats}
Style feats grant a character the ability to fight or cast spells in a particular style, granting them bonuses while in that style.
A character can only be in one style at once.
Initiating a style or changing to a different style is a swift action, but a style can be stopped as a free action.

Most style feats have requirements.
If a style requires specific equipment, such as a melee weapon, you must meet the requirements to activate the style.
If you fail to meet a style's requirements during a round, you leave the style at the end of the round.

\subsection{Strike Feats}\label{Strike Feats}
Strike feats grant a character the ability to make unusual physical attacks.
Most strike feats replace a certain number of \glossterm{strikes} you would normally make.
Usually, this means strike feats are used as part of a standard attack action.
Some strike feats instead require a specific action to use.

\section{Feat Tables}
\onecolumn

\subsection{General Feats}
% Feat names must follow ``have'', ``are (a)'', or ``can''.
\begin{longtabuwrapper}
    \begin{longtabu}{>{\lcol}p{10em} >{\lcol}p{15em} >{\lcol}X >{\lcol}p{8em} >{\lcol}p{3em}}
        \tb{General Feats} & \tb{Prerequisites} & \tb{Benefit} & \tb{Feat Types} & \tb{Page} \\
        \featref{Attribute Training} & \x & Gain \plus1 to an attribute           & \x & \featpref{Attribute Training} \\
        \featref{Arcane Spellgift} & 2nd level, Wil 1 & Gain arcane spell as spell-like ability & Spellgift & \featpref{Arcane Spellgift} \\
        %\featref{Destructive} & Str 3 & Gain \plus5 to breaking things            & \x & \featpref{Destructive} \\
        \featref{Diehard} & Con 10 & Halve penalties from critical damage         & \x & \featpref{Diehard} \\
        \featref{Divine Spellgift} & 2nd level, Wil 1 & Gain divine spell as spell-like ability & Spellgift & \featpref{Divine Spellgift} \\
        \featref{Fearless} & Wil 10 & Immune to fear effects                      & \x & \featpref{Fearless} \\
        \featref{Fortune's Friend} & Legend point & Use legend points to be lucky  & \x & \featpref{Fortune's Friend} \\
        \featref{Iron Will} & Int 3 or Wil 3 & \plus2 to Mental defense           & \x & \featpref{Iron Will} \\
        \featref{Lifebound} & Con 10 & Immune to hostile Life and Death effects   & \x & \featpref{Lifebound} \\
        \featref{Lightning Reflexes} & Dex 3 or Per 3 &  \plus2 to Reflex defense & \x & \featpref{Lightning Reflexes} \\
        \featref{Natural Spellgift} & 2nd level, Wil 1 & Gain natural spell as spell-like ability & Spellgift & \featpref{Natural Spellgift} \\
        \featref{Perfect Health} & Con 3 & Immune to disease, later poison        & \x & \featpref{Perfect Health} \\
        \featref{Pierce Delusions} & Per 5 & Immune to Delusion effects          & \x & \featpref{Pierce Delusions} \\
        %\featref{Rapid Recovery} & Con 3 & Naturally heal quickly & \x & \featpref{Rapid Recovery} \\
        \featref{Stubborn} & Wil 5 & Immune to Compulsion effects                 & \x & \featpref{Stubborn} \\
        \featref{Swift} & \x & Increase speed by 5 feet                           & Mobility & \featpref{Swift} \\
        \featref{Toughness} & Con 3 or Str 3 & \plus2 to Fortitude defense        & \x & \featpref{Toughness} \\

        \midrule
        \tb{Bloodline Feats} & \tb{Prerequisites} & \tb{Benefit} & \tb{Feat Types} & \tb{Page} \\
        \featref{Celestial Heritage} & Nonevil alignment & Gain retributive aura & Bloodline, Celestial & \featpref{Celestial Heritage} \\
            \tind \featref{Celestial Blessing} & Celestial Heritage & Gain ability to bless allies & Bloodline, Celestial & \featpref{Celestial Blessing} \\
            \tind \featref{Celestial Body} & Celestial Heritage & Gain physical damage reduction & Bloodline, Celestial & \featpref{Celestial Body} \\
            \tind \featref{Celestial Smiting} & Celestial Heritage & Smite evil more often, more accurately & Bloodline, Celestial & \featpref{Celestial Smiting} \\
            \tind \featref{Celestial Soul} & Any three celestial bloodline feats & Gain spell resistance against evil & Bloodline, Celestial & \featpref{Celestial Soul} \\
            \tind \featref{Celestial Wings} & Any three celestial bloodline feats & Gain wings to slow falls, glide, eventually fly & Bloodline, Celestial & \featpref{Celestial Wings} \\
        \featref{Draconic Heritage} & \x & Resist damage from chosen dragon's energy type & Bloodline, Draconic & \featpref{Draconic Heritage} \\
            \tind \featref{Draconic Breath} & Con 3, any three dragon bloodline feats & Gain dragon's breath weapon & Bloodline, Draconic & \featpref{Draconic Breath} \\
            \tind \featref{Draconic Might} & Any three dragon bloodline feats & Gain \plus1 to physical attribute & Bloodline, Draconic & \featpref{Draconic Might} \\
            \tind \featref{Draconic Mind} & Any three dragon bloodline feats & Gain \plus1 to mental attribute & Bloodline, Draconic & \featpref{Draconic Mind} \\
            \tind \featref{Draconic Scales} & Draconic Heritage & Gain natural armor & Bloodline, Draconic & \featpref{Draconic Scales} \\
            \tind \featref{Draconic Senses} & Draconic Heritage & Gain low-light vision, possibly darkvision & Bloodline, Draconic & \featpref{Draconic Senses} \\
            \tind \featref{Draconic Spellpower} & Draconic Heritage & Gain bonus to spellpower & Bloodline, Draconic & \featpref{Draconic Spellpower} \\
            \tind \featref{Draconic Voice} & Draconic Heritage & Gain bonus to Intimidate and Persuasion checks & Bloodline, Draconic & \featpref{Draconic Voice} \\
            \tind \featref{Draconic Weapons} & Draconic Heritage & Gain bite attack, possibly claws & Bloodline, Draconic & \featpref{Draconic Weapons} \\
            \tind \featref{Draconic Wings} & Any three dragon bloodline feats & Gain wings to slow falls, glide, eventually fly & Bloodline, Draconic & \featpref{Draconic Wings} \\
        \featref{Fae Heritage} & \x & Gain minor spell-like abilities & Bloodline & \featpref{Fae Heritage} \\

        \midrule
        \tb{Class Feats} \label{cap:Class Feats} & \tb{Prerequisites} & \tb{Benefit} & \tb{Feat Types} & \tb{Page} \\
        \featref{Accelerated Rage}            & Rage, barbarian 6                 & \plus30 foot speed while raging  & Rage & \featpref{Accelerated Rage} \\
        \featref{Athletic Rage}         & Rage, barbarian 2                 & Gain bonus to athletic skills while raging & Rage & \featpref{Athletic Rage} \\
        \featref{Arcane Resilience}     & Sorcerer 5                        & Gain damage reduction against arcane spells & \x & \featpref{Arcane Resilience} \\
        \featref{Battlemaster}          & Int 5, fighter 10                 & Grant ally a combat feat & Combat & \featpref{Battlemaster} \\
            \tind \featref{Battlemaster, Empowered} & Int 10, fighter 16                   & Grant ally two combat feats & Combat & \featpref{Battlemaster, Empowered} \\
        \featref{Battletrance}          & Rage, barbarian 2                 & Enter trance instead of angry rage & Rage & \featpref{Battletrance} \\
        \featref{Chaotic Mind}          & Paladin of chaos 3                & Alignment and lies cannot be detected & \x & \featpref{Chaotic Mind} \\
        \featref{Charming Wild Speech}  & Wild speech, druid 7              & Charm target with wild speech & \x & \featpref{Charming Wild Speech} \\
            \tind \featref{Dominating Wild Speech}  & Charming Wild Speech, druid 15       & Dominate target with wild speech & \x & \featpref{Dominating Wild Speech} \\
        \featref{Critical Rage}         & Rage, barbarian 10                & Increased critical range while raging & Rage & \featpref{Critical Rage} \\
        \featref{Destructive Rage}      & Rage, barbarian 6                 & Ignore hardness while raging & Rage & \featpref{Destructive Rage} \\
        \featref{Divine Health}         & Paladin 3                         & Immune to poison and disease & \x & \featpref{Divine Health} \\
        \featref{Elemental Speech}      & Wild speech, druid 9              & Speak with an element & \x & \featpref{Elemental Speech} \\
        \featref{Endless Rage}          & Rage, barbarian 18                & Rage lasts as long as you take damage & Rage & \featpref{Endless Rage} \\
        \featref{Extra Ability}         & Class ability with limited uses   & Gain additional uses of class ability & \x & \featpref{Extra Ability} \\
        \featref{Fearsome Rage}         & Rage, barbarian 6                 & Attacks make foes shaken while raging & Rage & \featpref{Fearsome Rage} \\
        \featref{Frenzy}                & Rage, barbarian 6                 & Enter frenzy instead of angry rage & Rage & \featpref{Frenzy} \\
        \featref{Furious Style}         & Rage, barbarian 6                 & Change combat styles while raging & Rage & \featpref{Furious Style} \\
        \featref{Insightful Fighter}    & Adaptive style                    & Use Perception for adaptive style (and battlemaster) & \x & \featpref{Insightful Fighter} \\
        \featref{Invulnerable Rage}     & Rage, barbarian 10                & Improve damage reduction while raging & Rage & \featpref{Invulnerable Rage} \\
        \featref{Insightful Rogue}      & Combat trick                      & Use Perception for combat tricks & \x & \featpref{Insightful Rogue} \\
        \featref{Legendary Versatility} & 5 levels in two classes           & Treat multiple classes as base classes & \x & \featpref{Legendary Versatility} \\
        \featref{Ki Focus}              & Manifest \ki, monk 8              & \Ki manifestation lasts longer & \x & \featpref{Ki Focus} \\
            \tind \featref{Ki Specialization}       & Ki Focus, monk 12                    & \Ki manifestation lasts all day & \x & \featpref{Ki Specialization} \\
        \featref{Ki Training}           & Manifest \ki                      & Learn additional ki manifestation & \x & \featpref{Ki Training} \\
        \featref{Mindless Rage}         & Rage, barbarian 10                & Immune to Mind effects while raging & Rage & \featpref{Mindless Rage} \\
        \featref{Overpowering Rage}     & Rage, barbarian 6                & Gain combat maneuver bonus while raging & Rage & \featpref{Overpowering Rage} \\
        \featref{Overwhelming Rage}     & Rage, barbarian 6                 & Increase overwhelm penalties while raging & Rage & \featpref{Overwhelming Rage} \\
        \featref{Patient Rage}          & Rage, barbarian 2                 & Don't take damage for not attacking while raging & Rage & \featpref{Patient Rage} \\
        \featref{Plant Speech}          & Wild speech, druid 5              & Speak with plants & \x & \featpref{Plant Speech} \\
        \featref{Rapid Adaptation}      & Adaptive combat, Int 5, fighter 8 & Change combat feats more quickly & \x & \featpref{Rapid Adaptation} \\
            \tind \featref{Instant Adaptation}      & Rapid Adaptation, Int 10, fighter 18 & Change combat feats instantly & \x & \featpref{Instant Adaptation} \\
        \featref{Resist Maneuver}       & Fighter                           & \plus5 to resist chosen maneuver & \x & \featpref{Resist Maneuver} \\
        \featref{Savage Rage}           & Rage, barbarian 2                 & Fight better unarmed while raging & Rage & \featpref{Savage Rage} \\
        \featref{Shield Discipline}     & Armor discipline                  & Apply armor discipline with shields \\
        \featref{Spellbreaker Rage}     & Rage, barbarian 10                & Gain spell resistance while raging & Rage & \featpref{Spellbreaker Rage} \\
        \featref{Spellwarped Ritualist} & Spellwarped                       & Gain ability to perform rituals using spellwarped powers & \x & \featpref{Spellwarped Ritualist} \\
        \featref{Unbending Resolve}     & Paladin 5                         & Immune to compulsion and delusion effects & \x & \featpref{Unbending Resolve} \\
        \featref{Taunting Rage}         & Rage, barbarian 10                & Attacks taunt foes while raging & Rage & \featpref{Taunting Rage} \\
        \featref{Terrain Lore} & Terrain lore, ranger 7 & Gain additional terrain lore & \x & \featpref{Terrain Lore} \\
        \featref{Vermin Speech}         & Wild speech, druid 3              & Speak with vermin & \x & \featpref{Vermin Speech} \\
        \featref{Wary Rage}             & Rage, barbarian 2                 & Reduce defense penalty while raging & Rage & \featpref{Wary Rage} \\
        \featref{Willful Rage}          & Rage, barbarian 2                 & Gain Mental defense bonus while raging & Rage & \featpref{Willful Rage} \\
        \featref{Wild Substitution}     & Wild magic, sorcerer 7            & Miscasting removes access to lower level spells & \x & \featpref{Wild Substitution} \\
        \featref{Wild Suppression}     & Wild magic, sorcerer 3            & Suppress spell instead of miscasting it & \x & \featpref{Wild Suppression} \\
        \featref{Versatility}           & Levels in two different classes   & Increase effective class level with abilities & \x & \featpref{Versatility} \\

        \midrule
        \tb{Item Creation Feats} & \tb{Prerequisites} & \tb{Benefit} & \tb{Feat Types} & \tb{Page} \\
        \featref{Imbue Magic} & 1st level spells or Craft (any) 5 ranks & Create magic items & Item Creation & \featpref{Imbue Magic} \\
        \featref{Imbuement Admixture} & 2nd level spells, 6th & Combine multiple spells to create items & Item Creation & \featpref{Imbuement Admixture} \\
        \featref{Versatile Crafter} & Craft (any) 10 ranks & Craft items from many schools of magic & Item Creation & \featpref{Versatile Crafter} \\

        \midrule
        \tb{Magic Feats} & \tb{Prerequisites} & \tb{Benefit} & \tb{Feat Types} & \tb{Page} \\
        \featref{Abjurant Shield} & 2nd level Shielding spell & \plus1 to physical defenses & \x & \featpref{Abjurant Shield} \\
        \featref{Augment Summoning} & Summoning spell, 2nd level spells & \plus2 to spellpower with Summoning spells & \x & \featpref{Augment Summoning} \\
        \featref{Bardic Spellpower} & Arcane spellcasting ability & Fighter and rogue levels improve some spells & \x & \featpref{Bardic Spellpower} \\
        \featref{Battlecaster} & Proficiency with light armor & Reduce arcane spell failure by 10\% & \x & \featpref{Battlecaster} \\
        \featref{Combat Casting} & \x &  \plus3 bonus to Concentration checks to cast spells & \x & \featpref{Combat Casting} \\
        \featref{Commanding Presence} & 1st level Compulsion spell & \plus2 to Intimidate and Persuasion, later \plus4 & \x & \featpref{Commanding Presence} \\
        \featref{Counterspell} & Spellcraft 10 ranks, 2nd level spells & Counter spells cast by foe & \x & \featpref{Counterspell} \\
            \tind \featref{Improved Counterspell} & Counterspell & Counter spells more easily & \x & \featpref{Improved Counterspell} \\
            \tind \featref{Mass Counterspell} & Spellcraft 15 ranks, 4th level spells, Counterspell & Counter spells cast by multiple foes & \x & \featpref{Mass Counterspell} \\
        \featref{Craftcaster} & Craft (any) 10 ranks, 2nd level spells & Increase spellpower when creating, transforming objects & \x & \featpref{Craftcaster} \\
        \featref{Delusive Guise} & 1st level Delusion spell & \plus2 to Bluff and Disguise, later \plus4 & \x & \featpref{Delusive Guise} \\
        \featref{Dispel Mastery} & 3rd level Antimagic spell & Dispel more effectively & \x & \featpref{Dispel Mastery} \\
        \featref{Enhanced Senses} & 1st level Awareness spell & \plus2 to Awareness and Sense Motive, later \plus4 & \x & \featpref{Enhanced Senses} \\
        \featref{Explosive Miscast} & 4th level spells & Miscast explosions are larger, target foes & \x & \featpref{Explosive Miscast} \\
        \featref{Fire Mastery} & 3rd level Fire spell & Fire spells can ignite foes & \x & \featpref{Fire Mastery} \\
        \featref{Foresight} & 3rd level Divination spell & Cannot be caught unaware & \x & \featpref{Foresight} \\
        \featref{Frigid Forge} & 2nd level Cold spell & Form ice into weapons and armor & \x & \featpref{Frigid Forge} \\
        \featref{Magical Synthesis} & Levels in two magical classes & Increase spellcasting ability in two magical classes at once & \x & \featpref{Magical Synthesis} \\
        \featref{Mass Transporter} & 5th level Conjuration (Translocation) spell & Transport many creatures at once & \x & \featpref{Mass Transporter} \\
        \featref{Personal Magic} & \x & Personalized magic grants various benefits & \x & \featpref{Personal Magic} \\
        \featref{Resilient Magic} & 3rd level Antimagic spell & Spells persist after being dispelled, later cannot be dispelled & \x & \featpref{Resilient Magic} \\
        \featref{Retargeting Miscast} & 3rd level spells & Roll twice for miscast targets & \x & \featpref{Retargeting Miscast} \\
        \featref{Ritual Caster} & Int 3 & Gain ability to perform rituals & \x & \featpref{Ritual Caster} \\
        \featref{Spell Focus}\fn{1} & 2nd level spells & Cheaper augment or metamagic with one school of magic & \x & \featpref{Spell Focus} \\
            \tind \featref{Spell Specialization} & 4th level spells, Spell Focus &  \plus2 spellpower with specific school of magic, \minus2 with other types.
        & \x & \featpref{Spell Specialization} \\
        \featref{Somatic Strike} & Combat prowess 8, 4th level spells & Make strike in place of somatic components & \x & \featpref{Somatic Strike} \\
        \featref{Summoning Mastery} & Summoning spell, 4th level spells & Maintain one Summoning spell indefinitely with no effort & \x & \featpref{Summoning Mastery} \\
        \featref{Sympathetic Resistance} & 2nd level Abjuration spell & Casting damaging spells grants brief damage reduction & \x & \featpref{Sympathetic Resistance} \\
        \featref{Unfettered Knowledge} & 2nd level Knowledge spell & \plus2 to all Knowledge checks, later \plus4 & \x & \featpref{Unfettered Knowledge} \\
        \featref{Vital Flow} & 4th level Positive spell & Automatically heal every round & \x & \featpref{Vital Flow} \\
        \featref{Wardwalker} & 3rd level Abjuration (Warding) spell & Wards you encounter trigger one round late & \x & \featpref{Wardwalker} \\
        \featref{Wired} & 1st level Electricity spell & Increase speed by 5 feet, later 10 feet & \x & \featpref{Wired} \\

        \midrule
        \tb{Magic Style Feats} & \tb{Prerequisites} & \tb{Benefit} & \tb{Feat Types} \tb{Page} \\
        \featref{Blood Magic} & 4th level Vivimancy spell & Gain spellpower bonus after taking damage & Style & \featpref{Blood Magic} \\
        \featref{Brutal Magic} & 4th level spells & Gain spellpower bonus after making physical strike & Style & \featpref{Brutal Magic} \\
        \featref{Creative Magic} & 4th level spells & Gain spellpower bonus with spells not cast recently & Style & \featpref{Creative Magic} \\
        \featref{Devastating Magic} & 8th level spells & Spells trigger critical successes more easily & Style & \featpref{Devastating Magic} \\
        \featref{Flameborne Magic} & 4th level Fire spell & Gain spellpower bonus when fire is near & Style & \featpref{Flameborne Magic} \\
        \featref{Peaceful Magic} & 4th level Abjuration spell & Gain spellpower bonus after not taking damage & Style & \featpref{Peaceful Magic} \\
        \featref{Unliving Conduit} & 4th level Vivimancy spell & Gain spellpower bonus when undead are near & Style & \featpref{Unliving Conduit} \\

        \midrule
        \tb{Metamagic Feats} & \tb{Prerequisites} & \tb{Benefit (Level Adjustment)} & \tb{Feat Types} & \tb{Page} \\
        \featref{Distant Spell} & 2nd level spells & Double spell's range (\plus1) & \x & \featpref{Distant Spell} \\
        \featref{Energetic Substitution} & 2nd level spells & Change spell's energy type (\plus1) & \x & \featpref{Energetic Substitution} \\
        \featref{Imbued Spellstrike} & 2nd level spells & Combine spell with weapon strike (\plus1) & \x & \featpref{Imbued Spellstrike} \\
            \tind \featref{Persistent Spellstrike} & 3rd level spells, Imbued Spellstrike & Imbue weapon with spell's power up to 5 minutes (\plus2) & \x & \featpref{Persistent Spellstrike} \\
        \featref{Quicken Spell} & 3rd level spells & Cast spell as swift action, but lose next action (\plus2) & \x & \featpref{Quicken Spell} \\
        \featref{Reach Spell} & 2nd level spells & Cast touch spell at Close range (\plus1) & \x & \featpref{Reach Spell} \\
        \featref{Shape Spell} & 3rd level spells & Exclude areas within spell's area (\plus2) & \x & \featpref{Shape Spell} \\
        \featref{Silent Spell} & 2nd level spells & Cast spell without verbal components (\plus1) & \x & \featpref{Silent Spell} \\
        \featref{Still Spell} & 2nd level spells & Cast spell without somatic components (\plus1) & \x & \featpref{Still Spell} \\
        \featref{Sustained Spell} & 2nd level spells & Maintain concentration as swift action (\plus1) & \x & \featpref{Sustained Spell} \\

        \midrule
        \tb{Performance Feats} & \tb{Prerequisites} & \tb{Benefit} & \tb{Feat Types} & \tb{Page} \\
        \featref{Dual Performance} & Perform 15 ranks, any three Performance feats & Sustain two performances at once & Skill & \featpref{Dual Performance} \\
        \featref{Fascinating Performance} & Perform 5 ranks & Fascinate creatures & Skill & \featpref{Fascinating Performance} \\
            \tind \featref{Suggestive Performance} & Perform 13 ranks, Fascinating Performance & Suggest action to fascinated creatures & Skill & \featpref{Suggestive Performance} \\
        \featref{Inspire Competence} & Perform 5 ranks  & Grant allies offensive legend point & Skill & \featpref{Inspire Competence} \\
        \featref{Inspire Courage} & Perform 9 ranks  & Grant allies temporary hit points, immunity to fear & Skill & \featpref{Inspire Courage} \\
        \featref{Inspire Resilience} & Perform 13 ranks  & Grant allies damage reduction, defensive legend point & Skill & \featpref{Inspire Resilience} \\
        \featref{Inspire Mastery} & Perform 17 ranks  & Grant allies legend points every round & Skill & \featpref{Inspire Mastery} \\
        \featref{Inspire Serenity} & Perform 13 ranks  & Dispel magical effects on ally & Skill & \featpref{Inspire Serenity} \\
        \featref{Mocking Performance} & Perform 5 ranks & Impair foes with mockery & Skill & \featpref{Mocking Performance} \\
        \featref{Rapid Performance} & Perform 15 ranks, any three Performance feats & Perform as a swift action & Skill & \featpref{Rapid Performance} \\
        \featref{Recover Performance} & Perform 11 ranks, any other Performance feat & Reroll failed performance attempt & Skill & \featpref{Recover Performance} \\
        \featref{Spellwoven Performance} & Perform 9 ranks, any other Performance feat & Cast some spells while performing & Skill & \featpref{Spellwoven Performance} \\
        \featref{Terrifying Performance} & Perform 15 ranks & Invoke fear in foes & Skill & \featpref{Terrifying Performance} \\

        \midrule
        \tb{Racial Feats} & \tb{Prerequisites} & \tb{Benefit} & \tb{Feat Types} & \tb{Page} \\
        \featref{Dwarven Resilience} & Dwarf & \plus2 to defenses against spells & \x & \featpref{Dwarven Resilience} \\
        \featref{Giantfighter} & Dwarf, gnome, or halfling & \plus2 to physical defenses against Large or larger creatures & Combat & \featpref{Giantfighter} \\
        \featref{Racial Acclimation} & Have a race other than your original race & Lose penalties for being in new body & \x & \featpref{Racial Acclimation} \\
        \featref{Stonecunning} & Dwarf & Gain a sixth sense about stonework & \x & \featpref{Stonecunning} \\

        \midrule
        \tb{Skill Feats} & \tb{Prerequisites} & \tb{Benefit} & \tb{Feat Types} & \tb{Page} \\
        \featref{Fungal Poisoner} & Survival 5 ranks & Use Survival to find and use fungal poisons & \x & \featpref{Fungal Poisoner} \\
        \featref{Hide in Plain Sight} & Stealth 20 ranks & Hide while being observed & \x & \featpref{Hide in Plain Sight} \\
        \featref{Ledgewalker} & Acrobatics 5 ranks & Balance easily on narrow surfaces & \x & \featpref{Ledgewalker} \\
        \featref{Legendary Awareness} & Awareness 15 ranks & Gain extrasensory perception & \x & \featpref{Legendary Awareness} \\
        \featref{Legendary Balance} & Acrobatics 15 ranks & Balance on impossible surfaces & \x & \featpref{Legendary Balance} \\
        \featref{Legendary Climber} & Climb 15 ranks & Gain climb speed, climb on impossible surfaces & \x & \featpref{Legendary Climber} \\
        \featref{Legendary Craftsman} & Craft 15 ranks & Craft items with fewer material components & \x & \featpref{Legendary Craftsman} \\
        \featref{Legendary Devicesmith} & Devices 15 ranks & Disable active spell effects & \x & \featpref{Legendary Devicesmith} \\
        \featref{Legendary Disguise} & Disguise 15 ranks & Alter magical auras with disguise & \x & \featpref{Legendary Disguise} \\
        \featref{Legendary Escapist} & Escape Artist 15 ranks & Escape from magical effects & \x & \featpref{Legendary Escapist} \\
        \featref{Legendary Handler} & Creature Handling 15 ranks & Train and command creatures faster & \x & \featpref{Legendary Handler} \\
        \featref{Legendary Healer} & Heal 15 ranks & Heal damage three times per day & \x & \featpref{Legendary Healer} \\
        \featref{Legendary Intimidation} & Intimidate 15 ranks & Panic intimidated foes & \x & \featpref{Legendary Intimidation} \\
        \featref{Legendary Knowledge} & Knowledge (any) 15 ranks & \plus2 against creatures you identify & \x & \featpref{Legendary Knowledge} \\
        \featref{Legendary Leaper} & Jump 15 ranks & Always treated as having running start & \x & \featpref{Legendary Leaper} \\
        \featref{Legendary Legerdemain} & Sleight of Hand 15 ranks & Hide objects in pocket dimension & \x & \featpref{Legendary Legerdemain} \\
        \featref{Legendary Liar} & Bluff 15 ranks & Lies appear to alter reality & \x & \featpref{Legendary Liar} \\
        \featref{Legendary Linguist} & Linguistics 15 ranks & Speak all languages & \x & \featpref{Legendary Linguist} \\
        \featref{Legendary Persuasion} & Persuasion 15 ranks & Compel listener to obey & \x & \featpref{Legendary Persuasion} \\
        \featref{Legendary Profession} & Profession (any) 15 ranks & Become famous for mastery of your profession & \x & \featpref{Legendary Profession} \\
        \featref{Legendary Rider} & Ride 15 ranks & Ride unwilling creatures & \x & \featpref{Legendary Rider} \\
        \featref{Legendary Sense Motive} & Sense Motive 15 ranks & Read creature's mind & \x & \featpref{Legendary Sense Motive} \\
        \featref{Legendary Spellcraft} & Spellcraft 15 ranks & Gain \plus2 against spells you identify & \x & \featpref{Legendary Spellcraft} \\
        \featref{Legendary Sprinter} & Sprint 15 ranks & You can sprint for much longer & \x & \featpref{Legendary Sprinter} \\
        \featref{Legendary Stealth} & Stealth 15 ranks & Hide in plain sight & \x & \featpref{Legendary Stealth} \\
        \featref{Legendary Survival} & Survival 15 ranks & Ignore harmful terrain and unerringly find your way & \x & \featpref{Legendary Survival} \\
        \featref{Legendary Tumbler} & Acrobatics 15 ranks & Tumble through enemies without provoking & \x & \featpref{Legendary Tumbler} \\
        \featref{Open Minded} & \x & Gain two skill points & \x & \featpref{Open Minded} \\
        \featref{Poison Expert} & Int 3, Craft (poison) 10 ranks & Apply poisons more effectively & \x & \featpref{Poison Expert} \\
        \featref{Ranged Legerdemain} &  2nd level Telekinesis spell & Use Disable Device or Sleight of Hand at range & \x & \featpref{Ranged Legerdemain} \\
        \featref{Scale the Beast} & Climb 10 ranks & Climb on massive creatures & \x & \featpref{Scale the Beast} \\
        \featref{Skill Focus}\fn{1} & Any skill 5 ranks & Roll twice for checks with selected skill & \x & \featpref{Skill Focus} \\
        \tind \featref{Skill Specialization}\fn{1} & Any skill 10 ranks, Skill Focus with skill & \plus5 to skill & \x & \featpref{Skill Specialization} \\
        \featref{Trapfinder} & Awareness 5 ranks & Find traps faster & \x & \featpref{Trapfinder} \\
        \featref{Trap Sense} & Awareness 10 ranks & Automatically detect traps & \x & \featpref{Trap Sense} \\
        \featref{Track} & Survival 5 ranks &  Use Survival skill to track & \x & \featpref{Track} \\
        \featref{Untrackable} & Survival 15 ranks & Become impossible to track & \x & \featpref{Untrackable} \\
        \featref{Veteran's Experience} & Combat prowess 8 & Identify monsters without Knowledge & \x & \featpref{Veteran's Experience} \\
        \featref{Wallpuncher} & Strength 5, Climb 10 & Punch walls to create handholds & \x & \featpref{Wallpuncher} \\
    \end{longtabu}
\end{longtabuwrapper}

\subsection{Combat Feat Table}\label{cap:Combat Feats}

Some combat feats require a specific number of physical attacks.\\
Only physical attacks that come from having a high \glossterm{combat prowess} count towards these prerequisites.\\
The table of combat feats is organized according to the combat prowess each feat requires.\\

\begin{longtabuwrapper}
    \begin{longtabu}{>{\lcol}p{10em} >{\lcol}p{15em} >{\lcol}X >{\lcol}p{8em} >{\lcol}p{3em}}
        \tb{Combat Feats} & \tb{Prerequisites} & \tb{Benefit} & \tb{Feat Types} & \tb{Page} \\
        \featref{Armor Familiarity} & Proficiency with armor & Reduce penalties from wearing armor & Equipment & \featpref{Armor Familiarity} \\
        \featref{Armor Proficiency} & Proficiency with lighter armor & Become proficient with type of armor & Equipment & \featpref{Armor Proficiency} \\
        \featref{Cleave} & Str 3 & Extra melee strike after dropping target & Power & \featpref{Cleave} \\
        \featref{Close-Quarters Fighting} & Dex 3 & No penalty for squeezing or attacking while grappling & Finesse & \featpref{Close-Quarters Fighting} \\
        \featref{Exotic Weapon Proficiency} & \x & Become proficient with exotic weapons & Equipment & \featpref{Exotic Weapon Proficiency} \\
        \featref{Far Shot} & Str 3 & Increase range increment & Power & \featpref{Far Shot} \\
        \featref{Mighty Blows} & Str 3 & \plus1 to physical damage & Power & \featpref{Mighty Blows} \\
        \featref{Mounted Archery} & Ride training & Reduce penalty for ranged accuracy while mounted & Mounted, Precision & \featpref{Mounted Archery} \\
        \featref{Mounted Defense} & Ride training & Negate hits on mount with Ride check & Defense, Mounted & \featpref{Mounted Defense} \\
        \featref{Mounted Trample} & Ride 10 ranks & Target cannot avoid mounted overrun & Maneuver, Mounted & \featpref{Mounted Trample} \\
        \featref{Quick Draw} & \x & Draw weapon as swift action & Reaction & \featpref{Quick Draw} \\
        \featref{Shield Proficiency} & \x & Become proficient with shields & Equipment & \featpref{Shield Proficiency} \\
            \tind \featref{Shield Expertise} & Shield Proficiency & \plus1 bonus to physical defenses while using a shield & Defense & \featpref{Shield Expertise} \\
        \featref{Tactical Analysis} & Int 3 & Identify foe's strengths and weaknesses & Awareness & \featpref{Tactical Analysis} \\
        \featref{Two-Weapon Fighting} & Dex 3 & Gain \plus2 bonus when fighting with two weapons & Finesse & \featpref{Two-Weapon Fighting} \\
            \tind \featref{Two-Weapon Defense} & Two-Weapon Fighting & Gain \plus1 to defenses when fighting with two weapons & Defense, Finesse & \featpref{Two-Weapon Defense} \\
        \featref{Unarmed Proficiency} & \x & Proficient with unarmed attacks & \x & \featpref{Unarmed Proficiency} \\
            \tind \featref{Deflect Arrows} & Dex 3, Unarmed Proficiency & Deflect ranged attacks with your bare hands & Finesse, Reaction & \featpref{Deflect Arrows} \\
        \featref{Weapon Proficiency}\fn{1} & \x & Gain proficiency with weapon group & Equipment & \featpref{Weapon Proficiency} \\
        \featref{Weapon Finesse} & Dex 3 & \plus1 damage with light weapons & Finesse & \featpref{Weapon Finesse} \\
        \featref{Weapon Focus} & Proficiency with weapon group & Special ability with weapon group & Equipment & \featpref{Weapon Focus} \\

        \midrule
        \tb{Combat Feats} & \tb{Prerequisites (Prowess 4)} & \tb{Benefit} & \tb{Feat Types} & \tb{Page} \\
        \featref{Combat Mobility} & Dex 5 & Move through foe & Defense, Mobility & \featpref{Combat Mobility} \\
        \featref{Gang Up} & \x & Increase overwhelm penalties foes suffer & \x & \featpref{Gang Up} \\
        \featref{Heavy Hitter} & Str 5 & \plus2 damage with heavy weapons & Power & \featpref{Heavy Hitter} \\
        \featref{Improved Dirty Trick} & \x & \plus2 to dirty trick accuracy; dirty tricks last longer & Finesse, Maneuver & \featpref{Improved Dirty Trick} \\
        \featref{Improved Disarm} & \x & \plus2 to disarm accuracy; knock foe's weapon away after disarming & Finesse, Maneuver & \featpref{Improved Disarm} \\
        \featref{Improved Feint} & \x & \plus2 to feint accuracy; feint lasts longer & Finesse, Maneuver & \featpref{Improved Feint} \\
        \featref{Improved Grapple} & \x & \plus2 to grapple accuracy; not defenseless while grappling & Maneuver, Power & \featpref{Improved Grapple} \\
        \featref{Improved Overrun} & \x & \plus2 to overrun accuracy; target can't avoid & Maneuver, Power & \featpref{Improved Overrun} \\
        \featref{Improved Shove} & \x & \plus2 to shove accuracy; move target away from you & Maneuver, Power & \featpref{Improved Shove} \\
        \featref{Improved Trip} & \x & \plus2 to trip accuracy; free strike against tripped foe & Finesse, Maneuver & \featpref{Improved Trip} \\
        \featref{Mounted Charge} & Ride 10 ranks & Gain bonus damage with mounted charge & Mounted, Power & \featpref{Mounted Charge} \\
        \featref{Tactical Prediction} & Int 5 & Predict foe's next action & Awareness & \featpref{Tactical Prediction} \\
        \featref{Unarmed Might} & Unarmed Proficiency, Str 5 & Unarmed attacks deal more damage & \x & \featpref{Unarmed Might} \\

        \midrule
        \tb{Combat Feats} & \tb{Prerequisites (Prowess 8)} & \tb{Benefit} & \tb{Feat Types} & \tb{Page} \\
        \featref{Armor Piercing Strike} & Per 5 & Attack targets Reflex defense & Precision, Strike & \featpref{Armor Piercing Strike} \\
        \featref{Calculated Strike} & Int 5 & Attack can deal critical damage & Precision, Strike & \featpref{Calculated Strike} \\
        \featref{Countering Strike} & Int 5 & Bonus damage if you are attacked & Reaction, Strike & \featpref{Countering Strike} \\
        \featref{Chargebreaker} & Dex 5 & Bonus damage if target approached you & Reaction, Strike & \featpref{Chargebreaker} \\
        \featref{Driving Assault} & Str 5 & Attack shoves creature  & Power, Maneuver, Strike & \featpref{Driving Assault} \\
        \featref{Focused Strike} & \x & Trade strikes for accuracy & Precision, Strike & \featpref{Focused Strike} \\
        \featref{Infuriating Strike} & \x & Melee attack taunts foe & Strike & \featpref{Infuriating Strike} \\
        \featref{Intimidating Strike} & Intimidate 10 ranks & Melee attack intimidates foe & Strike & \featpref{Intimidating Strike} \\
        \featref{Knockdown} & Str 5 & Shove foe and knock prone & Maneuver, Power, Strike & \featpref{Knockdown} \\
        \featref{Manyshot} & Dex 5 & Shoot two projectiles simultaneously & Finesse & \featpref{Manyshot} \\
        \featref{Pierce Wings} & Per 5 & Ranged attack removes foe's ability to fly & Precision, Strike & \featpref{Pierce Wings} \\
        \featref{Two-Weapon Rend} & Str 5, Two-Weapon Fighting & Gain bonus damage when fighting with two weapons & Power & \featpref{Two-Weapon Rend} \\
        \featref{Unarmed Dodge} & Unarmed Proficiency, Dex 5 & \plus1 physical defenses while unarmed & \x & \featpref{Unarmed Dodge} \\
        \featref{Wall Slam} & Str 5 & Slam opponent into wall to deal bonus damage & Maneuver, Power, Strike & \featpref{Wall Slam} \\
        \featref{Whirlwind Strike} & Dex 5 & One melee strike against each threatened opponent & Strike & \featpref{Whirlwind Strike} \\

        \midrule
        \tb{Combat Feats} & \tb{Prerequisites (Prowess 12)} & \tb{Benefit} & \tb{Feat Types} & \tb{Page} \\
        \featref{Disorienting Strike} & \x & Attack disorients foe & Power, Strike & \featpref{Disorienting Strike} \\
        \featref{Killing Blow} & Int 10 & Attack instantly kills foe without hit points & Precision, Strike & \featpref{Killing Blow} \\
        \featref{Hydra Strike} & Str 5, Dex 5, Con 5 & Attack all defenses for bonus damage & \x & \featpref{Hydra Strike} \\
        \featref{Legendary Strike} & Any three Strike feats & \plus2 to hit with strikes & \x & \featpref{Legendary Strike} \\
        \featref{Legendary Combat Awareness} & Per 10, any three Awareness feats & Immune to being overwhelmed & Awareness & \featpref{Legendary Combat Awareness} \\
        \featref{Legendary Defense} & Any three Defense feats & \plus2 to all defenses & Defense & \featpref{Legendary Defense} \\
        \featref{Legendary Finesse} & Dex 10, any three Finesse feats & Roll maximum damage again & Finesse & \featpref{Legendary Finesse} \\
        \featref{Legendary Maneuver Master} & Any three Maneuver feats & Deal damage with very successful maneuvers & Maneuver & \featpref{Legendary Maneuver Master} \\
        \featref{Legendary Mobility} & Dex 10, any three Mobility feats & Move through enemies & Mobility& \featpref{Legendary Mobility} \\
        \featref{Legendary Mounted Warrior} & Any three Mounted feats & Share damage with mount & Mounted& \featpref{Legendary Mounted Warrior} \\
        \featref{Legendary Power} & Str 10, any three Power feats & Wield weapons as if they were less encumbering & Power& \featpref{Legendary Power} \\
        \featref{Legendary Precision} & Per 10, any three Precision feats & Very accurate hits deal maximum damage & Precision & \featpref{Legendary Precision} \\
        \featref{Legendary Reaction} & Dex 10, any three Reaction feats & Extra immediate actions & Combat, Reaction & \featpref{Legendary Reaction} \\
        \featref{Legendary Unarmed Warrior} & Unarmed Dodge, Unarmed Might, Unarmed Proficiency & Unarmed attacks are incredibly strong & \x & \featpref{Legendary Unarmed Warrior} \\
        \featref{Patient Strike} & Int 10 & Bonus damage if you are not attacked & Reaction, Strike & \featpref{Patient Strike} \\
        \featref{Penetrating Strike} & \x & Attack removes foe's damage reduction & Strike & \featpref{Penetrating Strike} \\
        \featref{Staggering Strike} & \x & Attack staggers foe & Strike & \featpref{Staggering Strike} \\

        \midrule
        \tb{Combat Feats} & \tb{Prerequisites (Prowess 18)} & \tb{Benefit} & \tb{Feat Types} & \tb{Page} \\
        \featref{Executioner} & Any three Reaction feats & Gain free strikes against foes near death & Reaction & \featpref{Executioner} \\
        \featref{Hurricane Strike} & Dex 15, Whirlwind Strike & Two melee strikes against all threatened foes & Strike & \featpref{Hurricane Strike} \\
        \featref{Immobilizing Strike} & \x & Melee attack immobilizes foe & Strike & \featpref{Immobilizing Strike} \\
        \featref{Predictive Strike} & Int 15 & Bonus damage if foe takes predicted action & Reaction, Strike & \featpref{Predictive Strike} \\
    \end{longtabu}
    \begin{enumerate*}
        \item You can gain this feat multiple times.
            Each time you do, it has a different effect.
    \end{enumerate*}
\end{longtabuwrapper}

\subsection{Combat Style Feat Table}

\begin{longtabuwrapper}
    \begin{longtabu}{>{\lcol}p{10em} >{\lcol}p{15em} >{\lcol}X >{\lcol}p{8em} >{\lcol}p{3em}}
        \tb{Style Feats} & \tb{Prerequisites} & \tb{Benefit} & \tb{Feat Types} & \tb{Page} \\
        \featref{Blind-Fight} & Perception 3 & Fight better when unable to see & Awareness, Style & \featpref{Blind-Fight} \\
        \featref{Defensive Fighting} & \x & Gain bonus to physical defenses & Defense, Style & \featpref{Defensive Fighting} \\
        \featref{Distracting Foe} & \x & Threatened foes suffer Concentration penalties & Reaction, Style & \featpref{Distracting Foe} \\
        \featref{Eye of the Storm} & Per 5 & Reduce overwhelm penalties & Awareness, Defense, Style & \featpref{Eye of the Storm} \\
        \featref{Guardian} & \x & Adjacent allies suffer reduced overwhelm penalties & Defense, Style & \featpref{Guardian} \\
        \featref{Overpowering Assault} & Str 3 & Trade defense for maneuver bonus & Power, Style & \featpref{Overpowering Assault} \\
        \featref{Overwhelming Fire} & Dex 3 & Overwhelm foes with ranged weapons & Style & \featpref{Overwhelming Fire} \\
        \featref{Parry} & Dex 3 & Ready yourself to parry incoming blows & Defense, Reaction, Style & \featpref{Parry} \\
            \tind \featref{Shielded Parry} & Shield proficiency, Parry & Add shield bonus to parry attempts & Defense, Reaction & \featpref{Shielded Parry} \\
            \tind \featref{Riposte} & Parry & Foes are defenseless if you parry very well & Reaction & \featpref{Riposte} \\
        \featref{Point Blank Shot} & \x &  \plus2 bonus to ranged accuracy within half range increment & Precision, Style & \featpref{Point Blank Shot} \\

        \midrule
        \tb{Style Feats} & \tb{Prerequisites (Prowess 4)} & \tb{Benefit} & \tb{Feat Types} & \tb{Page} \\
        \featref{Covering Fire} & \x & Impair struck foes & Precision, Style & \featpref{Covering Fire} \\
        \featref{Deadly Aim} & Per 5 & Trade accuracy for ranged damage & Precision & \featpref{Deadly Aim} \\
        \featref{Power Attack} & Str 5 & Trade acccuracy for melee damage & Power & \featpref{Power Attack} \\
        \featref{Precise Shot} & Per 5 & Ignore less than perfect cover and concealment & Precision, Style & \featpref{Precise Shot} \\
        \featref{Predictive Defense} & Int 5 & Use Intelligence for defenses & Awareness, Defense, Style & \featpref{Predictive Defense} \\
        \featref{Reckless Attack} & \x & Trade ability to defend yourself for damage & Power, Style & \featpref{Reckless Attack} \\

        \midrule
        \tb{Style Feats} & \tb{Prerequisites (Prowess 8)} & \tb{Benefit} & \tb{Feat Types} & \tb{Page} \\
        \featref{Heartseeker} & \x & Increase critical range & Precision, Style & \featpref{Heartseeker} \\
        \featref{Inescapable} & \x & Enemies you are blocking cannot move & Reaction, Style & \featpref{Inescapable} \\
        \featref{Spring Attack} & Dex 5, Combat Mobility & Move before and after attacks & Mobility, Style & \featpref{Spring Attack} \\
        \featref{Threatening Fire} & Dex 5, Overwhelming Fire & Threaten adjacent creatures with a ranged weapon & Reaction, Style & \featpref{Threatening Fire} \\

        \midrule
        \tb{Style Feats} & \tb{Prerequisites (Prowess 12)} & \tb{Benefit} & \tb{Feat Types} & \tb{Page} \\
        \featref{Legendary Style} & Any three Style feats & Use two styles at once & \x & \featpref{Legendary Style} \\

        \midrule
        \tb{Style Feats} & \tb{Prerequisites (Prowess 16)} & \tb{Benefit} & \tb{Feat Types} & \tb{Page} \\
        \featref{Counterstorm} & Per 15 & Gain bonuses against foes that miss you & Awareness, Reaction, Style & \featpref{Counterstorm} \\
    \end{longtabu}
\end{longtabuwrapper}

\twocolumn

\section{Feat Descriptions}
Here is the format for feat descriptions.

\ssecfake{Feat Name [Type of Feat]}
\featpre A minimum attribute score, another feat or feats, a minimum combat prowess, a minimum number of ranks in one or more skills, or a class level that a character must have in order to acquire this feat.
This entry is absent if a feat has no prerequisite.
A feat may have more than one prerequisite.
\featben What the feat enables the character (``you'' in the feat description) to do.
If a character has the same feat more than once, its benefits do not stack unless indicated otherwise in the description.
\par In general, characters cannot gain the same feat twice.
\parhead{Normal}
What a character who does not have this feat is limited to or restricted from doing.
If not having the feat causes no particular drawback, this entry is absent.
\parhead{Special}
Additional facts about the feat that may be helpful when you decide whether to acquire the feat.

\feat{Abjurant Shield}{Magic}
\featpre 2nd level or higher \glossterm{Shielding} spell known.
\featben You gain a \plus1 bonus to physical defenses.

\feat{Accelerated Rage}{Class, Rage}
\featpre Rage ability, barbarian level 6.
\featben When you enter a rage, you can use this feat. If you do, during the rage you gain a \plus30 foot bonus to speed in all your movement modes, up to a maximum of double your original speed in that mode.
\parhead{Special} You can only gain the benefit of one Rage feat at a time.

\feat{Arcane Resilience}{Class}
\featpre Sorcerer level 5.
\featben You gain damage reduction against arcane spells equal to your sorcerer level or Constitution, whichever is higher.
This damage reduction applies against your arcane spells, including miscast effects.

\feat{Arcane Spellgift}{Spellgift}
\featpres 2nd level, Willpower 1.
\featben You have inherent magic in your body, granting you arcane power.
When you gain this feat, you choose a non-restricted arcane spell.
The spell's level must be no greater than half your total levels in classes that cannot cast spells.
In addition, its level cannot exceed your Willpower.
You can spend a spellgift point to use that spell as a spell-like ability.

Your spellpower with spellgifts is equal to your Willpower or your level, whichever is higher.
You have a pool with a number of spellgift points equal to twice the number of spellgift feats you possess.

\parhead{Special}
You may take this feat multiple times.
Each time, you choose a new spell to gain as a spell-like ability.
All spellgift feats you have use the same pool of spellgift points.
The number of spellgift feats you have cannot exceed your Willpower.

\feat{Armor Familiarity}{Combat}
\featpre Proficiency with light armor.
\featben You reduce your \glossterm{encumbrance penalty} by 2 and your arcane spell failure by 5\% while wearing body armor.
This only affects penalties from body armor you wear, not shields.

\feat{Armor Piercing Strike}{Combat, Precision}
\featpre Per 5, combat prowess 8.
\featben In place of two strikes, you can make a physical attack.
If you beat the target's Reflex defense, you deal damage normally.
If you also beat its Armor defense, you deal double damage.

\feat{Armor Proficiency}{Combat}
Choose a category of body armor: light, medium, or heavy.
\featpre Proficiency with all lighter types of body armor.
\featben You gain proficiency with your chosen armor type.
\parhead{Normal}
A character who is wearing armor with which she is not proficient applies its \glossterm{encumbrance penalty} to accuracy with physical attacks.
The character also suffers double the normal arcane spell failure chance for wearing the armor.

\feat{Athletic Rage}{Class, Rage}
\featpre Rage ability, barbarian level 2.
\featben When you enter a rage, you can use this feat. If you do, during the rage you add your rage bonus to your Climb, Jump, Sprint, and Swim checks.
\parhead{Special} You can only gain the benefit of one Rage feat at a time.

\feat{Attribute Training}{General}
\featben You increase one of your attributes by 1, up to a maximum value equal to your level.
Other bonuses, such as racial bonuses, increase the attribute beyond your level.
\parhead{Special} This feat can be taken multiple times. Its effects stack.

\feat{Augment Summoning}{Magic}
\featpre \glossterm{Summoning} spell known, 2nd level spells.
\featben You gain a \plus2 bonus to spellpower with Summoning spells.

\feat{Bardic Spellpower}{Magic}
\featpre Any arcane spell known.
\featben You can treat your fighter and rogue levels as magical classes when determining your spellpower with enchantment and illusion arcane spells.

\feat{Battlecaster}{Magic}
\featpre Proficiency with light armor.
\featben You reduce your chance of arcane spell failure from wearing armor by 10\%.

\feat{Battlemaster}{Class, Combat}
\featpre Intelligence 5, fighter level 10.
\featben As a swift action, you may grant the use of one of your combat feats to a willing creature within \rngmed range of you who can see and hear you.
The target must meet combat prowess prerequisites for the granted feat, but it can ignore all other prerequisites.
The effect lasts as long as you spend a swift action to maintain it, to a maximum number of rounds equal to your Intelligence.
After using this ability, you cannot use it again for 5 minutes.

\feat{Battlemaster, Empowered}{Class, Combat}
\featpre Intelligence 10, fighter level 16.
\featben When using your battlemaster ability, you can grant two feats instead of one.

\feat{Battletrance}{Class}
\featpre Rage ability, barbarian level 2.
\featben When you use your rage ability, you can use this feat. If you do, in place of the normal benefits and drawbacks from raging, you can use your Perception to determine your damage with light weapons.
In addition, you add your rage bonus to your Fortitude and Mental defenses.
If you do not spend a swift round each round to sustain the rage, it ends at the end of the round.
\parhead{Special} You can only gain the benefit of one Rage feat at a time.

\feat{Blind-Fight}{Awareness, Combat, Style}
\featpre Perception 3.
\featben While you are in this style, whenever you miss a melee attack because of a miss chance from because of being unable to see your opponent, you can reroll your miss chance one time to see if you actually hit.
\par If your Perception is 10, you gain blindsense into adjacent squares, allowing you to automatically pinpoint the location of invisible or concealed creatures.
\par If your Perception is 20, you gain blindsight into adjacent squares, allowing you to see invisible or concealed creatures perfectly.
\parhead{Normal}
You have a 50\% chance to miss opponents you can't see, and you are \defenseless against them.

\feat{Brutal Magic}{Magic, Style}
\featpres 4th level spells.
\featben At the end of each round, if you dealt damage with a \glossterm{strike} that round, you gain a \plus2 bonus to spellpower during the next round.
This bonus increases by \plus1 when you gain access to 6th and 8th level spells.

\feat{Calculated Strike}{Combat, Precision, Strike}
\featpre Intelligence 5, combat prowess 8.
\featben In place of two strikes, you can make a physical attack.
If you hit, you deal damage normally.
If your attack result also beats the target's Fortitude defense, any damage it takes this round in excess of its hit points is dealt as critical damage.
% clarify: healing magic offsets this, as normal. ``net'' damage?

\feat{Commanding Presence}{Magic}
\featpre 1st level or higher \glossterm{Compulsion} spell known.
\featben You gain a \plus2 bonus to Intimidate and Persuasion checks.
If you know a Compulsion spell of 5th level or higher, this bonus increases to \plus4.

\feat{Countering Strike}{Reaction, Strike}
\featpres Intelligence 5, combat prowess 8.
\featben In place of two strikes, you can make a physical attack.
If you hit, you deal damage normally.
If the target makes a physical attack against you during the current phase, this attack deals triple damage.

\feat{Celestial Blessing}{Bloodline, Celestial}
\featpres Nonevil alignment, Celestial Heritage.
\featben As a standard action, you can spend a celestial point to bless an ally, as the \spell{bless} spell.
If your celestial power is at least 10, this ability functions as \spell{mass bless} instead.

\feat{Celestial Body}{Bloodline, Celestial}
\featpres Nonevil alignment, Celestial Heritage.
\featben As a standard action, you can spend a celestial point to imbue your body with celestial resilience for a \durshort duration.
You gain damage reduction against physical damage equal to your celestial power.
This damage reduction allows you to ignore the first points of damage you take each round.
Evil-aligned attacks ignore this damage reduction and negate it for 1 round.
If you have five or more celestial feats, this ability becomes permanent.

\feat{Celestial Heritage}{Bloodline, Celestial}
\featpre Nonevil alignment.
\featben You have the blood of a celestial creature in your veins, granting you celestial power.
Your celestial power is equal to your Willpower, or your level \add the number of celestial bloodline feats you possess, whichever is higher.
You have a pool with a number of celestial points equal to the number of celestial bloodline feats you possess.

As a standard action, you can spend a celestial point to surround yourself in an aura of good for a \durshort duration.
During that time, whenever an evil creature within \rngclose range of you attacks you, make an attack against its Mental defense.
Your accuracy is equal to your celestial power.
Success means the attacking creature takes 1d8 divine damage per two celestial power.

If you have five or more celestial bloodline feats, this ability becomes permanent.

\feat{Celestial Smiting}{Bloodline, Celestial}
\featpres Nonevil alignment, Celestial Heritage.
\featben As a swift action, you can spend a celestial point to make your next physical attack a smite.
If your smite targets an evil creature, you gain a bonus to accuracy equal to the number of celestial bloodline feats you possess, and a damage bonus equal to your celestial power.

\feat{Celestial Soul}{Bloodline, Celestial}
\featpres Nonevil alignment, any three celestial feats.
\featben You gain spell resistance against evil spells and spells cast by evil creatures.
Your spell resistance is equal to 10 \add your celestial power.

\feat{Celestial Spell Conduit}{Bloodline, Celestial}
\featpres Nonevil alignment, Celestial Heritage.
\featben Whenever you cast a spell, you can heal an ally within \rngclose range for 1d6 hit points per level of the spell cast.
If you have five or more celestial bloodline feats, this healing increases to 1d8 hit points per level of the spell cast.

\feat{Celestial Wings}{Bloodline, Celestial}
\featpre Any three celestial bloodline feats.
\featben You gain feathery wings that sprout from your back.
These allow you to slow your fall so you fall no more than 60 feet per round, preventing you from taking falling damage.

If you are 6th level, you can use the wings to glide at a rate equal to your base land speed.
While gliding, you move forward each round at a rate equal to your base land speed, and you descend at a rate of 10 feet per round.

If you are 12th level, you gain a fly speed equal to your land speed with average maneuverability.
See \pcref{Flying}, for more details.
You can only fly for a number of rounds equal to 3 \add half your Willpower.
After that limit is reached, you must rest for 5 minutes before flying again.

If you are 18th level, you can fly for a number of minutes equal to 3 \add half your Willpower before resting.

\feat{Chaotic Mind}{Class}
\featpre Paladin level 3, Chaos devoted alignment.
\featben You are unaffected by effects which detect truth, lies, or alignment.
Such effects never detect you, just as if you were not there at all.

\feat{Chargebreaker}{Combat, Reaction, Strike}
\featpre Dex 5, combat prowess 8.
\featben In place of two strikes, you can make a physical attack.
If you hit, you deal damage normally.
If the target moved towards you during the current round, this attack deals triple damage.

\feat{Charming Wild Speech}{Class}
\featpre Wild speech ability, druid level 7
\featben  As a standard action, you can make a Nature power vs. Mental attack against one creature or object you are communicating with using your wild speech ability.
This consumes a use of your wild speech ability.
Success means the target is \charmed for the duration of the conversation, and for 1 hour thereafter.
Failure has no effect.

This ability is not a Mind effect, and can affect anything that the druid can use her wild speech to converse with.
The attack automatically succeeds against non-intelligent objects.

\feat{Cleave}{Combat, Power}
\featpre Str 3.
\featben If you make a melee \glossterm{strike} that knocks a creature unconscious or kills it, you can immediately make an extra melee strike against another creature you threaten.
The extra strike must be made with the same weapon, and uses the same accuracy, but it is otherwise treated as being a different attack.
You cannot move before making this extra strike.
There is no limit to the number of times you can use this feat per round.

\feat{Close-Quarters Fighting}{Combat, Finesse}
\featpre Dex 3.
\featben You reduce your penalties for \glossterm{squeezing} by 2. In addition, you reduce your penalty for attacking with non-light weapons in a grapple by 2.

\feat{Combat Casting}{Magic}
\featben You gain a \plus3 bonus to Concentration checks made to cast spells or use spell-like abilities.

\feat{Combat Mobility}{Combat, Mobility}
\featpre Dex 5, combat prowess 4.
\featben At the start of each phase, you may choose a creature you can see.
You can move through that creature's space this phase, treating it as difficult terrain.
\parhead{Special} If you have the Legendary Mobility feat, you do not treat that creature's space as difficult terrain.

\feat{Covering Fire}{Combat, Style}
\featpres Combat prowess 4.
\featben While in this style, if you hit a creature with a physical ranged attack, it is \impaired with physical attacks for 1 round.
\stylereq Wield a ranged weapon.

\feat{Counterspell}{Magic}
\featpres 2nd level spells, Spellcraft 10 ranks.
\featben As a standard action, you can attempt to counterspell a creature within \rngmed range of you.
If that creature casts a spell during the same phase, you can make a Spellcraft check to identify the spell as normal (see \pcref{Spellcraft}).

After attempting to identify the spell, you may cast any spell you know as a counterspell.
That spell's normal effect does not occur.
Instead, if you cast the same spell as your target, the target miscasts its spell.
If you cast a different spell, but one from the same spell school and of the same spell level or higher, you and your target make opposed spellpower checks.
If you win, the target miscasts its spell.
If you fail the spellpower check, if your spell is of a different spell school, or if your spell is lower level, the target's spell takes effect normally.

\parhead{Special}
Metamagic feats and similar spell augmentations are ignored for the purpose of countering spells.
The \spell{dispel magic} spell can be used to counter any spell, even higher level spells.

\feat{Counterstorm}{Awareness, Combat, Reaction, Style}
\featpres Perception 15, combat prowess 18.
\featben Whenever a foe misses you with a melee attack, you gain a cumulative \plus2 bonus to accuracy and damage on your next melee attack against that creature.
If this bonus is not used within 1 round, it is lost.
\stylereq Wield a melee weapon.

\feat{Craftcaster}{Magic}
\featpres Craft (any) 10 ranks, 2nd level spell known.
\featben When casting spells which create or transform a physical substance that you are trained to craft, you gain a \plus2 bonus to your spellpower.

\feat{Creative Magic}{Magic, Style}
\featpre 4th level spell known.
\featben You gain a \plus2 bonus to spellpower with all spells that you have not cast within the past 5 rounds.
This bonus increases by \plus1 when you gain access to 6th and 8th level spells.

\feat{Critical Rage}{Class, Rage}
\featpre Rage ability, barbarian level 10.
\featben When you enter a rage, you can use this feat. If you do, during the rage you increase your critical range by 1.
\parhead{Special} You can only gain the benefit of one Rage feat at a time.

\feat{Deadly Aim}{Combat, Precision, Style}
\featpres Perception 5, combat prowess 4.
\featben You gain a \plus2 bonus to physical ranged damage rolls.
In exchange, you take a \minus2 penalty to physical ranged accuracy.
This bonus increases by \plus1 for each strike you can make per round beyond the first.

\feat{Defensive Fighting}{Combat, Defense, Style}
\featben While in this style, you gain a \plus2 bonus to your physical defenses.
In exchange, you take a \minus2 penalty to your physical accuracy.
If your Intelligence is at least 5, you take no penalty to accuracy.
\stylereq Wield a melee weapon. You must make a melee attack or take the total defense action each round.

\feat{Deflect Arrows}{Combat, Finesse, Reaction}
\featpres Dex 3, Unarmed Proficiency.
\featben Once per round, when you would normally be hit with a ranged weapon, you may deflect it so that you take no damage from it.
You must have a free hand and be aware of the attack.
You can deflect a number of attacks equal to the number of strikes you can make.
\par Attempting to deflect a ranged weapon doesn't count as an action.
Unusually massive ranged weapons and ranged attacks generated by spell effects can't be deflected.
\par If your Dexterity is at least 10, you can catch the weapon instead of deflecting it.
As an immediate action, you can throw a caught weapon back at the original attacker.

\feat{Delusive Guise}{Magic}
\featpre 1st level or higher \glossterm{Delusion} spell known.
\featben You gain a \plus2 bonus to Bluff and Disguise checks.
If you know a Delusion spell of 5th level or higher, this bonus increases to \plus4.

\feat{Destructive Rage}{Class, Rage}
\featpre Rage ability, barbarian level 6.
\featben When you enter a rage, you can use this feat. If you do, during the rage your physical attacks ignore an amount of hardness equal to your Strength.
\parhead{Special} You can only gain the benefit of one Rage feat at a time.

\feat{Devastating Magic}{Magic, Style}
\featpre 8th level spells.
\featben Your spells always inflict their critical success effects if they succeed by any amount, rather than if they succeed by 10.
This has no effect on creatures whose level exceeds your spellpower.

\feat{Diehard}{General}
\featpre Con 10.
\featben You halve the penalties you take from having critical damage (see \pcref{Critical Damage}).

\feat{Dispel Mastery}{Magic}
\featpre 3rd level or higher \glossterm{Antimagic} spell known.
\featben You gain a \plus4 bonus to spellpower when dispelling effects and countering spells with \spell{dispel magic} and similar spells.

\feat{Disorienting Strike}{Combat, Strike}
\featpre Three strikes.
\featben In place of three strikes, you can make a physical attack.
If you hit, you deal damage normally.
If your attack result also beats the foe's Mental defense, it is \disoriented for 5 rounds.

\feat{Distracting Foe}{Combat, Reaction, Style}
\featben While in this style, foes you threaten take a \minus4 penalty to Concentration checks.
This penalty increases by 2 for each strike you can make per round beyond the first.
\stylereq Wield a melee weapon. You must make a melee attack each round.

\feat{Divine Health}{Class}
\featpre Paladin level 3.
\featben You are immune to poison and disease.

\feat{Divine Spellgift}{Spellgift}
\featpres 2nd level, Wil 1.
\featben You have inherent magic in your body, granting you divine power.
When you gain this feat, you choose a divine spell.
The spell's level must be no greater than half your total levels in classes that cannot cast spells.
In addition, its level cannot exceed your Willpower.
You can spend a spellgift point to use that spell as a spell-like ability.

Your spellpower with spellgifts is equal to your Willpower or your level, whichever is higher.
You have a pool with a number of spellgift points equal to twice the number of spellgift feats you possess.

\parhead{Special}
You may take this feat multiple times.
Each time, you choose a new spell to gain as a spell-like ability.
All spellgift feats you have use the same pool of spellgift points.
The number of spellgift feats you have cannot exceed your Willpower.

\feat{Dominating Wild Speech}{Class}
\featpre Wild speech ability, Charming Wild Speech, druid level 15
\featben When you use your Charming Wild Speech ability, if the attack succeeds, you can choose for the target to be \dominated instead of charmed.

\feat{Draconic Breath}{Bloodline, Dragon}
\featpres Con 3, any three dragon bloodline feats.
\featben You gain a breath weapon based on the type of dragon you chose for the Draconic Heritage feat.
The shape of the breath weapon is given on \trefnp{Dragon Types}: either a \arealarge, 5 ft. wide line or a \areamed cone.
At 11th level, the size increases to a 100 ft. long, 10 ft. wide line or a \arealarge cone.

When you use your breath weapon, you make a Reflex attack to deal damage to everything in the area.
It deals 1d8 damage per two levels.
A failed attack deals half damage.
Your accuracy is equal to your level or your Constitution, whichever is higher.
After using your breath weapon, you must wait 1d4 rounds before you can use it again.

\feat{Draconic Heritage}{Bloodline, Dragon}
\featben You have the blood of a dragon in your veins.
When you take this feat, choose a type of dragon.
You gain damage reduction against the damage type that dragon's breath weapon deals.
The value of the damage reduction is equal to 5 \mtimes the number of dragon bloodline feats that you have.
A list of dragons and their associated damage type is given below.

\begin{dtable}
    \lcaption{Dragon Types}
    \begin{dtabularx}{\columnwidth}{>{\lcol}X >{\lcol}X >{\lcol}X}
        \tb{Dragon} & \tb{Energy Type} & \tb{Breath Weapon} \\
        \hline
        Black & Acid & Line \\
        Blue & Electricity & Line \\
        Brass & Fire & Line \\
        Bronze & Electricity & Line \\
        Copper & Acid & Line \\
        Gold & Fire & Cone \\
        Green & Acid & Cone \\
        Red & Fire & Cone \\
        Silver & Cold & Cone \\
        White & Cold & Cone \\
    \end{dtabularx}
\end{dtable}

\feat{Draconic Might}{Bloodline, Dragon}
\featpre Any three dragon bloodline feats.
\featben Choose a physical attribute: Strength, Dexterity, or Constitution.
You gain a \plus1 bonus to that attribute.

\feat{Draconic Mind}{Bloodline, Dragon}
\featpre Any three dragon bloodline feats.
\featben Choose a mental attribute: Intelligence, Perception, or Willpower.
You gain a \plus1 bonus to that attribute.

\feat{Draconic Scales}{Bloodline, Dragon}
\featpre Draconic Heritage.
\featben You gain a \plus1 bonus to your Armor defense.

\feat{Draconic Senses}{Bloodline, Dragon}
\featpre Draconic Heritage.
\featben You gain low-light vision, allowing you to treat sources of light as if they had double their normal illumination range.
If you already have low-light vision, you double the benefit, allowing you to quadruple the illumination range of light sources.
If you have three or more dragon bloodline feats, you gain darkvision with a 50 foot range, or the range of your darkvision increases by 50 feet.

\feat{Draconic Spellpower}{Bloodline, Dragon}
\featpre Draconic Heritage.
\featben You gain a \plus1 bonus to spellpower with all spells and spell-like abilities.
If you have five or more dragon bloodline feats, this bonus increases to \plus2.
This bonus does not stack with the bonus from the Spell Specialization feat.

\feat{Draconic Voice}{Bloodline, Dragon}
\featpre Draconic Heritage.
\featben You gain a \plus2 bonus to Intimidate and Persuasion checks.
If you have three or more dragon bloodline feats, this bonus increases to \plus4.

\feat{Draconic Weapons}{Bloodline, Dragon}
\featpre Draconic Heritage.
\featben You gain a bite natural attack that deals 1d8 damage for a Medium creature.
If you have three or more dragon bloodline feats, you also gain a claw natural attack for each hand that deals 1d6 damage for a Medium creatures.

\feat{Draconic Wings}{Bloodline, Dragon}
\featpre Any three dragon bloodline feats.
\featben You gain leathery wings that sprout from your back.
These allow you to slow your fall so you fall no more than 60 feet per round, preventing you from taking falling damage.

If you are 6th level, you can use the wings to glide at a rate equal to your base land speed.
While gliding, you move forward each round at a rate equal to your base land speed, and you descend at a rate of 10 feet per round.

If you are 12th level, you gain a fly speed equal to your land speed with average maneuverability.
See \pcref{Flying}, for more details.
You can only fly for up to 5 rounds.
After that limit is reached, you must rest for 5 minutes before flying again.

If you are 18th level, you can fly for up to 5 minutes before resting.

\feat{Driving Assault}{Combat, Maneuver, Power, Strike}
\featpres Str 5, combat prowess 8.
\featben In place of two strikes, you can make a single melee attack.
If hit, you deal damage normally, and you can make an immediate shove attack against the struck foe.
You gain a bonus on the shove attack equal to the damage dealt.

\feat{Dwarven Resilience}{Racial}
\featpre Dwarf.
\featben You gain a \plus2 bonus to defenses against poisons, spells, and spell-like effects.

\feat{Elemental Speech}{Class}
\featpre Wild speech ability, druid level 9.
\featben You can use your wild speech ability to communicate with one element of the natural world: air, earth, fire or water.
You choose which element when you gain this feat.
A element's sense of its surroundings is limited, so it won't be able to give (or recognize) detailed descriptions of creatures or answer questions about events outside its immediate vicinity.
\parhead{Special} This feat can be taken multiple times.
Each time, you choose a different element.

\feat{Endless Rage}{Class, Rage}
\featpre Rage ability, barbarian level 18.
\featben When you enter a rage, you can use this feat. If you do, the rage becomes almost endless. At the end of each round, if you took damage and did not receive healing or gain temporary hit points, that round does not count against the duration of your rage.
\parhead{Special} You can only gain the benefit of one Rage feat at a time.

\feat{Energetic Substitution}{Metamagic}
\featpre 2nd level spell known.
\featben When casting a substituted spell, you can choose what kind of energy damage it deals: acid, cold, fire, or electricity.
This can only be applied to spells that originally dealt acid, cold, fire, or electricity damage.
A substituted spell uses up a spell slot one level higher than the spell's actual level.

\feat{Enhanced Senses}{Magic}
\featpre 1st level or higher \glossterm{Awareness} spell known.
\featben You gain a \plus2 bonus to Awareness and Sense Motive checks.
If you know a Divination spell of 5th level or higher, this bonus increases to \plus4.

\feat{Executioner}{Combat, Reaction}
\featpres Four strikes, any three Reaction feats.
\featben At the end of every round, you can make a free melee \glossterm{strike} against all creatures you threaten that have no hit points remaining.

\feat{Distant Spell}{Metamagic}
\featpre 2nd level spell known.
\featben A distant spell has its range doubled.
This metamagic can only be applied to spells with a range of \rngclose, \rngmed, \rnglong, or \rngext.
A distant spell uses up a spell slot one level higher than the spell's actual level.

\feat{Dual Performance}{Performance, Skill}
\featpres Perform 15 ranks, any three Performance feats.
\featben You can use two Performance feats as part of the same performance.
You can activate them both as part of the same action (if they require the same action to activate), and sustain both feats with a single performance.

\feat{Eye of the Storm}{Awareness, Combat, Defense, Style}
\featpre Perception 5, combat prowess 4.
\featben You reduce your overwhelm penalties by 1.
This penalty reduction increases to 2 at Perception 10, and to 3 at Perception 20.
If your overwhelm penalty is reduced to 0, you are not considered to be overwhelmed.
\featpre Wield a melee weapon.

\feat{Exotic Weapon Proficiency}{Combat}
You understand how to use exotic weapons in combat.
\featben You are proficient with exotic weapons from weapon groups that you are otherwise proficient with.
This allows you to defend yourself normally with melee weapons, and to attack with exotic ranged weapons without penalty.
\parhead{Normal}
You are not normally proficient with exotic weapons.
You cannot defend yourself with melee weapons you are not proficient with, and you take a \minus4 penalty to accuracy with ranged weapons you are not proficient with.

\feat{Explosive Miscast}{Magic}
\featpre 4th level spells.
\featben The area affected by your explosive miscasts increases to an \areasmall radius burst, centered on you.
In addition, your explosive miscasts do not affect your allies other than you.

\feat{Extra Ability}{Class}
Choose a class ability with limited uses.
\featpre The chosen class ability.
\featben You gain additional uses of your chosen class ability. The number of uses you gain depends on the class ability chosen, as shown on \trefnp{Extra Ability Uses}.
\begin{dtable}
    \lcaption{Extra Ability Uses}
    \begin{dtabularx}{\columnwidth}{l l X}
        \tb{Class} & \tb{Ability} & \tb{Effect} \\
        Barbarian & Rage & \plus1 use per day \\
        Cleric & Devotion & \plus2 devotion points \\
        Druid & Wild speech & \plus2 uses per day \\
        Fighter & Combat discipline & \plus2 uses per day \\
        Monk & Manifest \ki & \plus2 uses per day \\
        Paladin & Divine invocation & \plus2 uses per day \\
        Ranger & Wild speech & \plus2 uses per day \\
    \end{dtabularx}
\end{dtable}

\feat{Fae Heritage}{Bloodline}
\featben You have the blood of a fae creature in your veins, granting you fae power.
As a standard action, you can spend a fae point to create sound or light, as the \spell{create sound} or \spell{dancing light} spells.
Your spellpower with this spell-like ability is equal to your fae power.

Your fae power is equal to your Willpower, or your level \add the number of fae bloodline feats you possess, whichever is higher.
You have a pool with a number of fae points equal to the number of fae bloodline feats you possess.

In addition, you may use your fae power to determine your spellpower with spellgifts.

\feat{Far Shot}{Combat, Power}
\featpre Str 3.
\featben When you use a projectile weapon, such as a bow, its range increment increases by one-half (multiply by 1-1/2).
When you use a thrown weapon, its range increment is doubled.

\feat{Fascinating Performance}{Performance, Skill}
\featpre Perform 5 ranks.
\featben As a standard action, you can make a DR 8 Perform check to fascinate another creature within \rngmed range of you.
For every 5 points by which your Perform check succeeds, you can affect an additional target.
If your check result beats a target's Mental defense, it is \fascinated by you as long as you sustain the performance.
Otherwise, it is unaffected.
This is a \glossterm{Compulsion}, \glossterm{Mind} effect.
\parhead{Special}
You can use any combination of performance feats you possess a number of times per day equal to twice the number of performance feats you have.
If you fail the Perform check, the use of the ability is wasted.

\feat{Fearless}{General}
\featpre Willpower 10.
\featben You are immune to fear effects.

\feat{Fearsome Rage}{Class, Rage}
\featpre Rage ability, barbarian level 10.
\featben When you enter a rage, you can use this feat. If you do, during the rage your attacks can inflict fear on your foes.
Once per round, when you hit with a melee strike, you may cause the target to be \shaken by you for 2 rounds.
\parhead{Special} You can only gain the benefit of one Rage feat at a time.

\feat{Fire Mastery}{Magic}
\featpre 3rd level or higher \glossterm{Fire} spell known.
\featben Whenever you get a critical success with a Fire spell, the target of the spell is \ignited for 5 rounds.

\feat{Flameborne Magic}{Magic, Style}
\featpre 4th level or higher \glossterm{Fire} spell known.
\featben As long as you are within \rngmed range of an active fire of at least Medium size, you gain a \plus2 bonus to spellpower with Fire spells.
If the source is at least Large size, this bonus applies to all spells you cast.
This bonus increases by \plus1 when you gain access to 6th and 8th level spells.

\feat{Focused Strike}{Combat, Precision, Strike}
\featpre Two strikes.
\featben In place of two strikes, you can make a physical attack with a \plus5 bonus to accuracy.
If you hit, you deal damage normally.
You may trade additional strikes to make this attack.
For each strike you trade, you gain an additional \plus5 bonus to accuracy.

\feat{Foresight}{Magic}
\featpre 3rd level or higher Divination spell.
\featben You receive instant, intuitive warnings of incoming attacks, even attacks you can't see or detect.
As a result, you are normally never \unaware.
Some creatures immune to Divination spells may still catch you unaware.

\feat{Fortune's Friend}{General}
\featpre One general \glossterm{legend point}.
\featben You can spend a legend point to cause a fortunate turn of events for you or your allies.
The effects of this ability can vary, and are determined by the game master.
The event is always beneficial for you, and you can suggest possible lucky events.
In general, the power and scope of the event should not exceed that of a spell or ritual with a level equal to half your level.
Some examples of events this ability could cause are given below.
\begin{enumerate}
    \item A cart carrying gold overturns, distracting an angry mob so your party can escape.
    \item A guard accidentally drops the keys to the vault he is guarding.
    \item A conversation you are overhearing happens to include exactly the juicy gossip you are looking for.
\end{enumerate}

You can only use legend points gained from your character level for this ability.
Bonus legend points from any other source cannot be used.

\feat{Frenzy}{Class, Rage}
\featpre Rage ability, barbarian level 6.
\featben When you enter a rage, you can use this feat. If you do, you take no penalty to physical defenses while raging, but you take double damage for not attacking during a round.
\parhead{Special} You can only gain the benefit of one Rage feat at a time.

\feat{Frigid Forge}{Magic}
\featpre 2nd level or higher Cold spell known.
\featben You can use ice as a crafting material in place of bone, ceramics, metal, or stone.
Items created in this way follow the normal rules for crafting items (see \pcref{Craft}), with three exceptions.
First, these items melt if they leave your possession after approximately one hour, depending on the ambient temperature.
Second, creating items in this way requires water or ice in place of the normal equipment and materials (such as a forge).
Third, you gain a bonus on the Craft check equal to your spellpower with Cold spells.

If water enhanced with the \spell{enhance component} ritual is used, the item created will have an enhancement bonus.

\feat{Furious Style}{Class, Rage}
\featpre Rage ability, barbarian level 6.
\featben When you enter a rage, you can use this feat. If you do, during the rage you can initiate or change combat styles as part of the swift action you use to sustain your rage.
\parhead{Special} You can only gain the benefit of one Rage feat at a time.

\feat{Fungal Poisoner}
\featpre Survival 5 ranks.
\featben You can use Survival to find and use fungus-based poisons.
If you spend an hour harvesting fungi, you can make a Survival check to create fungal poisons (see \pcref{Creating Poisons}).
You can harvest fungi in almost all areas with plant life.
In addition, you can use Survival in place of Craft (poison) for the purposes of feats and abilities.
If you do, those feats and abilities only affect fungal poisons.

\feat{Gang Up}{Combat}
\featpre Combat prowess 4.
\featben You count as one additional creature for the purpose of determining overwhelm penalties.
This does not allow you to overwhelm a creature by yourself.

\feat{Giantfighter}{Combat, Racial}
\featpres Dwarf, gnome, or halfling.
\featben You gain a \plus2 bonus to physical defenses against creatures of size Large or larger.

\feat{Guardian}{Combat, Defense, Style}
\featben Allies adjacent to you reduce their overwhelm penalties by 2.
If this effect reduces an ally's overwhelm penalty to 0, the ally is not considered to be overwhelmed.
This penalty reduction increases by \plus1 for each strike you can make per round beyond the first.
\stylereq Wield a melee weapon.
You must make a melee attack each round.

\feat{Heartseeker}{Combat, Style}
\featpre Combat prowess 8.
\featben While in this style, you increase your critical range with any weapon you wield by 1.
\stylereq Wield a weapon.

\feat{Heavy Hitter}{Combat, Power}
\featpre Strength 5, combat prowess 4.
\featben You gain a \plus2 bonus to damage with heavy melee weapons.

\feat{Hide in Plain Sight}{Skill}
\featpre Stealth 20 ranks.
\featben You can use the Stealth skill to hide even while being observed, taking a \minus5 penalty to the Stealth check.
You still need cover or concealment to hide.

\feat{Hurricane Strike}{Combat, Strike}
\featpres Dex 15, Whirlwind Strike, combat prowess 18.
\featben As a full-round action, you can make two melee strikes against each foe you threaten. You cannot use Strike feats with these strikes.
If your Dexterity is at least 20, you can use this feat as a standard action.

\feat{Hydra Strike}{Combat, Strike}
\featpre Three strikes, Strength 5, Dexterity 5, Constitution 5.
\featben In place of three strikes, you can make a physical attack.
If you hit, you deal damage normally.
You deal double damage for each additional defense you beat (Maneuver, Reflex, Fortitude, and Mental), up to a maximum of five times your normal damage.
After you use this strike, you cannot use it again for 2 rounds.

\feat{Imbue Magic}{Item Creation}
\featpre 1st level spells or Craft (any) 5 ranks.
\featben You can imbue items with magic using your spells or crafting ability.
Imbuing an item with magic takes time and material components, as described in \pcref{Magic Item Creation}.

When you take this feat, you choose one school of magic for every 5 ranks that you have in each Craft skill.
You can craft items from those schools.
If you later gain additional Craft ranks, you gain new schools appropriately.

You can also mend a broken magic item if it is one that you could make.
Doing so costs a tenth of the raw materials and a quarter of the time it would take to craft that item in the first place.
You cannot mend a destroyed magic item.

\feat{Imbued Spellstrike}{Metamagic}
\featpre 2nd level spell known.
\featben As part of casting an imbued spellstrike spell, you can make a single attack with a weapon in your hand.
If the attack hits, and you maintain concentration on the spell, the spell takes effect.
The spell must target the struck creature, but otherwise functions as if it has been cast normally.
The imbuement fades away without effect at the end of the next round if you have not struck a foe.

Only spells which affect a single target and have a casting time no shorter than 1 standard action can be channeled in this way.
An imbued spellstrike uses up a spell slot one level higher than the spell's actual level.

\feat{Imbuement Admixture}{Item Creation}
\featpre 3rd level spells.
\featben You can blend two spells together to create magic items.
This allows you to meet item crafting prerequisites that you could not meet with a single spell.

\feat{Immobilizing Strike}{Combat, Strike}
\featpre Four strikes.
\featben In place of four strikes, you can make a single melee attack.
If you hit, you deal damage normally.
If your attack result also beats the foe's Fortitude defense, it is \immobilized for 5 rounds.

\feat{Improved Counterspell}{Magic}
\featpres Spellcraft 10 ranks, Counterspell.
\featben When counterspelling, you may attempt to counter spells of any school, regardless of which spell you cast.
In addition, your spells are treated as being one higher level than normal for the purpose of determining what spells they can counter.

\feat{Improved Dirty Trick}{Combat, Maneuver}
\featpre Combat prowess 4.
\featben When you perform a dirty trick on a cretaure successfully, the creature is \impaired for 1d4 rounds.
You also gain a \plus2 bonus to accuracy on dirty trick attacks.

\feat{Improved Disarm}{Combat, Finesse}
\featpre Combat prowess 4.
\featben When you disarm a creature, you can make the disarmed item land up to 15 feet away in a random direction.
You also gain a \plus2 bonus to accuracy on disarm attacks.

\feat{Improved Feint}{Combat, Finesse}
\featpre Combat prowess 4.
\featben When you feint against a creature, if you beat its Armor defense by 5 or more, you deal damage with your weapon normally.
You also gain a \plus2 bonus to accuracy on feint attacks.

\feat{Improved Grapple}{Combat, Maneuver}
\featpre Combat prowess 4.
\featben Grappling does not make you \defenseless against other creatures.
You also gain a \plus2 bonus to accuracy on grapple attacks.
This does not prevent you from being defenseless if you have no hands free to defend yourself with.
\parhead{Normal}
While grappling, you are defenseless against all creatures except the ones you are grappling.

\feat{Improved Overrun}{Combat, Maneuver}
\featpre Combat prowess 4.
\featben When you attempt to overrun a foe, the creature may not choose to avoid you unless you let it.
You also gain a \plus2 bonus to accuracy on overrun attacks.
\parhead{Normal}
Without this feat, the target of an overrun can choose to avoid you or to block you.

\feat{Improved Shove}{Combat, Maneuver}
\featpre Combat prowess 4.
\featben When you shove a creature, you can move it outside of your reach.
You also gain a \plus2 bonus to accuracy on shove attacks.

\feat{Improved Trip}{Combat, Maneuver}
\featpre Combat prowess 4.
\featben When you successfully trip a foe, you get a free melee attack against it.
This attack is made as the creature is being tripped, so it does not have penalties for being prone.
You also gain a \plus2 bonus to accuracy on trip attacks.

\feat{Inescapable}{Combat, Style}
\featpre Combat prowess 8.
\featben While in this style, all squares you threaten cost four times the normal movement cost to move out of.
\stylereq Wield a melee weapon.
You must make a melee attack each round.

\feat{Infuriating Strike}{Combat, Strike}
\featpre Combat prowess 8.
\featben In place of two strikes, you can make a single physical attack.
If you hit, you deal damage normally, and the target is \taunted by you for 5 rounds.

\feat{Insightful Fighter}{Class}
\featpre Adaptive style ability.
\featben You can use Perception in place of Intelligence for your adaptive style feat pool.
You can also use Perception in place of Intelligence for the battlemaster ability, if you have that ability.

\feat{Insightful Rogue}{Class}
\featpre Combat trick ability.
\featben You can use Perception in place of Intelligence for combat tricks that depend on Intelligence, and for your accuracy with combat tricks.

\feat{Inspire Competence}{Performance, Skill}
\featpre Perform 5 ranks.
\featben As a standard action, you can make a DR 8 Perform check to inspire competence in another willing creature within \rngmed range of you.
For every 5 points by which your Perform check succeeds, you can affect an additional target.
Each target gains a bonus offensive \glossterm{legend point} which it can spend as long as you sustain the performance.
If a target's legend point has not been used when you end the performance, the legend point is wasted.
This is a \glossterm{Mind} effect.

\parhead{Special} You can use any combination of performance feats you possess a number of times per day equal to twice the number of performance feats you have.
If you fail the Perform check, the use of the ability is wasted.

\feat{Inspire Courage}{Performance, Skill}
\featpre Perform 9 ranks.
\featben As a standard action, you can make a DR 16 Perform check to inpsire courage in another willing creature within \rngmed range of you.
For every 5 points by which your Perform check succeeds, you can affect an additional target.
Each target gains temporary hit points equal to twice your Perform ranks.
This is a \glossterm{Mind} effect.
\parhead{Special} You can use any combination of performance feats you possess a number of times per day equal to twice the number of performance feats you have.
If you fail the Perform check, the use of the ability is wasted.

\feat{Inspire Mastery}{Performance, Skill}
\featpre Perform 17 ranks, any three other Performance feats.
\featben As a standard action, you can make a DR 32 Perform check to inspire mastery in another willing creature within \rngmed range of you.
For every 5 points by which your Perform check succeeds, you can affect an additional target.
As long as you sustain the performance, at the start of each round, each target gains a bonus general \glossterm{legend point} which it can spend during that round.
If a target's legend point has not yet been used when the round ends, the legend point is wasted.
This is a \glossterm{Mind} effect.
\parhead{Special} You can use any combination of performance feats you possess a number of times per day equal to twice the number of performance feats you have.
If you fail the Perform check, the use of the ability is wasted.

\feat{Inspire Resilience}{Performance, Skill}
\featpre Perform 13 ranks.
\featben As a standard action, you can make a DR 24 Perform check to inspire resilience in another willing creature within \rngmed range of you.
For every 5 points by which your Perform check succeeds, you can affect an additional target.
Each target gains damage reduction against all damage equal to your Perform ranks.
In addition, it gains a bonus defensive \glossterm{legend point} which it can spend as long as you sustain the performnace.
If a target's legend point has not been used when you end the performance, the legend point is wasted.
This is a \glossterm{Mind} effect.
\parhead{Special} You can use any combination of performance feats you possess a number of times per day equal to twice the number of performance feats you have.
If you fail the Perform check, the use of the ability is wasted.

\feat{Inspire Serenity}{Performance, Skill}
\featpre Perform 13 ranks.
\featben As a standard action, you can make a DR 24 Perform check to clear the mind of a willing creature within \rngmed range of you.
For every 5 points by which your Perform check succeeds, you can affect an additional target.
Each target is immune to all hostile Mind effects as long as you sustain the performance.
This is a \glossterm{Mind} effect.
\parhead{Special}
You can use any combination of performance feats you possess a number of times per day equal to twice the number of performance feats you have.
If you fail the Perform check, the use of the ability is wasted.

\feat{Instant Adaptation}{Class}
\featpre Adaptive combat ability, Rapid Adaptation, Intelligence 10, fighter level 18.
\featben As a swift action, you can exchange a single adaptive combat feat from your adapative feat pool.

\feat{Intimidating Strike}{Combat, Strike}
\featpres Two strikes, Intimidate 10 ranks.
\featben In place of two strikes, you can make a single physical attack.
If you hit, you deal damage normally, and you can make an Intimidate check against the target to make it \shaken for 5 rounds.
You gain a bonus on the check equal to the damage dealt.

\feat{Invulnerable Rage}{Class, Rage}
\featpre Rage ability, barbarian level 10.
\featben When you enter a rage, you can use this feat. If you do, during the rage your barbarian damage reduction is doubled and applies to all damage, not just physical damage.
\parhead{Special} You can only gain the benefit of one Rage feat at a time.

\feat{Iron Will}{General}
\featpre Intelligence 3 or Willpower 3.
\featben You gain a \plus2 bonus to your Mental defense.
This bonus can increase your hit points (see \pcref{Hit Points}).
\parhead{Special} This feat can be taken multiple times. Its effects stack.
You may take this feat a maximum number of times equal to half your Intelligence or half your Willpower, whichever is higher.

\feat{Ki Focus}{Class}
\featpre Manifest \ki ability, monk level 8.
\featben Choose one \ki manifestation that you know.
The manifestation must normally last until the end of the round, and the minimum monk level required to learn the manifestation cannot exceed half your monk level.
When you activate it, you can spend two ki points instead of one.
If you do, the ability instead lasts for 5 rounds.
\parhead{Special} This feat can be taken multiple times.
Each time, you choose a different \ki manifestation you know.

\feat{Ki Specialization}{Class}
\featpre Ki Focus, Manifest \ki ability, monk level 12.
\featben Choose a \ki manifestation that you chose with the Ki Focus feat.
When you activate it, you can spend three ki points instead of one.
If you do, the ability instead lasts for 12 hours.

While you have the manifestation active, your maximum \ki points are reduced by 1.
You can dismiss the manifestation to restore your \ki points.

\feat{Ki Training}{Class}
\featpre Manifest \ki ability.
\featben You can learn an additional \ki manifestation.
You must meet all prerequisites for the ability, as normal.
\parhead{Special} This feat can be taken multiple times.
Each time, you choose a different \ki manifestation to learn.

\feat{Killing Blow}{Combat, Precision, Strike}
\featpres Intelligence 10, combat prowess 12.
\featben In place of three strikes, you can make a single physical attack.
If you hit, you deal damage normally.
If the target took damage from this strike and has no hit points remaining at the end of the current phase, it dies.

\feat{Knockdown}{Combat, Maneuver, Power, Strike}
\featpre Strength 5, combat prowess 8.
\featben In place of two strikes, you can attempt to shove an opponent.
If you succeed, it is knocked prone at the end of its movement, and you can make an extra melee attack against it if you still threaten it.
The free attack gains the normal benefit for striking a prone target (\minus4 to physical defenses).

\feat{Ledgewalker}{Skill}
\featpre Acrobatics 5 ranks.
\featben Using Acrobatics to move along narrow surfaces does not reduce your speed.
\parhead{Normal} You move at half speed while balancing on precarious surfaces.

\feat{Legendary Awareness}{Skill}
\featpre Awareness 15 ranks.
\featben You gain one of the following senses: blindsense (20 ft.), darkvision (50 ft.), scent, or tremorsense (20 ft.).
If you choose a sense you already possess, its range doubles.

\feat{Legendary Balance}{Skill}
\featpre Acrobatics 15 ranks.
\featben You can balance on surfaces that cannot support your weight.
The DR is 30 for liquids such as water, 40 for dense gases and raw energy, and 50 for ordinary air.
While balancing in this way, you must take a move action each round to continue moving; you cannot remain in the same place in consecutive rounds, or you will fall.
The DR increases by 2 for each consecutive round that you spend balancing in this way.
You gain a \plus4 bonus on this check per size category smaller than Medium, or a \minus4 penalty per size category larger than Medium.

\feat{Legendary Climber}{Skill}
\featpre Climb 15 ranks.
\featben You gain a \glossterm{climb speed} equal to your land speed.
This grants several benefits.
\begin{itemize}
    \item A successful Climb check allows you to move a distance equal to your climb speed.
    \item You gain a \plus10 bonus to Climb checks.
\end{itemize}

In addition, you can now climb surfaces that are perfectly smooth.
The DR is 30 for perfectly smooth vertical surfaces, and 40 to climb on a perfectly smooth ceiling.
You can also wallrun on ceilings.

\feat{Legendary Combat Awareness}{Awareness, Combat}
\featpres Perception 10, combat prowess 12, any three Awareness feats.
\featben You reduce your overwhelm penalties by 4.
If your overwhelm penalty is reduced to 0, you are not considered to be overwhelmed.

\feat{Legendary Constitution}{General}
\featpres 11th level, Constitution 6.
\featben You gain a \plus5 bonus to Constitution-based checks.
In addition, you treat your Constitution as being 5 higher for the purpose of feats of endurance such as holding your breath or sprinting.

\feat{Legendary Craftsman}{Skill}
\featpre Craft (any) 15 ranks.
\featben You can craft items with whatever materials you have on hand.
When making an item, if your Craft check is 30 or higher, you use half the costly material components to make the item, replacing them with other, less expensive components.
For every 10 points by which the check exceeds 30, you halve the required material components again.
This only applies once per item; use the highest result achieved while making that item.

This feat does not affect the material components required to craft magical items.

\feat{Legendary Devicesmith}{Skill}
\featpre Devices 15 ranks.
\featben You can disable spell effects on objects or areas as if they were merely complex devices.
You can make a Devices check against an active spell effect within your reach.
The DR to dispel the effect is equal to 20 \add the spells's spellpower, and the result must be at least 30.
Success means the spell is dispelled.

You must be aware of a spell to disable it, either through the Spellcraft skill or because the effect is noticeable.
You cannot disable spell effects on creatures.

\feat{Legendary Dexterity}{General}
\featpres 11th level, Dexterity 6.
\featben You gain a \plus5 bonus to Dexterity-based checks, as well as to initiative.

\feat{Legendary Disguise}{Skill}
\featpre Disguise 15 ranks.
\featben Your disguises can change the magic on a creature.
When you make a Disguise check, if the result is 30 or higher, you can decide how that creature and any items on the creature appear when examined by divination spells.
For example, you could cause all of its equipment to appear nonmagical, or you could cause it to have a strong aura of good when examined with \spell{detect good}.
You cannot create an aura of overwhelming strength with this skill.

Anyone using divination magic on the creature must make a spellpower check with a DR equal to your Disguise check in order to perceive the truth.
Regardless of the result of the check, the caster is not aware that the check was made.

\feat{Legendary Defense}{Combat, Defense}
\featpres Combat prowess 12, any three Defense feats.
\featben You gain a \plus2 bonus to physical defenses.

\feat{Legendary Escapist}{Skill}
\featpre Escape Artist 15 ranks.
\featben You can attempt to escape from magic itself, slipping hostile spells off of your body so they dissipate harmlessly.
As a standard action, you can make an Escape Artist check to throw off magical effects on you.
The DR to dispel the effect is equal to 20 \add the effect's spellpower, and the result must be at least 30.
Success means the spell is dispelled.

You must be aware of a spell to disable it, either through the Spellcraft skill or because the effect is noticeable.
You can only dispel spell effects which target you directly, not area effects which include you.
If a spell targets multiple creatures, you can only remove its effects on you.

\feat{Legendary Finesse}{Combat, Finesse}
\featpres Dexterity 10, combat prowess 12, any three Finesse feats.
\featben Whenever you miss with a \glossterm{strike}, the target still takes half damage from your attack.
This is not considered hitting with the attack for the purpose of effects which trigger on hitting a creature.

\feat{Legendary Handler}{Skill}
\featpre Creature Handling 15 ranks.
\featben You can pacify and push creatures as a swift action.
It takes no action for you to make a creature perform a trick it is trained in.

In addition, when training a creature, if your check result is 30 or higher, you can train it in only a hour.
The creature remembers the trick just as if you had spent the full amount of time training it.

\feat{Legendary Healer}{Skill}
\featpre Heal 15 ranks.
\featben As a standard action, you can make a Heal check on a creature you touch to heal wounds.
The target heals hit points equal to your check result.
For every five hit points you would restore, you can instead cure one point of critical damage.

You can use this ability three times per day.

\feat{Legendary Intelligence}{General}
\featpres 11th level, Intelligence 6.
\featben You gain a \plus5 bonus to Intelligence-based checks.
In addition, you can memorize any information perfectly to recall later.

\feat{Legendary Intimidation}{Skill}
\featpre Intimidate 15 ranks.
\featben When you demoralize an opponent, if your check result exceeds 30, the creature is \panicked instead of being shaken.

You can use this ability three times per day.

\feat{Legendary Knowledge}{Skill}
\featpre Knowledge (any) 15 ranks.
\featben You gain a \plus2 bonus on accuracy, checks, and defenses against non-humanoid creatures you identify with a successful Knowledge check.

\feat{Legendary Leaper}{Skill}
\featpre Jump 15 ranks.
\featben You are always treated as having a running start when jumping.

\feat{Legendary Legerdemain}{Skill}
\featpre Sleight of Hand 15 ranks.
\featben When you make a Sleight of Hand check to conceal or pickpocket an object, if the result is 30 or higher, you can hide the object into a pocket dimension.
You can retrieve the item later as a move action.
You may only have up to three items hidden in this way, none of which can be larger than one size category smaller than you.

\feat{Legendary Liar}{Skill}
\featpre Bluff 15 ranks.
\featben When you make a Bluff check, if the result is 30 or higher, you may treat it as an attack against the Mental defense of one creature that understands you.
If your attack succeeds against that creature, its sight, smell, hearing, and sense of temperature are altered so it perceives the world as you described it, as \spell{major image}.
This cannot remove things that do exist, but it can create new sensations where none existed.

You can use this ability three times per day.

\feat{Legendary Linguist}{Skill}
\featpre Linguistics 15 ranks.
\featben You can speak, read, and understand all languages.
This does not allow you to speak with creatures that lack a language.
Certain extremely obscure languages may be beyond your knowledge.

\feat{Legendary Maneuver Master}{Combat}
\featpres Combat prowess 12, any three Maneuver feats.
\featben If you succeed at a combat maneuver attack, if your attack result also beats the target's Armor defense, you deal normal damage with the weapon used to perform the maneuver in addition to gaining the successful effects of the maneuver.
If the maneuver was performed without a weapon, you deal damage equivalent to an unarmed attack.

\feat{Legendary Mobility}{Combat, Mobility}
\featpres Dexterity 10, combat prowess 12, any three Mobility feats.
\featben You can move through spaces occupied by your foes.
You treat spaces occupied by foes as difficult terrain.

\feat{Legendary Mounted Warrior}{Combat, Mounted}
\featpres Combat prowess 12, any three Mounted feats, Ride 10 ranks.
\featben When you take damage, you may choose to have your mount suffer half the damage instead of you (rounded down).
Likewise, when your mount takes damage, you may choose to suffer half of that damage instead of your mount (rounded down).

\feat{Legendary Persuasion}{Skill}
\featpre Persuasion 15 ranks.
\featben When you make a Persuasion check, if the result is 30 or higher, you may treat it as an attack against the Mental defense of one creature that understands you.
If your attack succeeds against that creature, it is compelled to obey you, as \spell{suggestion}.

You can use this ability three times per day.

\feat{Legendary Power}{Combat, Power}
\featpres Strength 10, combat prowess 12, any three Power feats.
\featben You can use weapons as if they were one category less encumbering than they actually are.
The weapon encumbrance categories are light, medium, and heavy.
For example, you can use a greatsword as a medium weapon, allowing you to wield it in one hand without suffering any penalties.

In addition, you increase your damage dice with unarmed attacks and natural weapons by one size category.

\feat{Legendary Precision}{Combat, Precison}
\featpres Perception 10, combat prowess 12, any three Precision feats.
\featben When you make a physical attack, if you hit by 5 or more, you deal maximum damage with your weapon.

\feat{Legendary Profession}{Skill}
\featpre Profession (any) 15 ranks.
\featben You are well-known for being an unparalleled master of your chosen profession.

\feat{Legendary Reaction}{Combat, Reaction}
\featpres Dexterity 10, combat prowess 12, any three Reaction feats.
\featben The first immediate action you take each round does not count against your actions that round.
This allows you to use two immediate actions in the same round, or to use both an immediate action and a swift action.
However, you still cannot use two immediate actions in response to the same triggering action.

\feat{Legendary Rider}{Skill}
\featpre Ride 15 ranks.
\featben You can attempt to ride unwilling creature.
You can only ride creatures of your size category or up to two size categories larger.
To ride an unwilling creature, you must beat its Maneuver defense with a Ride check.

Riding a creature is like grappling the creature, with the following changes.
\begin{itemize}
    \item You share space with the creature you ride, just like riding a normal mount.
        If your mount breaks the grapple, you move to an adjacent unoccupied square of your choice.
        If there are no adjacent unoccupied squares, you in the same space as your mount, squeezing as necessary.
    \item You make Ride checks instead of grapple attacks to remain in the grapple and take actions in the grapple.
    \item You cannot pin the creature.
\end{itemize}

\feat{Legendary Sense Motive}{Skill}
\featpre Sense Motive 15 ranks.
\featben When you make a Sense Motive check, if the result is 30 or higher, you may treat it as an attack against the Mental defense of one creature that you can see.
If your attack succeeds against that creature, you read its mind, as \spell{read mind}.

You can use this ability three times per day.

\feat{Legendary Spellcraft}{Skill}
\featpre Spellcraft 15 ranks.
\featben You gain a \plus2 bonus on accuracy, checks, and defenses against spells and magical effects you identify with a successful Spellcraft check.

\feat{Legendary Sprinter}{Skill}
\featpre Sprint 15 ranks.
\featben You can sprint for a number of minutes equal to 5 \add your Constitution.
After you finish sprinting, you are \fatigued for 1 round per minute of sprinting.

\feat{Legendary Stealth}{Skill}
\featpre Stealth 15 ranks.
\featben When you make a Stealth check to hide, if the result is 30 or higher, you can hide even from a creature of your choice that is observing you.
You must still have passive cover or concealment from that creature.
For every 5 points by which your check result exceeds 30, you can hide from an additional creature that is observing you.

\feat{Legendary Strength}{General}
\featpres 11th level, Strength 6.
\featben You gain a \plus5 bonus to Strength-based checks.
In addition, you treat your Strength as being 5 higher for the purpose of determining your carrying and lifting capacity.

\feat{Legendary Strike}{Combat}
\featpres Combat prowess 12, any three Strike feats.
\featben You gain a \plus2 bonus to accuracy with strikes.

\feat{Legendary Style}{Combat}
\featpres Combat prowess 12, any three Style feats.
\featben You may have two combat styles active at once.
Both styles can be initiated or changed as part of the same swift action.
You may sustain this ability for a number of rounds each day equal to the number of Style feats you have.

\feat{Legendary Survival}{Skill}
\featpre Survival 15 ranks.
\featben You ignore difficult terrain and harmful natural terrain of any kind, and are immune to harmful planar effects.

In addition, as a standard action, you can find your way to any location, as \spell{find the path}.
You may use this ability once per day.

\feat{Legendary Swimmer}{Skill}
\featpre Swim 15 ranks.
\featben You gain a \glossterm{swim speed} equal to your land speed.
This grants several benefits.
\begin{itemize}
    \item A successful Swim check allows you to move a distance equal to your swim speed.
    \item You gain a \plus10 bonus to Swim checks.
\end{itemize}

When swimming, if your check result is 30 or higher, you do not suffer any penalties to physical melee attacks, checks, or physical defenses for being underwater for that round.
You still suffer the normal penalty with underwater ranged attacks.

\feat{Legendary Tumbler}{Skill}
\featpre Acrobatics 15 ranks.
\featben When tumbling, if your check result is 30 or higher, you can move through squares occupied by foes or threatened by blocking foes without penalty.

\feat{Legendary Unarmed Warrior}{Combat}
\featpre Combat prowess 12, Unarmed Dodge, Unarmed Might, Unarmed Proficiency
\featben Your unarmed attacks deal damage as if you were four size categories larger than normal (1d10 for a Medium creature, or 1d8 for a small creature).
This does not stack with the benefits of Unarmed Might.

\feat{Legendary Versatility}{Class}
\featpre At least 5 levels in two different classes, one of which must be your base class.
\featben You treat all classes you have at least 5 levels in as base classes.
This allows you to gain the base class abilities of multiple classes.

\feat{Lifebound}{General}
\featpre Constitution 10.
\featben You become immune to hostile \glossterm{Life} and \glossterm{Death} effects.

\feat{Lightning Reflexes}{General}
\featpre Dexterity 3 or Perception 3.
\featben You gain a \plus2 bonus to your Reflex defense.
\parhead{Special} This feat can be taken multiple times. Its effects stack.
You may take this feat a maximum number of times equal to half your Dexterity or half your Perception, whichever is higher.

\feat{Magical Synthesis}{Magic}
Choose two magical classes you possess.
\featpres
Levels in two magical classes.
\featben When gaining levels in either of your chosen classes, you increase your spellcasting ability with the other class as if your class was nonmagical.
See \pcref{Spellcasters and Multiclassing}.

\feat{Manyshot}{Combat, Finesse}
\featpres Dex 5, combat prowess 8.
\featben When you attack with a projectile weapon, you may fire two projectiles at once.
If you do, you gain a \plus2 accuracy bonus, and you make a dual attack (see \pcref{Dual Attacking}).
Taking advantage of this style requires loading two projectiles at once, which may be difficult if you reload slowly.

\feat{Mass Counterspell}{Magic}
\featpres Spellcraft 15 ranks, 4th level spells, Counterspell.
\featben When you use the Counterspell feat, you may counter up to five creatures within \rngmed range of you.
You may still cast only one spell, but it is used to counter the spells cast by all of your targets.

\feat{Mass Transporter}{Magic}
\featpre 5th level or higher \glossterm{Teleportation} spell known.
\featben When casting Conjuration (Translocation) spells and rituals which affect multiple creatures, you can affect a number of creatures equal to your spellpower.

\feat{Mighty Blows}{Combat, Power}
\featpres Str 3.
\featben You gain a \plus1 bonus to physical damage rolls.
\parhead{Special} This feat can be taken multiple times. Its effects stack.
You may take this feat a maximum number of times equal to half your Strength.

\feat{Mindless Rage}{Class, Rage}
\featpre Rage ability, barbarian level 10.
\featben When you enter a rage, you can use this feat. If you do, during the rage you are immune to \glossterm{Mind} spells and effects.
\parhead{Special} You can only gain the benefit of one Rage feat at a time.

\feat{Miscast Immunity}{Magic}
\featpre 8th level spells.
\featben You become immune to the effects of miscast spells.
They are unable to target you, damage you, or affect you in any way.
If this would make a spell impossible to miscast, such as if you are the only valid target of a retargeting miscast, the miscast has no effect.
You may still choose to be affected by miscast effects, if desired.

\feat{Mocking Performance}{Performance, Skill}
\featpre Perform 5 ranks.
\featben As a standard action, you can make a DR 8 Perform check to mock a creature within \rngmed range of you.
For every 5 points by which your Perform check succeeds, you can affect an additional target.
If your check result beats a target's Mental defense, it is \impaired with attacks and checks as long as you sustain the performance.
Otherwise, it is unaffected.
This is a \glossterm{Mind} effect.
\parhead{Special}
You can use any combination of performance feats you possess a number of times per day equal to twice the number of performance feats you have.
If you fail the Perform check, the use of the ability is wasted.

\feat{Mounted Archery}{Combat, Mounted}
\featpre Training in the Ride skill.
\featben The penalty you take when using a ranged weapon while mounted is decreased by 4: \minus0 instead of \minus4 if your mount is taking a double move, and \minus4 instead of \minus8 if your mount is sprinting.

\feat{Mounted Charge}{Combat, Mounted}
\featpre Ride 10 ranks.
\featben When you charge a creature with your mount, your weapon's damage die size increases by one per size category that your mount is larger than you (minimum one die size increase).
For example, a Medium human charging on a Large horse would increase his weapon's damage die by one size.

\feat{Mounted Defense}{Combat, Defense, Mounted}
\featpre Training in Ride.
\featben Once per round when your mount is hit in combat, you may attempt a Ride check as an immediate action to negate the hit.
Your check result is used in place of the mount's physical defenses, if it is higher.
If that would make the attack fail, the attack fails.

\feat{Mounted Trample}{Combat, Mounted, Maneuver}
\featpre Ride 10 ranks.
\featben When you attempt to overrun an opponent while mounted, your target may not choose to avoid you.
Your mount may make one attack with an appropriate natural weapon (hoof, claw, or other leg-based attack) against any target you knock down, gaining the normal \plus4 bonus to accuracy against prone targets.

\feat{Natural Spellgift}{Spellgift}
\featpres 2nd level, Wil 1.
\featben You have inherent magic in your body, granting you natural power.
When you gain this feat, you choose a nature spell.
The spell's level must be no greater than half your total levels in classes that cannot cast spells.
In addition, its level cannot exceed your Willpower.
You can spend a spellgift point to use that spell as a spell-like ability.

Your spellpower with spellgifts is equal to your Willpower or your level, whichever is higher.
You have a pool with a number of spellgift points equal to twice the number of spellgift feats you possess.

\parhead{Special}
You may take this feat multiple times.
Each time, you choose a new spell to gain as a spell-like ability.
You may use any combination of the spells a number of times per day equal to the number of spellgift feats you have.
The number of spellgift feats you have cannot exceed your Willpower.

\feat{Open Minded}{Skill}
\featben You gain two skill points.
You may spend these skill points immediately.

\feat{Overpowering Assault}{Combat, Power, Style}
\featpre Str 3.
\featben While in this style, you gain a \plus2 bonus to accuracy with maneuvers.
This bonus increases by \plus1 for each strike you can make per round beyond the first.
\stylereq You must make a melee attack each round.

\feat{Overpowering Rage}{Class, Rage}
\featpre Rage ability, barbarian level 6.
\featben When you enter a rage, you can use this feat. If you do, during the rage you add your rage bonus to your Maneuver defense and your accuracy with combat maneuvers.
\parhead{Special} You can only gain the benefit of one Rage feat at a time.

\feat{Overwhelming Fire}{Combat, Style}
\featpre Dex 3.
\featben While in this style, you contribute to overwhelm penalties against all creatures within a single range increment of you.
You do not contribute to overwhelm penalties against creatures with cover from you.
For each strike you can make per round beyond the first, you threaten creatures an additional range increment away.
\stylereq Wield a ranged weapon.
You must make a physical ranged attack each round.

\feat{Overwhelming Rage}{Class, Rage}
\featpre Rage ability, barbarian level 6.
\featben When you enter a rage, you can use this feat. If you do, during the rage overwhelmed foes you threaten increase their overwhelm penalties by 1.
\parhead{Special} You can only gain the benefit of one Rage feat at a time.

\feat{Parry}{Combat, Defense, Reaction, Style}
\featpre Dexterity 3.
\featben While in this style, whenever a physical melee attack is made against you that you are aware of, you may make an attack roll to attempt to parry the attack.
You may use your attack result in place of any or all physical defenses against the parried attack.
You can make a maximum number of parry attempts per round equal to the number of attacks you can make.

In exchange, you take a \minus2 penalty to accuracy with physical attacks.

\stylereq Wield a melee weapon.
You must make a melee attack or take the total defense action each round.

\feat{Patient Rage}{Class, Rage}
\featpre Rage ability, barbarian level 2.
\featben When you enter a rage, you can use this feat. If you do, during the rage not attacking does not cause you to take damage at the end of each round.
\parhead{Special} You can only gain the benefit of one Rage feat at a time.

\feat{Patient Strike}{Reaction, Strike}
\featpre Intelligence 10, combat prowess 12.
\featben In place of three strikes, you can make a physical attack.
If you hit, you deal damage normally.
If the target does not attack you during the current phase, this attack deals quadruple damage.

\feat{Peaceful Magic}{Magic, Style}
\featpre 4th level or higher \glossterm{Abjuration} spell known.
\featben If you have not taken damage within the past 5 rounds, you gain a \plus2 bonus to spellpower.
This bonus increases by \plus1 when you gain access to 6th and 8th level spells.

\feat{Penetrating Strike}{Combat, Strike}
\featpre Three strikes.
\featben In place of three strikes, you can make a physical attack.
If you hit, you deal damage normally.
If your attack result also beats the foe's Fortitude defense, it loses any damage reduction it has for 5 rounds.

\feat{Personal Magic}{Magic}
\featben You have created or discovered a unique style of magic, with subtle differences from the magic most people have studied.
This has several benefits, which are described below.
\begin{itemize}
    \item You gain a \plus2 bonus to Concentration checks to cast spells and rituals.
    \item The DR to identify your spells and rituals with Spellcraft is increased by 10 (see \pcref{Spellcraft}).
    \item The cost for scribing rituals in your ritual book is halved, but the time required is doubled.
        This does not alter the cost to perform rituals.
\end{itemize}

\feat{Perfect Health}{General}
\featpre Constitution 3.
\featben You become immune to disease, except supernatural diseases such as mummy rot.
If your Constitution is 10, you are also immune to poison and supernatural diseases.

\feat{Persistent Spellstrike}{Metamagic}
\featpres 3rd level spell known, Imbued Spellstrike.
\featben This metamagic functions like Imbued Spellstrike, except that the imbuement lasts for 5 minutes if you have not struck a foe.
If the weapon leaves your hands or if you cast another spell, the imbuement fades away without effect.

A persistent spellstrike spell uses up a spell slot two levels higher than the spell's actual level.

\feat{Pierce Delusions}
\featpre Perception 5.
\featben You are immune to hostile \glossterm{Delusion} effects.

\feat{Pierce Wings}{Combat, Precision, Strike}
\featpre Perception 5, combat prowess 8.
\featben In place of two strikes, you can make a physical ranged attack.
If you hit, you deal damage normally.
If your attack result also beats the target's Fortitude defense, it loses its ability to fly for 5 rounds.
This only affects creatures who use wings or other physical means to fly, and has no effect on creatures with magical or supernatural flight.

\feat{Plant Speech}{Class}
\featpre Wild speech ability, druid level 5.
\featben You can use your wild speech ability to communicate with plants, in addition to animals.
A regular plant's sense of its surroundings is limited, so it won't be able to give (or recognize) detailed descriptions of creatures or answer questions about events outside its immediate vicinity.

\feat{Predictive Strike}{Reaction, Strike}
\featpre Intelligence 15, combat prowess 18.
\featben In place of four strikes, you can make a physical attack.
You also predict the type of action the foe will take during the current phase:
\begin{itemize}
    \item Physically attack
    \item Move
    \item Cast a spell
    \item None of the above
\end{itemize}
If you hit, you deal damage normally.
If the foe takes the type of action you predicted, your attack deals quintuple damage.
If the foe takes an action that fits into multiple action types, use all that apply.

\feat{Point Blank Shot}{Combat, Precision, Style}
\featben While in this style, you gain a \plus2 bonus to accuracy with physical ranged attacks when attacking targets within half of your range increment.
\stylereq Wield a ranged weapon.

\feat{Poison Expert}{Skill}
\featpres Intelligence 3, Craft (poison) 10 ranks.
\featben Whenever you use poisons, you can use your level or your Intelligence, whichever is higher, in place of the poison's normal potency.
In addition, you cannot accidentally poison yourself while applying poison to an object.

\feat{Power Attack}{Combat, Power, Style}
\featpres Str 5, combat prowess 4.
\featben You gain a \plus2 bonus to physical melee damage rolls.
In exchange, you take a \minus2 penalty to accuracy with physical melee attacks.
This bonus increases by \plus1 for each strike you can make per round beyond the first.

\feat{Precise Shot}{Combat, Precision, Style}
\featpres Perception 5, combat prowess 4.
\featben While in this style, your ranged attacks ignore cover and concealment, except total cover and total concealment.
In addition, when you shoot or throw ranged weapons at a grappling opponent, you automatically strike at the opponent you have chosen.
\stylereq Wield a ranged weapon.

\feat{Predictive Defense}{Awareness, Combat, Defense, Style}
\featpres Intelligence 5, combat prowess 4.
\featben You can use your Intelligence to determine your Armor and Maneuver defenses in place of your Dexterity or Constitution.
\stylereq Wield a melee weapon.
You must make a physical attack or take the total defense action each round.

\feat{Quick Draw}{Combat}
\featben You can draw light weapons as a free action and medium weapons as a swift action.
You can draw heavy weapons and hidden weapons of any type (see the Sleight of Hand skill) as a move action.
This allows you to throw light weapons at your full normal rate of attacks (much like a character with a bow).

If your \glossterm{combat prowess} is at least 8, you can draw light and medium weapons as an immediate action, and heavy and hidden weapons as a swift action.
\parhead{Normal}
Without this feat, you may draw light weapons as a swift action, medium and heavy weapons as a move action, and hidden weapons as a standard action.

\feat{Quicken Spell}{Metamagic}\label{Quicken Spell}
\featpre 3rd level spell known.
\featben Casting a quickened spell is a swift action.
You can perform another action, even casting another spell, in the same round as you cast a quickened spell.
However, casting a quickened spell is mentally exhausting.
In the round after you cast it, you cannot act during the action phase.

You may cast only one quickened spell per round.
A spell whose casting time is more than 1 standard action cannot be quickened.
A quickened spell uses up a spell slot two levels higher than the spell's actual level.

\feat{Racial Acclimation}{Racial}\label{Racial Acclimation}
\featpre Have a race other than your original race, such as by \spellindirect{reincarnate}{reincarnating}.
\featben You no longer suffer any penalties for being in a race other than your original race.
For example, this would remove the negative level for being \spellindirect{reincarnate}{reincarnated}.
\parhead{Special}
This feat can be retrained using normal retraining rules if you return to your original race.

\feat{Ranged Legerdemain}{Skill}
\featpre 2nd level or higher \glossterm{Telekinesis} spell known.
\featben By expending a Telekinesis spell of 2nd level or higher, you can use the Disable Device or Sleight of Hand skills at \rngclose range for a number of minutes equal to half the level of the spell slot.

\feat{Rapid Adaptation}{Class}
\featpre Adaptive combat ability, Intelligence 5, fighter level 8.
\featben You only need to spend 1 minute in training to change your adaptive combat feats.

\feat{Rapid Healing}{General}
\featpre Constitution 3.
\featben You naturally heal four times faster than normal.
This means you require only two hours of rest to heal half your hit points.
This stacks with the benefits of accelerating recovery with the Heal skill (see \pcref{Accelerate Recovery}).

\feat{Rapid Performance}{Performance, Skill}
\featpres Perform 15 ranks, any three Performance feats.
\featben You can use Performance feats as a swift action by taking a \minus10 penalty on the Perform check to activate the feat.

\feat{Reach Spell}{Metamagic}
\featpre 2nd level spell known.
\featben When casting a reach spell, you can use a spell with a range of touch on a target within \rngclose range.
You must succeed on a Spellpower vs. Reflex attack to touch the target.
A reach spell uses up a spell slot one level higher than the spell's actual level.

\feat{Reckless Attack}{Combat, Power, Style}
\featpre Combat prowess 4.
\featben While in this style, you are \defenseless.
In exchange, you gain a \plus2 bonus to damage with melee attacks.
\stylereq Wield a melee weapon.

\feat{Recover Performance}{Performance, Skill}
\featpres Perform 11 ranks, any other Performance feat.
\featben If you fail the Perform check to use a Performance feat, you can reroll the Perform check as an immediate action.

\feat{Resilient Magic}{Magic}
\featpre 3rd level or higher Antimagic spell known.
\featben Your spells persist for up to 5 rounds after being dispelled if they still have time remaining in their duration.
If you know an Antimagic spell of 7th level or higher, your spells cannot be dispelled by any means short of the \spell{disjoin magic} spell, or other effects of similar power.

\feat{Resist Maneuver}{Class, Fighter}
\featpre Fighter.
\featben Choose a combat maneuver.
You gain a \plus5 bonus to Maneuver Defense against that maneuver.

\feat{Retargeting Miscast}{Magic}
\featpre 3rd level spells.
\featben When you target a random creature with a spell's miscast effect, you may roll twice to determine which creature is affected.
You choose which result is used.

\feat{Retributive Counterspell}{Magic}
\featpre 4th level spell known.
\featben As part of the action to counter a spell, you may expend an Abjuration (Negation) spell of 4th level or higher.
If you do, the counterspelled spell is turned back on the caster as if it were affected by the \spell{spell turning}
spell.
If it cannot be affected by \spell{spell turning}, such as if it is a spell that only affects the caster, it is simply countered as normal.

\feat{Blood Magic}{Magic, Style}
\featpre 4th level Vivimancy spell known.
\featben At the end of each round, if you took at least as much lethal damage as your spellpower during that round, you gain a \plus2 bonus to spellpower during the next round.
In addition, as a swift action, you can deal yourself life damage equal to your spellpower.
This bonus increases by \plus1 when you gain access to 6th and 8th level spells.

\feat{Riposte}{Combat, Reaction}
\featpres Dexterity 3, Parry.
\featben When you successfully parry an attack from a foe you threaten, if your parry attempt exceeds your opponent's attack roll by 10 or more, your foe is \defenseless against you for 1 round.

\feat{Ritual Caster}{Magic}
\featpre Intelligence 3.
\featben You can learn and perform rituals as if you were an arcane caster with a spellpower equal to your level.
The maximum level of ritual that you can learn or perform is equal to half your level or your Intelligence, whichever is lower.

\feat{Savage Rage}{Class, Rage}
\featpre Rage ability, barbarian level 2.
\featben When you enter a rage, you can use this feat. If you do, during the rage benefit you gain the unarmed warrior monk ability (see Unarmed Warrior, \pref{Mnk:Unarmed Warrior}), increasing your power with unarmed attacks.
\parhead{Special} You can only gain the benefit of one Rage feat at a time.

\feat{Scale the Beast}{Skill}
\featpre Climb 10 ranks.
\featben As a standard action, you can make a Climb check against the Maneuver defense of a creature adjacent to you.
The creature must be three or more size categories larger than you.
You gain a \plus4 bonus to the check for every size category by which it exceeds your own.
If you succeed, you can climb the creature as if it were a solid object with a Climb DR equal to its Maneuver defense.
The creature takes a \minus4 penalty to accuracy on physical attacks against you, but is not otherwise encumbered by your presence unless it cannot support your weight.
It can attempt to remove you by attacking you, or with an appropriate maneuver, such as grappling or shoving.

\feat{Shape Spell}{Metamagic}
\featpre 3rd level spell known.
\featben When casting a shaped spell, you can exclude any number of 5-foot cubes within the spell's area.
This allows you to prevent the spell from affecting your allies, while still allowing it to affect your enemies.
The area affected by the spell must be contiguous.

Only area spells can be shaped.
A shaped spell uses a spell slot two levels higher than the spell's original level.

\feat{Shield Discipline}{Class, Defense, Combat}
\featpre Armor discipline ability.
\featben You apply the benefits of your armor discipline ability while using shields, as well as while wearing armor.
If you reduce the encumbrance category of body armor you wear, you also reduce the encumbrance category of shields you use by the same amount.
Reducing a shield's encumbrance category causes tower shields to be treated as heavy shields (and no longer impose a \minus2 penalty to accuracy), heavy shields to be treated as light shields, and both light shields and bucklers to be treated as being unarmored (allowing you to freely use other items in that hand).
As normal, if you use your shield hand for something other than defending yourself with your shield, you do not gain your shield's bonus to physical defenses that round.

\feat{Shield Expertise}{Defense, Combat}
\featpre Shield Proficiency.
\featben You gain a \plus1 bonus to physical defenses while using a shield.

\feat{Shield Proficiency}{Combat}
\featben You can use a shield and take only the standard penalties.
\parhead{Normal}
When you are using a shield with which you are not proficient, you apply the shield's encumbrance penalty to accuracy with physical attacks.
\parhead{Special}
Barbarians, clerics, druids, fighters, paladins, rangers, and spellwarped autogmatically have Shield Proficiency as a bonus feat.
They need not select it.

\feat{Shielded Parry}{Combat, Defense, Reaction}
\featpres Dexterity 3, shield proficiency, Parry.
\featben When readying yourself to parry incoming blows and using a shield, you may add your shield's defense bonus to your attack roll made to parry.

\feat{Staggering Strike}{Combat, Strike}
\featpre Three strikes.
\featben In place of two strikes, you can make a physical attack.
If you hit, you deal damage normally.
If your attack result also beats the foe's Fortitude defense, it is \staggered for 5 rounds.

\feat{Silent Spell}{Metamagic}
\featpre 2nd level spell known.
\featben A silent spell can be cast with no verbal components.
Spells without verbal components are not affected.
A silent spell uses up a spell slot one level higher than the spell's actual level.

\feat{Skill Focus}{Skill}
Choose a skill.
\featpre 5 ranks in the chosen skill.
\featben Whenever you make a check with the chosen skill, you may roll twice and take the higher result.
\parhead{Special}
You can gain this feat multiple times.
Its effects do not stack.
Each time you take the feat, it applies to a new skill.

\feat{Skill Specialization}{Skill}
Choose a skill.
\featpre 10 ranks in the chosen skill, Skill Focus in the chosen skill.
\featben You gain a \plus5 bonus to all checks involving that skill.
\parhead{Special}
You can gain this feat multiple times.
Its effects do not stack.
Each time you take the feat, it applies to a new skill.

\feat{Somatic Strike}{Magic}
\featpre Combat prowess 8, 4th level spell known.
\featben When you cast spells, you can make a single strike with a melee weapon in place of the somatic components for the spell.
The spell is otherwise cast as normal, regardless of whether the strike hits or misses.
This cannot be used in combination with other abilities that let you attack as part of casting a spell, such as Imbued Spellstrike.

\feat{Spell Focus}{Magic}
Choose a school of magic.
\featpre 2nd level spells.
\featben Whenever you cast a spell from the chosen school of magic, you can reduce the cost to apply one augment or metamagic feat to the spell by one spell level.
\parhead{Special}
You can gain this feat multiple times.
Its effects do not stack.
Each time you take the feat, it applies to a new school of magic.

\feat{Spell Specialization}{Magic}
Choose a school of magic.
\featpres 4th level spells, Spell Focus in the chosen school.
\featben You gain a \plus2 bonus to your spellpower when casting spells and using spell-like abilities from the chosen school of magic.
In exchange, you take a \minus2 penalty to spellpower with all other spells and spell-like abilities.

\feat{Spellbreaker Rage}{Class, Rage}
\featpre Rage ability, barbarian level 10.
\featben When you enter a rage, you can use this feat. If you do, during the rage you gain \glossterm{spell resistance} equal to 10 \add your Constitution or barbarian level.
\parhead{Special} You can only gain the benefit of one Rage feat at a time.

\feat{Spellwarped Ritualist}{Magic}
\featben You can learn and perform rituals as if you were an arcane caster with a spellpower equal to your spellwarped spellpower.
The maximum level of ritual that you can learn or perform is equal to half your spellwarped level or half your spellwarped key attribute, whichever is lower.

In addition, you gain a special ability based on which type of spellwarped you are.

\subcf{Alteration} \x
\subcf{Pyromancy}
You can use any combustable item as a ritual component.
It can replace an amount of normal ritual components equal to the value of the item.
It cannot replace special ritual components.
\subcf{Telekinesis}
You can perform rituals from up to 30 feet away from the ritual components.
\subcf{Temporal}
You perform rituals twice as quickly.

\feat{Spellwoven Performance}{Performance, Skill}
\featpres Perform 9 ranks, any other Performance feat.
\featben You can cast Enchantment and Illusion spells while sustaining a performance.
The Spellcraft DR to identify those spells as they are cast increases by 10, as the performance disguises the magic.

\feat{Spring Attack}{Combat}
\featpres Dex 5, Mobility, combat prowess 8.
\featben As a standard action, you can move up to your speed and make a standard attack.
You must move at least 5 feet after each strike you make before you can make another strike.
You can only use this style while \unencumbered.

\feat{Still Spell}{Metamagic}
\featpre 2nd level spell known.
\featben A stilled spell can be cast with no somatic components.
Spells without somatic components are not affected.
A stilled spell uses up a spell slot one level higher than the spell's actual level.

\feat{Stonecunning}{Racial}
\featpre Dwarf.
\featben You gain a \plus2 bonus to Craft and Awareness checks related to stone or metal.
In addition, if you come within 10 feet of unusual stonework, you can make a Awareness check to notice it as if you were actively searching.
Finally, you can also intuit depth, sensing your approximate depth underground as naturally as a human can sense which way is up.

\feat{Stubborn}{General}
\featpre Willpower 5.
\featben You are immune to hostile \glossterm{Compulsion} effects.

\feat{Suggestive Performance}{Performance, Skill}
\featpres Perform 13 ranks, Fascinating Performance.
\featben As a standard action, you can make a DR 24 Perform check to suggest an course of action to a creature within \rngmed range of you.
For every 5 points by which your Perform check succeeds, you can suggest the same course of action an additional target.
Each target must be already \fascinated by you using the Fascinating Performance feat.
If your check result beats a target's Mental defense, it thinks your suggestion is a good idea and will try to follow it as long as you sustain the performance.
Otherwise, it is unaffected, though it remains fascinated.
This is a \glossterm{Delusion}, \glossterm{Mind}, \glossterm{Speech} effect.

Using this ability is considered continuing your performance for the Fascinating Performance feat, and does not free creatures you have fascinated from that effect.

\parhead{Special} You can use any combination of performance feats you possess a number of times per day equal to twice the number of performance feats you have.
If you fail the Perform check, the use of the ability is wasted.

\feat{Summoning Mastery}{Magic}
\featpre \glossterm{Summoning} spell known, 4nd level spells.
\featben As you cast a Summoning spell, you can attune to it, allowing you to maintain focus on it indefinitely without expending effort.
This allows you to control the summoned creatures without spending an action, and can extend the duration indefinitely if the spell has \durshort duration.
However, you can only exercise this control as long as all summoned creatures remain within \rnglong range of you, ignoring intervening obstacles.
If this distance is exceeded, or if you become unconscious, you lose your attunement to the spell.

You can only attune to one spell in this way at once.
If you attune to another Summoning spell, the attunement to the previous spell is broken.

\feat{Sustained Spell}{Metamagic}
\featpre 2nd level spell known.
\featben You can maintain focus on a sustained spell as a swift action instead of as a standard action.
If you cast any other spell, you lose the ability to sustain the spell.
A sustained spell uses up a spell slot one level higher than the spell's actual level.

\feat{Swift}{General, Mobility}
\featben You increase your land speed by 5 feet.

\feat{Sympathetic Resistance}{Magic}
\featpre 2nd level or higher Abjuration spell.
\featben Whenever you cast a spell of at least 1st level that deals damage with a non-physical type, you gain damage reduction equal to your spellpower against damage of that type for 1 round.
If a spell deals multiple types of damage, you gain damage reduction that applies against all of those types.

\feat{Tactical Analysis}{Combat}
\featpre Intelligence 3.
\featben You can attempt to identify the strengths and weaknesses of creatures based on your combat experience.
As a swift action, you can make a special check with a bonus equal to your combat prowess or your Intelligence (whichever is higher) \add 2 per round you have seen the creature fight.
The DR is equal to 10 \add the creature's level.
If you succeed, you learn about the monster's combat abilities as if you had made a successful Knowledge check.

\feat{Tactical Prediction}{Combat}
\featpres Intelligence 5, combat prowess 4.
\featben You can attempt to predict what your opponent will do.
As a swift action, you can make a special check with a bonus equal to your combat prowess or your Intelligence (whichever is higher) \add 2 per round you have seen the creature fight.
The DR is equal to 15 \add the creature's level.
If you succeed, you learn in general terms what the creature is planning to do during the next phase.
Of course, it can change its plans, particularly if it hears you tell your allies what it will do.

\feat{Taunting Rage}{Class, Rage}
\featpre Rage ability, barbarian level 10.
\featben When you enter a rage, you can use this feat. If you do, during the rage you can taunt your foes.
Whenever you hit a creature with a melee strike, it is \taunted by you for 2 rounds.
\parhead{Special} You can only gain the benefit of one Rage feat at a time.

\feat{Terrain Lore}{Class}
\featpre Terrain lore ability, ranger 7.
\featben You gain an additional terrain lore ability that you qualify for.

\feat{Terrifying Performance}{Performance, Skill}
\featpre Perform 15 ranks.
\featben As a standard action, you can make a DR 20 Perform check to terrify a creature within \rngmed range of you.
For every 5 points by which your Perform check succeeds, you can affect an additional target.
If your check result beats a target's Mental defense, it is \frightened by you as long as you sustain the performance.
Otherwise, it is \shaken by you.
This is a \glossterm{Fear}, \glossterm{Mind} effect.

\feat{Threatening Fire}{Combat, Reaction, Style}
\featpres Dexterity 5, Overwhelming Fire, combat prowess 8.
\featben While in this style, you threaten creatures within your natural reach as if you were wielding a melee weapon, allowing you to contribute to overwhelm penalties and block the movement of other creatures.
(A Medium creature would threaten adjacent creatures).
In addition, you do not take a \minus4 penalty for using ranged weapons against adjacent creatures.
However, you still cannot used your ranged weapon to defend yourself, which can cause you to be \defenseless.
\stylereq Wield a ranged weapon.
You must make a physical ranged attack each round.

\feat{Trapfinder}{Skill}
\featpre Awareness 5 ranks.
\featben As a full-round action, you can move up to 10 feet while searching every square within 10 feet of you for traps with the Awareness skill (see \pcref{Awareness}).
If you detect a trap partway through your movement, you may immediately stop moving.

\feat{Toughness}{General}
\featpre Constitution 3 or Strength 3.
\featben You gain a \plus2 bonus to your Fortitude defense.
This bonus can increase your hit points (see \pcref{Hit Points}).
\parhead{Special} This feat can be taken multiple times. Its effects stack.
You may take this feat a maximum number of times equal to half your Strength or half your Constitution, whichever is higher.

\feat{Track}{Skill}
\featpre Survival 5 ranks.
\featben You can use the Survival skill to find or follow tracks.
Following tracks requires a successful Survival check.
The DR of the check depends on how easy the tracks are to notice, as shown on \trefnp{Track DRs} and \trefnp{Track Modifiers}.
You must make another Survival check after following the trail for 1 mile, or if it becomes especially difficult to follow.
While tracking, you move at half your normal speed (or at your normal speed with a \minus5 penalty on the check, or at up to twice your normal speed with a \minus20 penalty on the check).
The DR depends on the surface and the prevailing conditions, as given on the table below:

\begin{dtable}
    \lcaption{Track DRs}
    \begin{dtabularx}{\columnwidth}{l >{\lcol}X l}
        \tb{Surface} & \tb{Description} & \tb{Survival DR} \\
        \hline
        Very soft ground & Any surface (fresh snow, thick dust, wet mud) that holds deep, clear impressions of footprints. & 5 \\
        Soft ground & Any surface soft enough to yield to pressure, but firmer than wet mud or fresh snow, in which a creature leaves frequent but shallow footprints & 10 \\
        Firm ground & Most normal outdoor surfaces (such as lawns, fields, woods, and the like) or exceptionally soft or dirty indoor surfaces (thick rugs and very dirty or dusty floors) & 15 \\
        Hard ground & Any surface that doesn't hold footprints at all, such as bare rock or a streambed & 20
    \end{dtabularx}
\end{dtable}

\begin{dtable}
    \lcaption{Track Modifiers}
    \begin{dtabularx}{\columnwidth}{>{\lcol}X l}
        \tb{Condition} & \tb{DR Modifier} \\
        \hline
        Every three creatures in the group being tracked  & \minus1 \\
        Size of creature or creatures being tracked:\fn{1} &  \\
        Fine  & \plus16 \\
        Diminutive  & \plus12 \\
        Tiny  & \plus8 \\
        Small  & \plus4 \\
        Medium  & \plus0 \\
        Large  & \minus4 \\
        Huge  & \minus8 \\
        Gargantuan  & \minus12 \\
        Colossal  & \minus16 \\
        Every 24 hours since the trail was made  & \plus1 \\
        Every hour of rain since the trail was made  & \plus1 \\
        Fresh snow cover since the trail was made  & \plus10 \\
        \tb{Poor visibility:\fn{2}} &  \\
        Overcast or moonless night  & \plus6 \\
        Moonlight  & \plus3 \\
        Fog or precipitation  & \plus3 \\
        Tracked party hides trail (and moves at half speed)  & \plus5
    \end{dtabularx}
    1 For a group of mixed sizes, apply only the modifier for the largest size category. \\
    2 Apply only the largest modifier from this category. \\
\end{dtable}

If you fail a Survival check to track, you can retry after 1 hour (outdoors) or 10 minutes (indoors) of searching.
\parhead{Normal}
Without this feat, you can use the Survival skill to find tracks, but you can follow them only if the DR for the task is 10 or lower.
Alternatively, you can use the Awareness skill to find a footprint or similar sign of a creature's passage using the DRs given above, but you can't use Awareness to follow tracks, even if someone else has already found them.

\feat{Trap Sense}{Skill}
\featpre Awareness 10 ranks.
\featben Whenever you come within 10 feet of a trap, you receive an immediate Awareness check to notice the trap.
This check should be made secretly, so you do not know whether you failed to notice a trap.

\feat{Two-Weapon Defense}{Combat}
\featpres Dex 3, Two-Weapon Fighting.
\featben When wielding a weapon in both hands, you gain a \plus1 bonus to physical defenses.

\feat{Two-Weapon Fighting}{Combat}
You can fight with a weapon in each hand more effectively.
\featpre Dex 3.
\featben When wielding a weapon in both hands, you gain a \plus2 bonus to accuracy when dual attacking with those weapons (see \pcref{Dual Attacking}).

\feat{Two-Weapon Rend}{Combat, Power}
\featpres Str 5, Mighty Blows, Two-Weapon Fighting, combat prowess 8.
\featben When fighting with two weapons at once, you gain a \plus1 bonus to physical damage.
This bonus increases to \plus2 at Strength 10, and to \plus3 at Strength 20.

\feat{Unarmed Dodge}{Combat, Defense}
\featpre Combat prowess 8, Unarmed Proficiency, Dexterity 5.
\featben You gain a \plus1 bonus to physical defenses (Armor, Maneuver, Reflex) while both your hands are free.

\feat{Unarmed Might}{Combat, Power}
\featpres Combat prowess 4, Unarmed Proficiency, Strength 5.
\featben Your unarmed attacks deal damage as if you were two size categories larger than normal (1d6 for a Medium creature, or 1d4 for a small creature).

\feat{Unarmed Proficiency}{Combat}
\featben You gain proficiency with your unarmed attack.
This grants you a \plus4 bonus to accuracy with the weapon and allows you to defend yourself with it, just as if you were using another melee weapon you are proficient with.

In addition, your unarmed attacks can deal lethal or nonlethal damage as you choose.
\parhead{Normal} You are not proficient with your unarmed attack, and it deals nonlethal damage. See \pcref{Unarmed Combat}.

\feat{Unbending Resolve}{Class}
\featpre Paladin level 5.
\featben You are immune to \glossterm{Compulsion} and \glossterm{Delusion} effects.

\feat{Unfettered Knowledge}{Magic}
\featpre 2nd level or higher \glossterm{Knowledge} spell known.
\featben You gain a \plus2 bonus to all Knowledge checks.
If you know a Knowledge spell of 6th level or higher, this bonus increases to \plus4.

\feat{Unliving Conduit}{Magic, Style}
\featpre 4th level or higher \glossterm{Vivimancy} spell known.
\featben As long as you are within \rngclose range of an undead creature (other than yourself), you gain a \plus2 bonus to spellpower with Vivimancy spells.
If the undead creature is under your control, this bonus applies to all spells you cast.
This bonus increases by \plus1 when you gain access to 6th and 8th level spells.

\feat{Untrackable}{Skill}
\featpre Survival 15 ranks.
\featben You can choose to leave no trace of your passage as you move.
If you do, tracking you is impossible by any means.
This does not prevent effects that reveal your current location, such as the \spell{discern location} spell, from functioning.

\feat{Vermin Speech}{Class}
\featpre Wild speech ability, druid level 3.
\featben You can use your wild speech ability to communicate with vermin, in addition to animals.

\feat{Versatility}{Class}
\featpre Levels in two different classes.
\featben For the purpose of abilities based on your class level, you treat your class level as being equal to your total level.
This can increase the power of abilities you already have, but it does not affect the rate at which you gain new abilities.

\feat{Versatile Crafter}{Item Creation}
\featpre Craft (any) 10 ranks.
\featben You learn how to make items from any school of magic, including universal spells.
See the Craft skill, \pref{Craft}, for details.
\parhead{Normal}
You learn how to make items from one school for every five ranks you have in each Craft skill.

\feat{Veteran's Experience}{Skill}
\featpre Combat prowess 8.
\featben You can use your battlefield experience in place of learned knowledge to identify monsters.
When attempting to identify a monster, you may use your combat prowess in place of your skill modifier with the relevant Knowledge skill.

\feat{Vital Flow}{Magic}
\featpre 4th level or higher \glossterm{Positive} spell known.
\featben You gain \glossterm{fast healing} equal to the level of your highest level Positive spell known.
If you know a Positive spell of 8th level or higher, this fast healing increases to be equal to your spellpower with Positive spells.

\feat{Wallpuncher}{Skill}
\featpres Strength 5, Climb 10.
\featben You can punch holes in walls as you climb.
If you do, you create handholds, making the wall easier to climb (see \pcref{Climb}).
This does not slow your climbing, but you must have both hands free to climb in this way.
This ability works on any wall with \glossterm{hardness} no greater than your Strength.

\feat{Wall Slam}{Combat, Maneuver, Power, Strike}
\featpres Strength 5, combat prowess 8.
\featben As a standard action, you can attempt to shove an opponent into a solid obstacle.
If you succeed, the creature takes 1d6 bludgeoning damage per Strength.

\feat{Wardwalker}{Magic}
\featpre 3rd level or higher \glossterm{Warding} spell known.
\featben When you trigger a Warding spell, it does not take effect until one round after you triggered it. This can give you time to escape the spell's effects.

\feat{Wary Rage}{Class, Rage}
\featpre Rage ability, barbarian level 2.
\featben When you enter a rage, you can use this feat. If you do, during the rage you only suffer a \minus1 penalty to physical defenses for raging (instead of the normal \minus2).
\parhead{Special} You can only gain the benefit of one Rage feat at a time.

\feat{Weapon Focus}{Combat}
Choose one weapon group.
\featpre Proficiency with selected weapon group.
\featben You gain a special ability based on the weapon group chosen.
These special abilities only function when you are making attacks with weapons from your chosen weapon group.
\begin{itemize}
    \item Armor weapons: When you perform a shield bash, you still benefit from the shield's defense bonus.
        In addition, armor spikes no longer impose a penalty to your physical defenses.
    \item Axes: You gain a \plus2 bonus to damage against creatures not wearing medium or heavy armor.
    \item Blades, heavy: You gain a \plus2 bonus to damage against creatures not wearing medium or heavy armor.
    \item Blades, light: If you draw a light blade and attack with it in the same round, you gain a \plus2 bonus to damage.
        The creature you attack is unaffected by this ability for the rest of the encounter.
    \item Blunt weapons: When you deal damage to a creature, it takes a \minus2 penalty to Mental defense for 1 round.
        This penalty is not cumulative with itself.
    \item Bows: You can arc your shots, allowing you to ignore cover and treat total cover as total concealment (50\% failure chance) if you can shoot above the obstacle to reach your target.
    \item Crossbows: The time required for you to reload crossbows is reduced to a free action (for a hand or light crossbow) or a move action (for a heavy crossbow).
    \item Flexible weapons: You gain a \plus2 bonus to damage against creatures using a shield.
    \item Headed weapons: You gain a \plus2 bonus to damage against creatures wearing medium or heavy armor.
    \item Monk weapons: You gain a \plus2 bonus to damage against creatures using manufactured weapons (but not unarmed strikes or natural weapons).
    \item Polearms: You can switch grips to short haft or stop short hafting a polearm as a swift action, and you take no penalty while short hafting it.
    \item Spears: If a creature charges you, you gain a \plus5 bonus to damage against it until the end of the round.
    \item Thrown weapons: You can defend yourself with a thrown weapon as you throw it, preventing you from being \defenseless (see \pcref{Thrown Weapons in Melee}).
\end{itemize}
\parhead{Special}
You cannot choose simple weapons when you take this feat.

\feat{Weapon Finesse}{Combat, Finesse}
\featpres Dexterity 3
\featben You gain a \plus1 bonus to physical damage with light weapons.
\parhead{Special} This feat can be taken multiple times. Its effects stack.
You may take this feat a maximum number of times equal to half your Dexterity.

\feat{Weapon Proficiency}{Combat}
Choose one weapon group.
\featben You are proficient with weapons from your chosen weapon group.
This allows you to defend yourself normally with melee weapons, and to attack with exotic ranged weapons without penalty.
\parhead{Normal}
You cannot defend yourself with melee weapons you are not proficient with, and you take a \minus4 penalty to accuracy with ranged weapons you are not proficient with.
\parhead{Special}
You can gain Weapon Proficiency multiple times.
Each time you take the feat, it applies to a new weapon group.
You cannot choose simple weapons.

\feat{Wild Substitution}{Class, Magic}
\featpre Wild magic ability, sorcerer level 7.
\featben When you fail a wild magic roll, you may treat the spell as if it were one level lower (to a minimum of 1st level) for the purpose of determining which spell level you lose the ability to cast.
If you are already unable to cast that level of spells, you must treat the spell as if it were its normal level.

\feat{Wild Suppression}{Class, Magic}
\featpre Wild magic ability, sorcerer level 3.
\featben When you fail a wild magic roll, you may choose to suppress the magical energy released.
If you do, neither the spell nor its miscast effect occurs.
In addition, the time required to regain the ability to cast spells of the same level as the suppressed spell is halved.

\feat{Willful Rage}{Class, Rage}
\featpre Rage ability, barbarian level 2.
\featben When you enter a rage, you can use this feat. If you do, during the rage you add your rage bonus to your Mental defense.
\parhead{Special} You can only gain the benefit of one Rage feat at a time.

\feat{Wired}{Magic}
\featpre 1st level or higher Electricity spell known.
\featben You increase your speed with all forms of movement by 5 feet.
If you know an Electricity spell of 5th level or higher, this bonus increases to \plus10 feet.

\feat{Whirlwind Strike}{Combat, Strike}
\featpres Dex 5, combat prowess 8.
\featben As a full-round action, you can make a single melee strike against each opponent you threaten.
If your Dexterity is at least 10, you can use this feat as a standard action.

\section{Other Feat Rules}

\subsection{Bonus Feats}
Some abilities grant a character bonus feats.
Unless otherwise specified, the character must still meet any prerequisites for the feat.
If the character does not meet the prerequisites at the time the bonus feat is granted, the character does not gain the feat.
If the character later meets the prerequisites, the character immediately gains the benefit of the bonus feat.

If a character gains a feat as a bonus feat that he or she has already acquired through other means, the character may select instead any other feat for which she qualifies.

\subsection{Retraining Feats}
At every even level, your character can choose to retrain an old feat in exchange for a new feat.
You can only retrain feats for other feats you could have acquired at the time you took the original feat.
Thus, you cannot retrain feats gained through class abilities which give you a specific feat, since there were no other feats you could have taken.
For example, a 6th level fighter can retrain his 2nd level fighter bonus feat for any other combat feat that he qualified for at his 2nd fighter level.
