\ExplSyntaxOn

\newcommand{\tempderp}{''}

\def\replace#1#2#3{%
 \def\tmp##1#2{##1#3\tmp}%
   \tmp#1\stopreplace#2\stopreplace}
\def\stopreplace#1\stopreplace{}

% args:
%   '\spellname', 'Spell Name', pre-spell header,
%   spell short description, spell level
\DeclareDocumentCommand{\createspell}{m m o m o}
{
    \DeclareDocumentCommand{#1}{s o}
    {
        \IfBooleanTF{##1}
        {
            \spellhead[##2]*{#2}[#3]~#4
        }
        {
            \spellhead[##2]{#2}[#3]~#4
        }
    }
}

\newcommand{\makemacro}[2]
{
    \lowercase{
        \expandafter
        \newcommand\csname#1\endcsname
    }{#1:~#2}
}

\ExplSyntaxOff

% cantrips
\createspell{\SLacidorb}{Acid Orb}
    {Fling acid to deal damage}
\createspell{\SLaugmentattack}{Augment Attack}
    {Grant ally damage bonus on next attack}
\createspell{\SLcombattelekinesis}{Combat Telekinesis}
    {Wield weapon telekinetically.}
\createspell{\SLconfusionlesser}{Confusion}[Lesser]
    {Compel foe to briefly act randomly.}
\createspell{\SLconjureprojectile}{Conjure Projectile}
    {Fire projectile at foe to deal damage.}
\createspell{\SLdisplacementlesser}{Displacement}[Lesser]
    {Grant ally 20\% miss chance against physical attacks}
\createspell{\SLdrainingtouch}{Draining Touch}
    {Steal vital energy to deal damage and heal.}
\createspell{\SLexhaustion}{Exhaustion}
    {Fatigue or exhaust foe.}
\createspell{\SLfearlesser}{Fear}[Lesser]
    {Frighten foe briefly.}
\createspell{\SLflare}{Flare}
    {Create flash of light to impair foe's vision.}
\createspell{\SLmagicray}{Magic Ray}
    {Fire force ray to deal damage.}
\createspell{\SLresistdamage}{Resist Damage}
    {Grant ally brief damage reduction.}
\createspell{\SLslowlesser}{Slow}[Lesser]
    {Force foe to skip one movement phase.}
\createspell{\SLtwistfate}{Twist Fate}
    {Learn foe's future action, possibly apply penalty.}

% regular spells
\createspell{\SLablateimpact}{Ablate Impact}
    {Immediately reduce damage from physical attacks.}
\createspell{\SLablativespellshield}{Ablative Spellshield}
    {Immediately reduce damage from spells.}
\createspell{\SLacidarrowgreater}{Acid Arrow}[Greater]
    {Fling acid to deal lingering damage and stagger.}
\createspell{\SLacidarrow}{Acid Arrow}
    {Fling acid to deal lingering damage}
\createspell{\SLacidfog}{Acid Fog}
    {Create zone of acidic fog to deal damage.}
\createspell{\SLagony}{Agony}
    {Inflict pain to increase damage taken.}
\createspell{\SLaid}{Aid}
    {Grant temporary hit points and immunity to fear.}
\createspell{\SLairwalk}{Air Walk}
    {Grant ability to walk on air.}
\createspell{\SLantilifeshell}{Antilife Shell}
    {Prevent living creatures from entering zone.}
\createspell{\SLantimagicfield}{Antimagic Field}
    {Negate all magic in emanation.}
\createspell{\SLaqueousblade}{Aqueous Blade}
    {Transform weapon to attack Reflex defense.}
\createspell{\SLassimilate}{Assimilate}
    {Absorb creature into your body.}
\createspell{\SLaversion}{Aversion}
    {Compel creature to avoid something.}
\createspell{\SLbane}{Bane}
    {Impair foe.}
\createspell{\SLbanemass}{Bane}[Mass]
    {Impair multiple foes.}
\createspell{\SLbarkskin}{Barkskin}
    {Grant damage reduction.}
\createspell{\SLbestowcurse}{Bestow Curse}
    {\minus6 to an attribute score; \minus4 on attack rolls, saves, and checks; or 25\% chance of losing each action.}
\createspell{\SLblacktentacles}{Black Tentacles}
    {Grapple foes with tentacles from the ground.}
\createspell{\SLbladebarrier}{Blade Barrier}
    {Create wall of whirling blades to deal damage.}
\createspell{\SLblasphemy}{Blasphemy}
    {Damage and stagger nearby nonevil creatures.}
\createspell{\SLbless}{Bless}
    {Grant ally a legend point.}
    \createspell{\SLblessmass}{Bless}[Mass]
    {Grant multiple creatures a legend point.}
\createspell{\SLblink}{Blink}
    {Gain 50\% miss chance by disappearing and reappearing.}
\createspell{\SLblur}{Blur}
    {Grant Stealth bonus by blurring outline.}
\createspell{\SLburninghands}{Burning Hands}
    {Create damaging cone of fire.}
\createspell{\SLcacaphonicword}{Cacaphonic Word}
    {Make incoherent noise to damage and disorient non-chaotic creatures.}
\createspell{\SLcalllightning}{Call Lightning}
    {Call multiple lightning bolts from sky to deal damage.}
\createspell{\SLcalllightninggreater}{Call Lightning}[Greater]
    {Call multiple lightning bolts from sky to deal damage and stagger.}
\createspell{\SLcalmemotions}{Calm Emotions}
    {Calm multiple creatures to avoid violence.}
\createspell{\SLchainlightning}{Chain Lightning}
    {Create lightning that jumps between foes to deal damage.}
\createspell{\SLchaoshammer}{Chaos Hammer}
    {Damage and disorient nonchaotic creature.}
\createspell{\SLcharmmonster}{Charm Monster}
    {Delude creature into believing you are its ally.}
\createspell{\SLcharmmonstermass}{Charm Monster}[Mass]
    {Delude multiple creatures into believing you are their ally.}
\createspell{\SLcharmperson}{Charm Person}
    {Delude person into believing you are its ally.}
\createspell{\SLcharmpersonmass}{Charm Person}[Mass]
    {Delude multiple people into believing you are their ally.}
\createspell{\SLcloakofchaos}{Cloak of Chaos}
    {Grant multiple creatures spell resistance and retributive damage against lawful foes.}
\createspell{\SLcolorspray}{Color Spray}
    {Create cone of random colors to impair vision.}
\createspell{\SLcommand}{Command}
    {Speak command that creature must obey.}
\createspell{\SLcommandmass}{Command}[Mass]
    {Speak command that multiple creatures must obey.}
\createspell{\SLconeofcoldgreater}{Cone of Cold}[Greater]
    {Create large frigid cone to deal damage and inhibit movement.}
\createspell{\SLconeofcold}{Cone of Cold}
    {Create frigid cone to deal damage and inhibit movement.}
\createspell{\SLconfusionmass}{Confusion}[Mass]
    {Compel multiple foes to act randomly.}
\createspell{\SLconfusion}{Confusion}
    {Compel foe to act randomly.}
\createspell{\SLcreatesound}{Create Sound}
    {Create sounds from nowhere.}
\createspell{\SLcripple}{Cripple}
    {Cripple limbs to deal damage and stagger or paralyze.}
\createspell{\SLcurecriticalwoundsmass}{Cure Critical Wounds}[Mass]
    {Cure multiple allies with critical injuries.}
\createspell{\SLcurecriticalwounds}{Cure Critical Wounds}
    {Cure ally with critical injuries.}
\createspell{\SLcurelightwoundsmass}{Cure Light Wounds}[Mass]
    {Cure multiple allies with minor injuries.}
\createspell{\SLcurelightwounds}{Cure Light Wounds}
    {Cure ally with minor injuries.}
\createspell{\SLcuremoderatewoundsmass}{Cure Moderate Wounds}[Mass]
    {Cure multiple allies with moderate injuries.}
\createspell{\SLcuremoderatewounds}{Cure Moderate Wounds}
    {Cure ally with moderate injuries.}
\createspell{\SLcureseriouswoundsmass}{Cure Serious Wounds}[Mass]
    {Cure multiple allies with serious injuries.}
\createspell{\SLcureseriouswounds}{Cure Serious Wounds}
    {Cure ally with serious injuries.}
\createspell{\SLcurseofbloodandbone}{Curse of Blood and Bone}
    {Inflict curse to deal damage and reduce total hit points.}
\createspell{\SLcurseofenfeeblement}{Curse of Enfeeblement}
    {Inflict curse to penalize Strength or Dexterity.}
\createspell{\SLcurseofthewaywardmind}{Curse of the Wayward Mind}
    {Inflict curse to confuse or disorient foe.}
\createspell{\SLdancinglights}{Dancing Lights}
    {Create floating lights you control.}
\createspell{\SLdarkvision}{Darkvision}
    {Grant ability to see in total darkness.}
\createspell{\SLdeathknell}{Death Knell}
    {Inflict damage, accelerate death, and gain life if foe dies.}
\createspell{\SLdeathward}{Death Ward}
    {Grant ally immunity to negative energy and death effects.}
\createspell{\SLdeathwardmass}{Death Ward}[Mass]
    {Grant multiple allies immunity to negative energy and death effects.}
\createspell{\SLdeepslumber}{Deep Slumber}
    {Compel foe to close eyes and possibly sleep forever.}
\createspell{\SLdeflection}{Deflection}
    {Gain regenerating legend point for physical defenses.}
\createspell{\SLdelaydamage}{Delay Damage}
    {Delay half the damage you take until later.}
\createspell{\SLdelaypoison}{Delay Poison}
    {Quickly grant ally immunity to poison.}
\createspell{\SLdelayedblastfireball}{Delayed Blast Fireball}
    {Create fiery burst to deal damage; delaying blast increases power.}
\createspell{\SLdestruction}{Destruction}
    {Invoke divine power to deal damage and possibly kill.}
\createspell{\SLdetectalignment}{Detect Alignment}
    {Reveal entities with selected alignment within cone.}
\createspell{\SLdictum}{Dictum}
    {Damage and immobilize nearby nonlawful creatures.}
\createspell{\SLdimensiondoor}{Dimension Door}
    {Teleport anywhere within 1,000 feet.}
\createspell{\SLdimensiondoormass}{Dimension Door}[Mass]
    {Teleport anywhere within 1,000 feet with multiple allies.}
\createspell{\SLdimensionslide}{Dimension Slide}
    {Teleport creature short distance.}
\createspell{\SLdimensionalanchor}{Dimensional Anchor}
    {Block extradimensional movement.}
\createspell{\SLdiscernlies}{Discern Lies}
    {Reveal deliberate lies within cone.}
\createspell{\SLdiscernvulnerability}{Discern Vulnerability}
    {Quickly find weaknesses in foe's defenses.}
\createspell{\SLdiscordantsong}{Discordant Song}
    {Create music to compel creatures to act randomly.}
\createspell{\SLdisintegrate}{Disintegrate}
    {Deal damage, possibly transforming target into dust.}
\createspell{\SLdispelmagic}{Dispel Magic}
    {Negate magical effects on target.}
\createspell{\SLdispelmagicgreater}{Dispel Magic}[Greater]
    {Negate magical effects within area.}
\createspell{\SLdisjoinmagic}{Disjoin Magic}
    {Unconditionally negate magical effects on multiple targets.}
\createspell{\SLdisplacement}{Displacement}
    {Grant ally 50\% miss chance against attacks.}
\createspell{\SLdivinefavorgreater}{Divine Favor}[Greater]
    {Gain regenerating legend point.}
\createspell{\SLdivinefavor}{Divine Favor}
    {Gain legend point.}
\createspell{\SLdominatemonster}{Dominate Monster}
    {Compel creature to obey you completely.}
\createspell{\SLdominateperson}{Dominate Person}
    {Compel person to obey you completely.}
\createspell{\SLdrainlife}{Drain Life}
    {Steal vital energy to deal damage and heal.}
\createspell{\SLearthspull}{Earth's Pull}
    {Increase force of gravity on foe.}
\createspell{\SLearthenblade}{Earthen Blade}
    {Create magical weapon from the ground.}
\createspell{\SLearthglide}{Earth Glide}
    {Grant ally ability to glide through earth.}
\createspell{\SLearthquake}{Earthquake}
    {Shake ground to deal damage, knock prone, and immobilize.}
\createspell{\SLearthspike}{Earthspike}
    {Animate earthen spike to deal damage.}
\createspell{\SLearthspikemass}{Earthspike}[Mass]
    {Animate multiple earthen spikes to deal damage.}
\createspell{\SLelementalswarm}{Elemental Swarm}
    {Summon multiple elementals.}
\createspell{\SLenergyconversion}{Energy Conversion}
    {Absorb energy, fire energy bolts to deal damage.}
\createspell{\SLenervation}{Enervation}
    {Inflict three negative levels.}
\createspell{\SLenervationgreater}{Enervation}[Greater]
    {Inflict up to eight negative levels.}
\createspell{\SLenfeeblement}{Enfeeblement}
    {Target takes \minus4 to physical attribute.}
\createspell{\SLenlargemonster}{Enlarge Monster}
    {Double size of creature.}
\createspell{\SLenlargemonstermass}{Enlarge Monster}[Mass]
    {Double size of multiple creatures.}
\createspell{\SLenlargepersonmass}{Enlarge Person}[Mass]
    {Double size of multiple humanoids.}
\createspell{\SLenlargeperson}{Enlarge Person}
    {Double size of humanoid.}
\createspell{\SLentangle}{Entangle}
    {Immobilize foe with nearby plants.}
\createspell{\SLentanglegreater}{Entangling Growth}
    {Immobilize foe with newly grown plants.}
\createspell{\SLentropicshield}{Entropic Shield}
    {Grant ally 50\% miss chance against ranged attacks.}
\createspell{\SLetherealjaunt}{Ethereal Jaunt}
    {Travel to Ethereal Plane.}
\createspell{\SLetherealness}{Etherealness}
    {Travel to Ethereal Plane with multiple allies.}
\createspell{\SLeyebite}{Eyebite}
    {Inflict foe's eyes to deal damage and impair sight.}
\createspell{\SLfaeriefire}{Faerie Fire}
    {Create lights to negate concealment and stealth.}
\createspell{\SLfalsereality}{False Reality}
    {Create scripted image in massive area.}
\createspell{\SLfear}{Fear}
    {Terrify foe.}
\createspell{\SLfearmass}{Fear}[Mass]
    {Terrify multiple foes.}
\createspell{\SLfeatherfall}{Feather Fall}
    {Arrest fall of object or ally.}
\createspell{\SLfeeblemind}{Feeblemind}
    {Reduce foe's Intelligence to \minus9.}
\createspell{\SLfingerofdeath}{Finger of Death}
    {Deal damage, possibly instantly kill foe.}
\createspell{\SLfireball}{Fireball}
    {Create burst of fire to deal damage.}
\createspell{\SLfireseeds}{Fire Seeds}
    {Infuse berries with fiery energy, detonate to deal damage.}
\createspell{\SLfireshield}{Fire Shield}
    {Gain cold resistance and retributive damage against attacks.}
\createspell{\SLfirestorm}{Fire Storm}
    {Create storm of fire to deal damage to foes.}
\createspell{\SLflameblade}{Flame Blade}
    {Wield fire as scimitar.}
\createspell{\SLflamestrike}{Flame Strike}
    {Smite foes with divine fire.}
\createspell{\SLfly}{Fly}
    {Grant ally ability to fly.}
\createspell{\SLfogcloud}{Fog Cloud}
    {Create zone of fog that obscures vision.}
\createspell{\SLfogsea}{Fog Sea}
    {Create massive zone of fog that obscures vision.}
\createspell{\SLfogshield}{Fog Shield}
    {Emanate fog that follows you, obscuring vision.}
\createspell{\SLforcecage}{Forcecage}
    {Imprison foe in prison of force.}
\createspell{\SLforesightgreater}{Foresight}[Greater]
    {Gain immunity to being caught by surprise and massive initiative bonus.}
\createspell{\SLforesight}{Foresight}
    {Gain immunity to being caught by surprise.}
\createspell{\SLforesight}{Foresight}
    {``Sixth sense'' warns of impending danger.}
\createspell{\SLforget}{Forget}
    {Delude creature into forgetting something.}
\createspell{\SLfreedom}{Freedom}
    {Grant ally immunity to movement impediments.}
\createspell{\SLfreedommass}{Freedom}[Mass]
    {Grant multiple allies immunity to movement impediments.}
\createspell{\SLgaseousform}{Gaseous Form}
    {Transform ally into gas, granting flight.}
\createspell{\SLgentledescent}{Gentle Descent}
    {Grant ally ability to glide.}
\createspell{\SLgentledescentmass}{Gentle Descent}[Mass]
    {Grant multiple allies ability to glide.}
\createspell{\SLghoultouch}{Ghoul Touch}
    {Sicken or nauseate creauture.}
\createspell{\SLgiantvermin}{Giant Vermin}
    {Turns centipedes, scorpions, or spiders into giant vermin.}
\createspell{\SLglitterdust}{Glitterdust}
    {Create dust to negate concealment and stealth.}
\createspell{\SLgrease}{Grease}
    {Create grease to make area or object slippery.}
\createspell{\SLgustofwind}{Gust of Wind}
    {Create wind to shove creatures away.}
\createspell{\SLharm}{Harm}
    {Inflict damage and Constitution damage.}
\createspell{\SLhaste}{Haste}
    {Grant ally double movement speed and extra attack.}
\createspell{\SLhastelesser}{Haste}[Lesser]
    {Grant ally double movement speed.}
\createspell{\SLhastemass}{Haste}[Mass]
    {Grant multiple allies double movement speed, extra attack.}
\createspell{\SLheal}{Heal}
    {Cure ally of critical injuries and almost all afflictions.}
\createspell{\SLheatmetal}{Heat Metal}
    {Heat metal object to deal damage, possibly daze.}
\createspell{\SLheroism}{Heroism}
    {Grant temporary hit points, immunity to fear, legend point.}
\createspell{\SLheroismgreater}{Heroism}[Greater]
    {Gives combat bonuses, temporary hit points.}
\createspell{\SLholdmonster}{Hold Monster}
    {Immobilize a creature.}
\createspell{\SLholdmonstermass}{Hold Monster}[Mass]
    {Immobilize multiple creatures.}
\createspell{\SLholdperson}{Hold Person}
    {Immobilize a person.}
\createspell{\SLholdpersonmass}{Hold Person}[Mass]
    {Immobilize multiple people.}
\createspell{\SLholyaura}{Holy Aura}
    {Grant multiple creatures spell resistance and retributive damage against evil foes.}
\createspell{\SLholysmite}{Holy Smite}
    {Damage and daze nongood creature.}
\createspell{\SLholyword}{Holy Word}
    {Kills, paralyzes, blinds, or deafens nongood subjects in large radius.}
\createspell{\SLholyword}{Holy Word}
    {Damage and daze nearby nongood creatures.}
\createspell{\SLhorridwilting}{Horrid Wilting}
    {Create dessicating burst to deal damage.}
\createspell{\SLhypnoticpattern}{Hypnotic Pattern}
    {Create lights to fascinate creatures.}
\createspell{\SLhypnoticpatterngreater}{Hypnotic Pattern}
    {Create many lights to fascinate creatures.}
\createspell{\SLicestorm}{Ice Storm}
    {Create storm to deal damage.}
\createspell{\SLimplosion}{Implosion}
    {Concentrate to damage and possibly kill multiple creatures.}
\createspell{\SLimprisonment}{Imprisonment}
    {Entomb foe beneath the earth permanently.}
\createspell{\SLinertialshield}{Inertial Shield}
    {Grant damage reduction.}
\createspell{\SLinflictcriticalwoundsmass}{Inflict Critical Wounds}[Mass]
    {Inflict multiple foes with critical injuries.}
\createspell{\SLinflictcriticalwounds}{Inflict Critical Wounds}
    {Inflict foe with critical injuries.}
\createspell{\SLinflictlightwoundsmass}{Inflict Light Wounds}[Mass]
    {Inflict multiple foes with minor injuries.}
\createspell{\SLinflictlightwounds}{Inflict Light Wounds}
    {Inflict foe with minor injuries.}
\createspell{\SLinflictmoderatewoundsmass}{Inflict Moderate Wounds}[Mass]
    {Inflict multiple foes with moderate injuries.}
\createspell{\SLinflictmoderatewounds}{Inflict Moderate Wounds}
    {Inflict foe with moderate injuries.}
\createspell{\SLinflictseriouswoundsmass}{Inflict Serious Wounds}[Mass]
    {Inflict multiple foes with serious injuries.}
\createspell{\SLinflictseriouswounds}{Inflict Serious Wounds}
    {Inflict foe with serious injuries.}
\createspell{\SLinvisibility}{Invisibility}
    {Turn ally invisible until it attacks.}
\createspell{\SLinvisibilitymass}{Invisibility}[Mass]
    {Turn multiple allies invisible.}
\createspell{\SLinvisibilitygreater}{Invisibility}[Greater]
    {Turn ally invisible.}
\createspell{\SLironbody}{Iron Body}
    {Transform your body into living iron.}
\createspell{\SLirresistibledance}{Irresistible Dance}
    {Force foe to dance.}
\createspell{\SLknock}{Knock}
    {Open locked or barred object.}
\createspell{\SLlevitate}{Levitate}
    {Telekinetically move ally vertically.}
\createspell{\SLlifeseekingmissile}{Lifeseeking Missile}
    {Fire homing missiles to deal damage.}
\createspell{\SLlightningbolt}{Lightning Bolt}
    {Create line of lightning to deal damage.}
\createspell{\SLlongstrider}{Longstrider}
    {Double your land speed.}
\createspell{\SLmagearmor}{Mage Armor}
    {Create force armor to increase defenses.}
\createspell{\SLmagehand}{Mage Hand}
    {Telekinetically move objects.}
\createspell{\SLmagicmissile}{Magic Missile}
    {Fire unerring missiles to deal damage.}
\createspell{\SLmajorimage}{Major Image}
    {Create figment to fool all senses.}
\createspell{\SLmanipulateprobabilitylesser}{Manipulate Probability}[Lesser]
    {Know your next die results.}
\createspell{\SLmanipulateprobability}{Manipulate Probability}
    {Control creature's next die results.}
\createspell{\SLmaze}{Maze}
    {Trap foe in extradimensional maze.}
\createspell{\SLmeldintoplants}{Meld into Plants}
    {Become one with large plant.}
\createspell{\SLmeldintostone}{Meld into Stone}
    {Become one with large block of stone.}
\createspell{\SLmessage}{Message}
    {Gain ability to whisper conversation at range.}
\createspell{\SLmeteorswarm}{Meteor Swarm}
    {Create huge hail of meteors to deal damage and knock foes prone.}
\createspell{\SLminorimage}{Minor Image}
    {Create figment to fool eyes and ears.}
\createspell{\SLmirrorimage}{Mirror Image}
    {Create decoy duplicates of you.}
\createspell{\SLmirrorimagegreater}{Mirror Image}[Greater]
    {Create regenerating decoy duplicates of you.}
\createspell{\SLmislead}{Mislead}
    {Become invisible and create illusory double.}
\createspell{\SLmissilestorm}{Missile Storm}
    {Create swarm of homing missiles to damage multiple foes.}
\createspell{\SLmissilestormgreater}{Missile Storm}[Greater]
    {Create massive swarm of homing missiles to damage nearby foes.}
\createspell{\SLmomentofpresciencegreater}{Moment of Prescience}[Greater]
    {Immediately gain two legend points.}
\createspell{\SLmomentofprescience}{Moment of Prescience}
    {Immediately gain a legend point.}
\createspell{\SLorderswrath}{Order's Wrath}
    {Damage and immobilize nonlawful creature.}
\createspell{\SLpersistentimage}{Persistent Image}
    {Create scripted figment to fool all senses.}
\createspell{\SLphantasmalkiller}{Phantasmal Killer}
    {Frighten foe, possibly to death.}
\createspell{\SLphantasmalkillermass}{Phantasmal Killer}[Mass]
    {Frighten multiple foes, possibly to death.}
\createspell{\SLplanardisruption}{Planar Disruption}
    {Damage foe, possibly return it to its native plane.}
\createspell{\SLplanardisruptionmass}{Planar Disruption}[Mass]
    {Damage multiple foes, possibly return them to their native planes.}
\createspell{\SLpoison}{Poison}
    {Inflict deadly poison on foe.}
\createspell{\SLpolarray}{Polar Ray}
    {Fire frigid ray to deal damage and slow.}
\createspell{\SLpowerwordblind}{Power Word Blind}
    {Speak word to blind or visually impair foe.}
\createspell{\SLpowerwordconfuse}{Power Word Confuse}
    {Speak word to confuse or disorient foe.}
\createspell{\SLpowerworddaze}{Power Word Daze}
    {Speak word to daze foe.}
\createspell{\SLpowerwordimpair}{Power Word Impair}
    {Speak word to impair foe.}
\createspell{\SLpowerwordstun}{Power Word Stun}
    {Speak word to stun or daze foe.}
\createspell{\SLprecognitiongreater}{Precognition}[Greater]
    {Gain three legend points.}
\createspell{\SLprecognitionlesser}{Precognition}[Lesser]
    {Gain one legend point.}
\createspell{\SLprecognition}{Precognition}
    {Gain two legend points.}
\createspell{\SLprismaticbeam}{Prismatic Beam}
    {Fire multicolored ray with random effects.}
\createspell{\SLprismaticspray}{Prismatic Spray}
    {Fire multicolored cone with random effects.}
\createspell{\SLprismaticstorm}{Prismatic Storm}
    {Create multicolored blast with random effects.}
\createspell{\SLprismaticwall}{Prismatic Wall}
    {Create multicolored wall with random effects.}
\createspell{\SLprohibition}{Prohibition}
    {Punish nearby creatures that take specific action.}
\createspell{\SLprojectimage}{Project Image}
    {Create illusory double that can talk and cast spells.}
\createspell{\SLprotectionfromalignment}{Protection from Alignment}
    {Grant ally damage reduction against aligned foes.}
\createspell{\SLreadmindgreater}{Read Mind}[Greater]
    {Unavoidably learn a creature's surface thoughts.}
\createspell{\SLreadmindmass}{Read Mind}[Mass]
    {Learn surface thoughts from multiple creatures.}
\createspell{\SLreadmind}{Read Mind}
    {Learn a creature's surface thoughts.}
\createspell{\SLreducepersonmass}{Reduce Person}[Mass]
    {Halve size of multiple humanoids.}
\createspell{\SLreduceperson}{Reduce Person}
    {Halve size of humanoid.}
\createspell{\SLregeneration}{Regeneration}
    {Grant automatic healing and immunities}
\createspell{\SLregenerationlesser}{Regeneration}[Lesser]
    {Grant automatic healing each round.}
\createspell{\SLrepulsion}{Repulsion}
    {Prevent creatures from approaching you.}
\createspell{\SLresistenergygreater}{Resist Energy}[Greater]
    {Grant damage reduction against all energy types.}
\createspell{\SLresistenergy}{Resist Energy}
    {Grant damage reduction against energy type.}
\createspell{\SLretributiveshield}{Retributive Shield}
    {Grant damage reduction that damages attackers.}
\createspell{\SLretrieveally}{Retrieve Ally}
    {Teleport ally next to you.}
\createspell{\SLretrieveobjectgreater}{Retrieve Object}[Greater]
    {Teleport object to your hand.}
\createspell{\SLretrieveobject}{Retrieve Object}
    {Teleport unattended object to your hand.}
\createspell{\SLrevelation}{Revelation}
    {Grant target vision of one of three futures.}
\createspell{\SLreversegravity}{Reverse Gravity}
    {Reverse gravity in a zone.}
\createspell{\SLrevivify}{Revivify}
    {Resurrect recently deceased creature.}
\createspell{\SLdivinemight}{Divine Might}
    {Double in size and gain damage reduction.}
\createspell{\SLsanctuary}{Sanctuary}
    {Grant ally immunity to attack until it attacks.}
\createspell{\SLscorchingray}{Scorching Ray}
    {Fire ray of fire to deal damage and ignite.}
\createspell{\SLsearinglight}{Searing Light}
    {Fire ray of light deals 6d6 damage and blinds, more against undead.}
\createspell{\SLseeinvisibility}{See Invisibility}
    {Grant ability to see invisible creatures and objects.}
\createspell{\SLseeinvisibilitymass}{See Invisibility}[Mass]
    {Grant multiple allies ability to see invisible creatures and objects.}
\createspell{\SLshadowpuppet}{Shadow Puppet}
    {Step into Plane of Shadow and create shadow duplicate in your stead.}
\createspell{\SLshadowumbra}{Shadow Umbra}
    {Grant 50\% chance to absorb attacks.}
\createspell{\SLsharepain}{Share Pain}
    {Split damage between two allies.}
\createspell{\SLshieldoffaith}{Shield of Faith}
    {Create floating shield to increase defenses.}
\createspell{\SLshieldoflaw}{Shield of Law}
    {Grant multiple creatures spell resistance and retributive damage against chaotic foes.}
\createspell{\SLshockinggrasp}{Shocking Grasp}
    {Touch foe with electricity to deal damage and stagger.}
\createspell{\SLshout}{Shout}
    {Shout loudly to deal damage and deafen in cone.}
\createspell{\SLshoutgreater}{Shout}[Greater]
    {Shout loudly to deal damage and deafen in large cone.}
\createspell{\SLshrinkitem}{Shrink Item}
    {Shrink object to one-sixteenth size.}
\createspell{\SLsilence}{Silence}
    {Prevent target from making noise.}
\createspell{\SLsilencemass}{Silence}[Mass]
    {Prevent multiple targets from making noise.}
\createspell{\SLsilentimage}{Silent Image}
    {Creates figment to fool eyes.}
\createspell{\SLskysmite}{Skysmite}
    {Call unerring lightning bolt from sky to deal damage.}
\createspell{\SLsleepmass}{Sleep}[Mass]
    {Tire multiple creatures, possibly putting them to sleep.}
\createspell{\SLsleep}{Sleep}
    {Tire a creature, possibly putting it to sleep.}
\createspell{\SLslowmass}{Slow}[Mass]
    {Force multiple foes to skip movement phase.}
\createspell{\SLslow}{Slow}
    {Force foe to skip movement phase.}
\createspell{\SLsolidfog}{Solid Fog}
    {Create fog that blocks vision and slows movement.}
\createspell{\SLsoundburst}{Sound Burst}
    {Create blast of sound to deal damage.}
\createspell{\SLspellresistance}{Spell Resistance}
    {Grant spell resistance.}
\createspell{\SLspelltheftgreater}{Spelltheft}[Greater]
    {Steal magical effects on target.}
\createspell{\SLspelltheft}{Spelltheft}
    {Steal magical effects within area.}
\createspell{\SLspellturning}{Spell Turning}
    {Gain ability to reflect hostile spells.}
\createspell{\SLspiderclimb}{Spider Climb}
    {Grant ability to walk on walls and ceilings.}
\createspell{\SLspiritualweapon}{Spiritual Weapon}
    {Create floating weapon that attacks on its own.}
\createspell{\SLstinkingcloud}{Stinking Cloud}
    {Create fog that blocks sight and nauseates.}
\createspell{\SLstoneskin}{Stoneskin}
    {Grant damage reduction.}
\createspell{\SLstormlord}{Stormlord}
    {Gain whirlwind that blocks ranged attacks and deals retributive damage.}
\createspell{\SLstormofvengeance}{Storm of Vengeance}
    {Storm rains acid, lightning, and hail.}
\createspell{\SLstriptheflesh}{Strip the Flesh}
    {Rend foe's skin from its body.}
\createspell{\SLsuggestionmass}{Suggestion}[Mass]
    {Convince multiple creatures to obey suggestion.}
\createspell{\SLsuggestion}{Suggestion}
    {Convince creature to obey suggestion.}
\createspell{\SLsummonmonsteriii}{Summon Monster III}
    {Call extraplanar creature to fight for you.}
\createspell{\SLsummonmonsterii}{Summon Monster II}
    {Call extraplanar creature to fight for you.}
\createspell{\SLsummonmonsteriv}{Summon Monster IV}
    {Call extraplanar creature to fight for you.}
\createspell{\SLsummonmonsterix}{Summon Monster IX}
    {Call extraplanar creature to fight for you.}
\createspell{\SLsummonmonsteri}{Summon Monster I}
    {Call extraplanar creature to fight for you.}
\createspell{\SLsummonmonsterviii}{Summon Monster VIII}
    {Call extraplanar creature to fight for you.}
\createspell{\SLsummonmonstervii}{Summon Monster VII}
    {Call extraplanar creature to fight for you.}
\createspell{\SLsummonmonstervi}{Summon Monster VI}
    {Call extraplanar creature to fight for you.}
\createspell{\SLsummonmonsterv}{Summon Monster V}
    {Call extraplanar creature to fight for you.}
\createspell{\SLsummonnaturesallyiii}{Summon Nature's Ally III}
    {Call creature to fight.}
\createspell{\SLsummonnaturesallyii}{Summon Nature's Ally II}
    {Call creature to fight for you.}
\createspell{\SLsummonnaturesallyiv}{Summon Nature's Ally IV}
    {Call creature to fight for you.}
\createspell{\SLsummonnaturesallyix}{Summon Nature's Ally IX}
    {Call creature to fight for you.}
\createspell{\SLsummonnaturesallyi}{Summon Nature's Ally I}
    {Call creature to fight for you.}
\createspell{\SLsummonnaturesallyviii}{Summon Nature's Ally VIII}
    {Call creature to fight for you.}
\createspell{\SLsummonnaturesallyvii}{Summon Nature's Ally VII}
    {Call creature to fight for you.}
\createspell{\SLsummonnaturesallyvi}{Summon Nature's Ally VI}
    {Call creature to fight for you.}
\createspell{\SLsummonnaturesallyv}{Summon Nature's Ally V}
    {Call creature to fight for you.}
\createspell{\SLsummonnaturesarmy}{Summon Nature's Army}
    {Call many creatures to fight for you.}
\createspell{\SLsunbeam}{Sunbeam}
    {Create beam of light to deal damage and possibly blind.}
\createspell{\SLsunburst}{Sunburst}
    {Create burst of light to deal damage and possibly blind.}
\createspell{\SLtelepathy}{Telepathy}
    {Grant ability to communicate mentally.}
\createspell{\SLtemporalstasis}{Temporal Stasis}
    {Stop time for creature, possibly permanently.}
\createspell{\SLtimestop}{Time Stop}
    {Gain immense speed to take extra actions.}
\createspell{\SLtransmuteanyobject}{Transmute Any Object}
    {Transform objects or creatures into new forms.}
\createspell{\SLtransmutefleshandstone}{Transmute Flesh and Stone}
    {Transform foe to stone, or restore petrified ally.}
\createspell{\SLtreeshape}{Tree Shape}
    {Transform into a tree.}
\createspell{\SLtremorsense}{Tremorsense}
    {Grant ability to ``see'' through the ground.}
\createspell{\SLtrueseeing}{True Seeing}
    {Grant ability to see through all illusions and transformations.}
\createspell{\SLtruestrike}{True Strike}
    {Quickly grant brief legend point.}
\createspell{\SLunholyaura}{Unholy Aura}
    {Grant multiple creatures spell resistance and retributive damage against good foes.}
\createspell{\SLunholyblight}{Unholy Blight}
    {Damage and stagger nonevil creature.}
\createspell{\SLunlivingeyes}{Unliving Eyes}
    {Grant ability to perfectly see living creatures.}
\createspell{\SLunlivingheart}{Unliving Heart}
    {Gain temporary hit points, become undead.}
\createspell{\SLventriloquism}{Ventriloquism}
    {Gain ability to speak from anywhere.}
\createspell{\SLwailofthebanshee}{Wail of the Banshee}
    {Scream to damage and possibly kill nearby creatures.}
\createspell{\SLwalloffire}{Wall of Fire}
    {Create flaming wall that deals damage.}
\createspell{\SLwalloffiregreater}{Wall of Fire}[Greater]
    {Create huge flaming wall that deals damage.}
\createspell{\SLwallofforce}{Wall of Force}
    {Create force wall that is immune to damage.}
\createspell{\SLwallofthorns}{Wall of Thorns}
    {Create thorny wall that blocks sight and deals damage.}
\createspell{\SLwallofthornsgreater}{Wall of Thorns}[Greater]
    {Create huge thorny wall that blocks sight and deals damage.}
\createspell{\SLwaterwalk}{Water Walk}
    {Grant ability to walk on water.}
\createspell{\SLwaterwalkgreater}{Water Walk}[Greater]
    {Grant multiple allies ability to walk on water.}
\createspell{\SLwavesoffatigue}{Waves of Fatigue}
    {Fatigue creatures in large cone.}
\createspell{\SLwavesoffatiguegreater}{Waves of Fatigue}[Greater]
    {Fatigue or exhaust creatures in large cone.}
\createspell{\SLwavesoffatiguelesser}{Waves of Fatigue}[Lesser]
    {Fatigue creatures in cone.}
\createspell{\SLweb}{Web}
    {Create spiderwebs that immobilize.}
\createspell{\SLwindstrike}{Windstrike}
    {Bludgeon foe with wind to deal damage and shove.}
\createspell{\SLwindstrikegreater}{Windstrike}[Greater]
    {Bludgeon distant foe with wind to deal damage and shove.}
\createspell{\SLwordofchaos}{Word of Chaos}
    {Damage and disorient nearby nonchaotic creatures.}
\createspell{\SLwordofrecall}{Word of Recall}
    {Teleport back to designated place.}
\createspell{\SLzephyrbladegreater}{Zephyr Blade}[Greater]
    {Grant melee weapon ability to strike at range.}
\createspell{\SLzephyrblade}{Zephyr Blade}
    {Melee weapon can strike from a short distance for half damage.}
\createspell{\SLtotemicmindgreater}{Totemic Mind}[Greater]
    {Grant \plus4 to mental attribute}
\createspell{\SLtotemicmindmass}{Totemic Mind}[Mass]
    {Grant multiple allies \plus2 to mental attribute.}
\createspell{\SLtotemicmind}{Totemic Mind}
    {Grant \plus2 to mental attribute.}
\createspell{\SLtotemicpowergreater}{Totemic Power}[Greater]
    {Grant \plus4 to physical attribute.}
\createspell{\SLtotemicpowermass}{Totemic Power}[Mass]
    {Grant multiple allies \plus2 to physical attribute.}
\createspell{\SLtotemicpower}{Totemic Power}
    {Grant \plus2 to physical attribute.}
\createspell{\SLexcreteslime}{Excrete Slime}
    {Excrete slime to deal retributive damage.}
\createspell{\SLrockblast}{Rock Blast}
    {Fire rocks to deal damage.}
\createspell{\SLrottinggrasp}{Rotting Grasp}
    {Rot flesh of adjacent foe to deal lingering damage.}
\createspell{\SLrotburst}{Rotburst}
    {Rot flesh of nearby foes to deal lingering damage.}
\createspell{\SLrotburstgreater}{Rotburst}[Greater]
    {Rot flesh of many nearby foes to deal lingering damage.}
\createspell{\SLfungalgrowth}{Fungal Growth}
    {Grow fungus to sicken and damage foe.}
\createspell{\SLswarmofbats}{Swarm of Bats}
    {Summon bats to impair vision of foes.}
