\createspell{\SLalignweapon}{Align Weapon}
    {Weapon becomes \plus2 and chaotic.}
\createspell{\SLalignweapon}{Align Weapon}
    {Weapon becomes \plus2 and evil.}
\createspell{\SLalignweapon}{Align Weapon}
    {Weapon becomes \plus2 and good.}
\createspell{\SLalignweapon}{Align Weapon}
    {Weapon becomes \plus2 and lawful.}
\createspell{\SLanimalgrowth}{Animal Growth}
    {Enlarges multiple animals.}
\createspell{\SLbanishment}{Banishment}
    {Banish multiple extraplanar creatures. }
\createspell{\SLbestowcurse}{Bestow Curse}
    {\minus6 to an attribute; \minus4 on attack rolls, saves, checks, and AC; or 25\% chance of losing each action.}
\createspell{\SLblindnessdeafness}{Blindness/Deafness}
    {Creature is blinded or deafened.}
\createspell{\SLchangestaff}{Changestaff}
    {Your staff becomes a treant on command.}
\createspell{\SLclenchedfist}{Clenched Fist}
    {Large hand provides cover or attacks your foes.}
\createspell{\SLcloudkill}{Cloudkill}
    {Fog deals Con damage each round.}
\createspell{\SLcombattransformation}{Combat Transformation}
    {You gain combat bonuses.}
\createspell{\SLcontrolwater}{Control Water}
    {Raises or lowers bodies of water.}
\createspell{\SLcreepingdoom}{Creeping Doom}
    {Swarms of centipedes attack at your command.}
\createspell{\SLcrushingdespair}{Crushing Despair}
    {Subjects in medium cone are demoralized.}
\createspell{\SLcrushinghand}{Crushing Hand}
    {Large hand provides cover or crushes your foes.}
\createspell{\SLdaylight}{Daylight}
    {Large radius of bright light.}
\createspell{\SLdismissal}{Dismissal}
    {Forces a creature to return to native plane.}
\createspell{\SLdisruptingweapon}{Disrupting Weapon}
    {Melee weapon destroys undead.}
\createspell{\SLexpeditiousretreat}{Expeditious Retreat}
    {Your speed doubles.}
\createspell{\SLforeseeprobability}{Foresee Probability}
    {Know creature's next die results.}
\createspell{\SLfreezingsphere}{Freezing Sphere}
    {Freezes water or deals cold damage.}
\createspell{\SLgate}{Gate}
    {Connect two planes for travel.}
\createspell{\SLgraspinghand}{Grasping Hand}
    {Hand provides cover, pushes, or grapples.}
\createspell{\SLinsanity}{Insanity}
    {Subject indefinitely acts randomly.}
\createspell{\SLinterposinghand}{Interposing Hand}
    {Hand provides cover against one opponent at a time.}
\createspell{\SLinvestmagic}{Invest Magic}
    {All of subject's equipment is magical.}
\createspell{\SLlifebreakercurse}{Lifebreaker Curse}
    {Target takes permanent damage.}
\createspell{\SLlinkvitality}{Link Vitality}
    {Two targets share each others' damage and healing.}
\createspell{\SLlocatecreaturegreater}{Locate Creature}[Greater]
    {Senses direction toward distant visualized creatures.}
\createspell{\SLlocatecreature}{Locate Creature}
    {Senses direction toward visualized creature.}
\createspell{\SLlocatecreature}{Locate Creature}
    {[Greater] Senses direction toward distant visualized creatures.}
\createspell{\SLlocateentity}{Locate Entity}
    {Locates creatures or objects within 1 mile.}
\createspell{\SLlocateobjectgreater}{Locate Object}[Greater]
    {Senses direction toward distant visualized objects.}
\createspell{\SLlocateobjectlesser}{Locate Object}[Lesser]
    {Senses direction toward visualized object.}
\createspell{\SLlocateobject}{Locate Object}
    {Senses direction toward visualized object.}
\createspell{\SLmagesdisjunction}{Mage's Disjunction}
    {Dispels all magic.}
\createspell{\SLmagiccircleagainstchaos}{Magic Circle against Chaos}
    {Protect allies in a \areamed radius and hedge out opposing outsiders.}
\createspell{\SLmagiccircleagainstevil}{Magic Circle against Evil}
    {Protect allies in a \areamed radius and hedge out opposing outsiders.}
\createspell{\SLmagiccircleagainstgood}{Magic Circle against Good}
    {Protect allies in a \areamed radius and hedge out opposing outsiders.}
\createspell{\SLmagicfanggreater}{Magic Fang}[Greater]
    {All subject's natural weapons get \plus2 bonus.}
\createspell{\SLmagicfang}{Magic Fang}
    {One natural weapon of subject creature becomes \plus2 weapon.}
\createspell{\SLmagicvestment}{Magic Vestment}
    {Armor or shield becomes \plus2.}
\createspell{\SLmagicweapon}{Magic Weapon}
    {Weapon becomes \plus2.}
\createspell{\SLmomentofpresciencelesser}{Moment of Prescience}[Lesser]
    {Roll twice for single attack, save, or opposed check.}
\createspell{\SLobscuringmist}{Obscuring Mist}
    {Fog surrounds you.}
\createspell{\SLphantasmalmaze}{Phantasmal Maze}
    {Subject acts as if affected by \spell{maze}.}
\createspell{\SLphantasmalwound}{Phantasmal Wound}
    {Subject believes it is grievously wounded.}
\createspell{\SLphasedoor}{Phase Door}
    {Creates an invisible passage through wood or stone.}
\createspell{\SLpowerwordkill}{Power Word Kill}
    {Kills one bloodied creature.}
\createspell{\SLprohibitiongreater}{Prohibition}[Greater]
    {Forcibly prevent creatures from taking specific action.}
\createspell{\SLprotectionfromenergygreater}{Protection from Energy}[Greater]
    {As \spell{protection from energy}, but against all energy types.}
\createspell{\SLprotectionfromenergy}{Protection from Energy}
    {Reduce damage from one kind of energy.}
\createspell{\SLrepelmetalorstone}{Repel Metal or Stone}
    {Pushes away metal and stone in large line.}
\createspell{\SLresilientsphere}{Resilient Sphere}
    {Force globe protects but traps one creature or object.}
\createspell{\SLretributivebrilliance}{Retributive Brilliance}
    {Subject can dazzle or blind attacker once.}
\createspell{\SLscintillatingpattern}{Scintillating Pattern}
    {Twisting colors bewilder foes.}
\createspell{\SLseaoffog}{Sea of Fog}
    {200 ft. radius of fog obscures vision.}
\createspell{\SLseaoffog}{Sea of Fog}
    {500 ft. cylinder of fog obscures vision.}
\createspell{\SLshadowbody}{Shadow Body}
    {Transform into living shadow.}
\createspell{\SLshillelagh}{Shillelagh}
    {Cudgel or quarterstaff becomes \plus2 weapon.}
\createspell{\SLstampede}{Stampede}
    {Summon bison to trample foes.}
\createspell{\SLsummonswarm}{Summon Swarm}
    {Summons swarm of bats, rats, or spiders.}
\createspell{\SLtelekinesis}{Telekinesis}
    {Moves object, attacks creature, or hurls object or creature.}
\createspell{\SLtelekineticforce}{Telekinetic Force}
    {Move things with your mind.}
\createspell{\SLtelekineticmaneuver}{Telekinetic Maneuver}
    {Telekinetically dirty trick, disarm, grapple, shove, trip your target.}
\createspell{\SLtelekineticthrust}{Telekinetic Thrust}
    {Hurl objects with the force of your mind.}
\createspell{\SLtouchofidiocy}{Touch of Idiocy}
    {Subject takes \minus4 penalty to Int, Wis, and Cha.}
\createspell{\SLwallofstone}{Wall of Stone}
    {Creates a stone wall that can be shaped.}
\createspell{\SLwavesofexhaustion}{Waves of Exhaustion}
    {Creatures in medium cone become exhausted.}
\createspell{\SLwavesofexhaustion}{Waves of Exhaustion}
    {Creatures in medium cone become exhausted.}
\createspell{\SLweird}{Weird}
    {As \spell{phantasmal killer}, but affects multiple creatures}
