\chapter{Mystic Spheres}\label{Mystic Spheres}

\sectiongraphic*{Spell and Ritual Mechanics}{width=\columnwidth}{core mechanics/spell and ritual mechanics}
  Spells and rituals are common types of \magical abilities with some special rules.
  Every spell and ritual belongs to a thematically related grouping called a \glossterm{mystic sphere}.
  To learn a spell or lead a ritual, you must have access to the mystic sphere that it belongs to.

  \subsubsection{Magic Sources}
    There are four \glossterm{magic sources} that characters can use to cast spells and perform rituals: arcane (cast by sorcerers and wizards), divine (cast by clerics and paladins), nature (cast by druids), and pact (cast by votives).
    Each magic source has a set of associated \glossterm{mystic spheres}.

    \parhead{Characters with Multiple Magic Sources}
    Multiclass characters can have access to multiple magic sources.
    Their \glossterm{mystic spheres}, spells, and rituals are tracked separately for each source of magic they have access to.

  \subsection{Casting Components}\label{Casting Components}
    All rituals require both \glossterm{verbal components} and \glossterm{somatic components} (see \pcref{Ability Usage Components}).
    Unless otherwise noted, all spells require \glossterm{verbal components}.
    In addition, spells from the arcane and pact mystic sources require \glossterm{somatic components}.

  \subsection{Learning Spells}
    When you gain the ability to cast spells, you will also learn some number of spells.
    You choose those spells from among all spells offered by your \glossterm{mystic spheres}, up to the maximum spell rank that you have access to.
    If you know a spell, you can generally cast it as often as you want, unless the spell's description says otherwise.

    You can choose to learn the same spell multiple times.
    This does not allow you to bypass restrictions on repeatedly casting the same spell.
    For example, the \spell{accelerated dodge} spell \glossterm{briefly} prevents you from casting it again.
    That restriction would apply to all copies of the spell that you know.
    However, learning the same spell multiple times can be useful if you have abilities that customize the spell's effects, such as metamagic (see the cleric's Metamagic ability, \pref{Clr:Metamagic}).

  \subsectiongraphic*{Rituals}{width=\columnwidth}{core mechanics/rituals}
    Some characters have the ability to lead rituals.
    Like spells, rituals are \magical abilities that are grouped into \glossterm{mystic spheres}.

      % The fatigue level cost for 24 hour rituals is equal to (ritual rank ^ 2) * 2.
      % Should this be specified explicitly?
    \subsubsection{Ritual Requirements}\label{Ritual Requirements}
      Every ritual has three requirements to perform:
      \begin{raggeditemize}
        \item Fatigue: Every ritual takes at least one \glossterm{fatigue level}, and some require much more.
          Once per minute, the ritual leader can allocate one fatigue level to a creature that participated during that full minute.
          Non-leading ritual participants that are suffering a \glossterm{fatigue penalty} cannot have any additional fatigue levels allocated to them from a ritual.
        \item Leader: The ritual must be led by a spellcasting creature with access to one of the ritual's mystic spheres.
          They must also be able to cast a spell of the ritual's rank or higher.
          Finally, they must have knowledge of the ritual, either because they have it memorized or because they have a ritual book for the ritual.
          If at any time a ritual does not have a participating leader, it ends with no effect.
        \item Time: Every ritual takes at least one minute to perform, and some can take much longer.
      \end{raggeditemize}

    % TODO: how do ritual books interact with Craft (manuscript)?
    \subsubsection{Ritual Books}\label{Ritual Books}
      Unlike spells, you cannot learn additional rituals with \glossterm{insight points}.
      Instead, you must acquire ritual books to learn additional rituals.
      A ritual book contains the instructions for how to perform one or more rituals.
      A typical ritual book is Tiny, and can contain rituals with a maximum combined rank of 10.

      The act of inscribing such potent magical instructions imbues the books with intrinsic magic.
      That magic tends to escape or fade over time, so ritual books made of ordinary ink quickly become unreadable.
      Scribing a ritual book that can last requires expensive magical ink and thick, sturdy pages.
      The cost to scribe a ritual book of a given rank is equal to an item of the ritual's rank (see \pcref{Item Ranks}).
      High rank ritual books tend to be made of exotic materials, like dragonhide or mithral.

    \subsubsection{Ritual Participation}
      While a ritual is being performed, creatures can participate in the ritual.
      Participating in a ritual involves spending a standard action each round, which requires both \glossterm{verbal components} and \glossterm{somatic components}.
      Creatures must be able to speak at least one language to participate in a ritual.
      They must also be able to receive instruction about the appropriate actions to take during the ritual.

      Only the ritual leader needs to know how to perform the ritual, either by having it memorized or with a ritual book.
      Non-leading ritual participants must be given instructions by the leader on the steps they should perform during the ritual.
      Creatures can freely start or stop participating in rituals.
      The ritual can even change leaders, as long as it always has at least one ritual leader.

    \subsubsection{Ritual Mystic Spheres}
      Each ritual belongs to multiple mystic spheres.
      You can only lead rituals from \glossterm{mystic spheres} that you have access to.
      Some rituals have special effects based on which mystic sphere is used to access it.
      If you have access to a ritual through multiple different mystic spheres, you choose which mystic sphere you are using to perform the ritual when the ritual starts.

    \subsubsection{Ritual Scaling}
      Some rituals can be performed at a higher level, such as the \ritual{create water} ritual.
      You do not need to scribe each level of the ritual separately.
      If you know how to perform the normal version of the ritual, you can choose to perform any higher level version if you have access to a spell of that level.

      % \subsubsection{Attunement Rituals}
      %   Rituals with the \abilitytag{Attune} tag require a single ritual participant to \glossterm{attune} to the ritual's effect.
      %   Any ritual participant can attune to the effect, but only one ritual participant can attune to the effect unless otherwise noted in the ritual's description.
      %   For details, see \pcref{Attuned Abilities}.

\section{Magic Sources}\label{Magic Sources}

  \input{generated/mystic_sphere_lists.tex}

  \newpage
\section{Spell Lists}\label{Spell Lists}

  \input{generated/mystic_sphere_spell_summaries.tex}

  \input{generated/mystic_sphere_descriptions.tex}

\section{Ritual Lists}\label{Ritual Lists}

  \input{generated/mystic_sphere_ritual_summaries.tex}

  \input{generated/ritual_descriptions.tex}
